### 2017-05-04 draft/unique-linear.tex
first cut - incomplete

\documentclass{llncs}

\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage[utf8]{inputenc}
\usepackage{mathpartir}

\input{macros}

\begin{document}
Syntax
\begin{align*}
  e &::= x \mid \LAM xe \mid \APP ee \mid \LET xee \mid \REC{ \overline{a=e}} \mid \GET ea \mid \UPD eae \\
  \ell &::= \LIN \mid \UN & \UN & \sqsubset \LIN \\
  u &::= \READ \mid \WRITE & \WRITE & \sqsubset \READ \\
  t &::= b\,\ell\, u \\
  b &::= t \to t \mid \REC{ \overline{a : t}}
\end{align*}
Every type has two attributes, linearity $\ell$ and uniqueness $u$.

An environment $A$ is a partial mapping from variables to types. Environment overriding $A=A_1\OVERRIDE A_2$ is defined pointwise as a ternary  relation. We write $x:t\,-$ to indicate that no binding for $x$ exists. 
\begin{align*}
  x : t\,\UN\,\READ & =  x:t\,\UN\,\READ  \OVERRIDE x:t\,\UN\,\READ \\
  x : t\,\UN\,\WRITE & =  x:t\,\UN\,\WRITE  \OVERRIDE x:t\,- \\
  x : t\,\UN\,\WRITE & =  x:t\,\UN\,\READ  \OVERRIDE x:t\, \UN\,\WRITE \\
  x : t\,\LIN\,u & =  x:t\,\LIN\,u  \OVERRIDE x:t\,- \\
  x : t\,\LIN\,u & =  x:t\,-  \OVERRIDE x:t\, \LIN\,u \\
\end{align*}
For $n>2$ define $A= A_1 \OVERRIDE A_2 \OVERRIDE \dots \OVERRIDE A_n$ by $A= A_1 \OVERRIDE A_1'$ and $A_1' = A_2 \OVERRIDE \dots \OVERRIDE A_n$.

If any of the record components is linear, then the record must be linear, too.
If any of the components has write premission, then the record must have it, too.
\begin{mathpar}
  \inferrule{
    A = A_1 \OVERRIDE \dots \OVERRIDE A_n \\
    A_i \vdash e_i : t_i\,\ell_i\,u_i \\
    \ell_i \sqsubseteq \ell \\
    u \sqsubseteq u_i
  }{
    A \vdash \REC{ \overline{a_i=e_i}} : \REC{ \overline{a_i : t_i}}\,\ell\, u
  }
\end{mathpar}
To read from a record, the linearity or uniqueness of the record do not matter. However, the fields that are dropped must be unrestricted. The attributes of the field's type remain as they are.
\begin{mathpar}
  \inferrule{
    A \vdash e : \REC{ a:t \mid r}\,\ell\,u \\
    \UN (r)
  }{
    A \vdash \GET ea : t
  }
\end{mathpar}
To write to a record, we need write permission. The type of the overwritten field must be unrestricted because the old contents is dropped. (It would be good to have an update operation that works on linear fields.)
\begin{mathpar}
  \inferrule{
    A_0 \vdash e_0 : \REC{ a:t \mid r}\,\ell\,\WRITE \\
    A_1 \vdash e_1 : t \\
    \UN (t) \\
    A = A_0 \OVERRIDE A_1
  }{
    A \vdash \UPD {e_0}a{e_1} : \REC{ a:t \mid r}\,\ell\,u
  }
\end{mathpar}

\begin{mathpar}
  \inferrule{
  }{
    A, x: t \vdash x:t
  }
\end{mathpar}
Uniqueness does not matter for a function. However, a function that captures a linear variable or a variable with write permission must be linear. A function that captures a record with read permission can be influenced by subsequent writes to that record. Thus the following combinations are possible.
\begin{itemize}
\item $A$ is readonly, which means that no future writes through $A$ are possible. Currently, we have no notion of readonly.
\item $A$ has write permissions and $\ell = \LIN$, so that writes are only executed at most once.
\item $A$ contains linear stuff and $\ell = \LIN$, so that the linear stuff is used exactly once.
\end{itemize}
\begin{mathpar}
  \inferrule{
    A, x:t' \vdash e : t
  }{
    A \vdash \LAM x e : (t' \to t)\,\ell\,u
  }
\end{mathpar}
Function application has no particular restrictions, it seems.
\begin{mathpar}
  \inferrule{
    A_0 \vdash e_0 : (t_1 \to t_2)\, \ell\,u \\
    A_1 \vdash e_1 : t_1 \\
    A = A_0 \OVERRIDE A_1
  }{
    A \vdash \APP{e_0}e_1 :
  }
\end{mathpar}
\end{document}

### 2017-05-05 draft/unique-linear.tex
small fixes and clarifications

@@
 An environment $A$ is a partial mapping from variables to types.
-Environment overriding $A=A_1\OVERRIDE A_2$ is defined pointwise as a ternary  relation.
+Environment overriding $A=A'\OVERRIDE A''$ is defined as the pointwise extension of a ternary relation on types.
 We write $x:t\,-$ to indicate that no binding for $x$ exists.
-\begin{align*}
-  x : t\,\UN\,\READ & =  x:t\,\UN\,\READ  \OVERRIDE x:t\,\UN\,\READ \\
-  x : t\,\UN\,\WRITE & =  x:t\,\UN\,\WRITE  \OVERRIDE x:t\,- \\
-  x : t\,\UN\,\WRITE & =  x:t\,\UN\,\READ  \OVERRIDE x:t\, \UN\,\WRITE \\
-  x : t\,\LIN\,u & =  x:t\,\LIN\,u  \OVERRIDE x:t\,- \\
-  x : t\,\LIN\,u & =  x:t\,-  \OVERRIDE x:t\, \LIN\,u \\
-\end{align*}
+\begin{mathpar}
+  \inferrule{}{b\,\UN\,\READ  =  b\,\UN\,\READ   \OVERRIDE' b\,\UN\,\READ}
+
+  \inferrule{}{b\,\UN\,\WRITE  =  b\,\UN\,\WRITE  \OVERRIDE b\,-}
+  \\
+  \inferrule{b = b' \OVERRIDE' b''}
+  {b\,\UN\,\WRITE  =  b'\,\UN\,\READ   \OVERRIDE' b''\, \UN\,\WRITE}
+
+  \inferrule{t = t' \OVERRIDE' t''}{t = t' \OVERRIDE t''}
+  \\
+  \inferrule{}{t\,\LIN\,u           =  t\,\LIN\,u           \OVERRIDE t\,-}
+
+  \inferrule{}{t\,\LIN\,u           =  t\,-                     \OVERRIDE t\, \LIN\,u}
+  \\
+  \inferrule{
+    t_i = t_i' \OVERRIDE' t_i''
+  }{
+    \REC{ a_1:t_1, \dots } =     \REC{ a_1:t_1', \dots } \OVERRIDE'     \REC{ a_1:t_1'', \dots }
+  }
+
+  \inferrule{}{t_1\to t_2 = t_1 \to t_2 \OVERRIDE' t_1 \to t_2}
+\end{mathpar}
 For $n>2$ define $A= A_1 \OVERRIDE A_2 \OVERRIDE \dots \OVERRIDE A_n$ by $A= A_1 \OVERRIDE A_1'$ and $A_1' = A_2 \OVERRIDE \dots \OVERRIDE A_n$.
 
-If any of the record components is linear, then the record must be linear, too.
+\newpage
+If any record component is linear, then the record must be linear, too.
 If any of the components has write premission, then the record must have it, too.
@@
 \end{mathpar}
-To read from a record, the linearity or uniqueness of the record do not matter. However, the fields that are dropped must be unrestricted. The attributes of the field's type remain as they are.
+Reading from a record is a use of the record: If the record is linear, then there can be only one use.
+The fields that are dropped  must not be linear.
+Thus, a linear record can have at most a single linear field and no other field can be accessed.
+(That seems overly restrictive, but could be amended by a record elimination that reads more than one field at a time.)
+
+Uniqueness of the record does not matter for reading.
+The attributes of the field's type remain as they are, if the record has write permission.
+(If the record has only read permission, then none of the fields can have write permission, so the attributes remain the same, too.)
 \begin{mathpar}
@@
 \begin{itemize}
-\item $A$ is readonly, which means that no future writes through $A$ are possible. Currently, we have no notion of readonly.
-\item $A$ has write permissions and $\ell = \LIN$, so that writes are only executed at most once.
-\item $A$ contains linear stuff and $\ell = \LIN$, so that the linear stuff is used exactly once.
+\item $y:t \in A$ is readonly, which means that no future writes through $y$ are possible. Readonly variables impose no restriction on the function type, but currently we have no notion of readonly.
+\item $y:t \in A$ needs write permission to typecheck $e$.
+This requires $\ell = \LIN$, so that writes are only executed at  most once.
+If $\ell \ne\LIN$, then $y$ needs to be copied each time the closure is invoked.
+\item $y:t \in A$ is linear requires that $\ell = \LIN$, so that the linear stuff is used exactly once.
 \end{itemize}
@@
   }{
-    A \vdash \APP{e_0}e_1 :
+    A \vdash \APP{e_0}e_1 : t_2
   }

### 2017-05-27 draft/unique-linear.tex
sequential typing

@@
 \begin{document}
+\section{First Attempt: Uniqueness and Linearity}
+\label{sec:first-attempt:-uniq}
+
 Syntax
 \begin{align*}
-  e &::= x \mid \LAM xe \mid \APP ee \mid \LET xee \mid \REC{ \overline{a=e}} \mid \GET ea \mid \UPD eae \\
+  e &::= x \mid \LAM xte \mid \APP ee \mid \LET xee \mid \REC{ \overline{a=e}} \mid \GET ea \mid \UPD eae \\
   \ell &::= \LIN \mid \UN & \UN & \sqsubset \LIN \\
@@
   \inferrule{
-    A, x:t' \vdash e : t
+    A, x:t \vdash e : t'
   }{
-    A \vdash \LAM x e : (t' \to t)\,\ell\,u
+    A \vdash \LAM x t e : (t \to t')\,\ell\,u
   }
@@
 \end{mathpar}
+
+\clearpage{}
+
+\section{Second Attempt: Uniqueness Only}
+\label{sec:second-attempt:-uniq}
+
+Syntax
+\begin{align*}
+  e &::= x \mid \LAM xte \mid \APP ee \mid \LET xee  \mid c \mid e+e \mid \IF eee\\
+  &\mid \REC{ \overline{a=e}} \mid \GET ea \mid \UPD[u] eae \\
+  u,v &::= \READ \mid \WRITE & \WRITE & \sqsubset \READ \\
+  s,t &::= \NUM \mid s \to t \mid \REC[u]{ \overline{a : t}}
+\end{align*}
+Only record types have a uniqueness attribute $u$.
+The $u$ attribute on the update operation determines whether it is destructive ($u=\WRITE$) or copying ($u=\READ$).
+
+\begin{mathpar}
+  \inferrule{
+    A_2 \vdash e_0 : \REC[u]{ a_1:t \mid r} \dashv A_1 \\
+    A_1 \vdash e_1 : t \dashv A_0 
+  }{
+    A_2 \vdash
+    \UPD[u] {e_0}{a_1}{e_1} : \REC[v]{ a_1:t \mid r}
+    \dashv A_0
+  }
+
+  \inferrule{
+    A_1 \vdash e : \REC[\READ]{ a:t \mid r}
+    \dashv A_0
+  }{
+    A_1 \vdash \GET ea : t
+    \dashv A_0
+  }
+
+  \inferrule{
+    A_{i+1} \vdash e_i : t_i \dashv A_i \\
+  }{
+    A_{n+1} \vdash \REC{ \overline{a_i=e_i}} : \REC[u]{ \overline{a_i : t_i}}
+    \dashv A_0
+  }
+
+  \inferrule{
+    t_0 = t_1 \OVERRIDE t
+  }{
+    A, x: t_1 \vdash x : t \dashv A, x:t_0
+  }
+
+    \inferrule{
+    A_2 \vdash e_0 : (t_1 \to t_2) \dashv A_1 \\
+    A_1 \vdash e_1 : t_1 \dashv A_0
+  }{
+    A_2 \vdash
+    \APP{e_0}e_1 : t_2
+    \dashv A_0
+  }
+
+    \inferrule{
+      A_1, x:t_1 \vdash
+      e : t
+      \dashv A_0^{X:\READ}, x:t_0 \\
+      X = \FREE (\LAM x {t_0} e)
+  }{
+    A_0^{X:\READ} \vdash
+    \LAM x {t_0} e : (t_0 \to t)
+    \dashv A_0
+  }
+
+  \inferrule{
+    A_4 \vdash e_0 : \NUM \dashv A_3 \\
+    A_1 \vdash e_1 : t_1 \dashv A_0 \\
+    A_2 \vdash e_2 : t_2 \dashv A_0 \\
+    t_1, t_2 \sqsubseteq t_3 \\
+    A_1, A_2 \sqsubseteq A_3
+  }{
+    A_4 \vdash \IF {e_0}{e_1}{e_2} : t_3 \dashv A_0
+  }
+\end{mathpar}
+
+Operationally, the lambda takes a deep copy of all free variables.
+Thus, it must not propagate write permission to the left.
+As the lambda may be invoked multiple times, the free variables must
+not have write permission.
+The notation $A^{X:\READ}$ means that all permissions in the types of the
+variables in $X$ are set to $\READ$.
 \end{document}

### 2017-06-23 draft/unique-linear.tex
fine granular record permissions, take #1

@@
 variables in $X$ are set to $\READ$.
+
+\clearpage{}
+
+\section{Third Attempt: Field Effects}
+\label{sec:third-attempt:-field}
+
+\begin{mathpar}
+  \inferrule{
+    A = A_1 \OVERRIDE \dots \OVERRIDE A_n \\
+    A_i \vdash e_i : t_i 
+  }{
+    A \vdash \REC{ \overline{a_i=e_i}} : \REC{ \overline{a_i : t_i}}
+  }
+\end{mathpar}
+
+\begin{mathpar}
+  \inferrule
+  {
+    A \vdash e : \REC{ a:^\READ t \mid r}
+    % \\ \READ \in P
+  }
+  {
+    A \vdash \GET ea : t
+  }
+\end{mathpar}
+
+\begin{mathpar}
+  \inferrule{
+    A_0 \vdash e_0 : \REC{ a:^\WRITE t \mid r} \\
+    A_1 \vdash e_1 : t \\
+    % \WRITE \in P \\
+    A = A_0 \OVERRIDE A_1
+  }{
+    A \vdash \UPD {e_0}a{e_1} : \REC{ a:t \mid r}
+  }
+\end{mathpar}
+
+\begin{mathpar}
+  \inferrule{
+  }{
+    A, x: t \vdash x:t
+  }
+\end{mathpar}
+
+Look at combining $a :^O t \OVERRIDE a:^P t$ upwards to $a : ^Q t$
+\begin{itemize}
+\item $O\subseteq \{\READ\}$: $Q = O \cup P$,
+  no further action required as potential writing in $P$ happens after reading
+\item $\WRITE \in O$, $P \subseteq \{\READ\}$: two choices, not clear
+  which is better
+  \begin{itemize}
+  \item $Q = \READ$ insert copy, propagate left to the write operation
+  \item $Q = \WRITE$ insert copy, propagate right to the read operation
+  \end{itemize}
+\item $\WRITE\in O$, $\WRITE \in P$: $Q = \WRITE$ insert copy,
+  propagate left (or right)
+\end{itemize}
+
+\begin{mathpar}
+  \inferrule
+  { A \vdash e_1 : t_1 \\
+    A, x : t_1 \vdash e_2 : t_2
+  }
+  { A \vdash \LET x{e_1} e_2 : t}
+\end{mathpar}
+
 \end{document}

### 2017-06-23 draft/unique-linear.tex
monomorphic effects completed

--- 2017-05-27 draft/unique-linear.tex
@@
 variables in $X$ are set to $\READ$.
+
+\clearpage{}
+
+\section{Third Attempt: Field Effects}
+\label{sec:third-attempt:-field}
+
+\begin{mathpar}
+  \inferrule{
+    A = A_1 \OVERRIDE \dots \OVERRIDE A_n \\
+    A_i \vdash e_i : t_i 
+  }{
+    A \vdash \REC{ \overline{a_i=e_i}} : \REC{ \overline{a_i :^{P_i} t_i}}
+  }
+
+  \inferrule
+  {
+    A \vdash e : \REC{ a:^\READ t \mid r}
+    % \\ \READ \in P
+  }
+  {
+    A \vdash \GET ea : t
+  }
+
+  \inferrule{
+    A_0 \vdash e_0 : \REC{ a:^\WRITE t \mid r} \\
+    A_1 \vdash e_1 : t \\
+    % \WRITE \in P \\
+    A = A_0 \OVERRIDE A_1
+  }{
+    A \vdash \UPD {e_0}a{e_1} : \REC{ a:^P t \mid r}
+  }
+  \\
+  \inferrule{
+  }{
+    A, x: t \vdash x: t
+  }
+
+  \inferrule{
+    A_0, x:t \vdash e : t' \\
+    A = A_0^*
+  }{
+    A \vdash \LAM x t e : (t \to t')
+  }
+\end{mathpar}
+Function closures have ``infinite'' splitting $(a :^P t)^*$ into $(a :^Q t)$.
+\begin{itemize}
+\item $P \le \{\READ\}$ propagate $Q = \READ$, no action required
+\item $\WRITE \in P$: propagate $Q = \READ$, insert copy into closure
+\end{itemize}
+
+Function application has no particular restrictions, it seems.
+\begin{mathpar}
+  \inferrule{
+    A_0 \vdash e_0 : (t_1 \to t_2) \\
+    A_1 \vdash e_1 : t_1 \\
+    A = A_0 \OVERRIDE A_1
+  }{
+    A \vdash \APP{e_0}e_1 : t_2
+  }
+\end{mathpar}
+
+Look at combining $a :^O t \OVERRIDE a:^P t$ upwards to $a : ^Q t$
+\begin{itemize}
+\item if $O=\emptyset$ or $P=\emptyset$ (field only used in one branch), then $Q = O \cup P$, no
+  action required
+\item if $O\subseteq \{\READ\}$: $Q = O \cup P$,
+  no further action required as potential writing in $P$ happens after reading
+\item if $\WRITE \in O$, $P = \{\READ\}$: two choices, not clear
+  which is better
+  \begin{itemize}
+  \item $Q = \READ$ insert copy, propagate left to the write operation
+  \item $Q = \WRITE$ insert copy, propagate right to the read operation
+  \end{itemize}
+\item if $\WRITE\in O$, $\WRITE \in P$: $Q = \WRITE$ insert copy,
+  propagate left (or right)
+\end{itemize}
+
+\begin{mathpar}
+  \inferrule
+  { A_1 \vdash e_1 : t_1 \\
+    A_2, x : t_1 \vdash e_2 : t_2 \\
+    A = A_1 \OVERRIDE A_2
+  }
+  { A \vdash \LET x{e_1} e_2 : t}
+\end{mathpar}
+
 \end{document}

### 2017-06-30 draft/unique-linear.tex
snapshot

@@
 
+Effects attached to single record labels
+\begin{align*}
+  P, Q &\subseteq \{ \READ, \WRITE \}
+\end{align*}
+
+\begin{figure}[tp]
   \begin{mathpar}
@@
       A = A_1 \OVERRIDE \dots \OVERRIDE A_n \\
-    A_i \vdash e_i : t_i 
-  }{
-    A \vdash \REC{ \overline{a_i=e_i}} : \REC{ \overline{a_i :^{P_i} t_i}}
-  }
+      A_i \vdash e_i : t_i }{ A \vdash \REC{ \overline{a_i=e_i}} :
+      \REC{ \overline{a_i :^{P_i} t_i}} }
 
-  \inferrule
-  {
-    A \vdash e : \REC{ a:^\READ t \mid r}
+    \inferrule { A \vdash e : \REC{ a:^\READ t \mid r}
       % \\ \READ \in P
-  }
-  {
-    A \vdash \GET ea : t
-  }
+    } { A \vdash \GET ea : t }
 
@@
       % \WRITE \in P \\
-    A = A_0 \OVERRIDE A_1
-  }{
-    A \vdash \UPD {e_0}a{e_1} : \REC{ a:^P t \mid r}
-  }
+      A = A_0 \OVERRIDE A_1 }{ A \vdash \UPD {e_0}a{e_1} : \REC{ a:^P
+        t \mid r} }
     \\
-  \inferrule{
-  }{
-    A, x: t \vdash x: t
-  }
+    \inferrule{ }{ A, x: t \vdash x: t }
 
@@
       A_0, x:t \vdash e : t' \\
-    A = A_0^*
-  }{
-    A \vdash \LAM x t e : (t \to t')
-  }
-\end{mathpar}
-Function closures have ``infinite'' splitting $(a :^P t)^*$ into $(a :^Q t)$.
-\begin{itemize}
-\item $P \le \{\READ\}$ propagate $Q = \READ$, no action required
-\item $\WRITE \in P$: propagate $Q = \READ$, insert copy into closure
-\end{itemize}
+      A = A_0^* }{ A \vdash \LAM x t e : (t \to t') }
 
-Function application has no particular restrictions, it seems.
-\begin{mathpar}
   \inferrule{
@@
   }
+
+  \inferrule
+  { A_1 \vdash e_1 : t_1 \\
+    A_2, x : t_1 \vdash e_2 : t_2 \\
+    A = A_1 \OVERRIDE A_2
+  }
+  { A \vdash \LET x{e_1} e_2 : t}
   \end{mathpar}
+  \caption{Propagation rules}
+  \label{fig:field-effects}
+\end{figure}
+Function closures have ``infinite'' splitting $(a :^P t)^*$ into $(a :^Q t)$.
+\begin{itemize}
+\item $P \subseteq \{\READ\}$ propagate $Q = \READ$, no action required
+\item $\WRITE \in P$: propagate $Q = \READ$, insert copy into closure
+\end{itemize}
 
-Look at combining $a :^O t \OVERRIDE a:^P t$ upwards to $a : ^Q t$
+Function application has no particular restrictions, it seems.
+Let improves on lambda combined with app because infinite splitting is avoided.
+
+Look at combining $a :^O t \OVERRIDE a:^P t$ upwards to $a : ^Q t$:
+$Q= O \OVERRIDE P$
 \begin{itemize}
@@
 
+Combining environments: pointwise
+
+Combining records
 \begin{mathpar}
   \inferrule
-  { A_1 \vdash e_1 : t_1 \\
-    A_2, x : t_1 \vdash e_2 : t_2 \\
-    A = A_1 \OVERRIDE A_2
+  {}
+  {
+    \REC{ {a :^{Q_a} t_a}, {b :^{Q_b} t_b}}
+    =
+    \REC{ {a :^{O_a} t_a}, {b :^{O_b} t_b}}
+    \OVERRIDE
+    \REC{ {a :^{P_a} t_a}, {b :^{P_b} t_b}}
   }
-  { A \vdash \LET x{e_1} e_2 : t}
+\end{mathpar}
+
+\clearpage
+\begin{mathpar}
+  \begin{array}{|c|c||c|c||c|c|l}
+    Q_a&Q_b&O_a&O_b&P_a&P_b& \\
+    \hline
+    P_a &\emptyset& \emptyset & \emptyset & P_a & \emptyset & \\
+    O_a & \emptyset& O_a & \emptyset & \emptyset & \emptyset & \\
+    P_a\cup\READ   &\emptyset& \subseteq\READ & \emptyset & P_a & \emptyset & \\
+    \READ&\READ& \WRITE\in & \emptyset & \READ & \emptyset
+                       &  \text{copy, propagate left to write}\\
+    \WRITE &\READ& \WRITE\in & \emptyset & \WRITE\in & \emptyset
+                       &\text{copy, propagate left to write} \\
+  \end{array}
+
+  \begin{array}{|c|c||c|c||c|c|l}
+    Q_a&Q_b&O_a&O_b&P_a&P_b& \\
+    \hline
+    P_a & P_b & \emptyset & \emptyset & P_a & P_b & \\
+    O_a & P_b & O_a & \emptyset & \emptyset & P_b & \\
+    P_a\cup\READ   &P_b& \subseteq\READ & \emptyset & P_a & P_b & \\
+    \READ&P_b\cup\READ& \WRITE\in & \emptyset & \READ & P_b
+                       &  \text{copy, propagate left to write}\\
+    \WRITE & P_b\cup\READ & \WRITE\in & \emptyset & \WRITE\in & P_b
+                       &\text{copy, propagate left to write} \\
+  \end{array}
+
+  \begin{array}{|c|c||c|c||c|c|l}
+    Q_a&Q_b&O_a&O_b&P_a&P_b& \\
+    \hline
+    P_a & O_b & \emptyset & O_b & P_a & \emptyset & \\
+    O_a & O_b & O_a & O_b & \emptyset & \emptyset & \\
+    P_a\cup\READ  & O_b & \subseteq\READ & O_b & P_a & \emptyset & \\
+    \READ&O_b \cup\READ& \WRITE\in & O_b & \READ & \emptyset
+                       &  \text{copy, propagate left to write}\\
+    \WRITE &O_b\cup\READ& \WRITE\in & O_b & \WRITE\in & \emptyset
+                           &\text{copy, propagate left to write} \\
+  \end{array}
+
+\begin{array}{|c|c||c|c||c|c|l}
+    Q_a&Q_b&O_a&O_b&P_a&P_b& \\
+    \hline
+    P_a &\emptyset& \emptyset & \emptyset & P_a & \emptyset & \\
+    O_a & \emptyset& O_a & \emptyset & \emptyset & \emptyset & \\
+    P_a\cup\READ   &\emptyset& \subseteq\READ & \emptyset & P_a & \emptyset & \\
+    \READ&\READ& \WRITE\in & \emptyset & \READ & \emptyset
+                       &  \text{copy, propagate left to write}\\
+    \WRITE &\READ& \WRITE\in & \emptyset & \WRITE\in & \emptyset
+                       &\text{copy, propagate left to write} \\
+  \end{array}
 \end{mathpar}

### 2017-06-30 draft/unique-linear.tex
snapshot

@@
     O_a & \emptyset& O_a & \emptyset & \emptyset & \emptyset & \\
-    P_a\cup\READ   &\emptyset& \subseteq\READ & \emptyset & P_a & \emptyset & \\
+    P_a\cup\READ   &\emptyset& \READ & \emptyset & P_a & \emptyset & \\
     \READ&\READ& \WRITE\in & \emptyset & \READ & \emptyset
@@
     O_a & P_b & O_a & \emptyset & \emptyset & P_b & \\
-    P_a\cup\READ   &P_b& \subseteq\READ & \emptyset & P_a & P_b & \\
+    P_a\cup\READ   &P_b& \READ & \emptyset & P_a & P_b & \\
     \READ&P_b\cup\READ& \WRITE\in & \emptyset & \READ & P_b
@@
     O_a & O_b & O_a & O_b & \emptyset & \emptyset & \\
-    P_a\cup\READ  & O_b & \subseteq\READ & O_b & P_a & \emptyset & \\
+    P_a\cup\READ  & O_b & \READ & O_b & P_a & \emptyset & \\
     \READ&O_b \cup\READ& \WRITE\in & O_b & \READ & \emptyset
@@
     \hline
-    P_a &\emptyset& \emptyset & \emptyset & P_a & \emptyset & \\
-    O_a & \emptyset& O_a & \emptyset & \emptyset & \emptyset & \\
-    P_a\cup\READ   &\emptyset& \subseteq\READ & \emptyset & P_a & \emptyset & \\
-    \READ&\READ& \WRITE\in & \emptyset & \READ & \emptyset
+    P_a &P_b\cup\READ& \emptyset & \READ & P_a & P_b & \\
+    O_a & P_b\cup\READ& O_a & \READ & \emptyset & P_b & \\
+    P_a\cup\READ   &P_b\cup\READ& \READ & \READ & P_a & P_b & \\
+    \READ&P_b\cup\READ& \WRITE\in & \READ & \READ & P_b
                        &  \text{copy, propagate left to write}\\
-    \WRITE &\READ& \WRITE\in & \emptyset & \WRITE\in & \emptyset
+    \WRITE &P_b\cup\READ& \WRITE\in & \READ & \WRITE\in & P_b
+                       &\text{copy, propagate left to write} \\
+  \end{array}
+
+  \begin{array}{|c|c||c|c||c|c|l}
+    Q_a&Q_b&O_a&O_b&P_a&P_b& \\
+    \hline
+    P_a\cup\READ &\READ& \emptyset & \WRITE\in & P_a & \READ
+                           & \text{copy, propagate left to write} \\
+    \READ & \READ& O_a & \WRITE\in & \emptyset & \READ
+                           & \text{copy, propagate left to write} \\
+    P_a\cup\READ   &\READ& \READ & \WRITE\in & P_a & \READ
+                           &  \text{copy, propagate left to write}\\
+    \READ&\READ& \WRITE\in & \WRITE\in & \READ & \READ
+                       &  \text{copy, propagate left to write}\\
+    \WRITE &\READ& \WRITE\in & \WRITE\in & \WRITE\in & \READ
+                       &\text{copy, propagate left to write} \\
+  \end{array}
+
+  \begin{array}{|c|c||c|c||c|c|l}
+    Q_a&Q_b&O_a&O_b&P_a&P_b& \\
+    \hline
+    P_a\cup\READ &\WRITE& \emptyset & \WRITE\in & P_a & \WRITE\in
+                           & \text{copy, propagate left to write} \\
+    \READ & \WRITE& O_a & \WRITE\in & \emptyset & \WRITE\in
+                           & \text{copy, propagate left to write} \\
+    P_a\cup\READ   &\WRITE& \READ & \WRITE\in & P_a & \WRITE\in
+                           &  \text{copy, propagate left to write}\\
+    \READ&\WRITE& \WRITE\in & \WRITE\in & \READ & \WRITE\in
+                       &  \text{copy, propagate left to write}\\
+    \WRITE &\WRITE& \WRITE\in & \WRITE\in & \WRITE\in & \WRITE\in
                        &\text{copy, propagate left to write} \\
@@
 \end{mathpar}
+A copy operation reads \textbf{every} field. Hence, a copy operation
+caused by one field implies a read effect on all fields.
 

### 2017-06-30 draft/unique-linear.tex
effect propagation

--- 2017-06-23 draft/unique-linear.tex
@@
 
+Effects attached to single record labels, intuitively $P, Q \subseteq
+\{ \READ, \WRITE \}$ but it does not make sense to distinguish
+$\{\WRITE, \READ\}$ from $\{\WRITE\}$, so we take the three element lattice
+\begin{align*}
+  O, P, Q &::= \NOTHING \mid \READ \mid \WRITE
+  & \text{ordered by } \NOTHING
+  &\sqsubset \READ \sqsubset \WRITE
+\end{align*}
+
+\begin{figure}[tp]
   \begin{mathpar}
@@
   }
-\end{mathpar}
 
-Look at combining $a :^O t \OVERRIDE a:^P t$ upwards to $a : ^Q t$
-\begin{itemize}
-\item if $O=\emptyset$ or $P=\emptyset$ (field only used in one branch), then $Q = O \cup P$, no
-  action required
-\item if $O\subseteq \{\READ\}$: $Q = O \cup P$,
-  no further action required as potential writing in $P$ happens after reading
-\item if $\WRITE \in O$, $P = \{\READ\}$: two choices, not clear
-  which is better
-  \begin{itemize}
-  \item $Q = \READ$ insert copy, propagate left to the write operation
-  \item $Q = \WRITE$ insert copy, propagate right to the read operation
-  \end{itemize}
-\item if $\WRITE\in O$, $\WRITE \in P$: $Q = \WRITE$ insert copy,
-  propagate left (or right)
-\end{itemize}
-
-\begin{mathpar}
   \inferrule
@@
   \end{mathpar}
+  \caption{Propagation rules}
+  \label{fig:field-effects}
+\end{figure}
+Function closures have ``infinite'' splitting $(a :^P t)^*$ into $(a :^Q t)$.
+\begin{itemize}
+\item $P = \NOTHING$ propagate $Q=\NOTHING$.
+\item $P = \READ$ propagate $Q = \WRITE$, no action required inside closure.
+\item $P = \WRITE$ propagate $Q = \WRITE$, perform copy inside of closure.
+\end{itemize}
+ Propagating $\READ$ is not safe in both cases because that enables a write operation on the same
+ record after the closure has been constructed. Alternatively, we could have another READONLY mode,
+ which prevents subsequent write operations.
+
+Function application has no particular restrictions, it seems.
+Let improves on lambda combined with app because infinite splitting is avoided.
+
+Look at combining $a :^O t \OVERRIDE a:^P t$ upwards to $a : ^Q t$:
+
+$Q= O \OVERRIDE P$ is defined by:
+\begin{itemize}
+\item if $O=\NOTHING$ or $P=\NOTHING$ (field only used in one branch), then $Q = O \sqcup P$, no
+  action required
+\item if $O\sqsubseteq \READ$: $Q = O \sqcup P$,
+  no further action required as potential writing in $P$ happens after reading
+\item if $O = \WRITE$, $P = \READ$, then there are two choices
+  \begin{itemize}
+  \item $Q = \READ$ insert copy, propagate left to the write operation
+  \item $Q = \WRITE$ insert copy, propagate right to the read operation
+  \end{itemize}
+  First option is preferred because it keeps writes local and prevents
+  further copies downstream.
+\item if $O=\WRITE$, $P =\WRITE$: $Q = \WRITE$ insert copy,
+  propagate left (or right)
+\end{itemize}
+We need an operation that returns $\READ$ if a copy is required and
+$\NOTHING$ otherwise.
+\begin{align*}
+  \COPY (O, P) &=
+                 \begin{cases}
+                   \READ & O=\WRITE \wedge \READ \sqsubseteq P \\
+                   \NOTHING & \text{otherwise}
+                 \end{cases}
+  \\
+  \COPY (\overline{O}, \overline{P})
+               &= \bigsqcup_i \COPY (O_i, P_i)
+  \\
+  \COPY (\NUM, \NUM)
+               & = \NOTHING
+  \\
+  \COPY (t_1 \to t_2, t_1 \to t_2)
+               & = \NOTHING
+  \\
+  \COPY (\REC{ \overline{a_i :^{O_i} t_i'}}, \REC{ \overline{a_i :^{P_i} t_i''}})
+  &=\COPY (\overline{O_i}, \overline{P_i})
+    \sqcup \bigsqcup_i \overline{\COPY (t_i', t_i'')}
+\end{align*}
+
+Combining environments: pointwise
+
+Combining types and records
+\begin{mathpar}
+  \inferrule{}{ \NUM = \NUM \OVERRIDE \NUM}
+
+  \inferrule{}{t_1 \to t_2 = t_1 \to t_2 \OVERRIDE t_1 \to t_2}
+  \\
+  \inferrule{
+    Q_i = (O_i \OVERRIDE P_i)
+    \sqcup \COPY (\overline{O_i}, \overline{P_i})
+    \sqcup \bigsqcup_i \overline{\COPY (t_i', t_i'')} \\
+    t_i = t_i' \OVERRIDE t_i''
+  }{
+    \REC{ \overline{a_i :^{Q_i} t_i}}
+    =
+    \REC{ \overline{a_i :^{O_i} t_i'}}
+    \OVERRIDE
+    \REC{ \overline{a_i :^{P_i} t_i''}}
+  }
+\end{mathpar}
+
+\begin{mathpar}
+  \inferrule
+  {}
+  {
+    \REC{ {a :^{Q_a} t_a}, {b :^{Q_b} t_b}}
+    =
+    \REC{ {a :^{O_a} t_a}, {b :^{O_b} t_b}}
+    \OVERRIDE
+    \REC{ {a :^{P_a} t_a}, {b :^{P_b} t_b}}
+  }
+\end{mathpar}
+
+\clearpage
+\begin{mathpar}
+  \begin{array}{|c|c||c|c||c|c|l}
+    Q_a&Q_b&O_a&O_b&P_a&P_b& \\
+    \hline
+    P_a &\NOTHING& \NOTHING & \NOTHING & P_a & \NOTHING & \\
+    O_a & \NOTHING& O_a & \NOTHING & \NOTHING & \NOTHING & \\
+    P_a\sqcup\READ   &\NOTHING& \READ & \NOTHING & P_a & \NOTHING & \\
+    \READ&\READ& \WRITE & \NOTHING & \READ & \NOTHING
+                       &  \text{copy, propagate left to write}\\
+    \WRITE &\READ& \WRITE & \NOTHING & \WRITE & \NOTHING
+                       &\text{copy, propagate left to write} \\
+  \end{array}
+
+  \begin{array}{|c|c||c|c||c|c|l}
+    Q_a&Q_b&O_a&O_b&P_a&P_b& \\
+    \hline
+    P_a & P_b & \NOTHING & \NOTHING & P_a & P_b & \\
+    O_a & P_b & O_a & \NOTHING & \NOTHING & P_b & \\
+    P_a\sqcup\READ   &P_b& \READ & \NOTHING & P_a & P_b & \\
+    \READ&P_b\sqcup\READ& \WRITE & \NOTHING & \READ & P_b
+                       &  \text{copy, propagate left to write}\\
+    \WRITE & P_b\sqcup\READ & \WRITE & \NOTHING & \WRITE & P_b
+                       &\text{copy, propagate left to write} \\
+  \end{array}
+
+  \begin{array}{|c|c||c|c||c|c|l}
+    Q_a&Q_b&O_a&O_b&P_a&P_b& \\
+    \hline
+    P_a & O_b & \NOTHING & O_b & P_a & \NOTHING & \\
+    O_a & O_b & O_a & O_b & \NOTHING & \NOTHING & \\
+    P_a\sqcup\READ  & O_b & \READ & O_b & P_a & \NOTHING & \\
+    \READ&O_b \sqcup\READ& \WRITE & O_b & \READ & \NOTHING
+                       &  \text{copy, propagate left to write}\\
+    \WRITE &O_b\sqcup\READ& \WRITE & O_b & \WRITE & \NOTHING
+                           &\text{copy, propagate left to write} \\
+  \end{array}
+
+\begin{array}{|c|c||c|c||c|c|l}
+    Q_a&Q_b&O_a&O_b&P_a&P_b& \\
+    \hline
+    P_a &P_b\sqcup\READ& \NOTHING & \READ & P_a & P_b & \\
+    O_a & P_b\sqcup\READ& O_a & \READ & \NOTHING & P_b & \\
+    P_a\sqcup\READ   &P_b\sqcup\READ& \READ & \READ & P_a & P_b & \\
+    \READ&P_b\sqcup\READ& \WRITE & \READ & \READ & P_b
+                       &  \text{copy, propagate left to write}\\
+    \WRITE &P_b\sqcup\READ& \WRITE & \READ & \WRITE & P_b
+                       &\text{copy, propagate left to write} \\
+  \end{array}
+
+  \begin{array}{|c|c||c|c||c|c|l}
+    Q_a&Q_b&O_a&O_b&P_a&P_b& \\
+    \hline
+    P_a\sqcup\READ &\READ& \NOTHING & \WRITE & P_a & \READ
+                           & \text{copy, propagate left to write} \\
+    \READ & \READ& O_a & \WRITE & \NOTHING & \READ
+                           & \text{copy, propagate left to write} \\
+    P_a\sqcup\READ   &\READ& \READ & \WRITE & P_a & \READ
+                           &  \text{copy, propagate left to write}\\
+    \READ&\READ& \WRITE & \WRITE & \READ & \READ
+                       &  \text{copy, propagate left to write}\\
+    \WRITE &\READ& \WRITE & \WRITE & \WRITE & \READ
+                       &\text{copy, propagate left to write} \\
+  \end{array}
+
+  \begin{array}{|c|c||c|c||c|c|l}
+    Q_a&Q_b&O_a&O_b&P_a&P_b& \\
+    \hline
+    P_a\sqcup\READ &\WRITE& \NOTHING & \WRITE & P_a & \WRITE
+                           & \text{copy, propagate left to write} \\
+    \READ & \WRITE& O_a & \WRITE & \NOTHING & \WRITE
+                           & \text{copy, propagate left to write} \\
+    P_a\sqcup\READ   &\WRITE& \READ & \WRITE & P_a & \WRITE
+                           &  \text{copy, propagate left to write}\\
+    \READ&\WRITE& \WRITE & \WRITE & \READ & \WRITE
+                       &  \text{copy, propagate left to write}\\
+    \WRITE &\WRITE& \WRITE & \WRITE & \WRITE & \WRITE
+                       &\text{copy, propagate left to write} \\
+  \end{array}
+\end{mathpar}
+A copy operation reads \textbf{every} field. Hence, a copy operation
+caused by one field implies a read effect on all fields.
 

### 2017-07-13 draft/unique-linear.tex
snapshot after slight cleanup

@@
 
-Effects attached to single record labels, intuitively $P, Q \subseteq
+\begin{figure}[tp]
+  \begin{align*}
+    e &::= x \mid \LAM xte \mid \APP ee \mid \LET xee  \mid c \mid e+e \mid \IF eee\\
+      &\mid \REC{ \overline{a=e}} \mid \GET ea \mid \UPD eae \\
+    O,P,Q &::= \NOTHING \mid \READ \mid \WRITE & \NOTHING & \sqsubseteq\WRITE\sqsubset \READ \\
+    s,t &::= \NUM \mid s \to t \mid \REC{ \overline{a :^P t}} \\
+  \end{align*}
+  \caption{Syntax}
+  \label{fig:syntax-3}
+\end{figure}
+In the syntax in Figure~\ref{fig:syntax-3}, effects are attached to single record labels, intuitively $P, Q \subseteq
 \{ \READ, \WRITE \}$ but it does not make sense to distinguish
-$\{\WRITE, \READ\}$ from $\{\WRITE\}$, so we take the three element lattice
-\begin{align*}
-  O, P, Q &::= \NOTHING \mid \READ \mid \WRITE
-  & \text{ordered by } \NOTHING
-  &\sqsubset \READ \sqsubset \WRITE
-\end{align*}
+$\{\WRITE, \READ\}$ from $\{\WRITE\}$, so we take the three element lattice as indicated.
 
@@
   \end{mathpar}
-  \caption{Propagation rules}
+  \caption{Typing rules}
   \label{fig:field-effects}
 \end{figure}
+The novel ingredient in the typing rules are the propagation and splitting rules for the effect annotations on
+record labels.
 Function closures have ``infinite'' splitting $(a :^P t)^*$ into $(a :^Q t)$.

### 2017-07-13 draft/unique-linear.tex
snapshot

@@
   \begin{align*}
-    e &::= x \mid \LAM xte \mid \APP ee \mid \LET xee  \mid c \mid e+e \mid \IF eee\\
-      &\mid \REC{ \overline{a=e}} \mid \GET ea \mid \UPD eae \\
-    O,P,Q &::= \NOTHING \mid \READ \mid \WRITE & \NOTHING & \sqsubseteq\WRITE\sqsubset \READ \\
-    s,t &::= \NUM \mid s \to t \mid \REC{ \overline{a :^P t}} \\
+    e,f,g,r
+    &::= x \mid \LAM xte \mid \APP fe \mid \FIX xtf \\
+    & \mid \LET xef  \mid c \mid f+g \mid \IF efg\\
+    &\mid \REC{ \overline{a=e}} \mid \GET ra \mid \UPD rae \\
+    O,P,Q &::= \PVAR \mid \NOTHING \mid \READ \mid \WRITE & \NOTHING & \sqsubseteq\WRITE\sqsubset \READ \\
+    s,t &::= \TVAR \mid \NUM \mid s \to t \mid \REC{ \overline{a :^P t}} \\
+    S, T& ::= \forall \TVAR. T \mid \forall \PVAR. T \mid t
   \end{align*}
@@
 
+  \inferrule[PT: Subtyping needed?]
+  { A, x:s \vdash f :t  \\ s \SUB t}
+  { A \vdash \FIX xtf : t}
+
   \inferrule{
@@
   { A \vdash \LET x{e_1} e_2 : t}
+
+  \inferrule
+  {
+    A = A_1 \OVERRIDE (A_2 \sqcup A_3) \\
+    A_1 \vdash e : \NUM \\
+    A_2 \vdash f : t_2 \\
+    A_3 \vdash g : t_3 \\
+    \SPLIT{t_1}{  t_2}{ t_3}
+  }
+  {
+    A \vdash \IF e f g : t_1
+  }
+  % PT: is subtyping really needed? We are propagating demand information *backwards*, so (e.g.) the demand
+  % on the branches of a conditional is the same as the demand on the result of the conditional.
+  % \inferrule[Sub]
+  % {A \vdash e : s \\ s \SUB t}
+  % {A \vdash e : t}
   \end{mathpar}
-  \caption{Typing rules}
+  \caption{Typing rules.
+  It seems like the subtyping in the FIX rule can alway be eliminated by unfolding the fixed point
+  once.
+  If the demand on $t$ is stronger or weaker than $s$, then unfolding will mediate the demand.
+  NB the directionality of subtyping is wrong if interpolated from the lattice ordering (as it is presently).
+}
   \label{fig:field-effects}
 \end{figure}
+% \begin{figure}[tp]
+%   \begin{mathpar}
+%     \inferrule{}{\NUM \SUB \NUM}
+%
+%     \inferrule{
+%       s' \SUB s \\ t \SUB t'
+%     }{
+%       s \to t \SUB s' \to t' 
+%     }
+%
+%     \inferrule{
+%       \overline{s_i \SUB t_i} \\
+%       \overline{P_i \sqsubseteq Q_i}
+%     }{
+%       \REC{ \overline{a_i :^{P_i} s_i}}
+%       \SUB
+%       \REC{ \overline{a_i :^{Q_i} t_i}}
+%     }
+%   \end{mathpar}
+%   \caption{Subtyping}
+%   \label{fig:subtyping-3}
+% \end{figure}
+\begin{figure}[tp]
+  \begin{align*}
+    \NUM \sqcup \NUM
+    & = \NUM
+    \\
+    (s \to t) \sqcup (s' \to t')
+    & = (s \sqcap s') \to (t \sqcup t') \qquad ???
+    \\
+    \REC{ \overline{a_i :^{P_i} s_i}} \sqcup \REC{ \overline{a_i :^{Q_i} t_i}}
+    &= \REC{ \overline{a_i :^{P_i \sqcup Q_i} (s_i \sqcup t_i)}}
+  \end{align*}
+  \begin{mathpar}
+    \inferrule{}{
+      \SPLIT \NUM \NUM \NUM
+    }
+
+    \inferrule{}{
+      \JOIN \NUM \NUM \NUM
+    }
+    \\
+    \inferrule{
+      \JOIN{s}{ s_1}{ s_2} \\
+      \SPLIT{t}{t_1}{t_2}
+    }{
+      \SPLIT{s \to t}{s_1 \to t_1}{s_2 \to t_2}
+    }
+
+    \inferrule{
+      ??? \\
+      \JOIN{s}{ s_1}{ s_2} \\
+      \SPLIT{t}{t_1}{t_2}
+    }{
+      \JOIN{s \to t}{s_1 \to t_1}{s_2 \to t_2}
+    }
+    \\
+    \inferrule
+    {\overline{\SPLIT{s_i}{s_i'}{s_i''}}}
+    {\SPLIT{\REC{ \overline{a_i :^{P_i} s_i}}}{\REC{ \overline{a_i :^{P_i} s_i'}}}{\REC{ \overline{a_i :^{P_i} s_i''}}}}
+  \end{mathpar}
+  \caption{Bounding and splitting operations on types}
+  \label{fig:bounding-3}
+\end{figure}
 The novel ingredient in the typing rules are the propagation and splitting rules for the effect annotations on
-record labels.
+record labels indicated by the ternary override relation $\OVERRIDE$.
 Function closures have ``infinite'' splitting $(a :^P t)^*$ into $(a :^Q t)$.
@@
 caused by one field implies a read effect on all fields.
-
+\clearpage{}
+Consider the typing for 
+\begin{align*}
+  f
+  &=(\LAM r{\REC{ a:^\WRITE \NUM}} \UPD r a{42})
+  &&: \REC{ a:^\WRITE \NUM} \to \REC{ a:^\PVAR \NUM}
+  \\
+  g
+  &=(\LAM r{\REC{a :^\NOTHING \NUM}} \REC{a = 0})
+  &&: \REC{a :^\NOTHING \NUM} \to \REC{ a:^\PVAR \NUM}
+  \\
+  &\IF e{f}{g}
+  &&: \REC{ a:^\WRITE \NUM} \to \REC{ a:^\PVAR \NUM}
+\end{align*}
+This propagation is correct,
+\textbf{but} if the write is propagated, this may cause a useless copy if $e$ is false.
 \end{document}

### 2017-07-13 draft/unique-linear.tex
finished adding joins and meets to typing
comments

@@
   {
-    A = A_1 \OVERRIDE (A_2 \sqcup A_3) \\
-    A_1 \vdash e : \NUM \\
+    A = A_0 \OVERRIDE A_1 \\
+    \JOIN{A_1}{A_2}{ A_3} \\
+    \SPLIT{t_1}{  t_2}{ t_3} \\\\
+    A_0 \vdash e : \NUM \\
     A_2 \vdash f : t_2 \\
-    A_3 \vdash g : t_3 \\
-    \SPLIT{t_1}{  t_2}{ t_3}
+    A_3 \vdash g : t_3
   }
@@
 \begin{figure}[tp]
-  \begin{align*}
-    \NUM \sqcup \NUM
-    & = \NUM
-    \\
-    (s \to t) \sqcup (s' \to t')
-    & = (s \sqcap s') \to (t \sqcup t') \qquad ???
-    \\
-    \REC{ \overline{a_i :^{P_i} s_i}} \sqcup \REC{ \overline{a_i :^{Q_i} t_i}}
-    &= \REC{ \overline{a_i :^{P_i \sqcup Q_i} (s_i \sqcup t_i)}}
-  \end{align*}
+  % \begin{align*}
+  %   \NUM \sqcup \NUM
+  %   & = \NUM
+  %   \\
+  %   (s \to t) \sqcup (s' \to t')
+  %   & = (s \sqcap s') \to (t \sqcup t') \qquad ???
+  %   \\
+  %   \REC{ \overline{a_i :^{P_i} s_i}} \sqcup \REC{ \overline{a_i :^{Q_i} t_i}}
+  %   &= \REC{ \overline{a_i :^{P_i \sqcup Q_i} (s_i \sqcup t_i)}}
+  % \end{align*}
   \begin{mathpar}
@@
     \inferrule{
-      ??? \\
-      \JOIN{s}{ s_1}{ s_2} \\
-      \SPLIT{t}{t_1}{t_2}
+      \SPLIT{s}{ s_1}{ s_2} \\
+      \JOIN{t}{t_1}{t_2}
     }{
@@
     {\SPLIT{\REC{ \overline{a_i :^{P_i} s_i}}}{\REC{ \overline{a_i :^{P_i} s_i'}}}{\REC{ \overline{a_i :^{P_i} s_i''}}}}
+
+    \inferrule
+    {\overline{\JOIN{s_i}{s_i'}{s_i''}} \\
+    \overline{O_I = P_i \sqcup Q_i}}
+    {\JOIN{\REC{ \overline{a_i :^{O_i} s_i}}}{\REC{ \overline{a_i :^{P_i} s_i'}}}{\REC{ \overline{a_i :^{Q_i} s_i''}}}}
   \end{mathpar}
-  \caption{Bounding and splitting operations on types}
+  \caption{Bounding and splitting operations on types. The intuition is that \SPLITK{} propagates
+    information in the positive positions of a type downwards towards the subexpressions. At the
+    same time, it joins information in the negative positions and propagates the join upwards. The
+    \JOINK{} operation works dually.}
   \label{fig:bounding-3}

### 2017-07-15 draft/unique-linear.tex
polymorphic-woes

@@
 \end{align*}
-This propagation is correct,
-\textbf{but} if the write is propagated, this may cause a useless copy if $e$ is false.
+This propagation using $\WRITE\sqcup\NOTHING = \WRITE$ is correct,
+\textbf{but} the propagation of $\WRITE$  may cause a useless copy if $e$ is false.
+
+The typing is analogous to pushing the conditional into the lamdas.
+\begin{align*}
+  r : \REC{ a:^\WRITE \NUM} & \vdash \UPD r a{42} &&: \REC{ a:^\PVAR \NUM}
+  \\
+  r : \REC{ a:^\NOTHING \NUM} & \vdash \REC{a = 0} &&: \REC{ a:^\PVAR \NUM}
+  \\
+  r : \REC{ a:^\WRITE \NUM}& \vdash \IF e{( \UPD r a{42})}{(\REC{a = 0})} &&: \REC{ a:^\PVAR \NUM}
+  \\
+  \vdash \LAM r {\REC{ a:^\WRITE \NUM}} & \IF e{( \UPD r a{42})}{(\REC{a = 0})}
+                                                  &&: {\REC{ a:^\WRITE \NUM}} \to \REC{ a:^\PVAR \NUM}
+\end{align*}
+
+\textbf{Let's finally get down to polymorphism. }
+\begin{align*}
+  \mathit{id} =
+  \LAM x\TVAR x & : \forall\TVAR.\TVAR\to\TVAR
+  &&\text{seems fine, just instantiate}
+  \\
+  \mathit{dup} =
+  \LAM x\TVAR x & : \forall\TVAR.\TVAR\to(\TVAR,\TVAR)
+\end{align*}
+\begin{verbatim}
+lambda r.
+  let (r1, r2) = dup r 
+  in  (r1 { a = 42 }, r2.a)
+-- should behave like "write before read"
+lambda r. 
+      (r  { a = 42 }, r .a)
+\end{verbatim}
+\begin{align*}
+  r : \REC{ a:^\WRITE \NUM} & \vdash \UPD r a{42} &&: \REC{ a:^\PVAR \NUM}
+  \\
+  r : \REC{ a:^\READ \NUM} & \vdash  r. a &&: \NUM
+  \\
+  r : \REC{ a:^\READ \NUM} & \vdash (\UPD r a{42}, r. a)  &&: \REC{ a:^\PVAR \NUM} \times \NUM
+\end{align*}
+Using \textit{dup}
+\begin{align*}
+  r_1 : \REC{ a:^\WRITE \NUM} & \vdash \UPD r a{42} &&: \REC{ a:^\PVAR \NUM}
+  \\
+  r_2 : \REC{ a:^\READ \NUM} & \vdash  r. a &&: \NUM
+  \\
+  r_1 : \REC{ a:^\WRITE \NUM},
+  r_2 : \REC{ a:^\READ \NUM} & \vdash  (\UPD r a{42} , r. a) &&: \REC{ a:^\PVAR \NUM} \times \NUM
+\end{align*}
 \end{document}

### 2017-07-17 draft/unique-linear.tex
polymorphism, state monad

@@
 \end{align*}
+\textbf{Suggestion}
+Distiguish positive and negative occurrences of a type variable.
+\begin{itemize}
+\item If a negative occurrence of a type variable is instantiated with a record type, then its fields are all set to
+  \READ.
+\item Every instantiation of a positive occurrence of a type variable with a record type is allowed
+  to have \textbf{different field permissions}. If a record type ends up with permission $\WRITE$,
+  then the corresponding record is copied after instantiation.
+\item If there is only one positive occurrence under an affine type constructor, then its permission
+  is pushed through.
+\end{itemize}
+That does not quite solve the problem:
+\begin{verbatim}
+let (r1, r2) = (r, r)
+in  (r1 { a = 42 }, r2.a)
+\end{verbatim}
+Here the reads and writes to \texttt{r1} and \texttt{r2} get out of order due to aliasing in the
+pair.
+
+Practical instance: the state monad. Turns out that the bind cannot be handled satisfactorily by the
+suggestion on the table because we don't know that \texttt{s} is linear. Seems like linearity is
+useful knowledge for polymorphic arguments.
+\begin{verbatim}
+type ST s a = s -> (s,a)
+
+-- a: -, +
+-- s: -, +
+returnST :: a -> ST s a
+returnST a = \s -> (s, a)
+
+-- b: -, +
+-- a: -, +
+-- s: +, -, +, -, -, +
+bindST :: ST s a -> (a -> ST s b) -> ST s b
+bindST ma f = \s -> let (s1,a) = ma s 
+                    in  f s1 a
+\end{verbatim}
 \end{document}

### 2018-01-16 kindly/main.tex
First try.

\documentclass[sigconf,author]{acmart}\settopmatter{printfolios=true}

% Encoding and lang
\usepackage[T1]{fontenc}
\usepackage[utf8]{inputenc}
\usepackage[english]{babel}

% Graphical packages
\usepackage{graphicx}

% Math
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{amssymb}
\usepackage{amsthm}
% \usepackage{mathrsfs}
\usepackage{mathtools}
% \usepackage{textcomp}
% \usepackage{textgreek}

\usepackage{caption}
\usepackage{subcaption}

\usepackage{natbib}% Good citations and bibliography
\usepackage{mathpartir} % Syntax trees
\usepackage[nameinlink,capitalize]{cleveref}

\bibliographystyle{ACM-Reference-Format}
\citestyle{acmauthoryear}

\input{prelude}
\input{notations}

\begin{document}

\title{Kindly Bent to Free Us}
\maketitle

\begin{figure*}[h]
  \centering
  \input{syntax}
  \caption{Syntax}

  \input{kinds}
  \caption{Kind rules -- $\haskind{P}{\E}{\tau}{k}$ }

  \input{typing}
  \caption{Typing rules -- $\hastype{P}{\E}{e}{\tau}$ }

  \input{semantics}
  \caption{Reduction rules -- $\ered{I}{E}{e}{v}$ }

\end{figure*}

\end{document}

%%% Local Variables:
%%% mode: latex
%%% TeX-master: t
%%% End:

### 2018-05-24 kindly/main.tex
Inference.
More inference.
Add kind scheme and type validity.

@@
-\documentclass[sigconf,author]{acmart}\settopmatter{printfolios=true}
+\documentclass[manuscript,author]{acmart}\settopmatter{printfolios=true}
 
@@
   \caption{Syntax}
+\end{figure*}
+\begin{figure*}[h]
+  \input{kinds}
+  \caption{Constraint rules -- $\entail{C}{k}$ }
+\end{figure*}
+\begin{figure*}[h]
+  \input{types}
+  \caption{Type validity rules -- $\tyval{C}{\E}{\tau}$ }
+  \input{inference}
+  \caption{Inference rules -- $\inferW{\Sigma}{(C,\psi)}{\bf{\E}}{\bf{e}}{\tau}$ }
+\end{figure*}
+% \begin{figure*}[h]
+%   \input{semantics}
+%   \caption{Reduction rules -- $\ered{I}{E}{e}{v}$ }
 
-  \input{kinds}
-  \caption{Kind rules -- $\haskind{P}{\E}{\tau}{k}$ }
+% \end{figure*}
 
-  \input{typing}
-  \caption{Typing rules -- $\hastype{P}{\E}{e}{\tau}$ }
+\section*{Todo}
 
-  \input{semantics}
-  \caption{Reduction rules -- $\ered{I}{E}{e}{v}$ }
+\begin{itemize}
+\item Add type constructors
+\item Properly ensure that it respects HM(X) (cylindric, \dots)
+\item Show principal type inference:
+  \begin{itemize}
+  \item Principal constraint system
+  \item Regular constraint system: $\Ceq{\tau}{\tau'} \implies \fv{\tau} = \fv{\tau'}$.
+  \item Solved forms are in simplified form.
+    $C\in S, \entail{C}{\Ceq{\tau}{\tau'}} \implies \entail{}{\Ceq{\tau}{\tau'}}$.
+  \end{itemize}
+\item Show equivalence with the logic-based system.
 
-\end{figure*}
+\end{itemize}
+
+\begin{lemma}
+  Without loss of generality, we can consider that
+  kind inequalities in satisfiable constraints
+  are only done on kind variables.
+
+  \begin{proof}
+    Consider the constraint $\Cleq{k}{k'}\Cand C$.
+    \begin{itemize}
+    \item If $k$ and $k'$ are both constants, it can be removed.
+    \item If the constraint is $\Cleq{\kvar}{\kaff}$ or $\Cleq{\kun}{\kvar}$, it can be removed.
+    \item If the constraint is $\Cleq{\kvar}{\kun}$ or $\Cleq{\kaff}{\kvar}$, we
+      can substitute $\kvar$ by its value in $C$.
+    \end{itemize}
+  \end{proof}
+\end{lemma}
 

### 2018-11-06 kindly/lang.tex
Blip.
A bit more structure.

--- 2018-01-16 kindly/main.tex
@@
-\documentclass[sigconf,author]{acmart}\settopmatter{printfolios=true}
+\section{The \lang language}
 
-% Encoding and lang
-\usepackage[T1]{fontenc}
-\usepackage[utf8]{inputenc}
-\usepackage[english]{babel}
+\subsection{Syntax}
 
-% Graphical packages
-\usepackage{graphicx}
+The syntax of \lang is presented in \cref{grammar} and follows
+traditional ML languages with let bindings.
+Data constructors are introduced by $\app{C}{x}$ and eliminated
+with by $\matchin{C}{x}{e}{e'}$.
+Types are composed of type variable $\tvar$, parameterized type constructors
+$\T{t}$ and arrow types annotated with a kind, noted $\tau \tarr{k} \tau$.
+Kinds can be either a kind variable $\kvar$, affine $\kaff$ or unrestricted $kun$.
+Type schemes and kind schemes are guarded by a set of constraints.
 
-% Math
-\usepackage{amsmath}
-\usepackage{amsfonts}
-\usepackage{amssymb}
-\usepackage{amsthm}
-% \usepackage{mathrsfs}
-\usepackage{mathtools}
-% \usepackage{textcomp}
-% \usepackage{textgreek}
-
-\usepackage{caption}
-\usepackage{subcaption}
-
-\usepackage{natbib}% Good citations and bibliography
-\usepackage{mathpartir} % Syntax trees
-\usepackage[nameinlink,capitalize]{cleveref}
-
-\bibliographystyle{ACM-Reference-Format}
-\citestyle{acmauthoryear}
-
-\input{prelude}
-\input{notations}
-
-\begin{document}
-
-\title{Kindly Bent to Free Us}
-\maketitle
+Environments are composed of value and type bindings, along with type
+declarations. Type declarations, of the form
+$\tydecl{t}{\kschm}{C}{\tau}$, introduce both a new type constructor $\T{t}$ and
+a data constructor $C$.
 
@@
   \centering
-  \input{syntax}
+  \input{grammar}
   \caption{Syntax}
-
-  \input{kinds}
-  \caption{Kind rules -- $\haskind{P}{\E}{\tau}{k}$ }
-
-  \input{typing}
-  \caption{Typing rules -- $\hastype{P}{\E}{e}{\tau}$ }
-
-  \input{semantics}
-  \caption{Reduction rules -- $\ered{I}{E}{e}{v}$ }
-
+  \label{grammar}
 \end{figure*}
 
-\end{document}
+\subsection{Semantics}
+
+\TODO{}
+% \begin{figure*}[h]
+%   \input{semantics}
+%   \caption{Reduction rules -- $\ered{I}{E}{e}{v}$ }
+% \end{figure*}
+
+\subsection{Typing}
+
+\TODO{Add system-F like type rules}
 
@@
 %%% mode: latex
-%%% TeX-master: t
+%%% TeX-master: "main"
 %%% End:

### 2018-11-06 kindly/soa.tex
Blip.
A bit more structure.

--- 2018-01-16 kindly/main.tex
@@
-\documentclass[sigconf,author]{acmart}\settopmatter{printfolios=true}
+\section{Related works}
 
-% Encoding and lang
-\usepackage[T1]{fontenc}
-\usepackage[utf8]{inputenc}
-\usepackage[english]{babel}
+\subsection{Substructural type systems}
 
-% Graphical packages
-\usepackage{graphicx}
+\paragraph{Functional linear of affine type systems}
 
-% Math
-\usepackage{amsmath}
-\usepackage{amsfonts}
-\usepackage{amssymb}
-\usepackage{amsthm}
-% \usepackage{mathrsfs}
-\usepackage{mathtools}
-% \usepackage{textcomp}
-% \usepackage{textgreek}
+Alms~\citep{DBLP:conf/popl/TovP11},
+F$^\circ$~\citep{DBLP:conf/tldi/MazurakZZ10},
+Quill~\citep{DBLP:conf/icfp/Morris16},
+Linear Haskell~\citep{DBLP:journals/pacmpl/BernardyBNJS18}
 
-\usepackage{caption}
-\usepackage{subcaption}
+\paragraph{Object systems}
+\TODO{}
 
-\usepackage{natbib}% Good citations and bibliography
-\usepackage{mathpartir} % Syntax trees
-\usepackage[nameinlink,capitalize]{cleveref}
+\subsection{Qualified Types}
 
-\bibliographystyle{ACM-Reference-Format}
-\citestyle{acmauthoryear}
+HM(X)~\citep{DBLP:journals/tapos/OderskySW99} with extensions to
+bounded universal and existentials~\citep{DBLP:conf/icfp/Simonet03}
+and GADTs~\citep{DBLP:journals/toplas/SimonetP07},
+and syntactic proof of soundness~\citep{DBLP:journals/entcs/SkalkaP02}.
 
-\input{prelude}
-\input{notations}
+\subsection{Subtyping and constraints}
 
-\begin{document}
-
-\title{Kindly Bent to Free Us}
-\maketitle
-
-\begin{figure*}[h]
-  \centering
-  \input{syntax}
-  \caption{Syntax}
-
-  \input{kinds}
-  \caption{Kind rules -- $\haskind{P}{\E}{\tau}{k}$ }
-
-  \input{typing}
-  \caption{Typing rules -- $\hastype{P}{\E}{e}{\tau}$ }
-
-  \input{semantics}
-  \caption{Reduction rules -- $\ered{I}{E}{e}{v}$ }
-
-\end{figure*}
-
-\end{document}
+Flowcaml~\citep{DBLP:conf/popl/PottierS02},
+\citet{DBLP:conf/sas/TrifonovS96}.
 

### 2018-11-07 kindly/infer/infer.tex
Blip.
A bit more structure.
Start working on the inference spec.

--- 2018-05-24 kindly/main.tex
@@
-\documentclass[manuscript,author]{acmart}\settopmatter{printfolios=true}
+\section{Type inference}
 
-% Encoding and lang
-\usepackage[T1]{fontenc}
-\usepackage[utf8]{inputenc}
-\usepackage[english]{babel}
+We now formulate type inference for the \lang language. Our inference technique is based on the HM(X) framework~\citep{DBLP:journals/tapos/OderskySW99} which
+presents how to merge Hindley-Milner type inference for let-polymorphism and
+constraint solving. We first present our constraint language, we then show
+our slight extension of HM(X) that support a kind system and affine types.
 
-% Graphical packages
-\usepackage{graphicx}
+\subsection{Constraint language}
 
-% Math
-\usepackage{amsmath}
-\usepackage{amsfonts}
-\usepackage{amssymb}
-\usepackage{amsthm}
-% \usepackage{mathrsfs}
-\usepackage{mathtools}
-% \usepackage{textcomp}
-% \usepackage{textgreek}
+We consider $\mathcal C$ the set of constrains.
+The grammar of constrains, presented
+in \cref{grammar:constraint}, follows the traditional HM(X) formulation
+with conjunctions,projections and type inequalities. The only new
+element specific to our approach are kind equalities.
+Entailment is noted $\entail{C}{D}$, where $D$ is a consequence of the
+constraints $C$. The entailment rules are provided in \cref{rules:entail}.
 
-\usepackage{caption}
-\usepackage{subcaption}
-
-\usepackage{natbib}% Good citations and bibliography
-\usepackage{mathpartir} % Syntax trees
-\usepackage[nameinlink,capitalize]{cleveref}
-
-\bibliographystyle{ACM-Reference-Format}
-\citestyle{acmauthoryear}
-
-\input{prelude}
-\input{notations}
-
-\begin{document}
-
-\title{Kindly Bent to Free Us}
-\maketitle
+\begin{figure}[h]
+  \centering
+  \begin{align*}
+    C ::=&\ \Cleq{\tau_1}{\tau_2}\ |\ \Cleq{k_1}{k_2}\ |\ C_1 \Cand C_2\ |\ \Cproj{\alpha}{C}
+  \end{align*}
+  \caption{The constraint language}
+  \label{grammar:constraint}
+\end{figure}
 
 \begin{figure*}[h]
-  \centering
-  \input{syntax}
-  \caption{Syntax}
+  \input{infer/entails}
+  \caption{Entailment rules -- $\entail{C}{D}$ }
+  \label{rules:entail}
 \end{figure*}
+
+\paragraph{Solved forms}
+
+The set of solved forms, noted $\mathcal S$,
+is the set of constraints than can be used inside
+type and kind schemes. $\mathcal S$ is composed only of kind
+inequalities \emph{over variables}. For convenience, if $C\in\mathcal S$, we
+note $C$ as a list of kind inequalities: $\Cleq{\kvar_i}{\kvar_{i'}}^n$.
+\TODO{Extend the properties of solved forms}
+
+We consider the existence of a function $\normalize$ which takes
+a constraint in $\mathcal C$ and a unifier $\psi$ and returns a constraint
+in solved form $C'$, and an updated unifier.
+
+\subsection{Kind inference}
+
+We note $\inferK{(\unif,C)}{\E}{\tau}{k}$ the fact that type $\tau$ has kind $k$
+in environment $\E$ with constraints $C$ and unifier $\unif$. From an
+algorithmic point of view, $\E$ and $\tau$ are the input parameters of
+our inference procedure.
+
+\subsection{Type inference}
+
 \begin{figure*}[h]
-  \input{kinds}
-  \caption{Constraint rules -- $\entail{C}{k}$ }
-\end{figure*}
-\begin{figure*}[h]
-  \input{types}
+  \input{infer/validity}
   \caption{Type validity rules -- $\tyval{C}{\E}{\tau}$ }
-  \input{inference}
+  \input{infer/rules}
   \caption{Inference rules -- $\inferW{\Sigma}{(C,\psi)}{\bf{\E}}{\bf{e}}{\tau}$ }
 \end{figure*}
-% \begin{figure*}[h]
-%   \input{semantics}
-%   \caption{Reduction rules -- $\ered{I}{E}{e}{v}$ }
-
-% \end{figure*}
 
@@
 
-\end{document}
-
 %%% Local Variables:
 %%% mode: latex
-%%% TeX-master: t
+%%% TeX-master: "../main"
 %%% End:

### 2018-11-08 kindly/infer/infer.tex
Fix/Extend the type inference rules.

@@
 
-\subsection{Constraint language}
+\subsection{Preliminaries}
+
+\subsubsection{Constraint language}
 
@@
 
-\paragraph{Solved forms}
+\subsubsection{Solved forms}
 
@@
 
+\subsubsection{Usage maps}
+
+\TODO{Explain $\Sigma$}
+
 \subsection{Kind inference}
 
-We note $\inferK{(\unif,C)}{\E}{\tau}{k}$ the fact that type $\tau$ has kind $k$
-in environment $\E$ with constraints $C$ and unifier $\unif$. From an
+We note $\inferK{(C,\unif)}{\bf{\E}}{\bf{\tau}}{k}$ when type $\tau$ has kind $k$
+in environment $\E$ under constraints $C$ and unifier $\unif$. From an
 algorithmic point of view, $\E$ and $\tau$ are the input parameters of
 our inference procedure.
+We present the kind inference algorithm as a set of syntax-directed rules in
+\cref{rules:kinding}.
+Since higher-kinded types are not supported, the type application
+rule and the type constructor rule are merged in {\sc KApp}.
+Additionally, type variables must be of a simple kind in rule {\sc KVar}.
+Note that in the case of a type constructor with no argument, the {\sc KApp}
+rule degenerates to a simpler form which is similar to the {\sc KVar} rule.
+Kind schemes are instantiated in the {\sc KApp} and {\sc KVar} rules by creating
+fresh kind variables and the associated substitution.
+
+\TODO{Explain more?}
+
+\begin{figure}[h]
+  \centering
+  \input{infer/kinds}
+  \caption{Kind inference rules -- $\inferK{(C,\unif)}{\E}{\tau}{k}$}
+  \label{rules:kinding}
+\end{figure}
 
@@
 
+We reformulate the HM(X) type inference in the context of our affine type
+system. The main difference compared to HM(X) are noted in \addlin{blue}.
+We note $\inferW{\addlin{\Sigma}}{(C,\unif)}{\bf{\E}}{\bf{e}}{\tau}$ when
+$e$\ as type $\tau$ in $\E$ under the constraints $C$ and unifier $\unif$.
+$\Sigma$ is a map which associates free variables in $e$ to
+their kinds.
+As before, $\E$ and $e$ are the input parameters of the inference
+algorithm. The syntax-directed rules are shown in \cref{rules:typing}.
+
 \begin{figure*}[h]
-  \input{infer/validity}
-  \caption{Type validity rules -- $\tyval{C}{\E}{\tau}$ }
-  \input{infer/rules}
-  \caption{Inference rules -- $\inferW{\Sigma}{(C,\psi)}{\bf{\E}}{\bf{e}}{\tau}$ }
+  \input{infer/typing}
+  \caption{Type Inference rules -- $\inferW{\Sigma}{(C,\psi)}{\bf{\E}}{\bf{e}}{\tau}$ }
+  \label{rules:typing}
 \end{figure*}
 
-\section*{Todo}
+\subsection{Constraint normalization and generalization}
+
+\TODO{}
+
+\subsection{Principality}
+
+\TODO{}
 
 \begin{itemize}
-\item Add type constructors
 \item Properly ensure that it respects HM(X) (cylindric, \dots)

### 2018-11-08 kindly/lang.tex
Fix/Extend the type inference rules.

@@
 traditional ML languages with let bindings.
-Data constructors are introduced by $\app{C}{x}$ and eliminated
-with by $\matchin{C}{x}{e}{e'}$.
+Data constructors are introduced by $\introK{K}{e}$ and eliminated
+with by $\elimK{K}{e}$.
 Types are composed of type variable $\tvar$, parameterized type constructors
@@
 declarations. Type declarations, of the form
-$\tydecl{t}{\kschm}{C}{\tau}$, introduce both a new type constructor $\T{t}$ and
-a data constructor $C$.
+$\tydecl{t}{\kschm}{K}{\tau}$, introduce both a new type constructor $\T{t}$ and
+a data constructor $K$.
 

### 2018-11-09 kindly/infer/infer.tex
Add normalization procedure.

@@
 We now formulate type inference for the \lang language. Our inference technique is based on the HM(X) framework~\citep{DBLP:journals/tapos/OderskySW99} which
-presents how to merge Hindley-Milner type inference for let-polymorphism and
-constraint solving. We first present our constraint language, we then show
-our slight extension of HM(X) that support a kind system and affine types.
+presents how to implement Hindley-Milner type inference in a language
+with qualified types. We first present various preliminaries definitions,
+we then show our extension of HM(X) that support a kind system and affine
+types.
 
@@
 Entailment is noted $\entail{C}{D}$, where $D$ is a consequence of the
-constraints $C$. The entailment rules are provided in \cref{rules:entail}.
+constraints $C$.
+The base entailment rules are shown in \cref{rules:entail}, we consider
+the completion of that system to form a cylindric constraint system.
+\TODO{Give the properties}
+
+We say that $C$ and $D$ are equivalent, noted $C \equivC D$,
+when $\entail{C}{D}$ and $\entail{D}{C}$.
 
@@
   \input{infer/entails}
-  \caption{Entailment rules -- $\entail{C}{D}$ }
+  \caption{Base entailment rules -- $\entail{C}{D}$ }
   \label{rules:entail}
@@
 
-\subsubsection{Solved forms}
-
-The set of solved forms, noted $\mathcal S$,
-is the set of constraints than can be used inside
-type and kind schemes. $\mathcal S$ is composed only of kind
-inequalities \emph{over variables}. For convenience, if $C\in\mathcal S$, we
-note $C$ as a list of kind inequalities: $\Cleq{\kvar_i}{\kvar_{i'}}^n$.
-\TODO{Extend the properties of solved forms}
-
+We note $\mathcal S$ the set of solved forms
+which can be used inside type and kind schemes.
+We define $\mathcal S$ as $\mathcal C$ quotiented by the relation $\equivC$.
+%
 We consider the existence of a function $\normalize$ which takes
 a constraint in $\mathcal C$ and a unifier $\psi$ and returns a constraint
-in solved form $C'$, and an updated unifier.
+in solved form $C' \in \mathcal S$,
+and an updated unifier. We detail the implementation
+of the normalization function in \cref{sec:normalize}
+
+% $\mathcal S$ is composed only of kind
+% inequalities \emph{over variables}. For convenience, if $C\in\mathcal S$, we
+% note $C$ as a list of kind inequalities: $\Cleq{\kvar_i}{\kvar_{i'}}^n$.
+% \TODO{Extend the properties of solved forms}
 
@@
 
-\TODO{Explain $\Sigma$}
+In order to determine if a variable is used in an affine manner, we must track
+it's uses and the associated kinds. For instance, in the expression
+$(x,x)$, $x$ is used twice. If $x$ is of type $\tau$, which is of kind $k$,
+we must add the constraint $\Cleq{k}{\kun}$. In order to record uses and kinds,
+our type inference relation will return a ``usage map'', noted $\Sv$,
+which associates each variable to a list of kinds corresponding to each use site
+of the variable.
+We note $\Sone{x}{k_1,\dots,k_n}$ the singleton usage map.
+Common operations on usages maps are defined in \cref{op:usgmap}. We define
+regular set operations such as union
+($\Sv\cup\Sv'$) and intersection ($\Sv\cap\Sv'$), along
+with constraint creation $\Cleq{\Sv}{k_0}$ which indicates that all the kinds
+presents in the usage map $\Sv$ are constrained by the kind $k_0$.
+These operations corresponds roughly to the various parts of the join
+operator on sets $\esplit{}{}$.
+
+\begin{figure}[h]\centering
+\begin{align*}
+  \Sv\cap\Sv'
+  &= \left\{ x \to l,l'
+    \mid \Sone{x}{l}\in\Sv \text{ and } \Sone{x}{l'}\in\Sv' \right\} \\
+  \Sv\cup\Sv'
+  &= \Sv\cap\Sv' \sqcup
+    (\Sv\Sdel{x' \mid x'\in\Sv'}) \sqcup
+    (\Sv'\Sdel{x \mid x\in\Sv})
+  &\text{where $\sqcup$ is the disjoint union}\\
+  \Cleq{\Sv}{k_0}
+  &= \bigwedge_{\Sone{x}{l_x}\in\Sv} \bigwedge_{k\in l_x} \Cleq{k}{k_0}
+\end{align*}
+\caption{Operation on usage maps $\Sv$}
+\label{op:usgmap}
+\end{figure}
 
@@
 fresh kind variables and the associated substitution.
-
-\TODO{Explain more?}
 
@@
 
-\subsection{Constraint normalization and generalization}
+\subsection{Normal forms and generalization}
+\label{sec:normalize}
 
+\newcommand\lub\bigvee
+\newcommand\glb\bigwedge
+
+We now place ourselves in a more general context where kinds are either variables
+or constants belonging to a total bounded lattice $(\mathcal L, \leq)$ ie.,
+a lattice which admits a total order and upper and lower bounds.
+We note lattice elements $l$ and $\glb_i l_i$ (resp. $\lub_i l_i$)
+the greatest lower bound (resp. least upper bound) in $\mathcal L$.
+
+The normalization procedure is noted $\normalize{C_0}{\unif_0}$ where
+$C_0\in \mathcal C$ is a set of constraints and $\unif_0$ is a unifier.
+It returns a constraint $C \in \mathcal S$ in
+solved form and a unifier $\unif$.
+Our unification algorithm proceed as follows:
+\begin{enumerate}
+\item Solve all type equality constraints through Herband unification and
+  gather all existential quantifications at the front of the constraint.
+  We obtain a constraint $C = \exists \kvar_i,\ \Cleq{k_j}{k'_j}_j$ and
+  a unifier $\unif$.
+
+  We note $\mathcal R$ the relation $\Cleq{k_j}{k'_j}_j$,
+  $\mathcal G$ the underlying directed graph and $V$ its vertices.
+
+\item Saturate the lattice equalities in $\mathcal R$.
+
+  More precisely, for each kind variable $\kvar \in V$,
+  for each constant $l_i$ (resp. $l_j$) such that
+  there is a path from $l_i$ to $\kvar$ (resp. from $\kvar$ to $l_j$) in $\mathcal G$,
+  add an edge from $\lub l_i$ to $\kvar$
+  (resp. from $\kvar$ to $\glb l_j$).
+  This step is well defined since
+  our lattice is bounded and $\lub\emptyset$ and $\glb\emptyset$ are well defined.
+
+  We also complement $\mathcal R$ with $(\leq)$ by adding an edge
+  between related constants.
+\item For each strongly connected component in $\mathcal G$, unify all its vertices and replace it by a representative and complete the unifier accordingly.
+  The representative of a strongly connected component $g$ can be determined as follows:
+  \begin{itemize}
+  \item If $g$ does not contain any constant, then the representative
+    is a fresh kind variable.
+  \item If $g$ contains exactly one constant, it is the representative.
+  \item Otherwise, the initial constraint $C_0$ is not satisfiable.
+  \end{itemize}
+  Note that this step will also detect all unsatisfiable constraints.
+\item Take the transitive closure of $\mathcal R$.
+\item Remove all the vertices corresponding to the kind variables $\kvar_i$
+  that are existentially quantified in $C$.
+\item Remove the extremums of $\mathcal L$ and the edges of $(\leq)$
+  from $\mathcal R$.
+\item Take the transitive reduction of $\mathcal R$.
+  % Return $C = \left\{ k \leq k' \mid k \operatorname{\mathcal R}k' \right\}$
+\end{enumerate}
+
+\TODO{Add an example}
+
+\begin{lemma}
+  Given a constraint $C\in\mathcal C$, a substitution $\unif$ and
+  $(C',\unif') = \normalize{C}{\unif}$, we have the following properties:
+  \begin{enumerate}
+  \item $C'$ is of the form $\{ k_i \leq k'_i\}_i$.
+  \item For any disjoint subformulas $C_1$ and $C_2$ of $C'$,
+    then $C_1 \operatorname{\nvdash_e} C_2$ and $C_2 \operatorname{\nvdash_e} C_1$
+  \item For each $\kvar \in \operatorname{fv}(C)$,
+    there exists two constants $l_\bot,l_\top \in \mathcal L$  such that
+    $\entail{C}{\Cleq{l_\bot}{\kvar}}$ and $\entail{C}{\Cleq{\kvar}{l_\top}}$.
+  \end{enumerate}
+\end{lemma}
+% \begin{proof}
+
+%   \begin{enumerate}
+%   \item Directly follows from step (1) and (5).
+%   \end{enumerate}
+%   (1) is a direct consequence of step 
+% \end{proof}
+
+\begin{lemma}[Normal form]
+  Given a constraint $D\in\mathcal C$, a substitution $\phi$ and
+  $(C,\unif) = \normalize{D}{\phi}$,
+  then $\phi\leq\unif$,
+  $C \equivC \unif D$ and
+  $\unif C = C$.
+\end{lemma}
+\begin{proof}
   \TODO{}
+\end{proof}
+
+\begin{lemma}[Uniqueness]
+  Given $(C_1,\unif_1)$ and $(C_2,\unif_2)$ such that
+  $\unif_1 C_1 \equivC \unif_2 C_2$, then
+  $\normalize{C_1}{\unif_1}$ and $\normalize{C_2}{\unif_2}$
+  are identical up to $\alpha$-renaming.
+\end{lemma}
+\begin{proof}
+  \TODO{}
+\end{proof}
+
+\begin{theorem}
+  The procedure $\operatorname{normalize}$ is complete and computes
+  unique principal normal forms for our constraint system.
+\end{theorem}
 

### 2018-11-12 kindly/main.tex
Blip.
A bit more structure.
Add a crappy abstract.

--- 2018-01-16 kindly/main.tex
@@
-\documentclass[sigconf,author]{acmart}\settopmatter{printfolios=true}
+\documentclass[manuscript,author]{acmart}
 
@@
 \usepackage{graphicx}
+
+\usepackage{xspace}
 
@@
 \title{Kindly Bent to Free Us}
+\begin{abstract}
+  System programming often requires to manipulate resources that
+  cannot be duplicated such as file system handles, network connections
+  or even globally allocated memory.
+  Unfortunately such resources are often the root of bugs and security issues
+  through use-after-free errors or data races, even in the context
+  of programming language traditionally considered safe, such as statically typed
+  functional languages.
+  Linear and affine type systems propose to forbid such erroneous programming
+  patterns by tracking usage of variables through types.
+
+  We present \lang, an extension of ML with a kind system and qualified
+  types that allows to track affine types.
+  This language supports the defining features of ML family languages:
+  an impure, strict functional expression language with complete principal type
+  inference and type abstraction through modules.
+  Our language do not require any linearity annotation in
+  expressions and support common functional idioms.
+\end{abstract}
+
 \maketitle
 
-\begin{figure*}[h]
-  \centering
-  \input{syntax}
-  \caption{Syntax}
+\input{intro}
+\input{lang}
 
-  \input{kinds}
-  \caption{Kind rules -- $\haskind{P}{\E}{\tau}{k}$ }
+\input{infer/infer}
+\input{soa}
 
-  \input{typing}
-  \caption{Typing rules -- $\hastype{P}{\E}{e}{\tau}$ }
-
-  \input{semantics}
-  \caption{Reduction rules -- $\ered{I}{E}{e}{v}$ }
-
-\end{figure*}
+\bibliography{biblio}
 

### 2018-11-12 kindly/infer/infer.tex
Small fix for the normalization and proof of principality.

@@
 The normalization procedure is noted $\normalize{C_0}{\unif_0}$ where
-$C_0\in \mathcal C$ is a set of constraints and $\unif_0$ is a unifier.
+$C_0\in \mathcal C$ is a set of constraints and $\unif_0$ is a substitution.
 It returns a constraint $C \in \mathcal S$ in
@@
   gather all existential quantifications at the front of the constraint.
-  We obtain a constraint $C = \exists \kvar_i,\ \Cleq{k_j}{k'_j}_j$ and
-  a unifier $\unif$.
+  We obtain a constraint $C^k = \exists \kvar_i,\ \Cleq{k_j}{k'_j}_j$ and
+  a substitution $\unif_\tau$.
 
@@
   (resp. from $\kvar$ to $\glb l_j$).
-  This step is well defined since
-  our lattice is bounded and $\lub\emptyset$ and $\glb\emptyset$ are well defined.
+  This step is well defined since $\mathcal L$ is a bounded lattice
+  and $\lub\emptyset$ and $\glb\emptyset$ are well defined.
 
@@
   between related constants.
-\item For each strongly connected component in $\mathcal G$, unify all its vertices and replace it by a representative and complete the unifier accordingly.
+\item For each strongly connected component in $\mathcal G$, unify all its vertices and replace it by a representative.
+  We note $\unif_k$ the substitution that replaces a kind variable by
+  its representative.
   The representative of a strongly connected component $g$ can be determined as follows:
@@
 \item Remove all the vertices corresponding to the kind variables $\kvar_i$
-  that are existentially quantified in $C$.
+  that are existentially quantified in $C^k$.
+\item Take the transitive reduction of $\mathcal R$.
 \item Remove the extremums of $\mathcal L$ and the edges of $(\leq)$
   from $\mathcal R$.
-\item Take the transitive reduction of $\mathcal R$.
-  % Return $C = \left\{ k \leq k' \mid k \operatorname{\mathcal R}k' \right\}$
+\item Return $C = \left\{ k \leq k' \mid k \operatorname{\mathcal R}k' \right\}$
+  and $\unif =  \unif_\tau \meet \unif_k$.
 \end{enumerate}
@@
 \begin{lemma}[Normal form]
+  \label{lemma:normalform}
   Given a constraint $D\in\mathcal C$, a substitution $\phi$ and
@@
 \begin{proof}
-  \TODO{}
+  % For simplicity, we assume that any substitution has been already applied
+  % to $D$ and that $\phi = id$.
+  Let us partition $\phi$ into a part which affects type variables,
+  $\phi_\tau$, and a part which affects kind variables, $\phi_k$.
+
+  We note $(C^k,\unif_\tau)$ the result of
+  the modified Herbrand unification on $(D,\phi)$ in step (1).
+  Herbrand unification computes the most general
+  unifier. Our modified Herbrand unification only output additional
+  kind constraints for kind on the arrows and does not change
+  the result of the unification. Thus, we have
+  $\phi_\tau\leq\unif_\tau$,
+  $C^k \equivC \unif_\tau D$ and
+  $\unif_\tau C^k = C^k$.
+
+  Let $C^{k+}$ be the result after step (2), we trivially have that
+  $\fv{C^{k+}} = \fv{C^k}$ and that $C^{k+} \equivC C^k$.
+
+  Let $C^{A}$ and $\unif_k$ be the results after step (3).
+  By definition, we have $\unif_k C^{k+} \equivC C^{A}$ and
+  $\unif_k C^{A} = C^{A}$. Since $\phi_k$ has already be applied to $C$ before
+  unifying the strongly connected components,
+  we have that $\phi_k\leq\unif_k$.
+
+  Let $\unif = \unif_\tau \meet \unif_k$. Since $\unif_\tau$ and $\unif_k$
+  ave disjoint supports,
+  we have $C^{A} = \unif_\tau C^{A} \equivC \unif C^{k+} \equivC \unif C^{k} \equivC \unif D$
+  and $\unif C^{A} = C^{A}$.
+  Furthermore, $\phi_\tau \meet \phi_k \leq \unif_\tau \meet \unif_k$.
+
+  Steps (4) to (8) all preserve the free variables and the equivalence
+  of constraints, which concludes.
 \end{proof}
@@
 \begin{proof}
-  \TODO{}
+  In \cref{lemma:normalform}, we have showed that all the steps of the
+  normalization procedure preserve equivalence.
+  Since $\unif_1 C_1 \equivC \unif_2 C_2$, equivalence between
+  the two results of the normalization procedures is preserved for all steps.
+
+  We note $P(C_a)$ if for all $C = (k, k)'$
+  such that $\entail{C_a}{C}$ and $\nvdash_eC$,
+  we have $C \in {\mathcal R}_a$.
+
+  Let us note $C_1'$ and $C_2'$ the constraints after step (4). $P(C_1')$ and
+  $P(C_2')$ hold. Indeed, since $C_1'$ and $C_2'$ are only composed
+  of existential quantifications and kind inequalities, the only rules
+  that applies are transitivity and lattice inequalities.
+  After step (2) and (4), the associated relations are fully saturated for these
+  two rules, hence all inequalities that can be deduced from $C_a'$ are already
+  present in the relation.
+
+  The property $P$ is preserved by step (5) since we only remove
+  inequalities that involve existentially quantified variables. Such
+  inequalities could not be picked in $P$.
+  % In (6) we only remove
+  % trivial lattice inequalities.
+
+  Let us note $C''_a$ for $a\in\{1,2\}$ the constraints after step (5).
+  Since there are no more existential variables,
+  we have $C''_a=(k_i,k'_i)_i=\mathcal R''_a$.
+  For any $C=(k,k')$ such that $\entail{}{C}$ and $\entail{C''_a}{C}$,
+  then $C \in (\leq) \subset {\mathcal R}''_a$.
+  Indeed, the only trivial inequalities in our system are equalities of the form
+  $(\kvar, \kvar)$, which were removed in step (3) and the lattice inequalities.
+
+  Let us consider $C = (k,k') \in \mathcal R''_1$.
+  Since $C''_1\equivC C''_2$, we have $\entail{C''_2}{C}$.
+  If $\nvdash_e{C}$, by $P(C''_2)$ we have that
+  $C\in R''_2$.
+  If $\entail{}{C}$, then $C\in (\leq) \subset R''_2$.
+  We conclude that $R''_1 \subset R''_2$.
+  By symmetry, $R''_1 = R''_2$ and $C''_1 = C''_2$.
+
+  This equality is preserved by step (6) and (7)
+  since the transitive reduction of a directed acyclic graph is unique,
+  which concludes.
 \end{proof}
@@
 \begin{theorem}
-  The procedure $\operatorname{normalize}$ is complete and computes
+  The $\operatorname{normalize}$ procedure is complete and computes
   unique principal normal forms for our constraint system.

### 2018-11-12 kindly/infer/infer.tex
Remove useless property.

@@
 
-\begin{lemma}
-  Given a constraint $C\in\mathcal C$, a substitution $\unif$ and
-  $(C',\unif') = \normalize{C}{\unif}$, we have the following properties:
-  \begin{enumerate}
-  \item $C'$ is of the form $\{ k_i \leq k'_i\}_i$.
-  \item For any disjoint subformulas $C_1$ and $C_2$ of $C'$,
-    then $C_1 \operatorname{\nvdash_e} C_2$ and $C_2 \operatorname{\nvdash_e} C_1$
-  \item For each $\kvar \in \operatorname{fv}(C)$,
-    there exists two constants $l_\bot,l_\top \in \mathcal L$  such that
-    $\entail{C}{\Cleq{l_\bot}{\kvar}}$ and $\entail{C}{\Cleq{\kvar}{l_\top}}$.
-  \end{enumerate}
-\end{lemma}
-% \begin{proof}
-
-%   \begin{enumerate}
-%   \item Directly follows from step (1) and (5).
-%   \end{enumerate}
-%   (1) is a direct consequence of step 
-% \end{proof}
-
 \begin{lemma}[Normal form]

### 2018-11-12 kindly/infer/infer.tex
Slightly clearer explanations.

@@
 or constants belonging to a total bounded lattice $(\mathcal L, \leq)$ ie.,
-a lattice which admits a total order and upper and lower bounds.
+a lattice which admits a total order and upper and lower bounds ($l^\top$ and $l^\bot$).
 We note lattice elements $l$ and $\glb_i l_i$ (resp. $\lub_i l_i$)
 the greatest lower bound (resp. least upper bound) in $\mathcal L$.
+%
+Let $\mathcal C_{\mathcal L}$ the set of constraints in this lattice. We consider the
+following base entailment rules in this system:
+\begin{center}
+  \begin{mathpar}
+    \inferrule{l \leq l'}{\entail{}{\Cleq{l}{l'}}}
+    \and
+    \inferrule{}{\entail{}{\Cleq{k}{l^\top}}}
+    \and
+    \inferrule{}{\entail{}{\Cleq{l^\bot}{k}}}
+  \end{mathpar}
+\end{center}
 
 The normalization procedure is noted $\normalize{C_0}{\unif_0}$ where
-$C_0\in \mathcal C$ is a set of constraints and $\unif_0$ is a substitution.
+$C_0\in \mathcal C_{\mathcal L}$ is a set of constraints and $\unif_0$ is a substitution.
 It returns a constraint $C \in \mathcal S$ in
 solved form and a unifier $\unif$.
-Our unification algorithm proceed as follows:
+The main idea of the algorithm is to first remove all the type equalities
+by using regular Herbrand unification. After that, we only have
+a set of inequalities among kinds, which we can consider as a relation.
+We can then saturate the relation,
+unify all kinds that in the same equivalence classes to obtain
+a most general unifier on kind variables,
+remove all existentially quantified variables and
+then minimize back the relation.
+
+More precisely, we apply the following steps:
 \begin{enumerate}
@@
   Let $\unif = \unif_\tau \meet \unif_k$. Since $\unif_\tau$ and $\unif_k$
-  ave disjoint supports,
-  we have $C^{A} = \unif_\tau C^{A} \equivC \unif C^{k+} \equivC \unif C^{k} \equivC \unif D$
+  have disjoint supports,
+  we have $C^{A} = \unif_\tau C^{A} \equivC \unif C^{k+} \equivC \unif D$
   and $\unif C^{A} = C^{A}$.

### 2018-11-15 kindly/infer/infer.tex
Small fixes.

@@
 
-We now formulate type inference for the \lang language. Our inference technique is based on the HM(X) framework~\citep{DBLP:journals/tapos/OderskySW99} which
+We now formulate type inference for the \lang language. Type inference is
+based on the HM(X) framework~\citep{DBLP:journals/tapos/OderskySW99} which
 presents how to implement Hindley-Milner type inference in a language
@@
 
-We consider $\mathcal C$ the set of constrains.
-The grammar of constrains, presented
-in \cref{grammar:constraint}, follows the traditional HM(X) formulation
-with conjunctions,projections and type inequalities. The only new
-element specific to our approach are kind equalities.
+\newcommand\A{\mathcal A}
+\newcommand\SC{\mathcal S}
+
+Let us note $\A$ our constraint system. The grammar of constraints is
+given in  \cref{grammar:constraint}.
+$\A$ is defined as the smallest cylindric term constraint system that
+satisfies the axiom shown in \cref{rules:entail}.
+We follows the traditional HM(X) formulation
+with conjunctions, projections and type inequalities.
+The new element specific to our approach are kind inequalities.
 Entailment is noted $\entail{C}{D}$, where $D$ is a consequence of the
 constraints $C$.
-The base entailment rules are shown in \cref{rules:entail}, we consider
-the completion of that system to form a cylindric constraint system.
-\TODO{Give the properties}
-
 We say that $C$ and $D$ are equivalent, noted $C \equivC D$,
 when $\entail{C}{D}$ and $\entail{D}{C}$.
+\TODO{Give the cylindric properties ?}
 
@@
 
-We note $\mathcal S$ the set of solved forms
+We note $\SC$ the set of solved forms
 which can be used inside type and kind schemes.
-We define $\mathcal S$ as $\mathcal C$ quotiented by the relation $\equivC$.
+We define $\SC$ as $\A$ quotiented by the relation $\equivC$.
 %
-We consider the existence of a function $\normalize$ which takes
-a constraint in $\mathcal C$ and a unifier $\psi$ and returns a constraint
-in solved form $C' \in \mathcal S$,
-and an updated unifier. We detail the implementation
+We consider the existence of a function $\operatorname{normalize}$ which takes
+a constraint in $\A$ and a substitution $\psi$ and returns a constraint
+in solved form $C' \in \SC$,
+and an updated substitution. We detail the implementation
 of the normalization function in \cref{sec:normalize}
@@
 \newcommand\glb\bigwedge
+\newcommand\CL{{\mathcal C_{\mathcal L}}}
 
 We now place ourselves in a more general context where kinds are either variables
-or constants belonging to a total bounded lattice $(\mathcal L, \leq)$ ie.,
+or constants belonging to a total bounded lattice $(\mathcal L, \leq_{\mathcal L})$ ie.,
 a lattice which admits a total order and upper and lower bounds ($l^\top$ and $l^\bot$).
@@
 %
-Let $\mathcal C_{\mathcal L}$ the set of constraints in this lattice. We consider the
+Let $\CL$ the set of constraints in this lattice. We consider the
 following base entailment rules in this system:
@@
   \begin{mathpar}
-    \inferrule{l \leq l'}{\entail{}{\Cleq{l}{l'}}}
+    \inferrule{l \leq_{\mathcal L} l'}{\entail{}{\Cleq{l}{l'}}}
     \and
@@
 The normalization procedure is noted $\normalize{C_0}{\unif_0}$ where
-$C_0\in \mathcal C_{\mathcal L}$ is a set of constraints and $\unif_0$ is a substitution.
+$C_0\in \CL$ is a set of constraints and $\unif_0$ is a substitution.
 It returns a constraint $C \in \mathcal S$ in
@@
 
+\subsubsection{Soundness and Principality}
+
 \begin{lemma}[Normal form]
   \label{lemma:normalform}
-  Given a constraint $D\in\mathcal C$, a substitution $\phi$ and
+  Given a constraint $D\in\CL$, a substitution $\phi$ and
   $(C,\unif) = \normalize{D}{\phi}$,
@@
 \begin{theorem}
-  The $\operatorname{normalize}$ procedure is complete and computes
-  unique principal normal forms for our constraint system.
+  $\CL$ has the principal constraint property and
+  $\operatorname{normalize}$ computes principal normal forms for $\CL$.
+\end{theorem}
+
+\begin{lemma}
+  For all $C\in\mathcal S$, $\entail{C}{x = x}$ implies
+  $\entail{}{x = x}$.
+\end{lemma}
+\begin{proof}
+  By definition of $\operatorname{normalize}$, We have $C = \Cleq{k}{k'}_i$
+  such that the underlying relation has no cycles.
+  Thus, we can not deduce neither kind nor type equalities from $C$.
+\end{proof}
+
+\begin{lemma}
+  $\CL$ is regular, ie, for $x, x'$ two types or kinds,
+  $\entail{}{\Ceq{x}{x'}}$ implies
+  $\fv{x} = \fv{x'}$
+\end{lemma}
+\begin{proof}
+  The only equalities possibles are between variables (via symmetry) or
+  between constants.
+\end{proof}
+
+\begin{theorem}
+  Type inference in $HM(\CL)$ is sound, principal and complete.
 \end{theorem}

### 2018-11-16 kindly/infer/infer.tex
minor tweaks for readability

@@
 
-\begin{figure}[h]
+\begin{figure}[tp]
   \centering
@@
 
-\begin{figure*}[h]
+\begin{figure*}[tp]
   \input{infer/entails}
@@
 In order to determine if a variable is used in an affine manner, we must track
-it's uses and the associated kinds. For instance, in the expression
+its uses and the associated kinds. For instance, in the expression
 $(x,x)$, $x$ is used twice. If $x$ is of type $\tau$, which is of kind $k$,
@@
 We note $\Sone{x}{k_1,\dots,k_n}$ the singleton usage map.
-Common operations on usages maps are defined in \cref{op:usgmap}. We define
+Common operations on usage maps are defined in \cref{op:usgmap}. We define
 regular set operations such as union
@@
 
-\begin{figure}[h]\centering
+\begin{figure}[tp]\centering
 \begin{align*}

### 2019-02-20 kindly/motivation.tex
motivational section added

\section{Linearity, Affinity, and Borrows at Work}

\TODO{really good examples. first linearity/affinity (files?), then
  something with borrows (functional arrays), and then something
  really spaced out that pushes the limit}

%%% Local Variables:
%%% mode: latex
%%% TeX-master: "main"
%%% End:

### 2019-02-20 kindly/infer/infer.tex
Inference rules for regions.

@@
 
-\begin{lemma}
-  Without loss of generality, we can consider that
-  kind inequalities in satisfiable constraints
-  are only done on kind variables.
+% \begin{lemma}
+%   Without loss of generality, we can consider that
+%   kind inequalities in satisfiable constraints
+%   are only done on kind variables.
 
-  \begin{proof}
-    Consider the constraint $\Cleq{k}{k'}\Cand C$.
-    \begin{itemize}
-    \item If $k$ and $k'$ are both constants, it can be removed.
-    \item If the constraint is $\Cleq{\kvar}{\kaff}$ or $\Cleq{\kun}{\kvar}$, it can be removed.
-    \item If the constraint is $\Cleq{\kvar}{\kun}$ or $\Cleq{\kaff}{\kvar}$, we
-      can substitute $\kvar$ by its value in $C$.
-    \end{itemize}
-  \end{proof}
-\end{lemma}
+%   \begin{proof}
+%     Consider the constraint $\Cleq{k}{k'}\Cand C$.
+%     \begin{itemize}
+%     \item If $k$ and $k'$ are both constants, it can be removed.
+%     \item If the constraint is $\Cleq{\kvar}{\kaff}$ or $\Cleq{\kun}{\kvar}$, it can be removed.
+%     \item If the constraint is $\Cleq{\kvar}{\kun}$ or $\Cleq{\kaff}{\kvar}$, we
+%       can substitute $\kvar$ by its value in $C$.
+%     \end{itemize}
+%   \end{proof}
+% \end{lemma}
 

### 2019-02-20 kindly/soa.tex
vault

@@
-\section{Related works}
+\section{Related work}
 
@@
 Alms~\citep{DBLP:conf/popl/TovP11},
-F$^\circ$~\citep{DBLP:conf/tldi/MazurakZZ10},
+F\degree~\citep{DBLP:conf/tldi/MazurakZZ10},
 Quill~\citep{DBLP:conf/icfp/Morris16},
-Linear Haskell~\citep{DBLP:journals/pacmpl/BernardyBNJS18}
+Linear Haskell~\citep{DBLP:journals/pacmpl/BernardyBNJS18},
+Vault\citep{DBLP:conf/pldi/DeLineF01,DBLP:conf/pldi/FahndrichD02}
 

### 2019-02-21 kindly/infer/infer.tex
Add typing for pairs and complete the kinds/lattices.

@@
 \begin{figure*}[tp]
+  \input{infer/lattice}
+  \caption{Lattice inequalities -- $k \lk_\Lat k'$}
   \input{infer/entails}
@@
 \begin{figure}[tp]\centering
-\begin{align*}
-  \Sv\cap\Sv'
-  &= \left\{ x \to l,l'
-    \mid \Sone{x}{l}\in\Sv \text{ and } \Sone{x}{l'}\in\Sv' \right\} \\
-  \Sv\cup\Sv'
-  &= \Sv\cap\Sv' \sqcup
-    (\Sv\Sdel{x' \mid x'\in\Sv'}) \sqcup
-    (\Sv'\Sdel{x \mid x\in\Sv})
-  &\text{where $\sqcup$ is the disjoint union}\\
-  \Cleq{\Sv}{k_0}
-  &= \bigwedge_{\Sone{x}{l_x}\in\Sv} \bigwedge_{k\in l_x} \Cleq{k}{k_0}
-\end{align*}
+\begin{mathpar}
+  % \Sv\cap\Sv'
+  % &= \left\{ x \to l,l'
+  %   \mid \Sone{x}{l}\in\Sv \text{ and } \Sone{x}{l'}\in\Sv' \right\} \\
+  \infer
+  { \forall x \in \Dom{\Sv_1}\cup\Dom{\Sv_2}.\quad
+    \tsplit{C_x}{\Sv(x)}{\Sv_1(x)}{\Sv_2(x)}
+  } 
+  { \tsplit{\bigwedge C_x}{\Sv}{\Sv_1}{\Sv_2} }
+  % &= \Sv\cap\Sv' \sqcup
+  %   (\Sv\Sdel{x' \mid x'\in\Sv'}) \sqcup
+  %   (\Sv'\Sdel{x \mid x\in\Sv})
+  %   &\text{where $\sqcup$ is the disjoint union}\\
+  \and
+  \inferrule{}
+  { \Cleq{\Sv}{k_0}
+    = \bigwedge_{\Sone{x}{l_x}\in\Sv} \bigwedge_{k\in l_x} \Cleq{k}{k_0}
+  }
+\end{mathpar}
 \caption{Operation on usage maps $\Sv$}
@@
 
-\newcommand\lub\bigvee
-\newcommand\glb\bigwedge
-\newcommand\CL{{\mathcal C_{\mathcal L}}}
-
 We now place ourselves in a more general context where kinds are either variables
-or constants belonging to a total bounded lattice $(\mathcal L, \leq_{\mathcal L})$ ie.,
+or constants belonging to a total bounded lattice $(\mathcal L, \lk_\Lat)$ ie.,
 a lattice which admits a total order and upper and lower bounds ($l^\top$ and $l^\bot$).
@@
   \begin{mathpar}
-    \inferrule{l \leq_{\mathcal L} l'}{\entail{}{\Cleq{l}{l'}}}
+    \inferrule{l \lk_\Lat l'}{\entail{}{\Cleq{l}{l'}}}
     \and

### 2019-02-27 kindly/infer/infer.tex
Smooth over the borrow-related bits.
snapshot
Add checking during normalization.

@@
   } 
-  { \tsplit{\bigwedge C_x}{\Sv}{\Sv_1}{\Sv_2} }
+  { \tsplit{(\bigwedge C_x)}{\Sv}{\Sv_1}{\Sv_2} }
   % &= \Sv\cap\Sv' \sqcup
@@
     = \bigwedge_{\Sone{x}{l_x}\in\Sv} \bigwedge_{k\in l_x} \Cleq{k}{k_0}
+  }
+  \and
+  \inferrule
+  { \forall \bshadow{x} \in \Sv, b'\in\Sv'.\quad
+    \tregion{C_x}{y}{\bshadow{x}}{b'} }
+  { \tregion{(\bigwedge C_x)}{y}{\Sv}{\Sv'} }
+  \and
+  % \inferrule
+  % { \text{if } \operatorname{lenght}(\Sv(x))  = \Cempty{} \wedge  }
+  { \Weaken_{(x,k)}(\Sv) =
+    \begin{dcases}
+      \Ctrue& \operatorname{length}(\Sv(x)) = 1\\
+      {k}\lk\kaff_\infty& \text{ otherwise}\\
+    \end{dcases}
+
   }
@@
 
-\begin{figure}[h]
+\begin{figure}[ht]
   \centering
   \input{infer/kinds}
-  \caption{Kind inference rules -- $\inferK{(C,\unif)}{\E}{\tau}{k}$}
+  \caption{Kind inference algorithm -- $\inferK{(C,\unif)}{\E}{\tau}{k}$}
   \label{rules:kinding}
+\end{figure}
+
+\begin{figure}[ht]
+  \centering
+  \input{infer/sd-kinds}
+  \caption{Syntax-directed kind inference -- $\inferSK{C}{\E}{\tau}{k}$}
+  \label{rules:sd-kinding}
 \end{figure}
@@
   \input{infer/typing}
-  \caption{Type Inference rules -- $\inferW{\Sigma}{(C,\psi)}{\bf{\E}}{\bf{e}}{\tau}$ }
+  \caption{Type inference algorithm -- $\inferW{\Sigma}{(C,\psi)}{\bf{\E}}{\bf{e}}{\tau}$ }
   \label{rules:typing}
@@
 
+\begin{figure*}[ht]
+  \input{infer/sd-typing}
+  \caption{Syntax-directed typing rules}
+  \label{fig:syntax-directed-typing}
+\end{figure*}
 \subsection{Normal forms and generalization}
@@
   between related constants.
+\item
+  At this point, we can easily check for satisfiability: A constraint
+  is satisfiable (in the given environment) if and only if,
+  for any constants $l_1$ and $l_2$ such that
+  there is a path from $l_1$ to $l_2$ in $\mathcal G$, then $l_1\lk_\Lat l_2$.
+  If this is not the case, we return \textbf{fail}.
+
 \item For each strongly connected component in $\mathcal G$, unify all its vertices and replace it by a representative.
@@
 
-  Let $C^{A}$ and $\unif_k$ be the results after step (3).
+  Let $C^{A}$ and $\unif_k$ be the results after step (4).
   By definition, we have $\unif_k C^{k+} \equivC C^{A}$ and
@@
 
-  Steps (4) to (8) all preserve the free variables and the equivalence
+  Steps (5) to (9) all preserve the free variables and the equivalence
   of constraints, which concludes.
@@
   that applies are transitivity and lattice inequalities.
-  After step (2) and (4), the associated relations are fully saturated for these
+  After step (2) and (5), the associated relations are fully saturated for these
   two rules, hence all inequalities that can be deduced from $C_a'$ are already
@@
 
-  The property $P$ is preserved by step (5) since we only remove
+  The property $P$ is preserved by step (6) since we only remove
   inequalities that involve existentially quantified variables. Such
@@
   Indeed, the only trivial inequalities in our system are equalities of the form
-  $(\kvar, \kvar)$, which were removed in step (3) and the lattice inequalities.
+  $(\kvar, \kvar)$, which were removed in step (4) and the lattice inequalities.
 
@@
 
-  This equality is preserved by step (6) and (7)
+  This equality is preserved by step (7) and (8)
   since the transitive reduction of a directed acyclic graph is unique,

### 2019-03-11 kindly/motivation.tex
Start working on the examples.

@@
 
+\subsection{A session on linearity}
+
+A classical example of linearity at work are to control the uses of resources such
+as files or network handles.
+%
+Session types, notably, allow to statically type check protocols.
+For instance, the type \lstinline{!int.!int.?int.end} represent a session
+where the program must send two integers, receive an integer, then end the session.
+%
+In this context, it is essential that channels
+are used linearly, since every use of the channel
+will make the protocol progress and change the type of the channel.
+
+Session types are usually defined with the following operators:
+
+\begin{center}
+  \begin{tabular}{r l}
+    $\tau ? \mathcal S$ & Send a value of type $\tau$ then continue with protocol $S$.\\
+    $\tau ? \mathcal S$& Receive a value of type $\tau$ then continue with protocol $S$.\\
+    $\mathcal S \oplus \mathcal S'$& Choose between protocols $S$ and $S'$.\\
+    $\mathcal S \operatorname{\&} \mathcal S'$& Offer a choice between protocols $S$ and $S'$.
+  \end{tabular}
+\end{center}
+
+It is well known that binary session types can be embedded into ML languages
+\cite{DBLP:journals/jfp/Padovani17}. We can easily
+embed session types in our language and provide the following API:
+
+\begin{lstlisting}
+type 'S st : lin
+val receive: ('a ? 'S) st -> 'a * 'S st
+val send : ('a:'k). 'a -> ('a ! 'S) st -{'k}> 'S st
+val close : end st -> unit
+\end{lstlisting}
+
+\TODO{Explain}
+
+We can then use these primitives to implement a simple addition service.
+We do not need any extra linearity annotations in the code: the fact
+that the \texttt{st} type is declared as linear is sufficient.
+In the case below, \lstinline{add_service} is inferred to be of type
+\lstinline{(int ! int ! int ? end) st -> unit}
+
+\begin{lstlisting}
+let add_service ep =
+  let (x, ep) = receive ep in
+  let (y, ep) = receive ep in
+  let ep = send (x + y) ep in
+  close ep
+\end{lstlisting}
+
+% \begin{figure}
+%   \centering
+%   \begin{lstlisting}
+% val select : ('S st -{'k}> 'a) -> 'a ot -{'k}> ('S dual) st
+% val branch : 'm it -> 'm
+%   \end{lstlisting}
+%   \caption{Session primitives}
+%   \label{api:session}
+% \end{figure}
+
 %%% Local Variables:

### 2019-05-17 kindly/main.tex
motivational section added
vault
Add typing for pairs and complete the kinds/lattices.
Start working on the examples.
abstract take one

@@
-\documentclass[manuscript,author]{acmart}
+\documentclass[acmsmall,review,anonymous,table]{acmart}
+\settopmatter{printfolios=true,printccs=false,printacmref=false}
 
@@
 \usepackage{mathtools}
-% \usepackage{textcomp}
+\usepackage{textcomp}
+\usepackage{gensymb}
 % \usepackage{textgreek}
+
+\usepackage{listings}
+\usepackage[scaled=0.85]{DejaVuSansMono}
+
+\usepackage{tikz}
+\usetikzlibrary{arrows}
 
@@
 \usepackage{mathpartir} % Syntax trees
+
 \usepackage[nameinlink,capitalize]{cleveref}
+\addto\extrasenglish{%
+  \def\sectionautorefname{Section}
+  \def\subsectionautorefname{Section}
+  \def\subsubsectionautorefname{Section}
+  \def\lstnumberautorefname{Line}
+}
 
@@
 \input{prelude}
+\input{caml}
 \input{notations}
@@
   System programming often requires to manipulate resources that
-  cannot be duplicated such as file system handles, network connections
-  or even globally allocated memory.
-  Unfortunately such resources are often the root of bugs and security issues
-  through use-after-free errors or data races, even in the context
-  of programming language traditionally considered safe, such as statically typed
-  functional languages.
-  Linear and affine type systems propose to forbid such erroneous programming
-  patterns by tracking usage of variables through types.
+  cannot be duplicated such as file system handles, network connections,
+  or globally allocated memory.
+  Improper handling of such resources is often the cause of bugs and security issues
+  through use-after-free errors or data races. These problems are not
+  prevented by most statically typed programming languages that only
+  guarantee type soundness and memory safety.
+  Linear and affine type systems support tracking of single-threaded resource usage
+  through types, but they are often too restrictive for real-world applications.
 
@@
 \input{intro}
+\input{motivation}
 \input{lang}

### 2019-05-27 kindly/infer/infer.tex
quantify over kvars
semantics & syntax-driven typing
inconsistencies fixed
Get rid of one notion of spliting.
Simplify/factor the region/borrowing rules.
Use types in use-maps.
Use the (\E <= k) syntax everywhere.
Reordering.

@@
 \subsection{Preliminaries}
-
-\subsubsection{Constraint language}
-
-\newcommand\A{\mathcal A}
-\newcommand\SC{\mathcal S}
-
-Let us note $\A$ our constraint system. The grammar of constraints is
-given in  \cref{grammar:constraint}.
-$\A$ is defined as the smallest cylindric term constraint system that
-satisfies the axiom shown in \cref{rules:entail}.
-We follows the traditional HM(X) formulation
-with conjunctions, projections and type inequalities.
-The new element specific to our approach are kind inequalities.
-Entailment is noted $\entail{C}{D}$, where $D$ is a consequence of the
-constraints $C$.
-We say that $C$ and $D$ are equivalent, noted $C \equivC D$,
-when $\entail{C}{D}$ and $\entail{D}{C}$.
-\TODO{Give the cylindric properties ?}
-
-\begin{figure}[tp]
-  \centering
-  \begin{align*}
-    C ::=&\ \Cleq{\tau_1}{\tau_2}\ |\ \Cleq{k_1}{k_2}\ |\ C_1 \Cand C_2\ |\ \Cproj{\alpha}{C}
-  \end{align*}
-  \caption{The constraint language}
-  \label{grammar:constraint}
-\end{figure}
-
-\begin{figure*}[tp]
-  \input{infer/lattice}
-  \caption{Lattice inequalities -- $k \lk_\Lat k'$}
-  \input{infer/entails}
-  \caption{Base entailment rules -- $\entail{C}{D}$ }
-  \label{rules:entail}
-\end{figure*}
-
-We note $\SC$ the set of solved forms
-which can be used inside type and kind schemes.
-We define $\SC$ as $\A$ quotiented by the relation $\equivC$.
-%
-We consider the existence of a function $\operatorname{normalize}$ which takes
-a constraint in $\A$ and a substitution $\psi$ and returns a constraint
-in solved form $C' \in \SC$,
-and an updated substitution. We detail the implementation
-of the normalization function in \cref{sec:normalize}
-
-% $\mathcal S$ is composed only of kind
-% inequalities \emph{over variables}. For convenience, if $C\in\mathcal S$, we
-% note $C$ as a list of kind inequalities: $\Cleq{\kvar_i}{\kvar_{i'}}^n$.
-% \TODO{Extend the properties of solved forms}
 
@@
   %   \mid \Sone{x}{l}\in\Sv \text{ and } \Sone{x}{l'}\in\Sv' \right\} \\
-  \infer
-  { \forall x \in \Dom{\Sv_1}\cup\Dom{\Sv_2}.\quad
-    \tsplit{C_x}{\Sv(x)}{\Sv_1(x)}{\Sv_2(x)}
-  } 
-  { \tsplit{(\bigwedge C_x)}{\Sv}{\Sv_1}{\Sv_2} }
+  % \infer
+  % { \forall x \in \Dom{\Sv_1}\cup\Dom{\Sv_2}.\quad
+  %   \bsplit{C_x}{\Sv(x)}{\Sv_1(x)}{\Sv_2(x)}
+  % } 
+  % { \bsplit{(\bigwedge C_x)}{\Sv}{\Sv_1}{\Sv_2} }
   % &= \Sv\cap\Sv' \sqcup
@@
   %   &\text{where $\sqcup$ is the disjoint union}\\
-  \and
+  % \and
   \inferrule{}
-  { \Cleq{\Sv}{k_0}
-    = \bigwedge_{\Sone{x}{l_x}\in\Sv} \bigwedge_{k\in l_x} \Cleq{k}{k_0}
+  { \Cleq{\Sv}{k}
+    = \bigwedge_{\Sone{x}{\schm}\in\Sv} \Cleq{\schm}{k}
   }
   \and
-  \inferrule
-  { \forall \bshadow{x} \in \Sv, b'\in\Sv'.\quad
-    \tregion{C_x}{y}{\bshadow{x}}{b'} }
-  { \tregion{(\bigwedge C_x)}{y}{\Sv}{\Sv'} }
-  \and
+  % \inferrule
+  % { \forall \bshadow{x} \in \Sv, b'\in\Sv'.\quad
+  %   \tregion{C_x}{y}{\bshadow{x}}{b'} }
+  % { \bregion{(\bigwedge C_x)}{y}{\Sv}{\Sv'} }
+  % \and
   % \inferrule
@@
   \input{infer/kinds}
-  \caption{Kind inference algorithm -- $\inferK{(C,\unif)}{\E}{\tau}{k}$}
+  \caption{Kind inference algorithm -- $\inferK{(C,\unif)}{\E}{\tau}{(\Multi{k}) k}$}
   \label{rules:kinding}
-\end{figure}
-
-\begin{figure}[ht]
-  \centering
-  \input{infer/sd-kinds}
-  \caption{Syntax-directed kind inference -- $\inferSK{C}{\E}{\tau}{k}$}
-  \label{rules:sd-kinding}
 \end{figure}
@@
 
-\begin{figure*}[h]
+\begin{figure*}[htp]
   \input{infer/typing}
@@
 
-\begin{figure*}[ht]
-  \input{infer/sd-typing}
-  \caption{Syntax-directed typing rules}
-  \label{fig:syntax-directed-typing}
-\end{figure*}
 \subsection{Normal forms and generalization}

### 2019-05-27 kindly/lang.tex
minor tweaks for readability
Inference rules for regions.
semantics
Add typing for pairs and complete the kinds/lattices.
Smooth over the borrow-related bits.
update
Get rid of one notion of spliting.
Simplify/factor the region/borrowing rules.
Reordering.

@@
 $\T{t}$ and arrow types annotated with a kind, noted $\tau \tarr{k} \tau$.
-Kinds can be either a kind variable $\kvar$, affine $\kaff$ or unrestricted $kun$.
+Kinds can be either a kind variable $\kvar$, affine $\kaff$ or unrestricted $\kun$.
 Type schemes and kind schemes are guarded by a set of constraints.
@@
 
-\begin{figure*}[h]
+\begin{figure*}[t]
   \centering
@@
 \end{figure*}
+\input{internal-grammar}
 
+\clearpage
 \subsection{Semantics}
+\input{fig-reduction}
 
-\TODO{}
-% \begin{figure*}[h]
-%   \input{semantics}
-%   \caption{Reduction rules -- $\ered{I}{E}{e}{v}$ }
-% \end{figure*}
-
+\clearpage
 \subsection{Typing}
 
-\TODO{Add system-F like type rules}
+\subsubsection{Constraint language}
 
+\newcommand\A{\mathcal A}
+\newcommand\SC{\mathcal S}
+
+Let us note $\A$ our constraint system. The grammar of constraints is
+given in  \cref{grammar:constraint}.
+$\A$ is defined as the smallest cylindric term constraint system that
+satisfies the axiom shown in \cref{rules:entail}.
+We follows the traditional HM(X) formulation
+with conjunctions, projections and type inequalities.
+The new element specific to our approach are kind inequalities.
+Entailment is noted $\entail{C}{D}$, where $D$ is a consequence of the
+constraints $C$.
+We say that $C$ and $D$ are equivalent, noted $C \equivC D$,
+when $\entail{C}{D}$ and $\entail{D}{C}$.
+\TODO{Give the cylindric properties ?}
+
+\begin{figure}[tp]
+  \centering
+  \begin{align*}
+    C &::= \Cleq{\tau_1}{\tau_2}
+        \mid \Cleq{k_1}{k_2}
+        \mid C_1 \Cand C_2
+        \mid \Cproj{\tvar}{C}
+        \mid \Cproj{\kvar}{C}
+  \end{align*}
+  \caption{The constraint language}
+  \label{grammar:constraint}
+\end{figure}
+
+\begin{figure*}[tp]
+  \input{infer/lattice}
+  \caption{Lattice inequalities -- $k \lk_\Lat k'$}
+  \input{infer/entails}
+  \caption{Base entailment rules -- $\entail{C}{D}$ }
+  \label{rules:entail}
+\end{figure*}
+
+We note $\SC$ the set of solved forms
+which can be used inside type and kind schemes.
+We define $\SC$ as $\A$ quotiented by the relation $\equivC$.
+%
+We consider the existence of a function $\operatorname{normalize}$ which takes
+a constraint in $\A$ and a substitution $\psi$ and returns a constraint
+in solved form $C' \in \SC$,
+and an updated substitution. We detail the implementation
+of the normalization function in \cref{sec:normalize}
+
+% $\mathcal S$ is composed only of kind
+% inequalities \emph{over variables}. For convenience, if $C\in\mathcal S$, we
+% note $C$ as a list of kind inequalities: $\Cleq{\kvar_i}{\kvar_{i'}}^n$.
+% \TODO{Extend the properties of solved forms}
+
+\subsubsection{Kinding}
+
+\begin{figure}[ht]
+  \centering
+  \input{infer/sd-kinds}
+  \caption{Syntax-directed kind inference --
+    $\inferSK{C}{\E}{\tau}{(\Multi k)k}$}
+  \label{rules:sd-kinding}
+\end{figure}
+
+\subsubsection{Environments}
+
+\begin{figure}[tp]
+  \input{infer/sd-splitting}
+  \caption{Splitting --- environments $\lsplit
+    C\E\E\E$; binders $\bsplit Cbbb$}
+  \label{fig:sd-splitting}
+\end{figure}
+
+\begin{figure}[tp]
+  \input{infer/sd-borrowing}
+  \caption{Borrowing --- environments $\lsplit
+    C\E\E\E$; binders $\bsplit Cbbb$}
+  \label{fig:sd-borrowing}
+\end{figure}
+
+\subsubsection{Typing}
+
+\begin{figure*}[htp]
+  \input{infer/sd-typing}
+  \caption{Syntax-directed typing rules}
+  \label{fig:syntax-directed-typing}
+\end{figure*}
+
+\clearpage
 %%% Local Variables:

### 2019-05-31 kindly/motivation.tex
no blanking in store
Misc.
Reordering.
Keep working on the examples.

@@
   \begin{tabular}{r l}
-    $\tau ? \mathcal S$ & Send a value of type $\tau$ then continue with protocol $S$.\\
+    $\tau ! \mathcal S$ & Send a value of type $\tau$ then continue with protocol $S$.\\
     $\tau ? \mathcal S$& Receive a value of type $\tau$ then continue with protocol $S$.\\
@@
 embed session types in our language and provide the following API:
-
+%
 \begin{lstlisting}
@@
 val send : ('a:'k). 'a -> ('a ! 'S) st -{'k}> 'S st
+val create : unit -> 'S st * (dual 'S) st
 val close : end st -> unit
@@
 let add_service ep =
-  let (x, ep) = receive ep in
-  let (y, ep) = receive ep in
+  let x, ep = receive ep in
+  let y, ep = receive ep in
   let ep = send (x + y) ep in
   close ep
+# add_service : (int ! int ! int ? end) st -> unit
+\end{lstlisting}
+
+The following code then acts as a client, by sending two messages
+and receiving the result.
+This code is polymorphic in both argument and return types and could thus
+be used to any binary operator.
+%
+\begin{lstlisting}
+let op_client ep x y =
+  let ep = send x ep in
+  let ep = send y ep in
+  let result, ep = receive ep in
+  close ep;
+  result
+# op_client : ('a_1 ? 'a_2 ? 'b ! end) st -> 'a_1 -{lin}> 'a_2 -{lin}> 'b
+\end{lstlisting}
+% The actual answer from the typechecker looks like this:
+% # ('m2:^k2), ('m1:^k1).
+% # (^k1 < ^k) & (lin < ^k) =>
+% # ('m1 ?'m2 ? 'result ! end) st -> 'm1 -{lin}> 'm2 -{^k}> 'result
+
+In addition, this function can be partially
+applied to a channel. Since the closure returned by such a partial
+application captures the channel, it can only be used once.
+This is reflected by the arrow, \lstinline/-{lin}>/, which is the
+type of functions that can only be used once.
+The general form of arrows types in \lang is \lstinline/-{k}>/, where
+\lstinline/kk/ is an arbitrary kind that describes the behavior of
+underlying closure.
+For convenience, we shorten \lstinline/-{un}>/ as \lstinline/->/.
+Note again that the linearity on the arrows was inferred without
+any annotation from the user. In fact, the only difference between the code
+presented so far and the example by
+\citet{DBLP:journals/jfp/Padovani17}
+is the kind annotation on the type definition of \lstinline/st/.
+
+\TODO{more details?}
+
+Finally, we can compose both server and client.
+Failure to fully consume either \lstinline/a/ or \lstinline/b/
+would result in a type error.
+\begin{lstlisting}
+let main1 () =
+  let (a, b) = create () in
+  add_service a;
+  op_client b 1 2
+# main : unit -> int
 \end{lstlisting}
@@
 
+\subsection{Hybrid paradigms}
+
+The programming style used for channels so far, which are passed back
+and forth between functions, is appropriate for purely functional code.
+This style is however fairly bothersome in imperative contexts.
+
+ML languages are often characterised by their hybrid programming style
+that mixes functional styles with imperative side effects. In
+this context, the programming style used for channels is fairly verbose.
+Similarly to Rust, we use {\em borrows} to allow easy imperative
+programming with linear and affine resources.
+
 %%% Local Variables:

### 2019-06-03 kindly/motivation.tex
First shot at explaining imperative prog.

@@
 any annotation from the user. In fact, the only difference between the code
-presented so far and the example by
+presented so far and the examples by
 \citet{DBLP:journals/jfp/Padovani17}
@@
 
-The programming style used for channels so far, which are passed back
-and forth between functions, is appropriate for purely functional code.
-This style is however fairly bothersome in imperative contexts.
+The programming style used for channels so far is
+appropriate for purely functional code.
+This style is however fairly bothersome in mutable contexts.
+%
+To make linearity easier to use for imperative programming, we rely on
+{\em borrows}. We present a typical API for mutable affine arrays
+in \cref{sig:array} and an example of use in \cref{ex:array}.
+We can distinguish three type of operations. Firstly,
+function that manipulate the whole
+array such as \lstinline/create/ and \lstinline/free/, and behaves exactly
+like the functions on channel presented in the previous section.
+Second, read-only functions such as \lstinline/get/
+manipulates \emph{immutable} borrows. An immutable borrow over array \lstinline/a/
+is introduced with \lstinline/&a/ and represent a read-only view of the array.
+Finally, mutable functions, such as \lstinline/set/ manipulate
+a \emph{mutable borrow}. Mutable borrows, introduced with \lstinline/&!a/
+represent a read-write view of the array.
 
-ML languages are often characterised by their hybrid programming style
-that mixes functional styles with imperative side effects. In
-this context, the programming style used for channels is fairly verbose.
-Similarly to Rust, we use {\em borrows} to allow easy imperative
-programming with linear and affine resources.
+Multiple immutable borrows can be used concurrently and are considered
+unrestricted (\lstinline/un/).
+A mutable borrow is exclusive, and considered affine (\lstinline/aff/).
+Furthermore, any borrow prevents access to the ``raw'' access
+as long as they are in use. In \cref{ex:array}, on \cref{ex:array:get},
+two immutable borrows are used in the same expression. Once this operation
+is done, the borrows are not used anymore, which allows us to take
+an exclusive immutable borrow to mutate the array on \cref{ex:array:set}.
+Finally we can return the whole array on \cref{ex:array:return} since
+none of the borrows are used anymore.
+
+\begin{figure}
+  \centering
+  \begin{subfigure}{0.48\linewidth}
+\begin{lstlisting}
+module Array : sig
+  type ('a : un) t : aff
+
+  val create : ('a : un) => int -> 'a -> 'a t
+  val free : 'a t -> unit
+
+  val length : &('a t) -> int
+
+  val get : &('a t) * int -> 'a
+  val set : &!('a t) * int * 'a -> unit
+end
+\end{lstlisting}
+    \caption{Affine primitives}
+    \label{sig:array}
+  \end{subfigure}\hfill
+  \begin{subfigure}{0.5\linewidth}
+\begin{lstlisting}
+let mk_fib_array n =
+  let a = create n 1 in
+  for i = 2 to n - 1 do
+    let x1 = get &a (i-1) + get &a (i-2) in(*@\label{ex:array:get}*)
+    set &!a i(*@\label{ex:array:set}*)
+  done;
+  a(*@\label{ex:array:return}*)
+# mk_fib_array : int -> int Array.t
+\end{lstlisting}
+    \caption{Example of use of affine arrays}
+    \label{ex:array}
+  \end{subfigure}
+\end{figure}
+
+\begin{figure}
+  \centering
+  \begin{subfigure}{0.5\linewidth}
+\begin{lstlisting}
+let for f a  = 
+  let n = Array.length &a in
+  let rec loop f a i = 
+    if i < n then (
+      f &!a i;
+      loop f &!a (i+1)
+    )
+  in
+  loop f &!a 0
+# Borrowed m ('a Array.t) tt =>
+  (tt -> int -> unit) -> tt -> unit
+\end{lstlisting}
+    \caption{For loop function for affine arrays}
+    \label{ex:array}
+  \end{subfigure}
+\end{figure}
 

### 2019-06-06 kindly/motivation.tex
More motivations

@@
 
-\subsection{Hybrid paradigms}
+\subsection{Imperative programming}
 
@@
   for i = 2 to n - 1 do
-    let x1 = get &a (i-1) + get &a (i-2) in(*@\label{ex:array:get}*)
-    set &!a i(*@\label{ex:array:set}*)
+    let x = get &a (i-1) + get &a (i-2) in(*@\label{ex:array:get}*)
+    set &!a i x(*@\label{ex:array:set}*)
   done;
@@
   \end{subfigure}
+\end{figure}
+
+Naturally, \lang also checks that borrows do not escape their point
+of use through {\em syntactic regions}. The example
+above is reproduce in \cref{ex:array:region} with
+regions explicitly delimited by \lstinline/{| ... |}/.
+\lang ensures that borrows never escape the smallest enclosing
+region.
+For instance, On \cref{ex:array:region:get}, \lstinline/&a/ does
+not escape the region since the borrow is not captured, and the
+region only return integers.
+Regions annotations are automatically inserted by \lang using
+syntactic information such as borrows and binders.
+Users can also annotate regions manually to further restrict the range
+of a borrow. In practice, we have found that regions
+annotations are very rarely needed in common imperative usage.
+
+\begin{figure}
+  \centering
+  % \begin{subfigure}{0.6\linewidth}
+\begin{lstlisting}
+let mk_fib_array n =
+  let a = create n 1 in
+  for i = 2 to n - 1 do {|
+    let x = {| get &a (i-1) + get &a (i-2) |} in(*@\label{ex:array:region:get}*)
+    set &!a i x
+  |}
+  done;
+  a(*@\label{ex:array:return}*)
+# mk_fib_array : int -> int Array.t
+\end{lstlisting}
+    \caption{Example from \cref{ex:array} with explicit regions}
+    \label{ex:array:region}
+  % \end{subfigure}
 \end{figure}
@@
 
+\section{Hybrid programming styles}
+
+While purely functional data-structure present numerous advantages in term
+of correctness and ease of use, they sometimes offer worse performances
+than their imperative counterpart. On the other hand, reasoning about imperative
+data-structure in the context of sharing is notoriously difficult.
+%
+In this section, we show how linearity and borrowing allows us to define
+a safe API for hybrid arrays that support both immutable and mutable features,
+and how it can be used to implement a backtracking sudoku solver.
+
+\begin{figure}
+  \centering
+  \begin{subfigure}{0.48\linewidth}
+\begin{lstlisting}
+module type HybArray = sig
+  type ('a : un) t : aff
+
+  val create : ('a : un) => int -> 'a -> 'a t
+  val free : 'a t -> unit
+
+  val length : &('a t) -> int
+  val get : &('a t) * int -> 'a
+
+  val set : &('a t) * int * 'a -> 'a t
+  val set_immut : &!('a t) * int * 'a -> unit
+end
+\end{lstlisting}
+    \caption{Signature}
+    \label{sig:hybarray}
+  \end{subfigure}\hfill
+  \begin{subfigure}{0.5\linewidth}
+\begin{lstlisting}
+module Cow : HybArray = struct
+  include Array
+
+  let set (a, i, x) =
+    let f k =
+      if k = i then x else Array.get (a,k)
+    in
+    create (length a) f
+
+  let set_mut = Array.set
+end
+\end{lstlisting}
+    \caption{Implementation using copy-on-write}
+    \label{ex:cow}
+  \end{subfigure}
+  \caption{Hybrid arrays -- signature and example implementation}
+\end{figure}
+
+\begin{figure}
+  \centering
+\begin{lstlisting}[multicols=2]
+type board = IntSet.t Mat.t
+
+let propagate_line i0 j0 g n =
+  for j = j0+1 to size - 1 do
+    let cell = Mat.get &g (i0 , j) in
+    let cell' = IntSet.remove n cell in
+    Mat.set_mut &!g (i0, j) cell'
+  done
+
+let propagate i j g n =
+  propagate_line i j &!g n;
+  propagate_column i j &!g n;
+  propagate_square i j &!g n
+
+let rec solve i j g =
+  match is_solved &g with
+  | true -> Mat.print &g
+  | false ->
+    let (new_i, new_j) = next_pos (i,j) in
+    let try_solution n =
+      let new_g = 
+        Mat.set &g (i,j) (singleton n) in
+      propagate i j &!new_g n;
+      if is_valid &new_g when
+        solve new_i new_j new_g
+    in
+    let cell = Mat.get &g (i,j) in
+    IntSet.iter try_solution cell
+\end{lstlisting}
+  \caption{Chosen excerpt from the sudoku solver}
+  \label{ex:cow}
+\end{figure}
+
 %%% Local Variables:

### 2019-06-17 kindly/motivation.tex
Write the sudoku section.

@@
 
-While purely functional data-structure present numerous advantages in term
-of correctness and ease of use, they sometimes offer worse performances
-than their imperative counterpart. On the other hand, reasoning about imperative
-data-structure in the context of sharing is notoriously difficult.
+In the last few years, new persistent data-structures
+have been introduced which allow transient mutations, but discourage
+non-linear uses either through degraded performances
+\cite{DBLP:conf/ml/ConchonF07} or
+dynamic and static checks \cite{DBLP:journals/pacmpl/Puente17}.
+In particular, persistent Hash-Array-Mapped-Trie has been used with similar
+API in numerous non-pure functional languages (OCaml, Clojure, \dots).
+Affine types allow to formalize the performance contract between the programmer
+and the library, while borrows avoid the need to thread state explicitly,
+as usually required by an immutable API.
 %
 In this section, we show how linearity and borrowing allows us to define
-a safe API for hybrid arrays that support both immutable and mutable features,
+a safe API for persistent arrays that support both immutable and mutable features,
 and how it can be used to implement a backtracking sudoku solver.
+
+Our sudoku solver is implemented by maintaining a grid that represent the state of the game.
+In order to make backtracking easier, we want to modify the grid immutably.
+However, while some changes correspond to backtracking points,
+some do not.
+In order to improve performance, we want to implement modifications
+that do not correspond to backtracking points as mutations.
+Thanks to affine types and borrows, we can provide an API that guarantee correct
+uses of mutable and immutable modifications to the board.
+This API is represented by the signature \lstinline/HybArray/ in \cref{sig:hybarray}.
+It only differs with the \lstinline/Array/ module on two points.
+The mutation function is now called \lstinline/set_mut/, with
+a type \lstinline/&!('a t) * int * 'a -> unit/ and
+the \lstinline/set/ operation
+is now of type \lstinline/&('a t) * int * 'a -> 'a t/. It
+takes an \emph{immutable} borrow, which allows it to read the array, and
+return a fresh hybrid array.
+We showcase a very simple implementation of \lstinline/HybArray/ in
+\cref{ex:cow} where hybrid arrays are represented
+as regular arrays with copy-on-write for immutable
+modifications.
+A more elaborate implementation could use persistent arrays
+\cite{DBLP:conf/ml/ConchonF07}.
 
@@
   val set : &('a t) * int * 'a -> 'a t
-  val set_immut : &!('a t) * int * 'a -> unit
+  val set_mut : &!('a t) * int * 'a -> unit
 end
@@
     let f k =
-      if k = i then x else Array.get (a,k)
+      if k = i then x else get (a,k)
     in
@@
 
+Thanks to this API, modifications that correspond to backtracking points
+use \lstinline/set/, which allows arbitrary uses of the previous version
+of the array, while other modifications uses \lstinline/set_mut/, which
+forbids uses of the previous version.
+This can be seen in the sudoku solver implementation in \cref{ex:sudoku}.
+The board is represented as a matrix of cells (\cref{line:boardtype}).
+We assume the existence of a \lstinline/Matrix/ module which uses the same API
+as \lstinline/HybArray/ but with two indices.
+Each cell is an integer
+set that represent admissible solutions so far. The sudoku solver will
+iterate on the cells and try each possible solution (\cref{line:try_solution}).
+When a value is picked for the current cell, we use an immutable modification
+to create a backtracking point (\cref{line:immutset}) and propagate
+the changes with the \lstinline/propagate/ function. The \lstinline/propagate/
+function does not need to preserve the previous
+version of the board, and thus uses direct mutation.
+The implementation of \lstinline/propagate/ is split into three parts
+for lines, columns, and square. For concision, only \lstinline/propagate_lines/ is
+presented here (\cref{line:propline}).
+The typing ensures that all the mutations do not compromise
+backtracking, since sharing is only allowed through
+the immutable set functions, which is used for backtracking points.
+
+While we presented our API on simple copy-on-write arrays, this API
+could easily be adapted to any persistent data-structure with
+transient mutability such as RRB Vectors \cite{DBLP:journals/pacmpl/Puente17}
+and persistent HAMT (used notably as Clojure's default associative maps
+\cite{bagwell2001ideal,clojurehamt}) while providing a good performance profile
+and a convenient programming style.
+
 \begin{figure}
   \centering
-\begin{lstlisting}[multicols=2]
-type board = IntSet.t Mat.t
-
-let propagate_line i0 j0 g n =
-  for j = j0+1 to size - 1 do
-    let cell = Mat.get &g (i0 , j) in
-    let cell' = IntSet.remove n cell in
-    Mat.set_mut &!g (i0, j) cell'
-  done
-
-let propagate i j g n =
-  propagate_line i j &!g n;
-  propagate_column i j &!g n;
-  propagate_square i j &!g n
-
-let rec solve i j g =
-  match is_solved &g with
-  | true -> Mat.print &g
-  | false ->
-    let (new_i, new_j) = next_pos (i,j) in
-    let try_solution n =
-      let new_g = 
-        Mat.set &g (i,j) (singleton n) in
-      propagate i j &!new_g n;
-      if is_valid &new_g when
-        solve new_i new_j new_g
-    in
-    let cell = Mat.get &g (i,j) in
-    IntSet.iter try_solution cell
-\end{lstlisting}
+  \begin{minipage}[t]{0.45\linewidth}
+    \lstinputlisting[linerange=1-13]{code/sukodu.affe}
+  \end{minipage}\hfill
+  \begin{minipage}[t]{0.54\linewidth}
+    \lstinputlisting[linerange=15-28,firstnumber=last]{code/sukodu.affe}
+  \end{minipage}
   \caption{Chosen excerpt from the sudoku solver}
-  \label{ex:cow}
+  \label{ex:sudoku}
 \end{figure}

### 2019-06-17 kindly/motivation.tex
More examples.

@@
 
-\section{Hybrid programming styles}
+\subsection{Hybrid programming style}
 
@@
 
+\subsection{Type abstraction and linearity violation}
+
+While linearity allows to verify many properties, it also prevents
+the implementation of functions that are nevertheless safe. Rust addresses this
+issue through the \lstinline/unsafe/ construct, which allows to locally
+ignore linearity constraints. This is essential to build basic
+functions such as splitting strings.
+%
+\lang allows to partially emulate this behavior using type abstraction
+as provided by ML modules.
+In particular, data types can be internally defined as unrestricted but exposed
+as affine.
+
+\TODO{Show example with arrays}
+
+\subsection{Local data-structures with regions}
+
+\TODO{Show a queue attached to a file}
+
 %%% Local Variables:

### 2019-06-19 kindly/infer/infer.tex
Prepare.

@@
 We note $\inferW{\addlin{\Sigma}}{(C,\unif)}{\bf{\E}}{\bf{e}}{\tau}$ when
-$e$\ as type $\tau$ in $\E$ under the constraints $C$ and unifier $\unif$.
+$e$\ has type $\tau$ in $\E$ under the constraints $C$ and unifier $\unif$.
 $\Sigma$ is a map which associates free variables in $e$ to
@@
 We can then saturate the relation,
-unify all kinds that in the same equivalence classes to obtain
+unify all kinds that are in the same equivalence classes to obtain
 a most general unifier on kind variables,

### 2019-06-27 kindly/motivation.tex
Make match parameterized over its modality.
My brain is too warm.

@@
 We can then use these primitives to implement a simple addition service.
+Both client and server code are shown in
+\cref{fig:sessiontype}.
 We do not need any extra linearity annotations in the code: the fact
 that the \texttt{st} type is declared as linear is sufficient.
-In the case below, \lstinline{add_service} is inferred to be of type
-\lstinline{(int ! int ! int ? end) st -> unit}
-
+The server, \lstinline{add_service}, is inferred to be of type
+\lstinline{(int ! int ! int ? end) st -> unit}.
+The client, \lstinline/op_client/, operates by sending two messages
+and receiving the result.
+This code is polymorphic in both argument and return types and could thus
+be used to any binary operator.
+%
+\begin{figure}[!h]
+  \begin{subfigure}[t]{.5\linewidth}
 \begin{lstlisting}
@@
   close ep
-# add_service : (int ! int ! int ? end) st -> unit
+# add_service :
+  (int ! int ! int ? end) st -> unit
 \end{lstlisting}
-
-The following code then acts as a client, by sending two messages
-and receiving the result.
-This code is polymorphic in both argument and return types and could thus
-be used to any binary operator.
-%
+    \caption{Addition server}
+  \end{subfigure}~
+  \begin{subfigure}[t]{.5\linewidth}
 \begin{lstlisting}
@@
   result
-# op_client : ('a_1 ? 'a_2 ? 'b ! end) st -> 'a_1 -{lin}> 'a_2 -{lin}> 'b
+# op_client :
+  ('a_1 ? 'a_2 ? 'b ! end) st -> 'a_1 -{lin}> 'a_2 -{lin}> 'b
 \end{lstlisting}
@@
 % # ('m1 ?'m2 ? 'result ! end) st -> 'm1 -{lin}> 'm2 -{^k}> 'result
+    \caption{Binary operators client}
+  \end{subfigure}
+  \caption{Corresponding session type programs in \lang}
+  \label{fig:sessiontype}
+\end{figure}
 
-In addition, this function can be partially
+In addition, the \lstinline/op_client/ function can be partially
 applied to a channel. Since the closure returned by such a partial
@@
 %
-In this section, we show how linearity and borrowing allows us to define
+In this section, we show how \lang allows us to define
 a safe API for persistent arrays that support both immutable and mutable features,
-and how it can be used to implement a backtracking sudoku solver.
+and how it can be used to implement a simple backtracking sudoku solver.
 
@@
 the immutable set functions, which is used for backtracking points.
+In particular, the \lstinline/try_solution/ function can be used in a loop
+precisely because the \lstinline/set/ function only requires an
+\emph{immutable} borrow of \lstinline/new_g/.
+This is allowed, since immutable borrows are unrestricted.
 
@@
 
-\subsection{Type abstraction and linearity violation}
+\subsection{Type abstraction and linearity violations}
 

### 2019-06-28 kindly/motivation.tex
make the array example a bit more liberal.

@@
 module Array : sig
-  type ('a : un) t : aff
+  type ('a : 'k) t : 'k
 
@@
 
-  val get : &('a t) * int -> 'a
+  val get : ('a : un) => &('a t) * int -> 'a
   val set : &!('a t) * int * 'a -> unit
+
+  val map : (&'a -> 'b) -> &('a t) -> 'b t
 end
 \end{lstlisting}
-    \caption{Affine primitives}
+    \caption{Affine arrays}
     \label{sig:array}
@@
   for i = 2 to n - 1 do
-    let x = get &a (i-1) + get &a (i-2) in(*@\label{ex:array:get}*)
-    set &!a i x(*@\label{ex:array:set}*)
+    let x = get (&a, i-1) + get (&a, i-2) in(*@\label{ex:array:get}*)
+    set (&!a, i, x)(*@\label{ex:array:set}*)
   done;

### 2019-06-28 kindly/motivation.tex
Add the 4th "big" code example.

@@
 
-\subsection{Type abstraction and linearity violations}
+% \subsection{Type abstraction and linearity violations}
 
-While linearity allows to verify many properties, it also prevents
-the implementation of functions that are nevertheless safe. Rust addresses this
-issue through the \lstinline/unsafe/ construct, which allows to locally
-ignore linearity constraints. This is essential to build basic
-functions such as splitting strings.
-%
-\lang allows to partially emulate this behavior using type abstraction
-as provided by ML modules.
-In particular, data types can be internally defined as unrestricted but exposed
-as affine.
+% While linearity allows to verify many properties, it also prevents
+% the implementation of functions that are nevertheless safe. Rust addresses this
+% issue through the \lstinline/unsafe/ construct, which allows to locally
+% ignore linearity constraints. This is essential to build basic
+% functions such as splitting strings.
+% %
+% \lang allows to partially emulate this behavior using type abstraction
+% as provided by ML modules.
+% In particular, data types can be internally defined as unrestricted but exposed
+% as affine.
 
-\TODO{Show example with arrays}
+% \TODO{Show example with arrays}
 
@@
 
-\TODO{Show a queue attached to a file}
+We have seen that borrows are limited to a region. In the various examples so
+far, such regions didn't appear in the type signature explicitly.
+We know present an example implement of a \texttt{Pool} module
+that uses such region to enforce linearity of the objects in the pool.
+
+\TODO{finish explanation}
+
+\begin{figure}[!h]
+  \centering
+  \begin{subfigure}[t]{.45\linewidth}
+\begin{lstlisting}
+type ('a: 'k) pool : 'k = {
+  spawn : unit -> 'a ;
+  queue : 'a queue;
+}
+
+let create spawn =
+  { spawn ; queue = Queue.create () }
+let consume f c = Queue.iter f c.queue
+
+let use { spawn ; queue } f =
+  let o = match Queue.pop &queue with
+    | Some x -> x
+    | None () -> spawn ()
+  in
+  let r = f &!o in
+  Queue.push o &queue;
+  r
+\end{lstlisting}
+    \caption{Implementation}
+  \end{subfigure}~
+  \begin{subfigure}[t]{.55\linewidth}
+\begin{lstlisting}
+type ('a: 'k) pool : 'k
+
+create : (unit -> 'a) -> 'a pool
+consume : ('a -> unit) -> 'a pool -> unit
+use : ('a_1:'k_1),('a_2:'k_2),('k_1 < lin_r),('k_2 < lin_r) =>
+  &('a_1 pool) -> (&!(aff_r,'a_1) -{'k}> 'a_2) -{'k_1}> 'a_2
+\end{lstlisting}
+    \caption{Interface}
+
+\begin{lstlisting}
+let uri = "postgresql://localhost:5432"
+
+(* Create a database connection pool. *)
+let pool = Pool.create (fun () -> Db.connect uri)
+
+(* Use the pool in queries. *)
+let create_user name =
+  Pool.use &pool (fun connection ->
+    Db.insert "users" [("name", name)] connection)
+\end{lstlisting}
+    \caption{Example of use}
+  \end{subfigure}
+
+  \caption{The \texttt{Pool} module}
+\end{figure}
 

### 2019-06-28 kindly/motivation.tex
Explain the pool example a bit further.

@@
 \subsection{Local data-structures with regions}
+\label{tuto:pool}
 
 We have seen that borrows are limited to a region. In the various examples so
-far, such regions didn't appear in the type signature explicitly.
-We know present an example implement of a \texttt{Pool} module
-that uses such region to enforce linearity of the objects in the pool.
+far, such regions didn't appear in the type signatures explicitly.
+We know present an pool implementation which can store linear objects
+and such region to enforce proper uses of the objects.
+The implementation and interfaces of the \texttt{Pool} module
+are shown in \cref{impl:pool,intf:pool}. A simple example of use
+is shown in \cref{ex:pool}.
+
+A pool is parameterized by its content. Additional, the kind of the pool
+depends on the content: Linear content implies
+a linear pool, while unrestricted content implies an unrestricted pool.
+The implementation uses a concurrent queue
+as a bag of elements with atomic add and remove operations.
+The \lstinline/Pool.create/ and \lstinline/Pool.consume/ allow to build or destroy
+the pool by providing the necessary components for the elements
+of the loop.
+The main function of interest is the \lstinline/Pool.use/ function, which
+allow to use an mutable borrow of the object. The function proceeds by removing
+the element from the pool (or creating a fresh element),
+passing it down to the \lstinline/f/ function, and then adding
+it back to the pool.
+We explicitly delimit the region on \cref{line:pool:region} to ensure that
+the \lstinline/f/ function will not be able to capture \lstinline/&!o/.
 
@@
 let create spawn =
-  { spawn ; queue = Queue.create () }
-let consume f c = Queue.iter f c.queue
+  { spawn ; queue = CQueue.create () }
+let consume f c = CQueue.iter f c.queue
 
 let use { spawn ; queue } f =
-  let o = match Queue.pop &queue with
+  let o = match CQueue.pop &queue with
     | Some x -> x
@@
   in
-  let r = f &!o in
+  let r = {| f &!o |} in(*@\label{line:pool:region}*)
   Queue.push o &queue;
@@
     \caption{Implementation}
+    \label{impl:pool}
   \end{subfigure}~
@@
     \caption{Interface}
+    \label{intf:pool}
 
@@
     \caption{Example of use}
+    \label{ex:pool}
   \end{subfigure}
@@
   \caption{The \texttt{Pool} module}
+  \label{fig:pool}
 \end{figure}

### 2019-06-29 kindly/main.tex
metatheory started
More motivations
semantics and metatheory
Attempt at typesetting code-as-semrules.
Progress on inference soundness proof.
abstract

@@
 % \usepackage{textgreek}
+\usepackage{multicol}
 
@@
 \usepackage{subcaption}
+\usepackage{enumitem}
+%\setlist{leftmargin=*,noitemsep,topsep=5pt}
 
@@
 \input{caml}
+\input{lst-rule}
 \input{notations}
@@
 \begin{abstract}
-  System programming often requires to manipulate resources that
-  cannot be duplicated such as file system handles, network connections,
-  or globally allocated memory.
-  Improper handling of such resources is often the cause of bugs and security issues
-  through use-after-free errors or data races. These problems are not
-  prevented by most statically typed programming languages that only
-  guarantee type soundness and memory safety.
-  Linear and affine type systems support tracking of single-threaded resource usage
-  through types, but they are often too restrictive for real-world applications.
+  System programming often requires the manipulation of resources like
+  file system handles, network connections, or dynamically allocated
+  memory. Programmers need to follow certain protocols to handle
+  these resources correctly. Violating these protocols causes bugs
+  ranging from type mismatches over data races to use-after-free 
+  errors and memory leaks. These bugs often lead to security  vulnerabilities.
 
-  We present \lang, an extension of ML with a kind system and qualified
-  types that allows to track affine types.
-  This language supports the defining features of ML family languages:
+  Proper statically typed programming languages guarantee type soundness and memory 
+  safety by design, but most of them do not address issues arising
+  from improper resource handling. 
+  Linear and affine types guarantee single-threaded resource usage,
+  but they are rarely deployed as they are too restrictive for real-world applications.
+
+  We present \lang, an extension of ML with constrained types that
+  manages linearity and affinity properties through kinds. In addition
+  \lang{} supports the mutable and immutable borrowing of affine
+  resources, inspired by features of Rust.
+  Moreover, \lang{} retains the defining features of the ML family:
   an impure, strict functional expression language with complete principal type
   inference and type abstraction through modules.
-  Our language do not require any linearity annotation in
-  expressions and support common functional idioms.
+  Our language does not require any linearity annotations in
+  expressions and supports common functional programming idioms.
 \end{abstract}
@@
 \maketitle
+
+\input{rulemacros}
 
@@
 \input{infer/infer}
+\input{metatheory}
 \input{soa}

### 2019-07-01 kindly/motivation.tex
Improve the region explanation.

@@
 
-A pool is parameterized by its content. Additional, the kind of the pool
-depends on the content: Linear content implies
-a linear pool, while unrestricted content implies an unrestricted pool.
-The implementation uses a concurrent queue
+A pool is parameterized by its content. The kind of the pool
+depends on the content: linear content implies
+a linear pool while unrestricted content implies an unrestricted pool.
+The functions \lstinline/Pool.create/ and \lstinline/Pool.consume/ allow
+to build or destroy
+the pool by providing the corresponding functions for the elements
+of the pool.
+The main function is \lstinline/Pool.use/ which
+allow to use an object from the pool.
+\lstinline/Pool.use/ takes a shared borrow of a pool (in order to allow
+for concurrent access) and a function. This function will receive
+a mutable borrow of one arbitrary object of the pool. The typing ensure
+that this borrow is not captured nor kept by the function.
+This is done through a \emph{region} variable $r$ which annotates
+the kinds. Here, the borrow is of type \lstinline/&!(aff_r,'a_1 )/, which indicates
+a mutable borrow with an affine kind. The return type of \lstinline/f/ is bound
+to be of kind \lstinline/'k_2 < aff_r/. Since this is a strict inequality,
+types whose kind mention the region can not be returned.
+
+The implementation presented in \cref{impl:pool} uses a concurrent queue
 as a bag of elements with atomic add and remove operations.
-The \lstinline/Pool.create/ and \lstinline/Pool.consume/ allow to build or destroy
-the pool by providing the necessary components for the elements
-of the loop.
-The main function of interest is the \lstinline/Pool.use/ function, which
-allow to use an mutable borrow of the object. The function proceeds by removing
-the element from the pool (or creating a fresh element),
-passing it down to the \lstinline/f/ function, and then adding
+The implementation for the \lstinline/Pool.create/ and \lstinline/Pool.consume/
+functions are straightforward.
+The function \lstinline/Pool.use/ first draws
+an element from the pool (or creates a fresh element),
+pass it down to the \lstinline/f/ function, and then add
 it back to the pool.
 We explicitly delimit the region on \cref{line:pool:region} to ensure that
-the \lstinline/f/ function will not be able to capture \lstinline/&!o/.
-
-\TODO{finish explanation}
+the return value of \lstinline/f &!o/ do not capture \lstinline/&!o/.
 
@@
 \begin{lstlisting}
-type ('a: 'k) pool : 'k = {
+type ('a:'k) pool : 'k = {
   spawn : unit -> 'a ;
-  queue : 'a queue;
+  queue : 'a CQueue.t;
 }
@@
 \begin{lstlisting}
-type ('a: 'k) pool : 'k
+type ('a:'k) pool : 'k
 
@@
 consume : ('a -> unit) -> 'a pool -> unit
-use : ('a_1:'k_1),('a_2:'k_2),('k_1 < lin_r),('k_2 < lin_r) =>
-  &('a_1 pool) -> (&!(aff_r,'a_1) -{'k}> 'a_2) -{'k_1}> 'a_2
+use : ('a_1:'k_1),('a_2:'k_2),('k_1 < aff_r),('k_2 < aff_r) =>
+  &('a_1 pool) -> (&!(aff_r,'a_1) -{lin}> 'a_2) -{'k_1}> 'a_2
 \end{lstlisting}

### 2019-07-01 kindly/lang.tex
Add syntactic rules for automatic region annotation.
Attempt at typesetting code-as-semrules.
checkpoint
Start writing the text.

@@
 
+We now present a formalization of the \lang language featuring
+a minimal ML language with let-polymorphism, abstract types, pairs, and our novel
+linear type system with borrows which uses kinds and qualified types.
+For ease of presentation,
+our formalization makes the following simplifications.
+First, we only consider matching on pairs, instead of arbitrary algebraic
+datatypes. We consider separate operators for borrowing and reborrowing (taking
+the borrow of a borrow). We consider regions to be fully annotated
+and present a set of rules to infer their placement.
+Finally we emulates region variables
+as a series of nesting indices.
+
+In the rest of this section, we present the syntax (\cref{syntax}),
+a linearity-aware semantics (\cref{sem}), syntax directed typing rules
+(\cref{sdtyping}) and automatic rules for region annotations (\cref{regionannot}).
+
 \subsection{Syntax}
+\label{syntax}
 
@@
 \subsection{Semantics}
+\label{sem}
+
 \input{fig-reduction}
+\input{fig-code}
 
@@
 \subsection{Typing}
+\label{sdtyping}
 
@@
 
-\begin{figure*}[htp]
+\begin{figure*}[tp]
+  \input{infer/sd-environments}
+  \caption{Rewriting constraints on environments and types}
+  \label{fig:contraints-environments-types}
+\end{figure*}
+\begin{figure*}[tp]
   \input{infer/sd-typing}
@@
 \clearpage
+
+\subsection{Annotating regions}
+\label{regionannot}
+
+\begin{figure*}[htp]
+  \centering
+  \input{region-annot}
+  \caption{Automatic region annotation --- $\Rannot{e}{e'}{B}$}
+  \label{fig:region-annotation}
+\end{figure*}
+
 %%% Local Variables:

### 2019-07-01 kindly/lang.tex
Explain the region inference.

@@
 
+So far, all \lang programs have been fully annotated with regions information.
+We now show how to infer these regions annotations based on
+optionally-annotated programs.
+First, we extend the region annotation to $\region{S}{E}$ where $S$ is
+a set of variables. This annotation, defined below, is equivalent to nested
+region annotations for each individual variable.
+
+\begin{align*}
+  \region{x;S}{e} &= \region{x}{\region{S}{e}}& \region{\emptyset}{e} &= e\\
+\end{align*}
+
+\Cref{fig:region-annotation} define a rewriting relation $\RannotT{e}{e'}$
+which indicates that an optionally annotated term $e$ can be written
+in a fully annotated term $e'$.
+Through the rule \textsc{Rewrite-Top}, this is defined
+in term of an inductively defined relation
+$\Rannot{e}{e'}{S}$ where $n$ is the current nesting and $S$ is a set of
+variable that are not enclosed in a region yet.
+The base cases are constants, variables and borrows.
+The general idea is to start from the leafs of the syntax tree, create a
+region for each borrow, and enlarge the region as much as possible.
+This is implemented by a depth-first walk of the syntax
+tree which collects each variable that has a corresponding borrow.
+At each step, it rewrites the inner subterms,
+consider which borrow must be enclosed by a region now, and
+return the others for later enclosing. Binders force immediate
+enclosing of the bound variables, as demonstrated in rule \textsc{Rewrite-Lam}.
+For nodes with multiple children, we
+use a scope merge operator to decide if regions should be placed and where.
+This is shown in rule \textsc{Rewrite-Pair}.
+The merge operator, written $\getBorrows{B_l}{B_r}{(S_l,S,S_r)}$, takes
+the sets $B_l$ and $B_r$ returned by rewriting the subterms
+and returns three sets: $S_l$ and $S_r$ indicates the variables
+that should be immediately enclosed by a region on the left and right
+subterms and $S$ indicates the set of the yet-to-be-enclosed variables.
+As an example, the rule \textsc{AnnotRegion-MutLeft} is applied
+when there is an immutable borrow and a mutable borrow. In that case, a
+region is created to enclose the immutable borrow, while the mutable
+borrow is left to be closed later. This is coherent with the rules
+for environment splitting and suspended bindings from \cref{sdtyping}.
+%
+Explicitly annotated regions are handled specially through
+rule \textsc{Rewrite-Region}. In that case, we assume that all inner
+borrows should be enclosed immediately.
+
 \begin{figure*}[htp]
@@
   \input{region-annot}
-  \caption{Automatic region annotation --- $\Rannot{e}{e'}{B}$}
+  \caption{Automatic region annotation --- $\RannotT{e}{e'}$}
   \label{fig:region-annotation}

### 2019-07-02 kindly/lang.tex
Adjusts the kinding rules a little bit.

@@
 
+We note $\inferSK{C}{\E}{\tau}{(\Multi k)k}$
+when $\tau$ has kind $(\Multi k)k$ in environment $\E$ under constraints $C$.
+The rules are shown in \cref{rules:sd-kinding}.
+Kinds and types follow a small calculus with variables ($\tvar$,\dots),
+functions (type constructors $\T{t}$), application ($\tapp{t}{\Multi{\tau}}$)
+and primitives such as arrow ($\tau\tarr{k}\tau'$) and
+borrow ($\borrowty{k}{\tau}$).
+Kind inference is thus fairly straightforward and is done simply by following
+the syntax of the types. The only kind arrows can appear when looking
+up the kind scheme of a type constructor $\T t$. Kind arrows are forbidden
+in any other contexts.
+
 \begin{figure}[ht]
@@
   \input{infer/sd-kinds}
-  \caption{Syntax-directed kind inference --
-    $\inferSK{C}{\E}{\tau}{(\Multi k)k}$}
+  \caption{Syntax-directed kinding rule --
+    $\inferSK{C}{\E}{\tau}{k}$}
   \label{rules:sd-kinding}

### 2019-07-02 kindly/lang.tex
Misc fix.
Improve a bit the syntax presentation.

@@
 
+Let us first define our conventions and meta-syntactic variables.
+Expressions, types and kinds are noted respectively $e$, $\tau$, and $k$.
+Variables, type variables and kind variables are noted
+respectively $x$, $\tvar$ and $\kvar$. Types and kind schemes
+are noted $\schm$ and $\kschm$.
+In the rest of the text, we will write lists with a bar and an index, for
+instance $\Multi[i]{\tau}$.
+Kind constants are noted in bold capital letters, such as $\kaff$ for affine.
+
 The syntax of \lang is presented in \cref{grammar} and follows
 traditional ML languages with let bindings.
-Data constructors are introduced by $\introK{K}{e}$ and eliminated
-with by $\elimK{K}{e}$.
-Types are composed of type variable $\tvar$, parameterized type constructors
-$\T{t}$ and arrow types annotated with a kind, noted $\tau \tarr{k} \tau$.
-Kinds can be either a kind variable $\kvar$, affine $\kaff$ or unrestricted $\kun$.
+Borrows, noted $\borrow{e}$, are annotated with a specification indicating
+if a borrow is immutable ($\IBORROW$) or mutable ($\MBORROW$).
+We also distinguish reborrows, which are noted $\reborrow{e}$.
+Match are indexed with a specification which can be either $\&^b$
+(which $b$ a borrow specification) or $id$, indicating if the match
+should operate on borrows or not.
+Regions, noted $\region{x}{e}$ are annotated with their nesting ($n$)
+and the variable whose borrow they enclose ($x$).
+%
+$\tapp{t}{\Multi{\tau}}$ denotes a type application where
+$\T t$ is an abstract type constructor.
+Arrow and borrow types are annotated with a kind, noted $\tau \tarr{k} \tau$
+and $\borrowty{k}{\tau}$ respectively.
+%
+Kinds can be either a kind variable $\kvar$, or constants
+(linear $\klin$, affine $\kaff$ or unrestricted $\kun$) indexed
+by a nesting level represented by a member of $N \cup \{\infty\}$.
 Type schemes and kind schemes are guarded by a set of constraints.
+%
+Finally, constraints are composed of a list of kind inequalities.
+
+\TODO{Do we explain environments here, or push that for typing, later on ?}
 
@@
 
-We note $\inferSK{C}{\E}{\tau}{(\Multi k)k}$
-when $\tau$ has kind $(\Multi k)k$ in environment $\E$ under constraints $C$.
+We note $\inferSK{C}{\E}{\tau}{k}$
+when $\tau$ has kind $k$ in environment $\E$ under constraints $C$.
 The rules are shown in \cref{rules:sd-kinding}.

### 2019-07-03 kindly/lang.tex
Misc fixes for the kind/type rules.

@@
 functions (type constructors $\T{t}$), application ($\tapp{t}{\Multi{\tau}}$)
-and primitives such as arrow ($\tau\tarr{k}\tau'$) and
-borrow ($\borrowty{k}{\tau}$).
-Kind inference is thus fairly straightforward and is done simply by following
-the syntax of the types. The only kind arrows can appear when looking
+and primitives such as types for arrows ($\tau\tarr{k}\tau'$) and
+borrows ($\borrowty{k}{\tau}$).
+Kind checking can thus be done in a fairly straightforward, syntax-directed
+fashion by simply following
+the syntax of the types. Kind arrows can only appear when looking
 up the kind scheme of a type constructor $\T t$. Kind arrows are forbidden

### 2019-07-03 kindly/intro.tex
initial blurb

\section{Introduction}

A large proportion of systems programming is focused on the proper
handling of resources, like file handles, network connections, or
dynamically allocated memory. Each of these resources comes with a
protocol that prescribes the correct use of its API.
For examples, a file handle appears as the result of of opening a
file. If it was opened for reading, then read operations will suceed,
but write operations will fail. Once the handle is closed, it cannot
be used for reading or writing, anymore.

Dynamic allocation of memory is analogous. An API call returns a
pointer to a memory area, which can then be read and written to until
the memory is released by another API call.

In both cases, a resource is created in a certain state and a resource
handle is returned to the program. Depending on this state, certain API calls
can safely be applied to it. Finally, there is another API call to
release the resource, which yields the handle invalid.
Taken to the extreme, each API call changes the state so that a
different set of API calls is enabled afterwards.

Ignoring this protocol is a common source of errors \cite{is there
  some paper?}.

Most statically typed languages do not address this issue.

But linear types do.

But typestate does it. A popular example is session types.

Typestate needs to be supported by affinity (or linearity, or
uniqueness).

But these approaches are restrictive in pratice because [here is an
example that doesn't work well with existing systems].

%%% Local Variables:
%%% mode: latex
%%% TeX-master: "main"
%%% End:

### 2019-07-03 kindly/main.tex
wordsmithing the abstract

@@
   System programming often requires the manipulation of resources like
-  file system handles, network connections, or dynamically allocated
+  file handles, network connections, or dynamically allocated
   memory. Programmers need to follow certain protocols to handle
@@
 
-  Proper statically typed programming languages guarantee type soundness and memory 
-  safety by design, but most of them do not address issues arising
+  While statically typed programming languages guarantee type soundness and memory 
+  safety by design, most of them do not address issues arising
   from improper resource handling. 

### 2019-07-03 kindly/motivation.tex
motivation

@@
 
-A classical example of linearity at work are to control the uses of resources such
-as files or network handles.
+Session typing \cite{Honda1993,DBLP:conf/esop/HondaVK98} is a type
+discipline for checking protocols statically. A session type is
+ascribed to a communication channel and describes
+a sequence of interactions. For instance, the type
+\lstinline{!int.!int.?int.end} specifies a protocol
+where the program must send two integers, receive an integer, and then
+close the channel. 
 %
-Session types, notably, allow to statically type check protocols.
-For instance, the type \lstinline{!int.!int.?int.end} represent a session
-where the program must send two integers, receive an integer, then end the session.
-%
-In this context, it is essential that channels
-are used linearly, since every use of the channel
-will make the protocol progress and change the type of the channel.
+In this context, channels must be used linearly, because every use of
+a channel ``consumes'' one interaction and hence changes the type of the
+channel. That is, after sending two integers on the channel,
+the remaining channel has type \lstinline{?int.end}.
 
-Session types are usually defined with the following operators:
-
+Here are some standard type operators for session types \SType:
 \begin{center}
-  \begin{tabular}{r l}
-    $\tau ! \mathcal S$ & Send a value of type $\tau$ then continue with protocol $S$.\\
-    $\tau ? \mathcal S$& Receive a value of type $\tau$ then continue with protocol $S$.\\
-    $\mathcal S \oplus \mathcal S'$& Choose between protocols $S$ and $S'$.\\
-    $\mathcal S \operatorname{\&} \mathcal S'$& Offer a choice between protocols $S$ and $S'$.
+  $\SType ::=$
+  \begin{tabular}[t]{rl}
+    $\tau ! \SType$ & Send a value of type $\tau$ then continue with protocol $\SType$.\\
+    $\tau ? \SType$& Receive a value of type $\tau$ then continue with protocol $\SType$.\\
+    $\SType \oplus \SType'$& Internal choice between protocols $\SType$ and $\SType'$.\\
+    $\SType \operatorname{\&} \SType'$
+                    & Offer a choice between protocols $\SType$ and $\SType'$. 
   \end{tabular}
@@
 
-It is well known that binary session types can be embedded into ML languages
-\cite{DBLP:journals/jfp/Padovani17}. We can easily
-embed session types in our language and provide the following API:
+\citet{DBLP:journals/jfp/Padovani17}  has shown how to encode this
+style of session typing in ML-like languages, but his implementation
+downgrades linearity to a run-time check for affinity. Building on that
+encoding we can provide a safe API in \lang that statically enforces
+linear handling of channels:
 %
 \begin{lstlisting}
-type 'S st : lin
-val receive: ('a ? 'S) st -> 'a * 'S st
+type 'S st : lin (*@\label{line:session1}*)
+val receive: ('a ? 'S) st -> 'a * 'S st (*@\label{line:session2}*)
 val send : ('a:'k). 'a -> ('a ! 'S) st -{'k}> 'S st
+val send : ('a ! 'S) st -> 'a -{lin}> 'S st
 val create : unit -> 'S st * (dual 'S) st
@@
 \end{lstlisting}
+
+Line~\ref{line:session1} introduces a parameterized abstract
+type \lstinline{st} which is linear as indicated
+by its kind \lstinline{lin}. Its low-level implementation would wrap a
+handle for a socket, for example. The \lstinline{receive}  operation
+in Line~\ref{line:session2} takes a channel that is ready to receive a
+value of type \lstinline{'a} and returns a pair of the value a the
+channel at its residual type \lstinline{'S}. It does not matter
+whether \lstinline{'a} is restricted to be linear, in fact
+\lstinline{receive} is polymorphic in the kind of \lstinline{'a} and
+also in the kind of \lstinline{->}. This kind polymorphism is the
+default if no constraints are specified.
+
+The \lstinline{send} operation, however,
 

### 2019-07-03 kindly/infer/infer.tex
Fix weakening.
Progress on inference soundness proof.

@@
 presents how to implement Hindley-Milner type inference in a language
-with qualified types. We first present various preliminaries definitions,
+with qualified types and an arbitrary custom constraint language.
+We first present various preliminaries definitions,
 we then show our extension of HM(X) that support a kind system and affine
@@
   % \and
-  \inferrule{}
-  { \Cleq{\Sv}{k}
-    = \bigwedge_{\Sone{x}{\schm}\in\Sv} \Cleq{\schm}{k}
-  }
-  \and
+  % \inferrule{}
+  % { \Cleq{\Sv}{k}
+  %   = \bigwedge_{\Sone{x}{\schm}\in\Sv} \Cleq{\schm}{k}
+  % }
+  % \and
   % \inferrule
@@
   % { \text{if } \operatorname{lenght}(\Sv(x))  = \Cempty{} \wedge  }
-  { \Weaken_{(x,k)}(\Sv) =
+  { \Weaken_{\bvar{x}{\sigma}}(\Sv) =
     \begin{dcases}
-      \Ctrue& \operatorname{length}(\Sv(x)) = 1\\
-      {k}\lk\kaff_\infty& \text{ otherwise}\\
+      \Ctrue& \text{if } x\in\Sv\\
+      \Cleq{\sigma}{\kaff_\infty}&\\
     \end{dcases}
@@
   }
+
+  \inferrule{
+    \inferK{C \Cand C_x} \E \tau {k'} \\
+    D = C \Cand C_x \Cand \Cleq{k'}k
+  }{
+    \E \vdash \Cleq{\forall \kvar_i \forall (\tvar_j:k_j).\ \qual{C_x}{\tau}}{k} \Crewrite  D
+  }
 \end{mathpar}
-\caption{Operation on usage maps $\Sv$}
+\caption{Auxiliary rules for inference}
 \label{op:usgmap}
@@
 \cref{rules:kinding}.
-Since higher-kinded types are not supported, the type application
-rule and the type constructor rule are merged in {\sc KApp}.
-Additionally, type variables must be of a simple kind in rule {\sc KVar}.
-Note that in the case of a type constructor with no argument, the {\sc KApp}
-rule degenerates to a simpler form which is similar to the {\sc KVar} rule.
-Kind schemes are instantiated in the {\sc KApp} and {\sc KVar} rules by creating
+Similar to the syntax-directed rules, higher-kinds are not generally supported
+and can only appear by looking-up the kind scheme of a type constructor,
+for use in the type application rule {\sc KApp}.
+Type variables must be of a simple kind in rule {\sc KVar}.
+Kind schemes are instantiated in the {\sc KVar} rules by creating
 fresh kind variables and the associated substitution.
+{\sc KArr} and {\sc KBorrow} simply returns the kind of the primitive
+arrow and borrow types.
 

### 2019-07-03 kindly/motivation.tex
session types

@@
 val receive: ('a ? 'S) st -> 'a * 'S st (*@\label{line:session2}*)
-val send : ('a:'k). 'a -> ('a ! 'S) st -{'k}> 'S st
 val send : ('a ! 'S) st -> 'a -{lin}> 'S st
@@
 \end{lstlisting}
+% val send : ('a:'k). 'a -> ('a ! 'S) st -{'k}> 'S st
 
@@
 default if no constraints are specified.
+%
+The \lstinline{send} operation takes a linear channel and returns a
+single-use function that takes a value of type \lstinline{'a} suitable
+for sending and returns the channel with updated type.
+%
+The \lstinline{create} operation returns a pair of channel
+endpoints. They follow dual communication protocols, where the
+\lstinline{dual} operator swaps sending and receiving operations.
+%
+Finally, \lstinline{close} closes the channel.
 
-The \lstinline{send} operation, however,
+In \cref{fig:sessiontype} we show how to use these primitives to
+implement client and server for an addition service.
+No linearity annotations are needed in the code, as all linearity
+properties can be inferred from the linearity of the \texttt{st} type.
 
-\TODO{Explain}
-
-We can then use these primitives to implement a simple addition service.
-Both client and server code are shown in
-\cref{fig:sessiontype}.
-We do not need any extra linearity annotations in the code: the fact
-that the \texttt{st} type is declared as linear is sufficient.
-The server, \lstinline{add_service}, is inferred to be of type
+The inferred type of the server, \lstinline{add_service}, is
 \lstinline{(int ! int ! int ? end) st -> unit}.
-The client, \lstinline/op_client/, operates by sending two messages
+The client operates by sending two messages
 and receiving the result.
-This code is polymorphic in both argument and return types and could thus
-be used to any binary operator.
+This code is polymorphic in both argument and return types, so it could
+be used with any binary operator.
 %
@@
   let y, ep = receive ep in
-  let ep = send (x + y) ep in
+  let ep = send ep (x + y) in
   close ep
@@
 let op_client ep x y =
-  let ep = send x ep in
-  let ep = send y ep in
+  let ep = send ep x in
+  let ep = send ep y in
   let result, ep = receive ep in
@@
 \end{figure}
+%
+Moreover, the \lstinline/op_client/ function can be partially applied
+to a channel. Since the closure returned by such a partial application
+captures the channel, it can only be used once.  This restriction is
+reflected by the arrow of kind \lstinline{lin}, \lstinline/-{lin}>/,
+which is the type of a single-use function. 
+The general form of
+arrows types in \lang is \lstinline/-{k}>/, where \lstinline/kk/ is a
+kind that restricts the number of uses of the function.  For
+convenience, we shorten \lstinline/-{un}>/ to \lstinline/->/.  \lang
+infers the
+single-use property of the arrows without any user
+annotation. In fact, the only difference between the
+code presented here and Padovani's examples
+\cite{DBLP:journals/jfp/Padovani17} is the kind annotation on the
+type definition of \lstinline/st/.
 
-In addition, the \lstinline/op_client/ function can be partially
-applied to a channel. Since the closure returned by such a partial
-application captures the channel, it can only be used once.
-This is reflected by the arrow, \lstinline/-{lin}>/, which is the
-type of functions that can only be used once.
-The general form of arrows types in \lang is \lstinline/-{k}>/, where
-\lstinline/kk/ is an arbitrary kind that describes the behavior of
-underlying closure.
-For convenience, we shorten \lstinline/-{un}>/ as \lstinline/->/.
-Note again that the linearity on the arrows was inferred without
-any annotation from the user. In fact, the only difference between the code
-presented so far and the examples by
-\citet{DBLP:journals/jfp/Padovani17}
-is the kind annotation on the type definition of \lstinline/st/.
-
-\TODO{more details?}
-
-Finally, we can compose both server and client.
-Failure to fully consume either \lstinline/a/ or \lstinline/b/
+To run client and server, we can create a channel and apply
+\lstinline{add_service} to one end and \lstinline{op_client} to the other.
+Failure to consume either channel endpoints (\lstinline/a/ or \lstinline/b/)
 would result in a type error.
@@
   let (a, b) = create () in
-  add_service a;
+  fork add_service a;
   op_client b 1 2

### 2019-07-03 kindly/motivation.tex
more motivation

@@
 \subsection{A session on linearity}
+\label{sec:session-linearity}
 
@@
 \begin{lstlisting}
-let main1 () =
+let main () =
   let (a, b) = create () in
@@
 \subsection{Imperative programming}
+\label{sec:imper-progr}
 
-The programming style used for channels so far is
-appropriate for purely functional code.
-This style is however fairly bothersome in mutable contexts.
+The programming style demonstrated for channels is
+appropriate for purely functional code, but becomes tedious when
+manipulating mutable resources.
 %
 To make linearity easier to use for imperative programming, we rely on
-{\em borrows}. We present a typical API for mutable affine arrays
-in \cref{sig:array} and an example of use in \cref{ex:array}.
-We can distinguish three type of operations. Firstly,
-function that manipulate the whole
-array such as \lstinline/create/ and \lstinline/free/, and behaves exactly
-like the functions on channel presented in the previous section.
-Second, read-only functions such as \lstinline/get/
-manipulates \emph{immutable} borrows. An immutable borrow over array \lstinline/a/
-is introduced with \lstinline/&a/ and represent a read-only view of the array.
-Finally, mutable functions, such as \lstinline/set/ manipulate
-a \emph{mutable borrow}. Mutable borrows, introduced with \lstinline/&!a/
-represent a read-write view of the array.
+the notion of borrowing \cite{DBLP:conf/popl/BoylandR05}. As an example,
+\cref{sig:array} contains an API for mutable affine arrays with a use case in \cref{ex:array}.
 
-Multiple immutable borrows can be used concurrently and are considered
-unrestricted (\lstinline/un/).
-A mutable borrow is exclusive, and considered affine (\lstinline/aff/).
-Furthermore, any borrow prevents access to the ``raw'' access
-as long as they are in use. In \cref{ex:array}, on \cref{ex:array:get},
+We distinguish several kinds of operation: constructors, destructore,
+observers, and mutators. 
+Constructors and destructors like \lstinline/create/ and \lstinline/free/ manipulate the whole
+array. As with channels in \cref{sec:session-linearity},
+\lstinline/create/ yields a linear resource which is consumed by \lstinline/free/.
+During the lifetime of the array resource \lstinline/a/, we can split
+off \emph{immutable borrows} \lstinline/&a/ that provide a read-only
+view or \emph{mutable borrows} \lstinline/&!a/ for read-write views.
+Observer functions such as \lstinline/get/ expect an immutable borrow
+argument, mutator functions such a \lstinline/set/ expect
+a mutable borrow.
+
+Generally, each borrow is tied to a region whose lifetime is properly
+contained in the lifetime of the resource. 
+In a region, we can split off as many immutable borrows as we like
+from a resource, but we can have only one mutable borrow. In a
+subsidiary region, we can take further immutable borrows from any
+borrow or we can take a mutable borrow from a mutable borrow from a
+enclosing region. Borrows are confined to their regions. Inside,
+immutable borrows are unrestricted (\lstinline/un/) whereas mutable
+borrows are affine (\lstinline/aff/).
+
+Furthermore, any borrow prevents access to the ``raw''  resource
+as long as it is in use. In \cref{ex:array}, \cref{ex:array:get},
 two immutable borrows are used in the same expression. Once this operation
-is done, the borrows are not used anymore, which allows us to take
-an exclusive immutable borrow to mutate the array on \cref{ex:array:set}.
-Finally we can return the whole array on \cref{ex:array:return} since
-none of the borrows are used anymore.
+is done, the borrows are not used anymore, which allows us to use
+an exclusive mutable borrow in \cref{ex:array:set}.
+Finally, we return the whole array in \cref{ex:array:return} because
+none of the borrows are in scope anymore.
 
@@
 
+%%% TBC
 Naturally, \lang also checks that borrows do not escape their point
@@
     \caption{For loop function for affine arrays}
-    \label{ex:array}
+    \label{ex:array:for}
   \end{subfigure}

### 2019-07-04 kindly/motivation.tex
explanation array api

@@
 
+The intention of this API is to enable arrays to contain affine
+elements. One would first use \lstinline/create n v/ to create
+an (unrestricted) array of size \lstinline/n/ with some value
+\lstinline/v/. This value must be unrestricted as it is duplicated to
+initialize all array elements. Using the \lstinline/map/ function we
+can transform the unrestricted elements into linear affine)
+ones. This transformation yields a linear (affine) array that can be
+processed further using \lstinline/map/ (or \lstinline/set/ if
+affine). Before \lstinline/free/ing the array the elements must be
+affine, again.
+
 We distinguish several kinds of operation: constructors, destructore,
@@
 view or \emph{mutable borrows} \lstinline/&!a/ for read-write views.
-Observer functions such as \lstinline/get/ expect an immutable borrow
+Observer functions such as \lstinline/length/ and \lstinline/get/ expect an immutable borrow
 argument, mutator functions such a \lstinline/set/ expect
@@
   val create : ('a : un) => int -> 'a -> 'a t
-  val free : 'a t -> unit
+  val free : ('a : aff) => 'a t -> unit
 
@@
   val get : ('a : un) => &('a t) * int -> 'a
-  val set : &!('a t) * int * 'a -> unit
+  val set : ('a : aff) => &!('a t) * int * 'a -> unit
 

### 2019-07-04 kindly/motivation.tex
motivation: hybrid arrays

@@
 
-The programming style demonstrated for channels is
-appropriate for purely functional code, but becomes tedious when
-manipulating mutable resources.
+The functional resource-passing style demonstrated for channels in
+\cref{sec:session-linearity} becomes tedious when manipulating mutable
+resources like buffers or connection pools.
 %
-To make linearity easier to use for imperative programming, we rely on
-the notion of borrowing \cite{DBLP:conf/popl/BoylandR05}. As an example,
-\cref{sig:array} contains an API for mutable affine arrays with a use case in \cref{ex:array}.
+To make linearity easier to use with an imperative programming style, we rely on
+the notion of borrowing \cite{DBLP:conf/popl/BoylandR05}.
 
-The intention of this API is to enable arrays to contain affine
-elements. One would first use \lstinline/create n v/ to create
+As a first example, we consider an API for mutable linear arrays
+(\cref{sig:array}).
+The intention of this API is to safely handle manual allocation and
+deallocation of arrays that may contain affine elements.
+One would first use \lstinline/create n v/ to create
 an (unrestricted) array of size \lstinline/n/ with some value
@@
 initialize all array elements. Using the \lstinline/map/ function we
-can transform the unrestricted elements into linear affine)
-ones. This transformation yields a linear (affine) array that can be
-processed further using \lstinline/map/ (or \lstinline/set/ if
-affine). Before \lstinline/free/ing the array the elements must be
+can transform the unrestricted elements into linear (affine)
+ones. This array can be processed further using \lstinline/map/ (or \lstinline/set/ if
+affine). The \lstinline/get/ function only applies if the element type
+is unrestricted as one element is duplicated. The \lstinline/length/
+function applies anytime. Before \lstinline/free/ing the array the elements must be
 affine, again.
 
-We distinguish several kinds of operation: constructors, destructore,
+To manage the different kinds of accessing the array we distinguish between constructors, destructors, 
 observers, and mutators. 
@@
 
-Generally, each borrow is tied to a region whose lifetime is properly
+Each borrow is tied to a region whose lifetime is properly
 contained in the lifetime of the resource. 
 In a region, we can split off as many immutable borrows as we like
-from a resource, but we can have only one mutable borrow. In a
+from a resource, but we can take only one mutable borrow. In a
 subsidiary region, we can take further immutable borrows from any
 borrow or we can take a mutable borrow from a mutable borrow from a
-enclosing region. Borrows are confined to their regions. Inside,
+enclosing region. Borrows are confined to their regions. Inside the region,
 immutable borrows are unrestricted (\lstinline/un/) whereas mutable
@@
 
-Furthermore, any borrow prevents access to the ``raw''  resource
-as long as it is in use. In \cref{ex:array}, \cref{ex:array:get},
-two immutable borrows are used in the same expression. Once this operation
-is done, the borrows are not used anymore, which allows us to use
-an exclusive mutable borrow in \cref{ex:array:set}.
-Finally, we return the whole array in \cref{ex:array:return} because
-none of the borrows are in scope anymore.
+The code in \cref{ex:array} demonstrates a use of the API to create an
+array of Fibonacci numbers. After creation of the array, we can see
+that the presence of a borrow prevents access to the ``raw'' 
+resource inside the borrow's region. In \cref{ex:array:get} of the
+example, there are two immutable borrows in the same
+expression, which forms a region by itself. These borrows are split
+off the mutable borrow used in \cref{ex:array:set}, which belongs to
+the next enclosing region.
+Finally, the code returns the whole array in \cref{ex:array:return},
+which is possible because none of the borrows are in scope anymore.
 
-\begin{figure}
+\begin{figure}[tp]
   \centering
@@
 module Array : sig
-  type ('a : 'k) t : 'k
+  type ('a : 'k) t : lin
 
@@
 
-%%% TBC
-Naturally, \lang also checks that borrows do not escape their point
-of use through {\em syntactic regions}. The example
-above is reproduce in \cref{ex:array:region} with
-regions explicitly delimited by \lstinline/{| ... |}/.
-\lang ensures that borrows never escape the smallest enclosing
-region.
-For instance, On \cref{ex:array:region:get}, \lstinline/&a/ does
-not escape the region since the borrow is not captured, and the
-region only return integers.
-Regions annotations are automatically inserted by \lang using
-syntactic information such as borrows and binders.
-Users can also annotate regions manually to further restrict the range
-of a borrow. In practice, we have found that regions
-annotations are very rarely needed in common imperative usage.
+\cref{ex:array:region} makes explicit the regions from the example
+indicated by braces as in \lstinline/{| ... |}/.  \lang guarantees that borrows
+never escape the smallest enclosing region.  For instance, in
+\cref{ex:array:region:get}, \lstinline/&a/ does not escape because it
+is consumed in the calls to \lstinline/get/.
+Regions annotations are automatically inserted using
+syntactic information such as borrows and binders (see
+\cref{regionannot} for more information).  Users can also
+annotate regions manually to further restrict the range of a
+borrow. In practice, we have found that region annotations are
+rarely needed with common programming idioms. All regions for the
+examples in this section have been inferred.
 
@@
   loop f &!a 0
-# Borrowed m ('a Array.t) tt =>
+# for : Borrowed m ('a Array.t) tt =>
   (tt -> int -> unit) -> tt -> unit
@@
 
-In the last few years, new persistent data-structures
-have been introduced which allow transient mutations, but discourage
-non-linear uses either through degraded performances
+In the last few years, new persistent data structures
+have been introduced which permit transient mutations, but discourage
+non-linear uses either through degraded performance
 \cite{DBLP:conf/ml/ConchonF07} or
 dynamic and static checks \cite{DBLP:journals/pacmpl/Puente17}.
-In particular, persistent Hash-Array-Mapped-Trie has been used with similar
-API in numerous non-pure functional languages (OCaml, Clojure, \dots).
-Affine types allow to formalize the performance contract between the programmer
+In particular, persistent Hash-Array-Mapped-Tries have been used with similar
+APIs in several non-pure functional languages (OCaml, Clojure, \dots).
+Affine types allow us to formalize the performance contract between the programmer
 and the library, while borrows avoid the need to thread state explicitly,
-as usually required by an immutable API.
+as usually required by an API that enforces immutability.
 %
@@
 a safe API for persistent arrays that support both immutable and mutable features,
-and how it can be used to implement a simple backtracking sudoku solver.
+and how it can be used to implement a simple backtracking Sudoku solver.
 
-Our sudoku solver is implemented by maintaining a grid that represent the state of the game.
-In order to make backtracking easier, we want to modify the grid immutably.
-However, while some changes correspond to backtracking points,
-some do not.
-In order to improve performance, we want to implement modifications
-that do not correspond to backtracking points as mutations.
-Thanks to affine types and borrows, we can provide an API that guarantee correct
-uses of mutable and immutable modifications to the board.
-This API is represented by the signature \lstinline/HybArray/ in \cref{sig:hybarray}.
-It only differs with the \lstinline/Array/ module on two points.
-The mutation function is now called \lstinline/set_mut/, with
-a type \lstinline/&!('a t) * int * 'a -> unit/ and
-the \lstinline/set/ operation
-is now of type \lstinline/&('a t) * int * 'a -> 'a t/. It
-takes an \emph{immutable} borrow, which allows it to read the array, and
-return a fresh hybrid array.
-We showcase a very simple implementation of \lstinline/HybArray/ in
-\cref{ex:cow} where hybrid arrays are represented
-as regular arrays with copy-on-write for immutable
-modifications.
-A more elaborate implementation could use persistent arrays
-\cite{DBLP:conf/ml/ConchonF07}.
+The Sudoku solver maintains a 2D-array that represents the state of
+the game and uses backtracking when there are several alternatives to proceed.
+The use of backtracking suggests a persistent data structure for the 2D-array.
+However, only changes that correspond to a choice point need to use
+the persistence mechanism, others may be implemented as
+cheap in-place mutations.
 
-\begin{figure}
+\cref{sig:hybarray} contains an API \lstinline/HybArray/ 
+that guarantees correct use of mutable and immutable modifications to
+the board by relying on affine types and borrows.
+The signature differs slightly from the \lstinline/Array/
+signature. For simplicity, the array elements are unrestricted, 
+hence it is sufficent for the array to be affine.
+The in-place mutation function is called \lstinline/set_mut/ with type
+\lstinline/&!('a t) * int * 'a -> unit/ and the persistent
+\lstinline/set/ operation has type \lstinline/&('a t) * int * 'a -> 'a t/. It
+takes an \emph{immutable borrow} because it only reads from the
+argument array, but returns a fresh,  modified structure.
+The code in \cref{ex:cow} is a very simple implementation of
+\lstinline/HybArray/ that represents hybrid arrays
+as regular arrays with copy-on-write for persistent
+modifications. The function
+\lstinline/create_from : int -> (int -> 'a) -> 'a Array.t /
+initializes an array from a function.
+% A more elaborate implementation could use persistent arrays
+% \cite{DBLP:conf/ml/ConchonF07}.
+
+\begin{figure}[tp]
   \centering
@@
 
-  val create : ('a : un) => int -> 'a -> 'a t
+  val create : int -> 'a -> 'a t
   val free : 'a t -> unit
@@
     in
-    create (length a) f
+    create_from (length a) f
 
@@
 forbids uses of the previous version.
-This can be seen in the sudoku solver implementation in \cref{ex:sudoku}.
+This can be seen in the Sudoku solver implementation in \cref{ex:sudoku}.
 The board is represented as a matrix of cells (\cref{line:boardtype}).
@@
 Each cell is an integer
-set that represent admissible solutions so far. The sudoku solver will
+set that represent admissible solutions so far. The Sudoku solver will
 iterate on the cells and try each possible solution (\cref{line:try_solution}).
@@
   \end{minipage}
-  \caption{Chosen excerpt from the sudoku solver}
+  \caption{Excerpt of the Sudoku solver}
   \label{ex:sudoku}

### 2019-07-04 kindly/motivation.tex
sudoku

@@
 
-The Sudoku solver maintains a 2D-array that represents the state of
-the game and uses backtracking when there are several alternatives to proceed.
-The use of backtracking suggests a persistent data structure for the 2D-array.
+The Sudoku solver maintains an array to represent the state of
+the game and uses backtracking when there are several choices to proceed.
+The use of backtracking suggests a persistent data structure for the array.
 However, only changes that correspond to a choice point need to use
@@
 \cref{sig:hybarray} contains an API \lstinline/HybArray/ 
-that guarantees correct use of mutable and immutable modifications to
-the board by relying on affine types and borrows.
+that enables using mutable and immutable modifications to
+the board through affine types and borrows.
 The signature differs slightly from the \lstinline/Array/
@@
 hence it is sufficent for the array to be affine.
-The in-place mutation function is called \lstinline/set_mut/ with type
-\lstinline/&!('a t) * int * 'a -> unit/ and the persistent
+The in-place mutation function \lstinline/set_mut/ with type
+\lstinline/&!('a t) * int * 'a -> unit/ works on a mutable borrow  whereas the persistent
 \lstinline/set/ operation has type \lstinline/&('a t) * int * 'a -> 'a t/. It
-takes an \emph{immutable borrow} because it only reads from the
+takes an immutable borrow because it only reads from the
 argument array, but returns a fresh,  modified structure.
-The code in \cref{ex:cow} is a very simple implementation of
+The code in \cref{ex:cow} contains a very simple implementation of
 \lstinline/HybArray/ that represents hybrid arrays
-as regular arrays with copy-on-write for persistent
+as regular arrays and uses copy-on-write for persistent
 modifications. The function
@@
 
-Thanks to this API, modifications that correspond to backtracking points
-use \lstinline/set/, which allows arbitrary uses of the previous version
-of the array, while other modifications uses \lstinline/set_mut/, which
+Our implementation of a Sudoku solver (\cref{ex:sudoku}) performs
+modifications that correspond to choice points using \lstinline/set/,
+which makes it trivial to come back to the previous version
+of the array, while other modifications use \lstinline/set_mut/, which
 forbids uses of the previous version.
-This can be seen in the Sudoku solver implementation in \cref{ex:sudoku}.
-The board is represented as a matrix of cells (\cref{line:boardtype}).
-We assume the existence of a \lstinline/Matrix/ module which uses the same API
-as \lstinline/HybArray/ but with two indices.
-Each cell is an integer
-set that represent admissible solutions so far. The Sudoku solver will
-iterate on the cells and try each possible solution (\cref{line:try_solution}).
-When a value is picked for the current cell, we use an immutable modification
-to create a backtracking point (\cref{line:immutset}) and propagate
-the changes with the \lstinline/propagate/ function. The \lstinline/propagate/
-function does not need to preserve the previous
-version of the board, and thus uses direct mutation.
+
+The board is represented as a 2D-matrix (\cref{line:boardtype}), where
+the \lstinline/Matrix/ type uses the same API as \lstinline/HybArray/
+but with two indices. 
+Each cell contains an integer set that represent admissible solutions so far.
+The Sudoku solver iterates over the cells and tries each possible solution (\cref{line:try_solution}).
+When a value is picked for the current cell, we create a choicepoint,
+change the cell with an immutable modification (\cref{line:immutset}), and propagate
+the changes with the \lstinline/propagate/ function.
+The \lstinline/propagate/ function uses direct mutation through a
+mutable borrow of the matrix as it need not preserve the previous
+version of the board.  
 The implementation of \lstinline/propagate/ is split into three parts
-for lines, columns, and square. For concision, only \lstinline/propagate_lines/ is
-presented here (\cref{line:propline}).
-The typing ensures that all the mutations do not compromise
-backtracking, since sharing is only allowed through
-the immutable set functions, which is used for backtracking points.
+for lines, columns, and square, which are all very similar to function
+\lstinline/propagate_lines/ (\cref{line:propline}).
+
+The typing ensures that the mutations do not compromise the state at
+the choice point, because they operate on a new state \lstinline/new_g/ created for one
+particular branch of the choice.
 In particular, the \lstinline/try_solution/ function can be used in a loop
 precisely because the \lstinline/set/ function only requires an
-\emph{immutable} borrow of \lstinline/new_g/.
-This is allowed, since immutable borrows are unrestricted.
+unrestricted immutable borrow of \lstinline/g/, which can be used as
+many times as needed.
+We could even free \lstinline/new_g/ in \cref{line:free:new_g}.
 
-While we presented our API on simple copy-on-write arrays, this API
-could easily be adapted to any persistent data-structure with
+While presented for copy-on-write arrays, the API
+can easily be adapted to other persistent data structures with
 transient mutability such as RRB Vectors \cite{DBLP:journals/pacmpl/Puente17}
-and persistent HAMT (used notably as Clojure's default associative maps
-\cite{bagwell2001ideal,clojurehamt}) while providing a good performance profile
-and a convenient programming style.
+or persistent HAMT \cite{bagwell2001ideal,clojurehamt} to provide  a
+convenient programming style without compromising performance.
 

### 2019-07-04 kindly/motivation.tex
more sudoku

@@
 The signature differs slightly from the \lstinline/Array/
-signature. For simplicity, the array elements are unrestricted, 
-hence it is sufficent for the array to be affine.
+signature. For simplicity, the array elements are unrestricted, but
+the structure itself remains linear so as to be implemented in terms of \lstinline/Array/.
 The in-place mutation function \lstinline/set_mut/ with type
@@
 module type HybArray = sig
-  type ('a : un) t : aff
+  type ('a : un) t : lin
 
@@
 many times as needed.
-We could even free \lstinline/new_g/ in \cref{line:free:new_g}.
+As \lstinline/g/ is linear we must free it in \cref{line:free:g}.
 
@@
   \begin{minipage}[t]{0.54\linewidth}
-    \lstinputlisting[linerange=15-28,firstnumber=last]{code/sukodu.affe}
+    \lstinputlisting[linerange=15-29,firstnumber=last]{code/sukodu.affe}
   \end{minipage}

### 2019-07-04 kindly/lang.tex
Fix typos.
Move Typing stuff around.

--- 2019-07-02 kindly/lang.tex
@@
 
-\subsubsection{Constraint language}
+\subsubsection{Bindings and Weakening}
 
-\newcommand\A{\mathcal A}
-\newcommand\SC{\mathcal S}
+\subsubsection{Regions and Borrows}
 
-Let us note $\A$ our constraint system. The grammar of constraints is
-given in  \cref{grammar:constraint}.
-$\A$ is defined as the smallest cylindric term constraint system that
-satisfies the axiom shown in \cref{rules:entail}.
-We follows the traditional HM(X) formulation
-with conjunctions, projections and type inequalities.
-The new element specific to our approach are kind inequalities.
-Entailment is noted $\entail{C}{D}$, where $D$ is a consequence of the
-constraints $C$.
-We say that $C$ and $D$ are equivalent, noted $C \equivC D$,
-when $\entail{C}{D}$ and $\entail{D}{C}$.
-\TODO{Give the cylindric properties ?}
+\begin{figure}
+  \centering
+  \begin{subfigure}{0.4\linewidth}
+    \begin{mathpar}
+      \ruleSDRegion
+    \end{mathpar}
+    \caption{The {\sc Region} rule}
+  \end{subfigure}\hfill
+  \begin{subfigure}{0.6\linewidth}
+    \centering
+    \begin{tabular}
+      {@{}>{$}r<{$}@{ $\Lleftarrow$ }
+      >{$}c<{$}@{ $\rightsquigarrow_n^{x}$ }
+      >{$}l<{$}
+      r}
 
-\begin{figure}[tp]
-  \centering
-  \begin{align*}
-    C &::= \Cleq{\tau_1}{\tau_2}
-        \mid \Cleq{k_1}{k_2}
-        \mid C_1 \Cand C_2
-        \mid \Cproj{\tvar}{C}
-        \mid \Cproj{\kvar}{C}
-  \end{align*}
-  \caption{The constraint language}
-  \label{grammar:constraint}
+      {(\kun_n\lk k\lk\kun_\infty)}
+      &{\svar[\IBORROW]{x}{\tau}^n}
+      &{\bvar{\borrow[\IBORROW]{x}}{\borrowty[\IBORROW] k{\tau}}}
+      &Immut\\
+
+      (\kaff_n\lk k\lk\kaff_\infty)
+      &\svar[\MBORROW]{x}{\tau}^n
+      &\bvar{\borrow[\MBORROW]{x}}{\borrowty[\MBORROW] k{\tau}}
+      &Mut
+    \end{tabular}
+    \caption{Borrowing rules for bindings}
+  \end{subfigure}
 \end{figure}
 
-\begin{figure*}[tp]
-  \input{infer/lattice}
-  \caption{Lattice inequalities -- $k \lk_\Lat k'$}
-  \input{infer/entails}
-  \caption{Base entailment rules -- $\entail{C}{D}$ }
-  \label{rules:entail}
-\end{figure*}
+\subsubsection{Copying and Splitting}
 
-We note $\SC$ the set of solved forms
-which can be used inside type and kind schemes.
-We define $\SC$ as $\A$ quotiented by the relation $\equivC$.
-%
-We consider the existence of a function $\operatorname{normalize}$ which takes
-a constraint in $\A$ and a substitution $\psi$ and returns a constraint
-in solved form $C' \in \SC$,
-and an updated substitution. We detail the implementation
-of the normalization function in \cref{sec:normalize}
+\begin{figure}
+  \centering
+  \begin{subfigure}{0.35\linewidth}
+    \begin{mathpar}
+      \ruleSDPair
+    \end{mathpar}
+    \caption{The {\sc Pair} rule}
+  \end{subfigure}\hfill
+  \begin{subfigure}{0.6\linewidth}
+    \centering
+    \begin{tabular}
+      {@{}>{$}r<{$}@{ $\Lleftarrow$ }
+      >{$}c<{$}@{ $=$ }
+      >{$}c<{$}@{ $\ltimes$ }
+      >{$}c<{$}r}
 
-% $\mathcal S$ is composed only of kind
-% inequalities \emph{over variables}. For convenience, if $C\in\mathcal S$, we
-% note $C$ as a list of kind inequalities: $\Cleq{\kvar_i}{\kvar_{i'}}^n$.
-% \TODO{Extend the properties of solved forms}
+      \Cleq{\schm}{\kun_\infty}
+      &\bvar{x}{\schm}&\bvar{x}{\schm}&\bvar{x}{\schm}
+      &Both\\[2mm]
 
-\subsubsection{Kinding}
+      {\Cempty}&{\bvar{x}{\schm}}&{\bvar{x}{\schm}}&{\bnone}
+      &Left\\
+      {\Cempty}&{\bvar{x}{\schm}}&{\bnone}&{\bvar{x}{\schm}}
+      &Right\\[2mm]
 
-We note $\inferSK{C}{\E}{\tau}{k}$
-when $\tau$ has kind $k$ in environment $\E$ under constraints $C$.
-The rules are shown in \cref{rules:sd-kinding}.
-Kinds and types follow a small calculus with variables ($\tvar$,\dots),
-functions (type constructors $\T{t}$), application ($\tapp{t}{\Multi{\tau}}$)
-and primitives such as arrow ($\tau\tarr{k}\tau'$) and
-borrow ($\borrowty{k}{\tau}$).
-Kind inference is thus fairly straightforward and is done simply by following
-the syntax of the types. The only kind arrows can appear when looking
-up the kind scheme of a type constructor $\T t$. Kind arrows are forbidden
-in any other contexts.
+      {\Cempty}&{\bvar x \schm}&{\svar x \schm^n}&{\bvar x \schm}
+      &Susp\\
 
-\begin{figure}[ht]
-  \centering
-  \input{infer/sd-kinds}
-  \caption{Syntax-directed kinding rule --
-    $\inferSK{C}{\E}{\tau}{k}$}
-  \label{rules:sd-kinding}
+      {\Cempty}&
+      {\bvar{\borrow x} \schm}&{\svar[\IBORROW] x \schm^n}&{\bvar{\borrow x} \schm}
+      &SuspB\\[2mm]
+
+      {\Cempty}&
+      {\bvar{\borrow[i]{x}}{\schm}}&
+      {\bvar{\borrow[i]{x}}{\schm}}&{\bvar{\borrow[i]{x}}{\schm}}
+      &Borrow\\
+
+    \end{tabular}
+    \caption{Splitting rules for bindings}
+  \end{subfigure}
 \end{figure}
 
-\subsubsection{Environments}
-
-\begin{figure}[tp]
-  \input{infer/sd-splitting}
-  \caption{Splitting --- environments $\lsplit
-    C\E\E\E$; binders $\bsplit Cbbb$}
-  \label{fig:sd-splitting}
-\end{figure}
-
-\begin{figure}[tp]
-  \input{infer/sd-borrowing}
-  \caption{Borrowing --- environments $\lsplit
-    C\E\E\E$; binders $\bsplit Cbbb$}
-  \label{fig:sd-borrowing}
-\end{figure}
-
-\subsubsection{Typing}
-
-\begin{figure*}[tp]
-  \input{infer/sd-environments}
-  \caption{Rewriting constraints on environments and types}
-  \label{fig:contraints-environments-types}
-\end{figure*}
-\begin{figure*}[tp]
-  \input{infer/sd-typing}
-  \caption{Syntax-directed typing rules}
-  \label{fig:syntax-directed-typing}
-\end{figure*}
+\subsubsection{Constraints}
 
@@
 
-\begin{figure*}[htp]
+\begin{figure*}[!hbt]
   \centering

### 2019-07-04 kindly/sdtyping/extra.tex
Move Typing stuff around.

\section{Syntax directed typing}

\subsection{Constraint language}

\newcommand\A{\mathcal A}
\newcommand\SC{\mathcal S}

Let us note $\A$ our constraint system. The full grammar of constraints is
given in \cref{grammar:constraint}.
$\A$ is defined as the smallest cylindric term constraint system that
satisfies the axiom shown in \cref{rules:entail}.
We follows the traditional HM(X) formulation
with conjunctions, projections and type inequalities.
The new element specific to our approach are kind inequalities.
Entailment is noted $\entail{C}{D}$, where $D$ is a consequence of the
constraints $C$.
We say that $C$ and $D$ are equivalent, noted $C \equivC D$,
when $\entail{C}{D}$ and $\entail{D}{C}$.
\TODO{Give the cylindric properties ?}

\begin{figure}[tp]
  \centering
  \begin{align*}
    C &::= \Cleq{\tau_1}{\tau_2}
        \mid \Cleq{k_1}{k_2}
        \mid C_1 \Cand C_2
        \mid \Cproj{\tvar}{C}
        \mid \Cproj{\kvar}{C}
  \end{align*}
  \caption{The constraint language}
  \label{grammar:constraint}
\end{figure}

\begin{figure*}[tp]
  \input{infer/lattice}
  \caption{Lattice inequalities -- $k \lk_\Lat k'$}
  \input{infer/entails}
  \caption{Base entailment rules -- $\entail{C}{D}$ }
  \label{rules:entail}
\end{figure*}

We note $\SC$ the set of solved forms
which can be used inside type and kind schemes.
We define $\SC$ as $\A$ quotiented by the relation $\equivC$.
%
We consider the existence of a function $\operatorname{normalize}$ which takes
a constraint in $\A$ and a substitution $\psi$ and returns a constraint
in solved form $C' \in \SC$,
and an updated substitution. We detail the implementation
of the normalization function in \cref{sec:normalize}

% $\mathcal S$ is composed only of kind
% inequalities \emph{over variables}. For convenience, if $C\in\mathcal S$, we
% note $C$ as a list of kind inequalities: $\Cleq{\kvar_i}{\kvar_{i'}}^n$.
% \TODO{Extend the properties of solved forms}

\subsection{Kinding}

We note $\inferSK{C}{\E}{\tau}{k}$
when $\tau$ has kind $k$ in environment $\E$ under constraints $C$.
The rules are shown in \cref{rules:sd-kinding}.
Kinds and types follow a small calculus with variables ($\tvar$,\dots),
functions (type constructors $\T{t}$), application ($\tapp{t}{\Multi{\tau}}$)
and primitives such as types for arrows ($\tau\tarr{k}\tau'$) and
borrows ($\borrowty{k}{\tau}$).
Kind checking can thus be done in a fairly straightforward, syntax-directed
fashion by simply following
the syntax of the types. Kind arrows can only appear when looking
up the kind scheme of a type constructor $\T t$. Kind arrows are forbidden
in any other contexts.

\begin{figure}[ht]
  \centering
  \input{sdtyping/sd-kinds}
  \caption{Syntax-directed kinding rule --
    $\inferSK{C}{\E}{\tau}{k}$}
  \label{rules:sd-kinding}
\end{figure}

\subsection{Environments}

\begin{figure}[tp]
  \input{sdtyping/sd-splitting}
  \caption{Splitting --- environments $\lsplit
    C\E\E\E$; binders $\bsplit Cbbb$}
  \label{fig:sd-splitting}
\end{figure}

\begin{figure}[tp]
  \input{sdtyping/sd-borrowing}
  \caption{Borrowing --- environments $\lsplit
    C\E\E\E$; binders $\bsplit Cbbb$}
  \label{fig:sd-borrowing}
\end{figure}

\subsection{Typing}

\begin{figure*}[tp]
  \input{sdtyping/sd-environments}
  \caption{Rewriting constraints on environments and types}
  \label{fig:contraints-environments-types}
\end{figure*}
\begin{figure*}[tp]
  \input{sdtyping/sd-typing}
  \caption{Syntax-directed typing rules}
  \label{fig:syntax-directed-typing}
\end{figure*}

%%% Local Variables:
%%% mode: latex
%%% TeX-master: "../main"
%%% End:

### 2019-07-04 kindly/lang.tex
Add a section explaining typing of bindings and closures.

@@
 
-\subsubsection{Bindings and Weakening}
+We now present the essential parts of our \lang's type system. A complete
+description of the system is available in \cref{appendix:sdtyping}.
+In this section, we focus on chosen parts of our system
+through a set of selected rules.
+%
+We first define the following typing, kinding and constraint judgements:
+\begin{description}
+\item[$\inferS{C}{\E}{e}{\tau}$ --]
+  Expression $e$ has type $\tau$ in environment $\E$ under constraints $C$.
+\item[$\inferSK{C}{\E}{\tau}{k}$ --]
+  Type $\tau$ has kind $k$ in environment $\E$ under constraints $C$.
+\item[$\entail{D}{C}$ --] Constraint $D$ implies constraint $C$.
+\end{description}
+
+We will now present the novel aspects of our system which are necessary to handle
+three specific features: binders, borrows, and sharing.
 
@@
 
-\begin{figure}
+\begin{figure}[!h]
   \centering
@@
 
+\subsubsection{Bindings and Closures}
+
+One essential aspect of linear types is to control which variables
+are allowed to be unused. This is usually done through \emph{weakening},
+which allow to discard bindings from the environment.
+Another important feature is to control \emph{captures} done by closures.
+In \lang, we enforce both properties through the use of environment-wide
+kind inequalities, which we note $\Cleq{\E}{k}$.
+The intuitive meaning of $\Cleq{\E}{k}$ is that for all bindings
+$\bvar{x}{\tau}$ in $\E$,
+if type $\tau$ has kind $k'$ ($\inferSK{C}{\E}{\tau}{k'}$)
+then $\Cleq{k'}{k}$.
+While this definition is incomplete, it is sufficient to understand
+its use in the {\sc Abs} and {\sc Var} rules shown in \cref{selectrules:binders}.
+%
+The {\sc Var} rule looks up the type scheme of the variable $x$ in
+the environment $\E$
+and instantiate a concrete type through the substitution $\unif$. It then
+checks that the other bindings in $\E$ can safely be discarded by
+ensuring they are not linear. This is achieved through the constraint
+$\Cleq{\E\Sdel{X}}{\kaff_\infty}$.
+%
+The {\sc Lam} constraints must ensures that if the body of the lambda captures
+linear or affine variable, the content will be reflected in the kind
+annotated on the arrow, such as $\tau_2 \tarr{k} \tau_1$.
+In order to do so, we constraint all the bindings of $\E$ to be less than $k$.
+If one of the binding of $\E$ constraints a type whose kind is affine,
+we will derive the constraints $\Cleq{\kaff_n}{k}$, which ensures that
+the kind on the arrow is at least affine.
+
+\begin{figure}[!h]
+  \centering
+  \begin{mathpar}
+    \ruleSDLam
+
+    \ruleSDVar
+  \end{mathpar}
+  \caption{The {\sc Lam} and {\sc Var} rules}
+  \label{selectrules:binders}
+\end{figure}
+
 \subsubsection{Copying and Splitting}
 
-\begin{figure}
+\begin{figure}[!h]
   \centering

### 2019-07-04 kindly/lang.tex
Start explanation for borrow and regions.

@@
 
+In order to provide good support for imperative programming, \affe allows
+to borrows a resource. Such borrows are constrained to a given
+region.
+In order to support such features, we differentiate several sorts of
+bindings in typing environments.
+The grammar of type environment (noted $\E$) is shown in \cref{grammar:env}.
+On top of the traditional type and value bindings, noted respectively
+$\bvar{\tvar}{\kschm}$ and $\bvar{x}{\schm}$, we also consider two new
+bindings.
+Borrow bindings $\bvar{\borrow{x}}{\borrowty{k}{\tau}}$ which indicates
+that the borrow $\borrow{x}$ can be used directly, as implemented
+by the rule {\sc Borrow} in \cref{selectrules:borrow}
+Suspended bindings $\svar{x}{\schm}^n$ indicates that $x$ is alive, but
+can not be used immediately.
+In order to use a suspended binding for the variable $x$,
+we need to enter a \emph{region} associated with $x$,
+as shown in the {\sc Region} rule.
+
+\TODO{Explain region rule}
+
 \begin{figure}[!h]
   \centering
-  \begin{subfigure}{0.4\linewidth}
   \begin{mathpar}
+    \ruleSDBorrow
+
     \ruleSDRegion
   \end{mathpar}
-    \caption{The {\sc Region} rule}
-  \end{subfigure}\hfill
-  \begin{subfigure}{0.6\linewidth}
+  \caption{The {\sc Borrow} and {\sc Region} rules}
+  \label{selectrules:borrow}
+  \vspace{1em}
+
+  \begin{minipage}{0.35\linewidth}
+    \begin{align*}
+      \htag{Type environments}
+      \E &::=\ \Eempty \mid \E;B\\
+      \htag{Bindings}
+      B &::=\ \bnone \tag{Empty} \\
+      &\mid \bvar{x}{\schm}\tag{Variables}\\
+      &\mid \bvar{\borrow{x}}{\borrowty{k}{\tau}}\tag{Borrows}\\
+      &\mid \svar{x}{\schm}^n \tag{Suspended}\\
+      &\mid \bvar{\tvar}{\kschm}\tag{Types}
+      % |&\ \tydecl{t}{\kschm}{K}{\tau}\tag{Type Declaration}\\
+    \end{align*}
+    \caption{Type environments}
+    \label{grammar:env}
+  \end{minipage}\hfill
+  \begin{minipage}{0.6\linewidth}
     \centering
@@
     \caption{Borrowing rules for bindings}
-  \end{subfigure}
-\end{figure}
-
-\subsubsection{Bindings and Closures}
-
-One essential aspect of linear types is to control which variables
-are allowed to be unused. This is usually done through \emph{weakening},
-which allow to discard bindings from the environment.
-Another important feature is to control \emph{captures} done by closures.
-In \lang, we enforce both properties through the use of environment-wide
-kind inequalities, which we note $\Cleq{\E}{k}$.
-The intuitive meaning of $\Cleq{\E}{k}$ is that for all bindings
-$\bvar{x}{\tau}$ in $\E$,
-if type $\tau$ has kind $k'$ ($\inferSK{C}{\E}{\tau}{k'}$)
-then $\Cleq{k'}{k}$.
-While this definition is incomplete, it is sufficient to understand
-its use in the {\sc Abs} and {\sc Var} rules shown in \cref{selectrules:binders}.
-%
-The {\sc Var} rule looks up the type scheme of the variable $x$ in
-the environment $\E$
-and instantiate a concrete type through the substitution $\unif$. It then
-checks that the other bindings in $\E$ can safely be discarded by
-ensuring they are not linear. This is achieved through the constraint
-$\Cleq{\E\Sdel{X}}{\kaff_\infty}$.
-%
-The {\sc Lam} constraints must ensures that if the body of the lambda captures
-linear or affine variable, the content will be reflected in the kind
-annotated on the arrow, such as $\tau_2 \tarr{k} \tau_1$.
-In order to do so, we constraint all the bindings of $\E$ to be less than $k$.
-If one of the binding of $\E$ constraints a type whose kind is affine,
-we will derive the constraints $\Cleq{\kaff_n}{k}$, which ensures that
-the kind on the arrow is at least affine.
-
-\begin{figure}[!h]
-  \centering
-  \begin{mathpar}
-    \ruleSDLam
-
-    \ruleSDVar
-  \end{mathpar}
-  \caption{The {\sc Lam} and {\sc Var} rules}
-  \label{selectrules:binders}
+  \end{minipage}
 \end{figure}
@@
   \end{subfigure}
+\end{figure}
+
+\subsubsection{Bindings and Closures}
+
+One essential aspect of linear types is to control which variables
+are allowed to be unused. This is usually done through \emph{weakening},
+which allow to discard bindings from the environment.
+Another important feature is to control \emph{captures} done by closures.
+In \lang, we enforce both properties through the use of environment-wide
+kind inequalities, which we note $\Cleq{\E}{k}$.
+The intuitive meaning of $\Cleq{\E}{k}$ is that for all bindings
+$\bvar{x}{\tau}$ in $\E$,
+if type $\tau$ has kind $k'$ ($\inferSK{C}{\E}{\tau}{k'}$)
+then $\Cleq{k'}{k}$.
+While this definition is incomplete, it is sufficient to understand
+its use in the {\sc Abs} and {\sc Var} rules shown in \cref{selectrules:binders}.
+%
+The {\sc Var} rule looks up the type scheme of the variable $x$ in
+the environment $\E$
+and instantiate a concrete type through the substitution $\unif$. It then
+checks that the other bindings in $\E$ can safely be discarded by
+ensuring they are not linear. This is achieved through the constraint
+$\Cleq{\E\Sdel{X}}{\kaff_\infty}$.
+%
+The {\sc Lam} constraints must ensures that if the body of the lambda captures
+linear or affine variable, the content will be reflected in the kind
+annotated on the arrow, such as $\tau_2 \tarr{k} \tau_1$.
+In order to do so, we constraint all the bindings of $\E$ to be less than $k$.
+If one of the binding of $\E$ constraints a type whose kind is affine,
+we will derive the constraints $\Cleq{\kaff_n}{k}$, which ensures that
+the kind on the arrow is at least affine.
+
+\begin{figure}[!h]
+  \centering
+  \begin{mathpar}
+    \ruleSDLam
+
+    \ruleSDVar
+  \end{mathpar}
+  \caption{The {\sc Lam} and {\sc Var} rules}
+  \label{selectrules:binders}
 \end{figure}

### 2019-07-05 kindly/sdtyping/extra.tex
Add a section explaining typing of bindings and closures.
Extend/reorganize the syntax-directed typing section.

@@
+\section{Annotating regions}
+\label{regionannot}
+
+So far, all \lang programs have been fully annotated with regions information.
+We now show how to infer these regions annotations based on
+optionally-annotated programs.
+First, we extend the region annotation to $\region{S}{E}$ where $S$ is
+a set of variables. This annotation, defined below, is equivalent to nested
+region annotations for each individual variable.
+
+\begin{align*}
+  \region{x;S}{e} &= \region{x}{\region{S}{e}}& \region{\emptyset}{e} &= e\\
+\end{align*}
+
+\Cref{fig:region-annotation} define a rewriting relation $\RannotT{e}{e'}$
+which indicates that an optionally annotated term $e$ can be written
+in a fully annotated term $e'$.
+Through the rule \textsc{Rewrite-Top}, this is defined
+in term of an inductively defined relation
+$\Rannot{e}{e'}{S}$ where $n$ is the current nesting and $S$ is a set of
+variable that are not enclosed in a region yet.
+The base cases are constants, variables and borrows.
+The general idea is to start from the leafs of the syntax tree, create a
+region for each borrow, and enlarge the region as much as possible.
+This is implemented by a depth-first walk of the syntax
+tree which collects each variable that has a corresponding borrow.
+At each step, it rewrites the inner subterms,
+consider which borrow must be enclosed by a region now, and
+return the others for later enclosing. Binders force immediate
+enclosing of the bound variables, as demonstrated in rule \textsc{Rewrite-Lam}.
+For nodes with multiple children, we
+use a scope merge operator to decide if regions should be placed and where.
+This is shown in rule \textsc{Rewrite-Pair}.
+The merge operator, written $\getBorrows{B_l}{B_r}{(S_l,S,S_r)}$, takes
+the sets $B_l$ and $B_r$ returned by rewriting the subterms
+and returns three sets: $S_l$ and $S_r$ indicates the variables
+that should be immediately enclosed by a region on the left and right
+subterms and $S$ indicates the set of the yet-to-be-enclosed variables.
+As an example, the rule \textsc{AnnotRegion-MutLeft} is applied
+when there is an immutable borrow and a mutable borrow. In that case, a
+region is created to enclose the immutable borrow, while the mutable
+borrow is left to be closed later. This is coherent with the rules
+for environment splitting and suspended bindings from \cref{sdtyping}.
+%
+Explicitly annotated regions are handled specially through
+rule \textsc{Rewrite-Region}. In that case, we assume that all inner
+borrows should be enclosed immediately.
+
+\begin{figure*}[!hbt]
+  \centering
+  \input{region-annot}
+  \caption{Automatic region annotation --- $\RannotT{e}{e'}$}
+  \label{fig:region-annotation}
+\end{figure*}
+
 \section{Syntax directed typing}
+\label{appendix:sdtyping}
 

### 2019-07-05 kindly/motivation.tex
motivated

@@
   \centering
-  \begin{subfigure}{0.48\linewidth}
+  \begin{subfigure}[t]{0.48\linewidth}
 \begin{lstlisting}
@@
   \end{subfigure}\hfill
-  \begin{subfigure}{0.5\linewidth}
+  \begin{subfigure}[t]{0.5\linewidth}
 \begin{lstlisting}
@@
     \label{ex:array}
-  \end{subfigure}
-\end{figure}
-
-\cref{ex:array:region} makes explicit the regions from the example
-indicated by braces as in \lstinline/{| ... |}/.  \lang guarantees that borrows
-never escape the smallest enclosing region.  For instance, in
-\cref{ex:array:region:get}, \lstinline/&a/ does not escape because it
-is consumed in the calls to \lstinline/get/.
-Regions annotations are automatically inserted using
-syntactic information such as borrows and binders (see
-\cref{regionannot} for more information).  Users can also
-annotate regions manually to further restrict the range of a
-borrow. In practice, we have found that region annotations are
-rarely needed with common programming idioms. All regions for the
-examples in this section have been inferred.
-
-\begin{figure}
-  \centering
-  % \begin{subfigure}{0.6\linewidth}
 \begin{lstlisting}
-let mk_fib_array n =
 let a = create n 1 in
@@
   set &!a i x
-  |}
-  done;
-  a(*@\label{ex:array:return}*)
-# mk_fib_array : int -> int Array.t
+|} done;
 \end{lstlisting}
-    \caption{Example from \cref{ex:array} with explicit regions}
+    \caption{Excerpt of \cref{ex:array} with explicit regions}
     \label{ex:array:region}
-  % \end{subfigure}
-\end{figure}
-
-\begin{figure}
-  \centering
-  \begin{subfigure}{0.5\linewidth}
-\begin{lstlisting}
-let for f a  = 
-  let n = Array.length &a in
-  let rec loop f a i = 
-    if i < n then (
-      f &!a i;
-      loop f &!a (i+1)
-    )
-  in
-  loop f &!a 0
-# for : Borrowed m ('a Array.t) tt =>
-  (tt -> int -> unit) -> tt -> unit
-\end{lstlisting}
-    \caption{For loop function for affine arrays}
-    \label{ex:array:for}
   \end{subfigure}
@@
 
+\cref{ex:array:region} uses braces as in \lstinline/{| ... |}/ to
+explicitly state the regions of the example in \cref{ex:array}.  One
+region consists of the header expression of the \lstinline/let/ in
+\cref{ex:array:region:get}. It is contained in another region that
+comprises the body of the \lstinline/for/ loop. \lang guarantees that borrows 
+never escape the smallest enclosing region. It employs a system of
+\emph{indexed kinds} like \lstinline{aff_r} and \lstinline{un_r} where
+$r$ is a positive integer that corresponds to the lexical nesting
+depth of regions. For instance, the type of \lstinline{&!a} in
+\cref{ex:array:set} has kind \lstinline{aff_1} whereas the type of
+\lstinline{&a} in \cref{ex:array:region:get} has kind
+\lstinline{unr_2} and the typing of its region is such that types with
+kind indexes greater than or equal to $2$ cannot escape.
+
+In \cref{ex:array:region:get} of the example, there is no danger of \lstinline/&a/ escaping because it
+is consumed in the calls to \lstinline/get/.
+Region annotations are automatically inserted using
+syntactic cues such as borrows and binders (see 
+\cref{regionannot}).  Users may manually insert regions to further restrict the range of a
+borrow. In practice, region annotations are
+rarely needed with common programming idioms.
+% All regions for the examples in this section have been inferred.
+
+% \begin{figure}
+%   \centering
+%   % \begin{subfigure}{0.6\linewidth}
+% \begin{lstlisting}
+% let mk_fib_array n =
+%   let a = create n 1 in
+%   for i = 2 to n - 1 do {|
+%     let x = {| get &a (i-1) + get &a (i-2) |} in(*@\label{ex:array:region:get}*)
+%     set &!a i x
+%   |}
+%   done;
+%   a(*@\label{ex:array:return}*)
+% # mk_fib_array : int -> int Array.t
+% \end{lstlisting}
+%     \caption{Example from \cref{ex:array} with explicit regions}
+%     \label{ex:array:region}
+%   % \end{subfigure}
+% \end{figure}
+
 \subsection{Hybrid programming style}
 
-In the last few years, new persistent data structures
-have been introduced which permit transient mutations, but discourage
-non-linear uses either through degraded performance
-\cite{DBLP:conf/ml/ConchonF07} or
+Recently introduced persistent data structures
+permit transient mutations where
+non-linear uses lead to degraded performance
+\cite{DBLP:conf/ml/ConchonF07} or to
 dynamic and static checks \cite{DBLP:journals/pacmpl/Puente17}.
-In particular, persistent Hash-Array-Mapped-Tries have been used with similar
+In particular, persistent Hash-Array-Mapped-Tries (HAMT) have been used with similar
 APIs in several non-pure functional languages (OCaml, Clojure, \dots).
-Affine types allow us to formalize the performance contract between the programmer
+Affine types help formalize the performance contract between the programmer
 and the library, while borrows avoid the need to thread state explicitly,
-as usually required by an API that enforces immutability.
+as usually required by an API for immutable data types.
 %
-In this section, we show how \lang allows us to define
+
+In this section, we present
 a safe API for persistent arrays that support both immutable and mutable features,
-and how it can be used to implement a simple backtracking Sudoku solver.
-
+and use it to implement a backtracking Sudoku solver.
+% show how \lang allows us to define
 The Sudoku solver maintains an array to represent the state of
@@
 can easily be adapted to other persistent data structures with
-transient mutability such as RRB Vectors \cite{DBLP:journals/pacmpl/Puente17}
-or persistent HAMT \cite{bagwell2001ideal,clojurehamt} to provide  a
+transient mutability such as Relaxed-Radix Balance Vectors (RRB) \cite{DBLP:journals/pacmpl/Puente17}
+or persistent HAMTs \cite{bagwell2001ideal,clojurehamt} to provide  a
 convenient programming style without compromising performance.
@@
 
-We have seen that borrows are limited to a region. In the various examples so
-far, such regions didn't appear in the type signatures explicitly.
-We know present an pool implementation which can store linear objects
-and such region to enforce proper uses of the objects.
-The implementation and interfaces of the \texttt{Pool} module
-are shown in \cref{impl:pool,intf:pool}. A simple example of use
-is shown in \cref{ex:pool}.
+% We have seen that borrows are limited to a region.
+In the examples so far, regions do not appear in type signatures.
+But for certain programming idioms, we want to extend the scope of a
+region across a function boundary. As an example, 
+we present an interface and implementation of a pool of linear resources where the
+extended scope of the region enforces proper use of the resources.
 
+\cref{intf:pool} contains the interface of the \lstinline/Pool/ module.
 A pool is parameterized by its content. The kind of the pool
 depends on the content: linear content implies
-a linear pool while unrestricted content implies an unrestricted pool.
-The functions \lstinline/Pool.create/ and \lstinline/Pool.consume/ allow
-to build or destroy
-the pool by providing the corresponding functions for the elements
+a linear pool while unrestricted content yields an unrestricted pool.
+The functions \lstinline/Pool.create/ and \lstinline/Pool.consume/
+build/destroy a pool given creators/destructors for the elements
 of the pool.
-The main function is \lstinline/Pool.use/ which
-allow to use an object from the pool.
-\lstinline/Pool.use/ takes a shared borrow of a pool (in order to allow
-for concurrent access) and a function. This function will receive
-a mutable borrow of one arbitrary object of the pool. The typing ensure
-that this borrow is not captured nor kept by the function.
-This is done through a \emph{region} variable $r$ which annotates
-the kinds. Here, the borrow is of type \lstinline/&!(aff_r,'a_1 )/, which indicates
-a mutable borrow with an affine kind. The return type of \lstinline/f/ is bound
-to be of kind \lstinline/'k_2 < aff_r/. Since this is a strict inequality,
-types whose kind mention the region can not be returned.
+The function \lstinline/Pool.use/ is the workhorse of the API, which
+borrows a resource from the pool to a callback.
+It takes a shared borrow of a pool (to enable concurrent access) and a
+callback function.
+The callback receives a mutable borrow of an arbitrary resource from the pool. 
+The typing of the callback ensures
+that this borrow is neither captured nor returned by the function.
 
-The implementation presented in \cref{impl:pool} uses a concurrent queue
-as a bag of elements with atomic add and remove operations.
-The implementation for the \lstinline/Pool.create/ and \lstinline/Pool.consume/
-functions are straightforward.
+This encapsulation is implemented with a universally quantified \emph{kind index variable} $r$.
+The signature prescribes the type \lstinline/&!(aff_r1,'a_1 )/ for the
+mutable borrow of the resource with an affine kind at region nesting $r+1$. The return
+type of the callback is constrained to kind \lstinline/'k_2 <= aff_r/
+so that the callback certainly cannot return the borrowed argument. 
+In a specific use of \lstinline/Pool.use/, the index $r$ gets unified
+with the current nesting level of regions so that the region for the
+callback effectively gets ``inserted'' into the lexical nesting at the callsite.
+
+\cref{ex:pool} shows a simple example of using the \lstinline/Pool/ module.
+
+The implementation in \cref{impl:pool} represents a bag of resourfces
+using a concurrent queue with atomic add and remove operations. 
+The implementation of the \lstinline/Pool.create/ and \lstinline/Pool.consume/
+functions is straightforward.
 The function \lstinline/Pool.use/ first draws
 an element from the pool (or creates a fresh element),
-pass it down to the \lstinline/f/ function, and then add
-it back to the pool.
-We explicitly delimit the region on \cref{line:pool:region} to ensure that
-the return value of \lstinline/f &!o/ do not capture \lstinline/&!o/.
+passes it down to the callback function \lstinline/f/, and returns
+it to the pool afterwards.
+We explicitly delimit the region in \cref{line:pool:region} to ensure that
+the return value of \lstinline/f &!o/ does not capture \lstinline/&!o/.
 
-\begin{figure}[!h]
+\begin{figure}[tp]
   \centering
+  \begin{subfigure}[t]{.55\linewidth}
+\begin{lstlisting}
+type ('a:'k) pool : 'k
+
+create : (unit -> 'a) -> 'a pool
+consume : ('a -> unit) -> 'a pool -> unit
+use : ('a_1:'k_1),('a_2:'k_2),('k_2 <= aff_r) =>
+  &('a_1 pool) -> (&!(aff_r1,'a_1) -{lin}> 'a_2) -{'k_1}> 'a_2
+\end{lstlisting}
+    \caption{Interface}
+    \label{intf:pool}
+
+\begin{lstlisting}
+let uri = "postgresql://localhost:5432"
+
+(* Create a database connection pool. *)
+let pool = Pool.create (fun () -> Db.connect uri)
+
+(* Use the pool in queries. *)
+let create_user name =
+  Pool.use &pool (fun connection ->
+    Db.insert "users" [("name", name)] connection)
+\end{lstlisting}
+    \caption{Example of use}
+    \label{ex:pool}
+  \end{subfigure}~
   \begin{subfigure}[t]{.45\linewidth}
@@
     \label{impl:pool}
-  \end{subfigure}~
-  \begin{subfigure}[t]{.55\linewidth}
-\begin{lstlisting}
-type ('a:'k) pool : 'k
-
-create : (unit -> 'a) -> 'a pool
-consume : ('a -> unit) -> 'a pool -> unit
-use : ('a_1:'k_1),('a_2:'k_2),('k_1 < aff_r),('k_2 < aff_r) =>
-  &('a_1 pool) -> (&!(aff_r,'a_1) -{lin}> 'a_2) -{'k_1}> 'a_2
-\end{lstlisting}
-    \caption{Interface}
-    \label{intf:pool}
-
-\begin{lstlisting}
-let uri = "postgresql://localhost:5432"
-
-(* Create a database connection pool. *)
-let pool = Pool.create (fun () -> Db.connect uri)
-
-(* Use the pool in queries. *)
-let create_user name =
-  Pool.use &pool (fun connection ->
-    Db.insert "users" [("name", name)] connection)
-\end{lstlisting}
-    \caption{Example of use}
-    \label{ex:pool}
   \end{subfigure}

### 2019-07-05 kindly/lang.tex
Reordering.

--- 2019-07-04 kindly/lang.tex
@@
 
-Environments are composed of value and type bindings, along with type
-declarations. Type declarations, of the form
-$\tydecl{t}{\kschm}{K}{\tau}$, introduce both a new type constructor $\T{t}$ and
-a data constructor $K$.
+% Environments are composed of value and type bindings, along with type
+% declarations. Type declarations, of the form
+% $\tydecl{t}{\kschm}{K}{\tau}$, introduce both a new type constructor $\T{t}$ and
+% a data constructor $K$.
 
@@
 \end{figure*}
-\input{internal-grammar}
 
-\clearpage
-\subsection{Semantics}
-\label{sem}
-
-\input{fig-reduction}
-\input{fig-code}
-
-\clearpage
 \subsection{Typing}
@@
 
-\subsubsection{Regions and Borrows}
+\subsubsection{Environments and bindings}
 
-\begin{figure}[!h]
-  \centering
-  \begin{subfigure}{0.4\linewidth}
-    \begin{mathpar}
-      \ruleSDRegion
-    \end{mathpar}
-    \caption{The {\sc Region} rule}
-  \end{subfigure}\hfill
-  \begin{subfigure}{0.6\linewidth}
-    \centering
-    \begin{tabular}
-      {@{}>{$}r<{$}@{ $\Lleftarrow$ }
-      >{$}c<{$}@{ $\rightsquigarrow_n^{x}$ }
-      >{$}l<{$}
-      r}
+One main feature of our system is to control the usage of variables, either
+through
+linearity and affinity, which controls how variables are shared or ignored,
+or through borrows, which allow to circumvent the linearity rules
+in a controlled way.
+In order to check these properties, we introduce several new bindings
+in our typing environments.
+The grammar of type environment (noted $\E$) is shown in \cref{grammar:env}.
+On top of the traditional type and value bindings, noted respectively
+$\bvar{\tvar}{\kschm}$ and $\bvar{x}{\schm}$, we also consider two new
+bindings.
+Borrow bindings $\bvar{\borrow{x}}{\borrowty{k}{\tau}}$ which indicates
+that the borrow $\borrow{x}$ can be used directly, as implemented
+by the rule {\sc Borrow} in \cref{selectrules:borrow}.
+Suspended bindings, noted $\svar{x}{\schm}^n$, indicates that $x$
+cannot be used immediately, but will be usable in a region
+marked with $x$ (as detailed in \cref{sdtyp:regions}).
 
-      {(\kun_n\lk k\lk\kun_\infty)}
-      &{\svar[\IBORROW]{x}{\tau}^n}
-      &{\bvar{\borrow[\IBORROW]{x}}{\borrowty[\IBORROW] k{\tau}}}
-      &Immut\\
-
-      (\kaff_n\lk k\lk\kaff_\infty)
-      &\svar[\MBORROW]{x}{\tau}^n
-      &\bvar{\borrow[\MBORROW]{x}}{\borrowty[\MBORROW] k{\tau}}
-      &Mut
-    \end{tabular}
-    \caption{Borrowing rules for bindings}
-  \end{subfigure}
-\end{figure}
-
-\subsubsection{Bindings and Closures}
-
-One essential aspect of linear types is to control which variables
-are allowed to be unused. This is usually done through \emph{weakening},
-which allow to discard bindings from the environment.
-Another important feature is to control \emph{captures} done by closures.
-In \lang, we enforce both properties through the use of environment-wide
-kind inequalities, which we note $\Cleq{\E}{k}$.
+In \lang, many properties are encoded through the use of environment-wide
+constraints. Here we present how to control both unused variables and
+closure captures through such environment constraints, noted $\Cleq{\E}{k}$.
 The intuitive meaning of $\Cleq{\E}{k}$ is that for all bindings
@@
 \begin{figure}[!h]
-  \centering
+  \begin{minipage}{0.33\linewidth}
+    \begin{align*}
+      \E &::=\ \Eempty \mid \E;B\tag{Environments}\\
+      B &::=\ \bnone \tag{Empty} \\
+         &\mid \bvar{x}{\schm}\tag{Variables}\\
+         &\mid \bvar{\borrow{x}}{\borrowty{k}{\tau}}\tag{Borrows}\\
+         &\mid \svar{x}{\schm}^n \tag{Suspended}\\
+         &\mid \bvar{\tvar}{\kschm}\tag{Types}
+         % |&\ \tydecl{t}{\kschm}{K}{\tau}\tag{Type Declaration}\\
+    \end{align*}
+    \caption{Type environments}
+    \label{grammar:env}
+  \end{minipage}\hfill
+  \begin{minipage}{0.65\linewidth}
     \begin{mathpar}
+      \ruleSDBorrow
+      \and
       \ruleSDLam
-
+      \and
       \ruleSDVar
     \end{mathpar}
-  \caption{The {\sc Lam} and {\sc Var} rules}
-  \label{selectrules:binders}
+    \caption{The {\sc Borrow} rule}
+    \label{selectrules:borrow}
+  \end{minipage}
 \end{figure}
@@
 \subsubsection{Copying and Splitting}
+
+Linear and affine types also control the copy of variables.
+We now showcase how this is achieved in \lang through the {\sc Pair} typing rule
+in \cref{sdtyp:pair}.
+\TODO{Explain splitting}
 
@@
     \caption{The {\sc Pair} rule}
+    \label{sdtyp:pair}
   \end{subfigure}\hfill
@@
     \begin{tabular}
-      {@{}>{$}r<{$}@{ $\Lleftarrow$ }
+      {@{}>{$}r<{$}@{ $\vdash_e$ }
       >{$}c<{$}@{ $=$ }
@@
 
+\subsubsection{Regions}
+\label{sdtyp:regions}
+
+In order to provide good support for imperative programming, \affe allows
+to borrow resources.
+Such borrows are constrained to a given syntactic region. We note
+$\region{x}{e}$ for the region allowing borrows of $x$ to be used in $e$
+and where $n$ is the nesting of the region.
+Borrows can not be used outside of a corresponding region.
+These remarks lead to an
+algorithm for automatic region annotation: start from each borrow, create
+a region, and extend it as much as possible while respecting other borrows
+and binders.
+A complete specification of the algorithm is provided in \cref{regionannot}.
+In the rest of this section, we will only consider terms
+with fully annotated regions.
+
+\TODO{Transition}
+
+In order to use a suspended binding for the variable $x$,
+we need to enter a \emph{region} associated with $x$,
+noted $\region{x}{e}$.
+The main idea of region typing (implemented by the {\sc Region} rule)
+is to replace suspended bindings by borrow bindings, typecheck the body
+of the region, and finally ensure that the borrow made available inside
+the region does not leak outside.
+This last check is done by using integer indices that correspond to the nesting
+level of the region. The kind of borrow is annotated with the integer
+corresponding to their region, and a final constraint ensure that
+the return type of the region does not contain such borrow.
+
+We illustrate this on the expression $\region{x}{f(\borrow[\IBORROW]{c})}$.
+The first step is to check that
+$\svar[\IBORROW]{c}{\text{channel}}$ is in $\E$. We then define $\E'$ which
+corresponds to $\E$ where the suspended binding is replaced by a borrow binding.
+%
+This is done by the borrow relation, noted $\lregion{C}{x}{\E}{\E'}$ and
+whose main rules are shown in \cref{env:rule:region}.
+In our case, we consider an immutable binding, so we replace the suspended
+binding by $\bvar{\borrow[\IBORROW]{c}}{\borrowty[\IBORROW]{k}{\text{channel}}}$.
+We must also check that $\entail{C}{(\kun_n\lk k\lk\kun_\infty)}$.
+We can then typecheck the body of the region, $f(\borrow[\IBORROW]{c})$ for our example and check that the borrow made available in the region does not escape.
+This is done by the constraint $\Cleq{\tau}{\klin_{n-1}}$, which ensures
+that the kind of $\tau$ does not contain anything whose kind is ``above $n$''
+(for instance, $\borrowty[\IBORROW]{\kun_n}{\text{channel}}$ would have kind
+$\kun_n$).
+
+\begin{figure}[!h]
+  \begin{minipage}{0.35\linewidth}
+    \centering
+    \begin{mathpar}
+      \ruleSDRegion
+    \end{mathpar}
+    \caption{The {\sc Region} rule}
+    \label{selectrules:region}
+  \end{minipage}\hfill
+  \begin{minipage}{0.6\linewidth}
+    \centering
+    \begin{tabular}
+      {@{}>{$}r<{$}@{ $\vdash_e$ }
+      >{$}c<{$}@{ $\rightsquigarrow_n^{x}$ }
+      >{$}l<{$}
+      r}
+
+      {(\kun_n\lk k\lk\kun_\infty)}
+      &{\svar[\IBORROW]{x}{\tau}^n}
+      &{\bvar{\borrow[\IBORROW]{x}}{\borrowty[\IBORROW] k{\tau}}}
+      &Immut\\
+
+      (\kaff_n\lk k\lk\kaff_\infty)
+      &\svar[\MBORROW]{x}{\tau}^n
+      &\bvar{\borrow[\MBORROW]{x}}{\borrowty[\MBORROW] k{\tau}}
+      &Mut
+    \end{tabular}
+    \caption{Borrowing rules for bindings -- $\lregion{C}{x}{\E}{\E'}$}
+    \label{env:rule:borrow}
+  \end{minipage}
+\end{figure}
+
 \subsubsection{Constraints}
@@
 \clearpage
+\subsection{Semantics}
+\label{sem}
 
-\subsection{Annotating regions}
-\label{regionannot}
-
-So far, all \lang programs have been fully annotated with regions information.
-We now show how to infer these regions annotations based on
-optionally-annotated programs.
-First, we extend the region annotation to $\region{S}{E}$ where $S$ is
-a set of variables. This annotation, defined below, is equivalent to nested
-region annotations for each individual variable.
-
-\begin{align*}
-  \region{x;S}{e} &= \region{x}{\region{S}{e}}& \region{\emptyset}{e} &= e\\
-\end{align*}
-
-\Cref{fig:region-annotation} define a rewriting relation $\RannotT{e}{e'}$
-which indicates that an optionally annotated term $e$ can be written
-in a fully annotated term $e'$.
-Through the rule \textsc{Rewrite-Top}, this is defined
-in term of an inductively defined relation
-$\Rannot{e}{e'}{S}$ where $n$ is the current nesting and $S$ is a set of
-variable that are not enclosed in a region yet.
-The base cases are constants, variables and borrows.
-The general idea is to start from the leafs of the syntax tree, create a
-region for each borrow, and enlarge the region as much as possible.
-This is implemented by a depth-first walk of the syntax
-tree which collects each variable that has a corresponding borrow.
-At each step, it rewrites the inner subterms,
-consider which borrow must be enclosed by a region now, and
-return the others for later enclosing. Binders force immediate
-enclosing of the bound variables, as demonstrated in rule \textsc{Rewrite-Lam}.
-For nodes with multiple children, we
-use a scope merge operator to decide if regions should be placed and where.
-This is shown in rule \textsc{Rewrite-Pair}.
-The merge operator, written $\getBorrows{B_l}{B_r}{(S_l,S,S_r)}$, takes
-the sets $B_l$ and $B_r$ returned by rewriting the subterms
-and returns three sets: $S_l$ and $S_r$ indicates the variables
-that should be immediately enclosed by a region on the left and right
-subterms and $S$ indicates the set of the yet-to-be-enclosed variables.
-As an example, the rule \textsc{AnnotRegion-MutLeft} is applied
-when there is an immutable borrow and a mutable borrow. In that case, a
-region is created to enclose the immutable borrow, while the mutable
-borrow is left to be closed later. This is coherent with the rules
-for environment splitting and suspended bindings from \cref{sdtyping}.
-%
-Explicitly annotated regions are handled specially through
-rule \textsc{Rewrite-Region}. In that case, we assume that all inner
-borrows should be enclosed immediately.
-
-\begin{figure*}[!hbt]
-  \centering
-  \input{region-annot}
-  \caption{Automatic region annotation --- $\RannotT{e}{e'}$}
-  \label{fig:region-annotation}
-\end{figure*}
+\input{internal-grammar}
+\input{fig-reduction}
+\input{fig-code}
 

### 2019-07-05 kindly/motivation.tex
preserved linearity

@@
 \begin{lstlisting}
-let uri = "postgresql://localhost:5432"
-
-(* Create a database connection pool. *)
-let pool = Pool.create (fun () -> Db.connect uri)
-
-(* Use the pool in queries. *)
-let create_user name =
+(* Using the pool in queries. *)
+let create_user pool name =
   Pool.use &pool (fun connection ->
     Db.insert "users" [("name", name)] connection)
+
+let uri = "postgresql://localhost:5432"
+let main users =
+  (* Create a database connection pool. *)
+  let pool = Pool.create (fun _ -> Db.connect uri)
+  List.parallel_iter (create_user &pool) users
+  Pool.consume (fun c -> Db.close c)
 \end{lstlisting}

### 2019-07-06 kindly/lang.tex
Write the rest of the explanation for SD typing.

@@
 
-In the rest of this section, we present the syntax (\cref{syntax}),
-a linearity-aware semantics (\cref{sem}), syntax directed typing rules
-(\cref{sdtyping}) and automatic rules for region annotations (\cref{regionannot}).
+In the rest of this section, we present the syntax (\cref{syntax})
+the syntax directed typing rules (\cref{sdtyping})
+and a linearity-aware semantics (\cref{sem}) for \lang.
 
@@
 
-\TODO{Do we explain environments here, or push that for typing, later on ?}
-
 % Environments are composed of value and type bindings, along with type
@@
 
-We will now present the novel aspects of our system which are necessary to handle
-three specific features: binders, borrows, and sharing.
+\subsubsection{Kinds and constraints}
+
+One novel feature or \affe is the use of kinds and qualified types to encode
+linear and affine types. Before explaining the details of our type system, we
+give a quick presentation of the kinds and constraints.
+A kind $k$ can be either a kind variable, $\kvar$, or a constant.
+Each constant is composed of a ``quality'' and a ``level''.
+The quality describes the use pattern of the type and can be
+either unrestricted $\kun$, affine $\kaff$ or linear $\klin$. The level
+describes the nested regions in which it can be used. Levels are represented
+by a natural number of $\infty$.
+$0$ represents the normal scope outside of any region and is often omitted
+(for instance $\kaff_0$ is simply noted $\kaff$).
+$\infty$ is the empty region that is infinitely nested and is mostly used
+as an upper bound.
+For instance, the constraint $\Cleq{\kvar}{\kun_\infty}$ indicates that
+$\kvar$ must be unrestricted, but can be local to a region.
+%
+Just as presented before, constraints present in schemes and type signatures
+are simply a list of atomic constraints, which are composed
+of kind inequalities.
+%
+Our inequality constraints are based on an inequality relation on
+kind constants noted $\lk_\Lat$. This relation forms a lattice
+and  is summarized in
+\cref{sdtyp:lattice}. We have the basic inequalities on qualities:
+$\kun_n \lk_\Lat \kaff_n \lk_\Lat \klin_n$ for arbitrary regions $n$.
+Additionally, the inequality follows the region nesting, for
+instance $\kaff_n \lk_\Lat \kaff_{n+1}$.
+
+\begin{figure*}[!h]
+  \input{infer/lattice}
+  \caption{Lattice inequalities -- $k \lk_\Lat k'$}
+  \label{sdtyp:lattice}
+\end{figure*}
 
@@
 
-One main feature of our system is to control the usage of variables, either
+\lang aims to control the usage of variables, either
 through
-linearity and affinity, which controls how variables are shared or ignored,
-or through borrows, which allow to circumvent the linearity rules
+linearity and affinity which restrict how variables are shared and ignored,
+or through borrows which allow to circumvent the linearity rules
 in a controlled way.
@@
 bindings.
-Borrow bindings $\bvar{\borrow{x}}{\borrowty{k}{\tau}}$ which indicates
+Borrow bindings $\bvar{\borrow{x}}{\borrowty{k}{\tau}}$ indicates
 that the borrow $\borrow{x}$ can be used directly, as implemented
@@
 cannot be used immediately, but will be usable in a region
-marked with $x$ (as detailed in \cref{sdtyp:regions}).
+marked with $x$, which we detail in \cref{sdtyp:regions}.
 
-In \lang, many properties are encoded through the use of environment-wide
-constraints. Here we present how to control both unused variables and
+Environments are also used to encode several linearity-related properties
+through \emph{environment-wide constraints}.
+Here we present how to control both unused variables and
 closure captures through such environment constraints, noted $\Cleq{\E}{k}$.
@@
 $\bvar{x}{\tau}$ in $\E$,
-if type $\tau$ has kind $k'$ ($\inferSK{C}{\E}{\tau}{k'}$)
-then $\Cleq{k'}{k}$.
+if type $\tau$ has kind $k'$ (noted $\inferSK{C}{\E}{\tau}{k'}$)
+then the following constraints should hold: $\Cleq{k'}{k}$.
 While this definition is incomplete, it is sufficient to understand
@@
 ensuring they are not linear. This is achieved through the constraint
-$\Cleq{\E\Sdel{X}}{\kaff_\infty}$.
-%
-The {\sc Lam} constraints must ensures that if the body of the lambda captures
-linear or affine variable, the content will be reflected in the kind
+$\Cleq{\E\Sdel{x}}{\kaff_\infty}$ which replaces the traditional
+``weakening'' rule found in many type systems.
+
+The {\sc Abs} rule must ensures that if the body of the lambda captures
+linear or affine variables, the content will be reflected in the kind
 annotated on the arrow, such as $\tau_2 \tarr{k} \tau_1$.
 In order to do so, we constraint all the bindings of $\E$ to be less than $k$.
-If one of the binding of $\E$ constraints a type whose kind is affine,
-we will derive the constraints $\Cleq{\kaff_n}{k}$, which ensures that
+If one of the binding of $\E$ contains a type whose kind is affine,
+we will derive the constraints $\Cleq{\kaff}{k}$, which ensures that
 the kind on the arrow is at least affine.
+Note that $x$ is \emph{not} included in $\E$: since it's the argument of the
+closure, its use should not induce a capture.
 
@@
     \end{mathpar}
-    \caption{The {\sc Borrow} rule}
+    \caption{The {\sc Borrow}, {\sc Abs} and {\sc Var} rules}
     \label{selectrules:borrow}
+    \label{selectrules:binders}
   \end{minipage}
@@
 Linear and affine types also control the copy of variables.
-We now showcase how this is achieved in \lang through the {\sc Pair} typing rule
-in \cref{sdtyp:pair}.
-\TODO{Explain splitting}
+To demonstrate how this is achieve in \lang, we show the
+{\sc Pair} typing rule in \cref{sdtyp:pair}.
+In the {\sc Pair} rule, we split the type environment $\E$ in two parts,
+$\E_1$ and $\E_2$, which are used to typecheck the components of the pair.
+We note the splitting $\lsplit{C}{\E}{\E_1}{\E_2}$ where $C$ is a constraint
+that must be respected for the split to be valid.
+The main splitting rules are shown in \cref{sdtyp:split}. Each
+rule applies to a binding for a given variable. For instance, the
+{\sc Both} rule indicates that if we can prove that $x$ is unrestricted, we
+can duplicate it in both environments.
+Similarly, since immutable borrows are always unrestricted, the
+{\sc Borrow} rule allows to duplicate them arbitrarily.
+The {\sc Left} and {\sc Right} rule allow to move a binding
+to the left or right environment.
+The {\sc Susp} and {\sc SuspB} allow to split off suspended bindings to
+the left environment while conserving access to the binding on the right.
+Such suspended binding can later be turned
+into a borrow inside a region. The asymmetry here must always follow
+the order of execution of the language: in \affe, execution is from left to
+right, which means that a resource can be used as a borrow on the left
+and then as a full resource on the right, assuming that the borrow does not
+escape. The {\sc Susp} rule work with a full resource while the {\sc SuspB}
+rule allow to split off an immutable suspended binding based off a borrow.
+
+Using these various rules, we can distribute the bindings every time an
+expression has multiple sub-expressions such as pairs, function applications
+and let bindings. For instance, in the
+expression
+$\letin{a}{\text{create}\ 10\ x}
+{\introPair{\region[{}]{x}{\app{\text{length}}{\borrow[\IBORROW]{a}}}}{a}}$,
+we use the rule {\sc Susp} to detach a borrow on the left of the pair.
+Naturally, we must here ensure that the borrow on the left of the pair does not
+escape, which brings us to the typing rules for regions.
 
@@
     \caption{Splitting rules for bindings}
+    \label{sdtyp:split}
   \end{subfigure}
@@
 Such borrows are constrained to a given syntactic region. We note
-$\region{x}{e}$ for the region allowing borrows of $x$ to be used in $e$
+$\region{x}{e}$ the region which allow borrows of $x$ to be used in $e$
 and where $n$ is the nesting of the region.
@@
 
-\TODO{Transition}
-
-In order to use a suspended binding for the variable $x$,
-we need to enter a \emph{region} associated with $x$,
-noted $\region{x}{e}$.
 The main idea of region typing (implemented by the {\sc Region} rule)
@@
 This is done by the borrow relation, noted $\lregion{C}{x}{\E}{\E'}$ and
-whose main rules are shown in \cref{env:rule:region}.
+whose main rules are shown in \cref{selectrules:region}.
 In our case, we consider an immutable binding, so we replace the suspended
@@
 
-\subsubsection{Constraints}
-
 \clearpage

### 2019-07-06 kindly/lang.tex
Tiny fixes.

@@
 We must also check that $\entail{C}{(\kun_n\lk k\lk\kun_\infty)}$.
-We can then typecheck the body of the region, $f(\borrow[\IBORROW]{c})$ for our example and check that the borrow made available in the region does not escape.
+We can then typecheck the body of the region, $f(\borrow[\IBORROW]{c})$ in our example, and check that the borrow made available in the region does not escape.
 This is done by the constraint $\Cleq{\tau}{\klin_{n-1}}$, which ensures
@@
 
+\subsubsection{Kind checking}
+
+\TODO{Present this if we have enough space}
+
+\subsubsection{Pattern matching}
+
+\TODO{Present this if we have enough space}
+
 \clearpage

### 2019-07-06 kindly/infer/infer.tex
Start working on the inference section.

@@
-\section{Type inference}
+\section{Inference}
 
-We now formulate type inference for the \lang language. Type inference is
+We now formulate type inference for \lang. Type inference is
 based on the HM(X) framework~\citep{DBLP:journals/tapos/OderskySW99} which
@@
 with qualified types and an arbitrary custom constraint language.
-We first present various preliminaries definitions,
-we then show our extension of HM(X) that support a kind system and affine
-types.
+We adapt and extend the rules of HM(X) to support kind inference,
+track linearity and affinity and handle borrow and regions. Furthermore, we
+formulate an appropriate constraint solving algorithm for our constraint
+system such that inference is sound and complete.
+In the rest of this section, we first present various preliminaries definitions
+before showing our kind and type inference judgements
+and our constraint solving algorithm.
 
@@
 
-\subsubsection{Usage maps}
+Unlike the syntax-directed version of our type system, knowing which elements
+are input and output or our inference judgement is critical. In the rest
+of this section, when presenting a new judgement
+we will note in \textbf{bold} the input parameters. The rest will then be
+output parameters.
+
+\subsubsection{Usage environments}
+
+% One novelty of our inference  judgement now
+% returns a type environment, which we call ``usage environment'' and commonly
+% note $\Sv$, to summarize how variables and borrows are used inside
+% the expression.
 
@@
 $(x,x)$, $x$ is used twice. If $x$ is of type $\tau$, which is of kind $k$,
-we must add the constraint $\Cleq{k}{\kun}$. In order to record uses and kinds,
-our type inference relation will return a ``usage map'', noted $\Sv$,
-which associates each variable to a list of kinds corresponding to each use site
-of the variable.
-We note $\Sone{x}{k_1,\dots,k_n}$ the singleton usage map.
-Common operations on usage maps are defined in \cref{op:usgmap}. We define
-regular set operations such as union
-($\Sv\cup\Sv'$) and intersection ($\Sv\cap\Sv'$), along
-with constraint creation $\Cleq{\Sv}{k_0}$ which indicates that all the kinds
-presents in the usage map $\Sv$ are constrained by the kind $k_0$.
-These operations corresponds roughly to the various parts of the join
-operator on sets $\esplit{}{}$.
+we must add the constraint $\Cleq{k}{\kun}$.
+%
+In order to infer such constraints, our inference judgement will not only
+take an environment as parameter but also return an environment, which
+we call ``usage environment'', and which summarize how variables and borrows
+are used. Usage environments follow the exact same grammar
+as normal environment. In order to distinguish them more easily,
+we will general note them $\Sv$.
 
-\begin{figure}[tp]\centering
+In \cref{sdtyping}, we used various relations to split environments in two, or
+transform suspended bindings into borrow binding inside an environment.
+These relations took as argument a constraint which is used to validate
+the environment transformation.
+In the context of inference, we define a new judgements which \emph{infers}
+the constraints required.
+\begin{itemize}
+\item $\bsplit{C}{\Sv}{\bf{\Sv_1}}{\bf{\Sv_2}}$.
+  Given two usage environments (inferred for two subexpressions, for instance)
+  $\Sv_1$ and $\Sv_2$, we return $\Sv$, the merged environment, and $C$, a set
+  of constraints that must be respected.
+  This relation is total and non-ambiguous given $\Sv_1$ and $\Sv_2$
+  and uses the same rules as the one presented in \cref{sdtyping:split}.
+\item $\bregion{C}{\bf{x}}{\Sv}{\bf\Sv'}$.
+  Given a usage environment $\Sv'$ and a variable name $x$, we return
+  $\Sv$ where the borrow binding of $x$ in $\Sv'$, if it exists, is replaced by
+  a suspended binding. We also return $C$, a set of constraints that must
+  be respected.
+  Again, the relation is total and non-ambiguous for any given $\Sv'$ and $x$,
+  and uses the same rules as the one presented in \cref{sdtyping:regions}.
+\end{itemize}
+
+The relations used for syntax-directed typing can trivially be defined
+in term of these new relations and constraint entailment.
+All relations are fully described in \cref{typ:extra:envs}.
+
+\subsubsection{Constraint normalization}
+
+The HM(X) framework assumes the existence of a $\operatorname{normalize}$
+function which takes a constraint $C$ and a substitution $\psi$ and returns a
+constraint in solved form $C'$,
+and an updated substitution $\unif'$.
+Normalization is supposed to return the ``best'' solution $(C',\unif')$, for
+instance the most general unifier.
+We detail the implementation
+of the normalization function in \cref{infer:solving}. In the meantime, we just
+assume the existence of such a function for our constraint system.
+
+\subsection{Kind inference}
+
+We note $\inferK{(C,\unif)}{\bf{\E}}{\bf{\tau}}{k}$ when type $\tau$ has kind $k$
+in environment $\E$ under constraints $C$ and unifier $\unif$.
+$\E$ and $\tau$ are the input parameters of
+our inference procedure.
+We present the kind inference algorithm as a set of rules in
+\cref{rules:kinding}.
+Similar to the syntax-directed rules, higher-kinds are not generally supported
+and can only appear by looking-up the kind scheme of a type constructor,
+for use in the type application rule {\sc KApp}.
+Type variables must be of a simple kind in rule {\sc KVar}.
+Kind schemes are instantiated in the {\sc KVar} rules by creating
+fresh kind variables and the associated substitution.
+{\sc KArr} and {\sc KBorrow} simply returns the kind of the primitive
+arrow and borrow types.
+The $\operatorname{normalize}$ function is used every time several constraints
+must be composed in order to simplify the constraint and return a most general
+unifier.
+
+\begin{figure}[ht]
+  \centering
+  \input{infer/kinds}
+  \caption{Kind inference algorithm -- $\inferK{(C,\unif)}{\E}{\tau}{k}$}
+  \label{rules:kinding}
+\end{figure}
+
+\subsection{Type inference}
+
+We note $\inferW{\Sv}{(C,\unif)}{\bf{\E}}{\bf{e}}{\tau}$ when
+$e$\ has type $\tau$ in $\E$ under the constraints $C$ and unifier $\unif$
+with a usage environment $\Sv$. $\E$ and $e$ are the input parameters of our
+inference algorithm.
+The rules are shown in \cref{rules:typing}.
+
+\TODO{Explain/Cut}
+
+\begin{figure*}[htp]
+  \input{infer/typing}
+  \caption{Type inference algorithm -- $\inferW{\Sigma}{(C,\psi)}{\bf{\E}}{\bf{e}}{\tau}$ }
+  \label{rules:typing}
+\end{figure*}
+
+\begin{figure}[tp]
+  \centering
   \begin{mathpar}
-  % \Sv\cap\Sv'
-  % &= \left\{ x \to l,l'
-  %   \mid \Sone{x}{l}\in\Sv \text{ and } \Sone{x}{l'}\in\Sv' \right\} \\
-  % \infer
-  % { \forall x \in \Dom{\Sv_1}\cup\Dom{\Sv_2}.\quad
-  %   \bsplit{C_x}{\Sv(x)}{\Sv_1(x)}{\Sv_2(x)}
-  % } 
-  % { \bsplit{(\bigwedge C_x)}{\Sv}{\Sv_1}{\Sv_2} }
-  % &= \Sv\cap\Sv' \sqcup
-  %   (\Sv\Sdel{x' \mid x'\in\Sv'}) \sqcup
-  %   (\Sv'\Sdel{x \mid x\in\Sv})
-  %   &\text{where $\sqcup$ is the disjoint union}\\
-  % \and
-  % \inferrule{}
-  % { \Cleq{\Sv}{k}
-  %   = \bigwedge_{\Sone{x}{\schm}\in\Sv} \Cleq{\schm}{k}
-  % }
-  % \and
-  % \inferrule
-  % { \forall \bshadow{x} \in \Sv, b'\in\Sv'.\quad
-  %   \tregion{C_x}{y}{\bshadow{x}}{b'} }
-  % { \bregion{(\bigwedge C_x)}{y}{\Sv}{\Sv'} }
-  % \and
-  % \inferrule
-  % { \text{if } \operatorname{lenght}(\Sv(x))  = \Cempty{} \wedge  }
     { \Weaken_{\bvar{x}{\sigma}}(\Sv) =
@@
       \end{dcases}
-
     }
@@
 
-\subsection{Kind inference}
+\subsection{Constraint solving}
+\label{infer:normalize}
 
-We note $\inferK{(C,\unif)}{\bf{\E}}{\bf{\tau}}{k}$ when type $\tau$ has kind $k$
-in environment $\E$ under constraints $C$ and unifier $\unif$. From an
-algorithmic point of view, $\E$ and $\tau$ are the input parameters of
-our inference procedure.
-We present the kind inference algorithm as a set of syntax-directed rules in
-\cref{rules:kinding}.
-Similar to the syntax-directed rules, higher-kinds are not generally supported
-and can only appear by looking-up the kind scheme of a type constructor,
-for use in the type application rule {\sc KApp}.
-Type variables must be of a simple kind in rule {\sc KVar}.
-Kind schemes are instantiated in the {\sc KVar} rules by creating
-fresh kind variables and the associated substitution.
-{\sc KArr} and {\sc KBorrow} simply returns the kind of the primitive
-arrow and borrow types.
-
-\begin{figure}[ht]
-  \centering
-  \input{infer/kinds}
-  \caption{Kind inference algorithm -- $\inferK{(C,\unif)}{\E}{\tau}{(\Multi{k}) k}$}
-  \label{rules:kinding}
-\end{figure}
-
-\subsection{Type inference}
-
-We reformulate the HM(X) type inference in the context of our affine type
-system. The main difference compared to HM(X) are noted in \addlin{blue}.
-We note $\inferW{\addlin{\Sigma}}{(C,\unif)}{\bf{\E}}{\bf{e}}{\tau}$ when
-$e$\ has type $\tau$ in $\E$ under the constraints $C$ and unifier $\unif$.
-$\Sigma$ is a map which associates free variables in $e$ to
-their kinds.
-As before, $\E$ and $e$ are the input parameters of the inference
-algorithm. The syntax-directed rules are shown in \cref{rules:typing}.
-
-\begin{figure*}[htp]
-  \input{infer/typing}
-  \caption{Type inference algorithm -- $\inferW{\Sigma}{(C,\psi)}{\bf{\E}}{\bf{e}}{\tau}$ }
-  \label{rules:typing}
-\end{figure*}
-
-\subsection{Normal forms and generalization}
-\label{sec:normalize}
+\TODO{Present the full constraint system}
 
@@
 
-\subsubsection{Soundness and Principality}
+\subsection{Soundness and Principality}
+
+\TODO{Simplify/Cut}
 
@@
 
-\subsection{Principality}
-
 \TODO{}

### 2019-07-06 kindly/lang.tex
Start working on the inference section.
Fix missing refs.

@@
 %
-We first define the following typing, kinding and constraint judgements:
+We introduce the following typing, kinding and constraint judgements:
 \begin{description}
@@
 \subsubsection{Environments and bindings}
+\label{sdtyping:envs}
 
@@
 cannot be used immediately, but will be usable in a region
-marked with $x$, which we detail in \cref{sdtyp:regions}.
+marked with $x$, which we detail in \cref{sdtyping:regions}.
 
@@
 \subsubsection{Copying and Splitting}
+\label{sdtyping:split}
 
@@
 \subsubsection{Regions}
-\label{sdtyp:regions}
+\label{sdtyping:regions}
 

### 2019-07-06 kindly/intro.tex
more motivation
motivation: hybrid arrays
motivated
Fix missing refs.
intro

@@
 different set of API calls is enabled afterwards. 
+Ignoring such life cycle protocols is a common source of errors.
+%% this is about typestate \cite{DBLP:conf/ecoop/BeckmanKA11}.
 
-Ignoring this protocol is a common source of errors \cite{is there
-  some paper?}.
+Most type systems provide just type soundness, but neglect the
+protocol aspect. The systems that do build on linear types \cite{Girard} and/or
+uniqueness types\cite{DBLP:conf/plilp/BarendsenS95}. A value of linear
+type is guaranteed to be consumed 
+exactly once. That is, a file that has been opened must be closed and
+memory that has been allocated must be released. A value of unique
+type is guaranteed to have a single reference to it.
 
-Most statically typed languages do not address this issue.
+These systems work well if one is prepared to write programs
+functionally in resource-passing style. In this style, all operations
+in the resource's API take the resource as a parameter and return it
+in possibly modified state\cite{DBLP:journals/jfp/AchtenP95}. In
+typestate-oriented programming, they would also modify its type
+\cite{DBLP:conf/oopsla/AldrichSSS09}. Functional session types
+represent a popular example \cite{DBLP:journals/jfp/GayV10}.
 
-But linear types do.
+Explicit resource passing places a heavy burden on the programmer and
+complicates the program structure. For imperative APIs,
+resource-passing style is not an option at all. To this end,
+\citet{DBLP:conf/popl/BoylandR05}  proposed the notion of
+\emph{borrowing} a resource. The idea is that a linear resource can be
+borrowed to a function call, which can work with the resource, but
+cannot release it. Only the original owner of the resource has all
+rights to it and can release it.
 
-But typestate does it. A popular example is session types.
+The concepts of ownership and borrowing have grown popular over time
+and they form the foundation of the type system of the Rust language
+\cite{rust}. Rust supports two kinds of borrows of data structures, mutable and immutable
+ones. Mutable borrows enable modification of the data structure
+whereas immutable borrows come only with read access. Moreover, Rust
+manages the lifetime of data and makes sure that the lifetime of a
+borrow is properly contained in the lifetime of its lender.
 
-Typestate needs to be supported by affinity (or linearity, or
-uniqueness).
+The design of Rust is geared towards programmers with a low-level
+imperative programming background, like C or C++. Its management of
+lifetime supports the manual way of memory management customary in
+these languages very well and makes it safe. However, programmers from
+a managed language background feel alienated from the lack of garbage
+collected data. They would prefer a setting where automatic memory
+management with garbage collection was the default, but where they
+could seemlessly switch to safe, manual resource management if that
+was required.
 
-But these approaches are restrictive in pratice because [here is an
-example that doesn't work well with existing systems].
+As a concrete example, consider a functional programmer who wants to
+safely interact with a C library. Invoking the C functions is easy via
+the existing foreign function interface, but managing the underlying
+resources like malloc'd storage is not: It cannot be left to the
+garbage collector, but the proper calls to \texttt{free()} must be
+ensured by programming conventions.
+
+Our work provides a solution to programmers in this situation. We
+propose an extended type system for ML-like languages that comes with
+linear and affine types, mutable and immutable borrows, but all that
+integrated with full type inference, garbage collected data, and
+ANOTHER FEATURE.
+
+\subsection{A first example}
+\label{sec:first-example}
+\begin{figure}[tp]
+  \begin{subfigure}[t]{0.45\linewidth}
+    \lstinputlisting{code/writefiles.affe}
+    \caption{File API}
+    \label{fig:writing-files-api}
+  \end{subfigure}
+  ~
+  \begin{subfigure}[t]{0.45\linewidth}
+\begin{lstlisting}
+let main () =
+  let h = File.fopen "foo" in
+  write &!h "Hello ";
+  write &!h "world!";
+  File.close h
+\end{lstlisting}
+    \caption{File example}
+    \label{fig:writing-files-example}
+  \end{subfigure}
+  \caption{Writing files}
+  \label{fig:writing-fiules}
+\end{figure}
+
+As a first, well-trodden example we consider a very simple API for
+writing files in \cref{fig:writing-files-api}.
+The type \lstinline/File.t/ is linear. A call like
+\lstinline/fopen "foo"/ returns a linear handle to a newly created
+file, which \emph{must} be closed later using
+\lstinline/close/. Failing to do so is a static type error.
+To write to the file, we must take (mutable) borrows of the handle and
+pass them to the \lstinline/write/
+function. \Cref{fig:writing-files-example} contains an example where
+\lstinline/&!h/ stands for the mutable borrow.
+
+\subsection{Contributions}
+\label{sec:contributions}
+
+\begin{itemize}
+\item Type system that encodes linearity and affinity with borrowing
+\item Full type inference for \dots
+\item Automatic inference of regions for borrows
+\end{itemize}
 

### 2019-07-06 kindly/intro.tex
aff fixed

@@
 \lstinline/close/. Failing to do so is a static type error.
-To write to the file, we must take (mutable) borrows of the handle and
-pass them to the \lstinline/write/
-function. \Cref{fig:writing-files-example} contains an example where
+To write to the file, we must take a (mutable) borrow of the handle and
+pass it to the \lstinline/write/
+function. As mutable borrows must not be duplicated so they are
+classified as affine. This affinity shows up in the annotation of the
+second arrow in the type of \lstinline/write/: a partial application
+\lstinline/write &!h/ captures the affine borrow and hence the
+resulting function must also be handled in an affine way.
+\Cref{fig:writing-files-example} contains an example where
 \lstinline/&!h/ stands for the mutable borrow.
+
+This example demostrates two features of our system.
+\begin{enumerate}
+\item full type inference without annotations in user code
+\item types carry multiplicity annotations in the form of kinds
+\end{enumerate}
+
+More features: kinds can be polymorphic and constrained.
 

### 2019-07-07 kindly/intro.tex
intro-examples

--- 2019-07-06 kindly/intro.tex
@@
 in the resource's API take the resource as a parameter and return it
-in possibly modified state\cite{DBLP:journals/jfp/AchtenP95}. In
-typestate-oriented programming, they would also modify its type
-\cite{DBLP:conf/oopsla/AldrichSSS09}. Functional session types
-represent a popular example \cite{DBLP:journals/jfp/GayV10}.
+in possibly modified state~\cite{DBLP:journals/jfp/AchtenP95}. In
+typestate-oriented programming, they would also modify its
+type~\cite{DBLP:conf/oopsla/AldrichSSS09}. Functional session types 
+represent a popular example~\cite{DBLP:journals/jfp/GayV10}.
 
@@
 ones. Mutable borrows enable modification of the data structure
-whereas immutable borrows come only with read access. Moreover, Rust
-manages the lifetime of data and makes sure that the lifetime of a
+whereas immutable borrows grant read-only access. Mutable borrows are
+exclusive whereas several immutable borrows can be active at the same
+time. Moreover, Rust makes sure that the lifetime of a
 borrow is properly contained in the lifetime of its lender.
@@
 imperative programming background, like C or C++. Its management of
-lifetime supports the manual way of memory management customary in
-these languages very well and makes it safe. However, programmers from
-a managed language background feel alienated from the lack of garbage
+lifetimes supports the manual way of memory management customary in
+these languages very well and makes it safe. However, programmers with
+a background in managed languages feel alienated from the lack of garbage
 collected data. They would prefer a setting where automatic memory
@@
 
+Compared to the most closely related contender in this design space, Linear Haskell
+\cite{DBLP:journals/pacmpl/BernardyBNJS18}, their goals and means are
+similar as both systems permit abstraction over the number of uses of
+values and retain type inference, but the details are quite different.
+\begin{enumerate}
+\item Their multiplicities are either linear or unrestricted whereas
+  we also permit affine values.
+\item Our system supports mutable and immutable borrows, hence the
+  need for affinity.
+\item Our multiplicities are directly attached to the type of the
+  value. For example, in our system the function type \lstinline/'a-{lin}>'b/
+  denotes the type of a \emph{single-use function} that can be called
+  just once, whereas the multiplicities in Linear Haskell choose
+  between $\alpha\to\beta$ and $\alpha \multimap\beta$ where the
+  latter is a function that promises to \emph{use its argument exactly once}.
+\end{enumerate}
 \subsection{A first example}
@@
   let h = File.fopen "foo" in
-  write &!h "Hello ";
-  write &!h "world!";
+  File.write &!h "Hello ";
+  File.write &!h "world!";
   File.close h
@@
 
-As a first, well-trodden example we consider a very simple API for
-writing files in \cref{fig:writing-files-api}.
-The type \lstinline/File.t/ is linear. A call like
+As a first, well-known example we consider a simplified API for
+writing files shown in \cref{fig:writing-files-api}.  It introduces a
+linear abstract type \lstinline/File.t/. A call like
 \lstinline/fopen "foo"/ returns a linear handle to a newly created
-file, which \emph{must} be closed later using
-\lstinline/close/. Failing to do so is a static type error.
-To write to the file, we must take (mutable) borrows of the handle and
-pass them to the \lstinline/write/
-function. \Cref{fig:writing-files-example} contains an example where
-\lstinline/&!h/ stands for the mutable borrow.
+file, which \emph{must} be released later on through \lstinline/close/
+as shown in \cref{fig:writing-files-example}. Failing to do so is a
+static type error.  To write to the file, we must take a (mutable)
+borrow \lstinline/&!h/ of the handle and pass it to the
+\lstinline/write/ function. Mutable borrows are classified as affine:
+they must not be duplicated, but there is no requirement to use
+them. This affinity shows up in the annotation of the second arrow in
+the type of \lstinline/write/: a partial application like
+\lstinline/write &!h/ captures the affine borrow and hence the
+resulting function must also be handled in an affine way. That is, it
+would be an error to write
+\begin{lstlisting}
+  let w = File.write &!h in w "Hello "; w "world!" (* type error *)
+\end{lstlisting}
+as we would be using the affine closure twice.
+
+This example demonstrates two features of our system.
+\begin{enumerate}
+\item full type inference without annotations in user code
+\item types carry multiplicity annotations in the form of kinds
+\end{enumerate}
+
+In general, kinds can be polymorphic and constrained. Function
+application and composition are the archetypical
+examples for functions exploiting that feature.
+\begin{lstlisting}
+let app f x = f x
+# app : ('a -{'k}> 'b) -> ('a -{'k}> 'b)
+\end{lstlisting}
+The reading of the type inferred by our system is straightforward. If
+\lstinline/f/ is a \lstinline/'k/-restricted function, then so is
+\lstinline/app f/. The multiplicities of \lstinline/'a/ and
+\lstinline/'b/ play no role. As usual in ML-like languages, we
+implicitly assume prenex quantification by
+$\forall\kappa\forall\alpha\forall\beta$. In reality, the
+type checker also quantifies over the kinds of $\alpha$ and $\beta$,
+hence the full truth is
+$\forall\kappa\kappa_1\kappa_2\forall(\alpha:\kappa_1)\forall(\beta:\kappa_2)$,
+which is only revealed if necessary for understanding the type. 
+\begin{lstlisting}
+let compose f g x = f (g x) 
+# compose : ('k <= 'k_1) => ('b -{'k}> 'c) -> ('a -{'k_1}> 'b) -{'k}> ('a -{'k_1}> 'c)
+\end{lstlisting}
+Like in \lstinline/app/, the multiplicities of the type variables
+\lstinline/'a,'b,'c/ do not matter. However, the multiplicity
+\lstinline/'k/ of \lstinline/f/ reappears on the second to last arrow
+because \lstinline/compose f/ is a closure that inherits
+\lstinline/f/'s multiplicity. The multiplicities of \lstinline/g/  and
+\lstinline/f/ both influence the multiplicity of the last arrow, so
+that we would expect its annotation to be the least upper bound
+$\kappa \sqcup \kappa_1$. Thanks to subsumption of multiplicities, it
+is sufficient to assume $\kappa \le \kappa_1$ and \lstinline/g/'s
+actual multiplicity gets subsumed to $\kappa_1$. This constraint
+simplification technique is built into our type inference algorithm. Again, the printout
+of the type scheme only mentions the required constraint
+$\kappa\le\kappa_1$ and omits the prenex quantification over $\kappa,
+\kappa_1$ as well as the kinds of \lstinline/'a,'b,'c/.
 
@@
 \item Full type inference for \dots
+\item Constraint simplification
 \item Automatic inference of regions for borrows

### 2019-07-07 kindly/infer/infer.tex
Fix missing refs.
Small fixes.

@@
 
-We now formulate type inference for \lang. Type inference is
-based on the HM(X) framework~\citep{DBLP:journals/tapos/OderskySW99} which
-presents how to implement Hindley-Milner type inference in a language
+The main innovation of \affe is its principal type inference.
+We now formulate our inference algorithm
+based on the HM(X) framework~\citep{DBLP:journals/tapos/OderskySW99}.
+HM(X) presents how to implement Hindley-Milner type inference in a language
 with qualified types and an arbitrary custom constraint language.
 We adapt and extend the rules of HM(X) to support kind inference,
-track linearity and affinity and handle borrow and regions. Furthermore, we
+track linearity and handle borrows and regions. Furthermore, we
 formulate an appropriate constraint solving algorithm for our constraint
 system such that inference is sound and complete.
-In the rest of this section, we first present various preliminaries definitions
+We first present various preliminary definitions
 before showing our kind and type inference judgements
@@
 function which takes a constraint $C$ and a substitution $\psi$ and returns a
-constraint in solved form $C'$,
+simplified constraint $C'$,
 and an updated substitution $\unif'$.
@@
 We detail the implementation
-of the normalization function in \cref{infer:solving}. In the meantime, we just
+of the normalization function and its property in \cref{infer:solving}.
+In the meantime, we simply
 assume the existence of such a function for our constraint system.
@@
 \subsection{Constraint solving}
-\label{infer:normalize}
+\label{infer:solving}
 

### 2019-07-07 kindly/motivation.tex
Typo fixes in the motivation.
Small fixes.

@@
 whether \lstinline{'a} is restricted to be linear, in fact
-\lstinline{receive} is polymorphic in the kind of \lstinline{'a} and
-also in the kind of \lstinline{->}. This kind polymorphism is the
+\lstinline{receive} is polymorphic in the kind of \lstinline{'a}.
+This kind polymorphism is the
 default if no constraints are specified.
@@
 
-  val get : ('a : un) => &('a t) * int -> 'a
-  val set : ('a : aff) => &!('a t) * int * 'a -> unit
+  val get :
+    ('a : un) => &('a t) * int -> 'a
+  val set :
+    ('a : aff) => &!('a t) * int * 'a -> unit
 
@@
 \lstinline{&a} in \cref{ex:array:region:get} has kind
-\lstinline{unr_2} and the typing of its region is such that types with
+\lstinline{un_2} and the typing of its region is such that types with
 kind indexes greater than or equal to $2$ cannot escape.
@@
 modifications. The function
-\lstinline/create_from : int -> (int -> 'a) -> 'a Array.t /
-initializes an array from a function.
+\lstinline/mapi : (int -> &'a -> 'b) -> &('a t) -> 'b t /
+is a simple variation on \lstinline/Array.map/ where the mapping function
+takes the position of the element.
 % A more elaborate implementation could use persistent arrays
@@
 
-  let set (a, i, x) =
-    let f k =
-      if k = i then x else get (a,k)
+  let set (a, i0, x0) =
+    let f i x =
+      if i = i0 then x0 else x
     in
-    create_from (length a) f
+    Array.mapi f a
 
@@
 The Sudoku solver iterates over the cells and tries each possible solution (\cref{line:try_solution}).
-When a value is picked for the current cell, we create a choicepoint,
+When a value is picked for the current cell, we create a choice point,
 change the cell with an immutable modification (\cref{line:immutset}), and propagate
@@
 many times as needed.
+This would also easily allow parallel iteration on \cref{line:iter}.
 As \lstinline/g/ is linear we must free it in \cref{line:free:g}.
@@
 
-The implementation in \cref{impl:pool} represents a bag of resourfces
+The implementation in \cref{impl:pool} represents a bag of resources
 using a concurrent queue with atomic add and remove operations. 
@@
 it to the pool afterwards.
-We explicitly delimit the region in \cref{line:pool:region} to ensure that
+For clarity,
+we explicitly delimit the region in \cref{line:pool:region} to ensure that
 the return value of \lstinline/f &!o/ does not capture \lstinline/&!o/.
+In practice, such annotation is unnecessary, as it would be automatically
+annotated.
 
@@
   \centering
-  \begin{subfigure}[t]{.55\linewidth}
+  \begin{subfigure}[t]{.58\linewidth}
 \begin{lstlisting}
@@
   (* Create a database connection pool. *)
-  let pool = Pool.create (fun _ -> Db.connect uri)
-  List.parallel_iter (create_user &pool) users
+  let pool = Pool.create (fun _ -> Db.connect uri) in
+  List.parallel_iter (create_user &pool) users;
   Pool.consume (fun c -> Db.close c)
@@
   \end{subfigure}~
-  \begin{subfigure}[t]{.45\linewidth}
+  \begin{subfigure}[t]{.43\linewidth}
 \begin{lstlisting}
@@
   { spawn ; queue = CQueue.create () }
-let consume f c = CQueue.iter f c.queue
+let consume f c =
+  CQueue.iter f c.queue
 

### 2019-07-07 kindly/intro.tex
intro&motivation

@@
 \end{enumerate}
-\subsection{A first example}
+\subsection{First examples}
 \label{sec:first-example}
@@
 they must not be duplicated, but there is no requirement to use
-them. This affinity shows up in the annotation of the second arrow in
+them. This affinity shows up in the annotation \lstinline/-{aff}>/ of the second arrow in
 the type of \lstinline/write/: a partial application like
@@
 \end{lstlisting}
-as we would be using the affine closure twice.
+as we would be using the affine closure twice. The remaining arrows in
+the API are unrestricted and we write \lstinline/->/ instead of the
+explicitly annotated \lstinline/-{un}>/.
 
-This example demonstrates two features of our system.
+This example demonstrates two features of our system:
 \begin{enumerate}
-\item full type inference without annotations in user code
-\item types carry multiplicity annotations in the form of kinds
+\item full type inference without annotations in user code,
+\item types carry multiplicity annotations in the form of kinds.
 \end{enumerate}
@@
 type checker also quantifies over the kinds of $\alpha$ and $\beta$,
-hence the full truth is
+hence the full prefix is
 $\forall\kappa\kappa_1\kappa_2\forall(\alpha:\kappa_1)\forall(\beta:\kappa_2)$,
-which is only revealed if necessary for understanding the type. 
+which is only revealed as much as necessary for understanding the type. 
 \begin{lstlisting}

### 2019-07-07 kindly/motivation.tex
intro&motivation

@@
 In this context, channels must be used linearly, because every use of
-a channel ``consumes'' one interaction and hence changes the type of the
+a channel ``consumes'' one interaction and  changes the type of the
 channel. That is, after sending two integers on the channel,
@@
 the return value of \lstinline/f &!o/ does not capture \lstinline/&!o/.
-In practice, such annotation is unnecessary, as it would be automatically
-annotated.
+In practice, the type checker inserts this region automatically.
 

### 2019-07-07 kindly/infer/extra.tex
Select the various part of the type inference section.

--- 2019-07-03 kindly/infer/infer.tex
@@
-\section{Type inference}
+\section{Inference rules}
+\label{appendix:infer}
 
-We now formulate type inference for the \lang language. Type inference is
-based on the HM(X) framework~\citep{DBLP:journals/tapos/OderskySW99} which
-presents how to implement Hindley-Milner type inference in a language
-with qualified types and an arbitrary custom constraint language.
-We first present various preliminaries definitions,
-we then show our extension of HM(X) that support a kind system and affine
-types.
+\begin{figure*}[htp]
+  \input{infer/typing}
+  \caption{Type inference algorithm -- $\inferW{\Sigma}{(C,\psi)}{\bf{\E}}{\bf{e}}{\tau}$ }
+  \label{rules:typing}
+\end{figure*}
 
-\subsection{Preliminaries}
-
-\subsubsection{Usage maps}
-
-In order to determine if a variable is used in an affine manner, we must track
-its uses and the associated kinds. For instance, in the expression
-$(x,x)$, $x$ is used twice. If $x$ is of type $\tau$, which is of kind $k$,
-we must add the constraint $\Cleq{k}{\kun}$. In order to record uses and kinds,
-our type inference relation will return a ``usage map'', noted $\Sv$,
-which associates each variable to a list of kinds corresponding to each use site
-of the variable.
-We note $\Sone{x}{k_1,\dots,k_n}$ the singleton usage map.
-Common operations on usage maps are defined in \cref{op:usgmap}. We define
-regular set operations such as union
-($\Sv\cup\Sv'$) and intersection ($\Sv\cap\Sv'$), along
-with constraint creation $\Cleq{\Sv}{k_0}$ which indicates that all the kinds
-presents in the usage map $\Sv$ are constrained by the kind $k_0$.
-These operations corresponds roughly to the various parts of the join
-operator on sets $\esplit{}{}$.
-
-\begin{figure}[tp]\centering
+\begin{figure}[tp]
+  \centering
   \begin{mathpar}
-  % \Sv\cap\Sv'
-  % &= \left\{ x \to l,l'
-  %   \mid \Sone{x}{l}\in\Sv \text{ and } \Sone{x}{l'}\in\Sv' \right\} \\
-  % \infer
-  % { \forall x \in \Dom{\Sv_1}\cup\Dom{\Sv_2}.\quad
-  %   \bsplit{C_x}{\Sv(x)}{\Sv_1(x)}{\Sv_2(x)}
-  % } 
-  % { \bsplit{(\bigwedge C_x)}{\Sv}{\Sv_1}{\Sv_2} }
-  % &= \Sv\cap\Sv' \sqcup
-  %   (\Sv\Sdel{x' \mid x'\in\Sv'}) \sqcup
-  %   (\Sv'\Sdel{x \mid x\in\Sv})
-  %   &\text{where $\sqcup$ is the disjoint union}\\
-  % \and
-  % \inferrule{}
-  % { \Cleq{\Sv}{k}
-  %   = \bigwedge_{\Sone{x}{\schm}\in\Sv} \Cleq{\schm}{k}
-  % }
-  % \and
-  % \inferrule
-  % { \forall \bshadow{x} \in \Sv, b'\in\Sv'.\quad
-  %   \tregion{C_x}{y}{\bshadow{x}}{b'} }
-  % { \bregion{(\bigwedge C_x)}{y}{\Sv}{\Sv'} }
-  % \and
-  % \inferrule
-  % { \text{if } \operatorname{lenght}(\Sv(x))  = \Cempty{} \wedge  }
     { \Weaken_{\bvar{x}{\sigma}}(\Sv) =
@@
 
-\subsection{Kind inference}
-
-We note $\inferK{(C,\unif)}{\bf{\E}}{\bf{\tau}}{k}$ when type $\tau$ has kind $k$
-in environment $\E$ under constraints $C$ and unifier $\unif$. From an
-algorithmic point of view, $\E$ and $\tau$ are the input parameters of
-our inference procedure.
-We present the kind inference algorithm as a set of syntax-directed rules in
-\cref{rules:kinding}.
-Similar to the syntax-directed rules, higher-kinds are not generally supported
-and can only appear by looking-up the kind scheme of a type constructor,
-for use in the type application rule {\sc KApp}.
-Type variables must be of a simple kind in rule {\sc KVar}.
-Kind schemes are instantiated in the {\sc KVar} rules by creating
-fresh kind variables and the associated substitution.
-{\sc KArr} and {\sc KBorrow} simply returns the kind of the primitive
-arrow and borrow types.
-
-\begin{figure}[ht]
-  \centering
-  \input{infer/kinds}
-  \caption{Kind inference algorithm -- $\inferK{(C,\unif)}{\E}{\tau}{(\Multi{k}) k}$}
-  \label{rules:kinding}
-\end{figure}
-
-\subsection{Type inference}
-
-We reformulate the HM(X) type inference in the context of our affine type
-system. The main difference compared to HM(X) are noted in \addlin{blue}.
-We note $\inferW{\addlin{\Sigma}}{(C,\unif)}{\bf{\E}}{\bf{e}}{\tau}$ when
-$e$\ has type $\tau$ in $\E$ under the constraints $C$ and unifier $\unif$.
-$\Sigma$ is a map which associates free variables in $e$ to
-their kinds.
-As before, $\E$ and $e$ are the input parameters of the inference
-algorithm. The syntax-directed rules are shown in \cref{rules:typing}.
-
-\begin{figure*}[htp]
-  \input{infer/typing}
-  \caption{Type inference algorithm -- $\inferW{\Sigma}{(C,\psi)}{\bf{\E}}{\bf{e}}{\tau}$ }
-  \label{rules:typing}
-\end{figure*}
-
-\subsection{Normal forms and generalization}
-\label{sec:normalize}
-
-We now place ourselves in a more general context where kinds are either variables
-or constants belonging to a total bounded lattice $(\mathcal L, \lk_\Lat)$ ie.,
-a lattice which admits a total order and upper and lower bounds ($l^\top$ and $l^\bot$).
-We note lattice elements $l$ and $\glb_i l_i$ (resp. $\lub_i l_i$)
-the greatest lower bound (resp. least upper bound) in $\mathcal L$.
-%
-Let $\CL$ the set of constraints in this lattice. We consider the
-following base entailment rules in this system:
-\begin{center}
-  \begin{mathpar}
-    \inferrule{l \lk_\Lat l'}{\entail{}{\Cleq{l}{l'}}}
-    \and
-    \inferrule{}{\entail{}{\Cleq{k}{l^\top}}}
-    \and
-    \inferrule{}{\entail{}{\Cleq{l^\bot}{k}}}
-  \end{mathpar}
-\end{center}
-
-The normalization procedure is noted $\normalize{C_0}{\unif_0}$ where
-$C_0\in \CL$ is a set of constraints and $\unif_0$ is a substitution.
-It returns a constraint $C \in \mathcal S$ in
-solved form and a unifier $\unif$.
-The main idea of the algorithm is to first remove all the type equalities
-by using regular Herbrand unification. After that, we only have
-a set of inequalities among kinds, which we can consider as a relation.
-We can then saturate the relation,
-unify all kinds that are in the same equivalence classes to obtain
-a most general unifier on kind variables,
-remove all existentially quantified variables and
-then minimize back the relation.
-
-More precisely, we apply the following steps:
-\begin{enumerate}
-\item Solve all type equality constraints through Herband unification and
-  gather all existential quantifications at the front of the constraint.
-  We obtain a constraint $C^k = \exists \kvar_i,\ \Cleq{k_j}{k'_j}_j$ and
-  a substitution $\unif_\tau$.
-
-  We note $\mathcal R$ the relation $\Cleq{k_j}{k'_j}_j$,
-  $\mathcal G$ the underlying directed graph and $V$ its vertices.
-
-\item Saturate the lattice equalities in $\mathcal R$.
-
-  More precisely, for each kind variable $\kvar \in V$,
-  for each constant $l_i$ (resp. $l_j$) such that
-  there is a path from $l_i$ to $\kvar$ (resp. from $\kvar$ to $l_j$) in $\mathcal G$,
-  add an edge from $\lub l_i$ to $\kvar$
-  (resp. from $\kvar$ to $\glb l_j$).
-  This step is well defined since $\mathcal L$ is a bounded lattice
-  and $\lub\emptyset$ and $\glb\emptyset$ are well defined.
-
-  We also complement $\mathcal R$ with $(\leq)$ by adding an edge
-  between related constants.
-\item
-  At this point, we can easily check for satisfiability: A constraint
-  is satisfiable (in the given environment) if and only if,
-  for any constants $l_1$ and $l_2$ such that
-  there is a path from $l_1$ to $l_2$ in $\mathcal G$, then $l_1\lk_\Lat l_2$.
-  If this is not the case, we return \textbf{fail}.
-
-\item For each strongly connected component in $\mathcal G$, unify all its vertices and replace it by a representative.
-  We note $\unif_k$ the substitution that replaces a kind variable by
-  its representative.
-  The representative of a strongly connected component $g$ can be determined as follows:
-  \begin{itemize}
-  \item If $g$ does not contain any constant, then the representative
-    is a fresh kind variable.
-  \item If $g$ contains exactly one constant, it is the representative.
-  \item Otherwise, the initial constraint $C_0$ is not satisfiable.
-  \end{itemize}
-  Note that this step will also detect all unsatisfiable constraints.
-\item Take the transitive closure of $\mathcal R$.
-\item Remove all the vertices corresponding to the kind variables $\kvar_i$
-  that are existentially quantified in $C^k$.
-\item Take the transitive reduction of $\mathcal R$.
-\item Remove the extremums of $\mathcal L$ and the edges of $(\leq)$
-  from $\mathcal R$.
-\item Return $C = \left\{ k \leq k' \mid k \operatorname{\mathcal R}k' \right\}$
-  and $\unif =  \unif_\tau \meet \unif_k$.
-\end{enumerate}
-
-\TODO{Add an example}
-
-\subsubsection{Soundness and Principality}
-
-\begin{lemma}[Normal form]
-  \label{lemma:normalform}
-  Given a constraint $D\in\CL$, a substitution $\phi$ and
-  $(C,\unif) = \normalize{D}{\phi}$,
-  then $\phi\leq\unif$,
-  $C \equivC \unif D$ and
-  $\unif C = C$.
-\end{lemma}
-\begin{proof}
-  % For simplicity, we assume that any substitution has been already applied
-  % to $D$ and that $\phi = id$.
-  Let us partition $\phi$ into a part which affects type variables,
-  $\phi_\tau$, and a part which affects kind variables, $\phi_k$.
-
-  We note $(C^k,\unif_\tau)$ the result of
-  the modified Herbrand unification on $(D,\phi)$ in step (1).
-  Herbrand unification computes the most general
-  unifier. Our modified Herbrand unification only output additional
-  kind constraints for kind on the arrows and does not change
-  the result of the unification. Thus, we have
-  $\phi_\tau\leq\unif_\tau$,
-  $C^k \equivC \unif_\tau D$ and
-  $\unif_\tau C^k = C^k$.
-
-  Let $C^{k+}$ be the result after step (2), we trivially have that
-  $\fv{C^{k+}} = \fv{C^k}$ and that $C^{k+} \equivC C^k$.
-
-  Let $C^{A}$ and $\unif_k$ be the results after step (4).
-  By definition, we have $\unif_k C^{k+} \equivC C^{A}$ and
-  $\unif_k C^{A} = C^{A}$. Since $\phi_k$ has already be applied to $C$ before
-  unifying the strongly connected components,
-  we have that $\phi_k\leq\unif_k$.
-
-  Let $\unif = \unif_\tau \meet \unif_k$. Since $\unif_\tau$ and $\unif_k$
-  have disjoint supports,
-  we have $C^{A} = \unif_\tau C^{A} \equivC \unif C^{k+} \equivC \unif D$
-  and $\unif C^{A} = C^{A}$.
-  Furthermore, $\phi_\tau \meet \phi_k \leq \unif_\tau \meet \unif_k$.
-
-  Steps (5) to (9) all preserve the free variables and the equivalence
-  of constraints, which concludes.
-\end{proof}
-
-\begin{lemma}[Uniqueness]
-  Given $(C_1,\unif_1)$ and $(C_2,\unif_2)$ such that
-  $\unif_1 C_1 \equivC \unif_2 C_2$, then
-  $\normalize{C_1}{\unif_1}$ and $\normalize{C_2}{\unif_2}$
-  are identical up to $\alpha$-renaming.
-\end{lemma}
-\begin{proof}
-  In \cref{lemma:normalform}, we have showed that all the steps of the
-  normalization procedure preserve equivalence.
-  Since $\unif_1 C_1 \equivC \unif_2 C_2$, equivalence between
-  the two results of the normalization procedures is preserved for all steps.
-
-  We note $P(C_a)$ if for all $C = (k, k)'$
-  such that $\entail{C_a}{C}$ and $\nvdash_eC$,
-  we have $C \in {\mathcal R}_a$.
-
-  Let us note $C_1'$ and $C_2'$ the constraints after step (4). $P(C_1')$ and
-  $P(C_2')$ hold. Indeed, since $C_1'$ and $C_2'$ are only composed
-  of existential quantifications and kind inequalities, the only rules
-  that applies are transitivity and lattice inequalities.
-  After step (2) and (5), the associated relations are fully saturated for these
-  two rules, hence all inequalities that can be deduced from $C_a'$ are already
-  present in the relation.
-
-  The property $P$ is preserved by step (6) since we only remove
-  inequalities that involve existentially quantified variables. Such
-  inequalities could not be picked in $P$.
-  % In (6) we only remove
-  % trivial lattice inequalities.
-
-  Let us note $C''_a$ for $a\in\{1,2\}$ the constraints after step (5).
-  Since there are no more existential variables,
-  we have $C''_a=(k_i,k'_i)_i=\mathcal R''_a$.
-  For any $C=(k,k')$ such that $\entail{}{C}$ and $\entail{C''_a}{C}$,
-  then $C \in (\leq) \subset {\mathcal R}''_a$.
-  Indeed, the only trivial inequalities in our system are equalities of the form
-  $(\kvar, \kvar)$, which were removed in step (4) and the lattice inequalities.
-
-  Let us consider $C = (k,k') \in \mathcal R''_1$.
-  Since $C''_1\equivC C''_2$, we have $\entail{C''_2}{C}$.
-  If $\nvdash_e{C}$, by $P(C''_2)$ we have that
-  $C\in R''_2$.
-  If $\entail{}{C}$, then $C\in (\leq) \subset R''_2$.
-  We conclude that $R''_1 \subset R''_2$.
-  By symmetry, $R''_1 = R''_2$ and $C''_1 = C''_2$.
-
-  This equality is preserved by step (7) and (8)
-  since the transitive reduction of a directed acyclic graph is unique,
-  which concludes.
-\end{proof}
-
-\begin{theorem}
-  $\CL$ has the principal constraint property and
-  $\operatorname{normalize}$ computes principal normal forms for $\CL$.
-\end{theorem}
-
-\begin{lemma}
-  For all $C\in\mathcal S$, $\entail{C}{x = x}$ implies
-  $\entail{}{x = x}$.
-\end{lemma}
-\begin{proof}
-  By definition of $\operatorname{normalize}$, We have $C = \Cleq{k}{k'}_i$
-  such that the underlying relation has no cycles.
-  Thus, we can not deduce neither kind nor type equalities from $C$.
-\end{proof}
-
-\begin{lemma}
-  $\CL$ is regular, ie, for $x, x'$ two types or kinds,
-  $\entail{}{\Ceq{x}{x'}}$ implies
-  $\fv{x} = \fv{x'}$
-\end{lemma}
-\begin{proof}
-  The only equalities possibles are between variables (via symmetry) or
-  between constants.
-\end{proof}
-
-\begin{theorem}
-  Type inference in $HM(\CL)$ is sound, principal and complete.
-\end{theorem}
-
-\subsection{Principality}
-
-\TODO{}
-
-\begin{itemize}
-\item Properly ensure that it respects HM(X) (cylindric, \dots)
-\item Show principal type inference:
-  \begin{itemize}
-  \item Principal constraint system
-  \item Regular constraint system: $\Ceq{\tau}{\tau'} \implies \fv{\tau} = \fv{\tau'}$.
-  \item Solved forms are in simplified form.
-    $C\in S, \entail{C}{\Ceq{\tau}{\tau'}} \implies \entail{}{\Ceq{\tau}{\tau'}}$.
-  \end{itemize}
-\item Show equivalence with the logic-based system.
-
-\end{itemize}
-
-% \begin{lemma}
-%   Without loss of generality, we can consider that
-%   kind inequalities in satisfiable constraints
-%   are only done on kind variables.
-
-%   \begin{proof}
-%     Consider the constraint $\Cleq{k}{k'}\Cand C$.
-%     \begin{itemize}
-%     \item If $k$ and $k'$ are both constants, it can be removed.
-%     \item If the constraint is $\Cleq{\kvar}{\kaff}$ or $\Cleq{\kun}{\kvar}$, it can be removed.
-%     \item If the constraint is $\Cleq{\kvar}{\kun}$ or $\Cleq{\kaff}{\kvar}$, we
-%       can substitute $\kvar$ by its value in $C$.
-%     \end{itemize}
-%   \end{proof}
-% \end{lemma}
-
 %%% Local Variables:

### 2019-07-07 kindly/lang.tex
Small fixes.
cosmetics-syntax

@@
 
-We now present a formalization of the \lang language featuring
-a minimal ML language with let-polymorphism, abstract types, pairs, and our novel
-linear type system with borrows which uses kinds and qualified types.
-For ease of presentation,
-our formalization makes the following simplifications.
-First, we only consider matching on pairs, instead of arbitrary algebraic
-datatypes. We consider separate operators for borrowing and reborrowing (taking
-the borrow of a borrow). We consider regions to be fully annotated
-and present a set of rules to infer their placement.
-Finally we emulates region variables
-as a series of nesting indices.
+\lang is a minimal ML-like language with
+let-polymorphism, abstract types, and pairs. Its type system manages
+linearity and borrows using kinds and (kind-) constrained types.
+For ease of presentation, we keep the language as simple as possible.
+\begin{itemize}
+\item We showcase pattern matching on pairs rather than algebraic
+  datatypes.
+\item We have separate operators for borrowing and reborrowing (taking
+  the borrow of a borrow).
+\item We assume that regions are explicit in the code. (But we present
+  an algorithm to infer them in \cref{regionannot}.)
+\item We use de Bruijn-indexing to identify regions instead of region
+  variables. 
+\end{itemize}
 
-In the rest of this section, we present the syntax (\cref{syntax})
-the syntax directed typing rules (\cref{sdtyping})
-and a linearity-aware semantics (\cref{sem}) for \lang.
+The rest of this section formalizes  \lang: the syntax (\cref{syntax}),
+the statics in terms of syntax directed typing (\cref{sdtyping}),
+and a dynamics that is linearity-aware (\cref{sem}).
 
@@
 \label{syntax}
+
+\Cref{grammar} defines the syntax of \lang. Expressions are as usual
+in an ML-like language.  The novel aspects are match
+specifications, regions, and borrows.
+
+We start from the bottom.
+A borrow $\borrow{x}$ is always taken from a variable $x$. The
+\emph{borrow annotation} $\BORROW$ indicates whether the borrow is mutable
+($\MBORROW$) or immutable ($\IBORROW$). If $x$ is already borrowed,
+then we need to use the reborrow expression $\reborrow{x}$.
+
+A region $\region{x}{e}$ is annotated with its nesting $n$
+and the variable $x$ that may be borrowed in $e$.
+
+A match is indexed with a \emph{match specification} $\etransfm$ that indicates
+whether the match should work on borrows ($\etransfm=\&^b$) or not ($\etransfm=id$).
+
+Many types are indexed with kinds, ranged over by $k$.
+A kind is either a kind variable $\kvar$ or a constant
+(linear $\klin$, affine $\kaff$, or unrestricted $\kun$) indexed
+by a nesting level $n \in \Nat \cup \{\infty\}$.
+
+A type $\tau$ is either a type variable, a function type indexed by a
+kind, 
+a type application $\tapp{t}{\Multi{\tau}}$ of an abstract type constructor
+$\T t$, or a borrowed type  $\borrowty{k}{\tau}$.
+
+Type schemes $\schm$ add quantification over kind
+variables $\kappa$, kinded type variables $\alpha:k$, and constraints
+$C$ to a type, where a constraint is a list of inequalities over
+kinds.
+
+Abstract type constructors possess kind schemes $\kschm$ which relate
+the kinds of the type constructors arguments to the kind of the
+constructed type.
+
+Generally, we write lists with an overbar and (sometimes) an index as
+in $\Multi[i]\tau$.
+
+\paragraph{Previously}
 
@@
 
-We illustrate this on the expression $\region{x}{f(\borrow[\IBORROW]{c})}$.
+We illustrate this on the expression $\region{x}{f(\borrow[\IBORROW]{c})}$
+when $c$ is a linear channel.
 The first step is to check that
@@
 \begin{figure}[!h]
-  \begin{minipage}{0.35\linewidth}
+  \begin{minipage}{0.4\linewidth}
     \centering

### 2019-07-07 kindly/infer/infer.tex
Select the various part of the type inference section.
Start moving proof to the appendix.

@@
 
-The main innovation of \affe is its principal type inference.
+An important innovation of \affe is its principal type inference.
 We now formulate our inference algorithm
@@
 system such that inference is sound and complete.
+In addition, our constraint solving algorithm simplifies constraints in a
+principled way.
 We first present various preliminary definitions
@@
 
-\begin{figure*}[htp]
-  \input{infer/typing}
-  \caption{Type inference algorithm -- $\inferW{\Sigma}{(C,\psi)}{\bf{\E}}{\bf{e}}{\tau}$ }
-  \label{rules:typing}
-\end{figure*}
+\subsubsection{Environments and bindings}
+\label{infer:envs}
 
-\begin{figure}[tp]
-  \centering
+\begin{figure}[!h]
   \begin{mathpar}
@@
       \begin{dcases}
-        \Ctrue& \text{if } x\in\Sv\\
+        \Ctrue&\text{if } x\in\Sv\\
         \Cleq{\sigma}{\kaff_\infty}&\\
@@
     }
+    \and
+    \ruleIVar
+    \and
+    \ruleIAbs
+  \end{mathpar}
+  \caption{The {\sc Abs$_I$} and {\sc Var$_I$} rules}
+  \label{rule:infer:envs}
+\end{figure}
+
+\subsubsection{Copying and Splitting}
+\label{infer:split}
+
+\begin{figure}[!h]
+  \centering
+  \begin{mathpar}
+    \ruleIPair
+  \end{mathpar}
+  \caption{The {\sc Pair$_I$} rule}
+  \label{rule:infer:pair}
+\end{figure}
+
+\subsubsection{Regions}
+\label{infer:regions}
+
+\begin{figure}[!h]
+  \centering
+  \begin{mathpar}
+    \ruleIRegion
+  \end{mathpar}
+  \caption{The {\sc Region$_I$} rule}
+  \label{rule:infer:region}
+\end{figure}
+
+\subsubsection{Constraint desugaring}
+
+\begin{figure}[!htb]
+  \begin{mathpar}
+
+  \inferrule{}{ \Cleq{\Eempty}{k} \Crewrite  \Cempty}
+
+  \inferrule{
+    \Cleq\E k \Crewrite  C \\ \E \vdash \Cleq B k \Crewrite  D
+  }{
+    \Cleq{\E; B}{k} \Crewrite  C \Cand D
+  }
+
+  \inferrule{
+    \E \vdash \Cleq \schm k \Crewrite  C
+  }{ \E \vdash \Cleq{\bvar x \schm}{k} \Crewrite  C}
+
+  \inferrule{ }{
+    \E \vdash \Cleq{\svar x \schm^n} k \Crewrite  \Cfalse
+  }
+
+  % \inferrule{}{\Cleq{\IBORROW^n} k \Crewrite  \Cleq{\kun_n} k}
+
+  % \inferrule{}{\Cleq{\MBORROW^n} k \Crewrite  \Cleq{\kaff_n} k}
 
@@
     }{
-      \E \vdash \Cleq{\forall \kvar_i \forall (\tvar_j:k_j).\ \qual{C_x}{\tau}}{k} \Crewrite  D
+      \E \vdash
+      \Cleq{(\forall \kvar_i \forall (\tvar_j:k_j).\ \qual{C_x}{\tau})}{k} \Crewrite  D
     }
   \end{mathpar}
-  \caption{Auxiliary rules for inference}
-  \label{op:usgmap}
+
+  \caption{Rewriting constraints on environments and types}
+  \label{rule:typing:desugar}
 \end{figure}
@@
 remove all existentially quantified variables and
-then minimize back the relation.
+then minimize back the relation and apply various
+simplification rules to make the resulting type easier to understand to users.
 
@@
 
-\TODO{Simplify/Cut}
-
-\begin{lemma}[Normal form]
-  \label{lemma:normalform}
-  Given a constraint $D\in\CL$, a substitution $\phi$ and
-  $(C,\unif) = \normalize{D}{\phi}$,
-  then $\phi\leq\unif$,
-  $C \equivC \unif D$ and
-  $\unif C = C$.
-\end{lemma}
-\begin{proof}
-  % For simplicity, we assume that any substitution has been already applied
-  % to $D$ and that $\phi = id$.
-  Let us partition $\phi$ into a part which affects type variables,
-  $\phi_\tau$, and a part which affects kind variables, $\phi_k$.
-
-  We note $(C^k,\unif_\tau)$ the result of
-  the modified Herbrand unification on $(D,\phi)$ in step (1).
-  Herbrand unification computes the most general
-  unifier. Our modified Herbrand unification only output additional
-  kind constraints for kind on the arrows and does not change
-  the result of the unification. Thus, we have
-  $\phi_\tau\leq\unif_\tau$,
-  $C^k \equivC \unif_\tau D$ and
-  $\unif_\tau C^k = C^k$.
-
-  Let $C^{k+}$ be the result after step (2), we trivially have that
-  $\fv{C^{k+}} = \fv{C^k}$ and that $C^{k+} \equivC C^k$.
-
-  Let $C^{A}$ and $\unif_k$ be the results after step (4).
-  By definition, we have $\unif_k C^{k+} \equivC C^{A}$ and
-  $\unif_k C^{A} = C^{A}$. Since $\phi_k$ has already be applied to $C$ before
-  unifying the strongly connected components,
-  we have that $\phi_k\leq\unif_k$.
-
-  Let $\unif = \unif_\tau \meet \unif_k$. Since $\unif_\tau$ and $\unif_k$
-  have disjoint supports,
-  we have $C^{A} = \unif_\tau C^{A} \equivC \unif C^{k+} \equivC \unif D$
-  and $\unif C^{A} = C^{A}$.
-  Furthermore, $\phi_\tau \meet \phi_k \leq \unif_\tau \meet \unif_k$.
-
-  Steps (5) to (9) all preserve the free variables and the equivalence
-  of constraints, which concludes.
-\end{proof}
-
-\begin{lemma}[Uniqueness]
-  Given $(C_1,\unif_1)$ and $(C_2,\unif_2)$ such that
-  $\unif_1 C_1 \equivC \unif_2 C_2$, then
-  $\normalize{C_1}{\unif_1}$ and $\normalize{C_2}{\unif_2}$
-  are identical up to $\alpha$-renaming.
-\end{lemma}
-\begin{proof}
-  In \cref{lemma:normalform}, we have showed that all the steps of the
-  normalization procedure preserve equivalence.
-  Since $\unif_1 C_1 \equivC \unif_2 C_2$, equivalence between
-  the two results of the normalization procedures is preserved for all steps.
-
-  We note $P(C_a)$ if for all $C = (k, k)'$
-  such that $\entail{C_a}{C}$ and $\nvdash_eC$,
-  we have $C \in {\mathcal R}_a$.
-
-  Let us note $C_1'$ and $C_2'$ the constraints after step (4). $P(C_1')$ and
-  $P(C_2')$ hold. Indeed, since $C_1'$ and $C_2'$ are only composed
-  of existential quantifications and kind inequalities, the only rules
-  that applies are transitivity and lattice inequalities.
-  After step (2) and (5), the associated relations are fully saturated for these
-  two rules, hence all inequalities that can be deduced from $C_a'$ are already
-  present in the relation.
-
-  The property $P$ is preserved by step (6) since we only remove
-  inequalities that involve existentially quantified variables. Such
-  inequalities could not be picked in $P$.
-  % In (6) we only remove
-  % trivial lattice inequalities.
-
-  Let us note $C''_a$ for $a\in\{1,2\}$ the constraints after step (5).
-  Since there are no more existential variables,
-  we have $C''_a=(k_i,k'_i)_i=\mathcal R''_a$.
-  For any $C=(k,k')$ such that $\entail{}{C}$ and $\entail{C''_a}{C}$,
-  then $C \in (\leq) \subset {\mathcal R}''_a$.
-  Indeed, the only trivial inequalities in our system are equalities of the form
-  $(\kvar, \kvar)$, which were removed in step (4) and the lattice inequalities.
-
-  Let us consider $C = (k,k') \in \mathcal R''_1$.
-  Since $C''_1\equivC C''_2$, we have $\entail{C''_2}{C}$.
-  If $\nvdash_e{C}$, by $P(C''_2)$ we have that
-  $C\in R''_2$.
-  If $\entail{}{C}$, then $C\in (\leq) \subset R''_2$.
-  We conclude that $R''_1 \subset R''_2$.
-  By symmetry, $R''_1 = R''_2$ and $C''_1 = C''_2$.
-
-  This equality is preserved by step (7) and (8)
-  since the transitive reduction of a directed acyclic graph is unique,
-  which concludes.
-\end{proof}
-
-\begin{theorem}
-  $\CL$ has the principal constraint property and
-  $\operatorname{normalize}$ computes principal normal forms for $\CL$.
-\end{theorem}
-
-\begin{lemma}
-  For all $C\in\mathcal S$, $\entail{C}{x = x}$ implies
-  $\entail{}{x = x}$.
-\end{lemma}
-\begin{proof}
-  By definition of $\operatorname{normalize}$, We have $C = \Cleq{k}{k'}_i$
-  such that the underlying relation has no cycles.
-  Thus, we can not deduce neither kind nor type equalities from $C$.
-\end{proof}
-
-\begin{lemma}
-  $\CL$ is regular, ie, for $x, x'$ two types or kinds,
-  $\entail{}{\Ceq{x}{x'}}$ implies
-  $\fv{x} = \fv{x'}$
-\end{lemma}
-\begin{proof}
-  The only equalities possibles are between variables (via symmetry) or
-  between constants.
-\end{proof}
-
-\begin{theorem}
-  Type inference in $HM(\CL)$ is sound, principal and complete.
-\end{theorem}
-
-\TODO{}
-
+Our system is an extension of HM(X). Our goal is prove soundness and completeness
+of our inference algorithm with respect to our syntax-directed system, which
+assures us principal type inference.
+We prove two properties:
 \begin{itemize}
-\item Properly ensure that it respects HM(X) (cylindric, \dots)
-\item Show principal type inference:
-  \begin{itemize}
-  \item Principal constraint system
-  \item Regular constraint system: $\Ceq{\tau}{\tau'} \implies \fv{\tau} = \fv{\tau'}$.
-  \item Solved forms are in simplified form.
-    $C\in S, \entail{C}{\Ceq{\tau}{\tau'}} \implies \entail{}{\Ceq{\tau}{\tau'}}$.
-  \end{itemize}
-\item Show equivalence with the logic-based system.
-
+\item Our constraint system respects all the properties required by HM(X)
+  for principal type inference.
+\item Our extended inference algorithm still satisfies the soundness
+  and completeness properties of HM(X). This is achieved by extending
+  the original proofs.
 \end{itemize}
 
-% \begin{lemma}
-%   Without loss of generality, we can consider that
-%   kind inequalities in satisfiable constraints
-%   are only done on kind variables.
-
-%   \begin{proof}
-%     Consider the constraint $\Cleq{k}{k'}\Cand C$.
-%     \begin{itemize}
-%     \item If $k$ and $k'$ are both constants, it can be removed.
-%     \item If the constraint is $\Cleq{\kvar}{\kaff}$ or $\Cleq{\kun}{\kvar}$, it can be removed.
-%     \item If the constraint is $\Cleq{\kvar}{\kun}$ or $\Cleq{\kaff}{\kvar}$, we
-%       can substitute $\kvar$ by its value in $C$.
-%     \end{itemize}
-%   \end{proof}
-% \end{lemma}
+\TODO{State the theorems}
 

### 2019-07-07 kindly/sdtyping/extra.tex
Start working on the inference section.
Fix missing refs.
Select the various part of the type inference section.
Start moving proof to the appendix.

@@
 
-\begin{figure}[tp]
+\begin{figure}[btp]
   \centering
@@
   \label{grammar:constraint}
-\end{figure}
-
-\begin{figure*}[tp]
-  \input{infer/lattice}
-  \caption{Lattice inequalities -- $k \lk_\Lat k'$}
+  % \input{infer/lattice}
+  % \caption{Lattice inequalities -- $k \lk_\Lat k'$}
   \input{infer/entails}
@@
   \label{rules:entail}
-\end{figure*}
+\end{figure}
 
@@
 %
-We consider the existence of a function $\operatorname{normalize}$ which takes
-a constraint in $\A$ and a substitution $\psi$ and returns a constraint
-in solved form $C' \in \SC$,
-and an updated substitution. We detail the implementation
-of the normalization function in \cref{sec:normalize}
+The normalization function is described in \cref{infer:solving}.
+
+% We consider the existence of a function $\operatorname{normalize}$ which takes
+% a constraint in $\A$ and a substitution $\psi$ and returns a constraint
+% in solved form $C' \in \SC$,
+% and an updated substitution. We detail the implementation
+% of the normalization function in \cref{infer:solving}
 
@@
 % \TODO{Extend the properties of solved forms}
+
+Our goal is to prove that $\CL$ supports all the properties necessary for
+principal type inference, as defined by HM(X).
+
+\begin{lemma}[Normal form]
+  \label{lemma:normalform}
+  Given a constraint $D\in\CL$, a substitution $\phi$ and
+  $(C,\unif) = \normalize{D}{\phi}$,
+  then $\phi\leq\unif$,
+  $C \equivC \unif D$ and
+  $\unif C = C$.
+\end{lemma}
+\begin{proof}
+  % For simplicity, we assume that any substitution has been already applied
+  % to $D$ and that $\phi = id$.
+  Let us partition $\phi$ into a part which affects type variables,
+  $\phi_\tau$, and a part which affects kind variables, $\phi_k$.
+
+  We note $(C^k,\unif_\tau)$ the result of
+  the modified Herbrand unification on $(D,\phi)$ in step (1).
+  Herbrand unification computes the most general
+  unifier. Our modified Herbrand unification only output additional
+  kind constraints for kind on the arrows and does not change
+  the result of the unification. Thus, we have
+  $\phi_\tau\leq\unif_\tau$,
+  $C^k \equivC \unif_\tau D$ and
+  $\unif_\tau C^k = C^k$.
+
+  Let $C^{k+}$ be the result after step (2), we trivially have that
+  $\fv{C^{k+}} = \fv{C^k}$ and that $C^{k+} \equivC C^k$.
+
+  Let $C^{A}$ and $\unif_k$ be the results after step (4).
+  By definition, we have $\unif_k C^{k+} \equivC C^{A}$ and
+  $\unif_k C^{A} = C^{A}$. Since $\phi_k$ has already be applied to $C$ before
+  unifying the strongly connected components,
+  we have that $\phi_k\leq\unif_k$.
+
+  Let $\unif = \unif_\tau \meet \unif_k$. Since $\unif_\tau$ and $\unif_k$
+  have disjoint supports,
+  we have $C^{A} = \unif_\tau C^{A} \equivC \unif C^{k+} \equivC \unif D$
+  and $\unif C^{A} = C^{A}$.
+  Furthermore, $\phi_\tau \meet \phi_k \leq \unif_\tau \meet \unif_k$.
+
+  Steps (5) to (9) all preserve the free variables and the equivalence
+  of constraints, which concludes.
+\end{proof}
+
+\begin{lemma}[Uniqueness]
+  Given $(C_1,\unif_1)$ and $(C_2,\unif_2)$ such that
+  $\unif_1 C_1 \equivC \unif_2 C_2$, then
+  $\normalize{C_1}{\unif_1}$ and $\normalize{C_2}{\unif_2}$
+  are identical up to $\alpha$-renaming.
+\end{lemma}
+\begin{proof}
+  In \cref{lemma:normalform}, we have showed that all the steps of the
+  normalization procedure preserve equivalence.
+  Since $\unif_1 C_1 \equivC \unif_2 C_2$, equivalence between
+  the two results of the normalization procedures is preserved for all steps.
+
+  We note $P(C_a)$ if for all $C = (k, k)'$
+  such that $\entail{C_a}{C}$ and $\nvdash_eC$,
+  we have $C \in {\mathcal R}_a$.
+
+  Let us note $C_1'$ and $C_2'$ the constraints after step (4). $P(C_1')$ and
+  $P(C_2')$ hold. Indeed, since $C_1'$ and $C_2'$ are only composed
+  of existential quantifications and kind inequalities, the only rules
+  that applies are transitivity and lattice inequalities.
+  After step (2) and (5), the associated relations are fully saturated for these
+  two rules, hence all inequalities that can be deduced from $C_a'$ are already
+  present in the relation.
+
+  The property $P$ is preserved by step (6) since we only remove
+  inequalities that involve existentially quantified variables. Such
+  inequalities could not be picked in $P$.
+  % In (6) we only remove
+  % trivial lattice inequalities.
+
+  Let us note $C''_a$ for $a\in\{1,2\}$ the constraints after step (5).
+  Since there are no more existential variables,
+  we have $C''_a=(k_i,k'_i)_i=\mathcal R''_a$.
+  For any $C=(k,k')$ such that $\entail{}{C}$ and $\entail{C''_a}{C}$,
+  then $C \in (\leq) \subset {\mathcal R}''_a$.
+  Indeed, the only trivial inequalities in our system are equalities of the form
+  $(\kvar, \kvar)$, which were removed in step (4) and the lattice inequalities.
+
+  Let us consider $C = (k,k') \in \mathcal R''_1$.
+  Since $C''_1\equivC C''_2$, we have $\entail{C''_2}{C}$.
+  If $\nvdash_e{C}$, by $P(C''_2)$ we have that
+  $C\in R''_2$.
+  If $\entail{}{C}$, then $C\in (\leq) \subset R''_2$.
+  We conclude that $R''_1 \subset R''_2$.
+  By symmetry, $R''_1 = R''_2$ and $C''_1 = C''_2$.
+
+  This equality is preserved by step (7) and (8)
+  since the transitive reduction of a directed acyclic graph is unique,
+  which concludes.
+\end{proof}
+
+\begin{theorem}
+  $\CL$ has the principal constraint property and
+  $\operatorname{normalize}$ computes principal normal forms for $\CL$.
+\end{theorem}
+
+\begin{lemma}
+  For all $C\in\mathcal S$, $\entail{C}{x = x}$ implies
+  $\entail{}{x = x}$.
+\end{lemma}
+\begin{proof}
+  By definition of $\operatorname{normalize}$, We have $C = \Cleq{k}{k'}_i$
+  such that the underlying relation has no cycles.
+  Thus, we can not deduce neither kind nor type equalities from $C$.
+\end{proof}
+
+\begin{lemma}
+  $\CL$ is regular, ie, for $x, x'$ two types or kinds,
+  $\entail{}{\Ceq{x}{x'}}$ implies
+  $\fv{x} = \fv{x'}$
+\end{lemma}
+\begin{proof}
+  The only equalities possibles are between variables (via symmetry) or
+  between constants.
+\end{proof}
+
+\begin{theorem}
+  $\CL$ has the principal constraint property and
+  $\operatorname{normalize}$ computes principal normal forms for $\CL$.
+\end{theorem}
+
+This is sufficient to show that $HM(\CL)$ is principal. However,
+we do not use HM(X) directly, but an extended version with kind inference,
+linear and affine types, and borrow.
 
@@
 \subsection{Environments}
+\label{typ:extra:envs}
 

### 2019-07-08 kindly/motivation.tex
Misc code fixes in the motivations.

@@
   \centering
-  \begin{subfigure}[t]{0.48\linewidth}
+  \begin{subfigure}[t]{0.47\linewidth}
 \begin{lstlisting}
@@
   \end{subfigure}\hfill
-  \begin{subfigure}[t]{0.5\linewidth}
+  \begin{subfigure}[t]{0.53\linewidth}
 \begin{lstlisting}
@@
     \label{ex:array}
-\begin{lstlisting}
+\begin{lstlisting}[firstnumber=2]
   let a = create n 1 in
@@
   \begin{minipage}[t]{0.54\linewidth}
-    \lstinputlisting[linerange=15-29,firstnumber=last]{code/sukodu.affe}
+    \lstinputlisting[linerange=15-30,firstnumber=last]{code/sukodu.affe}
   \end{minipage}

### 2019-07-08 kindly/infer/infer.tex
State the theorems.

@@
 
-\TODO{State the theorems}
+This section only states the theorems (and the necessary definitions).
+Details and proofs are provided in \cref{appendix:constraints,appendix:infer}.
+
+\subsubsection{Constraint system}
+
+We requires three properties from our constraint solving algorithms: it should
+compute normal firms, such normal forms should be unique, and the
+system should be regular.
+
+\begin{property}[Principal normal form]
+  Normalization computes principal normal forms for $\CL$, i.e. 
+  given a constraint $D\in\CL$, a substitution $\phi$ and
+  $(C,\unif) = \normalize{D}{\phi}$,
+  then $\phi\leq\unif$,
+  $C \equivC \unif D$ and
+  $\unif C = C$.
+\end{property}
+
+\begin{property}[Principal constraints]
+  $\CL$ has the principal constraint property, i.e.
+  for every $D\in\CL$, and a substitution $\unif$,
+  either $D$ does not have a normal form, or it has
+  a principal normal form.
+\end{property}
+
+\begin{property}[Regular constraint system]
+  $\CL$ is regular, ie, for $x, x'$ two types or kinds,
+  $\entail{}{\Ceq{x}{x'}}$ implies
+  $\fv{x} = \fv{x'}$
+\end{property}
+
+The proofs are given in \cref{appendix:constraints}. These three properties
+are sufficient to state that $HM(\CL)$ provides principal type inference.
+However, we do not use HM(X) directly, but an extension with kind inference
+and affine types and borrows.
+We now show that HM(X)'s completeness and soundness theorems are still valid on
+this extended system.
+
+\subsection{Inference algorithm}
+
+The first theorem states that our inference algorithm is sound
+with respect to the syntax directed type system.
+
+\begin{theorem}[Soundness of inference]
+  Given a type environment $\E$ and a term $e$,\\
+  if $\inferW{\Sv}{(C,\unif)}{\E}{e}{\tau}$
+  then $\inferS{C}{\unif\Sv}{e}{\tau}$, $\unif C = C$ and $\unif \tau = \tau$.
+\end{theorem}
+
+Note here that the syntax-directed derivation holds with the usage environment $\Sv$ instead of the originally provided environment $\E$. Indeed,
+$\E$ does not contain suspended and borrow bindings, since those
+are discovered on the fly and recorded in $\Sv$.
+
+The second theorem states that our algorithm is complete: for any given
+syntax directed typing derivation, our inference algorithm can find
+a derivation that gives a type at least as general.
+For this, we need first to provide a few additional definitions.
+
+\begin{definition}
+  Given a set of variable $U$ and $\unif$, $\unif'$ and $\phi$
+  substitutions. \\
+  Then
+  $\unif \leq^{\phi}_{U} \unif'$ iff $(\phi \circ \unif)|_{U} = \unif'|_U$.
+\end{definition}
+
+\begin{definition}
+  Given a constraints $C$ and two schemes
+  $\schm = \forall \Multi\tvar. \qual{D}{\tau}$ and
+  $\schm' = \forall \Multi\tvar'. \qual{D'}{\tau'} $.
+  Then $\entail{C}{\schm \preceq \schm'}$
+  iff $\entail{C}{\subst{\tvar}{\tau''} D}$
+  and $\entail{C\Cand D'}{\Cleq{\subst{\tvar}{\tau''}\tau}{\tau'}}$
+\end{definition}
+
+\begin{definition}
+A flattened environment,
+noted $\Eflat\E$, is the environment
+where all the binders are replace by normal ones. More formally:
+\begin{align*}
+  \Eflat\E
+  &= \left\{ \bvar{x}{\tau} \mid
+    \bvar{x}{\tau}\in\E
+    \vee \bvar{\borrow{x}}{\borrowty{k}{\tau}}\in\E
+    \vee \svar{x}{\tau}^n\in\E
+    \right\} \cup \left\{ \bvar{\tvar}{k} \mid \bvar{\tvar}{k}\in\E \right\}
+\end{align*}
+\end{definition}
+
+We can then state the completeness theorem.
+
+\begin{theorem}[Completeness]
+  Let $\inferS{C'}{\unif'\E}{e}{\schm'}$ a typing derivation,
+  then $\inferW{\Sv}{(C,\unif)}{\Eflat\E}{e}{\tau}$
+  for some substitution $\unif$, constraint $C$ and type $\tau$ such
+  that
+  \begin{align*}
+    (C_\schm,\schm) &= \generalize{C}{\unif\E}{\tau}
+    &\unif &\leq^{\phi}_{\fv{\E}} \unif'
+    &\entail{C'&}{\phi C}
+    &\entail{C'&}{\phi \schm \preceq \schm'}
+    % &( C, \schm, \unif) &\leq (C',\schm',\unif')
+  \end{align*}
+\end{theorem}
+
+The proofs are given in \cref{appendix:infer}.
 

### 2019-07-08 kindly/sdtyping/extra.tex
State the theorems.

@@
 \subsection{Constraint language}
+\label{appendix:constraints}
 
@@
 principal type inference, as defined by HM(X).
+
+In order to prove this, we first prove that constraint solving
+does compute normal forms, and that such normal forms are unique.
 
@@
 
-\begin{theorem}
-  $\CL$ has the principal constraint property and
-  $\operatorname{normalize}$ computes principal normal forms for $\CL$.
-\end{theorem}
+We can now prove all the necessary high level properties.
 
@@
 
-\begin{lemma}
+\begin{property}
   $\CL$ is regular, ie, for $x, x'$ two types or kinds,
@@
   $\fv{x} = \fv{x'}$
-\end{lemma}
+\end{property}
 \begin{proof}
@@
 
+Finally, we can conclude with all the properties we need for
+HM(X):
+
 \begin{theorem}
-  $\CL$ has the principal constraint property and
-  $\operatorname{normalize}$ computes principal normal forms for $\CL$.
+  $\CL$ has the principal constraint property,
+  $\operatorname{normalize}$ computes principal normal forms for $\CL$
+  and $\CL$ is regular.
 \end{theorem}
@@
 This is sufficient to show that $HM(\CL)$ is principal. However,
-we do not use HM(X) directly, but an extended version with kind inference,
+we do not use HM(X) directly but an extended version with kind inference,
 linear and affine types, and borrow.

### 2019-07-08 kindly/intro.tex
minor cosmetics

@@
 Most type systems provide just type soundness, but neglect the
-protocol aspect. The systems that do build on linear types \cite{Girard} and/or
+protocol aspect. The systems that do build on linear types \cite{DBLP:journals/tcs/Girard87} and/or
 uniqueness types\cite{DBLP:conf/plilp/BarendsenS95}. A value of linear
@@
 \begin{itemize}
-\item Type system that encodes linearity and affinity with borrowing
-\item Full type inference for \dots
-\item Constraint simplification
-\item Automatic inference of regions for borrows
+\item A type system that encodes linearity and affinity with
+  borrowing. The type system is polymorphic with respect to types as
+  well as kinds that express constraint on the number of uses of a value.
+\item An extension of the HM(X) framework
+  \cite{DBLP:journals/tapos/OderskySW99} with kinds, regions, and
+  multiplicity constraints that supports full, prinicipal type inference.
+\item Constraint simplification techniques.
+\item Automatic inference of regions for borrows.
 \end{itemize}

### 2019-07-08 kindly/infer/infer.tex
fixup! State the theorems.

@@
 
-\subsection{Inference algorithm}
+\subsubsection{Inference algorithm}
 
@@
 
-\begin{definition}
+\begin{definition}[More general unifier]
   Given a set of variable $U$ and $\unif$, $\unif'$ and $\phi$
@@
 
-\begin{definition}
+\begin{definition}[Instance relation]
   Given a constraints $C$ and two schemes
@@
 
-\begin{definition}
+\begin{definition}[Flattened Environment]
 A flattened environment,

### 2019-07-08 kindly/infer/infer.tex
Inference and constraints

@@
 as normal environment. In order to distinguish them more easily,
-we will general note them $\Sv$.
+we note them $\Sv$.
 
@@
 inference algorithm.
-The rules are shown in \cref{rules:typing}.
-
-\TODO{Explain/Cut}
+We revisit the syntax-directed rules presented in \cref{sdtyping} to highlight the
+novelties
+of our inference algorithm and the differences with the syntax-directed
+system.
 
@@
 \label{infer:envs}
+
+In our syntax directed system presented in \cref{sdtyping}, we ensured
+that linear variables are not discarded at the \emph{leafs}, through
+the {\sc Var} rule. In the inference algorithm, we operate in the opposite
+direction: we collect data from the leafs, and only enforce linearity
+at \emph{binders}. This is reflected in the {\sc Var$_I$} and
+{\sc Abs$_I$} rules presented in \cref{rule:infer:envs}.
+Typing for variables is very similar to traditional (non-affine) ML
+typing. The only difference is that we collect
+the fact that $x$ was used with the scheme $\schm$ by returning
+a usage environment $\Sv = \{ \bvar{x}{\schm} \}$.
+%
+This usage environment is in turn used at each binder to enforce proper
+usage of linear variable via the $\Weaken$ property.
+We demonstrate how this is achieved for lambda expressions
+in the {\sc Abs$_I$} rule.
+First, we typecheck the body of the lambda and obtain a usage
+environment $\Sv_x$. Just like in the syntax directed,
+we introduce the constraint
+$\Cleq{\Sv\Sdel{x}}{\kvar}$ which properly accounts for captures in
+the body of the lambda expression. We also introduce the constraint
+$\Weaken_{\bvar{x}{\sigma}}(\Sv)$. If $x$ was used $e$, $x$ is present
+in $\Sv$ and this constraint is empty. Otherwise, if $x$ is
+\emph{not} present in $\Sv$, its type should not have a linear
+kind, which is enforced by the constraints $\Cleq{\sigma}{\kaff_\infty}$.
+The $\Weaken$ constraint is introduce at each binding point in the program
+such as let-bindings and pattern matching.
+
+Note again here the use of normalization on complex constraints to ensure
+that the inference algorithm always return the simplest possible
+constraints and unifiers.
 
@@
 
-\subsubsection{Copying and Splitting}
+\subsubsection{Splitting and Regions}
 \label{infer:split}
+\label{infer:regions}
+
+As noted previously, in our inference algorithm, we infer
+usage environments which describes how variables are used. In
+\cref{sdtyping}, we described the {\sc Pair} and {\sc Region} rules, and their
+respective actions on the environment. We now see the inference version
+of these rules in \cref{rule:infer:envrules}.
+While the rules are quite similar to the original version, the usage
+environment $\Sv$ is now an \emph{output} of the algorithm.
+As such, we use the ``inference'' version of the relations on
+environment,
+$\bsplit{C}{\Sv}{\bf{\Sv_1}}{\bf{\Sv_2}}$ and $\bregion{C}{\bf{x}}{\Sv}{\bf\Sv'}$,
+which returns the necessary constraints and
+are total and non ambiguous for their input parameters ($\Sv_1, \Sv_2$ and $\Sv'$, respectively).
+We then simply collect all the constraints and normalize them.
 
@@
     \ruleIPair
+
+    \ruleIRegion
   \end{mathpar}
-  \caption{The {\sc Pair$_I$} rule}
-  \label{rule:infer:pair}
+  \caption{The {\sc Pair$_I$} and {\sc Region$_I$} rules}
+  \label{rule:infer:envrules}
 \end{figure}
 
-\subsubsection{Regions}
-\label{infer:regions}
+\subsubsection{Generalization and constraints}
+
+So far, constraints have only been composed of a list of kind inequalities.
+The internal language of constraints used for inference is in fact
+richer.
+To illustrate this, we now show the inference rule for let-bindings
+in \cref{rule:infer:let}.
+This rule combines several ingredients previously seen:
+since let expressions are binders, we use $\Weaken$ on the bound
+identifier $x$. Since let-expressions contain
+two subexpressions, so we use the environment splitting relation,
+$\bsplit{C_s}{\Sv}{\Sv_1}{(\Sv_2 \Sdel{x})}$. We carefully remove the $x$ from
+the left environment, since it is not available outside of the expression
+$e_2$, and thus should not be part of the returned usage environment.
+
+As per tradition in ML languages, generalization is performed
+on let-bindings.
+Following HM(X), we note $(C_\schm, \schm) = \generalize{C}{\E}{\tau}$
+the pair of a constraint and a scheme resulting from
+generalization. The definition is provided in \cref{rule:infer:let}.
+The type scheme $\schm$ is created by quantifying over all the appropriate
+free variables and the current constraints.
+The generated constraint $C_\schm$ uses a new projection operator,
+noted $\Cproj{\tvar}{D}$, which
+allow the creation of local variables inside the constraints.
+This allow to encapsulate all the quantified variables in the global constraints.
+It also reflect the fact that there
+should exists at least one solution for $C$ for the scheme to be valid.
+\citet{DBLP:journals/tapos/OderskySW99} give a detailed account
+on the projection operators in HM inference.
 
@@
   \begin{mathpar}
-    \ruleIRegion
+    \ruleILet
+
+    {\generalize{C}{\E}{\tau} =
+      (\Cproj{\Multi{\kvar},\Multi{\tvar}}{C},
+      \forall \Multi{\kvar},\Multi{\tvar}.\qual{C}{\tau})
+
+      \text{ where }
+
+      \Multi{\kvar},\Multi{\tvar} = (\fv{\tau}\cup\fv{C})\setminus\fv{\E}
+    }
   \end{mathpar}
-  \caption{The {\sc Region$_I$} rule}
-  \label{rule:infer:region}
+  \caption{The {\sc Let$_I$} rule}
+  \label{rule:infer:let}
 \end{figure}
@@
 \subsubsection{Constraint desugaring}
+
+\TODO{Present this if there is space}
 
@@
 
-\TODO{Present the full constraint system}
+\newcommand\A{\mathcal A}
+\newcommand\SC{\mathcal S}
 
-We now place ourselves in a more general context where kinds are either variables
+Before presenting our constraint solving algorithms,
+we place ourselves in a more general context where kinds are either variables
 or constants belonging to a total bounded lattice $(\mathcal L, \lk_\Lat)$ ie.,
@@
 %
-Let $\CL$ the set of constraints in this lattice. We consider the
-following base entailment rules in this system:
-\begin{center}
+Let $\CL$ the set of constraints in this lattice.
+The full grammar of constraints is shown in \cref{grammar:constraint}.
+Constraints are made of kind inequalities, conjunctions and
+projections, as shown previously, along with type unification
+constraints. Since types might contain kinds (for instance, on the arrows),
+type unification is oriented and noted $\leq$.
+
+As before, entailment is noted $\entail{C}{D}$,
+where $D$ is a consequence of the constraints $C$.
+We say that $C$ and $D$ are equivalent, noted $C \equivC D$,
+when $\entail{C}{D}$ and $\entail{D}{C}$.
+The base rules for entailment are given in \cref{rules:entail}.
+The rules for classic Herbrand unification and
+projection operators are omitted for brevity (see
+\citep{DBLP:journals/tapos/OderskySW99} for
+details).
+
+\begin{figure}[!h]
+  \centering
+  \begin{minipage}{0.32\linewidth}
+    \begin{align*}
+      C &::= \Cleq{\tau_1}{\tau_2} \\
+        &\mid \Cleq{k_1}{k_2} \tag{Inequalities} \\
+        &\mid C_1 \Cand C_2  \tag{Conjunction}\\
+        &\mid \Cproj{\tvar}{C} 
+        \mid \Cproj{\kvar}{C} \tag{Projection} 
+    \end{align*}
+    \caption{The constraint language}
+    \label{grammar:constraint}
+  \end{minipage}\hfill
+  \begin{minipage}{0.64\linewidth}
     \begin{mathpar}
-    \inferrule{l \lk_\Lat l'}{\entail{}{\Cleq{l}{l'}}}
-    \and
       \inferrule{}{\entail{}{\Cleq{k}{l^\top}}}
@@
       \inferrule{}{\entail{}{\Cleq{l^\bot}{k}}}
+      \and
+      \inferrule
+      {}{ \entail{}{\Cleq{\kvar}{\kvar}} }
+      \and
+      \inferrule
+      { \forall i,\ \entail{C}{\Ceq{\tau_i}{\tau_i}}\\
+      }
+      { \entail{C}{\Cleq{\tapp{t}{(\tau_i)}}{\tapp{t}{(\tau'_i)}}} }
+      \and
+      \inferrule{l \lk_\Lat l'}{\entail{}{\Cleq{l}{l'}}}
+      \and
+      \inferrule
+      { \entail{C}{\Cleq{\tau'_1}{\tau_1}}\\
+        \entail{C}{\Cleq{\tau_2}{\tau'_2}}\\
+        \entail{C}{\Cleq{k}{k'}}
+      }
+      { \entail{C}{\Cleq{\tau_1\tarr{k}\tau_2}{\tau'_1\tarr{k'}\tau'_2}} }
     \end{mathpar}
-\end{center}
+    \caption{Base entailment rules -- $\entail{C}{D}$ }
+    \label{rules:entail}
+  \end{minipage}
+\end{figure}
 
-The normalization procedure is noted $\normalize{C_0}{\unif_0}$ where
+We define the set of solved formed
+$\mathcal S$ as the set of constraints $\CL$ quantified by $\equivC$.
+We will show later that such constraints are in fact only composed of
+kind inequalities, and thus correspond to the syntactic constraints
+used in type and kind schemes.
+%
+We now define our normalization procedure, noted $\normalize{C_0}{\unif_0}$ where
 $C_0\in \CL$ is a set of constraints and $\unif_0$ is a substitution.
@@
 
-\TODO{Add an example}
+\TODO{Add example}
 
-\subsection{Soundness and Principality}
+This algorithm is complete and computes principal normal forms.
+It can be completed by a set of ``best-effort'' simplification
+rules which help reduce the size of inferred signatures
+\citep{DBLP:conf/popl/PottierS02,DBLP:conf/aplas/Simonet03}.
+In particular, our implementation and all the signatures presented in
+\cref{motivation} use a variance-based simplification
+where all covariants variables are replaced by their upper bounds
+and contra-variant variables are replaced by their lower bounds.
 
-Our system is an extension of HM(X). Our goal is prove soundness and completeness
-of our inference algorithm with respect to our syntax-directed system, which
-assures us principal type inference.
-We prove two properties:
-\begin{itemize}
-\item Our constraint system respects all the properties required by HM(X)
-  for principal type inference.
-\item Our extended inference algorithm still satisfies the soundness
-  and completeness properties of HM(X). This is achieved by extending
-  the original proofs.
-\end{itemize}
-
-This section only states the theorems (and the necessary definitions).
-Details and proofs are provided in \cref{appendix:constraints,appendix:infer}.
-
-\subsubsection{Constraint system}
-
-We requires three properties from our constraint solving algorithms: it should
-compute normal firms, such normal forms should be unique, and the
-system should be regular.
+We now show that this algorithms is well-behaved with respect to entailment
+and computes unique normal forms.
 
@@
 
-\subsubsection{Inference algorithm}
+\subsection{Soundness and Principality}
 
+We now show that our extended inference algorithm still satisfies the soundness
+and completeness properties of HM(X). This is achieved by extending
+the original proofs.
+%
 The first theorem states that our inference algorithm is sound

### 2019-07-08 kindly/infer/infer.tex
Reformulation/mv/fixes.

@@
 We now formulate our inference algorithm
-based on the HM(X) framework~\citep{DBLP:journals/tapos/OderskySW99}.
-HM(X) presents how to implement Hindley-Milner type inference in a language
-with qualified types and an arbitrary custom constraint language.
-We adapt and extend the rules of HM(X) to support kind inference,
-track linearity and handle borrows and regions. Furthermore, we
-formulate an appropriate constraint solving algorithm for our constraint
-system such that inference is sound and complete.
+based on the $HM(X)$ framework~\citep{DBLP:journals/tapos/OderskySW99}.
+$HM(X)$ presents a Hindley-Milner type system for a language
+with qualified types where constraints are expressed in an arbitrary
+constraint $X$. In particular, given some properties on the constraint system $X$,
+$HM(X)$ ensures principal type inference.
+We apply HM(X) to a concrete constraint language which we name $\CL$.
+We adapt and extend its existing rules to support kind inference,
+track linearity and handle borrows and regions. Finally, we
+formulate an appropriate constraint solving algorithm for $\CL$.
 In addition, our constraint solving algorithm simplifies constraints in a
@@
 
-\subsubsection{Constraint desugaring}
-
-\TODO{Present this if there is space}
-
-\begin{figure}[!htb]
-  \begin{mathpar}
-
-  \inferrule{}{ \Cleq{\Eempty}{k} \Crewrite  \Cempty}
-
-  \inferrule{
-    \Cleq\E k \Crewrite  C \\ \E \vdash \Cleq B k \Crewrite  D
-  }{
-    \Cleq{\E; B}{k} \Crewrite  C \Cand D
-  }
-
-  \inferrule{
-    \E \vdash \Cleq \schm k \Crewrite  C
-  }{ \E \vdash \Cleq{\bvar x \schm}{k} \Crewrite  C}
-
-  \inferrule{ }{
-    \E \vdash \Cleq{\svar x \schm^n} k \Crewrite  \Cfalse
-  }
-
-  % \inferrule{}{\Cleq{\IBORROW^n} k \Crewrite  \Cleq{\kun_n} k}
-
-  % \inferrule{}{\Cleq{\MBORROW^n} k \Crewrite  \Cleq{\kaff_n} k}
-
-    \inferrule{
-      \inferK{C \Cand C_x} \E \tau {k'} \\
-      D = C \Cand C_x \Cand \Cleq{k'}k
-    }{
-      \E \vdash
-      \Cleq{(\forall \kvar_i \forall (\tvar_j:k_j).\ \qual{C_x}{\tau})}{k} \Crewrite  D
-    }
-  \end{mathpar}
-
-  \caption{Rewriting constraints on environments and types}
-  \label{rule:typing:desugar}
-\end{figure}
-
 \subsection{Constraint solving}
@@
 
-Before presenting our constraint solving algorithms,
+Before presenting our constraint solving algorithm,
 we place ourselves in a more general context where kinds are either variables
@@
 
-This algorithm is complete and computes principal normal forms.
-It can be completed by a set of ``best-effort'' simplification
-rules which help reduce the size of inferred signatures
-\citep{DBLP:conf/popl/PottierS02,DBLP:conf/aplas/Simonet03}.
+Our algorithm is complete, computes principal normal forms,
+and already simplifies constraints significantly
+(notably thanks to steps 6, 7 and 8).
+It can be easily extended with further simplification phases.
 In particular, our implementation and all the signatures presented in
 \cref{motivation} use a variance-based simplification
-where all covariants variables are replaced by their upper bounds
-and contra-variant variables are replaced by their lower bounds.
-
+where all covariant (resp. contravariant) variables are replaced by their
+lower (resp. upper) bounds.
+Remarkably, all the simplification mechanisms presented
+here (including the variance-based one) are complete.
+It is also possible to add ``best-effort'' simplification
+rules which help reduce the size of inferred signatures even further
+\citep{DBLP:conf/aplas/Simonet03}.
+%
 We now show that this algorithms is well-behaved with respect to entailment

### 2019-07-08 kindly/soa.tex
abstract
more motivation
Start explanation for borrow and regions.
Start working on the state of the art.

--- 2018-11-06 kindly/soa.tex
@@
 
-\paragraph{Functional linear of affine type systems}
+Haller, P., Odersky, M.: Capabilities for Uniqueness and Borrowing
+\cite{DBLP:conf/ecoop/HallerO10}
 
-Alms~\citep{DBLP:conf/popl/TovP11},
-F$^\circ$~\citep{DBLP:conf/tldi/MazurakZZ10},
-Quill~\citep{DBLP:conf/icfp/Morris16},
-Linear Haskell~\citep{DBLP:journals/pacmpl/BernardyBNJS18}
+Sing\#
 
-\paragraph{Object systems}
+John Tang Boyland and William Retert. Connecting effects and
+uniqueness with adoption. In POPL, pages 283–295. ACM, 2005.
+\cite{DBLP:conf/popl/BoylandR05}
+
+Manuel Fahndrich and Robert DeLine. Adoption and focus: Practical
+linear types for imperative programming. In PLDI, pages 13–24, 2002.
+
+\subsubsection{Functional linear of affine type systems}
+
+Many systems have attempted to combine
+functional programming and linear types in a practical setting.
+One contribution of \lang is to combine several ingredients
+from these different languages and combines them, while still preserving
+complete type inference.
+In addition, none of the following language supports borrows.
+
+System F\degree~\citep{DBLP:conf/tldi/MazurakZZ10}
+extends System F with uses kinds to distinguish
+between linear and unrestricted types.
+\citet{DBLP:conf/tldi/MazurakZZ10} also provide
+(and proves soundness) of a linearity-aware semantics.
+Unlike \lang, System F\degree does not allow
+quantification over kinds which limits its expressivity. For instace, it
+does not admit a most general type for function composition.
+Furthermore, since it is based on System F, it does not admit
+principal type inference.
+
+Quill~\citep{DBLP:conf/icfp/Morris16} is an Haskell-like language with affine
+types through the use of qualified types.
+Quill does not exposes a kind language, but
+uses typeclass-like mechanism and annotation on arrows.
+It supports both a most general type for function composition and
+principal type inference.
+The combination of type and kind schemes
+and subkinding ocasionally yields significantly simpler
+type in \lang and we posit it strictly subsumes Quill's type system.
+Unlike Quill,
+we provide a linearity-aware semantics and prove its soundness.
+Quill does not support borrows.
+
+% For instance, the type of the constructor
+% in Quill is $\qual{t \geq f}{t \to u \to t * u}$.
+% In Affe, it is simply
+% $\qual{(\alpha:\kvar)\implies \alpha \to \beta \tarr{\kvar} \alpha * \beta$
+% with the kind of $*$ being $\kvar\to\kvar\to\kvar$.
+
+Alms~\citep{DBLP:conf/popl/TovP11} is an ML-like language with rich, kind-based
+affine types and supports abstraction through modules.
+It relies on meet and join operators instead of qualified types, but still
+uses subkinding.
+Their system often relies on existential types to track the identity
+of objects. For instance, they present an array creation
+functions of the type \lstinline/int -> 'a -> \E 'b. ('a, 'b) array/ where
+\lstinline/'a/ are the types of the elements in the array and \lstinline/'b/
+is a type variable that uniquely identifies the array. Unfortunately, due
+to the reliance on existential types, Alms does not support type inference.
+Furthermore, Alms does not support borrow and often rely
+on manual passing of capabilities.
+In our experience, \affe's limited support for existential types through
+regions is sufficient for most of their examples.
+%
+Apart from existential types, Alms' kind system is very complex and
+rely on union and intersection operators and dependent kinds. We believe
+such complexity is unnecessary and that any Alms' kind signature
+can be translated into \lang. For instance the pair type constructor
+has kind $\Pi\alpha\Pi\beta. \langle\alpha\rangle \sqcup \langle\beta\rangle$
+(where $\alpha$ and $\beta$ are types and $\Pi$ is the dependent function)
+in Alms, but simply $\kvar\to\kvar\to\kvar$ in \lang.
+
+\paragraph{Linear Haskell~\citep{DBLP:journals/pacmpl/BernardyBNJS18}}
+
+\paragraph{Vault\citep{DBLP:conf/pldi/DeLineF01,DBLP:conf/pldi/FahndrichD02}}
+
+\subsubsection{Borrows}
+
+Rust, Non Lexical lifetimes
+
+\subsubsection{Object systems}
 \TODO{}
+
+Plaid \&Co by Aldrich and others
+
+\subsubsection{Uniqueness}
+
+\subsubsection{Others}
+
+\paragraph{Mezzo}
 
@@
 
-\subsection{Subtyping and constraints}
+\subsection{Subtyping}
 

### 2019-07-09 kindly/intro.tex
Reformulation/mv/fixes.
implementation available

@@
 Most type systems provide just type soundness, but neglect the
+
 protocol aspect. The systems that do build on linear types \cite{DBLP:journals/tcs/Girard87} and/or
@@
 \item Automatic inference of regions for borrows.
+\item We implemented the type inference algorithm including all
+  constraint simplification and extended with pattern matching. A
+  webfront for the algorithm is available, its URL is given in the
+  non-anonymous supplementary material, and we plan to submit it for
+  artifact evaluation.
 \end{itemize}

### 2019-07-09 kindly/soa.tex

@@
 One contribution of \lang is to combine several ingredients
-from these different languages and combines them, while still preserving
+from these different languages while still preserving
 complete type inference.
-In addition, none of the following language supports borrows.
+Many of the following languages supports linear or affine types, but rarely
+both. In many cases, it is easy to adapt a system to support both, as
+\lang does.
+None of the following languages support borrows.
 
 System F\degree~\citep{DBLP:conf/tldi/MazurakZZ10}
-extends System F with uses kinds to distinguish
+extends System F with kinds to distinguish
 between linear and unrestricted types.
@@
 (and proves soundness) of a linearity-aware semantics.
-Unlike \lang, System F\degree does not allow
-quantification over kinds which limits its expressivity. For instace, it
+Unlike \lang, System F\degree{} does not allow
+quantification over kinds which limits its expressivity. For instance, it
 does not admit a most general type for function composition.
@@
 principal type inference.
-The combination of type and kind schemes
-and subkinding ocasionally yields significantly simpler
-type in \lang and we posit it strictly subsumes Quill's type system.
+We believe every Quill's kind signature can be expressed in \lang by leveraging
+kind inequalities.
 Unlike Quill,
@@
 Alms~\citep{DBLP:conf/popl/TovP11} is an ML-like language with rich, kind-based
-affine types and supports abstraction through modules.
-It relies on meet and join operators instead of qualified types, but still
-uses subkinding.
+affine types and ML modules.
 Their system often relies on existential types to track the identity
 of objects. For instance, they present an array creation
-functions of the type \lstinline/int -> 'a -> \E 'b. ('a, 'b) array/ where
-\lstinline/'a/ are the types of the elements in the array and \lstinline/'b/
-is a type variable that uniquely identifies the array. Unfortunately, due
-to the reliance on existential types, Alms does not support type inference.
+functions of type \lstinline/int -> 'a -> \E 'b. ('a, 'b) array/ where
+\lstinline/'a/ is the elements' type and \lstinline/'b/
+is a unique identifier of the array.
+Due to the reliance on existentials, Alms does not support complete type inference.
 Furthermore, Alms does not support borrow and often rely
@@
 In our experience, \affe's limited support for existential types through
-regions is sufficient for most of their examples.
+regions is sufficient for a large majority of examples and provides
+a more convenient programming style for imperative code.
 %
-Apart from existential types, Alms' kind system is very complex and
-rely on union and intersection operators and dependent kinds. We believe
-such complexity is unnecessary and that any Alms' kind signature
-can be translated into \lang. For instance the pair type constructor
+Apart from existential types, Alms relies
+on union and intersection operators and dependent kinds while
+\lang uses qualified types.
+We believe most of Alms' kind signatures can be expressed equivalently in
+our system: for instance the pair type constructor
 has kind $\Pi\alpha\Pi\beta. \langle\alpha\rangle \sqcup \langle\beta\rangle$
 (where $\alpha$ and $\beta$ are types and $\Pi$ is the dependent function)
-in Alms, but simply $\kvar\to\kvar\to\kvar$ in \lang.
+in Alms and $\kvar\to\kvar\to\kvar$ in \lang thanks
+to subkinding.
+%
+Finally, Alms provides excellent support for abstraction through
+modules by allowing to keep some type unrestricted inside a module, but
+exposing it as affine. We support a similar programming style, which
+we showcase in \cref{motivation}.
 
-\paragraph{Linear Haskell~\citep{DBLP:journals/pacmpl/BernardyBNJS18}}
+Linear Haskell~\citep{DBLP:journals/pacmpl/BernardyBNJS18} aims
+to retrofit Haskell with linear types.
+Linear Haskell introduces a new notion
+of ``linear arrows'', noted $\multimap$ similar to linear logic,
+which are functions who use their argument exactly once.
+By contrast, in \lang linearity is only decided
+by the kinds, and $\tarr{\klin}$ simply denotes \emph{single-use functions}.
+This design, while easy to retrofit on top of an existing compiler
+such as GHC, has several disadvantages,
+as highlighted by the long discussion in the GHC proposal\footnote{
+  See the in-depth discussion starting here: \url{https://github.com/ghc-proposals/ghc-proposals/pull/111\#issuecomment-403349707}}.
+Most relevant to \lang:
+\begin{itemize}
+\item Their system does not admit subtyping for arrows and leverages
+  $\eta$-expansion to allow passing unrestricted functions in linear
+  contexts. This is acceptable in a non-strict language such as
+  Haskell but can not be used in a strict setting.
+\item
+  While the GHC proposal implements \emph{some} type inference,
+  their algorithm is not formalized and not proved complete.
+\item
+  Linear Haskell promotes a ``bracket pattern'', similar to CPS-style such as 
+  \lstinline/withFile : path -> (file ->. file) -> unit/.
+  \lang follows System F\degree, Quill and Alms and
+  adopts a more direct style, where types themselves are
+  described as affine/linear.
+  We believe this leads to a more natural programming style
+  and makes reasoning about linearity in the context of abstraction much easier.
+% \item
+%   Due to the previous remark, ensuring that a given object is never
+%   aliased is fairly difficult in Linear Haskell, as one would need to ensure
+%   that no non-linear function can manipulate it. On the other
+\end{itemize}
 

### 2019-07-09 kindly/lang.tex
notation of regions

@@
 
-A region $\region{x}{e}$ is annotated with its nesting $n$
-and the variable $x$ that may be borrowed in $e$.
+A region $\region{\Sone x\BORROW}{e}$ is annotated with its nesting $n$, the variable $x$ that may be borrowed in $e$, and the kind of borrow $\BORROW$.
 
@@
 
-\paragraph{Previously}
+% \paragraph{Previously}
 
-Let us first define our conventions and meta-syntactic variables.
-Expressions, types and kinds are noted respectively $e$, $\tau$, and $k$.
-Variables, type variables and kind variables are noted
-respectively $x$, $\tvar$ and $\kvar$. Types and kind schemes
-are noted $\schm$ and $\kschm$.
-In the rest of the text, we will write lists with a bar and an index, for
-instance $\Multi[i]{\tau}$.
-Kind constants are noted in bold capital letters, such as $\kaff$ for affine.
+% Let us first define our conventions and meta-syntactic variables.
+% Expressions, types and kinds are noted respectively $e$, $\tau$, and $k$.
+% Variables, type variables and kind variables are noted
+% respectively $x$, $\tvar$ and $\kvar$. Types and kind schemes
+% are noted $\schm$ and $\kschm$.
+% In the rest of the text, we will write lists with a bar and an index, for
+% instance $\Multi[i]{\tau}$.
+% Kind constants are noted in bold capital letters, such as $\kaff$ for affine.
 
-The syntax of \lang is presented in \cref{grammar} and follows
-traditional ML languages with let bindings.
-Borrows, noted $\borrow{e}$, are annotated with a specification indicating
-if a borrow is immutable ($\IBORROW$) or mutable ($\MBORROW$).
-We also distinguish reborrows, which are noted $\reborrow{e}$.
-Match are indexed with a specification which can be either $\&^b$
-(which $b$ a borrow specification) or $id$, indicating if the match
-should operate on borrows or not.
-Regions, noted $\region{x}{e}$ are annotated with their nesting ($n$)
-and the variable whose borrow they enclose ($x$).
-%
-$\tapp{t}{\Multi{\tau}}$ denotes a type application where
-$\T t$ is an abstract type constructor.
-Arrow and borrow types are annotated with a kind, noted $\tau \tarr{k} \tau$
-and $\borrowty{k}{\tau}$ respectively.
-%
-Kinds can be either a kind variable $\kvar$, or constants
-(linear $\klin$, affine $\kaff$ or unrestricted $\kun$) indexed
-by a nesting level represented by a member of $N \cup \{\infty\}$.
-Type schemes and kind schemes are guarded by a set of constraints.
-%
-Finally, constraints are composed of a list of kind inequalities.
+% The syntax of \lang is presented in \cref{grammar} and follows
+% traditional ML languages with let bindings.
+% Borrows, noted $\borrow{e}$, are annotated with a specification indicating
+% if a borrow is immutable ($\IBORROW$) or mutable ($\MBORROW$).
+% We also distinguish reborrows, which are noted $\reborrow{e}$.
+% Match are indexed with a specification which can be either $\&^b$
+% (which $b$ a borrow specification) or $id$, indicating if the match
+% should operate on borrows or not.
+% Regions, noted $\region{\Sone x\BORROW}{e}$ are annotated with their nesting ($n$)
+% and the variable whose borrow they enclose ($x$).
+% %
+% $\tapp{t}{\Multi{\tau}}$ denotes a type application where
+% $\T t$ is an abstract type constructor.
+% Arrow and borrow types are annotated with a kind, noted $\tau \tarr{k} \tau$
+% and $\borrowty{k}{\tau}$ respectively.
+% %
+% Kinds can be either a kind variable $\kvar$, or constants
+% (linear $\klin$, affine $\kaff$ or unrestricted $\kun$) indexed
+% by a nesting level represented by a member of $N \cup \{\infty\}$.
+% Type schemes and kind schemes are guarded by a set of constraints.
+% %
+% Finally, constraints are composed of a list of kind inequalities.
 
@@
 $\letin{a}{\text{create}\ 10\ x}
-{\introPair{\region[{}]{x}{\app{\text{length}}{\borrow[\IBORROW]{a}}}}{a}}$,
+{\introPair{\region[{}]{\Sone x\IBORROW}{\app{\text{length}}{\borrow[\IBORROW]{a}}}}{a}}$,
 we use the rule {\sc Susp} to detach a borrow on the left of the pair.
@@
 Such borrows are constrained to a given syntactic region. We note
-$\region{x}{e}$ the region which allow borrows of $x$ to be used in $e$
+$\region{\Sone x\BORROW}{e}$ the region which allow borrows of $x$ to be used in $e$
 and where $n$ is the nesting of the region.
@@
 
-We illustrate this on the expression $\region{x}{f(\borrow[\IBORROW]{c})}$
+We illustrate this on the expression $\region{\Sone x\IBORROW}{f(\borrow[\IBORROW]{c})}$
 when $c$ is a linear channel.

### 2019-07-09 kindly/soa.tex
SOA.

@@
 by the kinds, and $\tarr{\klin}$ simply denotes \emph{single-use functions}.
-This design, while easy to retrofit on top of an existing compiler
-such as GHC, has several disadvantages,
-as highlighted by the long discussion in the GHC proposal\footnote{
-  See the in-depth discussion starting here: \url{https://github.com/ghc-proposals/ghc-proposals/pull/111\#issuecomment-403349707}}.
+This design is easy to retrofit on top of an existing compiler
+such as GHC, but has several disadvantages\footnote{
+  See the in-depth discussion attached to the GHC proposal starting here: \url{https://github.com/ghc-proposals/ghc-proposals/pull/111\#issuecomment-403349707}}.
 Most relevant to \lang:
@@
 \item
-  While the GHC proposal implements \emph{some} type inference,
+  While the GHC proposal implements some type inference,
   their algorithm is not formalized and not proved complete.
 \item
-  Linear Haskell promotes a ``bracket pattern'', similar to CPS-style such as 
-  \lstinline/withFile : path -> (file ->. file) -> unit/.
+  Linear Haskell promotes a CPS-style such as 
+  \lstinline/withFile : path -> (file ->. file) ->. unit/
+  to ensure that files are only used linearly.
   \lang follows System F\degree, Quill and Alms and
-  adopts a more direct style, where types themselves are
+  adopts a direct style, where types themselves are
   described as affine/linear.
   We believe this leads to a more natural programming style
-  and makes reasoning about linearity in the context of abstraction much easier.
+  and ease modular reasoning about linearity.
+  In particular, using abstraction through modules,
+  programmers only need consider the module
+  implementation to ensure that some linear
+  resource is properly handled.
+\item
+  Linear Haskell introduces a notion of linear monad to express
+  imperative code conveniently. Again, this solution is suitable in Haskell,
+  but less appropriate in other contexts. Our borrow system allow
+  to provide a more free-form programming style for imperative code.
 % \item

### 2019-07-09 kindly/soa.tex
SOA²

@@
 
-Manuel Fahndrich and Robert DeLine. Adoption and focus: Practical
-linear types for imperative programming. In PLDI, pages 13–24, 2002.
-
-\subsubsection{Functional linear of affine type systems}
+\subsubsection{Functional languages}
 
@@
 which are functions who use their argument exactly once.
-By contrast, in \lang linearity is only decided
-by the kinds, and $\tarr{\klin}$ simply denotes \emph{single-use functions}.
+% By contrast, in \lang linearity is only decided
+% by the kinds, and $\tarr{\klin}$ simply denotes \emph{single-use functions}.
 This design is easy to retrofit on top of an existing compiler
@@
   While the GHC proposal implements some type inference,
-  their algorithm is not formalized and not proved complete.
+  their algorithm is not formalized and is not complete.
 \item
-  Linear Haskell promotes a CPS-style such as 
+  Linear Haskell promotes a CPS style with functions such as 
   \lstinline/withFile : path -> (file ->. file) ->. unit/
@@
   adopts a direct style, where types themselves are
-  described as affine/linear.
+  described as affine or linear.
   We believe this leads to a more natural programming style
@@
   resource is properly handled.
-\item
-  Linear Haskell introduces a notion of linear monad to express
-  imperative code conveniently. Again, this solution is suitable in Haskell,
-  but less appropriate in other contexts. Our borrow system allow
-  to provide a more free-form programming style for imperative code.
+% \item
+%   Linear Haskell introduces a notion of linear monad to express
+%   imperative code conveniently. Again, this solution is suitable in Haskell,
+%   but less appropriate in other contexts. Our borrow system allow
+%   to provide a more free-form programming style for imperative code.
 % \item
@@
 
-\paragraph{Vault\citep{DBLP:conf/pldi/DeLineF01,DBLP:conf/pldi/FahndrichD02}}
+Mezzo~\citep{DBLP:phd/hal/Protzenko14} is an ML-like language
+with an extremely rich capability system which allows it to encode numerous
+properties akin to separation logic~\citep{DBLP:conf/lics/Reynolds02}.
+The main contribution of Mezzo is to push the boundary of what is possible
+to do with a type system. This makes it far more expressive than \lang, but
+also quite harder to use, and precludes any form of complete type
+inference.
 
-\subsubsection{Borrows}
+\subsubsection{Other systems}
 
-Rust, Non Lexical lifetimes
+Vault~\citep{DBLP:conf/pldi/DeLineF01}
+is a low-level language which uses typestate along with \emph{keys}
+to track ownership. It also introduces a novel paradigm
+called \emph{adoption and focus} \citep{DBLP:conf/pldi/FahndrichD02}
+which allow restricting the scope of linear objects
+(adoption) or temporarily treating them as non-linear (focus).
+These mechanisms serve a similar purpose as borrows, but are tracked
+much more precisely thanks to the use of typestate and pre/post-conditions,
+which makes it very expressive (at the cost of some verbosity).
+However, Vault is mostly geared towards traditional imperative programming:
+functions can't capture linear objects and it does not support
+type inference.
 
-\subsubsection{Object systems}
-\TODO{}
+Rust~\citep{rust} is the first
+mainstream language to popularize the idea of borrowing and ownership
+for safe low-level programming.
+\lang heavily draws inspiration from Rust's borrowing system to apply
+it in a functional setting with type inference, a GC, and
+an ML module system.
+Rust's lifetime system is more explicit and more expressive than \lang,
+however Rust does not provide type inference
+and only provides \emph{partial} lifetime inference.
+Recently, \citet{DBLP:journals/corr/abs-1903-00982} presented Oxide,
+a core language for Rust which formalizes its ownership discipline
+including non-lexical lifetimes, which we look forward to use to improve
+our borrowing system.
 
@@
 
-\subsubsection{Uniqueness}
+\subsection{Type-system features}
 
-\subsubsection{Others}
+\lang strongly relies on qualified types~\cite{DBLP:journals/scp/Jones94}
+to introduce the necessary kind equalities in our system. In
+particular, we use HM(X)~\citep{DBLP:journals/tapos/OderskySW99} which
+allow to use qualified type in an ML-like language with complete
+type inference.
+HM(X) has been shown to be compatible with subtyping,
+bounded universal and existentials~\citep{DBLP:conf/icfp/Simonet03}
+GADTs~\citep{DBLP:journals/toplas/SimonetP07},
+and syntactic proof of soundness~\citep{DBLP:journals/entcs/SkalkaP02}.
+This makes us confident that the system developed in \lang
+could be applied to larger and more complex languages such as OCaml.
 
-\paragraph{Mezzo}
-
-\subsection{Qualified Types}
-
-HM(X)~\citep{DBLP:journals/tapos/OderskySW99} with extensions to
-bounded universal and existentials~\citep{DBLP:conf/icfp/Simonet03}
-and GADTs~\citep{DBLP:journals/toplas/SimonetP07},
-and syntactic proof of soundness~\citep{DBLP:journals/entcs/SkalkaP02}.
-
-\subsection{Subtyping}
-
-Flowcaml~\citep{DBLP:conf/popl/PottierS02},
-\citet{DBLP:conf/sas/TrifonovS96}.
+The subtyping discipline we develop in \lang is very similar
+to structural subtyping, where the only subtyping (or here, subkinding)
+is at the leafs.
+Such discipline is known to be friendly to inference and has be used in many
+contexts, including OCaml itself, and has been combined
+with constraints~\citet{DBLP:journals/tapos/OderskySW99,DBLP:conf/sas/TrifonovS96}.
+In particular, Flowcaml~\citep{DBLP:conf/popl/PottierS02}
+extends OCaml with security levels forming a lattice and supports type inference.
+In \lang, we repurpose these ideas to check for linearity and affinity.
+\citet{DBLP:conf/aplas/Simonet03} also presents various algorithmic
+notions to solve lattice-based constraints, which we partially use
+in our constraint solving algorithm. Our main contribution
+is the design of a kind language specialized for checking linearity
+and sufficient simple to make
+all simplification rules complete, which allows to keep our type signature simple.
 

### 2019-07-09 kindly/soa.tex
More SOA.

@@
 \section{Related work}
-
-\subsection{Substructural type systems}
 
@@
 
-\subsubsection{Functional languages}
+\subsection{Substructural type-system in functional languages}
 
@@
 
-\subsubsection{Other systems}
+\subsection{Other substructural systems}
 
-Vault~\citep{DBLP:conf/pldi/DeLineF01}
-is a low-level language which uses typestate along with \emph{keys}
-to track ownership. It also introduces a novel paradigm
-called \emph{adoption and focus} \citep{DBLP:conf/pldi/FahndrichD02}
-which allow restricting the scope of linear objects
-(adoption) or temporarily treating them as non-linear (focus).
-These mechanisms serve a similar purpose as borrows, but are tracked
-much more precisely thanks to the use of typestate and pre/post-conditions,
-which makes it very expressive (at the cost of some verbosity).
-However, Vault is mostly geared towards traditional imperative programming:
-functions can't capture linear objects and it does not support
-type inference.
+Linear and affine discipline have also been used in non-functional
+settings, notably in the context of low-level imperative programming
+and object systems. In particular, \lang uses borrows and regions
+which were initially developed in objects or
+imperatives settings~\citep{DBLP:conf/popl/BoylandR05,DBLP:conf/pldi/GrossmanMJHWC02}.
 
@@
 
+Vault~\citep{DBLP:conf/pldi/DeLineF01}
+is a low-level language which uses typestate along with \emph{keys}
+to track ownership. It also introduces a novel paradigm
+called \emph{adoption and focus} \citep{DBLP:conf/pldi/FahndrichD02}
+which allow restricting the scope of linear objects
+(adoption) or temporarily treating them as non-linear (focus).
+These mechanisms serve a similar purpose as borrows, but are tracked
+much more precisely thanks to the use of typestate and pre/post-conditions,
+which makes it very expressive (at the cost of some verbosity).
+However, Vault is mostly geared towards traditional imperative programming:
+functions can't capture linear objects and it does not support
+type inference.
+
 Plaid \&Co by Aldrich and others
@@
 
-\lang strongly relies on qualified types~\cite{DBLP:journals/scp/Jones94}
-to introduce the necessary kind equalities in our system. In
+\lang relies on qualified types~\cite{DBLP:journals/scp/Jones94}
+to introduce the necessary kind equalities for linear types. In
 particular, we use HM(X)~\citep{DBLP:journals/tapos/OderskySW99} which
@@
 
-The subtyping discipline we develop in \lang is very similar
+The subtyping discipline used in \lang is very similar
 to structural subtyping, where the only subtyping (or here, subkinding)
@@
 contexts, including OCaml itself, and has been combined
-with constraints~\citet{DBLP:journals/tapos/OderskySW99,DBLP:conf/sas/TrifonovS96}.
+with constraints~\citep{DBLP:journals/tapos/OderskySW99,DBLP:conf/sas/TrifonovS96}.
 In particular, Flowcaml~\citep{DBLP:conf/popl/PottierS02}
@@
 and sufficient simple to make
-all simplification rules complete, which allows to keep our type signature simple.
+all simplification rules complete, which allows to keep type signatures simple.
 

### 2019-07-10 kindly/lang.tex
Fix region annotation and make sure borrow capture is allowed.

@@
 then the following constraints should hold: $\Cleq{k'}{k}$.
+Borrow bindings follow the exact same rules, but suspended bindings
+are forbidden in such constraints.
 While this definition is incomplete, it is sufficient to understand
@@
 the kind on the arrow is at least affine.
+Capturing a borrow is perfectly allowed here: the kind of the borrow
+will be reflected on the arrow which will restrict the closure
+to the region of the borrow. Capturing a \emph{suspended} binding, however,
+is forbidden.
 Note that $x$ is \emph{not} included in $\E$: since it's the argument of the
@@
     \centering
-    \begin{tabular}
-      {@{}>{$}r<{$}@{ $\vdash_e$ }
-      >{$}c<{$}@{ $\rightsquigarrow_n^{x}$ }
-      >{$}l<{$}
-      r}
+    % \begin{tabular}
+    %   {@{}>{$}r<{$}@{ $\vdash_e$ }
+    %   >{$}c<{$}@{ $\rightsquigarrow_n^{x}$ }
+    %   >{$}l<{$}
+    %   r}
 
-      {(\kun_n\lk k\lk\kun_\infty)}
-      &{\svar[\IBORROW]{x}{\tau}^n}
-      &{\bvar{\borrow[\IBORROW]{x}}{\borrowty[\IBORROW] k{\tau}}}
-      &Immut\\
+    %   (\kun_n\lk k\lk\kun_\infty)
+    %   &{\svar[\IBORROW]{x}{\tau}^n}
+    %   &{\bvar{\borrow[\IBORROW]{x}}{\borrowty[\IBORROW] k{\tau}}}
+    %   &Immut\\
 
-      (\kaff_n\lk k\lk\kaff_\infty)
-      &\svar[\MBORROW]{x}{\tau}^n
-      &\bvar{\borrow[\MBORROW]{x}}{\borrowty[\MBORROW] k{\tau}}
-      &Mut
-    \end{tabular}
+    %   (\kaff_n\lk k\lk\kaff_\infty)
+    %   &\svar[\MBORROW]{x}{\tau}^n
+    %   &\bvar{\borrow[\MBORROW]{x}}{\borrowty[\MBORROW] k{\tau}}
+    %   &Mut
+    % \end{tabular}
+    \begin{mathpar}
+      \inferrule{
+        \entail{C}{(\BORROW_n\lk k\lk \BORROW_\infty)}\\
+        \BORROW\in\left\{\kun;\kaff\right\}
+      }{
+        \lregion{C}{}
+        {\svar[\BORROW]{x}{\tau}^n}
+        {\bvar{\borrow[\BORROW]{x}}{\borrowty[\BORROW] k{\tau}}}
+      }
+    \end{mathpar}
     \caption{Borrowing rules for bindings -- $\lregion{C}{x}{\E}{\E'}$}

### 2019-07-10 kindly/sdtyping/extra.tex
Inference and constraints
Reformulation/mv/fixes.
notation of regions
Fix region annotation and make sure borrow capture is allowed.
Misc fixes.

@@
 First, we extend the region annotation to $\region{S}{E}$ where $S$ is
-a set of variables. This annotation, defined below, is equivalent to nested
+a map from variables to borrow indicator $b$. This annotation, defined below, is equivalent to nested
 region annotations for each individual variable.
@@
 \begin{align*}
-  \region{x;S}{e} &= \region{x}{\region{S}{e}}& \region{\emptyset}{e} &= e\\
+  \region{\Sone x\BORROW;S}{e} &= \region{\Sone x\BORROW}{\region{S}{e}}& \region{\emptyset}{e} &= e\\
 \end{align*}
@@
 \Cref{fig:region-annotation} define a rewriting relation $\RannotT{e}{e'}$
-which indicates that an optionally annotated term $e$ can be written
-in a fully annotated term $e'$.
+which indicates that an optionally annotated term $e$ can be rewritten
+as a fully annotated term $e'$.
 Through the rule \textsc{Rewrite-Top}, this is defined
@@
 $\Rannot{e}{e'}{S}$ where $n$ is the current nesting and $S$ is a set of
-variable that are not enclosed in a region yet.
+variable that are not yet enclosed in a region.
 The base cases are constants, variables and borrows.
@@
 
-\section{Syntax directed typing}
-\label{appendix:sdtyping}
-
-\subsection{Constraint language}
+\section{Constraints}
 \label{appendix:constraints}
 
-\newcommand\A{\mathcal A}
-\newcommand\SC{\mathcal S}
+% \newcommand\A{\mathcal A}
+% \newcommand\SC{\mathcal S}
 
@@
 \begin{proof}
-  By definition of $\operatorname{normalize}$, We have $C = \Cleq{k}{k'}_i$
+  By definition of $\operatorname{normalize}$, We have $C = \Multi{\Cleq{k}{k'}}$
   such that the underlying relation has no cycles.
@@
 linear and affine types, and borrow.
+
+\section{Syntax directed typing}
+\label{appendix:sdtyping}
 
@@
   \input{sdtyping/sd-borrowing}
-  \caption{Borrowing --- environments $\lsplit
-    C\E\E\E$; binders $\bsplit Cbbb$}
+  \caption{Borrowing --- environments $\lregion
+    C{x}\E\E$; binders $\bregion C{x}bb$}
   \label{fig:sd-borrowing}

### 2019-07-10 kindly/motivation.tex
Misc fixes.
Make the sudoku example cooler.

@@
 \section{Linearity, Affinity, and Borrows at Work}
-
-\TODO{really good examples. first linearity/affinity (files?), then
-  something with borrows (functional arrays), and then something
-  really spaced out that pushes the limit}
+\label{motivation}
 
@@
 \end{lstlisting}
-    \caption{Example of use of affine arrays}
+    \caption{Example of use of linear arrays}
     \label{ex:array}
@@
 particular branch of the choice.
-In particular, the \lstinline/try_solution/ function can be used in a loop
-precisely because the \lstinline/set/ function only requires an
-unrestricted immutable borrow of \lstinline/g/, which can be used as
-many times as needed.
-This would also easily allow parallel iteration on \cref{line:iter}.
-As \lstinline/g/ is linear we must free it in \cref{line:free:g}.
+In particular,
+since the \lstinline/set/ function only requires an unrestricted immutable borrow,
+\lstinline/try_solution/ can captures the borrow \lstinline/&g/
+while staying unrestricted.
+This also make \lstinline/try_solution/ itself bound to the same
+region as \lstinline/&g/. In this example, our region inference
+will place the region exactly at the \lstinline/begin//\lstinline/end/
+scope.
+This allows us to use \lstinline/try_solution/ in
+the iteration loop on \cref{line:iter}.
+Since \lstinline/&g/ is unrestricted, we could even use a parallel iteration
+instead of a sequential one.
+As \lstinline/g/ is linear we must free it before returning,
+in \cref{line:free:g}, outside of the region.
 
@@
   \begin{minipage}[t]{0.54\linewidth}
-    \lstinputlisting[linerange=15-30,firstnumber=last]{code/sukodu.affe}
+    \lstinputlisting[linerange=15-31,firstnumber=last]{code/sukodu.affe}
   \end{minipage}

### 2019-07-10 kindly/lang.tex
kinds

@@
 \begin{itemize}
-\item We showcase pattern matching on pairs rather than algebraic
+\item Pattern matching is demonstrated on pairs rather than algebraic
   datatypes.
-\item We have separate operators for borrowing and reborrowing (taking
+\item There are separate operators for borrowing and reborrowing (taking
   the borrow of a borrow).
-\item We assume that regions are explicit in the code. (But we present
+\item Regions are assumed to be explicit in the code. (But we present
   an algorithm to infer them in \cref{regionannot}.)
-\item We use de Bruijn-indexing to identify regions instead of region
+\item Regions are identified using de Bruijn-indexing instead of region
   variables. 
@@
 the statics in terms of syntax directed typing (\cref{sdtyping}),
-and a dynamics that is linearity-aware (\cref{sem}).
+and a dynamics that is linearity- and resource-aware (\cref{sem}).
 
@@
 
-We now present the essential parts of our \lang's type system. A complete
-description of the system is available in \cref{appendix:sdtyping}.
-In this section, we focus on chosen parts of our system
-through a set of selected rules.
+In this section, we focus on the essential and novel parts of the type system. A complete
+description is available in \cref{appendix:sdtyping}.
+% In this section, we focus on chosen parts of our system
+% through a set of selected rules.
 %
-We introduce the following typing, kinding and constraint judgements:
+We discuss judgments for typing, kinding, and constraint entailment:
 \begin{description}
@@
   Type $\tau$ has kind $k$ in environment $\E$ under constraints $C$.
-\item[$\entail{D}{C}$ --] Constraint $D$ implies constraint $C$.
+\item[$\entail{D}{C}$ --] Constraint $D$ entails constraint $C$.
 \end{description}
@@
 
-One novel feature or \affe is the use of kinds and qualified types to encode
-linear and affine types. Before explaining the details of our type system, we
-give a quick presentation of the kinds and constraints.
-A kind $k$ can be either a kind variable, $\kvar$, or a constant.
+\affe uses of kinds and constrained types to indicate
+linear and affine types.
+% Before explaining the details of our type system, we
+% give a quick presentation of the kinds and constraints.
+A kind $k$ is either a kind variable, $\kvar$, or a constant.
 Each constant is composed of a ``quality'' and a ``level''.
-The quality describes the use pattern of the type and can be
-either unrestricted $\kun$, affine $\kaff$ or linear $\klin$. The level
-describes the nested regions in which it can be used. Levels are represented
-by a natural number of $\infty$.
-$0$ represents the normal scope outside of any region and is often omitted
-(for instance $\kaff_0$ is simply noted $\kaff$).
-$\infty$ is the empty region that is infinitely nested and is mostly used
-as an upper bound.
+The quality describes the use pattern of the type. It can be
+unrestricted $\kun$, affine $\kaff$, or linear $\klin$. The level  $n \in \Nat \cup \{\infty\}$
+describes the nested regions in which the value can be used.
+Level $0$ refers to the top-level scope outside any region; we often elide it
+and write $\kaff$ for $\kaff_0$. Level
+$\infty$ refers to an empty region that is infinitely nested.
 For instance, the constraint $\Cleq{\kvar}{\kun_\infty}$ indicates that
 $\kvar$ must be unrestricted, but can be local to a region.
-%
-Just as presented before, constraints present in schemes and type signatures
-are simply a list of atomic constraints, which are composed
-of kind inequalities.
-%
-Our inequality constraints are based on an inequality relation on
-kind constants noted $\lk_\Lat$. This relation forms a lattice
-and  is summarized in
-\cref{sdtyp:lattice}. We have the basic inequalities on qualities:
-$\kun_n \lk_\Lat \kaff_n \lk_\Lat \klin_n$ for arbitrary regions $n$.
-Additionally, the inequality follows the region nesting, for
-instance $\kaff_n \lk_\Lat \kaff_{n+1}$.
 
-\begin{figure*}[!h]
+Kinds form a lattice with an ordering that implements two principles
+(see \cref{sdtyp:lattice}). First, unrestricted values may be used
+where affine ones are expected and affine ones are less restricted
+than linear one as reflected in \TirName{Lat-UA} and \TirName{Lat-AL}.
+Second, values usable at level $n$ may be used at any more deeply
+nested level $n'$ (where $n\le\infty$ for any $n$) as defined in the \TirName{Lat-X-Level} axioms.
+
+Constraints are conjunctions of inequality constraints over this kind
+lattice. In most uses, a constraint gives an upper or lower bound for
+a kind variable or it relates two kind variables.
+
+\begin{figure*}[tp]
   \input{infer/lattice}
-  \caption{Lattice inequalities -- $k \lk_\Lat k'$}
+  \caption{Lattice ordering -- $k \lk_\Lat k'$}
   \label{sdtyp:lattice}

### 2019-07-10 kindly/lang.tex
type environments

@@
 \label{sdtyping:envs}
-
-\lang aims to control the usage of variables, either
-through
-linearity and affinity which restrict how variables are shared and ignored,
-or through borrows which allow to circumvent the linearity rules
-in a controlled way.
-In order to check these properties, we introduce several new bindings
-in our typing environments.
-The grammar of type environment (noted $\E$) is shown in \cref{grammar:env}.
-On top of the traditional type and value bindings, noted respectively
-$\bvar{\tvar}{\kschm}$ and $\bvar{x}{\schm}$, we also consider two new
-bindings.
-Borrow bindings $\bvar{\borrow{x}}{\borrowty{k}{\tau}}$ indicates
-that the borrow $\borrow{x}$ can be used directly, as implemented
-by the rule {\sc Borrow} in \cref{selectrules:borrow}.
-Suspended bindings, noted $\svar{x}{\schm}^n$, indicates that $x$
-cannot be used immediately, but will be usable in a region
-marked with $x$, which we detail in \cref{sdtyping:regions}.
-
-Environments are also used to encode several linearity-related properties
-through \emph{environment-wide constraints}.
-Here we present how to control both unused variables and
-closure captures through such environment constraints, noted $\Cleq{\E}{k}$.
-The intuitive meaning of $\Cleq{\E}{k}$ is that for all bindings
-$\bvar{x}{\tau}$ in $\E$,
-if type $\tau$ has kind $k'$ (noted $\inferSK{C}{\E}{\tau}{k'}$)
-then the following constraints should hold: $\Cleq{k'}{k}$.
-Borrow bindings follow the exact same rules, but suspended bindings
-are forbidden in such constraints.
-While this definition is incomplete, it is sufficient to understand
-its use in the {\sc Abs} and {\sc Var} rules shown in \cref{selectrules:binders}.
-%
-The {\sc Var} rule looks up the type scheme of the variable $x$ in
-the environment $\E$
-and instantiate a concrete type through the substitution $\unif$. It then
-checks that the other bindings in $\E$ can safely be discarded by
-ensuring they are not linear. This is achieved through the constraint
-$\Cleq{\E\Sdel{x}}{\kaff_\infty}$ which replaces the traditional
-``weakening'' rule found in many type systems.
-
-The {\sc Abs} rule must ensures that if the body of the lambda captures
-linear or affine variables, the content will be reflected in the kind
-annotated on the arrow, such as $\tau_2 \tarr{k} \tau_1$.
-In order to do so, we constraint all the bindings of $\E$ to be less than $k$.
-If one of the binding of $\E$ contains a type whose kind is affine,
-we will derive the constraints $\Cleq{\kaff}{k}$, which ensures that
-the kind on the arrow is at least affine.
-Capturing a borrow is perfectly allowed here: the kind of the borrow
-will be reflected on the arrow which will restrict the closure
-to the region of the borrow. Capturing a \emph{suspended} binding, however,
-is forbidden.
-Note that $x$ is \emph{not} included in $\E$: since it's the argument of the
-closure, its use should not induce a capture.
-
-\begin{figure}[!h]
+\begin{figure}[tp]
   \begin{minipage}{0.33\linewidth}
@@
 \end{figure}
+
+\lang controls the use of variables by supporting new modes of
+binding in a type environment $\E$.
+Its definition is given in  \cref{grammar:env}.
+In addition to standard bindings of type variables to kind schemes,
+$\bvar{\tvar}{\kschm}$,  and value bindings, $\bvar{x}{\schm}$, there
+are suspended bindings and borrow bindings.
+A suspended binding, $\svar{x}{\schm}^n$, indicates that $x$ is
+earmarked for a borrowing use in a nested region
+marked with $x$ (see \cref{sdtyping:regions}), but
+cannot be used directly.
+A borrow binding, $\bvar{\borrow{x}}{\borrowty{k}{\schm}}$, replaces
+such a suspended binding on entry to the $x$-region. It indicates 
+that the borrow $\borrow{x}$ can be used directly. The kind $k$
+restricts the lifetime of the borrow to the region (see \cref{selectrules:borrow}).
+
+% , either
+% through
+% linearity and affinity which restrict how variables are shared and ignored,
+% or through borrows which allow to circumvent the linearity rules
+% in a controlled way.
+% In order to check these properties, we introduce several new bindings
+% in our typing environments.
+% The grammar of type environment (noted $\E$) is shown in \cref{grammar:env}.
+
+% Environments are also used to encode several linearity-related properties
+% through \emph{environment-wide constraints}.
+Constraints on an environment control substructural properties by
+restricting the types of variables.  The constraint $\Cleq{\E}{k}$
+abbreviates $\Cleq\schm k$, for all $\bvar{x}{\schm}$
+% $\svar{x}\schm$, and $\bvar{\borrow{x}}{\borrowty{k'}{\schm}}$
+in $\E$, which in turn means that $k' \lk k$ where $k'$ is the kind of
+the variable's type scheme.
+Borrow bindings follow the exact same rules, but suspended bindings
+are forbidden in an environment $\E$ constrained like that.
+This intuitive explanation is sufficient to understand
+the {\sc Abs} and {\sc Var} rules shown in
+\cref{selectrules:binders}.
+
+The {\sc Var} rule looks up the type scheme of the variable $x$ in
+the environment $\E$
+and instantiates it using the substitution $\unif$. The rule also
+checks that the other bindings in $\E$ can be safely discarded by
+imposing the constraint $\Cleq{\E\Sdel{x}}{\kaff_\infty}$. 
+It enforces that all remaining bindings (except $x$) are affine or
+unrestricted and can therefore be discarded.
+
+The {\sc Abs} rule ensures that the kind annotation on the arrow type
+($\tau_2 \tarr{k} \tau_1$) reflects the restrictions on the captured variables.
+To do so, the rule imposes the constraint $\Cleq\E k$.
+If, for instance, any binding in $\E$ is affine, it gives rise to the
+constraint $\Cleq{\kaff_n}{k}$, so that  the arrow kind is at least
+affine at nesting level $n$.
+Capturing a borrow is perfectly fine: the kind of the borrow is also a
+lower bound of the arrow kind $k$ which restricts the closure
+to the region of the borrow.
+Capturing a suspended binding is forbidden.
+% Note that $x$ is \emph{not} included in $\E$: since it's the argument of the
+% closure, its use should not induce a capture.
 

### 2019-07-10 kindly/infer/extra.tex
Add the proof for completeness.

@@
-\section{Inference rules}
+\section{Type Inference}
 \label{appendix:infer}
 
-\begin{figure*}[htp]
+\begin{figure*}[p]
   \input{infer/typing}
-  \caption{Type inference algorithm -- $\inferW{\Sigma}{(C,\psi)}{\bf{\E}}{\bf{e}}{\tau}$ }
-  \label{rules:typing}
+  \caption{Type inference rules -- $\inferW{\Sigma}{(C,\psi)}{\bf{\E}}{\bf{e}}{\tau}$ }
+  \label{rules:typing:full}
 \end{figure*}
 
-\begin{figure}[tp]
-  \centering
-  \begin{mathpar}
-    { \Weaken_{\bvar{x}{\sigma}}(\Sv) =
-      \begin{dcases}
-        \Ctrue& \text{if } x\in\Sv\\
-        \Cleq{\sigma}{\kaff_\infty}&\\
-      \end{dcases}
-    }
+In this appendix, we provide the complete type inference rules
+and show that our type inference algorithm is sound and complete.
+The kinding and constraints rules are already shown in \cref{inference}.
+The type inference rules are shown in \cref{rules:typing:full}.
+The various theorems and their proofs are direct adaptations
+of the equivalent statements for HM(X) \citep{sulzmann1997proofs}.
 
-    \inferrule{
-      \inferK{C \Cand C_x} \E \tau {k'} \\
-      D = C \Cand C_x \Cand \Cleq{k'}k
-    }{
-      \E \vdash \Cleq{\forall \kvar_i \forall (\tvar_j:k_j).\ \qual{C_x}{\tau}}{k} \Crewrite  D
-    }
-  \end{mathpar}
-  \caption{Auxiliary rules for inference}
-  \label{op:usgmap}
-\end{figure}
+\clearpage
+% \begin{figure}[tp]
+%   \centering
+%   \begin{mathpar}
+
+%     % \inferrule{
+%     %   \inferK{C \Cand C_x} \E \tau {k'} \\
+%     %   D = C \Cand C_x \Cand \Cleq{k'}k
+%     % }{
+%     %   \E \vdash \Cleq{\forall \kvar_i \forall (\tvar_j:k_j).\ \qual{C_x}{\tau}}{k} \Crewrite  D
+%     % }
+%   \end{mathpar}
+%   \caption{Auxiliary rules for inference}
+%   \label{op:usgmap}
+% \end{figure}
+
+\input{infer/metatheory}
 

### 2019-07-10 kindly/infer/infer.tex
Add the proof for completeness.

@@
 \section{Inference}
+\label{inference}
 
-An important innovation of \affe is its principal type inference.
+An important contribution of \affe is its principal type inference.
 We now formulate our inference algorithm
 based on the $HM(X)$ framework~\citep{DBLP:journals/tapos/OderskySW99}.
-$HM(X)$ presents a Hindley-Milner type system for a language
+$HM(X)$ is a Hindley-Milner type system for a language
 with qualified types where constraints are expressed in an arbitrary
-constraint $X$. In particular, given some properties on the constraint system $X$,
+theory $X$.
+Most important, given some properties on the constraint system $X$,
 $HM(X)$ ensures principal type inference.
@@
 system.
+As before, the complete type inference rules are shown in \cref{appendix:infer}.
 
@@
     &\unif &\leq^{\phi}_{\fv{\E}} \unif'
-    &\entail{C'&}{\phi C}
+    &\entail{C'&}{\phi C_\schm}
     &\entail{C'&}{\phi \schm \preceq \schm'}

### 2019-07-10 kindly/lang.tex
.-.. regions

@@
 \label{sdtyping:split}
-
-Linear and affine types also control the copy of variables.
-To demonstrate how this is achieve in \lang, we show the
-{\sc Pair} typing rule in \cref{sdtyp:pair}.
-In the {\sc Pair} rule, we split the type environment $\E$ in two parts,
-$\E_1$ and $\E_2$, which are used to typecheck the components of the pair.
-We note the splitting $\lsplit{C}{\E}{\E_1}{\E_2}$ where $C$ is a constraint
-that must be respected for the split to be valid.
-The main splitting rules are shown in \cref{sdtyp:split}. Each
-rule applies to a binding for a given variable. For instance, the
-{\sc Both} rule indicates that if we can prove that $x$ is unrestricted, we
-can duplicate it in both environments.
-Similarly, since immutable borrows are always unrestricted, the
-{\sc Borrow} rule allows to duplicate them arbitrarily.
-The {\sc Left} and {\sc Right} rule allow to move a binding
-to the left or right environment.
-The {\sc Susp} and {\sc SuspB} allow to split off suspended bindings to
-the left environment while conserving access to the binding on the right.
-Such suspended binding can later be turned
-into a borrow inside a region. The asymmetry here must always follow
-the order of execution of the language: in \affe, execution is from left to
-right, which means that a resource can be used as a borrow on the left
-and then as a full resource on the right, assuming that the borrow does not
-escape. The {\sc Susp} rule work with a full resource while the {\sc SuspB}
-rule allow to split off an immutable suspended binding based off a borrow.
-
-Using these various rules, we can distribute the bindings every time an
-expression has multiple sub-expressions such as pairs, function applications
-and let bindings. For instance, in the
-expression
-$\letin{a}{\text{create}\ 10\ x}
-{\introPair{\region[{}]{\Sone x\IBORROW}{\app{\text{length}}{\borrow[\IBORROW]{a}}}}{a}}$,
-we use the rule {\sc Susp} to detach a borrow on the left of the pair.
-Naturally, we must here ensure that the borrow on the left of the pair does not
-escape, which brings us to the typing rules for regions.
-
-\begin{figure}[!h]
+\begin{figure}[tp]
   \centering
-  \begin{subfigure}{0.35\linewidth}
+  \begin{minipage}{0.35\linewidth}
     \begin{mathpar}
@@
     \label{sdtyp:pair}
-  \end{subfigure}\hfill
-  \begin{subfigure}{0.6\linewidth}
+  \end{minipage}\hfill
+  \begin{minipage}{0.6\linewidth}
     \centering
@@
 
+      {\Cempty}&
+      {\bvar{\borrow[\IBORROW]{x}}{\schm}}&
+      {\bvar{\borrow[\IBORROW]{x}}{\schm}}&{\bvar{\borrow[\IBORROW]{x}}{\schm}}
+      &Borrow\\[2mm]
+
       {\Cempty}&{\bvar{x}{\schm}}&{\bvar{x}{\schm}}&{\bnone}
@@
       {\bvar{\borrow x} \schm}&{\svar[\IBORROW] x \schm^n}&{\bvar{\borrow x} \schm}
-      &SuspB\\[2mm]
-
-      {\Cempty}&
-      {\bvar{\borrow[i]{x}}{\schm}}&
-      {\bvar{\borrow[i]{x}}{\schm}}&{\bvar{\borrow[i]{x}}{\schm}}
-      &Borrow\\
+      &SuspB\\
 
@@
     \label{sdtyp:split}
-  \end{subfigure}
+  \end{minipage}
 \end{figure}
+
+The {\sc Pair} typing rule in \cref{sdtyp:pair} demonstrates how \lang
+deals with duplication and dropping of values.
+The splitting $\lsplit{C}{\E}{\E_1}{\E_2}$ in the rule decomposes the
+type environment $\E$ in two parts, $\E_1$ and $\E_2$, which are used
+to typecheck the components of the pair. 
+The constraint $C$ characterizes a valid  splitting.
+
+\cref{sdtyp:split} shows the action of the splitting rules on single
+bindings. If $x$'s type is unrestricted, then
+the {\sc Both} rule indicates that we can duplicate it.
+Similarly, unrestricted borrows can be duplicated according to rule
+{\sc Borrow}.
+The {\sc Left} and {\sc Right} rules are always applicable and move a binding
+either to the left or right environment.
+The rules {\sc Susp} and {\sc SuspB} split off suspended bindings to
+the left while conserving access to the binding on the right.
+A suspended binding can later be turned
+into a borrow inside a region. Splitting of suspended bindings is
+asymmetric. It must follow the order of execution from left to right,
+which means that a resource can be used first as a borrow on the left 
+and then later as a full resource on the right. The {\sc Susp} rule
+works with a full resource while the {\sc SuspB} 
+rule splits off an immutable suspended binding from a borrow.
+
+Splitting applies whenever an
+expression has multiple subexpressions:  pairs, function applications,
+and let bindings. For instance, in the
+expression
+$\letin{a}{\text{create}\ 10\ x}
+{\introPair{\region[{}]{\Sone x\IBORROW}{\app{\text{length}}{\borrow[\IBORROW]{a}}}}{a}}$,
+the rule {\sc Susp} splits off a borrow from  the resource
+\lstinline/a/ to use it in the left component of the pair.
+As usual, a borrow cannot be active in the same scope as its resource.
+The \emph{region} around its use ensures that the borrow in the left component does not
+escape, which brings us to next topic.
 
@@
 \label{sdtyping:regions}
-
-In order to provide good support for imperative programming, \affe allows
-to borrow resources.
-Such borrows are constrained to a given syntactic region. We note
-$\region{\Sone x\BORROW}{e}$ the region which allow borrows of $x$ to be used in $e$
-and where $n$ is the nesting of the region.
-Borrows can not be used outside of a corresponding region.
-These remarks lead to an
-algorithm for automatic region annotation: start from each borrow, create
-a region, and extend it as much as possible while respecting other borrows
-and binders.
-A complete specification of the algorithm is provided in \cref{regionannot}.
-In the rest of this section, we will only consider terms
-with fully annotated regions.
-
-The main idea of region typing (implemented by the {\sc Region} rule)
-is to replace suspended bindings by borrow bindings, typecheck the body
-of the region, and finally ensure that the borrow made available inside
-the region does not leak outside.
-This last check is done by using integer indices that correspond to the nesting
-level of the region. The kind of borrow is annotated with the integer
-corresponding to their region, and a final constraint ensure that
-the return type of the region does not contain such borrow.
-
-We illustrate this on the expression $\region{\Sone x\IBORROW}{f(\borrow[\IBORROW]{c})}$
-when $c$ is a linear channel.
-The first step is to check that
-$\svar[\IBORROW]{c}{\text{channel}}$ is in $\E$. We then define $\E'$ which
-corresponds to $\E$ where the suspended binding is replaced by a borrow binding.
-%
-This is done by the borrow relation, noted $\lregion{C}{x}{\E}{\E'}$ and
-whose main rules are shown in \cref{selectrules:region}.
-In our case, we consider an immutable binding, so we replace the suspended
-binding by $\bvar{\borrow[\IBORROW]{c}}{\borrowty[\IBORROW]{k}{\text{channel}}}$.
-We must also check that $\entail{C}{(\kun_n\lk k\lk\kun_\infty)}$.
-We can then typecheck the body of the region, $f(\borrow[\IBORROW]{c})$ in our example, and check that the borrow made available in the region does not escape.
-This is done by the constraint $\Cleq{\tau}{\klin_{n-1}}$, which ensures
-that the kind of $\tau$ does not contain anything whose kind is ``above $n$''
-(for instance, $\borrowty[\IBORROW]{\kun_n}{\text{channel}}$ would have kind
-$\kun_n$).
-
-\begin{figure}[!h]
+\begin{figure}[tp]
   \begin{minipage}{0.4\linewidth}
@@
 
+Borrowing is crucial to support an imperative programming style.
+To guarantee the validity of a borrow, its lifetime must be properly contained in its
+ancestors lifetime. \lang ensures proper nesting of lifetimes by using
+regions. The expression $\region{\Sone x\BORROW}{e}$ indicates a 
+region at nesting level $n$ in which a $\BORROW$-borrow can be taken of $x$.
+
+The typing for a region (see the {\sc Region} rule in \ref{selectrules:region})
+replaces suspended bindings by borrow bindings, typechecks the body
+of the region, and finally ensures that the borrow does not leak outside.
+This last check is done with integer indices that correspond to the nesting
+level of the region. The kind of a borrow is indexed with the level
+corresponding to its region and the constraint $\Cleq{\tau}{\klin_{n-1}}$ ensures that
+the return type of the region must live at some enclosing, lower level.
+
+As an example, consider the expression $\region{\Sone x\IBORROW}{f(\borrow[\IBORROW]{c})}$
+where $c$ is a linear channel.
+The first step is to check that
+$\svar[\IBORROW]{c}{\text{channel}}$ is in $\E$. We then define $\E'$ which
+corresponds to $\E$ where the suspended binding is replaced by a borrow binding.
+%
+This is done by the borrow relation, noted $\lregion{C}{x}{\E}{\E'}$ and
+whose main rules are shown in \cref{selectrules:region}.
+In our case, we consider an immutable binding, so we replace the suspended
+binding by $\bvar{\borrow[\IBORROW]{c}}{\borrowty[\IBORROW]{k}{\text{channel}}}$.
+We must also check that $\entail{C}{(\kun_n\lk k\lk\kun_\infty)}$.
+We can then typecheck the body of the region, $f(\borrow[\IBORROW]{c})$ in our example, and check that the borrow made available in the region does not escape.
+This is done by the constraint $\Cleq{\tau}{\klin_{n-1}}$, which ensures
+that the kind of $\tau$ does not contain anything whose kind is ``above $n$''
+(for instance, $\borrowty[\IBORROW]{\kun_n}{\text{channel}}$ would have kind
+$\kun_n$).
+
+These remarks lead to an
+algorithm for automatic region annotation: start from each borrow, create
+a region, and extend it as much as possible while respecting other borrows
+and binders.
+A complete specification of the algorithm is provided in \cref{regionannot}.
+In the rest of this section, we will only consider terms
+with fully annotated regions.
+
 \subsubsection{Kind checking}

### 2019-07-10 kindly/lang.tex
regions

@@
       \inferrule{
-        \entail{C}{(\BORROW_n\lk k\lk \BORROW_\infty)}\\
-        \BORROW\in\left\{\kun;\kaff\right\}
+        \entail{C}{(\BORROW_n\lk k) \wedge (k \lk \BORROW_\infty)}\\
+        \BORROW\in\left\{\kun,\kaff\right\}
       }{
@@
 
-The typing for a region (see the {\sc Region} rule in \ref{selectrules:region})
-replaces suspended bindings by borrow bindings, typechecks the body
+The typing for a region (see the {\sc Region} rule in \cref{selectrules:region})
+replaces suspended bindings by borrow bindings
+(\cref{env:rule:borrow}), typechecks the body 
 of the region, and finally ensures that the borrow does not leak outside.
-This last check is done with integer indices that correspond to the nesting
-level of the region. The kind of a borrow is indexed with the level
-corresponding to its region and the constraint $\Cleq{\tau}{\klin_{n-1}}$ ensures that
+This last check is done with indices that correspond to the nesting
+level of the region. The kind $k$ of the borrow is indexed with the level $n$
+corresponding to its region, thanks to the constraint $(\BORROW_n\lk
+k)$, and the constraint $\Cleq{\tau}{\klin_{n-1}}$ ensures that 
 the return type of the region must live at some enclosing, lower level.
@@
 As an example, consider the expression $\region{\Sone x\IBORROW}{f(\borrow[\IBORROW]{c})}$
-where $c$ is a linear channel.
-The first step is to check that
-$\svar[\IBORROW]{c}{\text{channel}}$ is in $\E$. We then define $\E'$ which
-corresponds to $\E$ where the suspended binding is replaced by a borrow binding.
-%
-This is done by the borrow relation, noted $\lregion{C}{x}{\E}{\E'}$ and
-whose main rules are shown in \cref{selectrules:region}.
-In our case, we consider an immutable binding, so we replace the suspended
-binding by $\bvar{\borrow[\IBORROW]{c}}{\borrowty[\IBORROW]{k}{\text{channel}}}$.
-We must also check that $\entail{C}{(\kun_n\lk k\lk\kun_\infty)}$.
-We can then typecheck the body of the region, $f(\borrow[\IBORROW]{c})$ in our example, and check that the borrow made available in the region does not escape.
-This is done by the constraint $\Cleq{\tau}{\klin_{n-1}}$, which ensures
-that the kind of $\tau$ does not contain anything whose kind is ``above $n$''
-(for instance, $\borrowty[\IBORROW]{\kun_n}{\text{channel}}$ would have kind
-$\kun_n$).
+where $c$ is a linear channel in an environment $\E$.
+The first step is to check that $\svar[\IBORROW]{c}{\text{channel}}$
+is in $\E$.
+When entering the region, rule \TirName{Region} imposes
+$\lregion{C}{x}{\E}{\E'}$, which defines $\E'$
+corresponding to $\E$ where the suspended binding is replaced by the
+borrow binding  $\bvar{\borrow[\IBORROW]{c}}{\borrowty[\IBORROW]{k}{\text{channel}}}$.
+To constrain the borrow to this region we impose the constraint
+$\entail{C}{(\kun_n\lk k) \wedge (\lk\kun_\infty)}$, which affirms
+that the borrow is unrestricted, but can only be used in nesting
+levels $n$ and higher.  
+WIth that setup we typecheck the body of the region,
+$f(\borrow[\IBORROW]{c})$.
+The constraint $\Cleq{\tau}{\klin_{n-1}}$ ensures
+that the kind of $\tau$ does not contain anything whose kind is
+``above level $n$'' and thus confines the borrow to the region
+(for instance, $\borrowty[\IBORROW]{\kun_n}{\text{c}}$ has kind
+$\kun_n$ and thus cannot be returned).
 
-These remarks lead to an
-algorithm for automatic region annotation: start from each borrow, create
-a region, and extend it as much as possible while respecting other borrows
-and binders.
-A complete specification of the algorithm is provided in \cref{regionannot}.
-In the rest of this section, we will only consider terms
-with fully annotated regions.
+As it would be cumbersome to markup regions manually, we developed an
+algorithm that starts from each borrow, creates a region around it,
+and extends it as much as possible while respecting other borrows
+and binders. The algorithm (see \cref{regionannot}) also provides the region's level
+and variable annotation. 
+Henceforth, we consider terms with fully annotated regions.
 

### 2019-07-10 kindly/infer/infer.tex
Principality.

@@
 and completeness properties of HM(X). This is achieved by extending
-the original proofs.
+the original proofs from HM(X), which is done in \cref{appendix:infer}.
 %
@@
 
-The proofs are given in \cref{appendix:infer}.
+Finally, principality is a direct application of completeness to
+toplevel programs.
+
+\begin{corollary}[Principality]
+  Let $\inferS{\Ctrue}{\E}{e}{\schm}$ a closed typing judgement.\\
+  Then $\inferW{\Sv}{(C,\unif)}{\Eflat\E}{e}{\tau}$
+  such that:
+  \begin{align*}
+    (\Ctrue,\schm_o) &= \generalize{C}{\unif\E}{\tau}
+    &\entail{&}{\schm_o \preceq \schm}
+    % &( C, \schm, \unif) &\leq (C',\schm',\unif')
+  \end{align*}
+
+\end{corollary}
 

### 2019-07-10 kindly/lang.tex
checkpoint-semantics

@@
 \label{sem}
+\input{internal-grammar}
 
-\input{internal-grammar}
-\input{fig-reduction}
+\lstMakeShortInline[style=rule]@
+
+In this section, we present the dynamics of \lang in big-step
+functional
+style~\cite{siek13:_type_safet_three_easy_lemmas,DBLP:conf/esop/OwensMKT16,
+  DBLP:conf/popl/AminR17}. It is given as a function
+@eval \Store \Perm \VEnv i e@ 
+that manipulates the semantic objects defined in
+\cref{fig:syntax-internal-language}. 
+To demonstrate resource handling, the semantics comes with an abstract
+notion of linear resources that can be created, destroyed, observed
+(through unrestricted borrows), and
+updated (through affine borrows).
+
+Expressions $e$ need to be elaborted with some kind and constraint
+annotations. Polymorphic functions give rise to elaborated
+abstractions which abstract over kind variables $\Multi\kappa$ and
+which are tagged with a kind $k$ indicating their usage restriction
+(linear, affine, etc). Their use
+gives rise to elaborated application where the instantiation of these
+kind variables is given. Pairs also come with a kind tag $k$
+indicating the usage restriction.
+
+Addresses $\Addr$ are composed of a raw location $\Loc$, which is just
+a pointer into a store, and a stack of modifiers that indicates the
+borrows that have been taken from the original object. From the raw
+location, we may take affine borrows and reborrowed. Once we have
+taken an unrestricted borrow (from a raw location or a borrowed one),
+then we can take further unrestricted borrows from it, but no more
+affine ones.
+
+A permission $\Perm$ is a set of addresses. During evaluation it
+contains the set of addresses that may be accessed. For each raw
+location, there is only ever one address inside of $\Perm$.
+
+A result $r$ is either an address or a constant (e.g., a number). All
+non-trivial values are boxed in the sense that they are represented by
+the address of their actual content.
+
+A value environment $\VEnv$ is a mapping from variables to results.
+
+A storable $W$ is a particular object that can be stored. There are five
+kinds of storable. A poly closure represents a polymorphic
+function. It consists of an environment and the components of an
+elaborated abstraction. Similarly, a closure represents a monomorphic
+function and it elides kind variables and the constraint. A pair
+consists of a kind and two results. A resource is just an indirection
+to another result and the $\blob $ fills a released location.
+
+Finally, the store $\Store$ is a mapping from raw locations to
+storables. It comes with a function
+@salloc : store -> storable -> (loc * store)@ such that
+@salloc delta w@ allocates a free location in @delta@ and fills it with
+storable $w$. It returns the location and the updated store.
+
+The evaluation function is indexed by a step count so that each
+invocation is guaranteed to terminate either with an error, a timeout,
+or a meaningful result. Hence its return type is a monad
+\lstinline/'a sem/ which combines error reporting, timeout, and a
+result of type \lstinline/'a/:
+\begin{lstlisting}
+type 'a sem = Error of string | TimeOut | Ok of 'a
+# eval : store -> perm -> venv -> int -> exp -> (store * perm * result) sem
+\end{lstlisting}
+It evaluates the given expression in the context of an initial store, a
+permission to use addresses in the store, a value environment, and a
+step count. If successful, it returns the final store, the remaining
+permissions, and the actual result.
+
 \input{fig-code}
+
+We give some excerpts of the definition of @eval@ in
+\cref{fig:big-step-interpretation} and leave the full 
+definition for the appendix~\cref{sec:semant-defin}. The appendix also
+contains the semantics 
+in traditional notation, which does not make the monadic nature of
+evaluation explicit.
+
+The definition uses concrete OCaml syntax with some pretty
+printing. The pervasive \lstinline/let*/ operator acts as monadic bind
+for the \lstinline/sem/ monad. The function header of @eval@ checks
+whether time is up and otherwise proceeds processing the expression.
+
+The @Varinst@ case corresponds to the elaborated application. It
+obtains the variable's value, checks that it is a location, checks the
+permission (the @let*?@ clause), obtains the storable $w$ at that
+location, and checks that it is an STPOLY. Then it updates the
+permission: if the STPOLY closure is unrestricted, then the location
+remains in the permission set, otherwise it is removed. Finally, we
+allocate a new monomorphic closure, add it to the permissions, and
+return the pointer as the result along with the updated store and
+permissions.
+
+The @App@ case implements function application. Compared to the syntax
+it comes with an annotation @sp@ which contains evidence of the
+environment splitting in the application rule. This evidence takes the
+form of two sets of variables from the domain of @gamma@.
+
+To execute an application, we first apply the splitting to @gamma@ and
+evaluate subterm @e_1@ with its part of the environment and the
+decremented timer @i'@. The result must be a location that we are
+permitted to use. Moreover, there must be a monomorphic STCLOS stored
+at that location. The permission to further use this closure is
+withdrawn unless the closure is unrestricted. Finally, we evaluate the
+argument, then the function body, and return its result.
+
+The @Pair@ case does not present any new features, but it
+corresponds to the \TirName{Pair} typing rule.
+
+The @Region@ case implements a region. It obtains the address for @x@,
+the suspended binding, and extends it with the intended borrow
+$\BORROW$. This extension may fail if we try to take an affine borrow
+of an unrestricted borrow. Next, we rebind @x@ to the borrow's
+address, extend the permission accordingly, and execute the region's
+body.  Finally, we withdraw the permission and return the result.
+
+The @Borrow@ case obtains the address for @x@, checks that it is a
+borrow of the correct mode $\BORROW$ and whether it is permitted to
+use it. It just returns the address.
+
+\lstDeleteShortInline@
 

### 2019-07-10 kindly/sdtyping/extra.tex
Improve organization in the appendix.

@@
-\section{Annotating regions}
+\section{Automatic region annotation}
 \label{regionannot}
@@
 region annotations for each individual variable.
-
+%
 \begin{align*}
@@
 \end{align*}
-
+%
+\begin{figure*}[b]
+  \centering
+  \input{region-annot}
+  \caption{Automatic region annotation --- $\RannotT{e}{e'}$}
+  \label{fig:region-annotation}
+\end{figure*}
+%
 \Cref{fig:region-annotation} define a rewriting relation $\RannotT{e}{e'}$
@@
 
-\begin{figure*}[!hbt]
-  \centering
-  \input{region-annot}
-  \caption{Automatic region annotation --- $\RannotT{e}{e'}$}
-  \label{fig:region-annotation}
-\end{figure*}
-
 \section{Constraints}
@@
 
-\begin{figure}[btp]
+\begin{figure}[!btp]
   \centering
@@
 principal type inference, as defined by HM(X).
-
-In order to prove this, we first prove that constraint solving
+We first prove that constraint solving
 does compute normal forms, and that such normal forms are unique.
@@
 linear and affine types, and borrow.
+We extend the proofs of HM(X) to such a system in \cref{appendix:infer}.
 

### 2019-07-11 kindly/infer/infer.tex
Put the appendix into readable shape.

@@
 the greatest lower bound (resp. least upper bound) in $\mathcal L$.
+The lattice used in \lang which was presented
+in \cref{sdtyping} is a total bounded lattice with
+$l^\top = \klin_\infty$ and $l^\bot = \kun_0$.
 %
-Let $\CL$ the set of constraints in this lattice.
+Let $\CL$ the set of constraints in such arbitrary lattice $mathcal L$.
 The full grammar of constraints is shown in \cref{grammar:constraint}.
@@
 type unification is oriented and noted $\leq$.
+For simplicity, we consider all type constructors
+invariant in their parameters
+and define $\Ceq{\tau}{\tau'}$ as $\Cleq{\tau}{\tau'} \wedge \Cleq{\tau'}{\tau}$.
 
@@
 
-\begin{figure}[!h]
+\begin{figure}[!bt]
   \centering
@@
   \end{minipage}\hfill
-  \begin{minipage}{0.64\linewidth}
+  \begin{minipage}{0.66\linewidth}
     \begin{mathpar}
-      \inferrule{}{\entail{}{\Cleq{k}{l^\top}}}
-      \and
-      \inferrule{}{\entail{}{\Cleq{l^\bot}{k}}}
+      \inferrule
+      {\forall i,\ \entail{C}{\Cleq{l_i}{k}}}
+      {\entail{C}{\Cleq{\wedge_i l_i}{k}}}
       \and
       \inferrule
-      {}{ \entail{}{\Cleq{\kvar}{\kvar}} }
+      {\forall i,\ \entail{C}{\Cleq{k}{l_i}}}
+      {\entail{C}{\Cleq{k}{\vee_i l_i}}}
       \and

### 2019-07-11 kindly/sdtyping/extra.tex
Put the appendix into readable shape.

@@
 
-Let us note $\A$ our constraint system. The full grammar of constraints is
-given in \cref{grammar:constraint}.
-$\A$ is defined as the smallest cylindric term constraint system that
-satisfies the axiom shown in \cref{rules:entail}.
-We follows the traditional HM(X) formulation
-with conjunctions, projections and type inequalities.
-The new element specific to our approach are kind inequalities.
+Our system is defined in \cref{infer:solving}.
+As before, we note $\CL$ our constraint system for an arbitrary
+total bounded lattice $(\mathcal L, \lk_\Lat)$.
 Entailment is noted $\entail{C}{D}$, where $D$ is a consequence of the
@@
 when $\entail{C}{D}$ and $\entail{D}{C}$.
-\TODO{Give the cylindric properties ?}
+Following HM(X), we define a cylindric constraint system as
+a system which respects the following properties for any variable $x$:
+\begin{align*}
+  \entail{C&}{\Cproj{x}{C}}
+  &\entail{C}{D} &\implies \entail{\Cproj{x}C}{\Cproj{x}D}\\
+  \Cproj{x}{(C\wedge \Cproj{x}D)}&\equivC \Cproj{x}{C} \wedge \Cproj{x}D
+  & \Cproj{x}{\Cproj{y}D}&\equivC \Cproj{y}{\Cproj{x}D}
+\end{align*}
 
-\begin{figure}[!btp]
+Furthermore, we define a term rewriting system as a system where, for every
+types $\tau$,$\tau'$, there exists an equality predicates $\Ceq{\tau}{\tau'}$
+which is reflexive, transitive, stable under substitution and such that,
+for any predicate $P$:
+\begin{align*}
+  \entail{\Ceq{x}{y}\wedge \Cproj{x}C \wedge \Ceq{x}{y}&}{C}
+  &\subst{x}{\tau}{P} &\equivC \Cproj{x}P \wedge \Ceq{x}{\tau}
+                        \text{ where } x\notin\fv{\tau}
+\end{align*}
+$\CL$ is defined as the smallest cylindric term constraint system that
+satisfies the axiom shown in \cref{rules:entail}.
+We define  $\Ceq{\tau}{\tau'}$ as $\Cleq{\tau}{\tau'} \wedge \Cleq{\tau'}{\tau}$.
+%
+\begin{figure}[!b]
   \centering
@@
 \end{figure}
-
+%
 We note $\SC$ the set of solved forms
 which can be used inside type and kind schemes.
-We define $\SC$ as $\A$ quotiented by the relation $\equivC$.
+We define $\SC$ as $\CL$ quotiented by the relation $\equivC$.
 %
@@
 
+\begin{figure}[!h]
+  \centering
+  \input{sdtyping/sd-kinds}
+  \caption{Syntax-directed kinding rule --
+    $\inferSK{C}{\E}{\tau}{k}$}
+  \label{rules:sd-kinding}
+\end{figure}
+
 We note $\inferSK{C}{\E}{\tau}{k}$
@@
 
-\begin{figure}[ht]
-  \centering
-  \input{sdtyping/sd-kinds}
-  \caption{Syntax-directed kinding rule --
-    $\inferSK{C}{\E}{\tau}{k}$}
-  \label{rules:sd-kinding}
-\end{figure}
-
 \subsection{Environments}
@@
 
-\begin{figure}[tp]
+In \cref{sdtyping}, we only gave a partial description of
+the splitting and borrowing relations on environments,
+$\lsplit C\E\E\E$ and $\lregion C{x}\E\E$.
+The complete definitions are shown on \cref{fig:sd-splitting,fig:sd-borrowing}.
+All the definitions are made in term of the inference version, which
+returns a fresh constraints. The solving version then simply
+uses entailment, as shown in rule {\sc ESplit-Check} and
+{\sc EBorrow-Check}.
+The remaining new rules are dedicated to iterating over the environment.
+
+\cref{fig:contraints-environments-types} defines
+the rewriting relation on environment constraints,
+noted $\Cleq\E k \Crewrite  C$ which rewrites a constraint of the form
+$\Cleq\E k$ into $C$. It proceeds by iterating over the environment
+and expending the constraints for each binding.
+Suspended bindings are rejected ({\sc ConstrSusp}).
+Borrow bindings directly use the annotated kind ({\sc ConstrBorrow}).
+Other bindings uses the underlying type scheme ({\sc ConstrBinding}).
+Type schemes are constrained by first inferring the kind, and then
+emitting the constraint ({\sc ConstrSD} and {\sc ConstrI}).
+
+\begin{figure}[!tb]
   \input{sdtyping/sd-splitting}
@@
   \label{fig:sd-splitting}
-\end{figure}
-
-\begin{figure}[tp]
   \input{sdtyping/sd-borrowing}
@@
   \label{fig:sd-borrowing}
+  \input{sdtyping/sd-environments}
+  \caption{Rewriting constraints on environments --- $\Cleq\E k \Crewrite  C$}
+  \label{fig:contraints-environments-types}
 \end{figure}
@@
 
-\begin{figure*}[tp]
-  \input{sdtyping/sd-environments}
-  \caption{Rewriting constraints on environments and types}
-  \label{fig:contraints-environments-types}
-\end{figure*}
-\begin{figure*}[tp]
+The rules for syntax directed typing are shown in \cref{fig:syntax-directed-typing}
+and follow the presentation given in \cref{sdtyping}.
+As usual in HM type systems, introduction of type-schemes
+is included in the {\sc Let} rule via generalization.
+We define $\generalize{C}{\E}{\tau} =
+(\Cproj{\Multi{\kvar},\Multi{\tvar}}{C},
+\forall \Multi{\kvar},\Multi{\tvar}.\qual{C}{\tau})$
+where
+$\Multi{\kvar},\Multi{\tvar} = (\fv{\tau}\cup\fv{C})\setminus\fv{\E}$.
+
+\begin{figure*}[hbtp]
   \input{sdtyping/sd-typing}

### 2019-07-11 kindly/infer/infer.tex
Small patch to the soundness theorem.

@@
 \begin{theorem}[Soundness of inference]
-  Given a type environment $\E$ and a term $e$,\\
-  if $\inferW{\Sv}{(C,\unif)}{\E}{e}{\tau}$
-  then $\inferS{C}{\unif\Sv}{e}{\tau}$, $\unif C = C$ and $\unif \tau = \tau$.
+  Given a type environment $\E$ containing only value bindings,
+  $\E|_\tau$ containing only type binding and a term $e$,\\
+  if $\inferW{\Sv}{(C,\unif)}{\E;\E_\tau}{e}{\tau}$
+  then $\inferS{C}{\unif(\Sv;\E_\tau)}{e}{\tau}$, $\unif C = C$ and $\unif \tau = \tau$
 \end{theorem}
 
-Note here that the syntax-directed derivation holds with the usage environment $\Sv$ instead of the originally provided environment $\E$. Indeed,
+The syntax-directed derivation holds with the usage environment $\Sv$ instead of the originally provided environment $\E$. Indeed,
 $\E$ does not contain suspended and borrow bindings, since those
-are discovered on the fly and recorded in $\Sv$.
+are discovered on the fly and recorded in $\Sv$. The type binding, however,
+are directly taken from the syntax-directed derivation.
 

### 2019-07-11 kindly/lang.tex
Remove the TODOs.
Minor cosmetics.

@@
 
-\subsubsection{Kind checking}
+% \subsubsection{Kind checking}
 
-\TODO{Present this if we have enough space}
+% \TODO{Present this if we have enough space}
 
-\subsubsection{Pattern matching}
+% \subsubsection{Pattern matching}
 
-\TODO{Present this if we have enough space}
+% \TODO{Present this if we have enough space}
 
-\clearpage
+% \clearpage
 \subsection{Semantics}
@@
 
-\lstMakeShortInline[style=rule]@
+\lstMakeShortInline[style=rule,columns=flexible,basicstyle=\normalsize\normalfont]@
 
@@
 type 'a sem = Error of string | TimeOut | Ok of 'a
-# eval : store -> perm -> venv -> int -> exp -> (store * perm * result) sem
+val eval : store -> perm -> venv -> int -> exp -> (store * perm * result) sem
 \end{lstlisting}
@@
 The definition uses concrete OCaml syntax with some pretty
-printing. The pervasive \lstinline/let*/ operator acts as monadic bind
+printing. The pervasive @let*@ operator acts as monadic bind
 for the \lstinline/sem/ monad. The function header of @eval@ checks

### 2019-07-11 kindly/nonanon.tex
Add the non anonymous materials.

\documentclass{article}

\usepackage[T1]{fontenc}
\usepackage[utf8]{inputenc}
\usepackage[english]{babel}
\usepackage{hyperref}

\title{Non-anonymous material for the article\\
  ``Kindly Bent to Free Us''}
\date{}

\begin{document}

\maketitle

An online playground for the type-checker is available at:
\begin{center}
  \url{https://drup.github.io/pl-experiments/affe/}
\end{center}

The implementation is available at
\begin{center}
  \url{https://github.com/Drup/pl-experiments}
\end{center}

\end{document}

%%% Local Variables:
%%% mode: latex
%%% TeX-master: t
%%% End:

### 2019-07-11 kindly/infer/exampleconstr.tex

We now demonstrate the constraint solving
algorithm presented in \cref{infer:solving}.
Let us consider the expression $\lam{f}{\lam{x}{(\app{f}{x},x)}}$.
The application of our inference algorithm yields the following constraints:
%
\begin{align*}
  \E &= (\tvar_f : \kvar_f)
  (\tvar_x : \kvar_x)\dots
  &C &= (\tvar_f \leq \gamma \tarr{\kvar_1} \beta )
  \Cand
  (\gamma \leq \tvar_x)
  \Cand
  (\beta * \tvar_x \leq \alpha_r)
  \Cand
  (\kvar_x \leq \kun)
\end{align*}

The first step of the algorithm is to use Herbrand unification to obtain
a type skeleton. We thus obtain
$$\alpha_r =
(\gamma \tarr{\kvar_3} \beta) \tarr{\kvar_2} \gamma \tarr{\kvar_1} \beta * \gamma$$

and the following kind constraints: 
\begin{gather*}
  \Cproj{\kvar_r}\Cproj{\kvar_f}\Cproj{\kvar_x}
  (\kvar_x \leq \kun)
  \Cand
  (\kvar_\gamma \leq \kvar_x)
  \Cand
  (\kvar_x \leq \kvar_r)
  \Cand
  (\kvar_\beta \leq \kvar_r)
  \Cand
  (\kvar_3 \leq \kvar_f)
  \Cand
  (\kvar_f \leq \kvar_1)
\end{gather*}
Since $\kvar_r$, $\kvar_f$ and $\kvar_x$ are not present in the type skeleton,
they are existentially quantified.

We translate these constraints into a relation whose graph
is shown in \cref{example:graph}.
%
The algorithm then proceeds as follow:
\begin{itemize}[noitemsep]
\item[1] From the constraints above, we deduce the graph shown
  with plain arrows.
\item[2] We add all the dashed arrows by saturating
  lattice inequalities. For clarity, we only show $\kun$.
\item[4] We identify the connected component circled in
  {\color{green} green}.
  We deduce $\kvar_\gamma = \kvar_x = \kun$
\item[5] We take the transitive closure, which adds the
  arrow in {\color{blue} blue} from $\kvar_1$ to $\kvar_3$.
\item[6] We remove the remaining existentially quantified node in {\color{red} red}, $\kvar_r$ and $\kvar_f$.
\item[7, 8, 9] We clean up the graph, and obtain the graph shown
  in \cref{example:graph:final}.
  We deduce $\kvar_3 \leq \kvar_1$.
\end{itemize}

The final constraint is thus
$$\kvar_\gamma = \kvar_x = \kun \Cand \kvar_3 \leq \kvar_1$$
If we were to generalize, we would obtain the type scheme:
$$\forall \kvar_\beta \kvar_1 \kvar_2 \kvar_3
(\gamma : \kun) (\beta : \kvar_\beta).\ %
\qual
{\Cleq{\kvar_3}{\kvar_1}}
{(\gamma \tarr{\kvar_3} \beta) \tarr{\kvar_2} \gamma \tarr{\kvar_1} \beta * \gamma}$$

The algorithm we presented does not use variance, however we can use it to
simplify this type further by noticing that $\kvar_1$ and $\kvar_2$ are only
used in covariant position, and can thus be replaced by their lower bound.
By removing the unneeded quantifier, we obtain the equivalent type:
$$
\forall \kvar
(\gamma : \kun).
{(\gamma \tarr{\kvar} \beta) \tarr{} \gamma \tarr{\kvar} \beta * \gamma}$$

\begin{figure}[!hb]
  \centering
  \begin{minipage}{0.5\linewidth}
  \centering
  \begin{tikzpicture}[every edge/.style=[link,->,>=latex,thick]]
    \node (U) {$\kun$} ;
    \node[above left=of U] (x) {$\kvar_x$} ;
    \node[above=of x] (r) {$\kvar_r$};
    \node[left=of x] (g) {$\kvar_\gamma$} ;
    \node[right=of x] (b) {$\kvar_\beta$} ;
    \node[right=of b] (3) {$\kvar_3$} ;
    \node[above=of 3] (f) {$\kvar_f$} ;
    \node[above=of f] (1) {$\kvar_1$} ;

    \draw (x) to[bend right] (U) ;
    \draw (x) -> (r) ;
    \draw (b) -> (r) ;
    \draw (g) -> (x) ;
    \draw (3) -> (f) ;
    \draw (f) -> (1) ;

    \draw[blue] (3) to[bend right] (1);

    \begin{scope}[dashed,gray]
      \draw (U) to[bend right] (x) ;
      \draw (U) to (b) ;
      \draw (U) to[bend left] (g) ;
      \draw (U) to (r) ;
      \draw (U) to (1) ;
      \draw (U) to (3) ;
      \draw (U) to (f) ;
    \end{scope}

    \begin{scope}[on background layer]
      \node[fill=green!20,draw=green,
      inner sep=-1mm,ellipse,rotate fit=-20,fit=(U) (x) (g)] {};
      \node[fill=red!20,draw=red, inner sep=-2pt, circle,fit=(r)] {};
      \node[fill=red!20,draw=red, inner sep=-2pt, circle,fit=(f)] {};
    \end{scope}
  \end{tikzpicture}
    \caption{Graph representing the example constraints}
    \label{example:graph}
\end{minipage}~
\begin{minipage}{0.5\linewidth}
  \centering
  \begin{tikzpicture}[every edge/.style=[link,->,>=latex,thick]]
    \node (b) {$\kvar_\beta$} ;
    \node[right=of b] (3) {$\kvar_3$} ;
    \node[above=of 3] (f) {} ;
    \node[above=of f] (1) {$\kvar_1$} ;

    \draw (3) to[bend right] (1);

  \end{tikzpicture}
    \caption{Final state of the graph}
    \label{example:graph:final}
\end{minipage}
\end{figure}

%%% Local Variables:
%%% mode: latex
%%% TeX-master: "../main"
%%% End:

### 2019-07-11 kindly/sdtyping/extra.tex

@@
 types $\tau$,$\tau'$, there exists an equality predicates $\Ceq{\tau}{\tau'}$
-which is reflexive, transitive, stable under substitution and such that,
+which is symmetric, reflexive, transitive, stable under substitution and such that,
 for any predicate $P$:
@@
 %
-\begin{figure}[!b]
+We note $\SC$ the set of solved forms
+which can be used inside type and kind schemes.
+We define $\SC$ as $\CL$ quotiented by the relation $\equivC$.
+%
+The normalization function is described in \cref{infer:solving}.
+%
+\begin{figure}[!hb]
   \centering
@@
 \end{figure}
-%
-We note $\SC$ the set of solved forms
-which can be used inside type and kind schemes.
-We define $\SC$ as $\CL$ quotiented by the relation $\equivC$.
-%
-The normalization function is described in \cref{infer:solving}.
 
-% We consider the existence of a function $\operatorname{normalize}$ which takes
-% a constraint in $\A$ and a substitution $\psi$ and returns a constraint
-% in solved form $C' \in \SC$,
-% and an updated substitution. We detail the implementation
-% of the normalization function in \cref{infer:solving}
+\subsection{Example of constraint solving and simplification}
+\label{solving:example}
 
-% $\mathcal S$ is composed only of kind
-% inequalities \emph{over variables}. For convenience, if $C\in\mathcal S$, we
-% note $C$ as a list of kind inequalities: $\Cleq{\kvar_i}{\kvar_{i'}}^n$.
-% \TODO{Extend the properties of solved forms}
+\input{infer/exampleconstr}
+
+\subsection{Principal constraint system}
 

### 2019-07-11 kindly/lang.tex
checkpoint
Fix ref.

@@
 the statics in terms of syntax directed typing (\cref{sdtyping}),
-and a dynamics that is linearity- and resource-aware (\cref{sem}).
+and a dynamics that is linearity- and resource-aware (\cref{sec:sem}).
 
@@
 \subsection{Semantics}
-\label{sem}
+\label{sec:sem}
 \input{internal-grammar}

### 2019-07-11 kindly/soa.tex
tighten the SOA a bit.

@@
 
-\subsection{Substructural type-system in functional languages}
+\subsection{Substructural type-systems in functional languages}
 
@@
 principal type inference.
-We believe every Quill's kind signature can be expressed in \lang by leveraging
-kind inequalities.
+We were able to express the type signatures of all Quill examples
+in \lang by leveraging kind inequalities.
 Unlike Quill,
@@
 Alms~\citep{DBLP:conf/popl/TovP11} is an ML-like language with rich, kind-based
-affine types and ML modules.
+affine types and ML modules, similar to \lang.
 Their system often relies on existential types to track the identity
-of objects. For instance, they present an array creation
-functions of type \lstinline/int -> 'a -> \E 'b. ('a, 'b) array/ where
-\lstinline/'a/ is the elements' type and \lstinline/'b/
-is a unique identifier of the array.
+of objects. For instance
+\lstinline/Array.create : int -> 'a -> \E 'b. ('a, 'b) array/ where
+\lstinline/'b/ uniquely identify of the array.
 Due to the reliance on existentials, Alms does not support complete type inference.
@@
 In our experience, \affe's limited support for existential types through
-regions is sufficient for a large majority of examples and provides
+regions is sufficient to express Alms' examples and leads to
 a more convenient programming style for imperative code.
 %
-Apart from existential types, Alms relies
-on union and intersection operators and dependent kinds while
+Kind-wise, Alms relies
+on unions, intersections and dependent kinds while
 \lang uses qualified types.
@@
 Most relevant to \lang:
-\begin{itemize}
+\begin{itemize}[leftmargin=*]
 \item Their system does not admit subtyping for arrows and leverages
-  $\eta$-expansion to allow passing unrestricted functions in linear
-  contexts. This is acceptable in a non-strict language such as
+  $\eta$-expansion to pass unrestricted functions in linear
+  contexts. This works in a non-strict language such as
   Haskell but can not be used in a strict setting.
@@
   While the GHC proposal implements some type inference,
-  their algorithm is not formalized and is not complete.
+  it is neither formalized nor complete.
 \item
-  Linear Haskell promotes a CPS style with functions such as 
+  Linear Haskell promotes a continuation-passing style with functions such as 
   \lstinline/withFile : path -> (file ->. file) ->. unit/
@@
   In particular, using abstraction through modules,
-  programmers only need consider the module
-  implementation to ensure that some linear
-  resource is properly handled.
+  programmers only need to consider the module
+  implementation to ensure that linear
+  resources are properly handled.
 % \item
@@
 
-\subsection{Other substructural systems}
+\subsection{Other substructural type-systems}
 
@@
 and only provides \emph{partial} lifetime inference.
-Recently, \citet{DBLP:journals/corr/abs-1903-00982} presented Oxide,
-a core language for Rust which formalizes its ownership discipline
-including non-lexical lifetimes, which we look forward to use to improve
-our borrowing system.
+Recently, \citet{DBLP:journals/corr/abs-1903-00982}
+formalized Rust's ownership discipline, including non-lexical lifetimes.
+We believe this work
+could help us improve our handling of regions and of successive exclusive borrows.
 
 Vault~\citep{DBLP:conf/pldi/DeLineF01}
-is a low-level language which uses typestate along with \emph{keys}
-to track ownership. It also introduces a novel paradigm
-called \emph{adoption and focus} \citep{DBLP:conf/pldi/FahndrichD02}
-which allow restricting the scope of linear objects
-(adoption) or temporarily treating them as non-linear (focus).
-These mechanisms serve a similar purpose as borrows, but are tracked
-much more precisely thanks to the use of typestate and pre/post-conditions,
-which makes it very expressive (at the cost of some verbosity).
-However, Vault is mostly geared towards traditional imperative programming:
-functions can't capture linear objects and it does not support
-type inference.
-
-Plaid \&Co by Aldrich and others
+and Plaid~\citep{DBLP:conf/oopsla/AldrichSSS09}
+leverage typestate and capabilities
+to express very rich properties in objects and protocols.
+These systems are designed for either low-level or object-oriented
+programming, and do not immediately lend themselves to a more functional
+style. Furthermore, while their type-systems are much more
+powerful than \affe's, they require many annotations by the programmer
+and do not support any inference.
+It could however be interesting to expand \affe with limited
+forms of typestate as a local, opt-in feature that would provide
+more expressivity at the cost of inference.
 

### 2019-07-11 kindly/intro.tex
checkpoint

--- 2019-07-08 kindly/intro.tex
@@
 linear and affine types, mutable and immutable borrows, but all that
-integrated with full type inference, garbage collected data, and
-ANOTHER FEATURE.
+integrated with full principal type inference, garbage collected data, and
+automatic placement of borrowing regions.
 
-Compared to the most closely related contender in this design space, Linear Haskell
-\cite{DBLP:journals/pacmpl/BernardyBNJS18}, their goals and means are
+The most closely related contenders in this design space are Linear Haskell
+\cite{DBLP:journals/pacmpl/BernardyBNJS18}, henceforth called LH,
+Quill \cite{DBLP:conf/icfp/Morris16}, and ALMS \cite{DBLP:conf/popl/TovP11}.
+Compared to LH, the goals and means are
 similar as both systems permit abstraction over the number of uses of
@@
 \begin{enumerate}
-\item Their multiplicities are either linear or unrestricted whereas
+\item LH's multiplicities are either linear or unrestricted whereas
   we also permit affine values.
@@
   denotes the type of a \emph{single-use function} that can be called
-  just once, whereas the multiplicities in Linear Haskell choose
+  just once, whereas the multiplicities in LH choose
   between $\alpha\to\beta$ and $\alpha \multimap\beta$ where the
-  latter is a function that promises to \emph{use its argument exactly once}.
+  latter is a function that promises to \emph{use its argument exactly
+    once}.
+\item Similar to Quill, we make use of multiplicity contraints. We
+  rely on kind subsumption to simplify the inferred types.
 \end{enumerate}
+See \cref{sec:related-work} for further in-depth discussion of these
+and other related works.
 \subsection{First examples}
@@
 
-This example demonstrates two features of our system:
+This example demonstrates three features of our system:
 \begin{enumerate}
 \item full type inference without annotations in user code,
-\item types carry multiplicity annotations in the form of kinds.
+\item types carry multiplicity annotations in the form of kinds,
+\item resource APIs can be written in direct style as linearity is a
+  property of the type \lstinline/File.t/.
 \end{enumerate}
@@
   borrowing. The type system is polymorphic with respect to types as
-  well as kinds that express constraint on the number of uses of a value.
+  well as kinds that express constraints on the number of uses of a value.
 \item An extension of the HM(X) framework
   \cite{DBLP:journals/tapos/OderskySW99} with kinds, regions, and
-  multiplicity constraints that supports full, prinicipal type inference.
+  multiplicity constraints that supports full, prinicipal type
+  inference.
+\item Full account of the metatheory for type inference and type soundness.
 \item Constraint simplification techniques.
 \item Automatic inference of regions for borrows.
+\item An implementation of the type inference algorithm including all
+  constraint simplification and extended with pattern matching. A
+  webfront for the algorithm is available, its URL is given in the
+  non-anonymous supplementary material, and we plan to submit it for
+  artifact evaluation.
 \end{itemize}

### 2019-07-11 kindly/infer/infer.tex
Remove the TODOs.
Add an example for constraint solving.
Tighten everything a little bit.

@@
 theory $X$.
-Most important, given some properties on the constraint system $X$,
-$HM(X)$ ensures principal type inference.
+Most importantly,
+$HM(X)$ ensures principal type inference
+when $X$ respects the specified properties.
 We apply HM(X) to a concrete constraint language which we name $\CL$.
-We adapt and extend its existing rules to support kind inference,
+We adapt and extend HM(X) existing rules to support kind inference,
 track linearity and handle borrows and regions. Finally, we
-formulate an appropriate constraint solving algorithm for $\CL$.
-In addition, our constraint solving algorithm simplifies constraints in a
-principled way.
-We first present various preliminary definitions
-before showing our kind and type inference judgements
-and our constraint solving algorithm.
+formulate an appropriate constraint solving algorithm for $\CL$
+which solves and simplifies constraints in a principled way.
+We start by various preliminary definitions.
 
@@
 In order to infer such constraints, our inference judgement will not only
-take an environment as parameter but also return an environment, which
-we call ``usage environment'', and which summarize how variables and borrows
+take an environment as parameter but also return an environment which
+we call ``usage environment'' and which summarizes how variables and borrows
 are used. Usage environments follow the exact same grammar
@@
 
-In \cref{sdtyping}, we used various relations to split environments in two, or
-transform suspended bindings into borrow binding inside an environment.
-These relations took as argument a constraint which is used to validate
-the environment transformation.
-In the context of inference, we define a new judgements which \emph{infers}
-the constraints required.
-\begin{itemize}
+In \cref{sdtyping}, we used relations to split environments in two and to
+transform suspended bindings into borrows inside a region.
+These relations take as argument a constraint which validates
+the transformations.
+In the context of inference, we define new judgements which \emph{infers}
+the constraints.
+\begin{itemize}[leftmargin=*]
 \item $\bsplit{C}{\Sv}{\bf{\Sv_1}}{\bf{\Sv_2}}$.
-  Given two usage environments (inferred for two subexpressions, for instance)
-  $\Sv_1$ and $\Sv_2$, we return $\Sv$, the merged environment, and $C$, a set
+  Given two usage environments $\Sv_1$ and $\Sv_2$
+  (inferred from two subexpressions, for instance),
+  we return $\Sv$, the merged environment, and $C$, a set
   of constraints that must be respected.
@@
 The relations used for syntax-directed typing can trivially be defined
-in term of these new relations and constraint entailment.
+in term of these new relations by using constraint entailment.
 All relations are fully described in \cref{typ:extra:envs}.
@@
 The HM(X) framework assumes the existence of a $\operatorname{normalize}$
-function which takes a constraint $C$ and a substitution $\psi$ and returns a
-simplified constraint $C'$,
+function which takes a constraint $C$, a substitution $\psi$ and returns a
+simplified constraint $C'$
 and an updated substitution $\unif'$.
-Normalization is supposed to return the ``best'' solution $(C',\unif')$, for
-instance the most general unifier.
+Normalization must return a normal form such that $\unif'$ is a most general unifier.
 We detail the implementation
@@
 \cref{rules:kinding}.
-Similar to the syntax-directed rules, higher-kinds are not generally supported
+Higher-kinds are not generally supported
 and can only appear by looking-up the kind scheme of a type constructor,
@@
 inference algorithm.
+Unlike in the syntax-directed version, $\E$ only regular and type bindings.
+Suspended and borrow bindings can only be present in $\Sv$.
 We revisit the syntax-directed rules presented in \cref{sdtyping} to highlight the
@@
 
-In our syntax directed system presented in \cref{sdtyping}, we ensured
+In our syntax-directed system, we ensured
 that linear variables are not discarded at the \emph{leafs}, through
@@
 
-Note again here the use of normalization on complex constraints to ensure
+Finally, we use normalization on complex constraints to ensure
 that the inference algorithm always return the simplest possible
@@
 
-As noted previously, in our inference algorithm, we infer
-usage environments which describes how variables are used. In
-\cref{sdtyping}, we described the {\sc Pair} and {\sc Region} rules, and their
-respective actions on the environment. We now see the inference version
-of these rules in \cref{rule:infer:envrules}.
+We now present the inference version
+of the {\sc Pair} and {\sc Region} rules in \cref{rule:infer:envrules}.
 While the rules are quite similar to the original version, the usage
@@
 $\bsplit{C}{\Sv}{\bf{\Sv_1}}{\bf{\Sv_2}}$ and $\bregion{C}{\bf{x}}{\Sv}{\bf\Sv'}$,
-which returns the necessary constraints and
+which return the necessary constraints.
+As appropriate for inference, these relations 
 are total and non ambiguous for their input parameters ($\Sv_1, \Sv_2$ and $\Sv'$, respectively).
-We then simply collect all the constraints and normalize them.
+We then collect all constraints and normalize them.
 
@@
 richer.
-To illustrate this, we now show the inference rule for let-bindings
+To illustrate this, we show the inference rule for let-bindings
 in \cref{rule:infer:let}.
@@
 identifier $x$. Since let-expressions contain
-two subexpressions, so we use the environment splitting relation,
-$\bsplit{C_s}{\Sv}{\Sv_1}{(\Sv_2 \Sdel{x})}$. We carefully remove the $x$ from
-the left environment, since it is not available outside of the expression
-$e_2$, and thus should not be part of the returned usage environment.
+two subexpressions, we use the environment splitting relation,
+$\bsplit{C_s}{\Sv}{\Sv_1}{(\Sv_2 \Sdel{x})}$. We remove the $x$ from
+the right environment, since it is not available outside of the expression
+$e_2$, and should not be part of the returned usage environment.
 
@@
 
-Before presenting our constraint solving algorithm,
-we place ourselves in a more general context where kinds are either variables
-or constants belonging to a total bounded lattice $(\mathcal L, \lk_\Lat)$ ie.,
-a lattice which admits a total order and upper and lower bounds ($l^\top$ and $l^\bot$).
-We note lattice elements $l$ and $\glb_i l_i$ (resp. $\lub_i l_i$)
-the greatest lower bound (resp. least upper bound) in $\mathcal L$.
-The lattice used in \lang which was presented
-in \cref{sdtyping} is a total bounded lattice with
+We define the constraint solver in terms of an arbitrary bounded
+lattice $(\mathcal L, \lk_\Lat)$, i.e., 
+a lattice which has a minimal and a maximal element ($l^\top$ and $l^\bot$).
+We write lattice elements as $l$ and $\glb_i l_i$ (resp. $\lub_i l_i$)
+for the greatest lower bound (resp. least upper bound) in $\mathcal L$.
+The lattice for \lang (see \cref{sdtyping}) is a bounded lattice with
 $l^\top = \klin_\infty$ and $l^\bot = \kun_0$.
-%
-Let $\CL$ the set of constraints in such arbitrary lattice $mathcal L$.
+
+Let $\CL$ be the set of constraints in such a lattice $\mathcal L$.
 The full grammar of constraints is shown in \cref{grammar:constraint}.
@@
 
-\TODO{Add example}
-
+An example of this algorithm in action is shown in \cref{solving:example}.
 Our algorithm is complete, computes principal normal forms,
 and already simplifies constraints significantly
-(notably thanks to steps 6, 7 and 8).
-It can be easily extended with further simplification phases.
+(thanks to steps 6, 7 and 8).
+It can be extended with further simplification phases.
 In particular, our implementation and all the signatures presented in
@@
 lower (resp. upper) bounds.
-Remarkably, all the simplification mechanisms presented
-here (including the variance-based one) are complete.
+All the simplification mechanisms presented
+here, including the variance-based one, are complete.
 It is also possible to add ``best-effort'' simplification
@@
 are sufficient to state that $HM(\CL)$ provides principal type inference.
-However, we do not use HM(X) directly, but an extension with kind inference
+However, we use an extension of HM(X) with kind inference
 and affine types and borrows.
@@
 \begin{theorem}[Completeness]
-  Let $\inferS{C'}{\unif'\E}{e}{\schm'}$ a typing derivation,
-  then $\inferW{\Sv}{(C,\unif)}{\Eflat\E}{e}{\tau}$
-  for some substitution $\unif$, constraint $C$ and type $\tau$ such
+  Given $\inferS{C'}{\E'}{e}{\tau'}$ and
+  $\entail{}{\unif'\E \preceq \E'}$.
+  Then $$\inferW{\Sv}{(C,\unif)}{\Eflat\E}{e}{\tau}$$
+  for some environment $\Sv$,
+  substitution $\unif$, constraint $C$ and type $\tau$ such
   that
   \begin{align*}
-    (C_\schm,\schm) &= \generalize{C}{\unif\E}{\tau}
-    &\unif &\leq^{\phi}_{\fv{\E}} \unif'
-    &\entail{C'&}{\phi C_\schm}
-    &\entail{C'&}{\phi \schm \preceq \schm'}
+    \unif &\leq^{\phi}_{\fv{\E}} \unif'
+    &\entail{C'&}{\phi C}
+    &\entail{&}{\phi \schm \preceq \schm'}
+    &\Sv&\subset\E
     % &( C, \schm, \unif) &\leq (C',\schm',\unif')
   \end{align*}
+  where $\schm' = \generalize{C'}{\E'}{\tau'}$
+  and $\schm = \generalize{C}{\E}{\tau}$
 \end{theorem}

### 2019-07-11 kindly/intro.tex
Tighten everything a little bit.

@@
 protocol aspect. The systems that do build on linear types \cite{DBLP:journals/tcs/Girard87} and/or
-uniqueness types\cite{DBLP:conf/plilp/BarendsenS95}. A value of linear
+uniqueness types~\cite{DBLP:conf/plilp/BarendsenS95}. A value of linear
 type is guaranteed to be consumed 
@@
 and they form the foundation of the type system of the Rust language
-\cite{rust}. Rust supports two kinds of borrows of data structures, mutable and immutable
-ones. Mutable borrows enable modification of the data structure
-whereas immutable borrows grant read-only access. Mutable borrows are
-exclusive whereas several immutable borrows can be active at the same
-time. Moreover, Rust makes sure that the lifetime of a
+\cite{rust}. Rust supports two kinds of borrows of data structures, shared and
+exclusive ones.
+Exclusive borrows generally enable modification of the data structure
+whereas shared borrows only grant read-only access.
+Several shared borrows can be active at the same time, but
+can not be active while an exclusive borrow is.
+Moreover, Rust makes sure that the lifetime of a
 borrow is properly contained in the lifetime of its lender.
@@
   \cite{DBLP:journals/tapos/OderskySW99} with kinds, regions, and
-  multiplicity constraints that supports full, prinicipal type
-  inference.
+  multiplicity constraints that supports full, principal type inference.
 \item Full account of the metatheory for type inference and type soundness.

### 2019-07-11 kindly/lang.tex
Tighten everything a little bit.

@@
 
-\lstMakeShortInline[style=rule,columns=flexible,basicstyle=\normalsize\normalfont]@
+\lstMakeShortInline[style=rule,basicstyle=\normalsize\normalfont]@
 
@@
 borrows that have been taken from the original object. From the raw
-location, we may take affine borrows and reborrowed. Once we have
+location, we may take affine borrows and reborrows. Once we have
 taken an unrestricted borrow (from a raw location or a borrowed one),
@@
 storables. It comes with a function
-@salloc : store -> storable -> (loc * store)@ such that
-@salloc delta w@ allocates a free location in @delta@ and fills it with
+@salloc@ of type @store -> storable -> (loc * store)@ such that
+@salloc delta@ @w@ allocates a free location in @delta@ and fills it with
 storable $w$. It returns the location and the updated store.
@@
 \cref{fig:big-step-interpretation} and leave the full 
-definition for the appendix~\cref{sec:semant-defin}. The appendix also
+definition for \cref{sec:semant-defin}. The appendix also
 contains the semantics 

### 2019-07-11 kindly/infer/infer.tex
beginning of explanation constraint solving

--- 2019-07-11 kindly/infer/infer.tex
@@
 
-Before presenting our constraint solving algorithm,
-we place ourselves in a more general context where kinds are either variables
-or constants belonging to a total bounded lattice $(\mathcal L, \lk_\Lat)$ ie.,
-a lattice which admits a total order and upper and lower bounds ($l^\top$ and $l^\bot$).
-We note lattice elements $l$ and $\glb_i l_i$ (resp. $\lub_i l_i$)
-the greatest lower bound (resp. least upper bound) in $\mathcal L$.
-The lattice used in \lang which was presented
-in \cref{sdtyping} is a total bounded lattice with
+We define the constraint solver in terms of an arbitrary bounded
+lattice $(\mathcal L, \lk_\Lat)$, i.e., 
+a lattice which has a minimal and a maximal element ($l^\top$ and $l^\bot$).
+We write lattice elements as $l$ and $\glb_i l_i$ (resp. $\lub_i l_i$)
+for the greatest lower bound (resp. least upper bound) in $\mathcal L$.
+The lattice for \lang (see \cref{sdtyping}) is a bounded lattice with
 $l^\top = \klin_\infty$ and $l^\bot = \kun_0$.
 %
-Let $\CL$ the set of constraints in such arbitrary lattice $mathcal L$.
+Let $\CL$ be the set of constraints in such a lattice $\mathcal L$.
 The full grammar of constraints is shown in \cref{grammar:constraint}.
@@
 
-\TODO{Add example}
-
+An example of this algorithm in action is shown in \cref{solving:example}.
 Our algorithm is complete, computes principal normal forms,

### 2019-07-11 kindly/soa.tex
checkpoint
cleanup for pre-submission

@@
 \section{Related work}
+\label{sec:related-work}
 
-Haller, P., Odersky, M.: Capabilities for Uniqueness and Borrowing
-\cite{DBLP:conf/ecoop/HallerO10}
+% Haller, P., Odersky, M.: Capabilities for Uniqueness and Borrowing
+% \cite{DBLP:conf/ecoop/HallerO10}
 
-Sing\#
+% Sing\#
 
-John Tang Boyland and William Retert. Connecting effects and
-uniqueness with adoption. In POPL, pages 283–295. ACM, 2005.
-\cite{DBLP:conf/popl/BoylandR05}
+% John Tang Boyland and William Retert. Connecting effects and
+% uniqueness with adoption. In POPL, pages 283–295. ACM, 2005.
+% \cite{DBLP:conf/popl/BoylandR05}
 
@@
 
-Linear Haskell~\citep{DBLP:journals/pacmpl/BernardyBNJS18} aims
-to retrofit Haskell with linear types.
-Linear Haskell introduces a new notion
-of ``linear arrows'', noted $\multimap$ similar to linear logic,
-which are functions who use their argument exactly once.
+The goal of Linear Haskell~\citep{DBLP:journals/pacmpl/BernardyBNJS18}
+(LH) is to retrofit linear types to Haskell.
+LH introduces a new notion
+of ``linear arrows'', written $\multimap$ similar to linear logic,
+which are functions that \emph{use} their argument exactly once.
 % By contrast, in \lang linearity is only decided
@@
 This design is easy to retrofit on top of an existing compiler
-such as GHC, but has several disadvantages\footnote{
+such as GHC, but there are quite controversial discussions about it\footnote{
   See the in-depth discussion attached to the GHC proposal starting here: \url{https://github.com/ghc-proposals/ghc-proposals/pull/111\#issuecomment-403349707}}.
@@
 \begin{itemize}[leftmargin=*]
-\item Their system does not admit subtyping for arrows and leverages
+\item LH does not admit subtyping for arrows and requires
   $\eta$-expansion to pass unrestricted functions in linear
-  contexts. This works in a non-strict language such as
-  Haskell but can not be used in a strict setting.
+  contexts. This approach is acceptable in a non-strict language such as
+  Haskell but changes the semantics in a strict setting.
 \item
-  While the GHC proposal implements some type inference,
+  The GHC implementation of LH performs type inference, but
   it is neither formalized nor complete.
 \item
-  Linear Haskell promotes a continuation-passing style with functions such as 
+  LH promotes a continuation-passing style with functions such as 
   \lstinline/withFile : path -> (file ->. file) ->. unit/
-  to ensure that files are only used linearly.
-  \lang follows System F\degree, Quill and Alms and
-  adopts a direct style, where types themselves are
-  described as affine or linear.
-  We believe this leads to a more natural programming style
-  and ease modular reasoning about linearity.
+  to ensure linear use of resources. This style leads to problems with
+  placing the annotation on, e.g., the IO monad.
+  \lang follows System F\degree, Quill, and Alms, all of which support
+  resource handling in direct style, where types themselves are
+  described as affine or linear. (Or course, continuation-passing
+  style is also supported.)
+  We expect that the direct approach eases modular reasoning about linearity.
   In particular, using abstraction through modules,
@@
 Mezzo~\citep{DBLP:phd/hal/Protzenko14} is an ML-like language
-with an extremely rich capability system which allows it to encode numerous
+with an extremely rich capability system which is able to encode numerous
 properties akin to separation logic~\citep{DBLP:conf/lics/Reynolds02}.
-The main contribution of Mezzo is to push the boundary of what is possible
-to do with a type system. This makes it far more expressive than \lang, but
-also quite harder to use, and precludes any form of complete type
-inference.
+Mezzo explores the  boundaries of the design space of type systems for
+resources. Hence, it is more expressive than \lang, but
+much harder to use. The Mezzo typechecker relies on explicit
+annotations and it is not known whether type inference for Mezzo is possible.
 
@@
 and object systems. In particular, \lang uses borrows and regions
-which were initially developed in objects or
-imperatives settings~\citep{DBLP:conf/popl/BoylandR05,DBLP:conf/pldi/GrossmanMJHWC02}.
+which were initially developed in the context of imperative and
+object-oriented
+programming~\citep{DBLP:conf/popl/BoylandR05,DBLP:conf/pldi/GrossmanMJHWC02}.
 
@@
 \subsection{Type-system features}
-
-\lang relies on qualified types~\cite{DBLP:journals/scp/Jones94}
-to introduce the necessary kind equalities for linear types. In
+%
+\lang relies on constrained types
+to introduce the kind inequalities required for linear types. In
 particular, we use HM(X)~\citep{DBLP:journals/tapos/OderskySW99} which
-allow to use qualified type in an ML-like language with complete
+allow to use constrained types in an ML-like language with complete
 type inference.
 HM(X) has been shown to be compatible with subtyping,
-bounded universal and existentials~\citep{DBLP:conf/icfp/Simonet03}
+bounded universal and existentials~\citep{DBLP:conf/icfp/Simonet03},
 GADTs~\citep{DBLP:journals/toplas/SimonetP07},
-and syntactic proof of soundness~\citep{DBLP:journals/entcs/SkalkaP02}.
-This makes us confident that the system developed in \lang
+and there exists a syntactic soundness proof~\citep{DBLP:journals/entcs/SkalkaP02}.
+These results make us confident that the system developed in \lang
 could be applied to larger and more complex languages such as OCaml.
+
+Alternatively, we could have based the system on qualified
+types~\cite{DBLP:journals/scp/Jones94}, similarly to Quill. This
+choice would also be sustainable as qualified types is part of the
+foundation of Haskell's type system.
 

### 2019-07-11 kindly/split.sh
utility

@@ -1,0 +1,3 @@
+pdfjam -o paper.pdf main.pdf '1-27'
+pdfjam -o appendix.pdf main.pdf '28-'
+

### 2019-07-11 kindly/main.tex
Move Typing stuff around.
Start working on the inference section.
Select the various part of the type inference section.
Principality.
checkpoint-semantics
references before appendix
split-metatheory
Put the appendix into readable shape.
Removing \clearpage gains space, who knew?!
checkpoint
abstract

@@
 
+\usepackage{tikz}
+\usetikzlibrary{decorations.text,backgrounds,positioning,shapes,
+  shadings,shadows,arrows,decorations.markings,calc,fit,fadings,
+  tikzmark
+}
+
 \usepackage[nameinlink,capitalize]{cleveref}
@@
 }
+\newtheorem{property}{Property}
+%\newtheorem{corollary}{Corollary}
 
@@
 \begin{abstract}
-  System programming often requires the manipulation of resources like
+  Systems programming often requires the manipulation of resources like
   file handles, network connections, or dynamically allocated
@@
   manages linearity and affinity properties through kinds. In addition
-  \lang{} supports the mutable and immutable borrowing of affine
+  \lang{} supports the exclusive and unrestricted borrowing of affine
   resources, inspired by features of Rust.
@@
 \input{lang}
-
 \input{infer/infer}
@@
 \bibliography{biblio}
+
+\clearpage
+\appendix
+\input{sdtyping/extra}
+\clearpage
+\input{infer/extra}
+\clearpage
+\input{lang-extra}
+\input{metatheory-extra}
 

### 2019-07-11 kindly/soa.tex

@@
 
+\citet{DBLP:journals/corr/abs-1803-02796} presents
+an extension of OCaml for resource management in the style of C++'s RAII
+and Rust's lifetimes. This system assumes
+the existence of a linear type system and develop the associated compilation
+and runtime infrastructure. We believe our approaches are very
+complementary and aim to combine them in the future.
+
 \subsection{Other substructural type-systems}

### 2019-08-24 kindly/conclusion.tex
first draft

\section{Conclusions}
\label{sec:conclusions}

The \lang language is an extension of ML-like languages with sound handling of linear and affine resources. It's main novel feature is the combination of full type inference and a practically useful notion of shared and exclusive borrowing of linear and affine resources.
Although the inferred types are much richer internally than plain ML types, most of that complexity can be hidden from user-level programmers. On the other hand, programmers of libraries dealing with resources have sufficient expressiveness at their fingertips to express many resource management schemes.

The main restriction of the current system is that the lifetime of borrows is determined by lexical scoping. Overcoming this restriction is subject of future work and will probably require extending the type system by some notion of effect.

%%% Local Variables:
%%% mode: latex
%%% TeX-master: "main"
%%% End:

### 2019-08-26 kindly/conclusion.tex
more description

@@
 The main restriction of the current system is that the lifetime of borrows is determined by lexical scoping.
-Overcoming this restriction is subject of future work and will probably require extending the type system by some notion of effect.
+Overcoming this restriction is subject of future work and will probably require extending the type system by some notion of effect, which is currently discussed in the OCaml community.
+Moreover, other systems rely on existential types for extra expressiveness. We chose not to include existentials to preserve complete type inference, but our design can be extended in this direction.
 

### 2019-09-19 kindly/popl20-rebuttal.md
overhaul#1

Thanks to the reviewers for their thoughtful comments.

# Reviewer A

> How is proper nesting of regions enforced, for both the type system and region inference in Appendix A?  And for the example I asked about above ( $\{|\ldots\{|\ldots|\}^2|\}^3$ )? Or does the paper simply give a nesting intuition that's not necessary because the indices don't actually matter for soundness?

We assume that the type checker operates on a program where
region annotations are inserted by the automatic
annotation phase provided in Appendix A.
So our typing rules do not check the nesting because they are supposed
to work only on programs with well-nested regions (a similar situation
occurs in type-based program analysis, where the analysis assumes a
well-typed program and only reasons about additional properties on values).
Unfortunately, the annotation rules have a typo: The recursive
call in "Rewrite-Region" should be on $n+1$.
Insertion of annotations is completely deterministic (which solves principality concerns)
and ensures that indices grow as regions are nested (which rules out all the given
problematic examples).
This is sufficient since there is no way to hide the level of a borrow
and smuggle it past a region boundary. 
If that was the case, for instance with existential types, we would require a
more sophisticated region system with abstract variables, as suggested
by the reviewer.

We agree that our assumption should be made clearer and will improve
the respective  explanations.

> Please explain why/whether Theorem F.2's statement properly enforces linearity

Peter??

# Reviewer B

## Modules

Affe's usage of subsumption also interacts with type abstraction and modules
For instance, the `File.t` type presented in Figure 1a could be implemented
as a file descriptor (i.e., an integer), which is unrestricted. Subsumption
then allow the unrestricted type to be exposed as affine.
Such mechanism allow programmers to implement their own primitives on
abstract types.
As usual with type-level programming, they must also ensure
the soundness of such primitives
Modules and subsumption are used similarly in Alms, and allow
to emulate the `unsafe` construct in Rust.

# Reviewer C

## Simplification and subkinding

The reviewer ignores Affe's use of subsumption which allows us to
simplify many type signatures without loosing expressiveness.
In particular, as highlighted in section 1.1, the compose function

    let compose f g x = f (g x)
    # compose : (k <= k1) => (b -k-> c) -> (a -k1-> b) -k-> (a -k1-> c)

admits having `f` linear and `g` affine with a linear outcome:
Because affine is a subkind of linear, `g`'s multiplicity can be subsumed to
linear, both k and k1 are instantiated to linear, and the resulting
function type is annotated as linear. This approach follows the
treatment of structural subtyping in HM(X).

## Regions and borrowing

The session types interface that we present is using a functional
style with linear types to demonstrate that this application is in
scope of our system and that  Affe's type inference
handles it gracefully, without extra annotations.

Borrows are more useful for imperative
programming and we are not claiming to apply them to session types.
As explained in Section 2.3, the goal of Affe is to support both
styles, allowing the programmer to choose the most appropriate one for
the task at hand.

Dealing with borrows for session-typed channels would require a major
extension with typestates, which is notoriously incompatible with type
inference. 
However, one could contemplate operations to extract channel metadata
(IP , bytes sent, etc) from a  shared borrow of a channel as in 
`get_address: &(_ channel) -> ip_address`, 
which is typable in Affe.
As the main reference to the channel cannot be used while a borrow is active, such
operations would be sound.

## Formalization

We disagree that the formalization is incomplete.
In particular, the handling of the range of borrows is detailed in Section 3.2.4 on regions.
Furthermore, the lack of deconstruction form for borrows of abstract types is
intended: only the appropriate primitives can act on borrows, because the semantics
of borrow is specific to a given abstract type: For instance the
shared borrow in `read_line : &file -> string` may lead to data races,
whereas `&!file -> string` doesn't.

For space reasons and because the current article focuses on linearity, borrowing,
and inference, we decided not to detail the handling of matching on pairs.
We agree that such aspect is important, and we will expand on this
topic if the article is accepted.

> Related work

????

# Reviewer D

> Partial inference and GADT

Even if a language does not admit full type inference, we believe, 
there should be a well-defined core language which does.
This way, a programmer who steps over the boundary knows when type
annotations are needed.
In particular, it is possible (and desirable) to extend Affe with existential types
by requiring type annotations, similarly to Alms.
The results we present nevertheless garantee that
full type inference is supported in the absence of existentials.

### 2019-09-19 kindly/popl20-rebuttal.md
Add plan.

@@
-Thanks to the reviewers for their thoughtful comments.
+Thanks to the reviewers for their thoughtful comments. Were the paper to be accepted, we would take their comments into account and revise our article in the following manner:
+- Streamline the description of examples in Section 2 to focus on Affe's novelties
+- Present automatic region annotation in the body of the article, and clarifies its role in the type system.
+- Present the pattern matching rules, both typing and semantic.
+- Improve the explanation through new examples for the semantics and the type inference (an example for constraint solving is already present in the appendix).
 
@@
 We agree that our assumption should be made clearer and will improve
-the respective  explanations.
+the associated explanations.
 
@@
 Such mechanism allow programmers to implement their own primitives on
-abstract types.
-As usual with type-level programming, they must also ensure
-the soundness of such primitives
-Modules and subsumption are used similarly in Alms, and allow
-to emulate the `unsafe` construct in Rust.
+abstract types. Programmers are then in charge of ensuring soundness of such primitives. This feature, which is also present in Alms, allows to partially emulate Rust's `unsafe` construct.
 

### 2019-09-19 kindly/popl20-rebuttal.md
more comments

@@
-Thanks to the reviewers for their thoughtful comments. Were the paper to be accepted, we would take their comments into account and revise our article in the following manner:
+Thanks to the reviewers for their thoughtful comments. 
+Were the paper to be accepted, we would take their comments into account and revise our article in the following manner:
 - Streamline the description of examples in Section 2 to focus on Affe's novelties
-- Present automatic region annotation in the body of the article, and clarifies its role in the type system.
+- Present automatic region annotation in the body of the article and clarify its role in the type system.
 - Present the pattern matching rules, both typing and semantic.
@@
 
-Peter??
+We don't understand the construction outlined by the reviewer. We
+believe that the situation cannot arise because the
+different parts of the environment are always disjoint. The
+Region/SRegion rule moves a binding from the resource part to a borrow
+part (mutable or immutable). This part of the formalization is
+unfortunately a bit opaque as it is hidden in the environment
+formation (L 2113-2123).
+The Region typing rule transforms the typing context (for example, L
+2375); the corresponding transformation of the run-time environment is
+left implicit.
+
+We will try to improve the writeup for this issue in the revision.
+
+> The proof of F.2 seems to omit the cases of the semantics that might
+> modify the heap
+
+We put the cases that we felt were most important.
+
+> I cannot find type rules for Create/Observe/Destroy in the supplementary material (my apologies if I missed them somewhere).
+
+That's an unfortunate editing mistake.
 
@@
 
-????
+Correct, Quill comes with a usage-counting semantics. What we meant to
+say is that their semantics isn't store based.
+
+Alms is a seminal work that supports many interesting examples, but it
+comes with a heavy price in terms of complexity (e.g., dependent
+kinds, lub operator in types) and it is not suitable for type
+inference. Most of the examples in the Alms paper exploit the
+genericity of existentials to, e.g., connect resources with
+capabilities. As Affe doesn't have unrestricted existentials, the
+corresponding examples are unlikely to work in Affe as they are stated
+in the Alms paper.
+
+The examples in section 2 are not aimed at comparing with Alms, rather
+we want to demonstrate that the idea put forward in the Alms paper (to
+annotate abstract types in module signatures) also works with Affe's
+type inference.
+
+We will insert a citation of the Gan, Tov, and Morrisett paper with
+the copy-on-write array.
 

### 2019-09-20 kindly/popl20-rebuttal.md
Minor changes.

@@
-Thanks to the reviewers for their thoughtful comments. 
-Were the paper to be accepted, we would take their comments into account and revise our article in the following manner:
+We thanks the reviewers for their thoughtful comments. We will take their comments into account and revise our article in the following manner:
 - Streamline the description of examples in Section 2 to focus on Affe's novelties
@@
 Affe's usage of subsumption also interacts with type abstraction and modules
-For instance, the `File.t` type presented in Figure 1a could be implemented
+For instance, the `File.t` type presented in Section 1 could be implemented
 as a file descriptor (i.e., an integer), which is unrestricted. Subsumption
@@
 genericity of existentials to, e.g., connect resources with
-capabilities. As Affe doesn't have unrestricted existentials, the
+capabilities. In order to provide full type inference
+Affe doesn't have unrestricted existentials, and the
 corresponding examples are unlikely to work in Affe as they are stated
@@
 full type inference is supported in the absence of existentials.
+
+> the paper would strongly benefit from more examples that explain why the type inference algorithm works the way it does.
+
+One such example is available for constraint solving in the Appendix B.1, but we agree
+that it should be extended, and part of the main body.

### 2019-09-20 kindly/popl20-rebuttal.md
cosmetics

--- 2019-09-19 kindly/popl20-rebuttal.md
@@
-Thanks to the reviewers for their thoughtful comments. 
-Were the paper to be accepted, we would take their comments into account and revise our article in the following manner:
-- Streamline the description of examples in Section 2 to focus on Affe's novelties
-- Present automatic region annotation in the body of the article and clarify its role in the type system.
-- Present the pattern matching rules, both typing and semantic.
-- Improve the explanation through new examples for the semantics and the type inference (an example for constraint solving is already present in the appendix).
+We thank the reviewers for their thoughtful comments.
+We will take their comments into account and revise our article as follows.
+* Streamline the description of examples in Section 2 to focus on Affe's novelties
+* Present automatic region annotation in the body of the article and clarify its role in the type system.
+* Present the pattern matching rules, both typing and semantic.
+* Improve the explanation through new examples for the semantics and
+  the type inference (the appendix contains an example for constraint solving).
 
@@
 We assume that the type checker operates on a program where
-region annotations are inserted by the automatic
-annotation phase provided in Appendix A.
-So our typing rules do not check the nesting because they are supposed
+region annotations are inserted by the
+annotation algorithm provided in Appendix A.
+Hence, our typing rules do not check the nesting because they are supposed
 to work only on programs with well-nested regions (a similar situation
@@
 call in "Rewrite-Region" should be on $n+1$.
-Insertion of annotations is completely deterministic (which solves principality concerns)
-and ensures that indices grow as regions are nested (which rules out all the given
+Insertion of annotations is deterministic (which solves principality concerns)
+and ensures that indices grow as regions are nested (which rules out all the
 problematic examples).
-This is sufficient since there is no way to hide the level of a borrow
+This is sufficient because there is no way to hide the level of a borrow
 and smuggle it past a region boundary. 
@@
 
-We don't understand the construction outlined by the reviewer. We
+As far as we understand the construction outlined by the reviewer, we
 believe that the situation cannot arise because the
-different parts of the environment are always disjoint. The
+different parts of the environment are always disjoint. This is
+ensured at a high level by splitting (in the proof case for application). The
 Region/SRegion rule moves a binding from the resource part to a borrow
@@
 
-We will try to improve the writeup for this issue in the revision.
+We will improve the writeup for this issue in the revision.
 
@@
 
-Affe's usage of subsumption also interacts with type abstraction and modules
-For instance, the `File.t` type presented in Figure 1a could be implemented
+Affe's usage of subsumption interacts with type abstraction and modules.
+For instance, the `File.t` type presented in Section 1 could be implemented
 as a file descriptor (i.e., an integer), which is unrestricted. Subsumption
-then allow the unrestricted type to be exposed as affine.
-Such mechanism allow programmers to implement their own primitives on
-abstract types. Programmers are then in charge of ensuring soundness of such primitives. This feature, which is also present in Alms, allows to partially emulate Rust's `unsafe` construct.
+allow us to expose the unrestricted type as an affine abstract type.
+Such a mechanism allows programmers to implement their own primitives on
+abstract types. Programmers are then in charge of ensuring soundness
+of such primitives. This feature, which is also present in Alms,
+can emulate Rust's `unsafe` construct in many cases.
 
@@
 
-The reviewer ignores Affe's use of subsumption which allows us to
-simplify many type signatures without loosing expressiveness.
-In particular, as highlighted in section 1.1, the compose function
+The reviewer overlooks the presence of kind subsumption in Affe, which
+allows us to simplify type signatures without loosing
+expressiveness.  In particular, as highlighted in section 1.1, the
+compose function
 
@@
 
-The session types interface that we present is using a functional
+The session type interface that we present is using a functional
 style with linear types to demonstrate that this application is in
@@
 
-Borrows are more useful for imperative
+Borrows are primarily useful for imperative
 programming and we are not claiming to apply them to session types.
@@
 and inference, we decided not to detail the handling of matching on pairs.
-We agree that such aspect is important, and we will expand on this
-topic if the article is accepted.
+We agree that this aspect is important and we will expand on this
+topic in future revisions.
 
@@
 genericity of existentials to, e.g., connect resources with
-capabilities. As Affe doesn't have unrestricted existentials, the
-corresponding examples are unlikely to work in Affe as they are stated
-in the Alms paper.
+capabilities. 
+Affe doesn't have unrestricted existentials so as to support full type
+inference. The corresponding examples are unlikely to work in Affe as
+they are stated in the Alms paper.
 
@@
 
-We will insert a citation of the Gan, Tov, and Morrisett paper with
-the copy-on-write array.
+We will insert a citation of the Gan, Tov, and Morrisett paper along with
+the copy-on-write array example.
 
@@
 
-Even if a language does not admit full type inference, we believe, 
+Even if a language does not admit full type inference, we believe
 there should be a well-defined core language which does.
@@
 full type inference is supported in the absence of existentials.
+
+> the paper would strongly benefit from more examples that explain why the type inference algorithm works the way it does.
+
+One such example is available for constraint solving in the Appendix B.1, but we agree
+that it should be extended and part of the main body.

### 2019-09-20 kindly/popl20-rebuttal.md
typo.

@@
 The examples in section 2 are not aimed at comparing with Alms, rather
-we want to demonstrate that the idea put forward in the Alms paper (to
-annotate abstract types in module signatures) also works with Affe's
-type inference.
+we want to demonstrate that Alm's use of modules is compatible with Affe's inference, as highlighted in Reviewer B's answer.
 

### 2019-09-20 kindly/popl20-rebuttal.md
Minor changes.

@@
 * Improve the explanation through new examples for the semantics and
-  the type inference (the appendix contains an example for constraint solving).
+  the type inference.
 
@@
 
-## Regions and borrowing
+## Examples, Regions and borrowing
 
@@
 
-> Related work
+## Related work
 
@@
 
-> Partial inference and GADT
+## Partial inference and GADT
 
@@
 
-> the paper would strongly benefit from more examples that explain why the type inference algorithm works the way it does.
+## Examples
 
-One such example is available for constraint solving in the Appendix B.1, but we agree
-that it should be extended and part of the main body.
+We strongly believe programming examples are essential to demonstrate and explain
+Affe's features, as highlighted in reviewer C's answer. We do agree that semantics and inference require more examples, one of which is already available in Appendix B.1.

### 2019-10-25 kindly/infer/infer.tex
Improve placement of figures.
Fixed typos from reviews and own review.

--- 2019-07-11 kindly/infer/infer.tex
@@
 We apply HM(X) to a concrete constraint language which we name $\CL$.
-We adapt and extend HM(X) existing rules to support kind inference,
+We adapt and extend HM(X)'s existing rules to support kind inference,
 track linearity and handle borrows and regions. Finally, we
@@
 The relations used for syntax-directed typing can trivially be defined
-in term of these new relations by using constraint entailment.
+in terms of these new relations by using constraint entailment.
 All relations are fully described in \cref{typ:extra:envs}.
@@
 \subsection{Kind inference}
-
+%
+\begin{figure}[t]
+  \centering
+  \input{infer/kinds}
+  \caption{Kind inference algorithm -- $\inferK{(C,\unif)}{\E}{\tau}{k}$}
+  \label{rules:kinding}
+\end{figure}
+%
 We note $\inferK{(C,\unif)}{\bf{\E}}{\bf{\tau}}{k}$ when type $\tau$ has kind $k$
@@
 
-\begin{figure}[ht]
-  \centering
-  \input{infer/kinds}
-  \caption{Kind inference algorithm -- $\inferK{(C,\unif)}{\E}{\tau}{k}$}
-  \label{rules:kinding}
-\end{figure}
-
 \subsection{Type inference}
@@
 \label{infer:envs}
-
-In our syntax-directed system, we ensured
-that linear variables are not discarded at the \emph{leafs}, through
-the {\sc Var} rule. In the inference algorithm, we operate in the opposite
-direction: we collect data from the leafs, and only enforce linearity
-at \emph{binders}. This is reflected in the {\sc Var$_I$} and
-{\sc Abs$_I$} rules presented in \cref{rule:infer:envs}.
-Typing for variables is very similar to traditional (non-affine) ML
-typing. The only difference is that we collect
-the fact that $x$ was used with the scheme $\schm$ by returning
-a usage environment $\Sv = \{ \bvar{x}{\schm} \}$.
 %
-This usage environment is in turn used at each binder to enforce proper
-usage of linear variable via the $\Weaken$ property.
-We demonstrate how this is achieved for lambda expressions
-in the {\sc Abs$_I$} rule.
-First, we typecheck the body of the lambda and obtain a usage
-environment $\Sv_x$. Just like in the syntax directed,
-we introduce the constraint
-$\Cleq{\Sv\Sdel{x}}{\kvar}$ which properly accounts for captures in
-the body of the lambda expression. We also introduce the constraint
-$\Weaken_{\bvar{x}{\sigma}}(\Sv)$. If $x$ was used $e$, $x$ is present
-in $\Sv$ and this constraint is empty. Otherwise, if $x$ is
-\emph{not} present in $\Sv$, its type should not have a linear
-kind, which is enforced by the constraints $\Cleq{\sigma}{\kaff_\infty}$.
-The $\Weaken$ constraint is introduce at each binding point in the program
-such as let-bindings and pattern matching.
-
-Finally, we use normalization on complex constraints to ensure
-that the inference algorithm always return the simplest possible
-constraints and unifiers.
-
-\begin{figure}[!h]
+\begin{figure}[tp]
   \begin{mathpar}
@@
 \end{figure}
+%
+In our syntax-directed system, we ensured
+that linear variables are not discarded at the \emph{leaves}, through
+the {\sc Var} rule. In the inference algorithm, we operate in the opposite
+direction: we collect data from the leaves, and only enforce linearity
+at \emph{binders}. This is reflected in the {\sc Var$_I$} and
+{\sc Abs$_I$} rules presented in \cref{rule:infer:envs}.
+Typing for variables is very similar to traditional (non-affine) ML
+typing. The only difference is that we collect
+the fact that $x$ was used with the scheme $\schm$ by returning
+a usage environment $\Sv = \{ \bvar{x}{\schm} \}$.
+%
+This usage environment is in turn used at each binder to enforce proper
+usage of linear variable via the $\Weaken$ property.
+We demonstrate how this is achieved for lambda expressions
+in the {\sc Abs$_I$} rule.
+First, we typecheck the body of the lambda and obtain a usage
+environment $\Sv_x$. Just like in the syntax-directed type system,
+we introduce the constraint
+$\Cleq{\Sv\Sdel{x}}{\kvar}$ which properly accounts for captures in
+the body of the lambda expression. We also introduce the constraint
+$\Weaken_{\bvar{x}{\sigma}}(\Sv)$. If $x$ was used $e$, $x$ is present
+in $\Sv$ and this constraint is empty. Otherwise, if $x$ is
+\emph{not} present in $\Sv$, its type should not have a linear
+kind, which is enforced by the constraints $\Cleq{\sigma}{\kaff_\infty}$.
+The $\Weaken$ constraint is introduced at each binding point in the program
+such as let-bindings and pattern matching.
+
+Finally, we use normalization on complex constraints to ensure
+that the inference algorithm always return the simplest possible
+constraints and unifiers.
 
@@
 \label{infer:regions}
-
-We now present the inference version
-of the {\sc Pair} and {\sc Region} rules in \cref{rule:infer:envrules}.
-While the rules are quite similar to the original version, the usage
-environment $\Sv$ is now an \emph{output} of the algorithm.
-As such, we use the ``inference'' version of the relations on
-environment,
-$\bsplit{C}{\Sv}{\bf{\Sv_1}}{\bf{\Sv_2}}$ and $\bregion{C}{\bf{x}}{\Sv}{\bf\Sv'}$,
-which return the necessary constraints.
-As appropriate for inference, these relations 
-are total and non ambiguous for their input parameters ($\Sv_1, \Sv_2$ and $\Sv'$, respectively).
-We then collect all constraints and normalize them.
-
-\begin{figure}[!h]
+\begin{figure}[tb]
   \centering
@@
 \end{figure}
+%
+We now present the inference version
+of the {\sc Pair} and {\sc Region} rules in \cref{rule:infer:envrules}.
+While the rules are quite similar to the original version, the usage
+environment $\Sv$ is now an \emph{output} of the algorithm.
+As such, we use the ``inference'' version of the relations on
+the environment,
+$\bsplit{C}{\Sv}{\bf{\Sv_1}}{\bf{\Sv_2}}$ and $\bregion{C}{\bf{x}}{\Sv}{\bf\Sv'}$,
+which returns the necessary constraints.
+As appropriate for inference, these relations
+are total and non ambiguous for their input parameters ($\Sv_1, \Sv_2$ and $\Sv'$, respectively).
+We then collect all constraints and normalize them.
 
 \subsubsection{Generalization and constraints}
+%
+\begin{figure}[tbp]
+  \centering
+  \begin{mathpar}
+    \ruleILet
 
+    {\generalize{C}{\E}{\tau} =
+      (\Cproj{\Multi{\kvar},\Multi{\tvar}}{C},
+      \forall \Multi{\kvar},\Multi{\tvar}.\qual{C}{\tau})
+
+      \text{ where }
+
+      \Multi{\kvar},\Multi{\tvar} = (\fv{\tau}\cup\fv{C})\setminus\fv{\E}
+    }
+  \end{mathpar}
+  \caption{The {\sc Let$_I$} rule}
+  \label{rule:infer:let}
+\end{figure}
+%
 So far, constraints have only been composed of a list of kind inequalities.
@@
 allow the creation of local variables inside the constraints.
-This allow to encapsulate all the quantified variables in the global constraints.
-It also reflect the fact that there
-should exists at least one solution for $C$ for the scheme to be valid.
+This allows us to encapsulate all the quantified variables in the global constraints.
+It also reflects the fact that there
+should exist at least one solution for $C$ for the scheme to be valid.
 \citet{DBLP:journals/tapos/OderskySW99} give a detailed account
 on the projection operators in HM inference.
-
-\begin{figure}[!h]
-  \centering
-  \begin{mathpar}
-    \ruleILet
-
-    {\generalize{C}{\E}{\tau} =
-      (\Cproj{\Multi{\kvar},\Multi{\tvar}}{C},
-      \forall \Multi{\kvar},\Multi{\tvar}.\qual{C}{\tau})
-
-      \text{ where }
-
-      \Multi{\kvar},\Multi{\tvar} = (\fv{\tau}\cup\fv{C})\setminus\fv{\E}
-    }
-  \end{mathpar}
-  \caption{The {\sc Let$_I$} rule}
-  \label{rule:infer:let}
-\end{figure}
 
@@
 %
-We now show that this algorithms is well-behaved with respect to entailment
+We now show that this algorithm is well-behaved with respect to entailment
 and computes unique normal forms.
@@
 The first theorem states that our inference algorithm is sound
-with respect to the syntax directed type system.
+with respect to the syntax-directed type system.
 
@@
   Given a type environment $\E$ containing only value bindings,
-  $\E|_\tau$ containing only type binding and a term $e$,\\
+  $\E|_\tau$ containing only a type binding and a term $e$,\\
   if $\inferW{\Sv}{(C,\unif)}{\E;\E_\tau}{e}{\tau}$
@@
 The second theorem states that our algorithm is complete: for any given
-syntax directed typing derivation, our inference algorithm can find
+syntax-directed typing derivation, our inference algorithm can find
 a derivation that gives a type at least as general.

### 2019-10-25 kindly/intro.tex
Fixed typos from reviews and own review.

@@
 protocol that prescribes the correct use of its API.
-For examples, a file handle appears as the result of of opening a
-file. If it was opened for reading, then read operations will suceed,
+For examples, a file handle appears as the result of opening a
+file. If it was opened for reading, then read operations will succeed,
 but write operations will fail. Once the handle is closed, it cannot

### 2019-10-25 kindly/lang.tex
One more.
Fixed typos from reviews and own review.

@@
 The rest of this section formalizes  \lang: the syntax (\cref{syntax}),
-the statics in terms of syntax directed typing (\cref{sdtyping}),
+the statics in terms of syntax-directed typing (\cref{sdtyping}),
 and a dynamics that is linearity- and resource-aware (\cref{sec:sem}).
@@
 then we need to use the reborrow expression $\reborrow{x}$.
-
+%
 A region $\region{\Sone x\BORROW}{e}$ is annotated with its nesting $n$, the variable $x$ that may be borrowed in $e$, and the kind of borrow $\BORROW$.
-
+%
 A match is indexed with a \emph{match specification} $\etransfm$ that indicates
@@
 
-\begin{figure*}[t]
+\begin{figure*}[b]
   \centering
@@
 The \emph{region} around its use ensures that the borrow in the left component does not
-escape, which brings us to next topic.
+escape, which brings us to the next topic.
 
@@
 To guarantee the validity of a borrow, its lifetime must be properly contained in its
-ancestors lifetime. \lang ensures proper nesting of lifetimes by using
+ancestor's lifetime. \lang ensures proper nesting of lifetimes by using
 regions. The expression $\region{\Sone x\BORROW}{e}$ indicates a
@@
 To constrain the borrow to this region we impose the constraint
-$\entail{C}{(\kun_n\lk k) \wedge (\lk\kun_\infty)}$, which affirms
+$\entail{C}{(\kun_n\lk k) \wedge (k \lk\kun_\infty)}$, which affirms
 that the borrow is unrestricted, but can only be used in nesting
 levels $n$ and higher.
-WIth that setup we typecheck the body of the region,
+With that setup we typecheck the body of the region,
 $f(\borrow[\IBORROW]{c})$.
@@
 
-Expressions $e$ need to be elaborted with some kind and constraint
+Expressions $e$ need to be elaborated with some kind and constraint
 annotations. Polymorphic functions give rise to elaborated

### 2019-10-25 kindly/sdtyping/extra.tex
Fixed typos from reviews and own review.

@@
 The base cases are constants, variables and borrows.
-The general idea is to start from the leafs of the syntax tree, create a
+The general idea is to start from the leaves of the syntax tree, create a
 region for each borrow, and enlarge the region as much as possible.
@@
 \label{appendix:constraints}
+%
+\begin{figure}[tb]
+  \centering
+  \begin{align*}
+    C &::= \Cleq{\tau_1}{\tau_2}
+        \mid \Cleq{k_1}{k_2}
+        \mid C_1 \Cand C_2
+        \mid \Cproj{\tvar}{C}
+        \mid \Cproj{\kvar}{C}
+  \end{align*}
+  \caption{The constraint language}
+  \label{grammar:constraint}
+  % \input{infer/lattice}
+  % \caption{Lattice inequalities -- $k \lk_\Lat k'$}
+  \input{infer/entails}
+  \caption{Base entailment rules -- $\entail{C}{D}$ }
+  \label{rules:entail}
+\end{figure}
 
@@
 The normalization function is described in \cref{infer:solving}.
-%
-\begin{figure}[!hb]
-  \centering
-  \begin{align*}
-    C &::= \Cleq{\tau_1}{\tau_2}
-        \mid \Cleq{k_1}{k_2}
-        \mid C_1 \Cand C_2
-        \mid \Cproj{\tvar}{C}
-        \mid \Cproj{\kvar}{C}
-  \end{align*}
-  \caption{The constraint language}
-  \label{grammar:constraint}
-  % \input{infer/lattice}
-  % \caption{Lattice inequalities -- $k \lk_\Lat k'$}
-  \input{infer/entails}
-  \caption{Base entailment rules -- $\entail{C}{D}$ }
-  \label{rules:entail}
-\end{figure}
 
@@
 
-\section{Syntax directed typing}
+\section{Syntax-directed typing}
 \label{appendix:sdtyping}
@@
 \subsection{Kinding}
-
-\begin{figure}[!h]
+%
+\begin{figure}[tb]
   \centering
@@
 
-The rules for syntax directed typing are shown in \cref{fig:syntax-directed-typing}
+The rules for syntax-directed typing are shown in \cref{fig:syntax-directed-typing}
 and follow the presentation given in \cref{sdtyping}.

### 2019-10-25 kindly/soa.tex
Fixed typos from reviews and own review.

@@
 to structural subtyping, where the only subtyping (or here, subkinding)
-is at the leafs.
+is at the leaves.
 Such discipline is known to be friendly to inference and has be used in many
@@
 is the design of a kind language specialized for checking linearity
-and sufficient simple to make
+and sufficiently simple to make
 all simplification rules complete, which allows to keep type signatures simple.

### 2019-11-07 kindly/lang.tex
Adapt to the PLDI format.

@@
 rule splits off an immutable suspended binding from a borrow.
-
+%
 Splitting applies whenever an
 expression has multiple subexpressions:  pairs, function applications,
-and let bindings. For instance, in the
+and let bindings. In the
 expression

### 2019-11-07 kindly/motivation.tex
Fixed typos from reviews and own review.
Adapt to the PLDI format.

@@
 \end{lstlisting}
+    \vspace{-10pt}
     \caption{Addition server}
@@
 % # ('m1 ?'m2 ? 'result ! end) st -> 'm1 -{lin}> 'm2 -{^k}> 'result
+    \vspace{-10pt}
     \caption{Binary operators client}
   \end{subfigure}
+  \vspace{-10pt}
   \caption{Corresponding session type programs in \lang}
@@
 The general form of
-arrows types in \lang is \lstinline/-{k}>/, where \lstinline/kk/ is a
+arrow types in \lang is \lstinline/-{k}>/, where \lstinline/kk/ is a
 kind that restricts the number of uses of the function.  For
@@
   \centering
-  \begin{subfigure}[t]{0.47\linewidth}
+  \begin{subfigure}[t]{1\linewidth}
 \begin{lstlisting}
@@
   type ('a : 'k) t : lin
-
   val create : ('a : un) => int -> 'a -> 'a t
   val free : ('a : aff) => 'a t -> unit
-
   val length : &('a t) -> int
-
-  val get :
-    ('a : un) => &('a t) * int -> 'a
-  val set :
-    ('a : aff) => &!('a t) * int * 'a -> unit
-
+  val get : ('a : un) => &('a t) * int -> 'a
+  val set : ('a : aff) => &!('a t) * int * 'a -> unit
   val map : (&'a -> 'b) -> &('a t) -> 'b t
@@
 \end{lstlisting}
+    \vspace{-15pt}
     \caption{Affine arrays}
     \label{sig:array}
-  \end{subfigure}\hfill
-  \begin{subfigure}[t]{0.53\linewidth}
+  \end{subfigure}
+
+  \begin{subfigure}[t]{1\linewidth}
 \begin{lstlisting}
@@
 \end{lstlisting}
+    \vspace{-10pt}
     \caption{Example of use of linear arrays}
@@
 \end{lstlisting}
+    \vspace{-10pt}
     \caption{Excerpt of \cref{ex:array} with explicit regions}
@@
   \centering
-  \begin{subfigure}{0.48\linewidth}
+  \begin{subfigure}{1\linewidth}
 \begin{lstlisting}
@@
   type ('a : un) t : lin
-
   val create : int -> 'a -> 'a t
   val free : 'a t -> unit
-
   val length : &('a t) -> int
   val get : &('a t) * int -> 'a
-
   val set : &('a t) * int * 'a -> 'a t
@@
 \end{lstlisting}
+    \vspace{-15pt}
     \caption{Signature}
     \label{sig:hybarray}
-  \end{subfigure}\hfill
-  \begin{subfigure}{0.5\linewidth}
+  \end{subfigure}
+
+  \begin{subfigure}{1\linewidth}
 \begin{lstlisting}
@@
   include Array
-
   let set (a, i0, x0) =
-    let f i x =
-      if i = i0 then x0 else x
-    in
-    Array.mapi f a
-
+    Array.mapi (fun i x -> if i = i0 then x0 else x) a
   let set_mut = Array.set
@@
 \end{lstlisting}
+    \vspace{-15pt}
     \caption{Implementation using copy-on-write}
@@
   \end{subfigure}
-  \caption{Hybrid arrays -- signature and example implementation}
+  \caption{Hybrid arrays}
 \end{figure}
@@
   \centering
-  \begin{minipage}[t]{0.45\linewidth}
-    \lstinputlisting[linerange=1-13]{code/sukodu.affe}
-  \end{minipage}\hfill
-  \begin{minipage}[t]{0.54\linewidth}
-    \lstinputlisting[linerange=15-31,firstnumber=last]{code/sukodu.affe}
-  \end{minipage}
+  \lstinputlisting[linerange=1-31]{code/sukodu.affe}
+    \vspace{-15pt}
   \caption{Excerpt of the Sudoku solver}
@@
   \centering
-  \begin{subfigure}[t]{.58\linewidth}
+  \begin{subfigure}[t]{1\linewidth}
 \begin{lstlisting}
@@
 \end{lstlisting}
+    \vspace{-10pt}
     \caption{Interface}
@@
 \end{lstlisting}
+    \vspace{-10pt}
     \caption{Example of use}
     \label{ex:pool}
-  \end{subfigure}~
-  \begin{subfigure}[t]{.43\linewidth}
+  \end{subfigure}
+
+  \begin{subfigure}[t]{1\linewidth}
 \begin{lstlisting}
@@
 \end{lstlisting}
+    \vspace{-15pt}
     \caption{Implementation}

### 2019-11-07 kindly/infer/extra.tex
Put the appendix into readable shape.
Start improving the inference section.

@@
 
-\begin{figure*}[p]
-  \input{infer/typing}
-  \caption{Type inference rules -- $\inferW{\Sigma}{(C,\psi)}{\bf{\E}}{\bf{e}}{\tau}$ }
-  \label{rules:typing:full}
-\end{figure*}
-
 In this appendix, we provide the complete type inference rules
 and show that our type inference algorithm is sound and complete.
-The kinding and constraints rules are already shown in \cref{inference}.
+The constraints rules are already shown in \cref{inference}.
+Kind inference is presented in \cref{rules:kinding:full}.
 The type inference rules are shown in \cref{rules:typing:full}.
@@
 
-\clearpage
+\subsection{Kind inference}
+\label{rules:kinding:full}
+%
+\begin{figure}[tb]
+  \centering
+  \input{infer/kinds}
+  \caption{Kind inference rules -- $\inferK{(C,\unif)}{{\bm\E}}{{\bm\tau}}{k}$}
+  \label{rules:kinding}
+\end{figure}
+%
+We note $\inferK{(C,\unif)}{{\bm\E}}{{\bm\tau}}{k}$ when type $\tau$ has kind $k$
+in environment $\E$ under constraints $C$ and unifier $\unif$.
+$\bm\E$ and $\bm\tau$ are the input parameters of
+our inference procedure.
+We present the kind inference algorithm as a set of rules in
+\cref{rules:kinding}.
+Higher-kinds are not generally supported
+and can only appear by looking-up the kind scheme of a type constructor,
+for use in the type application rule {\sc KApp}.
+Type variables must be of a simple kind in rule {\sc KVar}.
+Kind schemes are instantiated in the {\sc KVar} rules by creating
+fresh kind variables and the associated substitution.
+{\sc KArr} and {\sc KBorrow} simply returns the kind of the primitive
+arrow and borrow types.
+The $\operatorname{normalize}$ function is used every time several constraints
+must be composed in order to simplify the constraint and return a most general
+unifier.
+
+\begin{figure*}[!btp]
+  \input{infer/typing}
+  \caption{Type inference rules --
+    $\inferW{\Sigma}{(C,\psi)}{{\bm\E}}{{\bm e}}{\tau}$ }
+  \label{rules:typing:full}
+\end{figure*}
+
 % \begin{figure}[tp]

### 2019-11-07 kindly/infer/infer.tex
Bring back many of the figures.
Start improving the inference section.

@@
 \begin{itemize}[leftmargin=*]
-\item $\bsplit{C}{\Sv}{\bf{\Sv_1}}{\bf{\Sv_2}}$.
+\item $\bsplit{C}{\Sv}{\bm{\Sv_1}}{\bm{\Sv_2}}$.
   Given two usage environments $\Sv_1$ and $\Sv_2$
@@
   and uses the same rules as the one presented in \cref{sdtyping:split}.
-\item $\bregion{C}{\bf{x}}{\Sv}{\bf\Sv'}$.
+\item $\bregion[\bm{n}]{C}{\bm{x}}{\Sv}{\bm\Sv'}$.
   Given a usage environment $\Sv'$ and a variable name $x$, we return
@@
 
-The HM(X) framework assumes the existence of a $\operatorname{normalize}$
+The HM(X) framework assumes the existence of a ``$\operatorname{normalize}$''
 function which takes a constraint $C$, a substitution $\psi$ and returns a
@@
 
-\subsection{Kind inference}
-%
-\begin{figure}[t]
-  \centering
-  \input{infer/kinds}
-  \caption{Kind inference algorithm -- $\inferK{(C,\unif)}{\E}{\tau}{k}$}
-  \label{rules:kinding}
-\end{figure}
-%
-We note $\inferK{(C,\unif)}{\bf{\E}}{\bf{\tau}}{k}$ when type $\tau$ has kind $k$
-in environment $\E$ under constraints $C$ and unifier $\unif$.
-$\E$ and $\tau$ are the input parameters of
-our inference procedure.
-We present the kind inference algorithm as a set of rules in
-\cref{rules:kinding}.
-Higher-kinds are not generally supported
-and can only appear by looking-up the kind scheme of a type constructor,
-for use in the type application rule {\sc KApp}.
-Type variables must be of a simple kind in rule {\sc KVar}.
-Kind schemes are instantiated in the {\sc KVar} rules by creating
-fresh kind variables and the associated substitution.
-{\sc KArr} and {\sc KBorrow} simply returns the kind of the primitive
-arrow and borrow types.
-The $\operatorname{normalize}$ function is used every time several constraints
-must be composed in order to simplify the constraint and return a most general
-unifier.
-
 \subsection{Type inference}
 
-We note $\inferW{\Sv}{(C,\unif)}{\bf{\E}}{\bf{e}}{\tau}$ when
+We note $\inferW{\Sv}{(C,\unif)}{\bm{\E}}{\bm{e}}{\tau}$ when
 $e$\ has type $\tau$ in $\E$ under the constraints $C$ and unifier $\unif$
-with a usage environment $\Sv$. $\E$ and $e$ are the input parameters of our
+with a usage environment $\Sv$. $\bm\E$ and $\bm e$ are the input parameters of our
 inference algorithm.
@@
 Suspended and borrow bindings can only be present in $\Sv$.
-We revisit the syntax-directed rules presented in \cref{sdtyping} to highlight the
-novelties
+We revisit some of the syntax-directed rules presented in \cref{sdtyping}
+to highlight the novelties
 of our inference algorithm and the differences with the syntax-directed
@@
 
+    \inferrule{}
     {\generalize{C}{\E}{\tau} =
       (\Cproj{\Multi{\kvar},\Multi{\tvar}}{C},
-      \forall \Multi{\kvar},\Multi{\tvar}.\qual{C}{\tau})
-
-      \text{ where }
-
+      \forall \Multi{\kvar},\Multi{\tvar}.\qual{C}{\tau})\\
+      \text{ where}\quad
       \Multi{\kvar},\Multi{\tvar} = (\fv{\tau}\cup\fv{C})\setminus\fv{\E}
@@
 
-\begin{figure}[!bt]
+\begin{figure*}[!bt]
   \centering
@@
   \end{minipage}
-\end{figure}
+\end{figure*}
 
@@
 We now show that our extended inference algorithm still satisfies the soundness
-and completeness properties of HM(X). This is achieved by extending
-the original proofs from HM(X), which is done in \cref{appendix:infer}.
+and completeness properties of HM(X).
+% This is achieved by extending
+% the original proofs from HM(X), which is done in \cref{appendix:infer}.
 %
@@
   Given a type environment $\E$ containing only value bindings,
-  $\E|_\tau$ containing only a type binding and a term $e$,\\
-  if $\inferW{\Sv}{(C,\unif)}{\E;\E_\tau}{e}{\tau}$
+  $\E|_\tau$ containing only type bindings, and a term $e$:\\
+  if $\inferW{\Sv}{(C,\unif)}{\E;\E_\tau}{e}{\tau}$\\
   then $\inferS{C}{\unif(\Sv;\E_\tau)}{e}{\tau}$, $\unif C = C$ and $\unif \tau = \tau$
@@
     \vee \svar{x}{\tau}^n\in\E
-    \right\} \cup \left\{ \bvar{\tvar}{k} \mid \bvar{\tvar}{k}\in\E \right\}
+    \right\}\\
+  &\cup \left\{ \bvar{\tvar}{k} \mid \bvar{\tvar}{k}\in\E \right\}
 \end{align*}

### 2019-11-11 kindly/lang.tex
Bring back many of the figures.
3.2 Typing: Fixed typos.

@@
 
-\begin{figure*}[b]
+\begin{figure*}[!tb]
   \centering
@@
 
-\affe uses of kinds and constrained types to indicate
+\affe uses kinds and constrained types to indicate
 linear and affine types.
@@
 where affine ones are expected and affine ones are less restricted
-than linear one as reflected in \TirName{Lat-UA} and \TirName{Lat-AL}.
+than linear ones as reflected in \TirName{Lat-UA} and \TirName{Lat-AL}.
 Second, values usable at level $n$ may be used at any more deeply
@@
 \label{sdtyping:envs}
-\begin{figure}[tp]
-  \begin{minipage}{0.33\linewidth}
+\begin{figure*}[tp]
+  \begin{minipage}{0.28\linewidth}
     \begin{align*}
@@
   \end{minipage}\hfill
-  \begin{minipage}{0.65\linewidth}
+  \begin{minipage}{0.7\linewidth}
     \begin{mathpar}
@@
 \label{sdtyping:split}
-\begin{figure}[tp]
+\begin{figure*}[tp]
   \centering

### 2019-11-11 kindly/infer/infer.tex
Fixed uses of 'note' instead of 'denote' and various typos.

@@
 are input and output or our inference judgement is critical. In the rest
-of this section, when presenting a new judgement
-we will note in \textbf{bold} the input parameters. The rest will then be
+of this section, when presenting a new judgement,
+we write input parameters in \textbf{bold}. The rest will then be
 output parameters.
@@
 % returns a type environment, which we call ``usage environment'' and commonly
-% note $\Sv$, to summarize how variables and borrows are used inside
+% write $\Sv$, to summarize how variables and borrows are used inside
 % the expression.
@@
 are used. Usage environments follow the exact same grammar
-as normal environment. In order to distinguish them more easily,
-we note them $\Sv$.
+as normal environments. In order to distinguish them more easily,
+we denote them by $\Sv$.
 
@@
 
-We note $\inferW{\Sv}{(C,\unif)}{\bm{\E}}{\bm{e}}{\tau}$ when
+We write $\inferW{\Sv}{(C,\unif)}{\bm{\E}}{\bm{e}}{\tau}$ when
 $e$\ has type $\tau$ in $\E$ under the constraints $C$ and unifier $\unif$
@@
 on let-bindings.
-Following HM(X), we note $(C_\schm, \schm) = \generalize{C}{\E}{\tau}$
-the pair of a constraint and a scheme resulting from
+Following HM(X), we write $(C_\schm, \schm) = \generalize{C}{\E}{\tau}$
+for the pair of a constraint and a scheme resulting from
 generalization. The definition is provided in \cref{rule:infer:let}.
@@
 The generated constraint $C_\schm$ uses a new projection operator,
-noted $\Cproj{\tvar}{D}$, which
+$\Cproj{\tvar}{D}$, which
 allow the creation of local variables inside the constraints.
@@
 constraints. Since types might contain kinds (for instance, on the arrows),
-type unification is oriented and noted $\leq$.
+type unification is oriented and written as $\leq$.
 For simplicity, we consider all type constructors
@@
 
-As before, entailment is noted $\entail{C}{D}$,
+As before, entailment is denoted by $\entail{C}{D}$,
 where $D$ is a consequence of the constraints $C$.
-We say that $C$ and $D$ are equivalent, noted $C \equivC D$,
+We say that $C$ and $D$ are equivalent, $C \equivC D$,
 when $\entail{C}{D}$ and $\entail{D}{C}$.
@@
 %
-We now define our normalization procedure, noted $\normalize{C_0}{\unif_0}$ where
+We now define our normalization procedure $\normalize{C_0}{\unif_0}$, where
 $C_0\in \CL$ is a set of constraints and $\unif_0$ is a substitution.
@@
 
-  We note $\mathcal R$ the relation $\Cleq{k_j}{k'_j}_j$,
+  We write $\mathcal R$ for the relation $\Cleq{k_j}{k'_j}_j$,
   $\mathcal G$ the underlying directed graph and $V$ its vertices.
@@
 \item For each strongly connected component in $\mathcal G$, unify all its vertices and replace it by a representative.
-  We note $\unif_k$ the substitution that replaces a kind variable by
+  We write $\unif_k$ for the substitution that replaces a kind variable by
   its representative.
@@
 A flattened environment,
-noted $\Eflat\E$, is the environment
+written as $\Eflat\E$, is the environment
 where all the binders are replace by normal ones. More formally:

### 2019-11-11 kindly/sdtyping/extra.tex
Reformat the annex a bit.
Start improving the inference section.
Fixed uses of 'note' instead of 'denote' and various typos.

@@
 %
-\begin{figure*}[b]
+\begin{figure*}[!tb]
   \centering
@@
 Our system is defined in \cref{infer:solving}.
-As before, we note $\CL$ our constraint system for an arbitrary
+As before, we write $\CL$ for our constraint system for an arbitrary
 total bounded lattice $(\mathcal L, \lk_\Lat)$.
-Entailment is noted $\entail{C}{D}$, where $D$ is a consequence of the
+Entailment is written $\entail{C}{D}$, where $D$ is a consequence of the
 constraints $C$.
-We say that $C$ and $D$ are equivalent, noted $C \equivC D$,
+We say that $C$ and $D$ are equivalent, written as $C \equivC D$,
 when $\entail{C}{D}$ and $\entail{D}{C}$.
@@
 %
-We note $\SC$ the set of solved forms
+We write $\SC$ for the set of solved forms,
 which can be used inside type and kind schemes.
@@
 
-  We note $(C^k,\unif_\tau)$ the result of
+  We write $(C^k,\unif_\tau)$ for the result of
   the modified Herbrand unification on $(D,\phi)$ in step (1).
@@
 
-  We note $P(C_a)$ if for all $C = (k, k)'$
+  We write $P(C_a)$ if for all $C = (k, k)'$
   such that $\entail{C_a}{C}$ and $\nvdash_eC$,
@@
 
-  Let us note $C_1'$ and $C_2'$ the constraints after step (4). $P(C_1')$ and
+  Let us write $C_1'$ and $C_2'$ for the constraints after step (4). $P(C_1')$ and
   $P(C_2')$ hold. Indeed, since $C_1'$ and $C_2'$ are only composed
@@
 
-  Let us note $C''_a$ for $a\in\{1,2\}$ the constraints after step (5).
+  Let us write $C''_a$ for $a\in\{1,2\}$ the constraints after step (5).
   Since there are no more existential variables,
@@
 %
-\begin{figure}[tb]
+\begin{figure}[bt]
   \centering
@@
 
-We note $\inferSK{C}{\E}{\tau}{k}$
-when $\tau$ has kind $k$ in environment $\E$ under constraints $C$.
+We write $\inferSK{C}{\E}{\tau}{k}$
+if $\tau$ has kind $k$ in environment $\E$ under constraints $C$.
 The rules are shown in \cref{rules:sd-kinding}.
@@
 the rewriting relation on environment constraints,
-noted $\Cleq\E k \Crewrite  C$ which rewrites a constraint of the form
+$\Cleq\E k \Crewrite C$, which rewrites a constraint of the form
 $\Cleq\E k$ into $C$. It proceeds by iterating over the environment
@@
 
-\begin{figure}[!tb]
+\begin{figure*}[!tb]
   \input{sdtyping/sd-splitting}
-  \caption{Splitting --- environments $\lsplit
-    C\E\E\E$; binders $\bsplit Cbbb$}
+  \caption{Splitting ---
+    environments $\lsplit C\E{\E_l}{\E_r}$;
+    inference $\bsplit C\E{\bm\E_l}{\bm\E_r}$;
+    binders $\bsplit Cb{\bm b_r}{\bm b_l}$}
   \label{fig:sd-splitting}
   \input{sdtyping/sd-borrowing}
-  \caption{Borrowing --- environments $\lregion
-    C{x}\E\E$; binders $\bregion C{x}bb$}
+  \caption{Borrowing ---
+    environments $\lregion[n]{C}{x}{\E}{\E'}$;
+    inference $\bregion[\bm n]{C}{\bm x}{\E}{\bm\E'}$;
+    binders $\bregion[\bm n]{C}{\bm x}{b}{\bm b'}$}
   \label{fig:sd-borrowing}
   \input{sdtyping/sd-environments}
-  \caption{Rewriting constraints on environments --- $\Cleq\E k \Crewrite  C$}
+  \caption{Rewriting constraints on environments --- $\bm{\Cleq{\E}{k}}\Crewrite  C$}
   \label{fig:contraints-environments-types}
-\end{figure}
+\end{figure*}
 

### 2019-11-11 kindly/soa.tex
SOA: Fixed typo.
SOA: Fixed typos.

@@
 types through the use of qualified types.
-Quill does not exposes a kind language, but
+Quill does not expose a kind language, but
 uses typeclass-like mechanism and annotation on arrows.
@@
 
-Linear and affine discipline have also been used in non-functional
+Linear and affine disciplines have also been used in non-functional
 settings, notably in the context of low-level imperative programming
@@
 particular, we use HM(X)~\citep{DBLP:journals/tapos/OderskySW99} which
-allow to use constrained types in an ML-like language with complete
+allows to use constrained types in an ML-like language with complete
 type inference.
 HM(X) has been shown to be compatible with subtyping,
-bounded universal and existentials~\citep{DBLP:conf/icfp/Simonet03},
+bounded quantification and existentials~\citep{DBLP:conf/icfp/Simonet03},
 GADTs~\citep{DBLP:journals/toplas/SimonetP07},
@@
 is at the leaves.
-Such discipline is known to be friendly to inference and has be used in many
+Such discipline is known to be friendly to inference and has been used in many
 contexts, including OCaml itself, and has been combined

### 2019-11-11 kindly/sdtyping/extra.tex
Fixed typos and renamed reborrow-rule from 'Borrow' to 'Reborrow'.

@@
 All the definitions are made in term of the inference version, which
-returns a fresh constraints. The solving version then simply
+returns fresh constraints. The solving version then simply
 uses entailment, as shown in rule {\sc ESplit-Check} and
@@
 $\Cleq\E k$ into $C$. It proceeds by iterating over the environment
-and expending the constraints for each binding.
+and expanding the constraints for each binding.
 Suspended bindings are rejected ({\sc ConstrSusp}).
 Borrow bindings directly use the annotated kind ({\sc ConstrBorrow}).
-Other bindings uses the underlying type scheme ({\sc ConstrBinding}).
+Other bindings use the underlying type scheme ({\sc ConstrBinding}).
 Type schemes are constrained by first inferring the kind, and then

### 2019-11-11 kindly/infer/infer.tex
Fixed typos.

@@
 written as $\Eflat\E$, is the environment
-where all the binders are replace by normal ones. More formally:
+where all the binders are replaced by normal ones. More formally:
 \begin{align*}

### 2019-11-12 kindly/lang.tex
Fixed uses of 'note' instead of 'denote' and various typos.
Added syntactic category for multiplicities, simplified lattice rules, fixed syntax-directed kinding rules to match syntax (applied type constructors instead of type constructors).

@@
 kind,
-a type application $\tapp{t}{\Multi{\tau}}$ of an abstract type constructor
+a type application $\tapp{\tcon}{\Multi{\tau}}$ of an abstract type constructor
 $\T t$, or a borrowed type  $\borrowty{k}{\tau}$.
@@
 % Let us first define our conventions and meta-syntactic variables.
-% Expressions, types and kinds are noted respectively $e$, $\tau$, and $k$.
-% Variables, type variables and kind variables are noted
-% respectively $x$, $\tvar$ and $\kvar$. Types and kind schemes
-% are noted $\schm$ and $\kschm$.
+% Expressions, types and kinds are denoted respectively as $e$, $\tau$, and $k$.
+% Variables, type variables and kind variables are denoted
+% respectively as $x$, $\tvar$ and $\kvar$. Types and kind schemes
+% are denoted by $\schm$ and $\kschm$.
 % In the rest of the text, we will write lists with a bar and an index, for
 % instance $\Multi[i]{\tau}$.
-% Kind constants are noted in bold capital letters, such as $\kaff$ for affine.
+% Kind constants are denoted in bold capital letters, such as $\kaff$ for affine.
 
@@
 % traditional ML languages with let bindings.
-% Borrows, noted $\borrow{e}$, are annotated with a specification indicating
+% Borrows $\borrow{e}$ are annotated with a specification indicating
 % if a borrow is immutable ($\IBORROW$) or mutable ($\MBORROW$).
-% We also distinguish reborrows, which are noted $\reborrow{e}$.
+% We also distinguish reborrows, which are written as $\reborrow{e}$.
 % Match are indexed with a specification which can be either $\&^b$
@@
 % should operate on borrows or not.
-% Regions, noted $\region{\Sone x\BORROW}{e}$ are annotated with their nesting ($n$)
+% Regions $\region{\Sone x\BORROW}{e}$ are annotated with their nesting ($n$)
 % and the variable whose borrow they enclose ($x$).
 % %
-% $\tapp{t}{\Multi{\tau}}$ denotes a type application where
+% $\tapp{\tcon}{\Multi{\tau}}$ denotes a type application where
 % $\T t$ is an abstract type constructor.
-% Arrow and borrow types are annotated with a kind, noted $\tau \tarr{k} \tau$
+% Arrow and borrow types are annotated with a kind, written as $\tau \tarr{k} \tau$
 % and $\borrowty{k}{\tau}$ respectively.
@@
 % in our typing environments.
-% The grammar of type environment (noted $\E$) is shown in \cref{grammar:env}.
+% The grammar of type environment (denoted $\E$) is shown in \cref{grammar:env}.
 

### 2019-11-12 kindly/sdtyping/extra.tex
Add kinding rules for pairs.
Added syntactic category for multiplicities, simplified lattice rules, fixed syntax-directed kinding rules to match syntax (applied type constructors instead of type constructors).

@@
 Kinds and types follow a small calculus with variables ($\tvar$,\dots),
-functions (type constructors $\T{t}$), application ($\tapp{t}{\Multi{\tau}}$)
+functions (type constructors $\T{t}$), application ($\tapp{\tcon}{\Multi{\tau}}$)
 and primitives such as types for arrows ($\tau\tarr{k}\tau'$) and

### 2019-11-12 kindly/infer/infer.tex
Added syntactic category for multiplicities, simplified lattice rules, fixed syntax-directed kinding rules to match syntax (applied type constructors instead of type constructors).
Fix the remaining typos/small comments from the reviews.

@@
 the transformations.
-In the context of inference, we define new judgements which \emph{infers}
+In the context of inference, we define new judgements which \emph{infer}
 the constraints.
@@
   Again, the relation is total and non-ambiguous for any given $\Sv'$ and $x$,
-  and uses the same rules as the one presented in \cref{sdtyping:regions}.
+  and uses the rules presented in \cref{sdtyping:regions}.
 \end{itemize}
@@
 the body of the lambda expression. We also introduce the constraint
-$\Weaken_{\bvar{x}{\sigma}}(\Sv)$. If $x$ was used $e$, $x$ is present
-in $\Sv$ and this constraint is empty. Otherwise, if $x$ is
-\emph{not} present in $\Sv$, its type should not have a linear
-kind, which is enforced by the constraints $\Cleq{\sigma}{\kaff_\infty}$.
+$\Weaken_{\bvar{x}{\sigma}}(\Sv)$, which is false only if weakening
+is necessary but impossible due to linearity.
 The $\Weaken$ constraint is introduced at each binding point in the program
@@
     {\generalize{C}{\E}{\tau} =
-      (\Cproj{\Multi{\kvar},\Multi{\tvar}}{C},
-      \forall \Multi{\kvar},\Multi{\tvar}.\qual{C}{\tau})\\
+      (\Cproj{(\Multi{\kvar},\Multi{\tvar})}{C},
+      \forall (\Multi{\kvar},\Multi{\tvar}).\qual{C}{\tau})\\
       \text{ where}\quad
@@
 The generated constraint $C_\schm$ uses a new projection operator,
-$\Cproj{\tvar}{D}$, which
+$\Cproj{x}{D}$ where $x$ can be either a type or a kind variable, which
 allow the creation of local variables inside the constraints.
@@
 We define the set of solved formed
-$\mathcal S$ as the set of constraints $\CL$ quantified by $\equivC$.
+$\mathcal S$ as the quotient set of $\CL$ by $\equivC$.
 We will show later that such constraints are in fact only composed of
@@
   gather all existential quantifications at the front of the constraint.
-  We obtain a constraint $C^k = \exists \kvar_i,\ \Cleq{k_j}{k'_j}_j$ and
+  We obtain a constraint $C^k = \exists \Multi\kvar,\ \Cleq{k_j}{k'_j}_j$ and
   a substitution $\unif_\tau$.
@@
 \item Take the transitive closure of $\mathcal R$.
-\item Remove all the vertices corresponding to the kind variables $\kvar_i$
+\item Remove all the vertices corresponding to the kind variables $\Multi\kvar$
   that are existentially quantified in $C^k$.
@@
 \begin{definition}[Instance relation]
-  Given a constraints $C$ and two schemes
+  Given a constraint $C$ and two schemes
   $\schm = \forall \Multi\tvar. \qual{D}{\tau}$ and

### 2019-11-12 kindly/lang.tex
Fix the remaining typos/small comments from the reviews.

@@
   an algorithm to infer them in \cref{regionannot}.)
-\item Regions are identified using de Bruijn-indexing instead of region
+\item Regions are identified using nesting levels instead of region
   variables.
@@
 kind,
-a type application $\tapp{\tcon}{\Multi{\tau}}$ of an abstract type constructor
+a type application $\tapp{t}{\Multi{\tau}}$ of an abstract type constructor
 $\T t$, or a borrowed type  $\borrowty{k}{\tau}$.
@@
 % %
-% $\tapp{\tcon}{\Multi{\tau}}$ denotes a type application where
+% $\tapp{t}{\Multi{\tau}}$ denotes a type application where
 % $\T t$ is an abstract type constructor.
@@
 
-A storable $W$ is a particular object that can be stored. There are five
+A storable $w$ is a particular object that can be stored. There are five
 kinds of storable. A poly closure represents a polymorphic

### 2019-11-12 kindly/motivation.tex
Fix the remaining typos/small comments from the reviews.

@@
 a sequence of interactions. For instance, the type
-\lstinline{!int.!int.?int.end} specifies a protocol
+\lstinline{int!int!int?end} specifies a protocol
 where the program must send two integers, receive an integer, and then
@@
 channel. That is, after sending two integers on the channel,
-the remaining channel has type \lstinline{?int.end}.
+the remaining channel has type \lstinline{int?end}.
 

### 2019-11-12 kindly/soa.tex
Fix the remaining typos/small comments from the reviews.

@@
 
-Quill~\citep{DBLP:conf/icfp/Morris16} is an Haskell-like language with affine
+Quill~\citep{DBLP:conf/icfp/Morris16} is a Haskell-like language with affine
 types through the use of qualified types.
@@
 \lstinline/Array.create : int -> 'a -> \E 'b. ('a, 'b) array/ where
-\lstinline/'b/ uniquely identify of the array.
+\lstinline/'b/ uniquely identify the array.
 Due to the reliance on existentials, Alms does not support complete type inference.
@@
   resource handling in direct style, where types themselves are
-  described as affine or linear. (Or course, continuation-passing
+  described as affine or linear. (Of course, continuation-passing
   style is also supported.)
@@
 particular, we use HM(X)~\citep{DBLP:journals/tapos/OderskySW99} which
-allows to use constrained types in an ML-like language with complete
+allows us to use constrained types in an ML-like language with complete
 type inference.
@@
 Such discipline is known to be friendly to inference and has been used in many
-contexts, including OCaml itself, and has been combined
+contexts, including OCaml, and has been combined
 with constraints~\citep{DBLP:journals/tapos/OderskySW99,DBLP:conf/sas/TrifonovS96}.
@@
 and sufficiently simple to make
-all simplification rules complete, which allows to keep type signatures simple.
+all simplification rules complete, which allows us to keep type signatures simple.
 

### 2019-11-12 kindly/intro.tex
Adapt to the PLDI format.

@@
     \lstinputlisting{code/writefiles.affe}
+    \vspace{-15pt}
     \caption{File API}
@@
   ~
-  \begin{subfigure}[t]{0.45\linewidth}
+  \begin{subfigure}[t]{0.5\linewidth}
 \begin{lstlisting}
@@
 \end{lstlisting}
+    \vspace{-10pt}
     \caption{File example}
@@
   \end{subfigure}
+  \vspace{-10pt}
   \caption{Writing files}
@@
 \begin{lstlisting}
-  let w = File.write &!h in w "Hello "; w "world!" (* type error *)
+let w = File.write &!h in w "Hello "; w "world!" (*error*)
 \end{lstlisting}
@@
   constraint simplification and extended with pattern matching. A
-  webfront for the algorithm is available, its URL is given in the
+  web implementation of the typechecker is available, its URL is given in the
   non-anonymous supplementary material, and we plan to submit it for

### 2019-11-12 kindly/intro.tex
Swap "mutable/immutable borrow" for "exclusive/shared borrow".

@@
 propose an extended type system for ML-like languages that comes with
-linear and affine types, mutable and immutable borrows, but all that
+linear and affine types, exclusive and shared borrows, but all that
 integrated with full principal type inference, garbage collected data, and
@@
   we also permit affine values.
-\item Our system supports mutable and immutable borrows, hence the
+\item Our system supports exclusive and shared borrows, hence the
   need for affinity.
@@
 as shown in \cref{fig:writing-files-example}. Failing to do so is a
-static type error.  To write to the file, we must take a (mutable)
+static type error.  To write to the file, we must take a (exclusive)
 borrow \lstinline/&!h/ of the handle and pass it to the
-\lstinline/write/ function. Mutable borrows are classified as affine:
+\lstinline/write/ function. Exclusive borrows are classified as affine:
 they must not be duplicated, but there is no requirement to use

### 2019-11-12 kindly/lang.tex
Swap "mutable/immutable borrow" for "exclusive/shared borrow".

--- 2019-11-12 kindly/lang.tex
@@
 A borrow $\borrow{x}$ is always taken from a variable $x$. The
-\emph{borrow annotation} $\BORROW$ indicates whether the borrow is mutable
-($\MBORROW$) or immutable ($\IBORROW$). If $x$ is already borrowed,
+\emph{borrow annotation} $\BORROW$ indicates whether the borrow is exclusive
+($\MBORROW$) or shared ($\IBORROW$). If $x$ is already borrowed,
 then we need to use the reborrow expression $\reborrow{x}$.
@@
 works with a full resource while the {\sc SuspB}
-rule splits off an immutable suspended binding from a borrow.
+rule splits off an shared suspended binding from a borrow.
 %

### 2019-11-12 kindly/motivation.tex
Swap "mutable/immutable borrow" for "exclusive/shared borrow".

@@
 During the lifetime of the array resource \lstinline/a/, we can split
-off \emph{immutable borrows} \lstinline/&a/ that provide a read-only
-view or \emph{mutable borrows} \lstinline/&!a/ for read-write views.
-Observer functions such as \lstinline/length/ and \lstinline/get/ expect an immutable borrow
+off \emph{shared borrows} \lstinline/&a/ that provide a read-only
+view or \emph{exclusive borrows} \lstinline/&!a/ for read-write views.
+Observer functions such as \lstinline/length/ and \lstinline/get/ expect an shared borrow
 argument, mutator functions such a \lstinline/set/ expect
-a mutable borrow.
+a exclusive borrow.
 
@@
 contained in the lifetime of the resource.
-In a region, we can split off as many immutable borrows as we like
-from a resource, but we can take only one mutable borrow. In a
-subsidiary region, we can take further immutable borrows from any
-borrow or we can take a mutable borrow from a mutable borrow from a
+In a region, we can split off as many shared borrows as we like
+from a resource, but we can take only one exclusive borrow. In a
+subsidiary region, we can take further shared borrows from any
+borrow or we can take a exclusive borrow from a exclusive borrow from a
 enclosing region. Borrows are confined to their regions. Inside the region,
-immutable borrows are unrestricted (\lstinline/un/) whereas mutable
+shared borrows are unrestricted (\lstinline/un/) whereas exclusive
 borrows are affine (\lstinline/aff/).
@@
 resource inside the borrow's region. In \cref{ex:array:get} of the
-example, there are two immutable borrows in the same
+example, there are two shared borrows in the same
 expression, which forms a region by itself. These borrows are split
-off the mutable borrow used in \cref{ex:array:set}, which belongs to
+off the exclusive borrow used in \cref{ex:array:set}, which belongs to
 the next enclosing region.
@@
 The in-place mutation function \lstinline/set_mut/ with type
-\lstinline/&!('a t) * int * 'a -> unit/ works on a mutable borrow  whereas the persistent
+\lstinline/&!('a t) * int * 'a -> unit/ works on a exclusive borrow  whereas the persistent
 \lstinline/set/ operation has type \lstinline/&('a t) * int * 'a -> 'a t/. It
-takes an immutable borrow because it only reads from the
+takes an shared borrow because it only reads from the
 argument array, but returns a fresh,  modified structure.
@@
 The \lstinline/propagate/ function uses direct mutation through a
-mutable borrow of the matrix as it need not preserve the previous
+exclusive borrow of the matrix as it need not preserve the previous
 version of the board.
@@
 In particular,
-since the \lstinline/set/ function only requires an unrestricted immutable borrow,
+since the \lstinline/set/ function only requires an unrestricted shared borrow,
 \lstinline/try_solution/ can captures the borrow \lstinline/&g/
@@
 callback function.
-The callback receives a mutable borrow of an arbitrary resource from the pool.
+The callback receives a exclusive borrow of an arbitrary resource from the pool.
 The typing of the callback ensures
@@
 The signature prescribes the type \lstinline/&!(aff_r1,'a_1 )/ for the
-mutable borrow of the resource with an affine kind at region nesting $r+1$. The return
+exclusive borrow of the resource with an affine kind at region nesting $r+1$. The return
 type of the callback is constrained to kind \lstinline/'k_2 <= aff_r/

### 2019-11-12 kindly/sdtyping/extra.tex
Swap "mutable/immutable borrow" for "exclusive/shared borrow".

@@
 As an example, the rule \textsc{AnnotRegion-MutLeft} is applied
-when there is an immutable borrow and a mutable borrow. In that case, a
-region is created to enclose the immutable borrow, while the mutable
+when there is an shared borrow and a exclusive borrow. In that case, a
+region is created to enclose the shared borrow, while the exclusive
 borrow is left to be closed later. This is coherent with the rules
@@
     environments $\lsplit C\E{\E_l}{\E_r}$;
-    inference $\bsplit C\E{\bm\E_l}{\bm\E_r}$;
-    binders $\bsplit Cb{\bm b_r}{\bm b_l}$}
+    inference $\bsplit C\E{\bm{\E_l}}{\bm{\E_r}}$;
+    binders $\bsplit Cb{\bm{b_r}}{\bm{b_l}}$}
   \label{fig:sd-splitting}
@@
     environments $\lregion[n]{C}{x}{\E}{\E'}$;
-    inference $\bregion[\bm n]{C}{\bm x}{\E}{\bm\E'}$;
-    binders $\bregion[\bm n]{C}{\bm x}{b}{\bm b'}$}
+    inference $\bregion[\bm n]{C}{\bm x}{\E}{\bm{\E'}}$;
+    binders $\bregion[\bm n]{C}{\bm x}{b}{\bm{b'}}$}
   \label{fig:sd-borrowing}
@@
   \input{sdtyping/sd-typing}
-  \caption{Syntax-directed typing rules}
+  \caption{Syntax-directed typing rules --
+    $\inferS{C}{\E}{e}{\tau}$}
   \label{fig:syntax-directed-typing}

### 2019-11-13 kindly/intro.tex
tightening up

@@
 The most closely related contenders in this design space are Linear Haskell
-\cite{DBLP:journals/pacmpl/BernardyBNJS18}, henceforth called LH,
+\cite{DBLP:journals/pacmpl/BernardyBNJS18}, henceforth LH,
 Quill \cite{DBLP:conf/icfp/Morris16}, and ALMS \cite{DBLP:conf/popl/TovP11}.
@@
 similar as both systems permit abstraction over the number of uses of
-values and retain type inference, but the details are quite different.
+values and retain type inference, but the details are different.
 \begin{enumerate}
 \item LH's multiplicities are either linear or unrestricted whereas
-  we also permit affine values.
-\item Our system supports exclusive and shared borrows, hence the
-  need for affinity.
-\item Our multiplicities are directly attached to the type of the
-  value. For example, in our system the function type \lstinline/'a-{lin}>'b/
+  we also distinguish affine values.
+\item \lang{} supports two flavors of borrowing: affine (exclusive,
+  mutable) and unrestricted (shared, immutable) borrows.
+\item Multiplicities are directly attached to the type of the
+  value. For example, in \lang{} the function type \lstinline/'a-{lin}>'b/
   denotes the type of a \emph{single-use function} that can be called
@@
     once}.
-\item Similar to Quill, we make use of multiplicity contraints. We
-  rely on kind subsumption to simplify the inferred types.
+\item Similar to Quill, we make use of multiplicity contraints. Kind
+  subsumption in \lang{} helps simplify inferred types.
 \end{enumerate}
@@
 \lstinline/fopen "foo"/ returns a linear handle to a newly created
-file, which \emph{must} be released later on through \lstinline/close/
+file, which \emph{must} be released later on with \lstinline/close/
 as shown in \cref{fig:writing-files-example}. Failing to do so is a
-static type error.  To write to the file, we must take a (exclusive)
+static type error.  To write to the file, we must take an exclusive
 borrow \lstinline/&!h/ of the handle and pass it to the
-\lstinline/write/ function. Exclusive borrows are classified as affine:
+\lstinline/write/ function. Exclusive borrows are affine:
 they must not be duplicated, but there is no requirement to use
@@
 \lstinline/write &!h/ captures the affine borrow and hence the
-resulting function must also be handled in an affine way. That is, it
-would be an error to write
+resulting function is also affine. It
+would be an error to use the affine closure twice as in
 \begin{lstlisting}
@@
 \end{lstlisting}
-as we would be using the affine closure twice. The remaining arrows in
+The remaining arrows in
 the API are unrestricted and we write \lstinline/->/ instead of the
@@
 actual multiplicity gets subsumed to $\kappa_1$. This constraint
-simplification technique is built into our type inference algorithm. Again, the printout
+simplification is part of our type inference algorithm. Again, the printout
 of the type scheme only mentions the required constraint

### 2019-11-13 kindly/motivation-linearity.tex
moved session type example

--- 2019-11-12 kindly/motivation.tex
@@
-\section{Linearity, Affinity, and Borrows at Work}
-\label{motivation}
-
 \subsection{A session on linearity}
@@
 
-\subsection{Imperative programming}
-\label{sec:imper-progr}
-
-The functional resource-passing style demonstrated for channels in
-\cref{sec:session-linearity} becomes tedious when manipulating mutable
-resources like buffers or connection pools.
-%
-To make linearity easier to use with an imperative programming style, we rely on
-the notion of borrowing \cite{DBLP:conf/popl/BoylandR05}.
-
-As a first example, we consider an API for mutable linear arrays
-(\cref{sig:array}).
-The intention of this API is to safely handle manual allocation and
-deallocation of arrays that may contain affine elements.
-One would first use \lstinline/create n v/ to create
-an (unrestricted) array of size \lstinline/n/ with some value
-\lstinline/v/. This value must be unrestricted as it is duplicated to
-initialize all array elements. Using the \lstinline/map/ function we
-can transform the unrestricted elements into linear (affine)
-ones. This array can be processed further using \lstinline/map/ (or \lstinline/set/ if
-affine). The \lstinline/get/ function only applies if the element type
-is unrestricted as one element is duplicated. The \lstinline/length/
-function applies anytime. Before \lstinline/free/ing the array the elements must be
-affine, again.
-
-To manage the different kinds of accessing the array we distinguish between constructors, destructors,
-observers, and mutators.
-Constructors and destructors like \lstinline/create/ and \lstinline/free/ manipulate the whole
-array. As with channels in \cref{sec:session-linearity},
-\lstinline/create/ yields a linear resource which is consumed by \lstinline/free/.
-During the lifetime of the array resource \lstinline/a/, we can split
-off \emph{shared borrows} \lstinline/&a/ that provide a read-only
-view or \emph{exclusive borrows} \lstinline/&!a/ for read-write views.
-Observer functions such as \lstinline/length/ and \lstinline/get/ expect an shared borrow
-argument, mutator functions such a \lstinline/set/ expect
-a exclusive borrow.
-
-Each borrow is tied to a region whose lifetime is properly
-contained in the lifetime of the resource.
-In a region, we can split off as many shared borrows as we like
-from a resource, but we can take only one exclusive borrow. In a
-subsidiary region, we can take further shared borrows from any
-borrow or we can take a exclusive borrow from a exclusive borrow from a
-enclosing region. Borrows are confined to their regions. Inside the region,
-shared borrows are unrestricted (\lstinline/un/) whereas exclusive
-borrows are affine (\lstinline/aff/).
-
-The code in \cref{ex:array} demonstrates a use of the API to create an
-array of Fibonacci numbers. After creation of the array, we can see
-that the presence of a borrow prevents access to the ``raw''
-resource inside the borrow's region. In \cref{ex:array:get} of the
-example, there are two shared borrows in the same
-expression, which forms a region by itself. These borrows are split
-off the exclusive borrow used in \cref{ex:array:set}, which belongs to
-the next enclosing region.
-Finally, the code returns the whole array in \cref{ex:array:return},
-which is possible because none of the borrows are in scope anymore.
-
-\begin{figure}[tp]
-  \centering
-  \begin{subfigure}[t]{1\linewidth}
-\begin{lstlisting}
-module Array : sig
-  type ('a : 'k) t : lin
-  val create : ('a : un) => int -> 'a -> 'a t
-  val free : ('a : aff) => 'a t -> unit
-  val length : &('a t) -> int
-  val get : ('a : un) => &('a t) * int -> 'a
-  val set : ('a : aff) => &!('a t) * int * 'a -> unit
-  val map : (&'a -> 'b) -> &('a t) -> 'b t
-end
-\end{lstlisting}
-    \vspace{-15pt}
-    \caption{Affine arrays}
-    \label{sig:array}
-  \end{subfigure}
-
-  \begin{subfigure}[t]{1\linewidth}
-\begin{lstlisting}
-let mk_fib_array n =
-  let a = create n 1 in
-  for i = 2 to n - 1 do
-    let x = get (&a, i-1) + get (&a, i-2) in(*@\label{ex:array:get}*)
-    set (&!a, i, x)(*@\label{ex:array:set}*)
-  done;
-  a(*@\label{ex:array:return}*)
-# mk_fib_array : int -> int Array.t
-\end{lstlisting}
-    \vspace{-10pt}
-    \caption{Example of use of linear arrays}
-    \label{ex:array}
-\begin{lstlisting}[firstnumber=2]
-  let a = create n 1 in
-  for i = 2 to n - 1 do {|
-    let x = {| get &a (i-1) + get &a (i-2) |} in(*@\label{ex:array:region:get}*)
-    set &!a i x
-  |} done;
-\end{lstlisting}
-    \vspace{-10pt}
-    \caption{Excerpt of \cref{ex:array} with explicit regions}
-    \label{ex:array:region}
-  \end{subfigure}
-\end{figure}
-
-\cref{ex:array:region} uses braces as in \lstinline/{| ... |}/ to
-explicitly state the regions of the example in \cref{ex:array}.  One
-region consists of the header expression of the \lstinline/let/ in
-\cref{ex:array:region:get}. It is contained in another region that
-comprises the body of the \lstinline/for/ loop. \lang guarantees that borrows
-never escape the smallest enclosing region. It employs a system of
-\emph{indexed kinds} like \lstinline{aff_r} and \lstinline{un_r} where
-$r$ is a positive integer that corresponds to the lexical nesting
-depth of regions. For instance, the type of \lstinline{&!a} in
-\cref{ex:array:set} has kind \lstinline{aff_1} whereas the type of
-\lstinline{&a} in \cref{ex:array:region:get} has kind
-\lstinline{un_2} and the typing of its region is such that types with
-kind indexes greater than or equal to $2$ cannot escape.
-
-In \cref{ex:array:region:get} of the example, there is no danger of \lstinline/&a/ escaping because it
-is consumed in the calls to \lstinline/get/.
-Region annotations are automatically inserted using
-syntactic cues such as borrows and binders (see
-\cref{regionannot}).  Users may manually insert regions to further restrict the range of a
-borrow. In practice, region annotations are
-rarely needed with common programming idioms.
-% All regions for the examples in this section have been inferred.
-
-% \begin{figure}
-%   \centering
-%   % \begin{subfigure}{0.6\linewidth}
-% \begin{lstlisting}
-% let mk_fib_array n =
-%   let a = create n 1 in
-%   for i = 2 to n - 1 do {|
-%     let x = {| get &a (i-1) + get &a (i-2) |} in(*@\label{ex:array:region:get}*)
-%     set &!a i x
-%   |}
-%   done;
-%   a(*@\label{ex:array:return}*)
-% # mk_fib_array : int -> int Array.t
-% \end{lstlisting}
-%     \caption{Example from \cref{ex:array} with explicit regions}
-%     \label{ex:array:region}
-%   % \end{subfigure}
-% \end{figure}
-
-\subsection{Hybrid programming style}
-
-Recently introduced persistent data structures
-permit transient mutations where
-non-linear uses lead to degraded performance
-\cite{DBLP:conf/ml/ConchonF07} or to
-dynamic and static checks \cite{DBLP:journals/pacmpl/Puente17}.
-In particular, persistent Hash-Array-Mapped-Tries (HAMT) have been used with similar
-APIs in several non-pure functional languages (OCaml, Clojure, \dots).
-Affine types help formalize the performance contract between the programmer
-and the library, while borrows avoid the need to thread state explicitly,
-as usually required by an API for immutable data types.
-%
-
-In this section, we present
-a safe API for persistent arrays that support both immutable and mutable features,
-and use it to implement a backtracking Sudoku solver.
-% show how \lang allows us to define
-The Sudoku solver maintains an array to represent the state of
-the game and uses backtracking when there are several choices to proceed.
-The use of backtracking suggests a persistent data structure for the array.
-However, only changes that correspond to a choice point need to use
-the persistence mechanism, others may be implemented as
-cheap in-place mutations.
-
-\cref{sig:hybarray} contains an API \lstinline/HybArray/
-that enables using mutable and immutable modifications to
-the board through affine types and borrows.
-The signature differs slightly from the \lstinline/Array/
-signature. For simplicity, the array elements are unrestricted, but
-the structure itself remains linear so as to be implemented in terms of \lstinline/Array/.
-The in-place mutation function \lstinline/set_mut/ with type
-\lstinline/&!('a t) * int * 'a -> unit/ works on a exclusive borrow  whereas the persistent
-\lstinline/set/ operation has type \lstinline/&('a t) * int * 'a -> 'a t/. It
-takes an shared borrow because it only reads from the
-argument array, but returns a fresh,  modified structure.
-The code in \cref{ex:cow} contains a very simple implementation of
-\lstinline/HybArray/ that represents hybrid arrays
-as regular arrays and uses copy-on-write for persistent
-modifications. The function
-\lstinline/mapi : (int -> &'a -> 'b) -> &('a t) -> 'b t /
-is a simple variation on \lstinline/Array.map/ where the mapping function
-takes the position of the element.
-% A more elaborate implementation could use persistent arrays
-% \cite{DBLP:conf/ml/ConchonF07}.
-
-\begin{figure}[tp]
-  \centering
-  \begin{subfigure}{1\linewidth}
-\begin{lstlisting}
-module type HybArray = sig
-  type ('a : un) t : lin
-  val create : int -> 'a -> 'a t
-  val free : 'a t -> unit
-  val length : &('a t) -> int
-  val get : &('a t) * int -> 'a
-  val set : &('a t) * int * 'a -> 'a t
-  val set_mut : &!('a t) * int * 'a -> unit
-end
-\end{lstlisting}
-    \vspace{-15pt}
-    \caption{Signature}
-    \label{sig:hybarray}
-  \end{subfigure}
-
-  \begin{subfigure}{1\linewidth}
-\begin{lstlisting}
-module Cow : HybArray = struct
-  include Array
-  let set (a, i0, x0) =
-    Array.mapi (fun i x -> if i = i0 then x0 else x) a
-  let set_mut = Array.set
-end
-\end{lstlisting}
-    \vspace{-15pt}
-    \caption{Implementation using copy-on-write}
-    \label{ex:cow}
-  \end{subfigure}
-  \caption{Hybrid arrays}
-\end{figure}
-
-Our implementation of a Sudoku solver (\cref{ex:sudoku}) performs
-modifications that correspond to choice points using \lstinline/set/,
-which makes it trivial to come back to the previous version
-of the array, while other modifications use \lstinline/set_mut/, which
-forbids uses of the previous version.
-
-The board is represented as a 2D-matrix (\cref{line:boardtype}), where
-the \lstinline/Matrix/ type uses the same API as \lstinline/HybArray/
-but with two indices.
-Each cell contains an integer set that represent admissible solutions so far.
-The Sudoku solver iterates over the cells and tries each possible solution (\cref{line:try_solution}).
-When a value is picked for the current cell, we create a choice point,
-change the cell with an immutable modification (\cref{line:immutset}), and propagate
-the changes with the \lstinline/propagate/ function.
-The \lstinline/propagate/ function uses direct mutation through a
-exclusive borrow of the matrix as it need not preserve the previous
-version of the board.
-The implementation of \lstinline/propagate/ is split into three parts
-for lines, columns, and square, which are all very similar to function
-\lstinline/propagate_lines/ (\cref{line:propline}).
-
-The typing ensures that the mutations do not compromise the state at
-the choice point, because they operate on a new state \lstinline/new_g/ created for one
-particular branch of the choice.
-In particular,
-since the \lstinline/set/ function only requires an unrestricted shared borrow,
-\lstinline/try_solution/ can captures the borrow \lstinline/&g/
-while staying unrestricted.
-This also make \lstinline/try_solution/ itself bound to the same
-region as \lstinline/&g/. In this example, our region inference
-will place the region exactly at the \lstinline/begin//\lstinline/end/
-scope.
-This allows us to use \lstinline/try_solution/ in
-the iteration loop on \cref{line:iter}.
-Since \lstinline/&g/ is unrestricted, we could even use a parallel iteration
-instead of a sequential one.
-As \lstinline/g/ is linear we must free it before returning,
-in \cref{line:free:g}, outside of the region.
-
-While presented for copy-on-write arrays, the API
-can easily be adapted to other persistent data structures with
-transient mutability such as Relaxed-Radix Balance Vectors (RRB) \cite{DBLP:journals/pacmpl/Puente17}
-or persistent HAMTs \cite{bagwell2001ideal,clojurehamt} to provide  a
-convenient programming style without compromising performance.
-
-\begin{figure}
-  \centering
-  \lstinputlisting[linerange=1-31]{code/sukodu.affe}
-    \vspace{-15pt}
-  \caption{Excerpt of the Sudoku solver}
-  \label{ex:sudoku}
-\end{figure}
-
-% \subsection{Type abstraction and linearity violations}
-
-% While linearity allows to verify many properties, it also prevents
-% the implementation of functions that are nevertheless safe. Rust addresses this
-% issue through the \lstinline/unsafe/ construct, which allows to locally
-% ignore linearity constraints. This is essential to build basic
-% functions such as splitting strings.
-% %
-% \lang allows to partially emulate this behavior using type abstraction
-% as provided by ML modules.
-% In particular, data types can be internally defined as unrestricted but exposed
-% as affine.
-
-% \TODO{Show example with arrays}
-
-\subsection{Local data-structures with regions}
-\label{tuto:pool}
-
-% We have seen that borrows are limited to a region.
-In the examples so far, regions do not appear in type signatures.
-But for certain programming idioms, we want to extend the scope of a
-region across a function boundary. As an example,
-we present an interface and implementation of a pool of linear resources where the
-extended scope of the region enforces proper use of the resources.
-
-\cref{intf:pool} contains the interface of the \lstinline/Pool/ module.
-A pool is parameterized by its content. The kind of the pool
-depends on the content: linear content implies
-a linear pool while unrestricted content yields an unrestricted pool.
-The functions \lstinline/Pool.create/ and \lstinline/Pool.consume/
-build/destroy a pool given creators/destructors for the elements
-of the pool.
-The function \lstinline/Pool.use/ is the workhorse of the API, which
-borrows a resource from the pool to a callback.
-It takes a shared borrow of a pool (to enable concurrent access) and a
-callback function.
-The callback receives a exclusive borrow of an arbitrary resource from the pool.
-The typing of the callback ensures
-that this borrow is neither captured nor returned by the function.
-
-This encapsulation is implemented with a universally quantified \emph{kind index variable} $r$.
-The signature prescribes the type \lstinline/&!(aff_r1,'a_1 )/ for the
-exclusive borrow of the resource with an affine kind at region nesting $r+1$. The return
-type of the callback is constrained to kind \lstinline/'k_2 <= aff_r/
-so that the callback certainly cannot return the borrowed argument.
-In a specific use of \lstinline/Pool.use/, the index $r$ gets unified
-with the current nesting level of regions so that the region for the
-callback effectively gets ``inserted'' into the lexical nesting at the callsite.
-
-\cref{ex:pool} shows a simple example of using the \lstinline/Pool/ module.
-
-The implementation in \cref{impl:pool} represents a bag of resources
-using a concurrent queue with atomic add and remove operations.
-The implementation of the \lstinline/Pool.create/ and \lstinline/Pool.consume/
-functions is straightforward.
-The function \lstinline/Pool.use/ first draws
-an element from the pool (or creates a fresh element),
-passes it down to the callback function \lstinline/f/, and returns
-it to the pool afterwards.
-For clarity,
-we explicitly delimit the region in \cref{line:pool:region} to ensure that
-the return value of \lstinline/f &!o/ does not capture \lstinline/&!o/.
-In practice, the type checker inserts this region automatically.
-
-\begin{figure}[tp]
-  \centering
-  \begin{subfigure}[t]{1\linewidth}
-\begin{lstlisting}
-type ('a:'k) pool : 'k
-
-create : (unit -> 'a) -> 'a pool
-consume : ('a -> unit) -> 'a pool -> unit
-use : ('a_1:'k_1),('a_2:'k_2),('k_2 <= aff_r) =>
-  &('a_1 pool) -> (&!(aff_r1,'a_1) -{lin}> 'a_2) -{'k_1}> 'a_2
-\end{lstlisting}
-    \vspace{-10pt}
-    \caption{Interface}
-    \label{intf:pool}
-
-\begin{lstlisting}
-(* Using the pool in queries. *)
-let create_user pool name =
-  Pool.use &pool (fun connection ->
-    Db.insert "users" [("name", name)] connection)
-
-let uri = "postgresql://localhost:5432"
-let main users =
-  (* Create a database connection pool. *)
-  let pool = Pool.create (fun _ -> Db.connect uri) in
-  List.parallel_iter (create_user &pool) users;
-  Pool.consume (fun c -> Db.close c)
-\end{lstlisting}
-    \vspace{-10pt}
-    \caption{Example of use}
-    \label{ex:pool}
-  \end{subfigure}
-
-  \begin{subfigure}[t]{1\linewidth}
-\begin{lstlisting}
-type ('a:'k) pool : 'k = {
-  spawn : unit -> 'a ;
-  queue : 'a CQueue.t;
-}
-
-let create spawn =
-  { spawn ; queue = CQueue.create () }
-let consume f c =
-  CQueue.iter f c.queue
-
-let use { spawn ; queue } f =
-  let o = match CQueue.pop &queue with
-    | Some x -> x
-    | None () -> spawn ()
-  in
-  let r = {| f &!o |} in(*@\label{line:pool:region}*)
-  Queue.push o &queue;
-  r
-\end{lstlisting}
-    \vspace{-15pt}
-    \caption{Implementation}
-    \label{impl:pool}
-  \end{subfigure}
-
-  \caption{The \texttt{Pool} module}
-  \label{fig:pool}
-\end{figure}
-
 %%% Local Variables:

### 2019-11-13 kindly/motivation.tex
moved session type example

@@
 \label{motivation}
-
-\subsection{A session on linearity}
-\label{sec:session-linearity}
-
-Session typing \cite{Honda1993,DBLP:conf/esop/HondaVK98} is a type
-discipline for checking protocols statically. A session type is
-ascribed to a communication channel and describes
-a sequence of interactions. For instance, the type
-\lstinline{int!int!int?end} specifies a protocol
-where the program must send two integers, receive an integer, and then
-close the channel.
-%
-In this context, channels must be used linearly, because every use of
-a channel ``consumes'' one interaction and  changes the type of the
-channel. That is, after sending two integers on the channel,
-the remaining channel has type \lstinline{int?end}.
-
-Here are some standard type operators for session types \SType:
-\begin{center}
-  $\SType ::=$
-  \begin{tabular}[t]{rl}
-    $\tau ! \SType$ & Send a value of type $\tau$ then continue with protocol $\SType$.\\
-    $\tau ? \SType$& Receive a value of type $\tau$ then continue with protocol $\SType$.\\
-    $\SType \oplus \SType'$& Internal choice between protocols $\SType$ and $\SType'$.\\
-    $\SType \operatorname{\&} \SType'$
-                    & Offer a choice between protocols $\SType$ and $\SType'$.
-  \end{tabular}
-\end{center}
-
-\citet{DBLP:journals/jfp/Padovani17}  has shown how to encode this
-style of session typing in ML-like languages, but his implementation
-downgrades linearity to a run-time check for affinity. Building on that
-encoding we can provide a safe API in \lang that statically enforces
-linear handling of channels:
-%
-\begin{lstlisting}
-type 'S st : lin (*@\label{line:session1}*)
-val receive: ('a ? 'S) st -> 'a * 'S st (*@\label{line:session2}*)
-val send : ('a ! 'S) st -> 'a -{lin}> 'S st
-val create : unit -> 'S st * (dual 'S) st
-val close : end st -> unit
-\end{lstlisting}
-% val send : ('a:'k). 'a -> ('a ! 'S) st -{'k}> 'S st
-
-Line~\ref{line:session1} introduces a parameterized abstract
-type \lstinline{st} which is linear as indicated
-by its kind \lstinline{lin}. Its low-level implementation would wrap a
-handle for a socket, for example. The \lstinline{receive}  operation
-in Line~\ref{line:session2} takes a channel that is ready to receive a
-value of type \lstinline{'a} and returns a pair of the value a the
-channel at its residual type \lstinline{'S}. It does not matter
-whether \lstinline{'a} is restricted to be linear, in fact
-\lstinline{receive} is polymorphic in the kind of \lstinline{'a}.
-This kind polymorphism is the
-default if no constraints are specified.
-%
-The \lstinline{send} operation takes a linear channel and returns a
-single-use function that takes a value of type \lstinline{'a} suitable
-for sending and returns the channel with updated type.
-%
-The \lstinline{create} operation returns a pair of channel
-endpoints. They follow dual communication protocols, where the
-\lstinline{dual} operator swaps sending and receiving operations.
-%
-Finally, \lstinline{close} closes the channel.
-
-In \cref{fig:sessiontype} we show how to use these primitives to
-implement client and server for an addition service.
-No linearity annotations are needed in the code, as all linearity
-properties can be inferred from the linearity of the \texttt{st} type.
-
-The inferred type of the server, \lstinline{add_service}, is
-\lstinline{(int ! int ! int ? end) st -> unit}.
-The client operates by sending two messages
-and receiving the result.
-This code is polymorphic in both argument and return types, so it could
-be used with any binary operator.
-%
-\begin{figure}[!h]
-  \begin{subfigure}[t]{.5\linewidth}
-\begin{lstlisting}
-let add_service ep =
-  let x, ep = receive ep in
-  let y, ep = receive ep in
-  let ep = send ep (x + y) in
-  close ep
-# add_service :
-  (int ! int ! int ? end) st -> unit
-\end{lstlisting}
-    \vspace{-10pt}
-    \caption{Addition server}
-  \end{subfigure}~
-  \begin{subfigure}[t]{.5\linewidth}
-\begin{lstlisting}
-let op_client ep x y =
-  let ep = send ep x in
-  let ep = send ep y in
-  let result, ep = receive ep in
-  close ep;
-  result
-# op_client :
-  ('a_1 ? 'a_2 ? 'b ! end) st -> 'a_1 -{lin}> 'a_2 -{lin}> 'b
-\end{lstlisting}
-% The actual answer from the typechecker looks like this:
-% # ('m2:^k2), ('m1:^k1).
-% # (^k1 < ^k) & (lin < ^k) =>
-% # ('m1 ?'m2 ? 'result ! end) st -> 'm1 -{lin}> 'm2 -{^k}> 'result
-    \vspace{-10pt}
-    \caption{Binary operators client}
-  \end{subfigure}
-  \vspace{-10pt}
-  \caption{Corresponding session type programs in \lang}
-  \label{fig:sessiontype}
-\end{figure}
-%
-Moreover, the \lstinline/op_client/ function can be partially applied
-to a channel. Since the closure returned by such a partial application
-captures the channel, it can only be used once.  This restriction is
-reflected by the arrow of kind \lstinline{lin}, \lstinline/-{lin}>/,
-which is the type of a single-use function.
-The general form of
-arrow types in \lang is \lstinline/-{k}>/, where \lstinline/kk/ is a
-kind that restricts the number of uses of the function.  For
-convenience, we shorten \lstinline/-{un}>/ to \lstinline/->/.  \lang
-infers the
-single-use property of the arrows without any user
-annotation. In fact, the only difference between the
-code presented here and Padovani's examples
-\cite{DBLP:journals/jfp/Padovani17} is the kind annotation on the
-type definition of \lstinline/st/.
-
-To run client and server, we can create a channel and apply
-\lstinline{add_service} to one end and \lstinline{op_client} to the other.
-Failure to consume either channel endpoints (\lstinline/a/ or \lstinline/b/)
-would result in a type error.
-\begin{lstlisting}
-let main () =
-  let (a, b) = create () in
-  fork add_service a;
-  op_client b 1 2
-# main : unit -> int
-\end{lstlisting}
-
-% \begin{figure}
-%   \centering
-%   \begin{lstlisting}
-% val select : ('S st -{'k}> 'a) -> 'a ot -{'k}> ('S dual) st
-% val branch : 'm it -> 'm
-%   \end{lstlisting}
-%   \caption{Session primitives}
-%   \label{api:session}
-% \end{figure}
 

### 2019-11-13 kindly/motivation.tex
short intro

@@
 
+\lang{} is amenable to a functional resource-passing style as it is
+common in functional encodings of session types (e.g.,
+\cite{DBLP:journals/jfp/Padovani17}; see also
+\cref{sec:session-linearity} in the supplement). But it really shines 
+when manipulating mutable resources like buffers or connection pools
+using an imperative programming style. 
+%
+To support this usage pattern of linearity, we rely on
+the notion of borrowing \cite{DBLP:conf/popl/BoylandR05}.
+
 \subsection{Imperative programming}
 \label{sec:imper-progr}
-
-The functional resource-passing style demonstrated for channels in
-\cref{sec:session-linearity} becomes tedious when manipulating mutable
-resources like buffers or connection pools.
-%
-To make linearity easier to use with an imperative programming style, we rely on
-the notion of borrowing \cite{DBLP:conf/popl/BoylandR05}.
 

### 2019-11-13 kindly/lang.tex
Minor layout changes.

@@
 \label{syntax}
+
+\begin{figure}[!tb]
+  \input{grammar}
+  \caption{Syntax}
+  \label{grammar}
+\end{figure}
 
@@
 
-\begin{figure*}[!tb]
-  \centering
-  \input{grammar}
-  \caption{Syntax}
-  \label{grammar}
-\end{figure*}
-
 \subsection{Typing}
@@
 where affine ones are expected and affine ones are less restricted
-than linear ones as reflected in \TirName{Lat-UA} and \TirName{Lat-AL}.
+than linear ones as reflected in \TirName{Lat-UAL}.
 Second, values usable at level $n$ may be used at any more deeply
-nested level $n'$ (where $n\le\infty$ for any $n$) as defined in the \TirName{Lat-X-Level} axioms.
+nested level $n'$ (where $n\le\infty$ for any $n$) as defined in the \TirName{Lat-Level} axioms.
 
@@
 
-\begin{figure*}[tp]
+\begin{figure}[tp]
   \input{infer/lattice}
@@
   \label{sdtyp:lattice}
-\end{figure*}
+\end{figure}
 
@@
 consists of a kind and two results. A resource is just an indirection
-to another result and the $\blob $ fills a released location.
+to another result and the $\StFreed$ fills a released location.
 

### 2019-11-14 kindly/motivation.tex
adapted motivation to not refer to session type example anymore

@@
 
-As a first example, we consider an API for mutable linear arrays
+A first example is an API for mutable linear arrays
 (\cref{sig:array}).
@@
 One would first use \lstinline/create n v/ to create
-an (unrestricted) array of size \lstinline/n/ with some value
+an (unrestricted) array of size \lstinline/n/ initialized with value
 \lstinline/v/. This value must be unrestricted as it is duplicated to
@@
 ones. This array can be processed further using \lstinline/map/ (or \lstinline/set/ if
-affine). The \lstinline/get/ function only applies if the element type
+affine). The \lstinline/get/ function is only applicable if the element type
 is unrestricted as one element is duplicated. The \lstinline/length/
-function applies anytime. Before \lstinline/free/ing the array the elements must be
-affine, again.
+function is always applicable. To \lstinline/free/ an array the elements must be
+affine.
 
@@
 Constructors and destructors like \lstinline/create/ and \lstinline/free/ manipulate the whole
-array. As with channels in \cref{sec:session-linearity},
+array. %As with channels in \cref{sec:session-linearity},
+The constructor
 \lstinline/create/ yields a linear resource which is consumed by \lstinline/free/.
@@
 contained in the lifetime of the resource.
-In a region, we can split off as many shared borrows as we like
-from a resource, but we can take only one exclusive borrow. In a
-subsidiary region, we can take further shared borrows from any
-borrow or we can take a exclusive borrow from a exclusive borrow from a
+In a region, we can split off as many shared borrows of a resource as
+we like, but we can take only one exclusive borrow. In a 
+subsidiary region, we can take shared borrows of any
+borrow or we can take an exclusive borrow of an exclusive borrow from an
 enclosing region. Borrows are confined to their regions. Inside the region,
@@
 The code in \cref{ex:array} demonstrates a use of the API to create an
-array of Fibonacci numbers. After creation of the array, we can see
-that the presence of a borrow prevents access to the ``raw''
-resource inside the borrow's region. In \cref{ex:array:get} of the
-example, there are two shared borrows in the same
-expression, which forms a region by itself. These borrows are split
-off the exclusive borrow used in \cref{ex:array:set}, which belongs to
-the next enclosing region.
-Finally, the code returns the whole array in \cref{ex:array:return},
-which is possible because none of the borrows are in scope anymore.
+array of Fibonacci numbers as the programmer would write it. After
+creation of the array, the presence of a borrow in the for loop
+prevents access to the ``raw'' resource inside the loop's
+body. \cref{ex:array:get} of the example contains two shared borrows
+in the same expression, which forms a region by itself. These borrows
+are split off the exclusive borrow used in \cref{ex:array:set}, which
+belongs to the next enclosing region corresponding to the loop body.
+The whole array can be returned in \cref{ex:array:return} because  the
+borrows are no longer in scope.
 
@@
 
-\cref{ex:array:region} uses braces as in \lstinline/{| ... |}/ to
-explicitly state the regions of the example in \cref{ex:array}.  One
+\cref{ex:array:region} uses braces \lstinline/{| ... |}/ to make the
+regions of the example in \cref{ex:array} explicit.  One
 region consists of the header expression of the \lstinline/let/ in
@@
 \lstinline{&a} in \cref{ex:array:region:get} has kind
-\lstinline{un_2} and the typing of its region is such that types with
+\lstinline{un_2} and the typing of the inner region is such that types with
 kind indexes greater than or equal to $2$ cannot escape.
+In the example, borrows cannot escape  because they are consumed
+immediately by \lstinline/get/ and \lstinline/set/.
 
-In \cref{ex:array:region:get} of the example, there is no danger of \lstinline/&a/ escaping because it
-is consumed in the calls to \lstinline/get/.
 Region annotations are automatically inserted using

### 2019-11-14 kindly/motivation.tex
fixed some grammar

@@
 particular branch of the choice.
-In particular,
-since the \lstinline/set/ function only requires an unrestricted shared borrow,
-\lstinline/try_solution/ can captures the borrow \lstinline/&g/
-while staying unrestricted.
-This also make \lstinline/try_solution/ itself bound to the same
-region as \lstinline/&g/. In this example, our region inference
-will place the region exactly at the \lstinline/begin//\lstinline/end/
-scope.
-This allows us to use \lstinline/try_solution/ in
-the iteration loop on \cref{line:iter}.
-Since \lstinline/&g/ is unrestricted, we could even use a parallel iteration
-instead of a sequential one.
-As \lstinline/g/ is linear we must free it before returning,
-in \cref{line:free:g}, outside of the region.
+As the \lstinline/set/ function only requires an unrestricted shared borrow,
+the closure \lstinline/try_solution/ remains unrestricted even though
+it captures the borrow \lstinline/&g/. 
+The price is that \lstinline/try_solution/ cannot escape from
+\lstinline/&g/'s region. In this example, inference places the region
+around the \lstinline/begin//\lstinline/end/ scope.
+Hence, \lstinline/try_solution/ can be used in
+the iteration in \cref{line:iter}.
+% As \lstinline/&g/ is unrestricted, we could even use a parallel iteration
+% instead of a sequential one.
+As \lstinline/g/ is linear we must free it outside of the region
+before returning (\cref{line:free:g}).
 

### 2019-11-14 kindly/lang.tex
formatting
section 3.3 and friends

@@
 linearity and borrows using kinds and (kind-) constrained types.
-For ease of presentation, we keep the language as simple as possible.
+For ease of presentation, we simplify the language.
 \begin{itemize}
@@
   \centering
-  \begin{minipage}{0.35\linewidth}
+  \begin{minipage}{0.49\linewidth}
     \begin{mathpar}
       \ruleSDPair
+
+      \ruleSDRegion
     \end{mathpar}
-    \caption{The {\sc Pair} rule}
+    % \caption{The {\sc Pair} rule}
+    \caption{Selected typing rules}
     \label{sdtyp:pair}
+    \label{selectrules:region}
   \end{minipage}\hfill
-  \begin{minipage}{0.6\linewidth}
+  \begin{minipage}{0.5\linewidth}
     \centering
@@
 \begin{figure}[tp]
-  \begin{minipage}{0.4\linewidth}
-    \centering
-    \begin{mathpar}
-      \ruleSDRegion
-    \end{mathpar}
-    \caption{The {\sc Region} rule}
-    \label{selectrules:region}
-  \end{minipage}\hfill
-  \begin{minipage}{0.6\linewidth}
+  % \begin{minipage}{0.4\linewidth}
+  %   \centering
+  %   \begin{mathpar}
+  %     \ruleSDRegion
+  %   \end{mathpar}
+  %   \caption{The {\sc Region} rule}
+  %   \label{selectrules:region}
+  % \end{minipage}\hfill
+  \begin{minipage}{\linewidth}
     \centering
@@
 
-In this section, we present the dynamics of \lang in big-step
+The dynamics of \lang is given in big-step
 functional
 style~\cite{siek13:_type_safet_three_easy_lemmas,DBLP:conf/esop/OwensMKT16,
-  DBLP:conf/popl/AminR17}. It is given as a function
+  DBLP:conf/popl/AminR17}.  A function
 @eval \Store \Perm \VEnv i e@
-that manipulates the semantic objects defined in
+manipulates the semantic objects defined in
 \cref{fig:syntax-internal-language}.
@@
 
-Expressions $e$ need to be elaborated with some kind and constraint
-annotations. Polymorphic functions give rise to elaborated
+The semantics is defined in terms of \emph{elaborated expressions} $e$
+with kind and constraint annotations inserted by the typechecker.
+Polymorphic functions give rise to elaborated
 abstractions which abstract over kind variables $\Multi\kappa$ and
-which are tagged with a kind $k$ indicating their usage restriction
-(linear, affine, etc). Their use
-gives rise to elaborated application where the instantiation of these
-kind variables is given. Pairs also come with a kind tag $k$
-indicating the usage restriction.
+which are tagged with a constraint $C$ and a kind $k$ indicating their
+usage restriction (linear, affine, etc). Their use
+gives rise to explicit instantiation of these kind variables. Pairs
+come with a kind tag $k$ indicating the usage restriction. Application
+is annotated with evidence $\Sp$ of the inferred splitting.
 
@@
 
-A permission $\Perm$ is a set of addresses. During evaluation it
-contains the set of addresses that may be accessed. For each raw
-location, there is only ever one address inside of $\Perm$.
+A permission $\Perm$ is a set of addresses that may be accessed during
+evaluation. A well-formed permission contains at most one address for each raw
+location.
 
-A result $r$ is either an address or a constant (e.g., a number). All
-non-trivial values are boxed in the sense that they are represented by
-the address of their actual content.
+Non-trivial results are boxed in the  semantics. So, a result
+$r$ is either an address or a primitive constant (e.g., a number).
 
-A value environment $\VEnv$ is a mapping from variables to results.
+A value environment $\VEnv$  maps variables to results.
 
-A storable $w$ is a particular object that can be stored. There are five
-kinds of storable. A poly closure represents a polymorphic
+A storable $w$ describes the content of a location in the store. There are five
+kinds of storables. A \emph{poly closure} represents a polymorphic
 function. It consists of an environment and the components of an
-elaborated abstraction. Similarly, a closure represents a monomorphic
-function and it elides kind variables and the constraint. A pair
-consists of a kind and two results. A resource is just an indirection
-to another result and the $\StFreed$ fills a released location.
+elaborated abstraction. A \emph{closure} represents a monomorphic
+function in the usual way. A \emph{pair}
+consists of a kind and two results. A \emph{resource} contains
+a result and the \emph{hole} $\StFreed$ fills a released location.
 
-Finally, the store $\Store$ is a mapping from raw locations to
-storables. It comes with a function
-@salloc@ of type @store -> storable -> (loc * store)@ such that
-@salloc delta@ @w@ allocates a free location in @delta@ and fills it with
-storable $w$. It returns the location and the updated store.
+A store $\Store$ is a partial map from raw locations to
+storables. The function
+@salloc: store -> storable -> (loc * store)@ is such that
+@salloc delta@ @w@ allocates an unused location in @delta@ and fills it with
+@w@. It returns the location and the extended store.
 
-The evaluation function is indexed by a step count so that each
+The evaluation function is indexed by a step count @i@ so that each
 invocation is guaranteed to terminate either with an error, a timeout,
-or a meaningful result. Hence its return type is a monad
-\lstinline/'a sem/ which combines error reporting, timeout, and a
-result of type \lstinline/'a/:
+or a result. Its return type is a monad
+\lstinline/'a sem/ which combines error reporting and timeout:
 \begin{lstlisting}
@@
 \cref{fig:big-step-interpretation} and leave the full
-definition for \cref{sec:semant-defin}. The appendix also
-contains the semantics
-in traditional notation, which does not make the monadic nature of
-evaluation explicit.
-
-The definition uses concrete OCaml syntax with some pretty
+definition for \cref{sec:semant-defin}.
+% The appendix also
+% contains the semantics in mathematical notation.
+The definition uses OCaml syntax with some pretty
 printing. The pervasive @let*@ operator acts as monadic bind
-for the \lstinline/sem/ monad. The function header of @eval@ checks
+for the \lstinline/sem/ monad. The operator
+\lstinline/let*? : bool -> (unit -> 'a sem) -> 'a sem/
+ converts a boolean
+argument into success or failure in the monad.
+ The function header of @eval@ checks
 whether time is up and otherwise proceeds processing the expression.
 
-The @Varinst@ case corresponds to the elaborated application. It
+The @Varinst@ case corresponds to instantiation. It
 obtains the variable's value, checks that it is a location, checks the
 permission (the @let*?@ clause), obtains the storable $w$ at that
-location, and checks that it is an STPOLY. Then it updates the
-permission: if the STPOLY closure is unrestricted, then the location
+location, and checks that it is a poly closure (STPOLY). Next, it updates the
+permission: if the poly closure is unrestricted, then the location
 remains in the permission set, otherwise it is removed. Finally, we
@@
 
-The @App@ case implements function application. Compared to the syntax
-it comes with an annotation @sp@ which contains evidence of the
-environment splitting in the application rule. This evidence takes the
-form of two sets of variables from the domain of @gamma@.
-
-To execute an application, we first apply the splitting to @gamma@ and
+The @App@ case implements (elaborated) function application.
+We first apply the splitting @sp@ to @gamma@ and
 evaluate subterm @e_1@ with its part of the environment and the
@@
 permitted to use. Moreover, there must be a monomorphic STCLOS stored
-at that location. The permission to further use this closure is
-withdrawn unless the closure is unrestricted. Finally, we evaluate the
+at that location. The permission to further use this closure  remains
+in force only if the closure is unrestricted. Finally, we evaluate the
 argument, then the function body, and return its result.
 
-The @Pair@ case does not present any new features, but it
-corresponds to the \TirName{Pair} typing rule.
+The @Pair@ case corresponds to the \TirName{Pair} typing rule. It
+requires no new features.
 

### 2019-11-14 kindly/lang.tex
Add description of the automatic region annotation.

@@
 % a data constructor $K$.
+
+\subsection{Automatic region annotation}
+\label{regionannot}
+
+In the examples presented in \cref{motivation}, region annotations are optional.
+In our formalization, regions must be syntactically explicit and annotated with
+a nesting index and a scoped variable.
+We now define a best-effort
+syntactic code transformation $\RannotT{p}{p'}$ which
+annotates programs automatically.
+The input $p$ is a program with optional region annotations of the form $\region[]{}{e}$.
+The output $p'$ is a program with explicit annotations
+of the form $\region{\Sone x\BORROW}{e}$.
+We give an informal presentation of our code transformation. The complete
+definition is given in \cref{appendix:regionannot}. This code transformation
+aims to find, for each borrow $\borrow{x}$, the biggest region satisfying
+the following rules:
+\begin{enumerate}
+\item The region should contain at least $\borrow{x}$.
+\item The region should never be larger than the scope of $x$.
+\item An exclusive borrow $\borrow[\kaff]{x}$ should never share a region with any other borrow of $x$.
+\item A use of $x$ can never be in the region of $\borrow{x}$.
+\end{enumerate}
+Following these rules, we start from the leafs and grow the region
+associated to each borrow until it can't grow anymore, either due to a binding
+or to conflicting uses.
+
+Let us consider the following program with binders and function applications:
+
+\[
+  \setlength{\jot}{-1pt}
+  \RannotT{
+\begin{aligned}
+  \lam{a}{}\ &
+  \letin{x}{\app{f}{\borrow[\kun]{a}}}{}\\
+  &g\ (\borrow[\kaff]{x});\\
+  &f\ (\borrow[\kun]{x})\ (\borrow[\kun]{x})\\
+\end{aligned}
+}{
+\begin{aligned}
+  \lam{a}{}\region[1]{\Sone a\MBORROW}{\ &%
+  \letin{x}{\app{f}{\borrow[\kaff]{a}}}{}\\
+  &\region[2]{\Sone x\MBORROW}{g\ (\borrow[\kaff]{x})};\\
+  &\region[2]{\Sone x\IBORROW}{f\ (\borrow[\kun]{x})\ (\borrow[\kun]{x})}\\
+  &\!\!\!}
+\end{aligned}
+}
+\]
+
+Since the only use of $a$ is a borrow,
+the region covers the entire lexical scope of $a$. For $x$ however, there are
+multiple conflicting borrow and we must be more careful. We place a first
+region around the exclusive borrow and its function application,
+and a second region around both shared borrows. This placement of region
+is optimal, as making any of the region bigger would cause a borrowing
+error.
+
+Naturally, these rules are best effort and it's sometimes necessary for the
+programmer to annotate regions explicitly. We take such manual annotations
+into account as an upper bound on all the contained regions.
+In the following program, a manual annotation has been inserted to ensure
+no borrow enters the reference $r$:
+
+\[
+  \setlength{\jot}{-1pt}
+  \RannotT{
+\begin{aligned}
+  \letin{&r}{\operatorname{ref}\ 0}{}\\
+  \lam{a}{}\ &
+  set\ r\ \region[]{}{g\ (\borrow[\kun]{a})};\\
+  &f\ (\borrow[\kun]{a}) 
+\end{aligned}
+}{
+\begin{aligned}
+  \letin{&r}{\operatorname{ref}\ 0}{}\\
+  \lam{a}{}\ &
+  set\ r\ \region[1]{\Sone a\IBORROW}{g\ (\borrow[\kun]{a})};\\
+  &\region[1]{\Sone a\IBORROW}{f\ (\borrow[\kun]{a})}
+\end{aligned}
+}
+\]
+
+Following the rules, we could merge the two regions around the borrows
+$\borrow[\kun]{a}$. However
+the explicit annotation indicates that the region should stay around the closure
+passed as argument of $set$. This is particularly useful to control captures
+by imperative APIs.
+
+This code transformation is purely syntactic and must be used before
+typing. It only produces well-nested annotations: if $\region[n]{b}{\dots}$
+is nested inside $\region[n']{b'}{\dots}$, then $n > n'$. Furthermore, there
+is at most one region per borrow, and exactly one region per exclusive borrow.
+In the rest of this article, we consider
+that all terms have been annotated with this code transformation and respect
+these properties.
 

### 2019-11-14 kindly/infer/extra.tex
Fixed uses of 'note' instead of 'denote' and various typos.
Add a macro for input parameters.

@@
   \input{infer/kinds}
-  \caption{Kind inference rules -- $\inferK{(C,\unif)}{{\bm\E}}{{\bm\tau}}{k}$}
+  \caption{Kind inference rules -- $\inferK{(C,\unif)}{{\inP\E}}{{\inP\tau}}{k}$}
   \label{rules:kinding}
@@
 %
-We note $\inferK{(C,\unif)}{{\bm\E}}{{\bm\tau}}{k}$ when type $\tau$ has kind $k$
+We write $\inferK{(C,\unif)}{{\inP\E}}{{\inP\tau}}{k}$ when type $\tau$ has kind $k$
 in environment $\E$ under constraints $C$ and unifier $\unif$.
-$\bm\E$ and $\bm\tau$ are the input parameters of
+$\inP\E$ and $\inP\tau$ are the input parameters of
 our inference procedure.
@@
   \caption{Type inference rules --
-    $\inferW{\Sigma}{(C,\psi)}{{\bm\E}}{{\bm e}}{\tau}$ }
+    $\inferW{\Sigma}{(C,\psi)}{{\inP\E}}{{\inP e}}{\tau}$ }
   \label{rules:typing:full}

### 2019-11-14 kindly/infer/infer.tex
Add a macro for input parameters.

@@
 \begin{itemize}[leftmargin=*]
-\item $\bsplit{C}{\Sv}{\bm{\Sv_1}}{\bm{\Sv_2}}$.
+\item $\bsplit{C}{\Sv}{\inP{\Sv_1}}{\inP{\Sv_2}}$.
   Given two usage environments $\Sv_1$ and $\Sv_2$
@@
   and uses the same rules as the one presented in \cref{sdtyping:split}.
-\item $\bregion[\bm{n}]{C}{\bm{x}}{\Sv}{\bm\Sv'}$.
+\item $\bregion[\inP{n}]{C}{\inP{x}}{\Sv}{\inP\Sv'}$.
   Given a usage environment $\Sv'$ and a variable name $x$, we return
@@
 
-We write $\inferW{\Sv}{(C,\unif)}{\bm{\E}}{\bm{e}}{\tau}$ when
+We write $\inferW{\Sv}{(C,\unif)}{\inP{\E}}{\inP{e}}{\tau}$ when
 $e$\ has type $\tau$ in $\E$ under the constraints $C$ and unifier $\unif$
-with a usage environment $\Sv$. $\bm\E$ and $\bm e$ are the input parameters of our
+with a usage environment $\Sv$. $\inP\E$ and $\inP{e}$ are the input parameters of our
 inference algorithm.
@@
 the environment,
-$\bsplit{C}{\Sv}{\bf{\Sv_1}}{\bf{\Sv_2}}$ and $\bregion{C}{\bf{x}}{\Sv}{\bf\Sv'}$,
+$\bsplit{C}{\Sv}{\inP{\Sv_1}}{\inP{\Sv_2}}$
+and $\bregion{C}{\inP{x}}{\Sv}{\inP\Sv'}$,
 which returns the necessary constraints.
@@
       }
-      { \entail{C}{\Cleq{\tapp{t}{(\tau_i)}}{\tapp{t}{(\tau'_i)}}} }
+      { \entail{C}{\Cleq{\tapp{\tcon}{(\tau_i)}}{\tapp{\tcon}{(\tau'_i)}}} }
       \and

### 2019-11-14 kindly/infer/infer.tex
Made a new pass on input parameters.

@@
 Unlike the syntax-directed version of our type system, knowing which elements
-are input and output or our inference judgement is critical. In the rest
+are input and output or our inference judgements is critical. In the rest
 of this section, when presenting a new judgement,
-we write input parameters in \textbf{bold}. The rest will then be
+we write input parameters in \textbf{\textcolor{DarkSlateBlue}{bold blue}}. The rest will then be
 output parameters.
@@
 \item $\bsplit{C}{\Sv}{\inP{\Sv_1}}{\inP{\Sv_2}}$.
-  Given two usage environments $\Sv_1$ and $\Sv_2$
+  Given two usage environments $\inP{\Sv_1}$ and $\inP{\Sv_2}$
   (inferred from two subexpressions, for instance),
@@
 \item $\bregion[\inP{n}]{C}{\inP{x}}{\Sv}{\inP\Sv'}$.
-  Given a usage environment $\Sv'$ and a variable name $x$, we return
-  $\Sv$ where the borrow binding of $x$ in $\Sv'$, if it exists, is replaced by
+  Given a usage environment $\inP{\Sv'}$, a nesting level $\inP{n}$,
+  and a variable name $\inP{x}$, we return
+  $\Sv$ where the borrow binding of $x$ in $\Sv'$, if it exists,
+  is replaced by
   a suspended binding. We also return $C$, a set of constraints that must
   be respected.
-  Again, the relation is total and non-ambiguous for any given $\Sv'$ and $x$,
+  Again, the relation is total and non-ambiguous for any given $\Sv'$,
+  $n$ and $x$,
   and uses the rules presented in \cref{sdtyping:regions}.
@@
 We write $\inferW{\Sv}{(C,\unif)}{\inP{\E}}{\inP{e}}{\tau}$ when
-$e$\ has type $\tau$ in $\E$ under the constraints $C$ and unifier $\unif$
+$\inP{e}$\ has type $\tau$ in $\inP{\E}$ under the constraints $C$ and unifier $\unif$
 with a usage environment $\Sv$. $\inP\E$ and $\inP{e}$ are the input parameters of our
@@
     \end{mathpar}
-    \caption{Base entailment rules -- $\entail{C}{D}$ }
+    \caption{Base entailment rules -- $\entail{\inP{C}}{\inP{D}}$ }
     \label{rules:entail}

### 2019-11-14 kindly/sdtyping/extra.tex
Add description of the automatic region annotation.
Add a macro for input parameters.
Made a new pass on input parameters.

@@
 \section{Automatic region annotation}
-\label{regionannot}
+\label{appendix:regionannot}
 
-So far, all \lang programs have been fully annotated with regions information.
-We now show how to infer these regions annotations based on
-optionally-annotated programs.
+In this section, we define our automatic region annotation
+which is presented in \cref{regionannot}.
 First, we extend the region annotation to $\region{S}{E}$ where $S$ is
@@
   \input{region-annot}
-  \caption{Automatic region annotation --- $\RannotT{e}{e'}$}
+  \caption{Automatic region annotation --- $\RannotT{\inP{e}}{e'}$}
   \label{fig:region-annotation}
@@
   \input{infer/entails}
-  \caption{Base entailment rules -- $\entail{C}{D}$ }
+  \caption{Base entailment rules -- $\entail{\inP{C}}{\inP{D}}$ }
   \label{rules:entail}
@@
 total bounded lattice $(\mathcal L, \lk_\Lat)$.
-Entailment is written $\entail{C}{D}$, where $D$ is a consequence of the
+Entailment is written $\entail{\inP{C}}{\inP{D}}$,
+where $D$ is a consequence of the
 constraints $C$.
@@
 \begin{align*}
-  \entail{\Ceq{x}{y}\wedge \Cproj{x}C \wedge \Ceq{x}{y}&}{C}
-  &\subst{x}{\tau}{P} &\equivC \Cproj{x}P \wedge \Ceq{x}{\tau}
+  &\entail{\Ceq{x}{y}\wedge \Cproj{x}C \wedge \Ceq{x}{y}}{C}\\
+  &\subst{x}{\tau}{P} \equivC \Cproj{x}P \wedge \Ceq{x}{\tau}
                         \text{ where } x\notin\fv{\tau}
@@
     environments $\lsplit C\E{\E_l}{\E_r}$;
-    inference $\bsplit C\E{\bm{\E_l}}{\bm{\E_r}}$;
-    binders $\bsplit Cb{\bm{b_r}}{\bm{b_l}}$}
+    inference $\bsplit C\E{\inP{\E_l}}{\inP{\E_r}}$;
+    binders $\bsplit Cb{\inP{b_r}}{\inP{b_l}}$}
   \label{fig:sd-splitting}
@@
     environments $\lregion[n]{C}{x}{\E}{\E'}$;
-    inference $\bregion[\bm n]{C}{\bm x}{\E}{\bm{\E'}}$;
-    binders $\bregion[\bm n]{C}{\bm x}{b}{\bm{b'}}$}
+    inference $\bregion[\inP n]{C}{\inP x}{\E}{\inP{\E'}}$;
+    binders $\bregion[\inP n]{C}{\inP x}{b}{\inP{b'}}$}
   \label{fig:sd-borrowing}
   \input{sdtyping/sd-environments}
-  \caption{Rewriting constraints on environments --- $\bm{\Cleq{\E}{k}}\Crewrite  C$}
+  \caption{Rewriting constraints on environments --- $\inP{\Cleq{\E}{k}}\Crewrite  C$}
   \label{fig:contraints-environments-types}

### 2019-11-14 kindly/infer/infer.tex
Non-commutative lattices are hell.

@@
 
-We define the constraint solver in terms of an arbitrary bounded
+We define the constraint solver in terms of an arbitrary commutative bounded
 lattice $(\mathcal L, \lk_\Lat)$, i.e.,
-a lattice which has a minimal and a maximal element ($l^\top$ and $l^\bot$).
+a lattice which has a minimal and a maximal element ($l^\top$ and $l^\bot$)
+and where meet and joins are commutative.
 We write lattice elements as $l$ and $\glb_i l_i$ (resp. $\lub_i l_i$)

### 2019-11-14 kindly/lang.tex
WIP pattern matching

@@
       \ruleSDVar
+      \and
+      \ruleSDPair
+      \and
+      \ruleSDRegion
+      \and
+      \ruleSDMatchPair
     \end{mathpar}
-    \caption{The {\sc Borrow}, {\sc Abs} and {\sc Var} rules}
+    \caption{Selected typing rules}
     \label{selectrules:borrow}
     \label{selectrules:binders}
+    \label{sdtyp:pair}
+    \label{selectrules:region}
   \end{minipage}
@@
 \label{sdtyping:split}
-\begin{figure*}[tp]
-  \centering
-  \begin{minipage}{0.49\linewidth}
-    \begin{mathpar}
-      \ruleSDPair
-
-      \ruleSDRegion
-    \end{mathpar}
-    % \caption{The {\sc Pair} rule}
-    \caption{Selected typing rules}
-    \label{sdtyp:pair}
-    \label{selectrules:region}
-  \end{minipage}\hfill
-  \begin{minipage}{0.5\linewidth}
+\begin{figure}[tp]
+  % \begin{minipage}{0.5\linewidth}
     \centering
@@
     \label{sdtyp:split}
-  \end{minipage}
-\end{figure*}
+  % \end{minipage}
+\end{figure}
 
@@
 
-As it would be cumbersome to markup regions manually, we developed an
-algorithm that starts from each borrow, creates a region around it,
-and extends it as much as possible while respecting other borrows
-and binders. The algorithm (see \cref{regionannot}) also provides the region's level
-and variable annotation.
-Henceforth, we consider terms with fully annotated regions.
+% As it would be cumbersome to markup regions manually, we developed an
+% algorithm that starts from each borrow, creates a region around it,
+% and extends it as much as possible while respecting other borrows
+% and binders. The algorithm (see \cref{regionannot}) also provides the region's level
+% and variable annotation.
+% Henceforth, we consider terms with fully annotated regions.
 
@@
 
-% \subsubsection{Pattern matching}
+\subsubsection{Pattern matching}
 
-% \TODO{Present this if we have enough space}
+\TODO{Present this if we have enough space}
 

### 2019-11-14 kindly/conclusion.tex
cosmetics-shrinking

@@
 
-The \lang language is an extension of ML-like languages with sound handling of linear and affine resources.
-It's main novel feature is the combination of full type inference and a practically useful notion of shared and exclusive borrowing of linear and affine resources.
+\lang is an ML-like language extended with sound handling of linear and affine resources.
+Its main novel feature is the combination of full type inference and a practically useful notion of shared and exclusive borrowing of linear and affine resources.
 Although the inferred types are much richer internally than plain ML types, most of that complexity can be hidden from user-level programmers.

### 2019-11-14 kindly/soa.tex
cosmetics-shrinking

@@
 between linear and unrestricted types.
-\citet{DBLP:conf/tldi/MazurakZZ10} also provide
-(and proves soundness) of a linearity-aware semantics.
+\citet{DBLP:conf/tldi/MazurakZZ10} provide
+a linearity-aware semantics with a soundness proof.
 Unlike \lang, System F\degree{} does not allow
@@
 does not admit a most general type for function composition.
-Furthermore, since it is based on System F, it does not admit
+Being based on System F, it does not admit
 principal type inference.
@@
 Mezzo~\citep{DBLP:phd/hal/Protzenko14} is an ML-like language
-with an extremely rich capability system which is able to encode numerous
+with a rich capability system which is able to encode numerous
 properties akin to separation logic~\citep{DBLP:conf/lics/Reynolds02}.
@@
 and Rust's lifetimes. This system assumes
-the existence of a linear type system and develop the associated compilation
-and runtime infrastructure. We believe our approaches are very
+the existence of a linear type system and develops the associated compilation
+and runtime infrastructure. We believe our approach is
 complementary and aim to combine them in the future.
@@
 
-Linear and affine disciplines have also been used in non-functional
-settings, notably in the context of low-level imperative programming
-and object systems. In particular, \lang uses borrows and regions
-which were initially developed in the context of imperative and
+% Linear and affine disciplines have  been used in non-functional
+% settings, notably in the context of low-level imperative programming
+% and object systems. In particular,
+\lang uses borrows and regions
+which were initially developed in the context of linear and affine
+typing for  imperative and
 object-oriented
@@
 for safe low-level programming.
-\lang heavily draws inspiration from Rust's borrowing system to apply
+\lang is inspired from Rust's borrowing system to apply
 it in a functional setting with type inference, a GC, and
@@
 Rust's lifetime system is more explicit and more expressive than \lang,
-however Rust does not provide type inference
+but Rust does not provide type inference
 and only provides \emph{partial} lifetime inference.
@@
 leverage typestate and capabilities
-to express very rich properties in objects and protocols.
+to express rich properties in objects and protocols.
 These systems are designed for either low-level or object-oriented
-programming, and do not immediately lend themselves to a more functional
-style. Furthermore, while their type-systems are much more
-powerful than \affe's, they require many annotations by the programmer
-and do not support any inference.
-It could however be interesting to expand \affe with limited
+programming and do not immediately lend themselves to a more functional
+style. While these systems are much more
+powerful than \affe's, they require programmer annotations
+and do not support inference.
+It  would be interesting to extend \lang with limited
 forms of typestate as a local, opt-in feature that would provide
@@
 \lang relies on constrained types
-to introduce the kind inequalities required for linear types. In
-particular, we use HM(X)~\citep{DBLP:journals/tapos/OderskySW99} which
+to introduce the kind inequalities required for linear types.
+HM(X)~\citep{DBLP:journals/tapos/OderskySW99} 
 allows us to use constrained types in an ML-like language with complete
@@
 could be applied to larger and more complex languages such as OCaml.
+Alternatively, we could have based \lang on qualified
+types~\cite{DBLP:journals/scp/Jones94}, similarly to Quill.
+% This
+% choice would also be sustainable as qualified types is part of the
+% foundation of Haskell's type system.
 
-Alternatively, we could have based the system on qualified
-types~\cite{DBLP:journals/scp/Jones94}, similarly to Quill. This
-choice would also be sustainable as qualified types is part of the
-foundation of Haskell's type system.
-
-The subtyping discipline used in \lang is very similar
+\lang's  subtyping discipline is similar
 to structural subtyping, where the only subtyping (or here, subkinding)

### 2019-11-15 kindly/infer/exampleconstr.tex
Improve placement of figures.
Reformat the annex a bit.
Fix all the remaining borked formating.

@@
 We now demonstrate the constraint solving
-algorithm presented in \cref{infer:solving}.
-Let us consider the expression $\lam{f}{\lam{x}{(\app{f}{x},x)}}$.
+algorithm presented in \cref{infer:solving}
+on the expression $\lam{f}{\lam{x}{(\app{f}{x},x)}}$.
 The application of our inference algorithm yields the following constraints:
@@
   \E &= (\tvar_f : \kvar_f)
-  (\tvar_x : \kvar_x)\dots
-  &C &= (\tvar_f \leq \gamma \tarr{\kvar_1} \beta )
+  (\tvar_x : \kvar_x)\dots\\
+  C &= (\tvar_f \leq \gamma \tarr{\kvar_1} \beta )
   \Cand
@@
 and the following kind constraints: 
-\begin{gather*}
-  \Cproj{\kvar_r}\Cproj{\kvar_f}\Cproj{\kvar_x}
+\[\begin{aligned}
+  \Cproj{\kvar_r}\Cproj{\kvar_f}\Cproj{\kvar_x}{}&
   (\kvar_x \leq \kun)
@@
   (\kvar_x \leq \kvar_r)
-  \Cand
-  (\kvar_\beta \leq \kvar_r)
+  \Cand\\
+  &(\kvar_\beta \leq \kvar_r)
   \Cand
@@
   (\kvar_f \leq \kvar_1)
-\end{gather*}
+\end{aligned}\]
 Since $\kvar_r$, $\kvar_f$ and $\kvar_x$ are not present in the type skeleton,
@@
 
-\begin{figure}[!hb]
-  \centering
-  \begin{minipage}{0.5\linewidth}
+\begin{figure}[tbp]
   \centering
@@
     \label{example:graph}
-\end{minipage}~
-\begin{minipage}{0.5\linewidth}
-  \centering
+
   \begin{tikzpicture}[every edge/.style=[link,->,>=latex,thick]]
@@
     \label{example:graph:final}
-\end{minipage}
 \end{figure}

### 2019-11-15 kindly/sdtyping/extra.tex
Fix all the remaining borked formating.

@@
   Given $(C_1,\unif_1)$ and $(C_2,\unif_2)$ such that
-  $\unif_1 C_1 \equivC \unif_2 C_2$, then
+  $\unif_1 C_1 \equivC \unif_2 C_2$,\\
+  then
   $\normalize{C_1}{\unif_1}$ and $\normalize{C_2}{\unif_2}$
@@
 \begin{theorem}
-  $\CL$ has the principal constraint property,
+  $\CL$ has the principal constraint property,\\
   $\operatorname{normalize}$ computes principal normal forms for $\CL$

### 2019-11-15 kindly/lang.tex
Pattern matching.

@@
     \label{selectrules:region}
+    \label{selectrules:matching}
   \end{minipage}
@@
 
+\subsubsection{Pattern matching}
+\label{sdtyping:matching}
+
+In order to deconstruct pairs in the presence of linear types, we require
+more than projection operators. Indeed, given a linear pair $x$,
+$\operatorname{fst} x$ would consume $x$ without giving access to the second
+member.
+Furthermore, we would like to be able to deconstruct a borrow of a pair
+and obtain borrows of the components.
+For these purposes, we provide the pattern matching construct
+$\matchin{x,x'}{e_1}{e_2}$. The typing rule is shown in \cref{selectrules:matching}.
+A syntactic marker $\etransfm$ indicates if it applies to
+a pair ($\etransfm = \operatorname{id}$) or a borrow ($\etransfm = \&^\BORROW$).
+If $\etransfm = \operatorname{id}$, it operators a regular destruct on pair.
+Otherwise, $e_1$ is expected to be of type
+$\borrowty{k}{\tyPair{\tau_1}{\tau'_1}}$
+while $x$ and $x'$ are of type respectively
+$\borrowty{k}{\tau_1}$ and $\borrowty{k}{\tau'_1}$.
+
+The typing of pattern matching is voluntarily simple here.
+Using a typeclass mechanism could avoid the need for explicit markers
+for borrow matches. Furthermore, a richer pattern language would allow
+finer-grained borrowing of inner components.
+We believe both extension are compatible with our system
+and are out of scope of this article.
+
 \subsubsection{Regions}
@@
 % \TODO{Present this if we have enough space}
-
-\subsubsection{Pattern matching}
-
-\TODO{Present this if we have enough space}
 

### 2019-11-15 kindly/lang.tex
Make the goal of the examples clearer.

@@
 Following these rules, we start from the leafs and grow the region
-associated to each borrow until it can't grow anymore, either due to a binding
+associated to each borrow until it can't grow anymore either due to a binding
 or to conflicting uses.
@@
 error.
+Notes that indices of regions are well nested, as the inner region have
+higher indices that the outer ones.
 
-Naturally, these rules are best effort and it's sometimes necessary for the
-programmer to annotate regions explicitly. We take such manual annotations
-into account as an upper bound on all the contained regions.
+Naturally, these rules are best effort and programmers must sometime
+annotate regions explicitly. We take such manual annotations
+into account as an upper bound on the regions of all contained borrows.
 In the following program, a manual annotation has been inserted to ensure
@@
 more than projection operators. Indeed, given a linear pair $x$,
-$\operatorname{fst} x$ would consume $x$ without giving access to the second
+$\app{\operatorname{fst}}{x}$ would consume $x$ without giving access to the second
 member.
@@
 and obtain borrows of the components.
-For these purposes, we provide the pattern matching construct
-$\matchin{x,x'}{e_1}{e_2}$. The typing rule is shown in \cref{selectrules:matching}.
+For these purposes, we provide the construct
+$(\matchin{x,x'}{e_1}{e_2})$ whose
+typing rule is shown in \cref{selectrules:matching}.
 A syntactic marker $\etransfm$ indicates if it applies to
 a pair ($\etransfm = \operatorname{id}$) or a borrow ($\etransfm = \&^\BORROW$).
-If $\etransfm = \operatorname{id}$, it operators a regular destruct on pair.
+If $\etransfm = \operatorname{id}$, the typing simplifies to
+a regular destruct on pairs.
 Otherwise, $e_1$ is expected to be of type

### 2019-11-15 kindly/motivation.tex
Make the goal of the examples clearer.

@@
 when manipulating mutable resources like buffers or connection pools
-using an imperative programming style. 
+using a mixed functional and imperative programming style.
 %
 To support this usage pattern of linearity, we rely on
-the notion of borrowing \cite{DBLP:conf/popl/BoylandR05}.
+the notion of borrowing \cite{DBLP:conf/popl/BoylandR05} which
+we demonstrate through a series of examples:
+\begin{itemize}
+\item An API for linear arrays to demonstrate simple borrowing and imperative programming.
+\item
+  A backtracking Sudoku solver to demonstrate
+  \lang's ability to build safe hybrid functional/imperative data-structures.
+\item A pool of linear objects to demonstrate advanced usage of regions.
+\end{itemize}
 
-\subsection{Imperative programming}
+\subsection{Imperative programming with linear arrays}
 \label{sec:imper-progr}
 
-A first example is an API for mutable linear arrays
+We now showcase an API for mutable linear arrays
 (\cref{sig:array}).
@@
     \vspace{-15pt}
-    \caption{Affine arrays}
+    \caption{Signature}
     \label{sig:array}
@@
     \vspace{-10pt}
-    \caption{Example of use of linear arrays}
+    \caption{Example of use}
     \label{ex:array}
@@
   \end{subfigure}
+  \vspace{-5pt}
+  \caption{Linear arrays}
+  \label{ex:array}
 \end{figure}
@@
 
-\subsection{Hybrid programming style}
+\subsection{Solving sudokus with hybrid data-structures}
 
@@
 \lstinline/set/ operation has type \lstinline/&('a t) * int * 'a -> 'a t/. It
-takes an shared borrow because it only reads from the
+takes a shared borrow because it only reads from the
 argument array, but returns a fresh,  modified structure.
@@
     \vspace{-10pt}
-    \caption{Interface}
+    \caption{Signature}
     \label{intf:pool}

### 2019-11-18 kindly/motivation.tex
Misc compression.

@@
 
-\lang{} is amenable to a functional resource-passing style as it is
+\lang{} is amenable to a functional resource-passing style
 common in functional encodings of session types (e.g.,
@@
   \lang's ability to build safe hybrid functional/imperative data-structures.
-\item A pool of linear objects to demonstrate advanced usage of regions.
+\item A pool of linear objects to demonstrate advanced use of regions.
 \end{itemize}
@@
 type ('a:'k) pool : 'k
-
 create : (unit -> 'a) -> 'a pool
@@
 \begin{lstlisting}
-type ('a:'k) pool : 'k = {
-  spawn : unit -> 'a ;
-  queue : 'a CQueue.t;
-}
-
-let create spawn =
-  { spawn ; queue = CQueue.create () }
-let consume f c =
-  CQueue.iter f c.queue
-
+type ('a:'k) pool : 'k = {spawn: unit -> 'a; queue: 'a CQueue.t}
+let create spawn = { spawn ; queue = CQueue.create () }
+let consume f c = CQueue.iter f c.queue
 let use { spawn ; queue } f =

### 2019-11-18 kindly/lang.tex
Fix Splitting rules on bindings.

@@
     \end{mathpar}
-    \caption{Selected typing rules}
+    \caption{Selected typing rules -- $\inferS{C}{\E}{e}{\tau}$}
     \label{selectrules:borrow}
@@
       &\bvar{x}{\schm}&\bvar{x}{\schm}&\bvar{x}{\schm}
-      &Both\\[2mm]
+      &Both\\
 
@@
       {\bvar{\borrow[\IBORROW]{x}}{\schm}}&{\bvar{\borrow[\IBORROW]{x}}{\schm}}
-      &Borrow\\[2mm]
+      &Borrow
+      % \\
 
-      {\Cempty}&{\bvar{x}{\schm}}&{\bvar{x}{\schm}}&{\bnone}
+      % {\Cempty}&
+      % \svar[\IBORROW]{x}{\schm}^n&\svar[\IBORROW]{x}{\schm}^n&\svar[\IBORROW]{x}{\schm}^n
+      % &BothS
+      \\[2mm]
+
+      {\Cempty}&{b_x}&{b_x}&{\bnone}
       &Left\\
-      {\Cempty}&{\bvar{x}{\schm}}&{\bnone}&{\bvar{x}{\schm}}
+      {\Cempty}&{b_x}&{\bnone}&{b_x}
       &Right\\[2mm]
@@
 
+      {\Cempty}&
+      {\svar{x} \schm}&{\svar[\IBORROW] x \schm^n}&{\svar{x} \schm}
+      &SuspS\\
     \end{tabular}
-    \caption{Splitting rules for bindings}
+    \caption{Splitting rules for bindings -- $\lsplit{C}{B}{B_l}{B_r}$}
     \label{sdtyp:split}
@@
 either to the left or right environment.
-The rules {\sc Susp} and {\sc SuspB} split off suspended bindings to
+The rules {\sc Susp}, {\sc SuspB} and {\sc SuspS}
+split off suspended bindings to
 the left while conserving access to the binding on the right.
@@
 and then later as a full resource on the right. The {\sc Susp} rule
-works with a full resource while the {\sc SuspB}
-rule splits off an shared suspended binding from a borrow.
+works with a full resource, the {\sc SuspB}
+with a borrow and {\sc SuspS} with a suspended binding.
 %

### 2019-11-18 kindly/infer/infer.tex
Gather the selected inference rules.
Compact grammar 2
Herbrand!
Remove some unnecessary metatheory from the body.

@@
 %
-\begin{figure}[tp]
+\begin{figure*}[tb]
   \begin{mathpar}
-    { \Weaken_{\bvar{x}{\sigma}}(\Sv) =
-      \begin{dcases}
-        \Ctrue&\text{if } x\in\Sv\\
-        \Cleq{\sigma}{\kaff_\infty}&\\
-      \end{dcases}
-    }
-    \and
     \ruleIVar
@@
     \ruleIAbs
+    \and
+    \ruleILet
+    \hfill
+    \ruleIRegion
+    \and
+    \ruleIPair
+    \and
+    \inferrule{}
+    { \Weaken_{\bvar{x}{\sigma}}(\Sv) =
+      \text{if } (x\in\Sv) \text{ then} \Ctrue \text{else }
+      \Cleq{\sigma}{\kaff_\infty}
+      % \begin{dcases}
+      %   \Ctrue&\text{if } x\in\Sv\\
+      %   \Cleq{\sigma}{\kaff_\infty}&\\
+      % \end{dcases}
+      \\\\\\
+      \generalize{C}{\E}{\tau} =
+      (\Cproj{(\Multi{\kvar},\Multi{\tvar})}{C},
+      \forall (\Multi{\kvar},\Multi{\tvar}).\qual{C}{\tau})\\
+      \text{ where}\\\\
+      \Multi{\kvar},\Multi{\tvar} = (\fv{\tau}\cup\fv{C})\setminus\fv{\E}
+    }
   \end{mathpar}
-  \caption{The {\sc Abs$_I$} and {\sc Var$_I$} rules}
+  \vspace{-15pt}
+  \caption{Selected inference rules}
   \label{rule:infer:envs}
-\end{figure}
+  \label{rule:infer:envrules}
+  \label{rule:infer:let}
+\end{figure*}
 %
@@
 \label{infer:regions}
-\begin{figure}[tb]
-  \centering
-  \begin{mathpar}
-    \ruleIPair
 
-    \ruleIRegion
-  \end{mathpar}
-  \caption{The {\sc Pair$_I$} and {\sc Region$_I$} rules}
-  \label{rule:infer:envrules}
-\end{figure}
-%
 We now present the inference version
@@
 \subsubsection{Generalization and constraints}
-%
-\begin{figure}[tbp]
-  \centering
-  \begin{mathpar}
-    \ruleILet
 
-    \inferrule{}
-    {\generalize{C}{\E}{\tau} =
-      (\Cproj{(\Multi{\kvar},\Multi{\tvar})}{C},
-      \forall (\Multi{\kvar},\Multi{\tvar}).\qual{C}{\tau})\\
-      \text{ where}\quad
-      \Multi{\kvar},\Multi{\tvar} = (\fv{\tau}\cup\fv{C})\setminus\fv{\E}
-    }
-  \end{mathpar}
-  \caption{The {\sc Let$_I$} rule}
-  \label{rule:infer:let}
-\end{figure}
-%
 So far, constraints have only been composed of a list of kind inequalities.
@@
     \begin{align*}
-      C &::= \Cleq{\tau_1}{\tau_2} \\
-        &\mid \Cleq{k_1}{k_2} \tag{Inequalities} \\
-        &\mid C_1 \Cand C_2  \tag{Conjunction}\\
+      C &::=  C_1 \Cand C_2  \tag{Conjunction}\\
+        &\mid \Cleq{\tau_1}{\tau_2} \mid \Cleq{k_1}{k_2} \tag{Inequalities} \\
         &\mid \Cproj{\tvar}{C}
-        \mid \Cproj{\kvar}{C} \tag{Projection}
+        \mid \Cproj{\kvar}{C} \tag{Projections}
     \end{align*}
@@
 \begin{enumerate}
-\item Solve all type equality constraints through Herband unification and
+\item Solve all type equality constraints through Herbrand unification and
   gather all existential quantifications at the front of the constraint.
@@
 An example of this algorithm in action is shown in \cref{solving:example}.
-Our algorithm is complete, computes principal normal forms,
+Our algorithm is complete, computes principal normal forms
+(cf \cref{appendix:constraints}),
 and already simplifies constraints significantly
@@
 \citep{DBLP:conf/aplas/Simonet03}.
+
 % 
-We now show that this algorithm is well-behaved with respect to entailment
-and computes unique normal forms.
+% We now show that this algorithm is well-behaved with respect to entailment
+% and computes unique normal forms.
 
-\begin{property}[Principal normal form]
-  Normalization computes principal normal forms for $\CL$, i.e.
-  given a constraint $D\in\CL$, a substitution $\phi$ and
-  $(C,\unif) = \normalize{D}{\phi}$,
-  then $\phi\leq\unif$,
-  $C \equivC \unif D$ and
-  $\unif C = C$.
-\end{property}
+% \begin{property}[Principal normal form]
+%   Normalization computes principal normal forms for $\CL$, i.e.
+%   given a constraint $D\in\CL$, a substitution $\phi$ and
+%   $(C,\unif) = \normalize{D}{\phi}$,
+%   then $\phi\leq\unif$,
+%   $C \equivC \unif D$ and
+%   $\unif C = C$.
+% \end{property}
 
-\begin{property}[Principal constraints]
-  $\CL$ has the principal constraint property, i.e.
-  for every $D\in\CL$, and a substitution $\unif$,
-  either $D$ does not have a normal form, or it has
-  a principal normal form.
-\end{property}
+% \begin{property}[Principal constraints]
+%   $\CL$ has the principal constraint property, i.e.
+%   for every $D\in\CL$, and a substitution $\unif$,
+%   either $D$ does not have a normal form, or it has
+%   a principal normal form.
+% \end{property}
 
-\begin{property}[Regular constraint system]
-  $\CL$ is regular, ie, for $x, x'$ two types or kinds,
-  $\entail{}{\Ceq{x}{x'}}$ implies
-  $\fv{x} = \fv{x'}$
-\end{property}
+% \begin{property}[Regular constraint system]
+%   $\CL$ is regular, ie, for $x, x'$ two types or kinds,
+%   $\entail{}{\Ceq{x}{x'}}$ implies
+%   $\fv{x} = \fv{x'}$
+% \end{property}
 
-The proofs are given in \cref{appendix:constraints}. These three properties
-are sufficient to state that $HM(\CL)$ provides principal type inference.
-However, we use an extension of HM(X) with kind inference
-and affine types and borrows.
-We now show that HM(X)'s completeness and soundness theorems are still valid on
-this extended system.
+% The proofs are given in \cref{appendix:constraints}. These three properties
+% are sufficient to state that $HM(\CL)$ provides principal type inference.
+% However, we use an extension of HM(X) with kind inference
+% and affine types and borrows.
+% We now show that HM(X)'s completeness and soundness theorems are still valid on
+% this extended system.
 
@@
 
-We now show that our extended inference algorithm still satisfies the soundness
+We now state that our extended inference algorithm still satisfies the soundness
 and completeness properties of HM(X).
@@
 
-\begin{definition}[More general unifier]
-  Given a set of variable $U$ and $\unif$, $\unif'$ and $\phi$
-  substitutions. \\
-  Then
-  $\unif \leq^{\phi}_{U} \unif'$ iff $(\phi \circ \unif)|_{U} = \unif'|_U$.
-\end{definition}
+% \begin{definition}[More general unifier]
+%   Given a set of variable $U$ and $\unif$, $\unif'$ and $\phi$
+%   substitutions. \\
+%   Then
+%   $\unif \leq^{\phi}_{U} \unif'$ iff $(\phi \circ \unif)|_{U} = \unif'|_U$.
+% \end{definition}
 
@@
   \Eflat\E
-  &= \left\{ \bvar{x}{\tau} \mid
-    \bvar{x}{\tau}\in\E
+  =& \left\{\bvar{x}{\tau}\in\E \mid
     \vee \bvar{\borrow{x}}{\borrowty{k}{\tau}}\in\E
@@
 
-We can then state the completeness theorem.
+We can then state principality.
 
-\begin{theorem}[Completeness]
-  Given $\inferS{C'}{\E'}{e}{\tau'}$ and
-  $\entail{}{\unif'\E \preceq \E'}$.
-  Then $$\inferW{\Sv}{(C,\unif)}{\Eflat\E}{e}{\tau}$$
-  for some environment $\Sv$,
-  substitution $\unif$, constraint $C$ and type $\tau$ such
-  that
-  \begin{align*}
-    \unif &\leq^{\phi}_{\fv{\E}} \unif'
-    &\entail{C'&}{\phi C}
-    &\entail{&}{\phi \schm \preceq \schm'}
-    &\Sv&\subset\E
-    % &( C, \schm, \unif) &\leq (C',\schm',\unif')
-  \end{align*}
-  where $\schm' = \generalize{C'}{\E'}{\tau'}$
-  and $\schm = \generalize{C}{\E}{\tau}$
-\end{theorem}
+% \begin{theorem}[Completeness]
+%   Given $\inferS{C'}{\E'}{e}{\tau'}$ and
+%   $\entail{}{\unif'\E \preceq \E'}$.
+%   Then $$\inferW{\Sv}{(C,\unif)}{\Eflat\E}{e}{\tau}$$
+%   for some environment $\Sv$,
+%   substitution $\unif$, constraint $C$ and type $\tau$ such
+%   that
+%   \begin{align*}
+%     \unif &\leq^{\phi}_{\fv{\E}} \unif'
+%     &\entail{C'&}{\phi C}
+%     &\entail{&}{\phi \schm \preceq \schm'}
+%     &\Sv&\subset\E
+%     % &( C, \schm, \unif) &\leq (C',\schm',\unif')
+%   \end{align*}
+%   where $\schm' = \generalize{C'}{\E'}{\tau'}$
+%   and $\schm = \generalize{C}{\E}{\tau}$
+% \end{theorem}
 
-Finally, principality is a direct application of completeness to
-toplevel programs.
+% Finally, principality is a direct application of completeness to
+% toplevel programs.
 
-\begin{corollary}[Principality]
+\begin{theorem}[Principality]
   Let $\inferS{\Ctrue}{\E}{e}{\schm}$ a closed typing judgement.\\
@@
 
-\end{corollary}
+\end{theorem}
 

### 2019-11-18 kindly/sdtyping/extra.tex
Remove some unnecessary metatheory from the body.

@@
 
-\begin{lemma}[Normal form]
+\begin{lemma}[Principal normal form]
   \label{lemma:normalform}
@@
 
-\begin{property}
+\begin{property}[Regular constraint system]
   $\CL$ is regular, ie, for $x, x'$ two types or kinds,
@@
 
-\begin{theorem}
+\begin{theorem}[Principal constraints]
   $\CL$ has the principal constraint property,\\

### 2019-11-18 kindly/infer/exampleconstr.tex
Put the constraint example in the body, and push all the constrain theory out.

@@
-We now demonstrate the constraint solving
-algorithm presented in \cref{infer:solving}
-on the expression $\lam{f}{\lam{x}{(\app{f}{x},x)}}$.
+\label{solving:example}
+
+\begin{figure}[tbp]
+  \centering
+  \begin{tikzpicture}[node distance=7mm,every edge/.style=[link,->,>=latex,thick]]
+    \node (U) {$\kun$} ;
+    \node[above left=of U] (x) {$\kvar_x$} ;
+    \node[above=of x] (r) {$\kvar_r$};
+    \node[left=of x] (g) {$\kvar_\gamma$} ;
+    \node[right=of x] (b) {$\kvar_\beta$} ;
+    \node[right=of b] (3) {$\kvar_3$} ;
+    \node[above=of 3] (f) {$\kvar_f$} ;
+    \node[above=of f] (1) {$\kvar_1$} ;
+
+    \draw (x) to[bend right] (U) ;
+    \draw (x) -> (r) ;
+    \draw (b) -> (r) ;
+    \draw (g) -> (x) ;
+    \draw (3) -> (f) ;
+    \draw (f) -> (1) ;
+
+    \draw[blue] (3) to[bend right] (1);
+
+    \node at (-0.5,-0.5) {Before};
+
+    \begin{scope}[dashed,gray]
+      \draw (U) to[bend right] (x) ;
+      \draw (U) to (b) ;
+      \draw (U) to[bend left] (g) ;
+      \draw (U) to (r) ;
+      \draw (U) to (1) ;
+      \draw (U) to (3) ;
+      \draw (U) to (f) ;
+    \end{scope}
+
+    \begin{scope}[on background layer]
+      \node[fill=green!20,draw=green,
+      inner sep=-1pt,ellipse,rotate fit=-20,fit=(U.south) (x) (g)] {};
+      \node[fill=red!20,draw=red, inner sep=-2pt, circle,fit=(r)] {};
+      \node[fill=red!20,draw=red, inner sep=-2pt, circle,fit=(f)] {};
+    \end{scope}
+  \end{tikzpicture}~
+  \vrule~
+  \begin{tikzpicture}[node distance=7mm,every edge/.style=[link,->,>=latex,thick]]
+    \node (b) {$\kvar_\beta$} ;
+    \node[right=of b] (3) {$\kvar_3$} ;
+    \node[above=of 3] (f) {} ;
+    \node[above=of f] (1) {$\kvar_1$} ;
+
+    \draw (3) to[bend right] (1);
+
+    \node at (0.5,-1.5) {After};
+  \end{tikzpicture}
+    \caption{Graph representing the example constraints}
+    \label{example:graph}
+    % \caption{Final state of the graph}
+    \label{example:graph:final}
+\end{figure}
+
+We consider the expression $\lam{f}{\lam{x}{(\app{f}{x},x)}}$.
 The application of our inference algorithm yields the following constraints:
@@
 \[\begin{aligned}
-  \Cproj{\kvar_r}\Cproj{\kvar_f}\Cproj{\kvar_x}{}&
-  (\kvar_x \leq \kun)
+    % \Cproj{\kvar_r}\Cproj{\kvar_f}\Cproj{\kvar_x}{}
+    &(\kvar_x \leq \kun)
     \Cand
@@
 \end{aligned}\]
-Since $\kvar_r$, $\kvar_f$ and $\kvar_x$ are not present in the type skeleton,
-they are existentially quantified.
+% Since $\kvar_r$, $\kvar_f$ and $\kvar_x$ are not present in the type skeleton,
+% they are existentially quantified.
 
@@
 \begin{itemize}[noitemsep]
-\item[1] From the constraints above, we deduce the graph shown
-  with plain arrows.
-\item[2] We add all the dashed arrows by saturating
+\item From the constraints above, we deduce the graph shown
+  with plain arrows on the left
+\item We add all the dashed arrows by saturating
   lattice inequalities. For clarity, we only show $\kun$.
-\item[4] We identify the connected component circled in
+\item We identify the connected component circled in
   {\color{green} green}.
   We deduce $\kvar_\gamma = \kvar_x = \kun$
-\item[5] We take the transitive closure, which adds the
+\item We take the transitive closure, which adds the
   arrow in {\color{blue} blue} from $\kvar_1$ to $\kvar_3$.
-\item[6] We remove the remaining existentially quantified node in {\color{red} red}, $\kvar_r$ and $\kvar_f$.
-\item[7, 8, 9] We clean up the graph, and obtain the graph shown
-  in \cref{example:graph:final}.
+\item We remove the remaining nodes not present in the type skeleton (colored in {\color{red} red}): $\kvar_r$ and $\kvar_f$.
+\item We clean up the graph (transitive reduction, remove unneeded constants, \dots),
+  and obtain the graph shown on the right.
   We deduce $\kvar_3 \leq \kvar_1$.
@@
 
-\begin{figure}[tbp]
-  \centering
-  \begin{tikzpicture}[every edge/.style=[link,->,>=latex,thick]]
-    \node (U) {$\kun$} ;
-    \node[above left=of U] (x) {$\kvar_x$} ;
-    \node[above=of x] (r) {$\kvar_r$};
-    \node[left=of x] (g) {$\kvar_\gamma$} ;
-    \node[right=of x] (b) {$\kvar_\beta$} ;
-    \node[right=of b] (3) {$\kvar_3$} ;
-    \node[above=of 3] (f) {$\kvar_f$} ;
-    \node[above=of f] (1) {$\kvar_1$} ;
-
-    \draw (x) to[bend right] (U) ;
-    \draw (x) -> (r) ;
-    \draw (b) -> (r) ;
-    \draw (g) -> (x) ;
-    \draw (3) -> (f) ;
-    \draw (f) -> (1) ;
-
-    \draw[blue] (3) to[bend right] (1);
-
-    \begin{scope}[dashed,gray]
-      \draw (U) to[bend right] (x) ;
-      \draw (U) to (b) ;
-      \draw (U) to[bend left] (g) ;
-      \draw (U) to (r) ;
-      \draw (U) to (1) ;
-      \draw (U) to (3) ;
-      \draw (U) to (f) ;
-    \end{scope}
-
-    \begin{scope}[on background layer]
-      \node[fill=green!20,draw=green,
-      inner sep=-1mm,ellipse,rotate fit=-20,fit=(U) (x) (g)] {};
-      \node[fill=red!20,draw=red, inner sep=-2pt, circle,fit=(r)] {};
-      \node[fill=red!20,draw=red, inner sep=-2pt, circle,fit=(f)] {};
-    \end{scope}
-  \end{tikzpicture}
-    \caption{Graph representing the example constraints}
-    \label{example:graph}
-
-  \begin{tikzpicture}[every edge/.style=[link,->,>=latex,thick]]
-    \node (b) {$\kvar_\beta$} ;
-    \node[right=of b] (3) {$\kvar_3$} ;
-    \node[above=of 3] (f) {} ;
-    \node[above=of f] (1) {$\kvar_1$} ;
-
-    \draw (3) to[bend right] (1);
-
-  \end{tikzpicture}
-    \caption{Final state of the graph}
-    \label{example:graph:final}
-\end{figure}
-
 %%% Local Variables:

### 2019-11-18 kindly/infer/infer.tex
Put the constraint example in the body, and push all the constrain theory out.

@@
 
-We define the constraint solver in terms of an arbitrary commutative bounded
-lattice $(\mathcal L, \lk_\Lat)$, i.e.,
-a lattice which has a minimal and a maximal element ($l^\top$ and $l^\bot$)
-and where meet and joins are commutative.
-We write lattice elements as $l$ and $\glb_i l_i$ (resp. $\lub_i l_i$)
-for the greatest lower bound (resp. least upper bound) in $\mathcal L$.
-The lattice for \lang (see \cref{sdtyping}) is a bounded lattice with
-$l^\top = \klin_\infty$ and $l^\bot = \kun_0$.
+%
+% \begin{figure}[tb]
+%   \centering
+%   \begin{align*}
+%     C &::= \Cleq{\tau_1}{\tau_2}
+%         \mid \Cleq{k_1}{k_2}
+%         \mid C_1 \Cand C_2
+%         \mid \Cproj{\tvar}{C}
+%         \mid \Cproj{\kvar}{C}
+%   \end{align*}
+%   \caption{The constraint language}
+%   \label{grammar:constraint}
+%   % \input{infer/lattice}
+%   % \caption{Lattice inequalities -- $k \lk_\Lat k'$}
+%   \input{infer/entails}
+%   \caption{Base entailment rules -- $\entail{\inP{C}}{\inP{D}}$ }
+%   \label{rules:entail}
+% \end{figure}
 
-Let $\CL$ be the set of constraints in such a lattice $\mathcal L$.
-The full grammar of constraints is shown in \cref{grammar:constraint}.
-Constraints are made of kind inequalities, conjunctions and
-projections, as shown previously, along with type unification
-constraints. Since types might contain kinds (for instance, on the arrows),
-type unification is oriented and written as $\leq$.
-For simplicity, we consider all type constructors
-invariant in their parameters
-and define $\Ceq{\tau}{\tau'}$ as $\Cleq{\tau}{\tau'} \wedge \Cleq{\tau'}{\tau}$.
+For concision, we demonstrate our constraint solving
+algorithm on an example. The complete constraint system is
+defined in \cref{appendix:constraints}.
+%
+\input{infer/exampleconstr}
 
-As before, entailment is denoted by $\entail{C}{D}$,
-where $D$ is a consequence of the constraints $C$.
-We say that $C$ and $D$ are equivalent, $C \equivC D$,
-when $\entail{C}{D}$ and $\entail{D}{C}$.
-The base rules for entailment are given in \cref{rules:entail}.
-The rules for classic Herbrand unification and
-projection operators are omitted for brevity (see
-\citep{DBLP:journals/tapos/OderskySW99} for
-details).
-
-\begin{figure*}[!bt]
-  \centering
-  \begin{minipage}{0.32\linewidth}
-    \begin{align*}
-      C &::=  C_1 \Cand C_2  \tag{Conjunction}\\
-        &\mid \Cleq{\tau_1}{\tau_2} \mid \Cleq{k_1}{k_2} \tag{Inequalities} \\
-        &\mid \Cproj{\tvar}{C}
-        \mid \Cproj{\kvar}{C} \tag{Projections}
-    \end{align*}
-    \caption{The constraint language}
-    \label{grammar:constraint}
-  \end{minipage}\hfill
-  \begin{minipage}{0.66\linewidth}
-    \begin{mathpar}
-      \inferrule
-      {\forall i,\ \entail{C}{\Cleq{l_i}{k}}}
-      {\entail{C}{\Cleq{\wedge_i l_i}{k}}}
-      \and
-      \inferrule
-      {\forall i,\ \entail{C}{\Cleq{k}{l_i}}}
-      {\entail{C}{\Cleq{k}{\vee_i l_i}}}
-      \and
-      \inferrule
-      { \forall i,\ \entail{C}{\Ceq{\tau_i}{\tau_i}}\\
-      }
-      { \entail{C}{\Cleq{\tapp{\tcon}{(\tau_i)}}{\tapp{\tcon}{(\tau'_i)}}} }
-      \and
-      \inferrule{l \lk_\Lat l'}{\entail{}{\Cleq{l}{l'}}}
-      \and
-      \inferrule
-      { \entail{C}{\Cleq{\tau'_1}{\tau_1}}\\
-        \entail{C}{\Cleq{\tau_2}{\tau'_2}}\\
-        \entail{C}{\Cleq{k}{k'}}
-      }
-      { \entail{C}{\Cleq{\tau_1\tarr{k}\tau_2}{\tau'_1\tarr{k'}\tau'_2}} }
-    \end{mathpar}
-    \caption{Base entailment rules -- $\entail{\inP{C}}{\inP{D}}$ }
-    \label{rules:entail}
-  \end{minipage}
-\end{figure*}
-
-We define the set of solved formed
-$\mathcal S$ as the quotient set of $\CL$ by $\equivC$.
-We will show later that such constraints are in fact only composed of
-kind inequalities, and thus correspond to the syntactic constraints
-used in type and kind schemes.
-%
-We now define our normalization procedure $\normalize{C_0}{\unif_0}$, where
-$C_0\in \CL$ is a set of constraints and $\unif_0$ is a substitution.
-It returns a constraint $C \in \mathcal S$ in
-solved form and a unifier $\unif$.
-The main idea of the algorithm is to first remove all the type equalities
-by using regular Herbrand unification. After that, we only have
-a set of inequalities among kinds, which we can consider as a relation.
-We can then saturate the relation,
-unify all kinds that are in the same equivalence classes to obtain
-a most general unifier on kind variables,
-remove all existentially quantified variables and
-then minimize back the relation and apply various
-simplification rules to make the resulting type easier to understand to users.
-
-More precisely, we apply the following steps:
-\begin{enumerate}
-\item Solve all type equality constraints through Herbrand unification and
-  gather all existential quantifications at the front of the constraint.
-  We obtain a constraint $C^k = \exists \Multi\kvar,\ \Cleq{k_j}{k'_j}_j$ and
-  a substitution $\unif_\tau$.
-
-  We write $\mathcal R$ for the relation $\Cleq{k_j}{k'_j}_j$,
-  $\mathcal G$ the underlying directed graph and $V$ its vertices.
-
-\item Saturate the lattice equalities in $\mathcal R$.
-
-  More precisely, for each kind variable $\kvar \in V$,
-  for each constant $l_i$ (resp. $l_j$) such that
-  there is a path from $l_i$ to $\kvar$ (resp. from $\kvar$ to $l_j$) in $\mathcal G$,
-  add an edge from $\lub l_i$ to $\kvar$
-  (resp. from $\kvar$ to $\glb l_j$).
-  This step is well defined since $\mathcal L$ is a bounded lattice
-  and $\lub\emptyset$ and $\glb\emptyset$ are well defined.
-
-  We also complement $\mathcal R$ with $(\leq)$ by adding an edge
-  between related constants.
-\item
-  At this point, we can easily check for satisfiability: A constraint
-  is satisfiable (in the given environment) if and only if,
-  for any constants $l_1$ and $l_2$ such that
-  there is a path from $l_1$ to $l_2$ in $\mathcal G$, then $l_1\lk_\Lat l_2$.
-  If this is not the case, we return \textbf{fail}.
-
-\item For each strongly connected component in $\mathcal G$, unify all its vertices and replace it by a representative.
-  We write $\unif_k$ for the substitution that replaces a kind variable by
-  its representative.
-  The representative of a strongly connected component $g$ can be determined as follows:
-  \begin{itemize}
-  \item If $g$ does not contain any constant, then the representative
-    is a fresh kind variable.
-  \item If $g$ contains exactly one constant, it is the representative.
-  \item Otherwise, the initial constraint $C_0$ is not satisfiable.
-  \end{itemize}
-  Note that this step will also detect all unsatisfiable constraints.
-\item Take the transitive closure of $\mathcal R$.
-\item Remove all the vertices corresponding to the kind variables $\Multi\kvar$
-  that are existentially quantified in $C^k$.
-\item Take the transitive reduction of $\mathcal R$.
-\item Remove the extremums of $\mathcal L$ and the edges of $(\leq)$
-  from $\mathcal R$.
-\item Return $C = \left\{ k \leq k' \mid k \operatorname{\mathcal R}k' \right\}$
-  and $\unif =  \unif_\tau \meet \unif_k$.
-\end{enumerate}
-
-An example of this algorithm in action is shown in \cref{solving:example}.
-Our algorithm is complete, computes principal normal forms
-(cf \cref{appendix:constraints}),
-and already simplifies constraints significantly
-(thanks to steps 6, 7 and 8).
-It can be extended with further simplification phases.
-In particular, our implementation and all the signatures presented in
-\cref{motivation} use a variance-based simplification
-where all covariant (resp. contravariant) variables are replaced by their
-lower (resp. upper) bounds.
+Our algorithm respect the properties
+of HM(X), computes principal normal forms
+and simplifies constraints significantly.
 All the simplification mechanisms presented

### 2019-11-18 kindly/sdtyping/extra.tex
Put the constraint example in the body, and push all the constrain theory out.

@@
 \label{appendix:constraints}
-%
+
+% \newcommand\A{\mathcal A}
+% \newcommand\SC{\mathcal S}
+
+We place our constraint system in a more general setting.
+We define the constraint solver in terms of an arbitrary commutative bounded
+lattice $(\mathcal L, \lk_\Lat)$, i.e.,
+a lattice which has a minimal and a maximal element ($l^\top$ and $l^\bot$)
+and where meet and joins are commutative.
+We write lattice elements as $l$ and $\glb_i l_i$ (resp. $\lub_i l_i$)
+for the greatest lower bound (resp. least upper bound) in $\mathcal L$.
+The lattice for \lang (see \cref{sdtyping}) is a bounded lattice with
+$l^\top = \klin_\infty$ and $l^\bot = \kun_0$.
+
+% Our system is defined in \cref{infer:solving}.
+% As before, we write $\CL$ for our constraint system for an arbitrary
+% total bounded lattice $(\mathcal L, \lk_\Lat)$.
+% Entailment is written $\entail{\inP{C}}{\inP{D}}$,
+% where $D$ is a consequence of the
+% constraints $C$.
+% We say that $C$ and $D$ are equivalent, written as $C \equivC D$,
+% when $\entail{C}{D}$ and $\entail{D}{C}$.
+
 \begin{figure}[tb]
@@
 
-% \newcommand\A{\mathcal A}
-% \newcommand\SC{\mathcal S}
+Let $\CL$ be the set of constraints in such a lattice $\mathcal L$.
+The full grammar of constraints is shown in \cref{grammar:constraint}.
+Constraints are made of kind inequalities, conjunctions and
+projections along with type unification
+constraints. Since types might contain kinds (for instance, on the arrows),
+type unification is oriented and written as $\leq$.
+For simplicity, we consider all type constructors
+invariant in their parameters
+and define $\Ceq{\tau}{\tau'}$ as $\Cleq{\tau}{\tau'} \wedge \Cleq{\tau'}{\tau}$.
 
-Our system is defined in \cref{infer:solving}.
-As before, we write $\CL$ for our constraint system for an arbitrary
-total bounded lattice $(\mathcal L, \lk_\Lat)$.
-Entailment is written $\entail{\inP{C}}{\inP{D}}$,
-where $D$ is a consequence of the
-constraints $C$.
-We say that $C$ and $D$ are equivalent, written as $C \equivC D$,
+Entailment is denoted by $\entail{C}{D}$,
+where $D$ is a consequence of the constraints $C$.
+We say that $C$ and $D$ are equivalent, $C \equivC D$,
 when $\entail{C}{D}$ and $\entail{D}{C}$.
+% The base rules for entailment are given in \cref{rules:entail}.
+% The rules for classic Herbrand unification and
+% projection operators are omitted for brevity (see
+% \citep{DBLP:journals/tapos/OderskySW99} for
+% details).
+%
 Following HM(X), we define a cylindric constraint system as
@@
 satisfies the axiom shown in \cref{rules:entail}.
-We define  $\Ceq{\tau}{\tau'}$ as $\Cleq{\tau}{\tau'} \wedge \Cleq{\tau'}{\tau}$.
+
+% \begin{figure*}[!bt]
+%   \centering
+%   \begin{minipage}{0.32\linewidth}
+%     \begin{align*}
+%       C &::=  C_1 \Cand C_2  \tag{Conjunction}\\
+%         &\mid \Cleq{\tau_1}{\tau_2} \mid \Cleq{k_1}{k_2} \tag{Inequalities} \\
+%         &\mid \Cproj{\tvar}{C}
+%         \mid \Cproj{\kvar}{C} \tag{Projections}
+%     \end{align*}
+%     \caption{The constraint language}
+%     \label{grammar:constraint}
+%   \end{minipage}\hfill
+%   \begin{minipage}{0.66\linewidth}
+%     \begin{mathpar}
+%       \inferrule
+%       {\forall i,\ \entail{C}{\Cleq{l_i}{k}}}
+%       {\entail{C}{\Cleq{\wedge_i l_i}{k}}}
+%       \and
+%       \inferrule
+%       {\forall i,\ \entail{C}{\Cleq{k}{l_i}}}
+%       {\entail{C}{\Cleq{k}{\vee_i l_i}}}
+%       \and
+%       \inferrule
+%       { \forall i,\ \entail{C}{\Ceq{\tau_i}{\tau_i}}\\
+%       }
+%       { \entail{C}{\Cleq{\tapp{\tcon}{(\tau_i)}}{\tapp{\tcon}{(\tau'_i)}}} }
+%       \and
+%       \inferrule{l \lk_\Lat l'}{\entail{}{\Cleq{l}{l'}}}
+%       \and
+%       \inferrule
+%       { \entail{C}{\Cleq{\tau'_1}{\tau_1}}\\
+%         \entail{C}{\Cleq{\tau_2}{\tau'_2}}\\
+%         \entail{C}{\Cleq{k}{k'}}
+%       }
+%       { \entail{C}{\Cleq{\tau_1\tarr{k}\tau_2}{\tau'_1\tarr{k'}\tau'_2}} }
+%     \end{mathpar}
+%     \caption{Base entailment rules -- $\entail{\inP{C}}{\inP{D}}$ }
+%     \label{rules:entail}
+%   \end{minipage}
+% \end{figure*}
+
+We define the set of solved formed
+$\mathcal S$ as the quotient set of $\CL$ by $\equivC$.
+We will show later that such constraints are in fact only composed of
+kind inequalities, and thus correspond to the syntactic constraints
+used in type and kind schemes.
 %
-We write $\SC$ for the set of solved forms,
-which can be used inside type and kind schemes.
-We define $\SC$ as $\CL$ quotiented by the relation $\equivC$.
-%
-The normalization function is described in \cref{infer:solving}.
+We now define our normalization procedure $\normalize{C_0}{\unif_0}$, where
+$C_0\in \CL$ is a set of constraints and $\unif_0$ is a substitution.
+It returns a constraint $C \in \mathcal S$ in
+solved form and a unifier $\unif$.
+The main idea of the algorithm is to first remove all the type equalities
+by using regular Herbrand unification. After that, we only have
+a set of inequalities among kinds, which we can consider as a relation.
+We can then saturate the relation,
+unify all kinds that are in the same equivalence classes to obtain
+a most general unifier on kind variables,
+remove all existentially quantified variables and
+then minimize back the relation and apply various
+simplification rules to make the resulting type easier to understand to users.
 
-\subsection{Example of constraint solving and simplification}
-\label{solving:example}
+More precisely, we apply the following steps:
+\begin{enumerate}
+\item Solve all type equality constraints through Herbrand unification and
+  gather all existential quantifications at the front of the constraint.
+  We obtain a constraint $C^k = \exists \Multi\kvar,\ \Cleq{k_j}{k'_j}_j$ and
+  a substitution $\unif_\tau$.
 
-\input{infer/exampleconstr}
+  We write $\mathcal R$ for the relation $\Cleq{k_j}{k'_j}_j$,
+  $\mathcal G$ the underlying directed graph and $V$ its vertices.
+
+\item Saturate the lattice equalities in $\mathcal R$.
+
+  More precisely, for each kind variable $\kvar \in V$,
+  for each constant $l_i$ (resp. $l_j$) such that
+  there is a path from $l_i$ to $\kvar$ (resp. from $\kvar$ to $l_j$) in $\mathcal G$,
+  add an edge from $\lub l_i$ to $\kvar$
+  (resp. from $\kvar$ to $\glb l_j$).
+  This step is well defined since $\mathcal L$ is a bounded lattice
+  and $\lub\emptyset$ and $\glb\emptyset$ are well defined.
+
+  We also complement $\mathcal R$ with $(\leq)$ by adding an edge
+  between related constants.
+\item
+  At this point, we can easily check for satisfiability: A constraint
+  is satisfiable (in the given environment) if and only if,
+  for any constants $l_1$ and $l_2$ such that
+  there is a path from $l_1$ to $l_2$ in $\mathcal G$, then $l_1\lk_\Lat l_2$.
+  If this is not the case, we return \textbf{fail}.
+
+\item For each strongly connected component in $\mathcal G$, unify all its vertices and replace it by a representative.
+  We write $\unif_k$ for the substitution that replaces a kind variable by
+  its representative.
+  The representative of a strongly connected component $g$ can be determined as follows:
+  \begin{itemize}
+  \item If $g$ does not contain any constant, then the representative
+    is a fresh kind variable.
+  \item If $g$ contains exactly one constant, it is the representative.
+  \item Otherwise, the initial constraint $C_0$ is not satisfiable.
+  \end{itemize}
+  Note that this step will also detect all unsatisfiable constraints.
+\item Take the transitive closure of $\mathcal R$.
+\item Remove all the vertices corresponding to the kind variables $\Multi\kvar$
+  that are existentially quantified in $C^k$.
+\item Take the transitive reduction of $\mathcal R$.
+\item Remove the extremums of $\mathcal L$ and the edges of $(\leq)$
+  from $\mathcal R$.
+\item Return $C = \left\{ k \leq k' \mid k \operatorname{\mathcal R}k' \right\}$
+  and $\unif =  \unif_\tau \meet \unif_k$.
+\end{enumerate}
+
+An example of this algorithm in action is shown in \cref{solving:example}.
+Our algorithm is complete, computes principal normal forms,
+and already simplifies constraints significantly
+(thanks to steps 6, 7 and 8).
+It can be extended with further simplification phases.
+In particular, our implementation and all the signatures presented in
+\cref{motivation} use a variance-based simplification
+where all covariant (resp. contravariant) variables are replaced by their
+lower (resp. upper) bounds.
+All the simplification mechanisms presented
+here, including the variance-based one, are complete.
+It is also possible to add ``best-effort'' simplification
+rules which help reduce the size of inferred signatures even further
+\citep{DBLP:conf/aplas/Simonet03}.
+
+% \subsection{Example of constraint solving and simplification}
+% \label{solving:example}
+
+% \input{infer/exampleconstr}
 
@@
 
-Our goal is to prove that $\CL$ supports all the properties necessary for
+We now prove that $\CL$ supports all the properties necessary for
 principal type inference, as defined by HM(X).

### 2019-11-18 kindly/infer/extra.tex
Move explanation of Let in the appendix.

@@
 The constraints rules are already shown in \cref{inference}.
-Kind inference is presented in \cref{rules:kinding:full}.
+Kind inference is presented in \cref{rules:kinding:full}
+and the detailed treatment of let-bindings in
+\cref{infer:let}.
 The type inference rules are shown in \cref{rules:typing:full}.
@@
 
+\subsection{Generalization and constraints}
+\label{infer:let}
+
+% So far, constraints have only been composed of a list of kind inequalities.
+% The internal language of constraints used for inference is in fact
+% richer.
+% To illustrate this, we show the inference rule for let-bindings
+% in \cref{rule:infer:let}.
+The {\sc Let} rule combines several ingredients previously seen:
+since let expressions are binders, we use $\Weaken$ on the bound
+identifier $x$. Since let-expressions contain
+two subexpressions, we use the environment splitting relation,
+$\bsplit{C_s}{\Sv}{\Sv_1}{(\Sv_2 \Sdel{x})}$. We remove the $x$ from
+the right environment, since it is not available outside of the expression
+$e_2$, and should not be part of the returned usage environment.
+
+As per tradition in ML languages, generalization is performed
+on let-bindings.
+Following HM(X), we write $(C_\schm, \schm) = \generalize{C}{\E}{\tau}$
+for the pair of a constraint and a scheme resulting from
+generalization. The definition is provided in \cref{rule:infer:let}.
+The type scheme $\schm$ is created by quantifying over all the appropriate
+free variables and the current constraints.
+The generated constraint $C_\schm$ uses a new projection operator,
+$\Cproj{x}{D}$ where $x$ can be either a type or a kind variable, which
+allow the creation of local variables inside the constraints.
+This allows us to encapsulate all the quantified variables in the global constraints.
+It also reflects the fact that there
+should exist at least one solution for $C$ for the scheme to be valid.
+\citet{DBLP:journals/tapos/OderskySW99} give a detailed account
+on the projection operators in HM inference.
+
 \begin{figure*}[!btp]

### 2019-11-18 kindly/infer/infer.tex
Move explanation of Let in the appendix.

@@
 
-\subsubsection{Usage environments}
+\paragraph{Usage environments}
 
@@
 
-\subsubsection{Constraint normalization}
+\paragraph{Constraint normalization}
 
@@
     \and
-    \ruleILet
-    \hfill
+    % \ruleILet
+    % \hfill
     \ruleIRegion
@@
       % \end{dcases}
-      \\\\\\
-      \generalize{C}{\E}{\tau} =
-      (\Cproj{(\Multi{\kvar},\Multi{\tvar})}{C},
-      \forall (\Multi{\kvar},\Multi{\tvar}).\qual{C}{\tau})\\
-      \text{ where}\\\\
-      \Multi{\kvar},\Multi{\tvar} = (\fv{\tau}\cup\fv{C})\setminus\fv{\E}
+      % \\\\\\
+      % \generalize{C}{\E}{\tau} =
+      % (\Cproj{(\Multi{\kvar},\Multi{\tvar})}{C},
+      % \forall (\Multi{\kvar},\Multi{\tvar}).\qual{C}{\tau})\\
+      % \text{ where}\\\\
+      % \Multi{\kvar},\Multi{\tvar} = (\fv{\tau}\cup\fv{C})\setminus\fv{\E}
     }
@@
   \vspace{-15pt}
-  \caption{Selected inference rules}
+  \caption{Selected inference rules -- $\inferW{\Sv}{(C,\unif)}{\inP{\E}}{\inP{e}}{\tau}$}
   \label{rule:infer:envs}
@@
 
-We now present the inference version
-of the {\sc Pair} and {\sc Region} rules in \cref{rule:infer:envrules}.
+The inference version
+of the {\sc Pair} and {\sc Region} rules is shown in \cref{rule:infer:envrules}.
 While the rules are quite similar to the original version, the usage
@@
 We then collect all constraints and normalize them.
-
-\subsubsection{Generalization and constraints}
-
-So far, constraints have only been composed of a list of kind inequalities.
-The internal language of constraints used for inference is in fact
-richer.
-To illustrate this, we show the inference rule for let-bindings
-in \cref{rule:infer:let}.
-This rule combines several ingredients previously seen:
-since let expressions are binders, we use $\Weaken$ on the bound
-identifier $x$. Since let-expressions contain
-two subexpressions, we use the environment splitting relation,
-$\bsplit{C_s}{\Sv}{\Sv_1}{(\Sv_2 \Sdel{x})}$. We remove the $x$ from
-the right environment, since it is not available outside of the expression
-$e_2$, and should not be part of the returned usage environment.
-
-As per tradition in ML languages, generalization is performed
-on let-bindings.
-Following HM(X), we write $(C_\schm, \schm) = \generalize{C}{\E}{\tau}$
-for the pair of a constraint and a scheme resulting from
-generalization. The definition is provided in \cref{rule:infer:let}.
-The type scheme $\schm$ is created by quantifying over all the appropriate
-free variables and the current constraints.
-The generated constraint $C_\schm$ uses a new projection operator,
-$\Cproj{x}{D}$ where $x$ can be either a type or a kind variable, which
-allow the creation of local variables inside the constraints.
-This allows us to encapsulate all the quantified variables in the global constraints.
-It also reflects the fact that there
-should exist at least one solution for $C$ for the scheme to be valid.
-\citet{DBLP:journals/tapos/OderskySW99} give a detailed account
-on the projection operators in HM inference.
 

### 2019-11-18 kindly/intro.tex
Debug lstinline's usage.

@@
 and other related works.
+
 \subsection{First examples}
 \label{sec:first-example}
+
+\lstMakeShortInline[keepspaces,basicstyle=\small\ttfamily]@
+
 \begin{figure}[tp]
-  \begin{subfigure}[t]{0.45\linewidth}
+  \begin{subfigure}[t]{0.53\linewidth}
     \lstinputlisting{code/writefiles.affe}
@@
     \label{fig:writing-files-api}
-  \end{subfigure}
-  ~
-  \begin{subfigure}[t]{0.5\linewidth}
+  \end{subfigure}~
+  \begin{subfigure}[t]{0.43\linewidth}
 \begin{lstlisting}
@@
 writing files shown in \cref{fig:writing-files-api}.  It introduces a
-linear abstract type \lstinline/File.t/. A call like
-\lstinline/fopen "foo"/ returns a linear handle to a newly created
-file, which \emph{must} be released later on with \lstinline/close/
+linear abstract type @File.t@. A call like
+@fopen "foo"@ returns a linear handle to a newly created
+file, which \emph{must} be released later on with @close@
 as shown in \cref{fig:writing-files-example}. Failing to do so is a
 static type error.  To write to the file, we must take an exclusive
-borrow \lstinline/&!h/ of the handle and pass it to the
-\lstinline/write/ function. Exclusive borrows are affine:
+borrow @&!h@ of the handle and pass it to the
+@write@ function. Exclusive borrows are affine:
 they must not be duplicated, but there is no requirement to use
-them. This affinity shows up in the annotation \lstinline/-{aff}>/ of the second arrow in
-the type of \lstinline/write/: a partial application like
-\lstinline/write &!h/ captures the affine borrow and hence the
+them. This affinity shows up in the annotation @-{aff}>@ of the second arrow in
+the type of @write@: a partial application like
+@write &!h@ captures the affine borrow and hence the
 resulting function is also affine. It
@@
 The remaining arrows in
-the API are unrestricted and we write \lstinline/->/ instead of the
-explicitly annotated \lstinline/-{un}>/.
+the API are unrestricted and we write @->@ instead of the
+explicitly annotated @-{un}>@.
 
@@
 \item resource APIs can be written in direct style as linearity is a
-  property of the type \lstinline/File.t/.
+  property of the type @File.t@.
 \end{enumerate}
@@
 The reading of the type inferred by our system is straightforward. If
-\lstinline/f/ is a \lstinline/'k/-restricted function, then so is
-\lstinline/app f/. The multiplicities of \lstinline/'a/ and
-\lstinline/'b/ play no role. As usual in ML-like languages, we
+@f@ is a @'k@-restricted function, then so is
+@app f@. The multiplicities of @'a@ and
+@'b@ play no role. As usual in ML-like languages, we
 implicitly assume prenex quantification by
@@
 \end{lstlisting}
-Like in \lstinline/app/, the multiplicities of the type variables
-\lstinline/'a,'b,'c/ do not matter. However, the multiplicity
-\lstinline/'k/ of \lstinline/f/ reappears on the second to last arrow
-because \lstinline/compose f/ is a closure that inherits
-\lstinline/f/'s multiplicity. The multiplicities of \lstinline/g/  and
-\lstinline/f/ both influence the multiplicity of the last arrow, so
+Like in @app@, the multiplicities of the type variables
+@'a,'b,'c@ do not matter. However, the multiplicity
+@'k@ of @f@ reappears on the second to last arrow
+because @compose f@ is a closure that inherits
+@f@'s multiplicity. The multiplicities of @g@  and
+@f@ both influence the multiplicity of the last arrow, so
 that we would expect its annotation to be the least upper bound
 $\kappa \sqcup \kappa_1$. Thanks to subsumption of multiplicities, it
-is sufficient to assume $\kappa \le \kappa_1$ and \lstinline/g/'s
+is sufficient to assume $\kappa \le \kappa_1$ and @g@'s
 actual multiplicity gets subsumed to $\kappa_1$. This constraint
@@
 $\kappa\le\kappa_1$ and omits the prenex quantification over $\kappa,
-\kappa_1$ as well as the kinds of \lstinline/'a,'b,'c/.
+\kappa_1$ as well as the kinds of @'a,'b,'c@.
+
+\lstDeleteShortInline@
 

### 2019-11-18 kindly/lang.tex
Use capital B for bindings.
Misc.
Debug lstinline's usage.

@@
 
-\subsubsection{Environments and bindings}
-\label{sdtyping:envs}
 \begin{figure*}[tp]
   \begin{minipage}{0.28\linewidth}
-    \begin{align*}
-      \E &::=\ \Eempty \mid \E;B\tag{Environments}\\
-      B &::=\ \bnone \tag{Empty} \\
-         &\mid \bvar{x}{\schm}\tag{Variables}\\
-         &\mid \bvar{\borrow{x}}{\borrowty{k}{\tau}}\tag{Borrows}\\
-         &\mid \svar{x}{\schm}^n \tag{Suspended}\\
-         &\mid \bvar{\tvar}{\kschm}\tag{Types}
-         % |&\ \tydecl{t}{\kschm}{K}{\tau}\tag{Type Declaration}\\
-    \end{align*}
-    \caption{Type environments}
-    \label{grammar:env}
   \end{minipage}\hfill
-  \begin{minipage}{0.7\linewidth}
+
+  % \begin{minipage}{0.7\linewidth}
     \begin{mathpar}
@@
       \and
+      \ruleSDRegion
+      \and
       \ruleSDVar
@@
       \ruleSDPair
-      \and
-      \ruleSDRegion
       \and
@@
     \label{selectrules:matching}
-  \end{minipage}
+  % \end{minipage}
 \end{figure*}
+
+\subsubsection{Environments and bindings}
+\label{sdtyping:envs}
+
+\begin{figure}[tp]
+  \begin{align*}
+    \E &::=\ \Eempty \mid \E;B\tag{Environments}\\
+    B &::=\ \bnone % \tag{Empty} \\
+    \mid \bvar{\tvar}{\kschm} % \tag{Types}
+    \mid \bvar{x}{\schm}% \tag{Variables}
+    \mid \bvar{\borrow{x}}{\borrowty{k}{\tau}}% \tag{Borrows}\\
+    \mid \svar{x}{\schm}^n % \tag{Suspended}
+    % |&\ \tydecl{t}{\kschm}{K}{\tau}\tag{Type Declaration}\\
+  \end{align*}
+  \vspace{-10pt}
+  \caption{Type environments}
+  \vspace{10pt}
+  \label{grammar:env}
+  % \begin{minipage}{0.5\linewidth}
+    \centering
+    \begin{tabular}
+      {@{}>{$}r<{$}@{ $\vdash_e$ }
+      >{$}c<{$}@{ $=$ }
+      >{$}c<{$}@{ $\ltimes$ }
+      >{$}c<{$}r}
+
+      \Cleq{\schm}{\kun_\infty}
+      &\bvar{x}{\schm}&\bvar{x}{\schm}&\bvar{x}{\schm}
+      &Both\\
+
+      {\Cempty}&
+      {\bvar{\borrow[\IBORROW]{x}}{\schm}}&
+      {\bvar{\borrow[\IBORROW]{x}}{\schm}}&{\bvar{\borrow[\IBORROW]{x}}{\schm}}
+      &Borrow
+      % \\
+
+      % {\Cempty}&
+      % \svar[\IBORROW]{x}{\schm}^n&\svar[\IBORROW]{x}{\schm}^n&\svar[\IBORROW]{x}{\schm}^n
+      % &BothS
+      \\[2mm]
+
+      {\Cempty}&{B_x}&{B_x}&{\bnone}
+      &Left\\
+      {\Cempty}&{B_x}&{\bnone}&{B_x}
+      &Right\\[2mm]
+
+      {\Cempty}&{\bvar x \schm}&{\svar x \schm^n}&{\bvar x \schm}
+      &Susp\\
+
+      {\Cempty}&
+      {\bvar{\borrow x} \schm}&{\svar[\IBORROW] x \schm^n}&{\bvar{\borrow x} \schm}
+      &SuspB\\
+
+      {\Cempty}&
+      {\svar{x} \schm}&{\svar[\IBORROW] x \schm^n}&{\svar{x} \schm}
+      &SuspS\\
+    \end{tabular}
+    \caption{Splitting rules for bindings -- $\lsplit{C}{B}{B_l}{B_r}$}
+    \label{sdtyp:split}
+  % \end{minipage}
+\end{figure}
 
@@
 \label{sdtyping:split}
-\begin{figure}[tp]
-  % \begin{minipage}{0.5\linewidth}
-    \centering
-    \begin{tabular}
-      {@{}>{$}r<{$}@{ $\vdash_e$ }
-      >{$}c<{$}@{ $=$ }
-      >{$}c<{$}@{ $\ltimes$ }
-      >{$}c<{$}r}
-
-      \Cleq{\schm}{\kun_\infty}
-      &\bvar{x}{\schm}&\bvar{x}{\schm}&\bvar{x}{\schm}
-      &Both\\
-
-      {\Cempty}&
-      {\bvar{\borrow[\IBORROW]{x}}{\schm}}&
-      {\bvar{\borrow[\IBORROW]{x}}{\schm}}&{\bvar{\borrow[\IBORROW]{x}}{\schm}}
-      &Borrow
-      % \\
-
-      % {\Cempty}&
-      % \svar[\IBORROW]{x}{\schm}^n&\svar[\IBORROW]{x}{\schm}^n&\svar[\IBORROW]{x}{\schm}^n
-      % &BothS
-      \\[2mm]
-
-      {\Cempty}&{b_x}&{b_x}&{\bnone}
-      &Left\\
-      {\Cempty}&{b_x}&{\bnone}&{b_x}
-      &Right\\[2mm]
-
-      {\Cempty}&{\bvar x \schm}&{\svar x \schm^n}&{\bvar x \schm}
-      &Susp\\
-
-      {\Cempty}&
-      {\bvar{\borrow x} \schm}&{\svar[\IBORROW] x \schm^n}&{\bvar{\borrow x} \schm}
-      &SuspB\\
-
-      {\Cempty}&
-      {\svar{x} \schm}&{\svar[\IBORROW] x \schm^n}&{\svar{x} \schm}
-      &SuspS\\
-    \end{tabular}
-    \caption{Splitting rules for bindings -- $\lsplit{C}{B}{B_l}{B_r}$}
-    \label{sdtyp:split}
-  % \end{minipage}
-\end{figure}
 
@@
 the rule {\sc Susp} splits off a borrow from  the resource
-\lstinline/a/ to use it in the left component of the pair.
+@a@ to use it in the left component of the pair.
 As usual, a borrow cannot be active in the same scope as its resource.
@@
 or a result. Its return type is a monad
-\lstinline/'a sem/ which combines error reporting and timeout:
+@'a sem@ which combines error reporting and timeout:
 \begin{lstlisting}
@@
 printing. The pervasive @let*@ operator acts as monadic bind
-for the \lstinline/sem/ monad. The operator
-\lstinline/let*? : bool -> (unit -> 'a sem) -> 'a sem/
+for the @sem@ monad. The operator
+@let*? : bool -> (unit -> 'a sem) -> 'a sem@
  converts a boolean

### 2019-11-18 kindly/motivation.tex
Debug lstinline's usage.

@@
 \label{motivation}
+
+\lstMakeShortInline[keepspaces,basicstyle=\small\ttfamily]@
 
@@
 deallocation of arrays that may contain affine elements.
-One would first use \lstinline/create n v/ to create
-an (unrestricted) array of size \lstinline/n/ initialized with value
-\lstinline/v/. This value must be unrestricted as it is duplicated to
-initialize all array elements. Using the \lstinline/map/ function we
+One would first use @create n v@ to create
+an (unrestricted) array of size @n@ initialized with value
+@v@. This value must be unrestricted as it is duplicated to
+initialize all array elements. Using the @map@ function we
 can transform the unrestricted elements into linear (affine)
-ones. This array can be processed further using \lstinline/map/ (or \lstinline/set/ if
-affine). The \lstinline/get/ function is only applicable if the element type
-is unrestricted as one element is duplicated. The \lstinline/length/
-function is always applicable. To \lstinline/free/ an array the elements must be
+ones. This array can be processed further using @map@ (or @set@ if
+affine). The @get@ function is only applicable if the element type
+is unrestricted as one element is duplicated. The @length@
+function is always applicable. To @free@ an array the elements must be
 affine.
@@
 observers, and mutators.
-Constructors and destructors like \lstinline/create/ and \lstinline/free/ manipulate the whole
+Constructors and destructors like @create@ and @free@ manipulate the whole
 array. %As with channels in \cref{sec:session-linearity},
 The constructor
-\lstinline/create/ yields a linear resource which is consumed by \lstinline/free/.
-During the lifetime of the array resource \lstinline/a/, we can split
-off \emph{shared borrows} \lstinline/&a/ that provide a read-only
-view or \emph{exclusive borrows} \lstinline/&!a/ for read-write views.
-Observer functions such as \lstinline/length/ and \lstinline/get/ expect an shared borrow
-argument, mutator functions such a \lstinline/set/ expect
+@create@ yields a linear resource which is consumed by @free@.
+During the lifetime of the array resource @a@, we can split
+off \emph{shared borrows} @&a@ that provide a read-only
+view or \emph{exclusive borrows} @&!a@ for read-write views.
+Observer functions such as @length@ and @get@ expect an shared borrow
+argument, mutator functions such a @set@ expect
 a exclusive borrow.
@@
 enclosing region. Borrows are confined to their regions. Inside the region,
-shared borrows are unrestricted (\lstinline/un/) whereas exclusive
-borrows are affine (\lstinline/aff/).
+shared borrows are unrestricted (@un@) whereas exclusive
+borrows are affine (@aff@).
 
@@
 
-\cref{ex:array:region} uses braces \lstinline/{| ... |}/ to make the
+\cref{ex:array:region} uses braces @{| ... |}@ to make the
 regions of the example in \cref{ex:array} explicit.  One
-region consists of the header expression of the \lstinline/let/ in
+region consists of the header expression of the @let@ in
 \cref{ex:array:region:get}. It is contained in another region that
-comprises the body of the \lstinline/for/ loop. \lang guarantees that borrows
+comprises the body of the @for@ loop. \lang guarantees that borrows
 never escape the smallest enclosing region. It employs a system of
-\emph{indexed kinds} like \lstinline{aff_r} and \lstinline{un_r} where
+\emph{indexed kinds} like @aff_r@ and @un_r@ where
 $r$ is a positive integer that corresponds to the lexical nesting
-depth of regions. For instance, the type of \lstinline{&!a} in
-\cref{ex:array:set} has kind \lstinline{aff_1} whereas the type of
-\lstinline{&a} in \cref{ex:array:region:get} has kind
-\lstinline{un_2} and the typing of the inner region is such that types with
+depth of regions. For instance, the type of @&!a@ in
+\cref{ex:array:set} has kind @aff_1@ whereas the type of
+@&a@ in \cref{ex:array:region:get} has kind
+@un_2@ and the typing of the inner region is such that types with
 kind indexes greater than or equal to $2$ cannot escape.
 In the example, borrows cannot escape  because they are consumed
-immediately by \lstinline/get/ and \lstinline/set/.
+immediately by @get@ and @set@.
 
@@
 
-\cref{sig:hybarray} contains an API \lstinline/HybArray/
+\cref{sig:hybarray} contains an API @HybArray@
 that enables using mutable and immutable modifications to
 the board through affine types and borrows.
-The signature differs slightly from the \lstinline/Array/
+The signature differs slightly from the @Array@
 signature. For simplicity, the array elements are unrestricted, but
-the structure itself remains linear so as to be implemented in terms of \lstinline/Array/.
-The in-place mutation function \lstinline/set_mut/ with type
-\lstinline/&!('a t) * int * 'a -> unit/ works on a exclusive borrow  whereas the persistent
-\lstinline/set/ operation has type \lstinline/&('a t) * int * 'a -> 'a t/. It
+the structure itself remains linear so as to be implemented in terms of @Array@.
+The in-place mutation function @set_mut@ with type
+@&!('a t) * int * 'a -> unit@ works on a exclusive borrow  whereas the persistent
+@set@ operation has type @&('a t) * int * 'a -> 'a t@. It
 takes a shared borrow because it only reads from the
@@
 The code in \cref{ex:cow} contains a very simple implementation of
-\lstinline/HybArray/ that represents hybrid arrays
+@HybArray@ that represents hybrid arrays
 as regular arrays and uses copy-on-write for persistent
 modifications. The function
-\lstinline/mapi : (int -> &'a -> 'b) -> &('a t) -> 'b t /
-is a simple variation on \lstinline/Array.map/ where the mapping function
+@mapi: (int -> &'a -> 'b) -> &('a t) -> 'b t@
+is a simple variation on @Array.map@ where the mapping function
 takes the position of the element.
@@
 Our implementation of a Sudoku solver (\cref{ex:sudoku}) performs
-modifications that correspond to choice points using \lstinline/set/,
+modifications that correspond to choice points using @set@,
 which makes it trivial to come back to the previous version
-of the array, while other modifications use \lstinline/set_mut/, which
+of the array, while other modifications use @set_mut@, which
 forbids uses of the previous version.
@@
 The board is represented as a 2D-matrix (\cref{line:boardtype}), where
-the \lstinline/Matrix/ type uses the same API as \lstinline/HybArray/
+the @Matrix@ type uses the same API as @HybArray@
 but with two indices.
@@
 change the cell with an immutable modification (\cref{line:immutset}), and propagate
-the changes with the \lstinline/propagate/ function.
-The \lstinline/propagate/ function uses direct mutation through a
+the changes with the @propagate@ function.
+The @propagate@ function uses direct mutation through a
 exclusive borrow of the matrix as it need not preserve the previous
 version of the board.
-The implementation of \lstinline/propagate/ is split into three parts
+The implementation of @propagate@ is split into three parts
 for lines, columns, and square, which are all very similar to function
-\lstinline/propagate_lines/ (\cref{line:propline}).
+@propagate_lines@ (\cref{line:propline}).
 
 The typing ensures that the mutations do not compromise the state at
-the choice point, because they operate on a new state \lstinline/new_g/ created for one
+the choice point, because they operate on a new state @new_g@ created for one
 particular branch of the choice.
-As the \lstinline/set/ function only requires an unrestricted shared borrow,
-the closure \lstinline/try_solution/ remains unrestricted even though
-it captures the borrow \lstinline/&g/. 
-The price is that \lstinline/try_solution/ cannot escape from
-\lstinline/&g/'s region. In this example, inference places the region
-around the \lstinline/begin//\lstinline/end/ scope.
-Hence, \lstinline/try_solution/ can be used in
+As the @set@ function only requires an unrestricted shared borrow,
+the closure @try_solution@ remains unrestricted even though
+it captures the borrow @&g@. 
+The price is that @try_solution@ cannot escape from
+@&g@'s region. In this example, inference places the region
+around the @begin@/@end@ scope.
+Hence, @try_solution@ can be used in
 the iteration in \cref{line:iter}.
-% As \lstinline/&g/ is unrestricted, we could even use a parallel iteration
+% As @&g@ is unrestricted, we could even use a parallel iteration
 % instead of a sequential one.
-As \lstinline/g/ is linear we must free it outside of the region
+As @g@ is linear we must free it outside of the region
 before returning (\cref{line:free:g}).
@@
 % the implementation of functions that are nevertheless safe. Rust addresses this
-% issue through the \lstinline/unsafe/ construct, which allows to locally
+% issue through the @unsafe@ construct, which allows to locally
 % ignore linearity constraints. This is essential to build basic
@@
 
-\cref{intf:pool} contains the interface of the \lstinline/Pool/ module.
+\cref{intf:pool} contains the interface of the @Pool@ module.
 A pool is parameterized by its content. The kind of the pool
@@
 a linear pool while unrestricted content yields an unrestricted pool.
-The functions \lstinline/Pool.create/ and \lstinline/Pool.consume/
+The functions @Pool.create@ and @Pool.consume@
 build/destroy a pool given creators/destructors for the elements
 of the pool.
-The function \lstinline/Pool.use/ is the workhorse of the API, which
+The function @Pool.use@ is the workhorse of the API, which
 borrows a resource from the pool to a callback.
@@
 This encapsulation is implemented with a universally quantified \emph{kind index variable} $r$.
-The signature prescribes the type \lstinline/&!(aff_r1,'a_1 )/ for the
+The signature prescribes the type @&!(aff_r1,'a_1)@ for the
 exclusive borrow of the resource with an affine kind at region nesting $r+1$. The return
-type of the callback is constrained to kind \lstinline/'k_2 <= aff_r/
+type of the callback is constrained to kind @'k_2 <= aff_r@
 so that the callback certainly cannot return the borrowed argument.
-In a specific use of \lstinline/Pool.use/, the index $r$ gets unified
+In a specific use of @Pool.use@, the index $r$ gets unified
 with the current nesting level of regions so that the region for the
@@
 
-\cref{ex:pool} shows a simple example of using the \lstinline/Pool/ module.
+\cref{ex:pool} shows a simple example of using the @Pool@ module.
 
@@
 using a concurrent queue with atomic add and remove operations.
-The implementation of the \lstinline/Pool.create/ and \lstinline/Pool.consume/
+The implementation of the @Pool.create@ and @Pool.consume@
 functions is straightforward.
-The function \lstinline/Pool.use/ first draws
+The function @Pool.use@ first draws
 an element from the pool (or creates a fresh element),
-passes it down to the callback function \lstinline/f/, and returns
+passes it down to the callback function @f@, and returns
 it to the pool afterwards.
@@
 we explicitly delimit the region in \cref{line:pool:region} to ensure that
-the return value of \lstinline/f &!o/ does not capture \lstinline/&!o/.
+the return value of @f &!o@ does not capture @&!o@.
 In practice, the type checker inserts this region automatically.
@@
 
+\lstDeleteShortInline@
+
 %%% Local Variables:

### 2019-11-18 kindly/intro.tex
Add link to the prototype.

@@
 be used for reading or writing, anymore.
-
+%
 Dynamic allocation of memory is analogous. An API call returns a
@@
 \item Automatic inference of regions for borrows.
-\item An implementation of the type inference algorithm including all
-  constraint simplification and extended with pattern matching. A
-  web implementation of the typechecker is available, its URL is given in the
-  non-anonymous supplementary material, and we plan to submit it for
-  artifact evaluation.
+\item A prototype implementation of the type inference algorithm, including all
+  constraint simplification and extended with pattern matching,
+  available at \url{https://affe.netlify.com/}.
 \end{itemize}

### 2019-11-18 kindly/motivation.tex
Misc.

@@
 
-We now showcase an API for mutable linear arrays
-(\cref{sig:array}).
-The intention of this API is to safely handle manual allocation and
+The API for mutable linear arrays (shown in \cref{sig:array})
+aims to safely handle manual allocation and
 deallocation of arrays that may contain affine elements.
@@
 callback effectively gets ``inserted'' into the lexical nesting at the callsite.
-
-\cref{ex:pool} shows a simple example of using the @Pool@ module.
+%
+\cref{ex:pool} shows a simple example using the @Pool@ module.
 

### 2019-11-18 kindly/lang.tex
Misc.
Explain the resource operation in syntax and typing.

@@
   the borrow of a borrow).
-\item Regions are assumed to be explicit in the code. (But we present
-  an algorithm to infer them in \cref{regionannot}.)
+\item Regions are explicit in the code and must be annotated using the
+  algorithms presented in \cref{regionannot}.
 \item Regions are identified using nesting levels instead of region
@@
 
-We start from the bottom.
 A borrow $\borrow{x}$ is always taken from a variable $x$. The
@@
 whether the match should work on borrows ($\etransfm=\&^b$) or not ($\etransfm=id$).
+%
+We consider four primitive operations to manipulate resources:
+$\create$, $\observe$, $\update$ and $\destroy$.
 
@@
 
+\subsubsection{Resource management}
+
+Our system would be incomplete without a way to actually manipulate resources.
+We consider the simple case of an abstract resource
+type $\tapp{\tres}{\tau}$ whose content of type $\tau$ must be $\kun_0$ and
+equipped with four operations:
+\begin{itemize}
+\item 
+$\create$:
+$\forall{\kvar_\tvar}\bvar{\tvar}{\kvar_\tvar}.\ 
+\qual{\Cleq \kvar {\kun_0}}
+{\tvar \tarr{} \tapp\tres\tvar}$\\
+Creates a new resource.
+
+\item
+$\observe$:
+$\forall{\kvar\kvar_\tvar}\bvar{\tvar}{\kvar_\tvar}.\ 
+\qual{\Cleq {\kvar_\tvar} {\kun_0}}
+{\borrowty[\IBORROW]{\kvar}{\tapp\tres\tvar} \tarr{} \tvar}$\\
+Reads of a shared borrow of a resource.
+
+\item
+$\update$:
+$\forall{\kvar\kvar_\tvar}\bvar{\tvar}{\kvar_\tvar}.\ 
+\qual{\Cleq {\kvar_\tvar} {\kun_0}}
+\borrowty[\MBORROW]{\kvar}{\tapp\tres\tvar} \tarr{} \tvar \tarr{\kaff} \tunit$\\
+Writes in an exclusive borrow of a resource.
+
+\item
+$\destroy$:
+$\forall{\kvar\kvar_\tvar}\bvar{\tvar}{\kvar_\tvar}.\ 
+\qual{\Cleq {\kvar_\tvar} {\kun_0}}
+\tapp\tres\tvar \tarr{} \tunit$\\
+Disposes of a resource.
+\end{itemize}
+
 % \clearpage

### 2019-11-19 kindly/motivation-pool.tex
Replace the pool example by an iterator one. Should be shorter.

--- 2019-11-18 kindly/motivation.tex
@@
-\section{Linearity, Affinity, and Borrows at Work}
-\label{motivation}
-
-\lstMakeShortInline[keepspaces,basicstyle=\small\ttfamily]@
-
-\lang{} is amenable to a functional resource-passing style
-common in functional encodings of session types (e.g.,
-\cite{DBLP:journals/jfp/Padovani17}; see also
-\cref{sec:session-linearity} in the supplement). But it really shines 
-when manipulating mutable resources like buffers or connection pools
-using a mixed functional and imperative programming style.
-%
-To support this usage pattern of linearity, we rely on
-the notion of borrowing \cite{DBLP:conf/popl/BoylandR05} which
-we demonstrate through a series of examples:
-\begin{itemize}
-\item An API for linear arrays to demonstrate simple borrowing and imperative programming.
-\item
-  A backtracking Sudoku solver to demonstrate
-  \lang's ability to build safe hybrid functional/imperative data-structures.
-\item A pool of linear objects to demonstrate advanced use of regions.
-\end{itemize}
-
-\subsection{Imperative programming with linear arrays}
-\label{sec:imper-progr}
-
-The API for mutable linear arrays (shown in \cref{sig:array})
-aims to safely handle manual allocation and
-deallocation of arrays that may contain affine elements.
-One would first use @create n v@ to create
-an (unrestricted) array of size @n@ initialized with value
-@v@. This value must be unrestricted as it is duplicated to
-initialize all array elements. Using the @map@ function we
-can transform the unrestricted elements into linear (affine)
-ones. This array can be processed further using @map@ (or @set@ if
-affine). The @get@ function is only applicable if the element type
-is unrestricted as one element is duplicated. The @length@
-function is always applicable. To @free@ an array the elements must be
-affine.
-
-To manage the different kinds of accessing the array we distinguish between constructors, destructors,
-observers, and mutators.
-Constructors and destructors like @create@ and @free@ manipulate the whole
-array. %As with channels in \cref{sec:session-linearity},
-The constructor
-@create@ yields a linear resource which is consumed by @free@.
-During the lifetime of the array resource @a@, we can split
-off \emph{shared borrows} @&a@ that provide a read-only
-view or \emph{exclusive borrows} @&!a@ for read-write views.
-Observer functions such as @length@ and @get@ expect an shared borrow
-argument, mutator functions such a @set@ expect
-a exclusive borrow.
-
-Each borrow is tied to a region whose lifetime is properly
-contained in the lifetime of the resource.
-In a region, we can split off as many shared borrows of a resource as
-we like, but we can take only one exclusive borrow. In a 
-subsidiary region, we can take shared borrows of any
-borrow or we can take an exclusive borrow of an exclusive borrow from an
-enclosing region. Borrows are confined to their regions. Inside the region,
-shared borrows are unrestricted (@un@) whereas exclusive
-borrows are affine (@aff@).
-
-The code in \cref{ex:array} demonstrates a use of the API to create an
-array of Fibonacci numbers as the programmer would write it. After
-creation of the array, the presence of a borrow in the for loop
-prevents access to the ``raw'' resource inside the loop's
-body. \cref{ex:array:get} of the example contains two shared borrows
-in the same expression, which forms a region by itself. These borrows
-are split off the exclusive borrow used in \cref{ex:array:set}, which
-belongs to the next enclosing region corresponding to the loop body.
-The whole array can be returned in \cref{ex:array:return} because  the
-borrows are no longer in scope.
-
-\begin{figure}[tp]
-  \centering
-  \begin{subfigure}[t]{1\linewidth}
-\begin{lstlisting}
-module Array : sig
-  type ('a : 'k) t : lin
-  val create : ('a : un) => int -> 'a -> 'a t
-  val free : ('a : aff) => 'a t -> unit
-  val length : &('a t) -> int
-  val get : ('a : un) => &('a t) * int -> 'a
-  val set : ('a : aff) => &!('a t) * int * 'a -> unit
-  val map : (&'a -> 'b) -> &('a t) -> 'b t
-end
-\end{lstlisting}
-    \vspace{-15pt}
-    \caption{Signature}
-    \label{sig:array}
-  \end{subfigure}
-
-  \begin{subfigure}[t]{1\linewidth}
-\begin{lstlisting}
-let mk_fib_array n =
-  let a = create n 1 in
-  for i = 2 to n - 1 do
-    let x = get (&a, i-1) + get (&a, i-2) in(*@\label{ex:array:get}*)
-    set (&!a, i, x)(*@\label{ex:array:set}*)
-  done;
-  a(*@\label{ex:array:return}*)
-# mk_fib_array : int -> int Array.t
-\end{lstlisting}
-    \vspace{-10pt}
-    \caption{Example of use}
-    \label{ex:array}
-\begin{lstlisting}[firstnumber=2]
-  let a = create n 1 in
-  for i = 2 to n - 1 do {|
-    let x = {| get &a (i-1) + get &a (i-2) |} in(*@\label{ex:array:region:get}*)
-    set &!a i x
-  |} done;
-\end{lstlisting}
-    \vspace{-10pt}
-    \caption{Excerpt of \cref{ex:array} with explicit regions}
-    \label{ex:array:region}
-  \end{subfigure}
-  \vspace{-5pt}
-  \caption{Linear arrays}
-  \label{ex:array}
-\end{figure}
-
-\cref{ex:array:region} uses braces @{| ... |}@ to make the
-regions of the example in \cref{ex:array} explicit.  One
-region consists of the header expression of the @let@ in
-\cref{ex:array:region:get}. It is contained in another region that
-comprises the body of the @for@ loop. \lang guarantees that borrows
-never escape the smallest enclosing region. It employs a system of
-\emph{indexed kinds} like @aff_r@ and @un_r@ where
-$r$ is a positive integer that corresponds to the lexical nesting
-depth of regions. For instance, the type of @&!a@ in
-\cref{ex:array:set} has kind @aff_1@ whereas the type of
-@&a@ in \cref{ex:array:region:get} has kind
-@un_2@ and the typing of the inner region is such that types with
-kind indexes greater than or equal to $2$ cannot escape.
-In the example, borrows cannot escape  because they are consumed
-immediately by @get@ and @set@.
-
-Region annotations are automatically inserted using
-syntactic cues such as borrows and binders (see
-\cref{regionannot}).  Users may manually insert regions to further restrict the range of a
-borrow. In practice, region annotations are
-rarely needed with common programming idioms.
-% All regions for the examples in this section have been inferred.
-
-% \begin{figure}
-%   \centering
-%   % \begin{subfigure}{0.6\linewidth}
-% \begin{lstlisting}
-% let mk_fib_array n =
-%   let a = create n 1 in
-%   for i = 2 to n - 1 do {|
-%     let x = {| get &a (i-1) + get &a (i-2) |} in(*@\label{ex:array:region:get}*)
-%     set &!a i x
-%   |}
-%   done;
-%   a(*@\label{ex:array:return}*)
-% # mk_fib_array : int -> int Array.t
-% \end{lstlisting}
-%     \caption{Example from \cref{ex:array} with explicit regions}
-%     \label{ex:array:region}
-%   % \end{subfigure}
-% \end{figure}
-
-\subsection{Solving sudokus with hybrid data-structures}
-
-Recently introduced persistent data structures
-permit transient mutations where
-non-linear uses lead to degraded performance
-\cite{DBLP:conf/ml/ConchonF07} or to
-dynamic and static checks \cite{DBLP:journals/pacmpl/Puente17}.
-In particular, persistent Hash-Array-Mapped-Tries (HAMT) have been used with similar
-APIs in several non-pure functional languages (OCaml, Clojure, \dots).
-Affine types help formalize the performance contract between the programmer
-and the library, while borrows avoid the need to thread state explicitly,
-as usually required by an API for immutable data types.
-%
-
-In this section, we present
-a safe API for persistent arrays that support both immutable and mutable features,
-and use it to implement a backtracking Sudoku solver.
-% show how \lang allows us to define
-The Sudoku solver maintains an array to represent the state of
-the game and uses backtracking when there are several choices to proceed.
-The use of backtracking suggests a persistent data structure for the array.
-However, only changes that correspond to a choice point need to use
-the persistence mechanism, others may be implemented as
-cheap in-place mutations.
-
-\cref{sig:hybarray} contains an API @HybArray@
-that enables using mutable and immutable modifications to
-the board through affine types and borrows.
-The signature differs slightly from the @Array@
-signature. For simplicity, the array elements are unrestricted, but
-the structure itself remains linear so as to be implemented in terms of @Array@.
-The in-place mutation function @set_mut@ with type
-@&!('a t) * int * 'a -> unit@ works on a exclusive borrow  whereas the persistent
-@set@ operation has type @&('a t) * int * 'a -> 'a t@. It
-takes a shared borrow because it only reads from the
-argument array, but returns a fresh,  modified structure.
-The code in \cref{ex:cow} contains a very simple implementation of
-@HybArray@ that represents hybrid arrays
-as regular arrays and uses copy-on-write for persistent
-modifications. The function
-@mapi: (int -> &'a -> 'b) -> &('a t) -> 'b t@
-is a simple variation on @Array.map@ where the mapping function
-takes the position of the element.
-% A more elaborate implementation could use persistent arrays
-% \cite{DBLP:conf/ml/ConchonF07}.
-
-\begin{figure}[tp]
-  \centering
-  \begin{subfigure}{1\linewidth}
-\begin{lstlisting}
-module type HybArray = sig
-  type ('a : un) t : lin
-  val create : int -> 'a -> 'a t
-  val free : 'a t -> unit
-  val length : &('a t) -> int
-  val get : &('a t) * int -> 'a
-  val set : &('a t) * int * 'a -> 'a t
-  val set_mut : &!('a t) * int * 'a -> unit
-end
-\end{lstlisting}
-    \vspace{-15pt}
-    \caption{Signature}
-    \label{sig:hybarray}
-  \end{subfigure}
-
-  \begin{subfigure}{1\linewidth}
-\begin{lstlisting}
-module Cow : HybArray = struct
-  include Array
-  let set (a, i0, x0) =
-    Array.mapi (fun i x -> if i = i0 then x0 else x) a
-  let set_mut = Array.set
-end
-\end{lstlisting}
-    \vspace{-15pt}
-    \caption{Implementation using copy-on-write}
-    \label{ex:cow}
-  \end{subfigure}
-  \caption{Hybrid arrays}
-\end{figure}
-
-Our implementation of a Sudoku solver (\cref{ex:sudoku}) performs
-modifications that correspond to choice points using @set@,
-which makes it trivial to come back to the previous version
-of the array, while other modifications use @set_mut@, which
-forbids uses of the previous version.
-
-The board is represented as a 2D-matrix (\cref{line:boardtype}), where
-the @Matrix@ type uses the same API as @HybArray@
-but with two indices.
-Each cell contains an integer set that represent admissible solutions so far.
-The Sudoku solver iterates over the cells and tries each possible solution (\cref{line:try_solution}).
-When a value is picked for the current cell, we create a choice point,
-change the cell with an immutable modification (\cref{line:immutset}), and propagate
-the changes with the @propagate@ function.
-The @propagate@ function uses direct mutation through a
-exclusive borrow of the matrix as it need not preserve the previous
-version of the board.
-The implementation of @propagate@ is split into three parts
-for lines, columns, and square, which are all very similar to function
-@propagate_lines@ (\cref{line:propline}).
-
-The typing ensures that the mutations do not compromise the state at
-the choice point, because they operate on a new state @new_g@ created for one
-particular branch of the choice.
-As the @set@ function only requires an unrestricted shared borrow,
-the closure @try_solution@ remains unrestricted even though
-it captures the borrow @&g@. 
-The price is that @try_solution@ cannot escape from
-@&g@'s region. In this example, inference places the region
-around the @begin@/@end@ scope.
-Hence, @try_solution@ can be used in
-the iteration in \cref{line:iter}.
-% As @&g@ is unrestricted, we could even use a parallel iteration
-% instead of a sequential one.
-As @g@ is linear we must free it outside of the region
-before returning (\cref{line:free:g}).
-
-While presented for copy-on-write arrays, the API
-can easily be adapted to other persistent data structures with
-transient mutability such as Relaxed-Radix Balance Vectors (RRB) \cite{DBLP:journals/pacmpl/Puente17}
-or persistent HAMTs \cite{bagwell2001ideal,clojurehamt} to provide  a
-convenient programming style without compromising performance.
-
-\begin{figure}
-  \centering
-  \lstinputlisting[linerange=1-31]{code/sukodu.affe}
-    \vspace{-15pt}
-  \caption{Excerpt of the Sudoku solver}
-  \label{ex:sudoku}
-\end{figure}
-
-% \subsection{Type abstraction and linearity violations}
-
-% While linearity allows to verify many properties, it also prevents
-% the implementation of functions that are nevertheless safe. Rust addresses this
-% issue through the @unsafe@ construct, which allows to locally
-% ignore linearity constraints. This is essential to build basic
-% functions such as splitting strings.
-% %
-% \lang allows to partially emulate this behavior using type abstraction
-% as provided by ML modules.
-% In particular, data types can be internally defined as unrestricted but exposed
-% as affine.
-
-% \TODO{Show example with arrays}
-
-\subsection{Local data-structures with regions}
+\subsection{Pool of linear resources}
 \label{tuto:pool}
 
-% We have seen that borrows are limited to a region.
-In the examples so far, regions do not appear in type signatures.
-But for certain programming idioms, we want to extend the scope of a
-region across a function boundary. As an example,
-we present an interface and implementation of a pool of linear resources where the
+We present an interface and implementation of a pool of linear resources where the
 extended scope of the region enforces proper use of the resources.
@@
 
-\lstDeleteShortInline@
-
 %%% Local Variables:

### 2019-11-19 kindly/motivation.tex
Replace the pool example by an iterator one. Should be shorter.

@@
 \begin{lstlisting}
-module Array : sig
+module type ARRAY = sig
   type ('a : 'k) t : lin
@@
 end
+
+module Array : ARRAY
 \end{lstlisting}
@@
   \centering
-  \begin{subfigure}{1\linewidth}
+  % \begin{subfigure}{1\linewidth}
 \begin{lstlisting}
 module type HybArray = sig
-  type ('a : un) t : lin
-  val create : int -> 'a -> 'a t
-  val free : 'a t -> unit
-  val length : &('a t) -> int
-  val get : &('a t) * int -> 'a
+  include ARRAY
   val set : &('a t) * int * 'a -> 'a t
@@
 end
-\end{lstlisting}
-    \vspace{-15pt}
-    \caption{Signature}
-    \label{sig:hybarray}
-  \end{subfigure}
 
-  \begin{subfigure}{1\linewidth}
-\begin{lstlisting}
 module Cow : HybArray = struct
@@
     \vspace{-15pt}
-    \caption{Implementation using copy-on-write}
+  %   \caption{Implementation using copy-on-write}
+  % \end{subfigure}
+  \caption{Hybrid arrays}
+    \label{sig:hybarray}
     \label{ex:cow}
-  \end{subfigure}
-  \caption{Hybrid arrays}
+
+  \lstinputlisting[linerange=1-31]{code/sukodu.affe}
+    \vspace{-15pt}
+  \caption{Excerpt of the Sudoku solver}
+  \label{ex:sudoku}
 \end{figure}
@@
 
-\begin{figure}
-  \centering
-  \lstinputlisting[linerange=1-31]{code/sukodu.affe}
-    \vspace{-15pt}
-  \caption{Excerpt of the Sudoku solver}
-  \label{ex:sudoku}
-\end{figure}
-
 % \subsection{Type abstraction and linearity violations}
@@
 
-\subsection{Local data-structures with regions}
-\label{tuto:pool}
+\subsection{Iterators and regions}
+\label{tuto:vector}
 
@@
 But for certain programming idioms, we want to extend the scope of a
-region across a function boundary. As an example,
-we present an interface and implementation of a pool of linear resources where the
-extended scope of the region enforces proper use of the resources.
+region across a function boundary.
+For instance, how should we fold on an array of linear objects?
+A naive fold function would only work on unrestricted content:
+\begin{lstlisting}
+val fold : ('a : un) => ('a -> 'b -> 'b) -> 'a array -> 'b -> 'b
+\end{lstlisting}
 
-\cref{intf:pool} contains the interface of the @Pool@ module.
-A pool is parameterized by its content. The kind of the pool
-depends on the content: linear content implies
-a linear pool while unrestricted content yields an unrestricted pool.
-The functions @Pool.create@ and @Pool.consume@
-build/destroy a pool given creators/destructors for the elements
-of the pool.
-The function @Pool.use@ is the workhorse of the API, which
-borrows a resource from the pool to a callback.
-It takes a shared borrow of a pool (to enable concurrent access) and a
-callback function.
-The callback receives a exclusive borrow of an arbitrary resource from the pool.
+However, if we want to work on linear and affine resources,
+we must make sure those resources are not leaked in the result.
+content. We obtain the following signature:
+
+\begin{lstlisting}
+val fold :
+  ('b:'k),('k <= aff_r) =>
+  (&(aff_r1,'a) -> 'b -{aff_r1}> 'b) -> &('a array) -> 'b -{lin}> 'b
+\end{lstlisting}
+
+The folded function receives a shared borrow of the element in the array.
 The typing of the callback ensures
 that this borrow is neither captured nor returned by the function.
+%
+This encapsulation is implemented with a universally quantified \emph{kind index variable} $r$.
+The signature prescribes the type @&!(aff_r1,'a)@ for the
+exclusive borrow of the resource with an affine kind at region nesting $r+1$. The return
+type of the callback is constrained to kind @'k <= aff_r@
+so that the callback certainly cannot return the borrowed argument.
+%
+The input of the fold is a shared borrow of the array,
+which ensures that we have the right to share borrows of the inner content and
+make multiple concurrent folds.
 
-This encapsulation is implemented with a universally quantified \emph{kind index variable} $r$.
-The signature prescribes the type @&!(aff_r1,'a_1)@ for the
-exclusive borrow of the resource with an affine kind at region nesting $r+1$. The return
-type of the callback is constrained to kind @'k_2 <= aff_r@
-so that the callback certainly cannot return the borrowed argument.
-In a specific use of @Pool.use@, the index $r$ gets unified
-with the current nesting level of regions so that the region for the
-callback effectively gets ``inserted'' into the lexical nesting at the callsite.
-%
-\cref{ex:pool} shows a simple example using the @Pool@ module.
-
-The implementation in \cref{impl:pool} represents a bag of resources
-using a concurrent queue with atomic add and remove operations.
-The implementation of the @Pool.create@ and @Pool.consume@
-functions is straightforward.
-The function @Pool.use@ first draws
-an element from the pool (or creates a fresh element),
-passes it down to the callback function @f@, and returns
-it to the pool afterwards.
-For clarity,
-we explicitly delimit the region in \cref{line:pool:region} to ensure that
-the return value of @f &!o@ does not capture @&!o@.
-In practice, the type checker inserts this region automatically.
-
-\begin{figure}[tp]
-  \centering
-  \begin{subfigure}[t]{1\linewidth}
-\begin{lstlisting}
-type ('a:'k) pool : 'k
-create : (unit -> 'a) -> 'a pool
-consume : ('a -> unit) -> 'a pool -> unit
-use : ('a_1:'k_1),('a_2:'k_2),('k_2 <= aff_r) =>
-  &('a_1 pool) -> (&!(aff_r1,'a_1) -{lin}> 'a_2) -{'k_1}> 'a_2
-\end{lstlisting}
-    \vspace{-10pt}
-    \caption{Signature}
-    \label{intf:pool}
+We can then easily use this function to fold over an array of files. Here
+we compute the sum of their lengths:
 
 \begin{lstlisting}
-(* Using the pool in queries. *)
-let create_user pool name =
-  Pool.use &pool (fun connection ->
-    Db.insert "users" [("name", name)] connection)
+let all_lengths = fold (fun f s -> read f + s) &all_files 0 
+\end{lstlisting}
 
-let uri = "postgresql://localhost:5432"
-let main users =
-  (* Create a database connection pool. *)
-  let pool = Pool.create (fun _ -> Db.connect uri) in
-  List.parallel_iter (create_user &pool) users;
-  Pool.consume (fun c -> Db.close c)
+Naturally, this is not sufficient if we want to mutate the array during iteration.
+For this purpose, we need to take an exclusive borrow of the structure to iterate
+on:
+
+\begin{lstlisting}
+val iter_mut : &!('a array) -> (&!(aff_r1,'a) -{un_r1}> unit) -{lin}> unit
 \end{lstlisting}
-    \vspace{-10pt}
-    \caption{Example of use}
-    \label{ex:pool}
-  \end{subfigure}
 
-  \begin{subfigure}[t]{1\linewidth}
-\begin{lstlisting}
-type ('a:'k) pool : 'k = {spawn: unit -> 'a; queue: 'a CQueue.t}
-let create spawn = { spawn ; queue = CQueue.create () }
-let consume f c = CQueue.iter f c.queue
-let use { spawn ; queue } f =
-  let o = match CQueue.pop &queue with
-    | Some x -> x
-    | None () -> spawn ()
-  in
-  let r = {| f &!o |} in(*@\label{line:pool:region}*)
-  Queue.push o &queue;
-  r
-\end{lstlisting}
-    \vspace{-15pt}
-    \caption{Implementation}
-    \label{impl:pool}
-  \end{subfigure}
+Here, the callback of the @iter@ function must be of kind @un_r1@: Indeed, this
+callback should be @un@, as it will be called multiple time, but it should also
+be prevented from capturing objects of the region $r+1$ to prevent the borrow
+from escaping.
 
-  \caption{The \texttt{Pool} module}
-  \label{fig:pool}
-\end{figure}
+While the distinction between mutable and immutable iteration function
+might seem unfortunate, it is typical of
+programming with borrows and is also present in the Rust standard library, for
+instance. It allows the programmer to explicitly state how various iterations
+are allowed to compose and be optimized.
+\lang's region variables ensures that the content iterated on can never
+be leaked outside of the iteration function.
+This pattern is essential in many use-cases of linearity such
+as pools of linear objects (see \cref{tuto:pool}).
 

### 2019-11-20 kindly/motivation.tex
Fix unsound typos.

@@
   ('b:'k),('k <= aff_r) =>
-  (&(aff_r1,'a) -> 'b -{aff_r1}> 'b) -> &('a array) -> 'b -{lin}> 'b
+  (&(aff_r1,'a) -> 'b -{aff_r}> 'b) -> &('a array) -> 'b -{lin}> 'b
 \end{lstlisting}
@@
 \begin{lstlisting}
-val iter_mut : &!('a array) -> (&!(aff_r1,'a) -{un_r1}> unit) -{lin}> unit
+val iter_mut : &!('a array) -> (&!(aff_r1,'a) -{un_r}> unit) -{lin}> unit
 \end{lstlisting}
 
-Here, the callback of the @iter@ function must be of kind @un_r1@: Indeed, this
+Here, the callback of the @iter@ function must be of kind @un_r@: Indeed, this
 callback should be @un@, as it will be called multiple time, but it should also

### 2019-11-20 kindly/infer/exampleconstr.tex
Small fixes in the example.

@@
 \item From the constraints above, we deduce the graph shown
-  with plain arrows on the left
+  with plain arrows on the left of \cref{example:graph}.
 \item We add all the dashed arrows by saturating
@@
   {\color{green} green}.
-  We deduce $\kvar_\gamma = \kvar_x = \kun$
+  We deduce $\kvar_\gamma = \kvar_x = \kun$.
 \item We take the transitive closure, which adds the
-  arrow in {\color{blue} blue} from $\kvar_1$ to $\kvar_3$.
+  arrow in {\color{blue} blue} from $\kvar_3$ to $\kvar_1$.
 \item We remove the remaining nodes not present in the type skeleton (colored in {\color{red} red}): $\kvar_r$ and $\kvar_f$.

### 2019-11-21 kindly/motivation.tex
Misc fixes for the iterator stuff.

@@
   ('b:'k),('k <= aff_r) =>
-  (&(aff_r1,'a) -> 'b -{aff_r}> 'b) -> &('a array) -> 'b -{lin}> 'b
+  (&(aff_r1,'a) -> 'b -{aff_r1}> 'b) -> &('k_1,'a array) -> 'b -{'k_1}> 'b
 \end{lstlisting}
@@
 This encapsulation is implemented with a universally quantified \emph{kind index variable} $r$.
-The signature prescribes the type @&!(aff_r1,'a)@ for the
-exclusive borrow of the resource with an affine kind at region nesting $r+1$. The return
+The signature prescribes the type @&(aff_r1,'a)@ for the
+shared borrow of the resource with an affine kind at region nesting $r+1$. The return
 type of the callback is constrained to kind @'k <= aff_r@
@@
 We can then easily use this function to fold over an array of files. Here
-we compute the sum of their lengths:
+we compute the sum of their sizes:
 
 \begin{lstlisting}
-let all_lengths = fold (fun f s -> read f + s) &all_files 0 
+let total_size = fold (fun f s -> File.size f + s) &all_files 0 
 \end{lstlisting}
@@
 \begin{lstlisting}
-val iter_mut : &!('a array) -> (&!(aff_r1,'a) -{un_r}> unit) -{lin}> unit
+val iter_mut : &!('a array) -> (&!(aff_r1,'a) -{un}> unit) -{aff}> unit
 \end{lstlisting}
 
-Here, the callback of the @iter@ function must be of kind @un_r@: Indeed, this
-callback should be @un@, as it will be called multiple time, but it should also
-be prevented from capturing objects of the region $r+1$ to prevent the borrow
-from escaping.
+Since the callback will be called multiple times, it is explicitly
+annotated with @un@.
 
@@
 might seem unfortunate, it is typical of
-programming with borrows and is also present in the Rust standard library, for
-instance. It allows the programmer to explicitly state how various iterations
+programming with borrows and is also present in the Rust standard library.
+It allows the programmer to explicitly state how various iterations
 are allowed to compose and be optimized.

### 2019-11-21 kindly/motivation-array.tex
Rewrite the intro example.

--- 2019-11-21 kindly/motivation.tex
@@
-\section{Linearity, Affinity, and Borrows at Work}
-\label{motivation}
-
-\lstMakeShortInline[keepspaces,basicstyle=\small\ttfamily]@
-
-\lang{} is amenable to a functional resource-passing style
-common in functional encodings of session types (e.g.,
-\cite{DBLP:journals/jfp/Padovani17}; see also
-\cref{sec:session-linearity} in the supplement). But it really shines 
-when manipulating mutable resources like buffers or connection pools
-using a mixed functional and imperative programming style.
-%
-To support this usage pattern of linearity, we rely on
-the notion of borrowing \cite{DBLP:conf/popl/BoylandR05} which
-we demonstrate through a series of examples:
-\begin{itemize}
-\item An API for linear arrays to demonstrate simple borrowing and imperative programming.
-\item
-  A backtracking Sudoku solver to demonstrate
-  \lang's ability to build safe hybrid functional/imperative data-structures.
-\item A pool of linear objects to demonstrate advanced use of regions.
-\end{itemize}
-
 \subsection{Imperative programming with linear arrays}
@@
 
-\subsection{Solving sudokus with hybrid data-structures}
-
-Recently introduced persistent data structures
-permit transient mutations where
-non-linear uses lead to degraded performance
-\cite{DBLP:conf/ml/ConchonF07} or to
-dynamic and static checks \cite{DBLP:journals/pacmpl/Puente17}.
-In particular, persistent Hash-Array-Mapped-Tries (HAMT) have been used with similar
-APIs in several non-pure functional languages (OCaml, Clojure, \dots).
-Affine types help formalize the performance contract between the programmer
-and the library, while borrows avoid the need to thread state explicitly,
-as usually required by an API for immutable data types.
-%
-
-In this section, we present
-a safe API for persistent arrays that support both immutable and mutable features,
-and use it to implement a backtracking Sudoku solver.
-% show how \lang allows us to define
-The Sudoku solver maintains an array to represent the state of
-the game and uses backtracking when there are several choices to proceed.
-The use of backtracking suggests a persistent data structure for the array.
-However, only changes that correspond to a choice point need to use
-the persistence mechanism, others may be implemented as
-cheap in-place mutations.
-
-\cref{sig:hybarray} contains an API @HybArray@
-that enables using mutable and immutable modifications to
-the board through affine types and borrows.
-The signature differs slightly from the @Array@
-signature. For simplicity, the array elements are unrestricted, but
-the structure itself remains linear so as to be implemented in terms of @Array@.
-The in-place mutation function @set_mut@ with type
-@&!('a t) * int * 'a -> unit@ works on a exclusive borrow  whereas the persistent
-@set@ operation has type @&('a t) * int * 'a -> 'a t@. It
-takes a shared borrow because it only reads from the
-argument array, but returns a fresh,  modified structure.
-The code in \cref{ex:cow} contains a very simple implementation of
-@HybArray@ that represents hybrid arrays
-as regular arrays and uses copy-on-write for persistent
-modifications. The function
-@mapi: (int -> &'a -> 'b) -> &('a t) -> 'b t@
-is a simple variation on @Array.map@ where the mapping function
-takes the position of the element.
-% A more elaborate implementation could use persistent arrays
-% \cite{DBLP:conf/ml/ConchonF07}.
-
-\begin{figure}[tp]
-  \centering
-  % \begin{subfigure}{1\linewidth}
-\begin{lstlisting}
-module type HybArray = sig
-  include ARRAY
-  val set : &('a t) * int * 'a -> 'a t
-  val set_mut : &!('a t) * int * 'a -> unit
-end
-
-module Cow : HybArray = struct
-  include Array
-  let set (a, i0, x0) =
-    Array.mapi (fun i x -> if i = i0 then x0 else x) a
-  let set_mut = Array.set
-end
-\end{lstlisting}
-    \vspace{-15pt}
-  %   \caption{Implementation using copy-on-write}
-  % \end{subfigure}
-  \caption{Hybrid arrays}
-    \label{sig:hybarray}
-    \label{ex:cow}
-
-  \lstinputlisting[linerange=1-31]{code/sukodu.affe}
-    \vspace{-15pt}
-  \caption{Excerpt of the Sudoku solver}
-  \label{ex:sudoku}
-\end{figure}
-
-Our implementation of a Sudoku solver (\cref{ex:sudoku}) performs
-modifications that correspond to choice points using @set@,
-which makes it trivial to come back to the previous version
-of the array, while other modifications use @set_mut@, which
-forbids uses of the previous version.
-
-The board is represented as a 2D-matrix (\cref{line:boardtype}), where
-the @Matrix@ type uses the same API as @HybArray@
-but with two indices.
-Each cell contains an integer set that represent admissible solutions so far.
-The Sudoku solver iterates over the cells and tries each possible solution (\cref{line:try_solution}).
-When a value is picked for the current cell, we create a choice point,
-change the cell with an immutable modification (\cref{line:immutset}), and propagate
-the changes with the @propagate@ function.
-The @propagate@ function uses direct mutation through a
-exclusive borrow of the matrix as it need not preserve the previous
-version of the board.
-The implementation of @propagate@ is split into three parts
-for lines, columns, and square, which are all very similar to function
-@propagate_lines@ (\cref{line:propline}).
-
-The typing ensures that the mutations do not compromise the state at
-the choice point, because they operate on a new state @new_g@ created for one
-particular branch of the choice.
-As the @set@ function only requires an unrestricted shared borrow,
-the closure @try_solution@ remains unrestricted even though
-it captures the borrow @&g@. 
-The price is that @try_solution@ cannot escape from
-@&g@'s region. In this example, inference places the region
-around the @begin@/@end@ scope.
-Hence, @try_solution@ can be used in
-the iteration in \cref{line:iter}.
-% As @&g@ is unrestricted, we could even use a parallel iteration
-% instead of a sequential one.
-As @g@ is linear we must free it outside of the region
-before returning (\cref{line:free:g}).
-
-While presented for copy-on-write arrays, the API
-can easily be adapted to other persistent data structures with
-transient mutability such as Relaxed-Radix Balance Vectors (RRB) \cite{DBLP:journals/pacmpl/Puente17}
-or persistent HAMTs \cite{bagwell2001ideal,clojurehamt} to provide  a
-convenient programming style without compromising performance.
-
-% \subsection{Type abstraction and linearity violations}
-
-% While linearity allows to verify many properties, it also prevents
-% the implementation of functions that are nevertheless safe. Rust addresses this
-% issue through the @unsafe@ construct, which allows to locally
-% ignore linearity constraints. This is essential to build basic
-% functions such as splitting strings.
-% %
-% \lang allows to partially emulate this behavior using type abstraction
-% as provided by ML modules.
-% In particular, data types can be internally defined as unrestricted but exposed
-% as affine.
-
-% \TODO{Show example with arrays}
-
-\subsection{Iterators and regions}
-\label{tuto:vector}
-
-% We have seen that borrows are limited to a region.
-In the examples so far, regions do not appear in type signatures.
-But for certain programming idioms, we want to extend the scope of a
-region across a function boundary.
-For instance, how should we fold on an array of linear objects?
-A naive fold function would only work on unrestricted content:
-\begin{lstlisting}
-val fold : ('a : un) => ('a -> 'b -> 'b) -> 'a array -> 'b -> 'b
-\end{lstlisting}
-
-However, if we want to work on linear and affine resources,
-we must make sure those resources are not leaked in the result.
-content. We obtain the following signature:
-
-\begin{lstlisting}
-val fold :
-  ('b:'k),('k <= aff_r) =>
-  (&(aff_r1,'a) -> 'b -{aff_r1}> 'b) -> &('k_1,'a array) -> 'b -{'k_1}> 'b
-\end{lstlisting}
-
-The folded function receives a shared borrow of the element in the array.
-The typing of the callback ensures
-that this borrow is neither captured nor returned by the function.
-%
-This encapsulation is implemented with a universally quantified \emph{kind index variable} $r$.
-The signature prescribes the type @&(aff_r1,'a)@ for the
-shared borrow of the resource with an affine kind at region nesting $r+1$. The return
-type of the callback is constrained to kind @'k <= aff_r@
-so that the callback certainly cannot return the borrowed argument.
-%
-The input of the fold is a shared borrow of the array,
-which ensures that we have the right to share borrows of the inner content and
-make multiple concurrent folds.
-
-We can then easily use this function to fold over an array of files. Here
-we compute the sum of their sizes:
-
-\begin{lstlisting}
-let total_size = fold (fun f s -> File.size f + s) &all_files 0 
-\end{lstlisting}
-
-Naturally, this is not sufficient if we want to mutate the array during iteration.
-For this purpose, we need to take an exclusive borrow of the structure to iterate
-on:
-
-\begin{lstlisting}
-val iter_mut : &!('a array) -> (&!(aff_r1,'a) -{un}> unit) -{aff}> unit
-\end{lstlisting}
-
-Since the callback will be called multiple times, it is explicitly
-annotated with @un@.
-
-While the distinction between mutable and immutable iteration function
-might seem unfortunate, it is typical of
-programming with borrows and is also present in the Rust standard library.
-It allows the programmer to explicitly state how various iterations
-are allowed to compose and be optimized.
-\lang's region variables ensures that the content iterated on can never
-be leaked outside of the iteration function.
-This pattern is essential in many use-cases of linearity such
-as pools of linear objects (see \cref{tuto:pool}).
-
-\lstDeleteShortInline@
-
 %%% Local Variables:

### 2019-11-21 kindly/motivation.tex
Iter Example: Removed empty newlines before lstlistings
Rewrite the intro example.

@@
 the notion of borrowing \cite{DBLP:conf/popl/BoylandR05} which
-we demonstrate through a series of examples:
-\begin{itemize}
-\item An API for linear arrays to demonstrate simple borrowing and imperative programming.
-\item
-  A backtracking Sudoku solver to demonstrate
-  \lang's ability to build safe hybrid functional/imperative data-structures.
-\item A pool of linear objects to demonstrate advanced use of regions.
-\end{itemize}
+we demonstrate through a series of examples.
 
-\subsection{Imperative programming with linear arrays}
-\label{sec:imper-progr}
+As running theme of our examples, we use an API for persistent arrays.
+Recently introduced persistent data structures
+permit transient mutations where
+non-linear uses lead to degraded performance
+\cite{DBLP:conf/ml/ConchonF07} or to
+dynamic and static checks \cite{DBLP:journals/pacmpl/Puente17}.
+In particular, persistent Hash-Array-Mapped-Tries (HAMT) have been used with similar
+APIs in several non-pure functional languages (OCaml, Clojure, \dots).
+Affine types help formalize the performance contract between the programmer
+and the library, while borrows avoid the need to thread state explicitly,
+as usually required by an API for immutable data types.
 
-The API for mutable linear arrays (shown in \cref{sig:array})
+\cref{sig:hybarray} contains an API @PersistArray@
+that enables using mutable and immutable modifications to
+the board through affine types and borrows.
+%
+We first present how to use it imperatively, as if it was a simple
+mutable linear array. We then see how to exploit its persistent
+properties to write a backtracking sudoku solver. Finally, we consider
+iterators on such linear structures.
+
+% \begin{itemize}
+% \item An API for linear arrays to demonstrate simple borrowing and imperative programming.
+% \item
+%   A backtracking Sudoku solver to demonstrate
+%   \lang's ability to build safe hybrid functional/imperative data-structures.
+% \item A pool of linear objects to demonstrate advanced use of regions.
+% \end{itemize}
+
+\begin{figure}[tp]
+  \centering
+  % \begin{subfigure}{1\linewidth}
+\begin{lstlisting}
+module PersistArray : sig
+  type ('a : 'k) t : lin
+  val create : ('a : un) => int -> 'a -> 'a t
+  val free : ('a : aff) => 'a t -> unit
+  val length : &('a t) -> int
+  val get : ('a : un) => &('a t) * int -> 'a
+  val set_mut : ('a : aff) => &!('a t) * int * 'a -> unit
+  val set : ('a : un) => &('a t) * int * 'a -> 'a t
+  val map : (&'a -> 'b) -> &('a t) -> 'b t
+end
+\end{lstlisting}
+% \begin{lstlisting}
+% (* A copy-on-write implementation *)
+% module Cow : PersistArray = struct
+%   type 'a t = 'a array
+%   ...
+%   let set (a, i0, x0) =
+%     mapi (fun i x -> if i = i0 then x0 else x) a
+% end
+% \end{lstlisting}
+  \vspace{-15pt}
+  % \caption{Implementation using copy-on-write}
+  % \end{subfigure}
+  \caption{Hybrid arrays}
+  \label{sig:array}
+  \label{sig:hybarray}
+  \label{ex:cow}
+\end{figure}
+
+\subsection{Imperative programming}
+
+Our API for linear arrays
 aims to safely handle manual allocation and
@@
 can transform the unrestricted elements into linear (affine)
-ones. This array can be processed further using @map@ (or @set@ if
-affine). The @get@ function is only applicable if the element type
+ones. This array can be processed further using @map@ (or @set_mut@ if
+affine). The @get@ and @set@ function are only applicable if the element type
 is unrestricted as one element is duplicated. The @length@
@@
 view or \emph{exclusive borrows} @&!a@ for read-write views.
-Observer functions such as @length@ and @get@ expect an shared borrow
-argument, mutator functions such a @set@ expect
+Observer functions such as @length@, @get@ and @set@ expect a shared borrow
+argument, mutator functions such a @set_mut@ expect
 a exclusive borrow.
@@
 
-The code in \cref{ex:array} demonstrates a use of the API to create an
-array of Fibonacci numbers as the programmer would write it. After
+We can use these arrays in a purely mutable fashion
+to create an
+array of Fibonacci numbers using code very similar to an usual imperative style:
+\begin{lstlisting}
+let mk_fib_array n =
+  let a = create n 1 in
+  for i = 2 to n - 1 do
+    let x = get (&a, i-1) + get (&a, i-2) in(*@\label{ex:array:get}*)
+    set_mut (&!a, i, x)(*@\label{ex:array:set}*)
+  done;
+  a(*@\label{ex:array:return}*)
+# mk_fib_array : int -> int Array.t
+\end{lstlisting}
+
+After
 creation of the array, the presence of a borrow in the for loop
@@
 
-\begin{figure}[tp]
-  \centering
-  \begin{subfigure}[t]{1\linewidth}
-\begin{lstlisting}
-module type ARRAY = sig
-  type ('a : 'k) t : lin
-  val create : ('a : un) => int -> 'a -> 'a t
-  val free : ('a : aff) => 'a t -> unit
-  val length : &('a t) -> int
-  val get : ('a : un) => &('a t) * int -> 'a
-  val set : ('a : aff) => &!('a t) * int * 'a -> unit
-  val map : (&'a -> 'b) -> &('a t) -> 'b t
-end
+More precisely, the code below is an annotated excerpt with explicit regions
+marked in braces @{| ... |}@:
 
-module Array : ARRAY
-\end{lstlisting}
-    \vspace{-15pt}
-    \caption{Signature}
-    \label{sig:array}
-  \end{subfigure}
-
-  \begin{subfigure}[t]{1\linewidth}
-\begin{lstlisting}
-let mk_fib_array n =
-  let a = create n 1 in
-  for i = 2 to n - 1 do
-    let x = get (&a, i-1) + get (&a, i-2) in(*@\label{ex:array:get}*)
-    set (&!a, i, x)(*@\label{ex:array:set}*)
-  done;
-  a(*@\label{ex:array:return}*)
-# mk_fib_array : int -> int Array.t
-\end{lstlisting}
-    \vspace{-10pt}
-    \caption{Example of use}
-    \label{ex:array}
-\begin{lstlisting}[firstnumber=2]
-  let a = create n 1 in
+\begin{lstlisting}[firstnumber=3]
   for i = 2 to n - 1 do {|
@@
 \end{lstlisting}
-    \vspace{-10pt}
-    \caption{Excerpt of \cref{ex:array} with explicit regions}
-    \label{ex:array:region}
-  \end{subfigure}
-  \vspace{-5pt}
-  \caption{Linear arrays}
-  \label{ex:array}
-\end{figure}
 
-\cref{ex:array:region} uses braces @{| ... |}@ to make the
-regions of the example in \cref{ex:array} explicit.  One
+One
 region consists of the header expression of the @let@ in
@@
 In the example, borrows cannot escape  because they are consumed
-immediately by @get@ and @set@.
-
-Region annotations are automatically inserted using
-syntactic cues such as borrows and binders (see
-\cref{regionannot}).  Users may manually insert regions to further restrict the range of a
-borrow. In practice, region annotations are
-rarely needed with common programming idioms.
-% All regions for the examples in this section have been inferred.
-
-% \begin{figure}
-%   \centering
-%   % \begin{subfigure}{0.6\linewidth}
-% \begin{lstlisting}
-% let mk_fib_array n =
-%   let a = create n 1 in
-%   for i = 2 to n - 1 do {|
-%     let x = {| get &a (i-1) + get &a (i-2) |} in(*@\label{ex:array:region:get}*)
-%     set &!a i x
-%   |}
-%   done;
-%   a(*@\label{ex:array:return}*)
-% # mk_fib_array : int -> int Array.t
-% \end{lstlisting}
-%     \caption{Example from \cref{ex:array} with explicit regions}
-%     \label{ex:array:region}
-%   % \end{subfigure}
-% \end{figure}
+immediately by @get@ and @set_mut@.
 
@@
 
-Recently introduced persistent data structures
-permit transient mutations where
-non-linear uses lead to degraded performance
-\cite{DBLP:conf/ml/ConchonF07} or to
-dynamic and static checks \cite{DBLP:journals/pacmpl/Puente17}.
-In particular, persistent Hash-Array-Mapped-Tries (HAMT) have been used with similar
-APIs in several non-pure functional languages (OCaml, Clojure, \dots).
-Affine types help formalize the performance contract between the programmer
-and the library, while borrows avoid the need to thread state explicitly,
-as usually required by an API for immutable data types.
+\begin{figure}[tp]
+  \lstinputlisting[linerange=1-31]{code/sukodu.affe}
+  \vspace{-15pt}
+  \caption{Excerpt of the Sudoku solver}
+  \label{ex:sudoku}
+\end{figure}
+
+On top of the mutation function @set_mut@, our array implementation
+also provides @set@ of type @&('a t) * int * 'a -> 'a t@. It
+takes a shared borrow because it only reads from the
+argument array, but returns a fresh, modified structure.
 %
+This API could be implemented in a wide variety of ways, such as
+copy-on-write (which copies the whole array on each @set_mut@)
+or something more sophisticated\cite{DBLP:conf/ml/ConchonF07}.
 
-In this section, we present
-a safe API for persistent arrays that support both immutable and mutable features,
-and use it to implement a backtracking Sudoku solver.
+We can use our persistent arrays
+to implement a backtracking Sudoku solver.
 % show how \lang allows us to define
@@
 cheap in-place mutations.
-
-\cref{sig:hybarray} contains an API @HybArray@
-that enables using mutable and immutable modifications to
-the board through affine types and borrows.
-The signature differs slightly from the @Array@
-signature. For simplicity, the array elements are unrestricted, but
-the structure itself remains linear so as to be implemented in terms of @Array@.
-The in-place mutation function @set_mut@ with type
-@&!('a t) * int * 'a -> unit@ works on a exclusive borrow  whereas the persistent
-@set@ operation has type @&('a t) * int * 'a -> 'a t@. It
-takes a shared borrow because it only reads from the
-argument array, but returns a fresh,  modified structure.
-The code in \cref{ex:cow} contains a very simple implementation of
-@HybArray@ that represents hybrid arrays
-as regular arrays and uses copy-on-write for persistent
-modifications. The function
-@mapi: (int -> &'a -> 'b) -> &('a t) -> 'b t@
-is a simple variation on @Array.map@ where the mapping function
-takes the position of the element.
-% A more elaborate implementation could use persistent arrays
-% \cite{DBLP:conf/ml/ConchonF07}.
-
-\begin{figure}[tp]
-  \centering
-  % \begin{subfigure}{1\linewidth}
-\begin{lstlisting}
-module type HybArray = sig
-  include ARRAY
-  val set : &('a t) * int * 'a -> 'a t
-  val set_mut : &!('a t) * int * 'a -> unit
-end
-
-module Cow : HybArray = struct
-  include Array
-  let set (a, i0, x0) =
-    Array.mapi (fun i x -> if i = i0 then x0 else x) a
-  let set_mut = Array.set
-end
-\end{lstlisting}
-    \vspace{-15pt}
-  %   \caption{Implementation using copy-on-write}
-  % \end{subfigure}
-  \caption{Hybrid arrays}
-    \label{sig:hybarray}
-    \label{ex:cow}
-
-  \lstinputlisting[linerange=1-31]{code/sukodu.affe}
-    \vspace{-15pt}
-  \caption{Excerpt of the Sudoku solver}
-  \label{ex:sudoku}
-\end{figure}
 
@@
 
-While presented for copy-on-write arrays, the API
+While presented for persistent arrays, this API
 can easily be adapted to other persistent data structures with
@@
 we must make sure those resources are not leaked in the result.
-content. We obtain the following signature:
-
+We obtain the following signature:
 \begin{lstlisting}
@@
 we compute the sum of their sizes:
-
 \begin{lstlisting}
@@
 
-Naturally, this is not sufficient if we want to mutate the array during iteration.
+Naturally, this is not sufficient if we want to mutate the array during iteration,
+as is often the case with @iter@ functions.
 For this purpose, we need to take an exclusive borrow of the structure to iterate
 on:
-
 \begin{lstlisting}
-val iter_mut : &!('a array) -> (&!(aff_r1,'a) -{un}> unit) -{aff}> unit
+val iter : &!('a array) -> (&!(aff_r1,'a) -{un}> unit) -{aff}> unit
 \end{lstlisting}
 
-Since the callback will be called multiple times, it is explicitly
-annotated with @un@.
-
-While the distinction between mutable and immutable iteration function
+While the distinction between mutable and immutable iteration functions
 might seem unfortunate, it is typical of
@@
 This pattern is essential in many use-cases of linearity such
-as pools of linear objects (see \cref{tuto:pool}).
+as pools of linear objects (see example in \cref{tuto:pool}).
 

### 2019-11-21 kindly/intro.tex
Hunting lonely words.

@@
 was required.
-
+%
 As a concrete example, consider a functional programmer who wants to
@@
 \begin{itemize}
-\item A type system that encodes linearity and affinity with
-  borrowing. The type system is polymorphic with respect to types as
-  well as kinds that express constraints on the number of uses of a value.
+\item A type system encoding linearity with
+  borrowing. The type system is polymorphic for types and
+  for kinds that express constraints on the number of uses of a value.
 \item An extension of the HM(X) framework
@@
   multiplicity constraints that supports full, principal type inference.
-\item Full account of the metatheory for type inference and type soundness.
+\item Proof of soundness for the semantics and the inference.
 \item Constraint simplification techniques.

### 2019-11-21 kindly/motivation.tex
Hunting lonely words.

@@
 and the library, while borrows avoid the need to thread state explicitly,
-as usually required by an API for immutable data types.
+as often required by APIs for immutable datatypes.
 
-\cref{sig:hybarray} contains an API @PersistArray@
-that enables using mutable and immutable modifications to
-the board through affine types and borrows.
+\cref{sig:hybarray} contains a simple API for Persistent arrays
+with mutable and immutable modifications through affine types and borrows.
 %
-We first present how to use it imperatively, as if it was a simple
-mutable linear array. We then see how to exploit its persistent
+We first present how to use it imperatively as for
+mutable linear arrays. We then see how to exploit its persistent
 properties to write a backtracking sudoku solver. Finally, we consider
@@
   % \end{subfigure}
-  \caption{Hybrid arrays}
+  \caption{Persistent arrays}
   \label{sig:array}
@@
 
-Our API for linear arrays
+Our API
 aims to safely handle manual allocation and
@@
 
-We can use these arrays in a purely mutable fashion
+We can use the API in a mutable fashion
 to create an
-array of Fibonacci numbers using code very similar to an usual imperative style:
+array of Fibonacci numbers though an imperative coding style:
 \begin{lstlisting}
@@
 
-One
-region consists of the header expression of the @let@ in
-\cref{ex:array:region:get}. It is contained in another region that
-comprises the body of the @for@ loop. \lang guarantees that borrows
+A region consists of the header expression of the @let@ in
+\cref{ex:array:region:get}. It is contained in another region
+spanning the body of the @for@ loop. \lang guarantees that borrows
 never escape the smallest enclosing region. It employs a system of
@@
 % show how \lang allows us to define
-The Sudoku solver maintains an array to represent the state of
+The solver maintains an array to represent the state of
 the game and uses backtracking when there are several choices to proceed.
@@
 The board is represented as a 2D-matrix (\cref{line:boardtype}), where
-the @Matrix@ type uses the same API as @HybArray@
+the @Matrix@ type uses the same API as @PersistArray@
 but with two indices.

### 2019-11-21 kindly/lang.tex
Make sure splitting rules are all the same.
3.2 Automatic Region Annotations: Removed excessive vspace.
XXX Do something horrifying to whitespaces.

@@
   \label{grammar}
+  \vspace{-10pt}
 \end{figure}
@@
 Let us consider the following program with binders and function applications:
-
+%
 \[
@@
 \]
-
+%
 Since the only use of $a$ is a borrow,
@@
 no borrow enters the reference $r$:
-
+%
 \[
@@
 \]
-
+%
 Following the rules, we could merge the two regions around the borrows
@@
   \input{infer/lattice}
+  \vspace{-5pt}
   \caption{Lattice ordering -- $k \lk_\Lat k'$}
   \label{sdtyp:lattice}
+  \vspace{-10pt}
 \end{figure}
@@
 \begin{figure*}[tp]
-  \begin{minipage}{0.28\linewidth}
-  \end{minipage}\hfill
-
   % \begin{minipage}{0.7\linewidth}
@@
     \end{mathpar}
+    \vspace{-10pt}
     \caption{Selected typing rules -- $\inferS{C}{\E}{e}{\tau}$}
@@
     \label{selectrules:matching}
+    \vspace{-5pt}
   % \end{minipage}
@@
   \caption{Type environments}
-  \vspace{10pt}
   \label{grammar:env}
+  \vspace{5pt}
   % \begin{minipage}{0.5\linewidth}
   \centering
-    \begin{tabular}
-      {@{}>{$}r<{$}@{ $\vdash_e$ }
-      >{$}c<{$}@{ $=$ }
-      >{$}c<{$}@{ $\ltimes$ }
-      >{$}c<{$}r}
-
-      \Cleq{\schm}{\kun_\infty}
-      &\bvar{x}{\schm}&\bvar{x}{\schm}&\bvar{x}{\schm}
-      &Both\\
-
-      {\Cempty}&
-      {\bvar{\borrow[\IBORROW]{x}}{\schm}}&
-      {\bvar{\borrow[\IBORROW]{x}}{\schm}}&{\bvar{\borrow[\IBORROW]{x}}{\schm}}
-      &Borrow
-      % \\
-
-      % {\Cempty}&
-      % \svar[\IBORROW]{x}{\schm}^n&\svar[\IBORROW]{x}{\schm}^n&\svar[\IBORROW]{x}{\schm}^n
-      % &BothS
-      \\[2mm]
-
-      {\Cempty}&{B_x}&{B_x}&{\bnone}
-      &Left\\
-      {\Cempty}&{B_x}&{\bnone}&{B_x}
-      &Right\\[2mm]
-
-      {\Cempty}&{\bvar x \schm}&{\svar x \schm^n}&{\bvar x \schm}
-      &Susp\\
-
-      {\Cempty}&
-      {\bvar{\borrow x} \schm}&{\svar[\IBORROW] x \schm^n}&{\bvar{\borrow x} \schm}
-      &SuspB\\
-
-      {\Cempty}&
-      {\svar{x} \schm}&{\svar[\IBORROW] x \schm^n}&{\svar{x} \schm}
-      &SuspS\\
-    \end{tabular}
+    \input{sdtyping/split-binder}
+    \vspace{-5pt}
     \caption{Splitting rules for bindings -- $\lsplit{C}{B}{B_l}{B_r}$}
     \label{sdtyp:split}
+  \vspace{-10pt}
   % \end{minipage}
@@
 the rule {\sc Susp} splits off a borrow from  the resource
-@a@ to use it in the left component of the pair.
+$a$ to use it in the left component of the pair.
 As usual, a borrow cannot be active in the same scope as its resource.
@@
 \label{sdtyping:regions}
+
 \begin{figure}[tp]
@@
   % \end{minipage}\hfill
-  \begin{minipage}{\linewidth}
+  % \begin{minipage}{\linewidth}
     \centering
@@
     \end{mathpar}
+    \vspace{-10pt}
     \caption{Borrowing rules for bindings -- $\lregion{C}{x}{\E}{\E'}$}
     \label{env:rule:borrow}
-  \end{minipage}
+  \vspace{-10pt}
+  % \end{minipage}
 \end{figure}
@@
 $\forall{\kvar_\tvar}\bvar{\tvar}{\kvar_\tvar}.\
-\qual{\Cleq \kvar {\kun_0}}
+\qual{\Cleq {\kvar_\tvar} {\kun_0}}
 {\tvar \tarr{} \tapp\tres\tvar}$\\
@@
 {\borrowty[\IBORROW]{\kvar}{\tapp\tres\tvar} \tarr{} \tvar}$\\
-Reads of a shared borrow of a resource.
+Reads from a shared borrow of a resource.
 
@@
 \borrowty[\MBORROW]{\kvar}{\tapp\tres\tvar} \tarr{} \tvar \tarr{\kaff} \tunit$\\
-Writes in an exclusive borrow of a resource.
+Writes to an exclusive borrow of a resource.
 
@@
 \tapp\tres\tvar \tarr{} \tunit$\\
-Disposes of a resource.
+Destroys a resource.
 \end{itemize}

### 2019-11-21 kindly/lang.tex
3.3.3 Copying \& Splitting: replaced @a@ with $a$.

@@
   \label{grammar}
-  \vspace{-10pt}
 \end{figure}
@@
   \input{infer/lattice}
-  \vspace{-5pt}
   \caption{Lattice ordering -- $k \lk_\Lat k'$}
   \label{sdtyp:lattice}
-  \vspace{-10pt}
 \end{figure}
@@
 \begin{figure*}[tp]
+  \begin{minipage}{0.28\linewidth}
+  \end{minipage}\hfill
+
   % \begin{minipage}{0.7\linewidth}
@@
     \end{mathpar}
-    \vspace{-10pt}
     \caption{Selected typing rules -- $\inferS{C}{\E}{e}{\tau}$}
@@
     \label{selectrules:matching}
-    \vspace{-5pt}
   % \end{minipage}
@@
   \caption{Type environments}
+  \vspace{10pt}
   \label{grammar:env}
-  \vspace{5pt}
   % \begin{minipage}{0.5\linewidth}
@@
     \input{sdtyping/split-binder}
-    \vspace{-5pt}
     \caption{Splitting rules for bindings -- $\lsplit{C}{B}{B_l}{B_r}$}
     \label{sdtyp:split}
-  \vspace{-10pt}
   % \end{minipage}
@@
 \label{sdtyping:regions}
-
 \begin{figure}[tp]
@@
   % \end{minipage}\hfill
-  % \begin{minipage}{\linewidth}
+  \begin{minipage}{\linewidth}
     \centering
@@
     \end{mathpar}
-    \vspace{-10pt}
     \caption{Borrowing rules for bindings -- $\lregion{C}{x}{\E}{\E'}$}
     \label{env:rule:borrow}
-  \vspace{-10pt}
-  % \end{minipage}
+  \end{minipage}
 \end{figure}
@@
 $\forall{\kvar_\tvar}\bvar{\tvar}{\kvar_\tvar}.\
-\qual{\Cleq {\kvar_\tvar} {\kun_0}}
+\qual{\Cleq \kvar {\kun_0}}
 {\tvar \tarr{} \tapp\tres\tvar}$\\
@@
 {\borrowty[\IBORROW]{\kvar}{\tapp\tres\tvar} \tarr{} \tvar}$\\
-Reads from a shared borrow of a resource.
+Reads of a shared borrow of a resource.
 
@@
 \borrowty[\MBORROW]{\kvar}{\tapp\tres\tvar} \tarr{} \tvar \tarr{\kaff} \tunit$\\
-Writes to an exclusive borrow of a resource.
+Writes in an exclusive borrow of a resource.
 
@@
 \tapp\tres\tvar \tarr{} \tunit$\\
-Destroys a resource.
+Disposes of a resource.
 \end{itemize}

### 2019-11-21 kindly/lang.tex
Misc.

--- 2019-11-21 kindly/lang.tex
@@
 or to conflicting uses.
-
+%
 Let us consider the following program with binders and function applications:
@@
 
-\subsubsection{Kinds and constraints}
+\paragraph{Kinds and constraints}
 
@@
 
-Kinds form a lattice with an ordering that implements two principles
-(see \cref{sdtyp:lattice}). First, unrestricted values may be used
+Kinds form a lattice with an ordering implementing two principles
+(see \cref{sdtyp:lattice}). Unrestricted values may be used
 where affine ones are expected and affine ones are less restricted
 than linear ones as reflected in \TirName{Lat-UAL}.
-Second, values usable at level $n$ may be used at any more deeply
+Values usable at level $n$ may be used at any more deeply
 nested level $n'$ (where $n\le\infty$ for any $n$) as defined in the \TirName{Lat-Level} axioms.
@@
 Constraints are conjunctions of inequality constraints over this kind
-lattice. In most uses, a constraint gives an upper or lower bound for
-a kind variable or it relates two kind variables.
+lattice. In most uses, constraints give upper or lower bounds for
+kind variables or relate two kind variables.
 
@@
   \vspace{-10pt}
+\end{figure}
+
+\paragraph{Environments and bindings}
+\label{sdtyping:envs}
+
+\begin{figure}[tp]
+  \begin{align*}
+    \E &::=\ \Eempty \mid \E;B\tag{Environments}\\
+    B &::=\ \bnone % \tag{Empty} \\
+    \mid \bvar{\tvar}{\kschm} % \tag{Types}
+    \mid \bvar{x}{\schm}% \tag{Variables}
+    \mid \bvar{\borrow{x}}{\borrowty{k}{\tau}}% \tag{Borrows}\\
+    \mid \svar{x}{\schm}^n % \tag{Suspended}
+    % |&\ \tydecl{t}{\kschm}{K}{\tau}\tag{Type Declaration}\\
+  \end{align*}
+  \vspace{-10pt}
+  \caption{Type environments}
+  \label{grammar:env}
+  \vspace{5pt}
+  % \begin{minipage}{0.5\linewidth}
+  \centering
+    \input{sdtyping/split-binder}
+    \vspace{-5pt}
+    \caption{Splitting rules for bindings -- $\lsplit{C}{B}{B_l}{B_r}$}
+    \label{sdtyp:split}
+  \vspace{-10pt}
+  % \end{minipage}
 \end{figure}
@@
 
-\subsubsection{Environments and bindings}
-\label{sdtyping:envs}
-
-\begin{figure}[tp]
-  \begin{align*}
-    \E &::=\ \Eempty \mid \E;B\tag{Environments}\\
-    B &::=\ \bnone % \tag{Empty} \\
-    \mid \bvar{\tvar}{\kschm} % \tag{Types}
-    \mid \bvar{x}{\schm}% \tag{Variables}
-    \mid \bvar{\borrow{x}}{\borrowty{k}{\tau}}% \tag{Borrows}\\
-    \mid \svar{x}{\schm}^n % \tag{Suspended}
-    % |&\ \tydecl{t}{\kschm}{K}{\tau}\tag{Type Declaration}\\
-  \end{align*}
-  \vspace{-10pt}
-  \caption{Type environments}
-  \label{grammar:env}
-  \vspace{5pt}
-  % \begin{minipage}{0.5\linewidth}
-  \centering
-    \input{sdtyping/split-binder}
-    \vspace{-5pt}
-    \caption{Splitting rules for bindings -- $\lsplit{C}{B}{B_l}{B_r}$}
-    \label{sdtyp:split}
-  \vspace{-10pt}
-  % \end{minipage}
-\end{figure}
-
 \lang controls the use of variables by supporting new modes of
@@
 such a suspended binding on entry to the $x$-region. It indicates
-that the borrow $\borrow{x}$ can be used directly. The kind $k$
+that the borrow $\borrow{x}$ can be used directly. Kind $k$
 restricts the lifetime of the borrow to the region (see \cref{selectrules:borrow}).
@@
 
-\subsubsection{Copying and Splitting}
+\paragraph{Copying and Splitting}
 \label{sdtyping:split}
@@
 
-\subsubsection{Pattern matching}
+\paragraph{Pattern matching}
 \label{sdtyping:matching}
@@
 
-\subsubsection{Regions}
+\paragraph{Regions}
 \label{sdtyping:regions}
@@
 
-% \subsubsection{Kind checking}
+% \paragraph{Kind checking}
 
@@
 
-\subsubsection{Resource management}
+\paragraph{Resource management}
 

### 2019-11-21 kindly/motivation.tex
XXX Do something horrifying to whitespaces.
Misc.

@@
 borrows are no longer in scope. 
-
-More precisely, the code below is an annotated excerpt with explicit regions
+%
+More precisely, here is is an annotated excerpt with explicit regions
 marked in braces @{| ... |}@:
@@
   \label{ex:sudoku}
+  \vspace{-10pt}
 \end{figure}
@@
 On top of the mutation function @set_mut@, our array implementation
-also provides @set@ of type @&('a t) * int * 'a -> 'a t@. It
+also provides @set: &('a t) * int * 'a -> 'a t@. It
 takes a shared borrow because it only reads from the

### 2019-11-21 kindly/lang.tex
3.3.6 Resource management: Fixed typo in type signature of create.
3.3.6 Resource management: fixed typo.
Misc.

@@
 
-In the examples presented in \cref{motivation}, region annotations are optional.
+In \cref{motivation}, region annotations are optional.
 In our formalization, regions must be syntactically explicit and annotated with
@@
 of the form $\region{\Sone x\BORROW}{e}$.
-We give an informal presentation of our code transformation. The complete
-definition is given in \cref{appendix:regionannot}. This code transformation
+We give an informal presentation of our code transformation and defer
+the complete definition to \cref{appendix:regionannot}. This code transformation
 aims to find, for each borrow $\borrow{x}$, the biggest region satisfying
@@
 \end{enumerate}
-Following these rules, we start from the leafs and grow the region
+We then start from the leafs and grow the region
 associated to each borrow until it can't grow anymore either due to a binding
-or to conflicting uses.
-%
-Let us consider the following program with binders and function applications:
+or to conflicting uses. We demonstrate
+on the following program:
 %
@@
 %
-Since the only use of $a$ is a borrow,
-the region covers the entire lexical scope of $a$. For $x$ however, there are
-multiple conflicting borrow and we must be more careful. We place a first
+$a$ only has one borrow, its region covers its whole lexical scope.
+$x$ has multiple conflicting borrow and requires more consideration.
+We place a first
 region around the exclusive borrow and its function application,
 and a second region around both shared borrows. This placement of region
-is optimal, as making any of the region bigger would cause a borrowing
+is optimal: making any region bigger would cause a borrowing
 error.
@@
 
-Naturally, these rules are best effort and programmers must sometime
-annotate regions explicitly. We take such manual annotations
-into account as an upper bound on the regions of all contained borrows.
+Programmers might also want to annotate regions explicitly.
+Such annotations are considered as
+upper bound on the contained regions.
 In the following program, a manual annotation has been inserted to ensure
@@
 %
-Following the rules, we could merge the two regions around the borrows
+The rules allow merging the two regions around the borrows
 $\borrow[\kun]{a}$. However
 the explicit annotation indicates that the region should stay around the closure
-passed as argument of $set$. This is particularly useful to control captures
+passed as argument of $set$. This is useful to control captures
 by imperative APIs.
@@
 Each constant is composed of a ``quality'' and a ``level''.
-The quality describes the use pattern of the type. It can be
+The quality describes the use pattern of the type:
 unrestricted $\kun$, affine $\kaff$, or linear $\klin$. The level  $n \in \Nat \cup \{\infty\}$
@@
 $\kvar$ must be unrestricted, but can be local to a region.
-
-Kinds form a lattice with an ordering implementing two principles
-(see \cref{sdtyp:lattice}). Unrestricted values may be used
+%
+Kinds form a lattice described in \cref{sdtyp:lattice}.
+Unrestricted values may be used
 where affine ones are expected and affine ones are less restricted
@@
 Values usable at level $n$ may be used at any more deeply
-nested level $n'$ (where $n\le\infty$ for any $n$) as defined in the \TirName{Lat-Level} axioms.
-
+nested level $n'$ as defined in the \TirName{Lat-Level} axioms.
+%
 Constraints are conjunctions of inequality constraints over this kind
@@
 \lang controls the use of variables by supporting new modes of
-binding in a type environment $\E$.
-Its definition is given in  \cref{grammar:env}.
-In addition to standard bindings of type variables to kind schemes,
-$\bvar{\tvar}{\kschm}$,  and value bindings, $\bvar{x}{\schm}$, there
-are suspended bindings and borrow bindings.
+binding in type environments $\E$, as defined in \cref{grammar:env}.
+Environments contain standard bindings of type variables to kind schemes,
+$\bvar{\tvar}{\kschm}$, value bindings $\bvar{x}{\schm}$, but also
+suspended and borrow bindings.
 A suspended binding, $\svar{x}{\schm}^n$, indicates that $x$ is
 earmarked for a borrowing use in a nested region
-marked with $x$ (see \cref{sdtyping:regions}), but
+marked with $x$ % (see \cref{sdtyping:regions}),
+but
 cannot be used directly.
@@
 
-In order to deconstruct pairs in the presence of linear types, we require
-more than projection operators. Indeed, given a linear pair $x$,
-$\app{\operatorname{fst}}{x}$ would consume $x$ without giving access to the second
-member.
-Furthermore, we would like to be able to deconstruct a borrow of a pair
-and obtain borrows of the components.
-For these purposes, we provide the construct
+Elimination of pairs is done using a matching construct 
 $(\matchin{x,x'}{e_1}{e_2})$ whose
 typing rule is shown in \cref{selectrules:matching}.
+This construction can eliminate both normal pairs, but also borrows
+of pairs.
 A syntactic marker $\etransfm$ indicates if it applies to
@@
 
-The typing of pattern matching is voluntarily simple here.
-Using a typeclass mechanism could avoid the need for explicit markers
-for borrow matches. Furthermore, a richer pattern language would allow
-finer-grained borrowing of inner components.
-We believe both extension are compatible with our system
-and are out of scope of this article.
+Matching is voluntarily simple here and
+could be improved using a richer pattern language
+to access inner components and a typeclass mechanisms
+to avoid the need for explicit markers for borrow matches.
+Both extensions are compatible with our system and have been partially
+implemented in our prototype type-checker.
 
@@
 
-Our system would be incomplete without a way to actually manipulate resources.
-We consider the simple case of an abstract resource
+As a model of resource, we consider a simple abstract
 type $\tapp{\tres}{\tau}$ whose content of type $\tau$ must be $\kun_0$ and
@@
 The semantics is defined in terms of \emph{elaborated expressions} $e$
-with kind and constraint annotations inserted by the typechecker.
-Polymorphic functions give rise to elaborated
-abstractions which abstract over kind variables $\Multi\kappa$ and
-which are tagged with a constraint $C$ and a kind $k$ indicating their
-usage restriction (linear, affine, etc). Their use
-gives rise to explicit instantiation of these kind variables. Pairs
-come with a kind tag $k$ indicating the usage restriction. Application
-is annotated with evidence $\Sp$ of the inferred splitting.
+with kind, constraint, and splitting annotations inserted by the typechecker.
+A splitting $\Sp$ is evidence of the splitting relation for type environments
+used in the typing rules.
+
+Let-Polymorphism is demonstrated for functions giving rise to elaborated
+$\eletfun$ expressions annotated with a type scheme $\schm$ and a kind $k$ indicating their
+usage restriction (linear, affine, etc) relative to the variables and
+constraints from $\schm$. Their use
+gives rise to explicit instantiation of the kind and type variables. Pairs
+come with a kind tag $k$ indicating the usage restriction.
 

### 2019-11-21 kindly/soa.tex
Misc.

@@
 It  would be interesting to extend \lang with limited
-forms of typestate as a local, opt-in feature that would provide
+forms of typestate as a local, opt-in feature to provide
 more expressivity at the cost of inference.

### 2019-11-22 kindly/infer/infer.tex
Misc.
Remove pairs from the body.

@@
 We now formulate our inference algorithm
-based on the $HM(X)$ framework~\citep{DBLP:journals/tapos/OderskySW99}.
-$HM(X)$ is a Hindley-Milner type system for a language
+based on the $HM(X)$ framework~\citep{DBLP:journals/tapos/OderskySW99},
+a Hindley-Milner type system for a language
 with qualified types where constraints are expressed in an arbitrary
@@
 which solves and simplifies constraints in a principled way.
-We start by various preliminary definitions.
 
@@
 
-Unlike the syntax-directed version of our type system, knowing which elements
-are input and output or our inference judgements is critical. In the rest
-of this section, when presenting a new judgement,
-we write input parameters in \textbf{\textcolor{DarkSlateBlue}{bold blue}}. The rest will then be
+In the context of inference, it's critical to know which elements
+are input and output of inference judgements.
+In the following, when presenting a new judgement,
+we write input parameters in \textbf{\textcolor{DarkSlateBlue}{bold blue}}. The rest will be
 output parameters.
@@
 
-In order to determine if a variable is used in an affine manner, we must track
-its uses and the associated kinds. For instance, in the expression
-$(x,x)$, $x$ is used twice. If $x$ is of type $\tau$, which is of kind $k$,
-we must add the constraint $\Cleq{k}{\kun}$.
+To determine if a variable is used in an affine manner, we must track
+its uses and the associated kinds. In the expression
+$f\ x\ x$, $x$ is used twice. If $x$ is of type $\tau$, which is of kind $k$,
+we add the constraint $\Cleq{k}{\kun}$.
 %
-In order to infer such constraints, our inference judgement will not only
-take an environment as parameter but also return an environment which
-we call ``usage environment'' and which summarizes how variables and borrows
-are used. Usage environments follow the exact same grammar
-as normal environments. In order to distinguish them more easily,
-we denote them by $\Sv$.
-
+To infer such constraints, our inference judgement not only
+take an environment as parameter but also return an environment called
+``usage environment'', denoted $\Sv$,
+which summarizes usages of variables and borrows.
+Usage environments follow the grammar
+of normal environments.
+%
 In \cref{sdtyping}, we used relations to split environments in two and to
@@
 \item $\bsplit{C}{\Sv}{\inP{\Sv_1}}{\inP{\Sv_2}}$.
-  Given two usage environments $\inP{\Sv_1}$ and $\inP{\Sv_2}$
-  (inferred from two subexpressions, for instance),
+  Given two usage environments $\inP{\Sv_1}$ and $\inP{\Sv_2}$,
   we return $\Sv$, the merged environment, and $C$, a set
   of constraints that must be respected.
-  This relation is total and non-ambiguous given $\Sv_1$ and $\Sv_2$
-  and uses the same rules as the one presented in \cref{sdtyping:split}.
 \item $\bregion[\inP{n}]{C}{\inP{x}}{\Sv}{\inP\Sv'}$.
@@
   is replaced by
-  a suspended binding. We also return $C$, a set of constraints that must
-  be respected.
-  Again, the relation is total and non-ambiguous for any given $\Sv'$,
-  $n$ and $x$,
-  and uses the rules presented in \cref{sdtyping:regions}.
+  a suspended binding. We also return the constraints $C$.
 \end{itemize}
-
+Both relations are total and non-ambiguous in term of their input
+(i.e., functions), and use
+the rules presented in \cref{sdtyping:split,sdtyping:regions}.
+%
 The relations used for syntax-directed typing can trivially be defined
@@
 The HM(X) framework assumes the existence of a ``$\operatorname{normalize}$''
-function which takes a constraint $C$, a substitution $\psi$ and returns a
+function which takes a constraint $\inP{C}$, a substitution $\inP\psi$ and returns a
 simplified constraint $C'$
 and an updated substitution $\unif'$.
-Normalization must return a normal form such that $\unif'$ is a most general unifier.
-We detail the implementation
-of the normalization function and its property in \cref{infer:solving}.
-In the meantime, we simply
-assume the existence of such a function for our constraint system.
+Normalization returns a normal form such that $\unif'$ is a most general unifier.
+For now, we simply
+assume the existence of such a function for our constraint system
+and delay details to \cref{infer:solving}.
 
@@
 inference algorithm.
-Unlike in the syntax-directed version, $\E$ only regular and type bindings.
+Unlike in the syntax-directed version, $\E$ contains only regular and type bindings.
 Suspended and borrow bindings can only be present in $\Sv$.
@@
 of our inference algorithm and the differences with the syntax-directed
-system.
-As before, the complete type inference rules are shown in \cref{appendix:infer}.
+system in \cref{rule:infer:envs}.
+The complete type inference rules are shown in \cref{appendix:infer}.
 
-\subsubsection{Environments and bindings}
+\paragraph{Environments and bindings}
 \label{infer:envs}
@@
     \ruleIRegion
+    % \and
+    % \ruleIPair
     \and
-    \ruleIPair
+    \ruleIApp
     \and
@@
 %
-In our syntax-directed system, we ensured
-that linear variables are not discarded at the \emph{leaves}, through
-the {\sc Var} rule. In the inference algorithm, we operate in the opposite
-direction: we collect data from the leaves, and only enforce linearity
+In the syntax-directed system, the {\sc Var} rule ensure
+that linear variables are not discarded at the \emph{leaves}.
+In the inference algorithm, we operate in the opposite
+direction: we collect data from the leaves and enforce linearity
 at \emph{binders}. This is reflected in the {\sc Var$_I$} and
-{\sc Abs$_I$} rules presented in \cref{rule:infer:envs}.
-Typing for variables is very similar to traditional (non-affine) ML
-typing. The only difference is that we collect
-the fact that $x$ was used with the scheme $\schm$ by returning
+{\sc Abs$_I$} rules.
+Typing for variables is very similar to traditional (non-affine) HM.
+The difference is that we record
+that $x$ was used with the scheme $\schm$ by returning
 a usage environment $\Sv = \{ \bvar{x}{\schm} \}$.
@@
 usage of linear variable via the $\Weaken$ property.
-We demonstrate how this is achieved for lambda expressions
+We demonstrate this for lambda expressions
 in the {\sc Abs$_I$} rule.
 First, we typecheck the body of the lambda and obtain a usage
-environment $\Sv_x$. Just like in the syntax-directed type system,
+environment $\Sv_x$. Like in the syntax-directed type system,
 we introduce the constraint
 $\Cleq{\Sv\Sdel{x}}{\kvar}$ which properly accounts for captures in
-the body of the lambda expression. We also introduce the constraint
+the body of the lambda expression. We then introduce the constraint
 $\Weaken_{\bvar{x}{\sigma}}(\Sv)$, which is false only if weakening
-is necessary but impossible due to linearity.
-The $\Weaken$ constraint is introduced at each binding point in the program
-such as let-bindings and pattern matching.
-
-Finally, we use normalization on complex constraints to ensure
+is necessary-yet-impossible due to linearity.
+The $\Weaken$ constraint is introduced at each binding point in the program.
+%
+Finally, we normalize constraints to ensure
 that the inference algorithm always return the simplest possible
@@
 
-\subsubsection{Splitting and Regions}
+\paragraph{Splitting and Regions}
 \label{infer:split}
@@
 
-The inference version
-of the {\sc Pair} and {\sc Region} rules is shown in \cref{rule:infer:envrules}.
-While the rules are quite similar to the original version, the usage
-environment $\Sv$ is now an \emph{output} of the algorithm.
+Inference versions
+of the {\sc App} and {\sc Region} rules
+are similar to the original ones, but now \emph{return} the usage
+environment $\Sv$.
 As such, we use the ``inference'' version of the relations on
@@
 which returns the necessary constraints.
-As appropriate for inference, these relations
-are total and non ambiguous for their input parameters ($\Sv_1, \Sv_2$ and $\Sv'$, respectively).
 We then collect all constraints and normalize them.

### 2019-11-22 kindly/lang.tex
Remove pairs from the body.

@@
     \begin{mathpar}
+      \ruleSDVar
+      \and
+      \ruleSDApp
+      \and
+      \ruleSDRegion
+      \and
       \ruleSDBorrow
@@
       \and
-      \ruleSDRegion
-      \and
-      \ruleSDVar
-      \and
-      \ruleSDPair
-      \and
-      \ruleSDMatchPair
+      \inferrule[BorrowBinding]{
+        \entail{C}{(\BORROW_n\lk k) \wedge (k \lk \BORROW_\infty)}\\
+        \BORROW\in\left\{\kun,\kaff\right\}
+      }{
+        \lregion{C}{}
+        {\svar[\BORROW]{x}{\tau}^n}
+        {\bvar{\borrow[\BORROW]{x}}{\borrowty[\BORROW] k{\tau}}}
+      }
+      % \and
+      % \ruleSDPair
+      % \and
+      % \ruleSDMatchPair
     \end{mathpar}
     \vspace{-10pt}
-    \caption{Selected typing rules -- $\inferS{C}{\E}{e}{\tau}$}
+    \caption{Selected typing rules ($\inferS{C}{\E}{e}{\tau}$)
+      and borrowing rules ($\lregion{C}{x}{\E}{\E'}$)} 
     \label{selectrules:borrow}
     \label{selectrules:binders}
-    \label{sdtyp:pair}
+    \label{sdtyp:app}
     \label{selectrules:region}
-    \label{selectrules:matching}
+    \label{env:rule:borrow}
     \vspace{-5pt}
@@
 
-The {\sc Pair} typing rule in \cref{sdtyp:pair} demonstrates how \lang
+The {\sc App} typing rule in \cref{sdtyp:app} demonstrates how \lang
 deals with duplication and dropping of values.
@@
 
-\cref{sdtyp:split} shows the action of the splitting rules on single
-bindings. If $x$'s type is unrestricted, then
-the {\sc Both} rule indicates that we can duplicate it.
-Similarly, unrestricted borrows can be duplicated according to rule
+\cref{sdtyp:split} shows the action of splitting rules on single
+bindings. If $x$'s type is unrestricted, 
+rule {\sc Both} indicates that we can duplicate it.
+Similarly, unrestricted borrows can be duplicated with rule
 {\sc Borrow}.
-The {\sc Left} and {\sc Right} rules are always applicable and move a binding
+{\sc Left} and {\sc Right} rules are always applicable and move a binding
 either to the left or right environment.
@@
 with a borrow and {\sc SuspS} with a suspended binding.
-%
+
 Splitting applies whenever an
-expression has multiple subexpressions:  pairs, function applications,
+expression has multiple subexpressions:  function applications,
 and let bindings. In the
 expression
-$\letin{a}{\text{create}\ 10\ x}
-{\introPair{\region[{}]{\Sone x\IBORROW}{\app{\text{length}}{\borrow[\IBORROW]{a}}}}{a}}$,
+$\letin{a}{\text{create}\ 8\ x}
+{f\ \region[{}]{\Sone a\IBORROW}{\text{length}\ \borrow[\IBORROW]{a}}\ a}$,
 the rule {\sc Susp} splits off a borrow from  the resource
-$a$ to use it in the left component of the pair.
+$a$ to use it in the left argument.
 As usual, a borrow cannot be active in the same scope as its resource.
-The \emph{region} around its use ensures that the borrow in the left component does not
+The \emph{region} around its use ensures that the borrow in the left argument
+does not
 escape, which brings us to the next topic.
 
-\paragraph{Pattern matching}
-\label{sdtyping:matching}
+% \paragraph{Pattern matching}
+% \label{sdtyping:matching}
 
-Elimination of pairs is done using a matching construct 
-$(\matchin{x,x'}{e_1}{e_2})$ whose
-typing rule is shown in \cref{selectrules:matching}.
-This construction can eliminate both normal pairs, but also borrows
-of pairs.
-A syntactic marker $\etransfm$ indicates if it applies to
-a pair ($\etransfm = \operatorname{id}$) or a borrow ($\etransfm = \&^\BORROW$).
-If $\etransfm = \operatorname{id}$, the typing simplifies to
-a regular destruct on pairs.
-Otherwise, $e_1$ is expected to be of type
-$\borrowty{k}{\tyPair{\tau_1}{\tau'_1}}$
-while $x$ and $x'$ are of type respectively
-$\borrowty{k}{\tau_1}$ and $\borrowty{k}{\tau'_1}$.
+% Elimination of pairs is done using a matching construct 
+% $(\matchin{x,x'}{e_1}{e_2})$ whose
+% typing rule is shown in \cref{selectrules:matching}.
+% This construction can eliminate both normal pairs, but also borrows
+% of pairs.
+% A syntactic marker $\etransfm$ indicates if it applies to
+% a pair ($\etransfm = \operatorname{id}$) or a borrow ($\etransfm = \&^\BORROW$).
+% If $\etransfm = \operatorname{id}$, the typing simplifies to
+% a regular destruct on pairs.
+% Otherwise, $e_1$ is expected to be of type
+% $\borrowty{k}{\tyPair{\tau_1}{\tau'_1}}$
+% while $x$ and $x'$ are of type respectively
+% $\borrowty{k}{\tau_1}$ and $\borrowty{k}{\tau'_1}$.
 
-Matching is voluntarily simple here and
-could be improved using a richer pattern language
-to access inner components and a typeclass mechanisms
-to avoid the need for explicit markers for borrow matches.
-Both extensions are compatible with our system and have been partially
-implemented in our prototype type-checker.
+% Matching is voluntarily simple here and
+% could be improved using a richer pattern language
+% to access inner components and a typeclass mechanisms
+% to avoid the need for explicit markers for borrow matches.
+% Both extensions are compatible with our system and have been partially
+% implemented in our prototype type-checker.
 
@@
 
-\begin{figure}[tp]
-  % \begin{minipage}{0.4\linewidth}
+% \begin{figure}[tp]
+%   % \begin{minipage}{0.4\linewidth}
+%   %   \centering
+%   %   \begin{mathpar}
+%   %     \ruleSDRegion
+%   %   \end{mathpar}
+%   %   \caption{The {\sc Region} rule}
+%   %   \label{selectrules:region}
+%   % \end{minipage}\hfill
+%   % \begin{minipage}{\linewidth}
 %     \centering
+%     % \begin{tabular}
+%     %   {@{}>{$}r<{$}@{ $\vdash_e$ }
+%     %   >{$}c<{$}@{ $\rightsquigarrow_n^{x}$ }
+%     %   >{$}l<{$}
+%     %   r}
+
+%     %   (\kun_n\lk k\lk\kun_\infty)
+%     %   &{\svar[\IBORROW]{x}{\tau}^n}
+%     %   &{\bvar{\borrow[\IBORROW]{x}}{\borrowty[\IBORROW] k{\tau}}}
+%     %   &Immut\\
+
+%     %   (\kaff_n\lk k\lk\kaff_\infty)
+%     %   &\svar[\MBORROW]{x}{\tau}^n
+%     %   &\bvar{\borrow[\MBORROW]{x}}{\borrowty[\MBORROW] k{\tau}}
+%     %   &Mut
+%     % \end{tabular}
 %     \begin{mathpar}
-  %     \ruleSDRegion
 %     \end{mathpar}
-  %   \caption{The {\sc Region} rule}
-  %   \label{selectrules:region}
-  % \end{minipage}\hfill
-  % \begin{minipage}{\linewidth}
-    \centering
-    % \begin{tabular}
-    %   {@{}>{$}r<{$}@{ $\vdash_e$ }
-    %   >{$}c<{$}@{ $\rightsquigarrow_n^{x}$ }
-    %   >{$}l<{$}
-    %   r}
-
-    %   (\kun_n\lk k\lk\kun_\infty)
-    %   &{\svar[\IBORROW]{x}{\tau}^n}
-    %   &{\bvar{\borrow[\IBORROW]{x}}{\borrowty[\IBORROW] k{\tau}}}
-    %   &Immut\\
-
-    %   (\kaff_n\lk k\lk\kaff_\infty)
-    %   &\svar[\MBORROW]{x}{\tau}^n
-    %   &\bvar{\borrow[\MBORROW]{x}}{\borrowty[\MBORROW] k{\tau}}
-    %   &Mut
-    % \end{tabular}
-    \begin{mathpar}
-      \inferrule{
-        \entail{C}{(\BORROW_n\lk k) \wedge (k \lk \BORROW_\infty)}\\
-        \BORROW\in\left\{\kun,\kaff\right\}
-      }{
-        \lregion{C}{}
-        {\svar[\BORROW]{x}{\tau}^n}
-        {\bvar{\borrow[\BORROW]{x}}{\borrowty[\BORROW] k{\tau}}}
-      }
-    \end{mathpar}
-    \vspace{-10pt}
-    \caption{Borrowing rules for bindings -- $\lregion{C}{x}{\E}{\E'}$}
-    \label{env:rule:borrow}
-  \vspace{-10pt}
-  % \end{minipage}
-\end{figure}
+%     \vspace{-10pt}
+%     \caption{Borrowing rules for bindings -- }
+%   \vspace{-10pt}
+%   % \end{minipage}
+% \end{figure}
 
@@
 \cref{fig:syntax-internal-language}.
-To demonstrate resource handling, the semantics comes with an abstract
-notion of linear resources that can be created, destroyed, observed
-(through unrestricted borrows), and
-updated (through affine borrows).
+% To demonstrate resource handling, the semantics comes with an abstract
+% notion of linear resources that can be created, destroyed, observed
+% (through unrestricted borrows), and
+% updated (through affine borrows).
 
@@
 
-The @Pair@ case corresponds to the \TirName{Pair} typing rule. It
-requires no new features.
+% The @Pair@ case corresponds to the \TirName{Pair} typing rule. It
+% requires no new features.
 

### 2019-11-22 kindly/motivation-persistent.tex
Reshuffle the intro again.

--- 2019-11-21 kindly/motivation.tex
@@
-\section{Linearity, Affinity, and Borrows at Work}
-\label{motivation}
+\subsection{Solving sudokus with hybrid data-structures}
 
-\lstMakeShortInline[keepspaces,basicstyle=\small\ttfamily]@
-
-\lang{} is amenable to a functional resource-passing style
-common in functional encodings of session types (e.g.,
-\cite{DBLP:journals/jfp/Padovani17}; see also
-\cref{sec:session-linearity} in the supplement). But it really shines
-when manipulating mutable resources like buffers or connection pools
-using a mixed functional and imperative programming style.
-%
-To support this usage pattern of linearity, we rely on
-the notion of borrowing \cite{DBLP:conf/popl/BoylandR05} which
-we demonstrate through a series of examples.
-
-As running theme of our examples, we use an API for persistent arrays.
 Recently introduced persistent data structures
@@
 and the library, while borrows avoid the need to thread state explicitly,
-as often required by APIs for immutable datatypes.
+as usually required by an API for immutable data types.
+%
 
-\cref{sig:hybarray} contains a simple API for Persistent arrays
-with mutable and immutable modifications through affine types and borrows.
-%
-We first present how to use it imperatively as for
-mutable linear arrays. We then see how to exploit its persistent
-properties to write a backtracking sudoku solver. Finally, we consider
-iterators on such linear structures.
-
-% \begin{itemize}
-% \item An API for linear arrays to demonstrate simple borrowing and imperative programming.
-% \item
-%   A backtracking Sudoku solver to demonstrate
-%   \lang's ability to build safe hybrid functional/imperative data-structures.
-% \item A pool of linear objects to demonstrate advanced use of regions.
-% \end{itemize}
-
-\begin{figure}[tp]
-  \centering
-  % \begin{subfigure}{1\linewidth}
-\begin{lstlisting}
-module PersistArray : sig
-  type ('a : 'k) t : lin
-  val create : ('a : un) => int -> 'a -> 'a t
-  val free : ('a : aff) => 'a t -> unit
-  val length : &('a t) -> int
-  val get : ('a : un) => &('a t) * int -> 'a
-  val set_mut : ('a : aff) => &!('a t) * int * 'a -> unit
-  val set : ('a : un) => &('a t) * int * 'a -> 'a t
-  val map : (&'a -> 'b) -> &('a t) -> 'b t
-end
-\end{lstlisting}
-% \begin{lstlisting}
-% (* A copy-on-write implementation *)
-% module Cow : PersistArray = struct
-%   type 'a t = 'a array
-%   ...
-%   let set (a, i0, x0) =
-%     mapi (fun i x -> if i = i0 then x0 else x) a
-% end
-% \end{lstlisting}
-  \vspace{-15pt}
-  % \caption{Implementation using copy-on-write}
-  % \end{subfigure}
-  \caption{Persistent arrays}
-  \label{sig:array}
-  \label{sig:hybarray}
-  \label{ex:cow}
-\end{figure}
-
-\subsection{Imperative programming}
-
-Our API
-aims to safely handle manual allocation and
-deallocation of arrays that may contain affine elements.
-One would first use @create n v@ to create
-an (unrestricted) array of size @n@ initialized with value
-@v@. This value must be unrestricted as it is duplicated to
-initialize all array elements. Using the @map@ function we
-can transform the unrestricted elements into linear (affine)
-ones. This array can be processed further using @map@ (or @set_mut@ if
-affine). The @get@ and @set@ function are only applicable if the element type
-is unrestricted as one element is duplicated. The @length@
-function is always applicable. To @free@ an array the elements must be
-affine.
-
-To manage the different kinds of accessing the array we distinguish between constructors, destructors,
-observers, and mutators.
-Constructors and destructors like @create@ and @free@ manipulate the whole
-array. %As with channels in \cref{sec:session-linearity},
-The constructor
-@create@ yields a linear resource which is consumed by @free@.
-During the lifetime of the array resource @a@, we can split
-off \emph{shared borrows} @&a@ that provide a read-only
-view or \emph{exclusive borrows} @&!a@ for read-write views.
-Observer functions such as @length@, @get@ and @set@ expect a shared borrow
-argument, mutator functions such a @set_mut@ expect
-a exclusive borrow.
-
-Each borrow is tied to a region whose lifetime is properly
-contained in the lifetime of the resource.
-In a region, we can split off as many shared borrows of a resource as
-we like, but we can take only one exclusive borrow. In a
-subsidiary region, we can take shared borrows of any
-borrow or we can take an exclusive borrow of an exclusive borrow from an
-enclosing region. Borrows are confined to their regions. Inside the region,
-shared borrows are unrestricted (@un@) whereas exclusive
-borrows are affine (@aff@).
-
-We can use the API in a mutable fashion
-to create an
-array of Fibonacci numbers though an imperative coding style:
-\begin{lstlisting}
-let mk_fib_array n =
-  let a = create n 1 in
-  for i = 2 to n - 1 do
-    let x = get (&a, i-1) + get (&a, i-2) in(*@\label{ex:array:get}*)
-    set_mut (&!a, i, x)(*@\label{ex:array:set}*)
-  done;
-  a(*@\label{ex:array:return}*)
-# mk_fib_array : int -> int Array.t
-\end{lstlisting}
-
-After
-creation of the array, the presence of a borrow in the for loop
-prevents access to the ``raw'' resource inside the loop's
-body. \cref{ex:array:get} of the example contains two shared borrows
-in the same expression, which forms a region by itself. These borrows
-are split off the exclusive borrow used in \cref{ex:array:set}, which
-belongs to the next enclosing region corresponding to the loop body.
-The whole array can be returned in \cref{ex:array:return} because  the
-borrows are no longer in scope. 
-%
-More precisely, here is is an annotated excerpt with explicit regions
-marked in braces @{| ... |}@:
-
-\begin{lstlisting}[firstnumber=3]
-  for i = 2 to n - 1 do {|
-    let x = {| get &a (i-1) + get &a (i-2) |} in(*@\label{ex:array:region:get}*)
-    set &!a i x
-  |} done;
-\end{lstlisting}
-
-A region consists of the header expression of the @let@ in
-\cref{ex:array:region:get}. It is contained in another region
-spanning the body of the @for@ loop. \lang guarantees that borrows
-never escape the smallest enclosing region. It employs a system of
-\emph{indexed kinds} like @aff_r@ and @un_r@ where
-$r$ is a positive integer that corresponds to the lexical nesting
-depth of regions. For instance, the type of @&!a@ in
-\cref{ex:array:set} has kind @aff_1@ whereas the type of
-@&a@ in \cref{ex:array:region:get} has kind
-@un_2@ and the typing of the inner region is such that types with
-kind indexes greater than or equal to $2$ cannot escape.
-In the example, borrows cannot escape  because they are consumed
-immediately by @get@ and @set_mut@.
-
-\subsection{Solving sudokus with hybrid data-structures}
-
-\begin{figure}[tp]
-  \lstinputlisting[linerange=1-31]{code/sukodu.affe}
-  \vspace{-15pt}
-  \caption{Excerpt of the Sudoku solver}
-  \label{ex:sudoku}
-  \vspace{-10pt}
-\end{figure}
-
-On top of the mutation function @set_mut@, our array implementation
-also provides @set: &('a t) * int * 'a -> 'a t@. It
-takes a shared borrow because it only reads from the
-argument array, but returns a fresh, modified structure.
-%
-This API could be implemented in a wide variety of ways, such as
-copy-on-write (which copies the whole array on each @set_mut@)
-or something more sophisticated\cite{DBLP:conf/ml/ConchonF07}.
-
-We can use our persistent arrays
-to implement a backtracking Sudoku solver.
+In this section, we present
+a safe API for persistent arrays that support both immutable and mutable features,
+and use it to implement a backtracking Sudoku solver.
 % show how \lang allows us to define
@@
 cheap in-place mutations.
+
+\cref{sig:hybarray} contains an API @HybArray@
+that enables using mutable and immutable modifications to
+the board through affine types and borrows.
+The signature differs slightly from the @Array@
+signature. For simplicity, the array elements are unrestricted, but
+the structure itself remains linear so as to be implemented in terms of @Array@.
+The in-place mutation function @set_mut@ with type
+@&!('a t) * int * 'a -> unit@ works on a exclusive borrow  whereas the persistent
+@set@ operation has type @&('a t) * int * 'a -> 'a t@. It
+takes a shared borrow because it only reads from the
+argument array, but returns a fresh,  modified structure.
+The code in \cref{ex:cow} contains a very simple implementation of
+@HybArray@ that represents hybrid arrays
+as regular arrays and uses copy-on-write for persistent
+modifications. The function
+@mapi: (int -> &'a -> 'b) -> &('a t) -> 'b t@
+is a simple variation on @Array.map@ where the mapping function
+takes the position of the element.
+% A more elaborate implementation could use persistent arrays
+% \cite{DBLP:conf/ml/ConchonF07}.
+
+\begin{figure}[tp]
+  \centering
+  % \begin{subfigure}{1\linewidth}
+\begin{lstlisting}
+module type HybArray = sig
+  include ARRAY
+  val set : &('a t) * int * 'a -> 'a t
+  val set_mut : &!('a t) * int * 'a -> unit
+end
+
+module Cow : HybArray = struct
+  include Array
+  let set (a, i0, x0) =
+    Array.mapi (fun i x -> if i = i0 then x0 else x) a
+  let set_mut = Array.set
+end
+\end{lstlisting}
+    \vspace{-15pt}
+  %   \caption{Implementation using copy-on-write}
+  % \end{subfigure}
+  \caption{Hybrid arrays}
+    \label{sig:hybarray}
+    \label{ex:cow}
+
+  \lstinputlisting[linerange=1-31]{code/sukodu.affe}
+    \vspace{-15pt}
+  \caption{Excerpt of the Sudoku solver}
+  \label{ex:sudoku}
+\end{figure}
 
@@
 
-While presented for persistent arrays, this API
+While presented for copy-on-write arrays, the API
 can easily be adapted to other persistent data structures with
@@
 
-\subsection{Iterators and regions}
-\label{tuto:vector}
-
-% We have seen that borrows are limited to a region.
-In the examples so far, regions do not appear in type signatures.
-But for certain programming idioms, we want to extend the scope of a
-region across a function boundary.
-For instance, how should we fold on an array of linear objects?
-A naive fold function would only work on unrestricted content:
-\begin{lstlisting}
-val fold : ('a : un) => ('a -> 'b -> 'b) -> 'a array -> 'b -> 'b
-\end{lstlisting}
-
-However, if we want to work on linear and affine resources,
-we must make sure those resources are not leaked in the result.
-We obtain the following signature:
-\begin{lstlisting}
-val fold :
-  ('b:'k),('k <= aff_r) =>
-  (&(aff_r1,'a) -> 'b -{aff_r1}> 'b) -> &('k_1,'a array) -> 'b -{'k_1}> 'b
-\end{lstlisting}
-
-The folded function receives a shared borrow of the element in the array.
-The typing of the callback ensures
-that this borrow is neither captured nor returned by the function.
-%
-This encapsulation is implemented with a universally quantified \emph{kind index variable} $r$.
-The signature prescribes the type @&(aff_r1,'a)@ for the
-shared borrow of the resource with an affine kind at region nesting $r+1$. The return
-type of the callback is constrained to kind @'k <= aff_r@
-so that the callback certainly cannot return the borrowed argument.
-%
-The input of the fold is a shared borrow of the array,
-which ensures that we have the right to share borrows of the inner content and
-make multiple concurrent folds.
-
-We can then easily use this function to fold over an array of files. Here
-we compute the sum of their sizes:
-\begin{lstlisting}
-let total_size = fold (fun f s -> File.size f + s) &all_files 0
-\end{lstlisting}
-
-Naturally, this is not sufficient if we want to mutate the array during iteration,
-as is often the case with @iter@ functions.
-For this purpose, we need to take an exclusive borrow of the structure to iterate
-on:
-\begin{lstlisting}
-val iter : &!('a array) -> (&!(aff_r1,'a) -{un}> unit) -{aff}> unit
-\end{lstlisting}
-
-While the distinction between mutable and immutable iteration functions
-might seem unfortunate, it is typical of
-programming with borrows and is also present in the Rust standard library.
-It allows the programmer to explicitly state how various iterations
-are allowed to compose and be optimized.
-\lang's region variables ensures that the content iterated on can never
-be leaked outside of the iteration function.
-This pattern is essential in many use-cases of linearity such
-as pools of linear objects (see example in \cref{tuto:pool}).
-
-\lstDeleteShortInline@
-
 %%% Local Variables:

### 2019-11-22 kindly/motivation.tex
Reshuffle the intro again.

@@
 the notion of borrowing \cite{DBLP:conf/popl/BoylandR05} which
-we demonstrate through a series of examples.
+we demonstrate first through the example of linear arrays to demonstrate
+simple borrowing and imperative programming and second with iterators
+on linear values to demonstrate advanced uses of regions.
+Further examples are also available in \cref{sec:extra-example}.
 
-As running theme of our examples, we use an API for persistent arrays.
-Recently introduced persistent data structures
-permit transient mutations where
-non-linear uses lead to degraded performance
-\cite{DBLP:conf/ml/ConchonF07} or to
-dynamic and static checks \cite{DBLP:journals/pacmpl/Puente17}.
-In particular, persistent Hash-Array-Mapped-Tries (HAMT) have been used with similar
-APIs in several non-pure functional languages (OCaml, Clojure, \dots).
-Affine types help formalize the performance contract between the programmer
-and the library, while borrows avoid the need to thread state explicitly,
-as often required by APIs for immutable datatypes.
-
-\cref{sig:hybarray} contains a simple API for Persistent arrays
-with mutable and immutable modifications through affine types and borrows.
-%
-We first present how to use it imperatively as for
-mutable linear arrays. We then see how to exploit its persistent
-properties to write a backtracking sudoku solver. Finally, we consider
-iterators on such linear structures.
-
-% \begin{itemize}
-% \item An API for linear arrays to demonstrate simple borrowing and imperative programming.
-% \item
-%   A backtracking Sudoku solver to demonstrate
-%   \lang's ability to build safe hybrid functional/imperative data-structures.
-% \item A pool of linear objects to demonstrate advanced use of regions.
-% \end{itemize}
+\subsection{Imperative programming with linear arrays}
+\label{sec:imper-progr}
 
@@
   \centering
-  % \begin{subfigure}{1\linewidth}
 \begin{lstlisting}
-module PersistArray : sig
+module Array : sig
   type ('a : 'k) t : lin
@@
   val get : ('a : un) => &('a t) * int -> 'a
-  val set_mut : ('a : aff) => &!('a t) * int * 'a -> unit
-  val set : ('a : un) => &('a t) * int * 'a -> 'a t
+  val set : ('a : aff) => &!('a t) * int * 'a -> unit
   val map : (&'a -> 'b) -> &('a t) -> 'b t
@@
 \end{lstlisting}
-% \begin{lstlisting}
-% (* A copy-on-write implementation *)
-% module Cow : PersistArray = struct
-%   type 'a t = 'a array
-%   ...
-%   let set (a, i0, x0) =
-%     mapi (fun i x -> if i = i0 then x0 else x) a
-% end
-% \end{lstlisting}
   \vspace{-15pt}
-  % \caption{Implementation using copy-on-write}
-  % \end{subfigure}
-  \caption{Persistent arrays}
+  \caption{Linear arrays}
+  \label{ex:array}
   \label{sig:array}
-  \label{sig:hybarray}
-  \label{ex:cow}
 \end{figure}
 
-\subsection{Imperative programming}
-
-Our API
+The API for mutable linear arrays (shown in \cref{sig:array})
 aims to safely handle manual allocation and
@@
 can transform the unrestricted elements into linear (affine)
-ones. This array can be processed further using @map@ (or @set_mut@ if
-affine). The @get@ and @set@ function are only applicable if the element type
+ones. This array can be processed further using @map@ (or @set@ if
+affine). The @get@ function is only applicable if the element type
 is unrestricted as one element is duplicated. The @length@
@@
 view or \emph{exclusive borrows} @&!a@ for read-write views.
-Observer functions such as @length@, @get@ and @set@ expect a shared borrow
-argument, mutator functions such a @set_mut@ expect
+Observer functions such as @length@ and @get@ expect an shared borrow
+argument, mutator functions such a @set@ expect
 a exclusive borrow.
@@
 In the example, borrows cannot escape  because they are consumed
-immediately by @get@ and @set_mut@.
-
-\subsection{Solving sudokus with hybrid data-structures}
-
-\begin{figure}[tp]
-  \lstinputlisting[linerange=1-31]{code/sukodu.affe}
-  \vspace{-15pt}
-  \caption{Excerpt of the Sudoku solver}
-  \label{ex:sudoku}
-  \vspace{-10pt}
-\end{figure}
-
-On top of the mutation function @set_mut@, our array implementation
-also provides @set: &('a t) * int * 'a -> 'a t@. It
-takes a shared borrow because it only reads from the
-argument array, but returns a fresh, modified structure.
-%
-This API could be implemented in a wide variety of ways, such as
-copy-on-write (which copies the whole array on each @set_mut@)
-or something more sophisticated\cite{DBLP:conf/ml/ConchonF07}.
-
-We can use our persistent arrays
-to implement a backtracking Sudoku solver.
-% show how \lang allows us to define
-The solver maintains an array to represent the state of
-the game and uses backtracking when there are several choices to proceed.
-The use of backtracking suggests a persistent data structure for the array.
-However, only changes that correspond to a choice point need to use
-the persistence mechanism, others may be implemented as
-cheap in-place mutations.
-
-Our implementation of a Sudoku solver (\cref{ex:sudoku}) performs
-modifications that correspond to choice points using @set@,
-which makes it trivial to come back to the previous version
-of the array, while other modifications use @set_mut@, which
-forbids uses of the previous version.
-
-The board is represented as a 2D-matrix (\cref{line:boardtype}), where
-the @Matrix@ type uses the same API as @PersistArray@
-but with two indices.
-Each cell contains an integer set that represent admissible solutions so far.
-The Sudoku solver iterates over the cells and tries each possible solution (\cref{line:try_solution}).
-When a value is picked for the current cell, we create a choice point,
-change the cell with an immutable modification (\cref{line:immutset}), and propagate
-the changes with the @propagate@ function.
-The @propagate@ function uses direct mutation through a
-exclusive borrow of the matrix as it need not preserve the previous
-version of the board.
-The implementation of @propagate@ is split into three parts
-for lines, columns, and square, which are all very similar to function
-@propagate_lines@ (\cref{line:propline}).
-
-The typing ensures that the mutations do not compromise the state at
-the choice point, because they operate on a new state @new_g@ created for one
-particular branch of the choice.
-As the @set@ function only requires an unrestricted shared borrow,
-the closure @try_solution@ remains unrestricted even though
-it captures the borrow @&g@.
-The price is that @try_solution@ cannot escape from
-@&g@'s region. In this example, inference places the region
-around the @begin@/@end@ scope.
-Hence, @try_solution@ can be used in
-the iteration in \cref{line:iter}.
-% As @&g@ is unrestricted, we could even use a parallel iteration
-% instead of a sequential one.
-As @g@ is linear we must free it outside of the region
-before returning (\cref{line:free:g}).
-
-While presented for persistent arrays, this API
-can easily be adapted to other persistent data structures with
-transient mutability such as Relaxed-Radix Balance Vectors (RRB) \cite{DBLP:journals/pacmpl/Puente17}
-or persistent HAMTs \cite{bagwell2001ideal,clojurehamt} to provide  a
-convenient programming style without compromising performance.
-
-% \subsection{Type abstraction and linearity violations}
-
-% While linearity allows to verify many properties, it also prevents
-% the implementation of functions that are nevertheless safe. Rust addresses this
-% issue through the @unsafe@ construct, which allows to locally
-% ignore linearity constraints. This is essential to build basic
-% functions such as splitting strings.
-% %
-% \lang allows to partially emulate this behavior using type abstraction
-% as provided by ML modules.
-% In particular, data types can be internally defined as unrestricted but exposed
-% as affine.
-
-% \TODO{Show example with arrays}
+immediately by @get@ and @set@.
 
@@
 we must make sure those resources are not leaked in the result.
-We obtain the following signature:
+content. We obtain the following signature:
 \begin{lstlisting}
@@
 
-Naturally, this is not sufficient if we want to mutate the array during iteration,
-as is often the case with @iter@ functions.
+Naturally, this is not sufficient if we want to mutate the array during iteration.
 For this purpose, we need to take an exclusive borrow of the structure to iterate
@@
 \begin{lstlisting}
-val iter : &!('a array) -> (&!(aff_r1,'a) -{un}> unit) -{aff}> unit
+val iter_mut : &!('a array) -> (&!(aff_r1,'a) -{un}> unit) -{aff}> unit
 \end{lstlisting}
 
-While the distinction between mutable and immutable iteration functions
+Since the callback will be called multiple times, it is explicitly
+annotated with @un@.
+
+While the distinction between mutable and immutable iteration function
 might seem unfortunate, it is typical of
@@
 This pattern is essential in many use-cases of linearity such
-as pools of linear objects (see example in \cref{tuto:pool}).
+as pools of linear objects (see \cref{tuto:pool}).
 

### 2019-11-22 kindly/infer/infer.tex
More shaving.

@@
 the constraints.
-\begin{itemize}[leftmargin=*]
+\begin{itemize}[leftmargin=*,topsep=0pt]
 \item $\bsplit{C}{\Sv}{\inP{\Sv_1}}{\inP{\Sv_2}}$.
@@
 a derivation that gives a type at least as general.
-For this, we need first to provide a few additional definitions.
+% For this, we need first to provide a few additional definitions.
 
@@
 
-We can then state principality.
+% We can then state principality.
 
@@
 \begin{theorem}[Principality]
-  Let $\inferS{\Ctrue}{\E}{e}{\schm}$ a closed typing judgement.\\
+  Let $\inferS{\Ctrue}{\E}{e}{\schm}$ a closed typing judgement.
   Then $\inferW{\Sv}{(C,\unif)}{\Eflat\E}{e}{\tau}$

### 2019-11-22 kindly/soa.tex
More shaving.

@@
 
-\subsection{Substructural type-systems in functional languages}
+\paragraph{Substructural type-systems in functional languages}
 
@@
 modules by allowing to keep some type unrestricted inside a module, but
-exposing it as affine. We support a similar programming style, which
-we showcase in \cref{motivation}.
+exposing it as affine. \lang supports
+such programming style thanks to subsumption.
 
@@
 This design is easy to retrofit on top of an existing compiler
-such as GHC, but there are quite controversial discussions about it\footnote{
-  See the in-depth discussion attached to the GHC proposal starting here: \url{https://github.com/ghc-proposals/ghc-proposals/pull/111\#issuecomment-403349707}}.
+such as GHC, but has proven quite controversial\footnote{
+  See the in-depth discussion attached to the GHC proposal starting  \href{https://github.com/ghc-proposals/ghc-proposals/pull/111\#issuecomment-403349707}{here}.}.
 Most relevant to \lang:
@@
 
-\subsection{Other substructural type-systems}
+\paragraph{Other substructural type-systems}
 
@@
 
-\subsection{Type-system features}
+\paragraph{Type-system features}
 %
@@
 could be applied to larger and more complex languages such as OCaml.
-Alternatively, we could have based \lang on qualified
-types~\cite{DBLP:journals/scp/Jones94}, similarly to Quill.
+% Alternatively, we could have based \lang on qualified
+% types~\cite{DBLP:journals/scp/Jones94}, similarly to Quill.
 % This
@@
 with constraints~\citep{DBLP:journals/tapos/OderskySW99,DBLP:conf/sas/TrifonovS96}.
-In particular, Flowcaml~\citep{DBLP:conf/popl/PottierS02}
-extends OCaml with security levels forming a lattice and supports type inference.
-In \lang, we repurpose these ideas to check for linearity and affinity.
-\citet{DBLP:conf/aplas/Simonet03} also presents various algorithmic
-notions to solve lattice-based constraints, which we partially use
-in our constraint solving algorithm. Our main contribution
-is the design of a kind language specialized for checking linearity
-and sufficiently simple to make
+% In particular, Flowcaml~\citep{DBLP:conf/popl/PottierS02}
+% extends OCaml with security levels forming a lattice and supports type inference.
+% In \lang, we repurpose these ideas to check for linearity and affinity.
+It also admits classical simplification rules
+\citep{DBLP:conf/aplas/Simonet03,DBLP:conf/popl/PottierS02} which we partially use
+in our constraint solving algorithm.
+\affe's novelty is a kind language
+sufficiently simple to make
 all simplification rules complete, which allows us to keep type signatures simple.

### 2019-11-22 kindly/lang.tex
fixed lstlisting bugs by using 'keepspaces' option for all \lstMakeShortInline commands.
Shaved section 3. It fits!

--- 2019-11-22 kindly/lang.tex
@@
 \item Pattern matching is demonstrated on pairs rather than algebraic
-  datatypes.
+  datatypes, and is covered in \cref{appendix:pairs}.
 \item There are separate operators for borrowing and reborrowing (taking
@@
   \input{grammar}
+  \vspace{-5pt}
   \caption{Syntax}
@@
 %
-A match is indexed with a \emph{match specification} $\etransfm$ that indicates
-whether the match should work on borrows ($\etransfm=\&^b$) or not ($\etransfm=id$).
+% A match is indexed with a \emph{match specification} $\etransfm$ that indicates
+% whether the match should work on borrows ($\etransfm=\&^b$) or not ($\etransfm=id$).
 %
@@
 
-Many types are indexed with kinds, ranged over by $k$.
-A kind is either a kind variable $\kvar$ or a constant
+Many types are indexed with kinds.
+A kind $k$ is either a kind variable $\kvar$ or a constant
 (linear $\klin$, affine $\kaff$, or unrestricted $\kun$) indexed
@@
 % give a quick presentation of the kinds and constraints.
-A kind $k$ is either a kind variable, $\kvar$, or a constant.
-Each constant is composed of a ``quality'' and a ``level''.
-The quality describes the use pattern of the type:
+A kind $k$ is either a kind variable, $\kvar$, or a constant $\mul_n$.
+The quality $\mul$ describes the use pattern of the type:
 unrestricted $\kun$, affine $\kaff$, or linear $\klin$. The level $n \in \Nat \cup \{\infty\}$
@@
 to typecheck the components of the pair.
-The constraint $C$ characterizes a valid  splitting.
+% The constraint $C$ characterizes a valid  splitting.
 
@@
 levels $n$ and higher.
-With that setup we typecheck the body of the region,
-$f(\borrow[\IBORROW]{c})$.
-The constraint $\Cleq{\tau}{\klin_{n-1}}$ ensures
-that the kind of $\tau$ does not contain anything whose kind is
-``above level $n$'' and thus confines the borrow to the region
-(for instance, $\borrowty[\IBORROW]{\kun_n}{\text{c}}$ has kind
-$\kun_n$ and thus cannot be returned).
+%
+% With that setup we typecheck the body of the region,
+% $f(\borrow[\IBORROW]{c})$.
+% The constraint $\Cleq{\tau}{\klin_{n-1}}$ ensures
+% that the kind of $\tau$ does not contain anything whose kind is
+% ``above level $n$'' and thus confines the borrow to the region
+% (for instance, $\borrowty[\IBORROW]{\kun_n}{\text{c}}$ has kind
+% $\kun_n$ and thus cannot be returned).
+%
+The \TirName{Region} rule also imposes the constraint
+$\Cleq{\tau}{\klin_{n-1}}$, which prevents the borrow, having kind $k$ of level
+$\geq n$, from escaping the region's body of type $\tau$.
 
@@
 
-\lstMakeShortInline[style=rule,basicstyle=\normalsize\normalfont]@
+\lstMakeShortInline[keepspaces,style=rule,basicstyle=\normalsize\normalfont]@
 
@@
 % updated (through affine borrows).
-
+%
 The semantics is defined in terms of \emph{elaborated expressions} $e$
@@
 
-Let-Polymorphism is demonstrated for functions giving rise to elaborated
+Let-polymorphism is demonstrated for functions giving rise to elaborated
 $\eletfun$ expressions annotated with a type scheme $\schm$ and a kind $k$ indicating their
 usage restriction (linear, affine, etc) relative to the variables and
-constraints from $\schm$. Their use
-gives rise to explicit instantiation of the kind and type variables. Pairs
-come with a kind tag $k$ indicating the usage restriction.
+constraints of $\schm$. Their use
+gives rise to explicit instantiation of the kind and type variables.
+% Pairs come with a kind tag $k$ indicating the usage restriction.
 
@@
 type 'a sem = Error of string | TimeOut | Ok of 'a
-val eval : store -> perm -> venv -> int -> exp -> (store * perm * result) sem
+val eval: store->perm->venv->int->exp->(store*perm*result) sem
 \end{lstlisting}

### 2019-11-22 kindly/lang.tex
Fixed typos.

@@
 \end{enumerate}
-We then start from the leafs and grow the region
+We then start from the leaves and grow the region
 associated to each borrow until it can't grow anymore either due to a binding
@@
 $a$ only has one borrow, its region covers its whole lexical scope.
-$x$ has multiple conflicting borrow and requires more consideration.
+$x$ has multiple conflicting borrows and requires more consideration.
 We place a first
@@
 error.
-Notes that indices of regions are well nested, as the inner region have
-higher indices that the outer ones.
+Note that indices of regions are well nested, as the inner region have
+higher indices than the outer ones.
 

### 2019-11-22 kindly/main.tex
conclusion
Add authors.
Adapt to the PLDI format.
Start improving the inference section.
moved session type example
formatting
Made a new pass on input parameters.
Gather the selected inference rules.
checkpoint - restructuring needed
Replace the pool example by an iterator one. Should be shorter.
Misc.
Metatheory: Introduced macro for the soundness theorem and moved it into 'metatheory-common.tex', as we want to state the theorem in both 'metatheory.tex' and 'metatheory-extra.tex'.
Reshuffle the intro again.
Metatheory: Used 'restatable' and removed metatheory-common again.
Fixed typos.

@@
-\documentclass[acmsmall,review,anonymous,table]{acmart}
+\PassOptionsToPackage{svgnames}{xcolor}
+\documentclass[sigplan,10pt,review,anonymous]{acmart}
 \settopmatter{printfolios=true,printccs=false,printacmref=false}
@@
 \usepackage{amsthm}
+\usepackage{thm-restate}
 % \usepackage{mathrsfs}
@@
 \usepackage{subcaption}
-\usepackage{enumitem}
-%\setlist{leftmargin=*,noitemsep,topsep=5pt}
+\usepackage[inline,shortlabels]{enumitem}
+\setlist{leftmargin=*,noitemsep}
+\usepackage{array}
+\usepackage{bm}%Decent bolding for math symbols
 
@@
 \usepackage{mathpartir} % Syntax trees
+
+\mprset{sep=1.2em}
 
@@
 \bibliographystyle{ACM-Reference-Format}
-\citestyle{acmauthoryear}
+
+\author{Gabriel Radanne}
+\affiliation{
+  \institution{University of Freiburg}
+  \country{Germany}
+}
+\email{radanne@informatik.uni-freiburg.de}
+
+\author{Hannes Saffrich}
+\affiliation{
+  \institution{University of Freiburg}
+  \country{Germany}
+}
+\email{saffrich@informatik.uni-freiburg.de}
+
+\author{Peter Thiemann}
+\affiliation{
+  \institution{University of Freiburg}
+  \country{Germany}
+}
+\email{thiemann@informatik.uni-freiburg.de}
 
@@
   manages linearity and affinity properties through kinds. In addition
-  \lang{} supports the exclusive and unrestricted borrowing of affine
+  \lang{} supports the exclusive and shared borrowing of affine
   resources, inspired by features of Rust.
   Moreover, \lang{} retains the defining features of the ML family:
-  an impure, strict functional expression language with complete principal type
+  an impure, strict, functional expression language with complete principal type
   inference and type abstraction through modules.
@@
 \input{soa}
+\input{conclusion}
 
@@
 \appendix
+
+\section{Further Examples}
+\label{sec:extra-example}
+\lstMakeShortInline[keepspaces,basicstyle=\small\ttfamily]@
+\input{motivation-linearity}
+\input{motivation-persistent}
+\input{motivation-pool}
+\lstDeleteShortInline@
+
 \input{sdtyping/extra}

### 2019-11-22 kindly/motivation-persistent.tex
Fixed typos.

@@
 
-\cref{sig:hybarray} contains an API @HybArray@
+\cref{sig:hybarray} contains an API @HYBARRAY@
 that enables using mutable and immutable modifications to
@@
 The code in \cref{ex:cow} contains a very simple implementation of
-@HybArray@ that represents hybrid arrays
+@HYBARRAY@ that represents hybrid arrays
 as regular arrays and uses copy-on-write for persistent
@@
 \begin{lstlisting}
-module type HybArray = sig
+module type HYBARRAY = sig
   include ARRAY
@@
 
-module Cow : HybArray = struct
+module Cow : HYBARRAY = struct
   include Array

### 2019-11-23 kindly/lang.tex
Make sure there are no reference to pairs in the body for now.

@@
 \lang is a minimal ML-like language with
-let-polymorphism, abstract types, and pairs. Its type system manages
+let-polymorphism and abstract types. Its type system manages
 linearity and borrows using kinds and (kind-) constrained types.
@@
 \begin{itemize}
-\item Pattern matching is demonstrated on pairs rather than algebraic
-  datatypes, and is covered in \cref{appendix:pairs}.
+% \item Pattern matching is demonstrated on pairs rather than algebraic
+%   datatypes, and is covered in \cref{appendix:pairs}.
 \item There are separate operators for borrowing and reborrowing (taking
@@
 type environment $\E$ in two parts, $\E_1$ and $\E_2$, which are used
-to typecheck the components of the pair.
+to typecheck the components of the application.
 % The constraint $C$ characterizes a valid  splitting.
@@
 elaborated abstraction. A \emph{closure} represents a monomorphic
-function in the usual way. A \emph{pair}
-consists of a kind and two results. A \emph{resource} contains
+function in the usual way.
+% A \emph{pair}
+% consists of a kind and two results.
+A \emph{resource} contains
 a result and the \emph{hole} $\StFreed$ fills a released location.

### 2019-11-23 kindly/split.sh
Update split script.

@@ -1,3 +1,2 @@
-pdfjam -o paper.pdf main.pdf '1-27'
-pdfjam -o appendix.pdf main.pdf '28-'
-
+qpdf --pages main.pdf 1-14 -- main.pdf paper.pdf
+qpdf --pages main.pdf 15-z -- main.pdf appendix.pdf

### 2019-11-23 kindly/sdtyping/extra.tex
Move explanation of Let in the appendix.
Added typing for internal language.
Make sure splitting rules are all the same.
Added typing for internal language in A-normal form.
Partially fixed layout of appendix.

@@
 
-\subsection{Kinding}
-%
 \begin{figure}[bt]
@@
 \end{figure}
+
+\begin{figure}[!tbp]
+  \input{sdtyping/sd-splitting}
+  \caption{Splitting ---
+    environments $\lsplit C\E{\E_l}{\E_r}$;
+    inference $\bsplit C\E{\inP{\E_l}}{\inP{\E_r}}$;
+    binders $\bsplit Cb{\inP{b_r}}{\inP{b_l}}$}
+  \label{fig:sd-splitting}
+\end{figure}
+\begin{figure*}[!tbp]
+  \input{sdtyping/sd-borrowing}
+  \caption{Borrowing ---
+    environments $\lregion[n]{C}{x}{\E}{\E'}$;
+    inference $\bregion[\inP n]{C}{\inP x}{\E}{\inP{\E'}}$;
+    binders $\bregion[\inP n]{C}{\inP x}{b}{\inP{b'}}$}
+  \label{fig:sd-borrowing}
+\end{figure*}
+\begin{figure*}[!tbp]
+  \input{sdtyping/sd-environments}
+  \caption{Rewriting constraints on environments --- $\inP{\Cleq{\E}{k}}\Crewrite  C$}
+  \label{fig:contraints-environments-types}
+\end{figure*}
+\begin{figure*}[!tbp]
+  \input{sdtyping/sd-typing}
+  \caption{Syntax-directed typing rules --
+    $\inferS{C}{\E}{e}{\tau}$}
+  \label{fig:syntax-directed-typing}
+\end{figure*}
+\begin{figure*}[!btp]
+  \input{sdtyping/sd-typing-internal}
+  \caption{Syntax-directed typing rules for internal language --
+    $\inferS{C}{\E}{e}{\tau}$}
+  \label{fig:syntax-directed-typing-internal}
+\end{figure*}
+\begin{figure*}[!btp]
+  \input{sdtyping/sd-typing-internal-a}
+  \caption{Syntax-directed typing rules for internal language in A-normal form --
+    $\inferS{C}{\E}{e}{\tau}$}
+  \label{fig:syntax-directed-typing-internal-a}
+\end{figure*}
+
+\subsection{Kinding}
 
@@
 
-\begin{figure*}[!tb]
-  \input{sdtyping/sd-splitting}
-  \caption{Splitting ---
-    environments $\lsplit C\E{\E_l}{\E_r}$;
-    inference $\bsplit C\E{\inP{\E_l}}{\inP{\E_r}}$;
-    binders $\bsplit Cb{\inP{b_r}}{\inP{b_l}}$}
-  \label{fig:sd-splitting}
-  \input{sdtyping/sd-borrowing}
-  \caption{Borrowing ---
-    environments $\lregion[n]{C}{x}{\E}{\E'}$;
-    inference $\bregion[\inP n]{C}{\inP x}{\E}{\inP{\E'}}$;
-    binders $\bregion[\inP n]{C}{\inP x}{b}{\inP{b'}}$}
-  \label{fig:sd-borrowing}
-  \input{sdtyping/sd-environments}
-  \caption{Rewriting constraints on environments --- $\inP{\Cleq{\E}{k}}\Crewrite  C$}
-  \label{fig:contraints-environments-types}
-\end{figure*}
-
 \subsection{Typing}
@@
 $\Multi{\kvar},\Multi{\tvar} = (\fv{\tau}\cup\fv{C})\setminus\fv{\E}$.
-
-\begin{figure*}[hbtp]
-  \input{sdtyping/sd-typing}
-  \caption{Syntax-directed typing rules --
-    $\inferS{C}{\E}{e}{\tau}$}
-  \label{fig:syntax-directed-typing}
-\end{figure*}
+The typing rules specific to the internal language are shown in
+\cref{fig:syntax-directed-typing-internal}.
+The typing rules specific to the internal language in A-normal form are shown in
+\cref{fig:syntax-directed-typing-internal-a}.
 

### 2019-11-23 kindly/lang.tex
Pairs are back!

@@
 %
-% A match is indexed with a \emph{match specification} $\etransfm$ that indicates
-% whether the match should work on borrows ($\etransfm=\&^b$) or not ($\etransfm=id$).
+A match is indexed with a \emph{match specification} $\etransfm$ that indicates
+whether the match should work on borrows ($\etransfm=\&^b$) or not ($\etransfm=id$).
 %
@@
 expression has multiple subexpressions:  function applications,
-and let bindings. In the
+let bindings and pairs. In the
 expression
@@
 gives rise to explicit instantiation of the kind and type variables.
-% Pairs come with a kind tag $k$ indicating the usage restriction.
+Pairs come with a kind tag $k$ indicating the usage restriction.
 

### 2019-11-23 kindly/lang.tex
And the § on pattern matching!

@@
 The rest of this section formalizes  \lang: the syntax (\cref{syntax}),
-the statics in terms of syntax-directed typing (\cref{sdtyping}),
+the statics in terms of a region annotation pass (\cref{regionannot}) and
+syntax-directed typing (\cref{sdtyping}),
 and a dynamics that is linearity- and resource-aware (\cref{sec:sem}).
@@
   \input{grammar}
-  \vspace{-5pt}
+  \vspace{-7pt}
   \caption{Syntax}
@@
 
-In this section, we focus on the essential and novel parts of the type system. A complete
+We now focus on the essential and novel parts of the type system. A complete
 description is available in \cref{appendix:sdtyping}.
@@
 %
-We discuss judgments for typing, kinding, and constraint entailment:
+We discuss the following judgments:
 \begin{description}
@@
   \input{infer/lattice}
-  \vspace{-5pt}
+  \vspace{-10pt}
   \caption{Lattice ordering -- $k \lk_\Lat k'$}
@@
 
-The {\sc Var} rule looks up the type scheme of the variable $x$ in
+Rule {\sc Var} looks up the type scheme of the variable $x$ in
 the environment $\E$
@@
 
-The {\sc Abs} rule ensures that the kind annotation on the arrow type
-($\tau_2 \tarr{k} \tau_1$) reflects the restrictions on the captured variables.
-To do so, the rule imposes the constraint $\Cleq\E k$.
+Rule {\sc Abs} ensures the kind annotation on the arrow type
+($\tau_2 \tarr{k} \tau_1$) reflects the restrictions on captured variables
+via the constraint $\Cleq\E k$.
 If, for instance, any binding in $\E$ is affine, it gives rise to the
-constraint $\Cleq{\kaff_n}{k}$, so that  the arrow kind is at least
+constraint $\Cleq{\kaff_n}{k}$ and the arrow kind is at least
 affine at nesting level $n$.
@@
 
-% \paragraph{Pattern matching}
-% \label{sdtyping:matching}
-
-% Elimination of pairs is done using a matching construct
-% $(\matchin{x,x'}{e_1}{e_2})$ whose
-% typing rule is shown in \cref{selectrules:matching}.
-% This construction can eliminate both normal pairs, but also borrows
-% of pairs.
-% A syntactic marker $\etransfm$ indicates if it applies to
-% a pair ($\etransfm = \operatorname{id}$) or a borrow ($\etransfm = \&^\BORROW$).
-% If $\etransfm = \operatorname{id}$, the typing simplifies to
-% a regular destruct on pairs.
-% Otherwise, $e_1$ is expected to be of type
-% $\borrowty{k}{\tyPair{\tau_1}{\tau'_1}}$
-% while $x$ and $x'$ are of type respectively
-% $\borrowty{k}{\tau_1}$ and $\borrowty{k}{\tau'_1}$.
-
-% Matching is voluntarily simple here and
-% could be improved using a richer pattern language
-% to access inner components and a typeclass mechanisms
-% to avoid the need for explicit markers for borrow matches.
-% Both extensions are compatible with our system and have been partially
-% implemented in our prototype type-checker.
-
 \paragraph{Regions}
@@
 
-The typing for a region (see the {\sc Region} rule in \cref{selectrules:region})
+The typing for a region (rule {\sc Region} in \cref{selectrules:region})
 replaces suspended bindings by borrow bindings
-(\cref{env:rule:borrow}), typechecks the body
-of the region, and finally ensures that the borrow does not leak outside.
+(rule {\sc BorrowBinding}), typechecks the body
+of the region, and ensures that the borrow does not leak outside.
 This last check is done with indices that correspond to the nesting
@@
 corresponding to its region, thanks to the constraint $(\BORROW_n\lk
-k)$, and the constraint $\Cleq{\tau}{\klin_{n-1}}$ ensures that
+k)$. The constraint $\Cleq{\tau}{\klin_{n-1}}$ ensures that
 the return type of the region must live at some enclosing, lower level.
@@
 %
-The \TirName{Region} rule also imposes the constraint
+Rule {\sc Region} also imposes the constraint
 $\Cleq{\tau}{\klin_{n-1}}$, which prevents the borrow, having kind $k$ of level
@@
 
+\paragraph{Pattern matching}
+\label{sdtyping:matching}
+
+Elimination of pairs is done using a matching construct
+$(\matchin{x,x'}{e_1}{e_2})$.
+This construct is mostly standard, except it can operate both
+on normal pairs and borrows of pairs.
+The intuition is as follow:
+A syntactic marker $\etransfm$ indicates if it applies to
+a pair ($\etransfm = \operatorname{id}$) or a borrow ($\etransfm = \&^\BORROW$).
+If $\etransfm = \operatorname{id}$, the typing simplifies to
+a regular destruct on pairs.
+Otherwise, $e_1$ is expected to be of type
+$\borrowty{k}{\tyPair{\tau_1}{\tau'_1}}$
+while $x$ and $x'$ are of type respectively
+$\borrowty{k}{\tau_1}$ and $\borrowty{k}{\tau'_1}$.
+
+% Matching is voluntarily simple here and
+% could be improved using a richer pattern language
+% to access inner components and a typeclass mechanisms
+% to avoid the need for explicit markers for borrow matches.
+% Both extensions are compatible with our system and have been partially
+% implemented in our prototype type-checker.
+
 \paragraph{Resource management}
@@
 type $\tapp{\tres}{\tau}$ whose content of type $\tau$ must be $\kun_0$ and
-equipped with four operations:
-\begin{itemize}
+equipped with the four class of operations describe in \cref{motivation}:
+\begin{itemize}[topsep=0pt]
 \item
@@
 \qual{\Cleq {\kvar_\tvar} {\kun_0}}
-{\tvar \tarr{} \tapp\tres\tvar}$\\
-Creates a new resource.
-
+{\tvar \tarr{} \tapp\tres\tvar}$
+% Creates a new resource.
 \item
@@
 \qual{\Cleq {\kvar_\tvar} {\kun_0}}
-{\borrowty[\IBORROW]{\kvar}{\tapp\tres\tvar} \tarr{} \tvar}$\\
-Reads from a shared borrow of a resource.
-
+{\borrowty[\IBORROW]{\kvar}{\tapp\tres\tvar} \tarr{} \tvar}$
+% Reads from a shared borrow of a resource.
 \item
@@
 \qual{\Cleq {\kvar_\tvar} {\kun_0}}
-\borrowty[\MBORROW]{\kvar}{\tapp\tres\tvar} \tarr{} \tvar \tarr{\kaff} \tunit$\\
-Writes to an exclusive borrow of a resource.
-
+\borrowty[\MBORROW]{\kvar}{\tapp\tres\tvar} \tarr{} \tvar \tarr{\kaff} \tunit$
+% Writes to an exclusive borrow of a resource.
 \item
 $\destroy$:
-$\forall{\kvar\kvar_\tvar}\bvar{\tvar}{\kvar_\tvar}.\
+$\forall{\kvar_\tvar}\bvar{\tvar}{\kvar_\tvar}.\
 \qual{\Cleq {\kvar_\tvar} {\kun_0}}
-\tapp\tres\tvar \tarr{} \tunit$\\
-Destroys a resource.
+\tapp\tres\tvar \tarr{} \tunit$
+% Destroys a resource.
 \end{itemize}

### 2019-11-23 kindly/main.tex
Many misc changes.

@@
   an impure, strict, functional expression language with complete principal type
-  inference and type abstraction through modules.
-  Our language does not require any linearity annotations in
+  inference and type abstraction.
+  \lang does not require any linearity annotations in
   expressions and supports common functional programming idioms.

### 2019-11-23 kindly/sdtyping/extra.tex
Many misc changes.

@@
 
-In this section, we define our automatic region annotation
+We now define our automatic region annotation
 which is presented in \cref{regionannot}.
@@
 %
-Following HM(X), we define a cylindric constraint system as
+Following \hmx, we define a cylindric constraint system as
 a system which respects the following properties for any variable $x$:
@@
 We now prove that $\CL$ supports all the properties necessary for
-principal type inference, as defined by HM(X).
+principal type inference, as defined by \hmx.
 We first prove that constraint solving
@@
 Finally, we can conclude with all the properties we need for
-HM(X):
+\hmx:
 
@@
 This is sufficient to show that $HM(\CL)$ is principal. However,
-we do not use HM(X) directly but an extended version with kind inference,
+we do not use \hmx directly but an extended version with kind inference,
 linear and affine types, and borrow.
-We extend the proofs of HM(X) to such a system in \cref{appendix:infer}.
+We extend the proofs of \hmx to such a system in \cref{appendix:infer}.
 
@@
 \end{figure}
-\begin{figure*}[!tbp]
+\begin{figure}[!tbp]
   \input{sdtyping/sd-borrowing}
@@
   \label{fig:sd-borrowing}
-\end{figure*}
-\begin{figure*}[!tbp]
+\end{figure}
+\begin{figure*}[tbp]
   \input{sdtyping/sd-environments}
@@
   \label{fig:contraints-environments-types}
-\end{figure*}
-\begin{figure*}[!tbp]
+% \end{figure*}
+% \begin{figure*}[!tbp]
   \input{sdtyping/sd-typing}
@@
 \end{figure*}
-\begin{figure*}[!btp]
-  \input{sdtyping/sd-typing-internal-a}
-  \caption{Syntax-directed typing rules for internal language in A-normal form --
-    $\inferS{C}{\E}{e}{\tau}$}
-  \label{fig:syntax-directed-typing-internal-a}
-\end{figure*}
+% \begin{figure*}[!btp]
+%   \input{sdtyping/sd-typing-internal-a}
+%   \caption{Syntax-directed typing rules for internal language in A-normal form --
+%     $\inferS{C}{\E}{e}{\tau}$}
+%   \label{fig:syntax-directed-typing-internal-a}
+% \end{figure*}
 
@@
 \cref{fig:syntax-directed-typing-internal}.
-The typing rules specific to the internal language in A-normal form are shown in
-\cref{fig:syntax-directed-typing-internal-a}.
+% The typing rules specific to the internal language in A-normal form are shown in
+% \cref{fig:syntax-directed-typing-internal-a}.
 

### 2019-11-23 kindly/infer/infer.tex
Make a pass over the inference section
Partially fixed layout of appendix.

@@
 We now formulate our inference algorithm
-based on the $HM(X)$ framework~\citep{DBLP:journals/tapos/OderskySW99},
+based on the \hmx framework~\citep{DBLP:journals/tapos/OderskySW99},
 a Hindley-Milner type system for a language
@@
 Most importantly,
-$HM(X)$ ensures principal type inference
+\hmx ensures principal type inference
 when $X$ respects the specified properties.
-We apply HM(X) to a concrete constraint language which we name $\CL$.
-We adapt and extend HM(X)'s existing rules to support kind inference,
+We apply \hmx to a concrete constraint language which we name $\CL$.
+We adapt and extend \hmx's existing rules to support kind inference,
 track linearity and handle borrows and regions. Finally, we
@@
 
-The HM(X) framework assumes the existence of a ``$\operatorname{normalize}$''
+The \hmx framework assumes the existence of a ``$\operatorname{normalize}$''
 function which takes a constraint $\inP{C}$, a substitution $\inP\psi$ and returns a
@@
 Our algorithm respect the properties
-of HM(X), computes principal normal forms
+of \hmx, computes principal normal forms
 and simplifies constraints significantly.
@@
 % are sufficient to state that $HM(\CL)$ provides principal type inference.
-% However, we use an extension of HM(X) with kind inference
+% However, we use an extension of \hmx with kind inference
 % and affine types and borrows.
-% We now show that HM(X)'s completeness and soundness theorems are still valid on
+% We now show that \hmx's completeness and soundness theorems are still valid on
 % this extended system.
@@
 We now state that our extended inference algorithm still satisfies the soundness
-and completeness properties of HM(X).
+and completeness properties of \hmx.
 % This is achieved by extending
-% the original proofs from HM(X), which is done in \cref{appendix:infer}.
+% the original proofs from \hmx, which is done in \cref{appendix:infer}.
 %

### 2019-11-23 kindly/motivation.tex
More shaving.
Motivation: Fixed vspace before listing.
Fixed typos.

@@
   \label{sig:array}
+  \vspace{-10pt}
 \end{figure}
@@
 argument, mutator functions such a @set@ expect
-a exclusive borrow.
+an exclusive borrow.
 
@@
 
-We can use the API in a mutable fashion
+We can use the API 
 to create an
-array of Fibonacci numbers though an imperative coding style:
+array of Fibonacci numbers following an imperative coding style:
 \begin{lstlisting}
@@
     let x = get (&a, i-1) + get (&a, i-2) in(*@\label{ex:array:get}*)
-    set_mut (&!a, i, x)(*@\label{ex:array:set}*)
+    set (&!a, i, x)(*@\label{ex:array:set}*)
   done;
@@
 prevents access to the ``raw'' resource inside the loop's
-body. \cref{ex:array:get} of the example contains two shared borrows
-in the same expression, which forms a region by itself. These borrows
-are split off the exclusive borrow used in \cref{ex:array:set}, which
+body. \cref{ex:array:get} contains two shared borrows
+in the same expression which forms a region by itself. These borrows
+are split off the exclusive borrow used in \cref{ex:array:set} which
 belongs to the next enclosing region corresponding to the loop body.
@@
 marked in braces @{| ... |}@:
-
 \begin{lstlisting}[firstnumber=3]
@@
 
-However, if we want to work on linear and affine resources,
+If we want to work on linear and affine resources,
 we must make sure those resources are not leaked in the result.
@@
 
-While the distinction between mutable and immutable iteration function
+While the distinction between mutable and immutable iteration functions
 might seem unfortunate, it is typical of

### 2019-11-23 kindly/soa.tex
Add some notes about pattern matching in the conclusion.

@@
 Most relevant to \lang:
-\begin{itemize}[leftmargin=*]
+\begin{itemize}[leftmargin=*,topsep=0pt]
 \item LH does not admit subtyping for arrows and requires
@@
 to introduce the kind inequalities required for linear types.
-HM(X)~\citep{DBLP:journals/tapos/OderskySW99} 
+\hmx~\citep{DBLP:journals/tapos/OderskySW99} 
 allows us to use constrained types in an ML-like language with complete
 type inference.
-HM(X) has been shown to be compatible with subtyping,
+\hmx has been shown to be compatible with subtyping,
 bounded quantification and existentials~\citep{DBLP:conf/icfp/Simonet03},

### 2019-11-23 kindly/split-jam.sh
another splitter

@@ -1,0 +1,2 @@
+pdfjam -o paper.pdf main.pdf '1-14'
+pdfjam -o appendix.pdf main.pdf '15-'

### 2019-11-27 kindly/infer/extra.tex
Make sure splitting rules are all the same.
Partially fixed layout of appendix.
Small layout improvements for the appendix.

@@
 \label{appendix:infer}
+
+\begin{figure*}[tbp]
+  \centering
+  \input{infer/kinds}
+  \caption{Kind inference rules -- $\inferK{(C,\unif)}{{\inP\E}}{{\inP\tau}}{k}$}
+  \label{rules:kinding}
+% \end{figure*}
+
+% \begin{figure*}[!tbp]
+  \input{infer/typing}
+  \caption{Type inference rules --
+    $\inferW{\Sigma}{(C,\psi)}{{\inP\E}}{{\inP e}}{\tau}$ }
+  \label{rules:typing:full}
+\end{figure*}
 
@@
 The various theorems and their proofs are direct adaptations
-of the equivalent statements for HM(X) \citep{sulzmann1997proofs}.
+of the equivalent statements for \hmx \citep{sulzmann1997proofs}.
 
@@
 \label{rules:kinding:full}
-%
-\begin{figure}[tb]
-  \centering
-  \input{infer/kinds}
-  \caption{Kind inference rules -- $\inferK{(C,\unif)}{{\inP\E}}{{\inP\tau}}{k}$}
-  \label{rules:kinding}
-\end{figure}
-%
+
 We write $\inferK{(C,\unif)}{{\inP\E}}{{\inP\tau}}{k}$ when type $\tau$ has kind $k$
@@
 on let-bindings.
-Following HM(X), we write $(C_\schm, \schm) = \generalize{C}{\E}{\tau}$
+Following \hmx, we write $(C_\schm, \schm) = \generalize{C}{\E}{\tau}$
 for the pair of a constraint and a scheme resulting from
@@
 on the projection operators in HM inference.
-
-\begin{figure*}[!btp]
-  \input{infer/typing}
-  \caption{Type inference rules --
-    $\inferW{\Sigma}{(C,\psi)}{{\inP\E}}{{\inP e}}{\tau}$ }
-  \label{rules:typing:full}
-\end{figure*}
 

### 2019-11-27 kindly/lang.tex
one more pair note.

@@
 For ease of presentation, we simplify the language.
-\begin{itemize}
-% \item Pattern matching is demonstrated on pairs rather than algebraic
-%   datatypes, and is covered in \cref{appendix:pairs}.
+\begin{itemize}[topsep=0pt]
+\item Pattern matching is demonstrated on pairs rather than algebraic
+  datatypes.
 \item There are separate operators for borrowing and reborrowing (taking
@@
 a type application $\tapp{\tcon}{\Multi{\tau}}$ of an abstract type constructor
-$\T t$, or a borrowed type  $\borrowty{k}{\tau}$.
+$\T T$, or a borrowed type  $\borrowty{k}{\tau}$.
 
 Type schemes $\schm$ add quantification over kind
-variables $\kappa$, kinded type variables $\alpha:k$, and constraints
+variables $\kappa$, kinded type variables $(\alpha:k)$, and constraints
 $C$ to a type, where a constraint is a list of inequalities over
@@
 %
-$a$ only has one borrow, its region covers its whole lexical scope.
+Since $a$ only has one borrow, its region covers its whole lexical scope.
 $x$ has multiple conflicting borrows and requires more consideration.

### 2019-11-27 kindly/motivation-pool.tex

@@
   let r = {| f &!o |} in(*@\label{line:pool:region}*)
-  Queue.push o &queue;
+  CQueue.push o &queue;
   r

### 2020-02-05 kindly/pldi20-rebuttal.md
first shot at rebuttal.

We thank the reviewers for their thoughtful comments.
We will take their comments into account and revise our article as
follows.
* Provide a more detailed comparison with the state of the art
* Discuss the limits of our system

# Inference

Reviewer A argues that inference is not such an important property and
would forbid other interesting features, notably for programming in the large.
This is partially a cultural question: Haskell and Rust users are used
to annotations while ML programmers are not.
Nevertheless, several 
reasons leads us to consider complete inference as essential:

- Even if a language does not admit full type inference, we believe
  there should be a well-defined core language which does.
  This way, a programmer who steps over the boundary and use these features
  knows when type annotations are needed. 
  In particular, it is possible (and desirable) to extend Affe with
  existential types by requiring type annotations, similarly to Alms.
  The results we present nevertheless garantee that
  full type inference is supported in the absence of existentials.
  This is similar to the current state of OCaml.

- Contrary to what Reviewer A states, ad-hoc polymorphism (type
  classes, traits, implicits, ...) supports complete inference, and such
  extension of Affe would be highly desirable.
  Affe is also designed to work in tandem with a rich, ML-like module
  system, which has been shown support programming in the large very well.

- Linear types and ownerships are known for being fairly heavy to use.
  Rust is no exception and complex elision rules
  for lifetimes were instrumental in increasing ease-of-use. Affe shows
  that we can go much further thanks to inference. This could be used
  to improve ease-of-use of systems which, like rust, only support
  partial inference.

# Related works

As reviewer C points out, the related works is very rich. Due to
limited space, we only gave a small comparison to each relevant work.

## Linear Haskell
LH indeed does have a formalized core calculus. However, only type checking is
formalized, not inference. The inference implemented is
partial (as pointed out by the GHC maintainers themselves), which
is precisely one of the contentious point to consider inclusion in
GHC.
Not only is our inference algorithm formalized, but we
prove its correctness and completeness.

## Limitations and comparison with Rust

Rust is an established industrial language with more than 10 years of
improvements. Affe does not aim to match its range in term of
use-cases, it simply aims to provide a more accessible point in the
design-space.

Nevertheless, several features of Rust pointed by reviewer C
can be adapted to our system. In many cases, we believe these features
can be added by only modifying the "constraint generation" part of Affe. 
The constraint language and its solver are unchanged, which is
precisely the strengths of the typing-by-constraints approach which we use.

- Adapting NNL to our system would requires inferring regions
  that are not lexical scopes. In many cases, this does not require
  changing our typing rules, only our region inference to obtain
  piecewise lexical scopes.
  We aim to leverage the work by Weiss et Al in this direction.
- In rust, borrows of borrows are done through ad-hoc polymorphism
  (the Borrow and Defer traits) which inserts as many
  borrows as necessary.
  The exact same solution works perfectly in Affe: it does not
  compromise inference, works well in a functional setting and
  integrates in our theoretical framework thanks to qualified types.
- Affe does not need to support successive borrows, as our region
  inference will simply place appropriate region around each of them.

# Functional idioms and combinators

Affe supports currying. More generally, any valid core-ML code is valid in
Affe and the inferred type will be more general. This is immediate by
our use of HM(X). Typical functional combinators can all be written,
but their types might not allow manipulating
linear resources or borrows. 
In general, Affe will support borrowing patterns that are not
flow-sensitive. For instance combinators that always borrow each
element, such as our fold, can be written. 
Combinators that might or might not borrow,
depending on their argument (for instance, using permission witnesses)
requires a richer logic (such as mezzo).

### 2020-02-05 kindly/pldi20-rebuttal.md
Polish some parts of the rebuttal.

@@
 
-As reviewer C points out, the related works is very rich. Due to
+As several reviewers point out, the related works is very rich. Due to
 limited space, we only gave a small comparison to each relevant work.
+We plan to expand our article in this direction
+and answer here to specific questions.
 
 ## Linear Haskell
-LH indeed does have a formalized core calculus. However, only type checking is
-formalized, not inference. The inference implemented is
+LH indeed does indeed have a formalized core calculus. 
+However, only type checking is formalized, not inference. 
+The inference implemented is
 partial (as pointed out by the GHC maintainers themselves), which
-is precisely one of the contentious point to consider inclusion in
-GHC.
+is precisely one of the contentious point for inclusion in GHC.
 Not only is our inference algorithm formalized, but we
@@
 
-## Limitations and comparison with Rust
-
+## Rust
 Rust is an established industrial language with more than 10 years of
-improvements. Affe does not aim to match its range in term of
-use-cases, it simply aims to provide a more accessible point in the
+improvements. Affe does not (yet!) aim to match its range in term of
+use-cases, it simply provides a more accessible point in the
 design-space.
 
-Nevertheless, several features of Rust pointed by reviewer C
+Nevertheless, several features of Rust pointed by reviewer D
 can be adapted to our system. In many cases, we believe these features
 can be added by only modifying the "constraint generation" part of Affe. 
-The constraint language and its solver are unchanged, which is
-precisely the strengths of the typing-by-constraints approach which we use.
+The constraint language and its solver would be unchanged, which is
+precisely the strengths of the typing-by-constraints approach we use.
 
-- Adapting NNL to our system would requires inferring regions
-  that are not lexical scopes. In many cases, this does not require
-  changing our typing rules, only our region inference to obtain
-  piecewise lexical scopes.
-  We aim to leverage the work by Weiss et Al in this direction.
-- In rust, borrows of borrows are done through ad-hoc polymorphism
+- Adapting Non-lexical lifetimes (NNL) to our system would requires 
+  inferring regions that are not lexical scopes. 
+  Concretely, this means that a region is a cover tree of the
+  control flow graph, instead of the graph of all paths between two points.
+  Inferring such precise regions is very desirable and we believe
+  it can be done by only improving our region inference, without affecting the
+  rest of the system.
+- In rust, borrows of borrows are often done through ad-hoc polymorphism
   (the Borrow and Defer traits) which inserts as many
@@
   integrates in our theoretical framework thanks to qualified types.
-- Affe does not need to support successive borrows, as our region
-  inference will simply place appropriate region around each of them.
+- Affe does not need to support successive borrows (i.e., borrows in
+  successive expressions), as region
+  inference will place appropriate regions around each of them.
+- The extension to algebraic data types is easy and done in our
+  prototype. The only difficulty is to verify nesting, in particular
+  linear objects must not live inside unrestricted types.
+- Interior mutability is the ability to mutate an object from a shared
+  borrow. In rust, this is provided by an unsafe primitive.
+  Affe would have the same constraints. However, Affe can safely
+  express types such as Cell using mutable records:
+
+  type ('a : 'k) ref : 'k = { mutable content : 'a } where ('k <= Un_infty)
 
@@
 
+Reviewer C wonders about Affe's support for functional combinators and idioms.
 Affe supports currying. More generally, any valid core-ML code is valid in
@@
 linear resources or borrows. 
-In general, Affe will support borrowing patterns that are not
+Affe also supports borrowing patterns that are not
 flow-sensitive. For instance combinators that always borrow each
 element, such as our fold, can be written.
-Combinators that might or might not borrow,
-depending on their argument (for instance, using permission witnesses)
-requires a richer logic (such as mezzo).
+
+# Limitations
+
+Reviewer C and D asks to details the limitations of our system.
+The main limitations are due to the fact that Affe's type algebra
+is voluntarily simple:
+
+- Affe is not flow sensitive, and will not be able to express
+functions that might or might use something depending on their
+argument.
+This requires richer logic such as Mezzo, or languages with typestates.
+It prevents typing functions such as the merge on linear lists:
+
+  let rec merge l1 l2 = match l1, l2 with
+    | h1::t1, h2::t2 ->
+      if &h1 < &h2 
+      then h1::(merge t1 l2) (* Must expand l2 to h2::t2 here *)
+      else h2::(merge l1 t2)
+    | ....
+
+- In Rust, many primitives are written using unsafe code which is
+  then abstracted. Such unsafe code can be then proven correct (cf
+  RustBelt).
+  Affe does allow using abstraction in a similar way, but is much
+  more limited and doesn't provide such a language-integrate
+  mechanism.

### 2020-02-05 kindly/pldi20-rebuttal.md
More polish.

@@
 Reviewer A argues that inference is not such an important property and
-would forbid other interesting features, notably for programming in the large.
+would preclude other interesting features, notably for programming in the large.
 This is partially a cultural question: Haskell and Rust users are used
@@
   Affe is also designed to work in tandem with a rich, ML-like module
-  system, which has been shown support programming in the large very well.
+  system, which has been shown to support programming in the large very well.
 
@@
   that we can go much further thanks to inference. This could be used
-  to improve ease-of-use of systems which, like rust, only support
+  to improve ease-of-use of systems which, like Rust, only support
   partial inference.
@@
 limited space, we only gave a small comparison to each relevant work.
-We plan to expand our article in this direction
-and answer here to specific questions.
+We will expand our article in this direction and
+now answer specific questions.
 
@@
   Inferring such precise regions is very desirable and we believe
-  it can be done by only improving our region inference, without affecting the
+  it can be done by improving our region inference, without affecting the
   rest of the system.
@@
 - Affe does not need to support successive borrows (i.e., borrows in
-  successive expressions), as region
+  successive expressions) as region
   inference will place appropriate regions around each of them.
 - The extension to algebraic data types is easy and done in our
-  prototype. The only difficulty is to verify nesting, in particular
+  prototype. The only difficulty is to verify nesting: notably
   linear objects must not live inside unrestricted types.
@@
   borrow. In rust, this is provided by an unsafe primitive.
-  Affe would have the same constraints. However, Affe can safely
+  Affe would generally have the same constraints. However, Affe can safely
   express types such as Cell using mutable records:
 
-  type ('a : 'k) ref : 'k = { mutable content : 'a } where ('k <= Un_infty)
+  type ('a : 'k) ref : 'k = { mutable content : 'a } where ('k <= un)
 
@@
 Reviewer C and D asks to details the limitations of our system.
-The main limitations are due to the fact that Affe's type algebra
+Most limitations are due to the fact that Affe's type algebra
 is voluntarily simple:
@@
 functions that might or might use something depending on their
-argument.
-This requires richer logic such as Mezzo, or languages with typestates.
-It prevents typing functions such as the merge on linear lists:
+argument. More generally, we don't support borrowing pattern
+whose correctness depends on internal invariants of a data-structure.
+This requires richer logic such as Mezzo, typestates, etc.
+As a simple example, Affe can not directly type functions such as the 
+merge on linear lists:
 
@@
 - In Rust, many primitives are written using unsafe code which is
-  then abstracted. Such unsafe code can be then proven correct (cf
+  then abstracted. Such unsafe code can be then proven correct (see
   RustBelt).
-  Affe does allow using abstraction in a similar way, but is much
+  Affe aims to use module abstraction in a similar way, but is much
   more limited and doesn't provide such a language-integrate
-  mechanism.
+  unsafe mechanism.
+
+- Affe's semantics is similar to ML: pass by reference and GC.
+  Many linear languages, and Rust in particular, supports very precise
+  control over memory structures and call conventions which we do not
+  aim to provide.

### 2020-02-05 kindly/pldi20-rebuttal.md
Spellcheck.

@@
   existential types by requiring type annotations, similarly to Alms.
-  The results we present nevertheless garantee that
+  The results we present nevertheless guarantee that
   full type inference is supported in the absence of existentials.
@@
 
-- Linear types and ownerships are known for being fairly heavy to use.
+- Linear types and ownership are known for being fairly heavy to use.
   Rust is no exception and complex elision rules
@@
 ## Linear Haskell
-LH indeed does indeed have a formalized core calculus. 
+LH does indeed have a formalized core calculus. 
 However, only type checking is formalized, not inference. 
@@
 
-- Adapting Non-lexical lifetimes (NNL) to our system would requires 
+- Adapting Non-lexical lifetimes (NNL) to our system would require 
   inferring regions that are not lexical scopes. 
@@
   control flow graph, instead of the graph of all paths between two points.
-  Inferring such precise regions is very desirable and we believe
+  Inferring such precise regions is very desirable, and we believe
   it can be done by improving our region inference, without affecting the
@@
 
-- Affe is not flow sensitive, and will not be able to express
+- Affe is not flow-sensitive, and will not be able to express
 functions that might or might use something depending on their
@@
   RustBelt).
-  Affe aims to use module abstraction in a similar way, but is much
+  Affe aims to use module abstraction similarly, but is much
   more limited and doesn't provide such a language-integrate
@@
 - Affe's semantics is similar to ML: pass by reference and GC.
-  Many linear languages, and Rust in particular, supports very precise
+  Many linear languages, and Rust in particular, support very precise
   control over memory structures and call conventions which we do not

### 2020-02-05 kindly/pldi20-rebuttal.md
rewrites

--- 2020-02-05 kindly/pldi20-rebuttal.md
@@
 to annotations while ML programmers are not.
-Nevertheless, several 
-reasons leads us to consider complete inference as essential:
+We consider type inference as essential for several reaons:
 
-- Even if a language does not admit full type inference, we believe
-  there should be a well-defined core language which does.
-  This way, a programmer who steps over the boundary and use these features
-  knows when type annotations are needed. 
-  In particular, it is possible (and desirable) to extend Affe with
-  existential types by requiring type annotations, similarly to Alms.
-  The results we present nevertheless garantee that
-  full type inference is supported in the absence of existentials.
+- Type annotations complement and empower type inference.
+  Our proposal gives a well-defined core language with full type
+  inference. In an extended language, type annotations will be
+  required where a programmer uses the extensions, but the rest will
+  be inferred.
+  For example, it is possible (and desirable) to extend Affe with
+  existential types (as in Alms), which requires type annotations, but
+  *only* in places that deal with existentials.
   This is similar to the current state of OCaml.
@@
   extension of Affe would be highly desirable.
-  Affe is also designed to work in tandem with a rich, ML-like module
-  system, which has been shown support programming in the large very well.
+  Affe is designed to work with an ML-like module
+  system, which is proven to support programming in the large.
 
-- Linear types and ownerships are known for being fairly heavy to use.
+- Linear types and ownership are notoriously tricky concepts.
   Rust is no exception and complex elision rules
@@
   that we can go much further thanks to inference. This could be used
-  to improve ease-of-use of systems which, like rust, only support
+  to improve ease-of-use of systems which, like Rust, only support
   partial inference.
 
-# Related works
+# Related work (Reviewer C)
 
-As several reviewers point out, the related works is very rich. Due to
-limited space, we only gave a small comparison to each relevant work.
-We plan to expand our article in this direction
-and answer here to specific questions.
+Due to limited space, we can only compare with a selection of the most
+relevant work. We will incude more if there is more space in the final paper.
 
 ## Linear Haskell
-LH indeed does indeed have a formalized core calculus. 
-However, only type checking is formalized, not inference. 
-The inference implemented is
-partial (as pointed out by the GHC maintainers themselves), which
-is precisely one of the contentious point for inclusion in GHC.
-Not only is our inference algorithm formalized, but we
+
+LH comes with the metatheory culminating in type soundness. However,
+*type inference* for LH is not formalized and only sketched in the
+paper. (This is particularly concerning in the light of two POPL2020
+papers that highlight subtle issues in typechecking Haskell.)
+In contrast, our inference algorithm is formalized and we
 prove its correctness and completeness.
 
-## Rust
-Rust is an established industrial language with more than 10 years of
-improvements. Affe does not (yet!) aim to match its range in term of
-use-cases, it simply provides a more accessible point in the
-design-space.
+## Limitations and comparison with Rust
 
-Nevertheless, several features of Rust pointed by reviewer D
-can be adapted to our system. In many cases, we believe these features
-can be added by only modifying the "constraint generation" part of Affe. 
-The constraint language and its solver would be unchanged, which is
-precisely the strengths of the typing-by-constraints approach we use.
+Rust is an established industrial language which evolved over more
+than 10 years. Affe highlights a useful accessible point in the design
+space, it does not claim to address all of Rust's use cases.
 
-- Adapting Non-lexical lifetimes (NNL) to our system would requires 
-  inferring regions that are not lexical scopes. 
-  Concretely, this means that a region is a cover tree of the
-  control flow graph, instead of the graph of all paths between two points.
-  Inferring such precise regions is very desirable and we believe
-  it can be done by only improving our region inference, without affecting the
-  rest of the system.
-- In rust, borrows of borrows are often done through ad-hoc polymorphism
+Nevertheless, several features of Rust mentioned by reviewer C
+can be adapted to our system, sometimes by just modifying the
+"constraint generation" part of Affe.  
+The constraint language and its solver are unchanged, which is
+a key strength of the constraint-based approach to typing.
+
+- Adapting NNL to our system requires inferring regions
+  that are not lexical scopes. Mostly, this wouldn't affect our typing
+  rules, but the region inference to obtain
+  piecewise lexical scopes.
+  We aim to leverage the work by Weiss et al in this direction.
+- In Rust, borrows of borrows are done through ad-hoc polymorphism
   (the Borrow and Defer traits) which inserts as many
   borrows as necessary.
-  The exact same solution works perfectly in Affe: it does not
-  compromise inference, works well in a functional setting and
-  integrates in our theoretical framework thanks to qualified types.
-- Affe does not need to support successive borrows (i.e., borrows in
-  successive expressions), as region
-  inference will place appropriate regions around each of them.
-- The extension to algebraic data types is easy and done in our
-  prototype. The only difficulty is to verify nesting, in particular
-  linear objects must not live inside unrestricted types.
-- Interior mutability is the ability to mutate an object from a shared
-  borrow. In rust, this is provided by an unsafe primitive.
-  Affe would have the same constraints. However, Affe can safely
-  express types such as Cell using mutable records:
-
-  type ('a : 'k) ref : 'k = { mutable content : 'a } where ('k <= Un_infty)
+  The Affe core language in the paper has two different operators, but
+  a surface language would map them to one overloaded operator,
+  without compromising any of our results. It also fits with the
+  framework of constrained types, but our presentation aims at
+  avoiding unnecessary complexity.
+- Affe does not require successive borrows, as region
+  inference simply places appropriate regions around each of them.
+- reviewer D mentions pointers into enums. This concept does not exist
+  in ML; only explicit reference types are mutable.
 
@@
 
-Reviewer C wonders about Affe's support for functional combinators and idioms.
-Affe supports currying. More generally, any valid core-ML code is valid in
-Affe and the inferred type will be more general. This is immediate by
-our use of HM(X). Typical functional combinators can all be written,
-but their types might not allow manipulating
-linear resources or borrows. 
-Affe also supports borrowing patterns that are not
-flow-sensitive. For instance combinators that always borrow each
+Affe supports currying as it is a strict superset of core-ML. The
+additional constraints only affect linearity.
+
+In particular, all familiar combinators are typable with their familiar
+ML type; their principal linearity constraints will indicate whether
+the combinator is compatible with linear resources and/or borrows.
+
+In general, Affe supports borrowing patterns that are not
+flow-sensitive. For instance, combinators that always borrow each
 element, such as our fold, can be written. 
+A function that may or may not borrow,
+depending on an argument (for instance, using permission witnesses)
+requires a richer logic (such as Mezzo).
 
-# Limitations
+# Please explain in author response (reviewer D):
 
-Reviewer C and D asks to details the limitations of our system.
-The main limitations are due to the fact that Affe's type algebra
-is voluntarily simple:
+## how to read
+    "let*? () = rho ? b && rho \in pi ..."
 
-- Affe is not flow sensitive, and will not be able to express
-functions that might or might use something depending on their
-argument.
-This requires richer logic such as Mezzo, or languages with typestates.
-It prevents typing functions such as the merge on linear lists:
+This is explained in the text. "let*? () = bool-expression" evaluates
+the boolean expression and reduces to "return ()" when true and fails
+otherwise.
 
-  let rec merge l1 l2 = match l1, l2 with
-    | h1::t1, h2::t2 ->
-      if &h1 < &h2 
-      then h1::(merge t1 l2) (* Must expand l2 to h2::t2 here *)
-      else h2::(merge l1 t2)
-    | ....
+## stack of modifiers
 
-- In Rust, many primitives are written using unsafe code which is
-  then abstracted. Such unsafe code can be then proven correct (cf
-  RustBelt).
-  Affe does allow using abstraction in a similar way, but is much
-  more limited and doesn't provide such a language-integrate
-  mechanism.
+Each borrow adds its kind to the address. The addition checks whether
+the new borrow is permissible (the "rho ? b" operator): we cannot take
+a mutable borrow of an immutable borrow. This happens on entry to the
+borrow's region. The soundness proof maintains the currently
+accessible borrows in the set of permissions, which is also managed by
+the operational semantics (however, all of this manipulation can be
+elided in an impementation; the instrumented semantics is only needed
+for the soundness proof).

### 2020-02-05 kindly/pldi20-rebuttal.md

@@
 
-## Limitations and comparison with Rust
+## Rust
 
 Rust is an established industrial language which evolved over more
-than 10 years. Affe highlights a useful accessible point in the design
+than 10 years. Affe highlights an important point in the design
 space, it does not claim to address all of Rust's use cases.
 
-Nevertheless, several features of Rust mentioned by reviewer C
+Nevertheless, several features of Rust mentioned by reviewer D
 can be adapted to our system, sometimes by just modifying the
 "constraint generation" part of Affe.  
-The constraint language and its solver are unchanged, which is
+The constraint language and its solver remain the same, which is
 a key strength of the constraint-based approach to typing.
@@
 - Adapting NNL to our system requires inferring regions
-  that are not lexical scopes. Mostly, this wouldn't affect our typing
-  rules, but the region inference to obtain
-  piecewise lexical scopes.
-  We aim to leverage the work by Weiss et al in this direction.
+  that are not lexical scopes. 
+  Concretely, a region is a cover tree of the
+  control flow graph, instead of the graph of all paths between two points.
+  Inferring such precise regions is very desirable, and we believe
+  it can be done by improving our region inference, without affecting the
+  rest of the system.
 - In Rust, borrows of borrows are done through ad-hoc polymorphism
@@
 - Affe does not require successive borrows, as region
-  inference simply places appropriate regions around each of them.
-- reviewer D mentions pointers into enums. This concept does not exist
-  in ML; only explicit reference types are mutable.
+  inference places appropriate regions around each of them.
+- The extension to algebraic data types is easy and done in our
+  prototype. The only difficulty is to verify nesting: 
+  linear objects must not live inside unrestricted types.
+- Interior mutability is the ability to mutate an object from a shared
+  borrow. This is an usafe primitive In Rust.
+  Affe would have similar constraints. However, Affe can safely
+  express types such as Cell using mutable records:
 
-# Functional idioms and combinators
+  type ('a : 'k) ref : 'k = { mutable content : 'a } where ('k <= un)
+
+# Functional idioms and combinators (reviewer C)
 
@@
 requires a richer logic (such as Mezzo).
+
+# Limitations
+
+Reviewer C and D asks to details the limitations of our system.
+Most limitations are due to the fact that Affe's type algebra
+is voluntarily simple:
+
+- Affe is not flow-sensitive, and will not be able to express
+functions that might or might use something depending on their
+argument. More generally, we don't support borrowing pattern
+whose correctness depends on internal invariants of a data-structure.
+This requires richer logic such as Mezzo, typestates, etc.
+As a simple example, Affe can not directly type functions such as the 
+merge on linear lists:
+
+  let rec merge l1 l2 = match l1, l2 with
+    | h1::t1, h2::t2 ->
+      if &h1 < &h2 
+      then h1::(merge t1 l2) (* Must expand l2 to h2::t2 here *)
+      else h2::(merge l1 t2)
+    | ....
+
+- In Rust, many primitives are written using unsafe code which is
+  then abstracted. Such unsafe code can be then proven correct (see
+  RustBelt).
+  Affe aims to use module abstraction similarly, but is much
+  more limited and doesn't provide such a language-integrate
+  unsafe mechanism.
+
+- Affe's semantics is similar to ML: pass by reference and GC.
+  Many linear languages, and Rust in particular, support very precise
+  control over memory structures and call conventions which we do not
+  aim to provide.
 

### 2020-02-05 kindly/lang.tex

@@
 
-A type $\tau$ is either a type variable, a function type indexed by a
+A type $\tau$ is either a type variable, a pair type, a function type indexed by a
 kind,
@@
 Abstract type constructors possess kind schemes $\kschm$ which relate
-the kinds of the type constructors arguments to the kind of the
+the kinds of the type constructors' arguments to the kind of the
 constructed type.

### 2020-02-05 kindly/pldi20-rebuttal.md
Fixed 3 typos.

@@
 
-Reviewer C and D asks to details the limitations of our system.
+Reviewer C and D asks to detail the limitations of our system.
 Most limitations are due to the fact that Affe's type algebra
@@
 - Affe is not flow-sensitive, and will not be able to express
-functions that might or might use something depending on their
-argument. More generally, we don't support borrowing pattern
+functions that might or might not use something depending on their
+argument. More generally, we don't support borrowing patterns
 whose correctness depends on internal invariants of a data-structure.

### 2020-02-05 kindly/pldi20-rebuttal.md
Last fixes.

@@
 
-# Inference
-
-Reviewer A argues that inference is not such an important property and
-would preclude other interesting features, notably for programming in the large.
-This is partially a cultural question: Haskell and Rust users are used
-to annotations while ML programmers are not.
-We consider type inference as essential for several reaons:
-
-- Type annotations complement and empower type inference.
-  Our proposal gives a well-defined core language with full type
-  inference. In an extended language, type annotations will be
-  required where a programmer uses the extensions, but the rest will
-  be inferred.
-  For example, it is possible (and desirable) to extend Affe with
-  existential types (as in Alms), which requires type annotations, but
-  *only* in places that deal with existentials.
-  This is similar to the current state of OCaml.
-
-- Contrary to what Reviewer A states, ad-hoc polymorphism (type
-  classes, traits, implicits, ...) supports complete inference, and such
-  extension of Affe would be highly desirable.
-  Affe is designed to work with an ML-like module
-  system, which is proven to support programming in the large.
-
-- Linear types and ownership are notoriously tricky concepts.
-  Rust is no exception and complex elision rules
-  for lifetimes were instrumental in increasing ease-of-use. Affe shows
-  that we can go much further thanks to inference. This could be used
-  to improve ease-of-use of systems which, like Rust, only support
-  partial inference.
-
-# Related work (Reviewer C)
+# Related work (Reviewers A, C and D)
 
@@
   that are not lexical scopes. 
-  Concretely, a region is a cover tree of the
-  control flow graph, instead of the graph of all paths between two points.
+  Concretely, a region would be a cover tree of the
+  control flow graph, instead of the subgraph of all paths between two points.
   Inferring such precise regions is very desirable, and we believe
-  it can be done by improving our region inference, without affecting the
+  it can be done by improving our region inference without affecting the
   rest of the system.
@@
   The Affe core language in the paper has two different operators, but
-  a surface language would map them to one overloaded operator,
-  without compromising any of our results. It also fits with the
+  a surface language would map them to one overloaded operator using a
+  similar technique   without compromising any of our results. 
+  It also fits with the
   framework of constrained types, but our presentation aims at
@@
 - Interior mutability is the ability to mutate an object from a shared
-  borrow. This is an usafe primitive In Rust.
-  Affe would have similar constraints. However, Affe can safely
+  borrow. This is an usafe primitive in Rust.
+  Affe would have similar constraints. However, Affe could safely
   express types such as Cell using mutable records:
-
-  type ('a : 'k) ref : 'k = { mutable content : 'a } where ('k <= un)
+  ```
+  type ('a : 'k) cell : 'k = { mutable content : 'a } where ('k <= un)
+  ```
 
@@
 element, such as our fold, can be written.
-A function that may or may not borrow,
-depending on an argument (for instance, using permission witnesses)
-requires a richer logic (such as Mezzo).
 
@@
   merge on linear lists:
-
+  ```
   let rec merge l1 l2 = match l1, l2 with
@@
     | ....
+  ```
 
@@
 
-# Please explain in author response (reviewer D):
+# Inference (Reviewer A)
 
-## how to read
-    "let*? () = rho ? b && rho \in pi ..."
+Reviewer A argues that inference is not such an important property and
+would preclude other interesting features, notably for programming in the large.
+We consider type inference as essential for several reasons:
 
-This is explained in the text. "let*? () = bool-expression" evaluates
-the boolean expression and reduces to "return ()" when true and fails
+- Type annotations complement and empower type inference.
+  Our proposal gives a well-defined core language with full type
+  inference. In an extended language, type annotations will be
+  required where a programmer uses the extensions, but the rest will
+  be inferred.
+  For example, it is possible (and desirable) to extend Affe with
+  existential types (as in Alms), which requires type annotations, but
+  *only* in places that deal with existentials.
+  This is similar to the current state of OCaml.
+
+- Contrary to what Reviewer A states, ad-hoc polymorphism (type
+  classes, traits, implicits, ...) supports complete inference, and such
+  extension of Affe would be highly desirable.
+  Additionally, Affe is designed to work with an ML-like module
+  system, which is proven to support programming in the large.
+
+- Linear types and ownership are notoriously tricky concepts.
+  Rust is no exception and complex elision rules
+  for lifetimes were instrumental in increasing ease-of-use. Affe shows
+  that we can go much further thanks to inference. This could be used
+  to improve ease-of-use of systems which, like Rust, only support
+  partial inference.
+
+# Please explain in author response (Reviewer D):
+
+## how to read `let*? () = rho ? b && rho \in pi ...`
+
+This is explained in the text. `let*? () = bool-expression` evaluates
+the boolean expression and reduces to `return ()` when true and fails
 otherwise.

### 2020-02-26 kindly/motivation-persistent.tex
Move back the sudoku example.

@@
   \centering
-  % \begin{subfigure}{1\linewidth}
+  \begin{subfigure}[t]{0.45\linewidth}
 \begin{lstlisting}
@@
 end
-
+\end{lstlisting}
+  \end{subfigure}\hfill
+  \begin{subfigure}[t]{0.55\linewidth}
+\begin{lstlisting}
 module Cow : HYBARRAY = struct
@@
 \end{lstlisting}
+  \end{subfigure}
   \vspace{-15pt}
-  %   \caption{Implementation using copy-on-write}
-  % \end{subfigure}
-  \caption{Hybrid arrays}
+  \caption{Signature and Implementation of hybrid arrays}
   \label{sig:hybarray}
@@
 
-  \lstinputlisting[linerange=1-31]{code/sukodu.affe}
+  \begin{subfigure}[t]{0.45\linewidth}
+  \lstinputlisting[linerange=1-13]{code/sukodu.affe}
+  \end{subfigure}\hfill
+  \begin{subfigure}[t]{0.55\linewidth}
+  \lstinputlisting[linerange=14-31]{code/sukodu.affe}
+  \end{subfigure}
+  % \lstinputlisting[linerange=1-31,multicols=2]{code/sukodu.affe}
   \vspace{-15pt}

### 2020-02-28 kindly/main.tex
Use the ICFP format.
Adapt the figures to fit better.
Move back the sudoku example.
abstract

@@
 \PassOptionsToPackage{svgnames}{xcolor}
-\documentclass[sigplan,10pt,review,anonymous]{acmart}
+\documentclass[acmsmall,review,anonymous]{acmart}
 \settopmatter{printfolios=true,printccs=false,printacmref=false}
@@
 
-\mprset{sep=1.2em}
+\mprset{sep=1.5em}
 
@@
   safety by design, most of them do not address issues arising
-  from improper resource handling.
-  Linear and affine types guarantee single-threaded resource usage,
-  but they are rarely deployed as they are too restrictive for real-world applications.
+  from improper handling of resources.
+  An important step towards handling resources is the adoption of 
+  linear and affine types that enforce single-threaded resource usage.
+  However, the few languages supporting such types require heavy type
+  annotations.
 
-  We present \lang, an extension of ML with constrained types that
-  manages linearity and affinity properties through kinds. In addition
+  We present \lang, an extension of ML that
+  manages linearity and affinity properties using kinds and
+  constrained types. In addition
   \lang{} supports the exclusive and shared borrowing of affine
@@
   Moreover, \lang{} retains the defining features of the ML family:
-  an impure, strict, functional expression language with complete principal type
+  it is an impure, strict, functional expression language with complete principal type
   inference and type abstraction.
-  \lang does not require any linearity annotations in
+  \lang{} does not require any linearity annotations in
   expressions and supports common functional programming idioms.
@@
 \input{motivation-linearity}
-\input{motivation-persistent}
 \input{motivation-pool}

### 2020-02-28 kindly/intro.tex
intro

@@
 %
-Dynamic allocation of memory is analogous. An API call returns a
+Dynamic allocation of memory is similar. An API call returns a
 pointer to a memory area, which can then be read and written to until
-the memory is released by another API call.
+the area is released by another API call.
 
@@
 can safely be applied to it. Finally, there is another API call to
-release the resource, which yields the handle invalid.
+release the resource, which renders the handle invalid.
 Taken to the extreme, each API call changes the state so that a
@@
 
-Most type systems provide just type soundness, but neglect the
-protocol aspect. The systems that do build on linear types \cite{DBLP:journals/tcs/Girard87} and/or
+Most type systems provide type soundness and memory safety, but neglect the
+protocol aspect. Systems that can support reasoning about protocols
+build on linear types \cite{DBLP:journals/tcs/Girard87} and/or 
 uniqueness types~\cite{DBLP:conf/plilp/BarendsenS95}. A value of linear
@@
 memory that has been allocated must be released. A value of unique
-type is guaranteed to have a single reference to it.
+type is guaranteed to have a single reference to it. Thus, memory can
+be reused on consumption of the value.
 
@@
 type~\cite{DBLP:conf/oopsla/AldrichSSS09}. Functional session types
-represent a popular example~\cite{DBLP:journals/jfp/GayV10}.
+represent a popular
+example~\cite{DBLP:journals/jfp/GayV10,lindley17:_light_funct_session_types}.
 
@@
 \emph{borrowing} a resource. The idea is that a linear resource can be
-borrowed to a function call, which can work with the resource, but
-cannot release it. Only the original owner of the resource has all
-rights to it and can release it.
+borrowed to a function call. The function can work with a borrow of
+the resource, but it cannot release the resource. Only the original
+owner of the resource has all rights to it and can release it.
 
@@
 and they form the foundation of the type system of the Rust language
-\cite{rust}. Rust supports two kinds of borrows of data structures, shared and
-exclusive ones.
-Exclusive borrows generally enable modification of the data structure
-whereas shared borrows only grant read-only access.
-Several shared borrows can be active at the same time, but
-can not be active while an exclusive borrow is.
+\cite{rust}, which considers any memory-allocated data structure a
+resource. Rust supports two kinds of borrows,
+shared and exclusive ones.
+Exclusive borrows enable modification of the data structure
+whereas shared borrows only grant read access.
+At any given time, either a single exclusive borrow is active or 
+any number of shared borrows can be active.
 Moreover, Rust makes sure that the lifetime of a
@@
 As a concrete example, consider a functional programmer who wants to
-safely interact with a C library. Invoking the C functions is easy via
+safely interact with a C library. Invoking a C function is easy via
 the existing foreign function interface, but managing the underlying
 resources like malloc'd storage is not: It cannot be left to the
-garbage collector, but the proper calls to \texttt{free()} must be
+garbage collector, but proper release of the storage via calls to \texttt{free()} must be
 ensured by programming conventions.
 
-Our work provides a solution to programmers in this situation. We
+Our work provides a safe solution to programmers in this situation. We
 propose an extended type system for ML-like languages that comes with
@@
 automatic placement of borrowing regions.
+In our system, it is a type error to omit the call to release the
+storage given a suitably typed API for storage allocation.
 
@@
 Quill \cite{DBLP:conf/icfp/Morris16}, and ALMS \cite{DBLP:conf/popl/TovP11}.
-Compared to LH, the goals and means are
-similar as both systems permit abstraction over the number of uses of
+Compared to LH and Quill, the goals and means are
+similar as these systems also permit abstraction over the number of uses of
 values and retain type inference, but the details are different.
 \begin{enumerate}
-\item LH's multiplicities are either linear or unrestricted whereas
+\item Multiplicities in LH and Quill are either linear or unrestricted whereas
   we also distinguish affine values.
-\item \lang{} supports two flavors of borrowing: affine (exclusive,
-  mutable) and unrestricted (shared, immutable) borrows.
-\item Multiplicities are directly attached to the type of the
+\item In \lang{} and in Quill multiplicities are directly attached to the type of a
   value. For example, in \lang{} the function type \lstinline/'a-{lin}>'b/
@@
     once}.
-\item Similar to Quill, we make use of multiplicity contraints. Kind
-  subsumption in \lang{} helps simplify inferred types.
+\item \lang{} makes use of multiplicity contraints (like Quill) and kind
+  subsumption (unlike Quill). Kind subsumption results in significantly simpler, more readable
+  inferred types.
+\item Neither LH nor Quill have borrowing whereas \lang{} supports
+  two flavors: affine (exclusive, mutable) and unrestricted (shared, immutable) borrows.
 \end{enumerate}
@@
   \caption{Writing files}
-  \label{fig:writing-fiules}
+  \label{fig:writing-files}
 \end{figure}
@@
 linear abstract type @File.t@. A call like
-@fopen "foo"@ returns a linear handle to a newly created
-file, which \emph{must} be released later on with @close@
+@File.fopen "foo"@ returns a linear handle to a newly created
+file, which \emph{must} be released later on with @File.close@
 as shown in \cref{fig:writing-files-example}. Failing to do so is a
@@
 borrow @&!h@ of the handle and pass it to the
-@write@ function. Exclusive borrows are affine:
+@File.write@ function. Exclusive borrows are affine:
 they must not be duplicated, but there is no requirement to use
 them. This affinity shows up in the annotation @-{aff}>@ of the second arrow in
-the type of @write@: a partial application like
-@write &!h@ captures the affine borrow and hence the
+the type of @File.write@: a partial application like
+@File.write &!h@ captures the affine borrow and hence the
 resulting function is also affine. It
@@
 \begin{lstlisting}
-let w = File.write &!h in w "Hello "; w "world!" (*error*)
+let w = File.write &!h in w "Hello "; w "world!" (*type error*)
 \end{lstlisting}
@@
 \begin{enumerate}
-\item full type inference without annotations in user code,
+\item full type inference without annotations in user code (\cref{fig:writing-files-example}),
 \item types carry multiplicity annotations in the form of kinds,
@@
 
+Direct style means that there is a function like @fopen@ that creates
+and returns a linear resource. In contrast, LH forces programmers to
+use resource-passing style because, in LH, linearity is a property of
+a function, rather than a property of a value that restricts the way
+that value can be handled (as in \lang). An LH API analogous to @File@
+might provide functions like
+\begin{itemize}
+\item @withFile : path -> (handle o-> r) -> r@, which creates a new
+  file handle and takes a continuation that uses the @handle@ linearly,
+\item @writeFile : string -> handle o-> handle@, which
+  returns the transformed resource @handle@, and
+\item @closeFile : handle o-> unit@, which consumes the @handle@ by
+  closing the file.
+\end{itemize}
+
 In general, kinds can be polymorphic and constrained. Function
 application and composition are the archetypical
-examples for functions exploiting that feature.
+examples for functions exploiting that feature.\footnote{Compared to
+  Quill \cite{DBLP:conf/icfp/Morris16} the signatures of application and
+  composition are simpler because \lang{} supports kind subsumption.}
+For application, \lang{} infers the following type.
 \begin{lstlisting}
@@
 implicitly assume prenex quantification by
-$\forall\kappa\forall\alpha\forall\beta$. In reality, the
+$\forall\kappa\forall\alpha\forall\beta$. Internally, the
 type checker also quantifies over the kinds of $\alpha$ and $\beta$,
-hence the full prefix is
-$\forall\kappa\kappa_1\kappa_2\forall(\alpha:\kappa_1)\forall(\beta:\kappa_2)$,
-which is only revealed as much as necessary for understanding the type.
+but the full prefix
+$\forall\kappa\kappa_1\kappa_2\forall(\alpha:\kappa_1)\forall(\beta:\kappa_2)$
+of the type of @app@ is only revealed as much as necessary for
+understanding the type.
+
+For @compose@, \lang{} infers this type.
 \begin{lstlisting}
@@
 @f@ both influence the multiplicity of the last arrow, so
-that we would expect its annotation to be the least upper bound
+we would expect its annotation to be the least upper bound
 $\kappa \sqcup \kappa_1$. Thanks to subsumption of multiplicities, it
@@
 actual multiplicity gets subsumed to $\kappa_1$. This constraint
-simplification is part of our type inference algorithm. Again, the printout
-of the type scheme only mentions the required constraint
+simplification is part of our type inference algorithm. As before,
+printing the type scheme only mentions the required constraint 
 $\kappa\le\kappa_1$ and omits the prenex quantification over $\kappa,
@@
 \begin{itemize}
-\item A type system encoding linearity with
+\item A type system encoding linearity and affinity with
   borrowing. The type system is polymorphic for types and
   for kinds that express constraints on the number of uses of a value.
-\item An extension of the HM(X) framework
+\item An extension of the \hmx framework
   \cite{DBLP:journals/tapos/OderskySW99} with kinds, regions, and
   multiplicity constraints that supports full, principal type inference.
-\item Proof of soundness for the semantics and the inference.
+\item Soundness proof of the inference algorithm.
+\item Big-step linearity aware semantics with an expressive soundness theorem.
 \item Constraint simplification techniques.
@@
 
+As \lang{} is built on top of the \hmx{} framework, which is a general
+framework for expressing constraint-based typing and type inference,
+the extension of our work with features like typeclasses, ad-hoc overloading,
+traits, etc is possible and orthogonal to the presentation in this paper.
+
 %%% Local Variables:

### 2020-02-28 kindly/intro.tex
region

@@
   let h = File.fopen "foo" in
-  File.write &!h "Hello ";
-  File.write &!h "world!";
+  File.write &!h "Hello ";(*@\label{line:hello}*)
+  File.write &!h "world!";(*@\label{line:world}*)
   File.close h
@@
 
+Every borrow is restricted to a \emph{region}, i.e., a lexically
+scoped program fragment from which the borrow must not escape. In
+\cref{fig:writing-files-example}, there are two regions, one 
+consisting of \cref{line:hello} and another consisting of
+\cref{line:world}. Both are fully contained in the scope of the linear
+handle @h@, hence we can take one exclusive borrow @&!h@ in each
+region. In both regions the borrow is consumed immediately by passing
+it to @File.write@. Regions are generally inferred by \lang.
+
 This example demonstrates three features of our system:
 \begin{enumerate}
-\item full type inference without annotations in user code (\cref{fig:writing-files-example}),
+\item type and region inference without annotations in user code (\cref{fig:writing-files-example}),
 \item types carry multiplicity annotations in the form of kinds,

### 2020-02-28 kindly/motivation-persistent.tex
motivation cosmetics

@@
 @mapi: (int -> &'a -> 'b) -> &('a t) -> 'b t@
-is a simple variation on @Array.map@ where the mapping function
+is a simple variation on @Array.map@ where the mapping function also
 takes the position of the element.
@@
 of the array, while other modifications use @set_mut@, which
-forbids uses of the previous version.
+cannot be reverted.
 
@@
 but with two indices.
-Each cell contains an integer set that represent admissible solutions so far.
+Each cell contains an integer set that represents admissible solutions so far.
 The Sudoku solver iterates over the cells and tries each possible solution (\cref{line:try_solution}).
@@
 the changes with the @propagate@ function.
-The @propagate@ function uses direct mutation through a
+The @propagate@ function uses direct mutation through an
 exclusive borrow of the matrix as it need not preserve the previous
@@
 The price is that @try_solution@ cannot escape from
-@&g@'s region. In this example, inference places the region
-around the @begin@/@end@ scope.
+@&g@'s region. In this example, the inferred region corresponds to the
+@begin@/@end@ scope. 
 Hence, @try_solution@ can be used in

### 2020-02-28 kindly/motivation.tex
Move back the sudoku example.
motivation cosmetics

@@
 \cite{DBLP:journals/jfp/Padovani17}; see also
-\cref{sec:session-linearity} in the supplement). But it really shines
+\cref{sec:session-linearity} in the supplement) as well as other
+functional resource handling. But it really shines
 when manipulating mutable resources like buffers or connection pools
-using a mixed functional and imperative programming style.
+using a mix of functional and imperative programming styles.
 %
-To support this usage pattern of linearity, we rely on
-the notion of borrowing \cite{DBLP:conf/popl/BoylandR05} which
-we demonstrate first through the example of linear arrays to demonstrate
-simple borrowing and imperative programming and second with iterators
-on linear values to demonstrate advanced uses of regions.
-Further examples are also available in \cref{sec:extra-example}.
+To support this usage pattern of linearity, \lang{} relies on
+the notion of borrowing \cite{DBLP:conf/popl/BoylandR05}. Our first
+example of linear arrays demonstrates simple borrowing and
+imperative programming; the second example is a Sudoko solver using a
+hybrid copy-on-write data structure; the third example demonstrates advanced uses
+of regions with iterators on linear values.
+Further examples are available in \cref{sec:extra-example}.
 
@@
 
-The API for mutable linear arrays (shown in \cref{sig:array})
+The API for mutable linear arrays (\cref{sig:array})
 aims to safely handle manual allocation and
 deallocation of arrays that may contain affine elements.
-One would first use @create n v@ to create
-an (unrestricted) array of size @n@ initialized with value
-@v@. This value must be unrestricted as it is duplicated to
+A program would first use @create n v@ to create
+an array of size @n@ initialized with value
+@v@. The value @v@ must be unrestricted as it is duplicated to
 initialize all array elements. Using the @map@ function we
 can transform the unrestricted elements into linear (affine)
-ones. This array can be processed further using @map@ (or @set@ if
+ones. The resulting array can be processed further using @map@ (or @set@ if
 affine). The @get@ function is only applicable if the element type
@@
 view or \emph{exclusive borrows} @&!a@ for read-write views.
-Observer functions such as @length@ and @get@ expect an shared borrow
+Observer functions such as @length@ and @get@ expect a shared borrow
 argument, mutator functions such a @set@ expect
@@
 
-We can use the API 
-to create an
-array of Fibonacci numbers following an imperative coding style:
+Using the API 
+we can create an
+array of Fibonacci numbers in an imperative coding style:
 \begin{lstlisting}
@@
 %
-More precisely, here is is an annotated excerpt with explicit regions
-marked in braces @{| ... |}@:
+More precisely, here is is an annotated excerpt with regions explicitly
+marked by braces @{| ... |}@:
 \begin{lstlisting}[firstnumber=3]
@@
 
-A region consists of the header expression of the @let@ in
+One region consists of the header expression of the @let@ in
 \cref{ex:array:region:get}. It is contained in another region
@@
 immediately by @get@ and @set@.
+
+\input{motivation-persistent}
 
@@
 
-We can then easily use this function to fold over an array of files. Here
-we compute the sum of their sizes:
+As an easy example, we fold over an array of files to compute the sum of their sizes:
 \begin{lstlisting}
@@
 
-Naturally, this is not sufficient if we want to mutate the array during iteration.
-For this purpose, we need to take an exclusive borrow of the structure to iterate
+This approach is not sufficient if we want to mutate the array during iteration.
+To this end, we need to take an exclusive borrow of the structure to iterate
 on:
@@
 
-Since the callback will be called multiple times, it is explicitly
-annotated with @un@.
+The callback is explicitly annotated with @un@ because  it will be called multiple times.
 
 While the distinction between mutable and immutable iteration functions
-might seem unfortunate, it is typical of
+seems unfortunate, it is typical of
 programming with borrows and is also present in the Rust standard library.
-It allows the programmer to explicitly state how various iterations
-are allowed to compose and be optimized.
-\lang's region variables ensures that the content iterated on can never
+It enables the programmer to explicitly state how different iterations
+may be composed and optimized.
+\lang's region variables ensure that the content iterated on can never
 be leaked outside of the iteration function.
-This pattern is essential in many use-cases of linearity such
+This pattern is essential in many use cases of linearity such
 as pools of linear objects (see \cref{tuto:pool}).

### 2020-02-28 kindly/soa.tex
Small precision on LH.
Add small extensions to the state of the art.

@@
 (LH) is to retrofit linear types to Haskell.
-LH introduces a new notion
-of ``linear arrows'', written $\multimap$ similar to linear logic,
-which are functions that \emph{use} their argument exactly once.
-% By contrast, in \lang linearity is only decided
-% by the kinds, and $\tarr{\klin}$ simply denotes \emph{single-use functions}.
+As opposed to approaches presented previously, including \lang,
+LH annotates linearity on the arrow, similarly to linear logic.
+LH's ``linear arrows'', written $\multimap$ 
+are functions that \emph{use} their argument exactly once
 This design is easy to retrofit on top of an existing compiler
@@
   it is neither formalized nor complete.
+  Only type \emph{checking} is formalized.
+  Subsequent work~\cite{DBLP:journals/corr/abs-1911-00268}
+  formalizes the inference for rank 1 qualified-types.
 \item
@@
 formalized Rust's ownership discipline, including non-lexical lifetimes.
-We believe this work
-could help us improve our handling of regions and of successive exclusive borrows.
+Another important difference is that Rust allows
+to specify precisely the memory layout of objects
+and to pass arguments by value.
+These features are crucial for the efficiency goals of Rust.
+In \lang, we assume that all arguments are passed by reference and
+that a garbage collector is still available. This forgo numerous
+issues regarding interior mutability and algebraic data types.
+In particular, it
+allows us to easily nest mutable references inside objects, regardless
+if they are linear or unrestricted.
 

### 2020-02-28 kindly/lang.tex
Adapt the figures to fit better.
cosmetics

@@
 let-polymorphism and abstract types. Its type system manages
-linearity and borrows using kinds and (kind-) constrained types.
-For ease of presentation, we simplify the language.
+linearity and borrowing using kinds and (kind-) constrained types.
+For ease of presentation, we consider a simplified internal language.
 \begin{itemize}[topsep=0pt]
@@
 \item There are separate operators for borrowing and reborrowing (taking
-  the borrow of a borrow).
+  the borrow of a borrow); the surface language unifies these
+  operators using ad-hoc polymorphism / typeclasses.
 \item Regions are explicit in the code and must be annotated using the
@@
 A match is indexed with a \emph{match specification} $\etransfm$ that indicates
-whether the match should work on borrows ($\etransfm=\&^b$) or not ($\etransfm=id$).
+whether the match operates on borrows ($\etransfm=\&^b$) or not ($\etransfm=id$).
 %
@@
 
-In \cref{motivation}, region annotations are optional.
-In our formalization, regions must be syntactically explicit and annotated with
-a nesting index and a scoped variable.
-We now define a best-effort
-syntactic code transformation $\RannotT{p}{p'}$ which
-annotates programs automatically.
+In the surface language (\cref{motivation}) region annotations are optional.
+In the internal language, regions must be syntactically explicit and
+annotated with a nesting index and a scoped variable.
+This section defines a transformation $\RannotT{p}{p'}$ which
+automatically inserts region annotations in  programs.
 The input $p$ is a program with optional region annotations of the form $\region[]{}{e}$.
 The output $p'$ is a program with explicit annotations
-of the form $\region{\Sone x\BORROW}{e}$.
+of the form $\region{\Sone x\BORROW}{e}$ such that no borrow occurs
+outside a region.
 We give an informal presentation of our code transformation and defer
@@
 \item The region should contain at least $\borrow{x}$.
-\item The region should never be larger than the scope of $x$.
+\item The region must be contained in the scope of $x$.
 \item An exclusive borrow $\borrow[\kaff]{x}$ should never share a region with any other borrow of $x$.
-\item A use of $x$ can never be in the region of $\borrow{x}$.
+\item The variable $x$ cannot occur in the region of $\borrow{x}$.
 \end{enumerate}
-We then start from the leaves and grow the region
-associated to each borrow until it can't grow anymore either due to a binding
-or to conflicting uses. We demonstrate
-on the following program:
+The transformation starts from each borrow $\borrow x$ and grows its associated
+region until enlarging it would include the binding for $x$ or lead to a
+conflicting use of $x$ or a conflicting borrow for $x$.
+As an example, consider the following program:
 %
@@
 %
-Since $a$ only has one borrow, its region covers its whole lexical scope.
-$x$ has multiple conflicting borrows and requires more consideration.
+As variable $a$ only has one borrow, its region covers its whole lexical scope.
+Variable $x$ has multiple conflicting borrows and requires more consideration.
 We place a first
@@
 error.
-Note that indices of regions are well nested, as the inner region have
+Region indices are assigned after placement, so it is trivial to
+ensure well-nested regions where the inner regions have
 higher indices than the outer ones.
 
-Programmers might also want to annotate regions explicitly.
-Such annotations are considered as
+Programmers may also annotate regions explicitly.
+The transformation considers an annotation as an
 upper bound on the contained regions.
@@
 the explicit annotation indicates that the region should stay around the closure
-passed as argument of $set$. This is useful to control captures
+passed as argument of $set$. This feature is useful to control captures
 by imperative APIs.
 
-This code transformation is purely syntactic and must be used before
+The code transformation is purely syntactic and must be used before
 typing. It only produces well-nested annotations: if $\region[n]{b}{\dots}$
@@
 is at most one region per borrow, and exactly one region per exclusive borrow.
-In the rest of this article, we consider
-that all terms have been annotated with this code transformation and respect
+In the rest of this article, we assume
+that all terms have been annotated by this code transformation and respect
 these properties.
@@
 
-We now focus on the essential and novel parts of the type system. A complete
+To avoid distraction, this section focuses on the essential and novel
+parts of the type system. A complete 
 description is available in \cref{appendix:sdtyping}.
@@
 %
-We discuss the following judgments:
-\begin{description}
-\item[$\inferS{C}{\E}{e}{\tau}$ --]
-  Expression $e$ has type $\tau$ in environment $\E$ under constraints $C$.
-\item[$\inferSK{C}{\E}{\tau}{k}$ --]
-  Type $\tau$ has kind $k$ in environment $\E$ under constraints $C$.
-\item[$\entail{D}{C}$ --] Constraint $D$ entails constraint $C$.
-\end{description}
+Here we only discuss the following judgments:
+\begin{flushleft}
+  \begin{tabular}{l@{ --- }l}
+    $\inferS{C}{\E}{e}{\tau}$&
+                               Expression $e$ has type $\tau$ in
+                               environment $\E$ under constraints
+                        $C$. \\
+    $\inferSK{C}{\E}{\tau}{k}$&
+                                Type $\tau$ has kind $k$ in
+                                environment $\E$ under constraints
+                                $C$. \\
+    $\entail{D}{C}$ & Constraint $D$ entails constraint $C$.
+  \end{tabular}
+\end{flushleft}
 
@@
 Constraints are conjunctions of inequality constraints over this kind
-lattice. In most uses, constraints give upper or lower bounds for
+lattice, i.e., they specify upper or lower bounds for
 kind variables or relate two kind variables.
@@
 \begin{figure}[tp]
+  \begin{minipage}{0.36\linewidth}
   \begin{align*}
-    \E &::=\ \Eempty \mid \E;B\tag{Environments}\\
-    B &::=\ \bnone % \tag{Empty} \\
-    \mid \bvar{\tvar}{\kschm} % \tag{Types}
-    \mid \bvar{x}{\schm}% \tag{Variables}
-    \mid \bvar{\borrow{x}}{\borrowty{k}{\tau}}% \tag{Borrows}\\
-    \mid \svar{x}{\schm}^n % \tag{Suspended}
+    \E ::=&\ \Eempty \mid \E;B\tag{Environments}\\
+    B ::=&\ \bnone \tag{Empty} \\
+    |&\ \bvar{\tvar}{\kschm} \tag{Types}\\
+    |&\ \bvar{x}{\schm} \tag{Variables} \\
+    |&\ \bvar{\borrow{x}}{\borrowty{k}{\tau}} \tag{Borrows}\\
+    |&\ \svar{x}{\schm}^n \tag{Suspended}
     % |&\ \tydecl{t}{\kschm}{K}{\tau}\tag{Type Declaration}\\
@@
   \label{grammar:env}
-  \vspace{5pt}
-  % \begin{minipage}{0.5\linewidth}
-  \centering
+  \end{minipage}\hfill
+  \begin{minipage}{0.6\linewidth}
     \input{sdtyping/split-binder}
@@
     \label{sdtyp:split}
-  \vspace{-10pt}
-  % \end{minipage}
+  \end{minipage}
 \end{figure}
@@
 restricting the types of variables.  The constraint $\Cleq{\E}{k}$
-abbreviates $\Cleq\schm k$, for all $\bvar{x}{\schm}$
+stands for the conjunction of $\Cleq\schm k$, for all $\bvar{x}{\schm}$
 % $\svar{x}\schm$, and $\bvar{\borrow{x}}{\borrowty{k'}{\schm}}$
 in $\E$, which in turn means that $k' \lk k$ where $k'$ is the kind of
-the variable's type scheme.
-Borrow bindings follow the exact same rules, but suspended bindings
+the variable's type scheme $\schm$.
+Borrow bindings follow the same rules, but suspended bindings
 are forbidden in an environment $\E$ constrained like that.
 This intuitive explanation is sufficient to understand
-the {\sc Abs} and {\sc Var} rules shown in
+the \TirName{Var} and \TirName{Abs} rules shown in
 \cref{selectrules:binders}.
@@
 This construct is mostly standard, except it can operate both
-on normal pairs and borrows of pairs.
-The intuition is as follow:
+on a normal pair and a borrow of a pair.
+The intuition is as follows:
 A syntactic marker $\etransfm$ indicates if it applies to
@@
 If $\etransfm = \operatorname{id}$, the typing simplifies to
-a regular destruct on pairs.
-Otherwise, $e_1$ is expected to be of type
+the usual elimination of a pair.
+Otherwise, $e_1$ is expected to be a borrow of type
 $\borrowty{k}{\tyPair{\tau_1}{\tau'_1}}$
-while $x$ and $x'$ are of type respectively
-$\borrowty{k}{\tau_1}$ and $\borrowty{k}{\tau'_1}$.
+and the variables $x$ and $x'$ have type
+$\borrowty{k}{\tau_1}$ and $\borrowty{k}{\tau'_1}$,
+respectively. Thus, the borrow of a pair is considered as a pair of
+borrows of its components.
 
@@
 
-As a model of resource, we consider a simple abstract
-type $\tapp{\tres}{\tau}$ whose content of type $\tau$ must be $\kun_0$ and
-equipped with the four class of operations describe in \cref{motivation}:
+To demonstrate how \lang{} deals with resources, we introduce an abstract
+type $\tapp{\tres}{\tau}$ whose content of type $\tau$ must be
+unrestricted ($\kun_0$) and which is equipped with the four
+operations described in \cref{motivation}: 
 \begin{itemize}[topsep=0pt]
@@
 
-\lstMakeShortInline[keepspaces,style=rule,basicstyle=\normalsize\normalfont]@
+% PT: removed style=rule
+\lstMakeShortInline[keepspaces,basicstyle=\normalsize\normalfont]@
 
@@
   DBLP:conf/popl/AminR17}.  A function
-@eval \Store \Perm \VEnv i e@
+@eval@
 manipulates the semantic objects defined in
@@
 
-Let-polymorphism is demonstrated for functions giving rise to elaborated
+Let-polymorphism in the surface language gives rise to elaborated
 $\eletfun$ expressions annotated with a type scheme $\schm$ and a kind $k$ indicating their
@@
 storables. The function
-@salloc: store -> storable -> (loc * store)@ is such that
+@salloc: store -> storable -> (loc *store)@ is such that
 @salloc delta@ @w@ allocates an unused location in @delta@ and fills it with
@@
 type 'a sem = Error of string | TimeOut | Ok of 'a
-val eval: store->perm->venv->int->exp->(store*perm*result) sem
+val eval: store->perm->venv->int->exp->(store *perm *result) sem
 \end{lstlisting}
-It evaluates the given expression in the context of an initial store, a
+Function @eval@ evaluates the given expression in the context of an initial store, a
 permission to use addresses in the store, a value environment, and a
@@
 % contains the semantics in mathematical notation.
-The definition uses OCaml syntax with some pretty
+The definition uses OCaml syntax with extensive pretty
 printing. The pervasive @let*@ operator acts as monadic bind
@@
 argument into success or failure in the monad.
+\begin{lstlisting}
+let (let*?) : bool -> (unit -> 'b sem) -> 'b sem =
+  fun b f -> if b then f () else Error ("test failed")
+\end{lstlisting}
 The function header of @eval@ checks

### 2020-02-28 kindly/infer/exampleconstr.tex
Adapt the figures to fit better.
soundness

@@
   \centering
-  \begin{tikzpicture}[node distance=7mm,every edge/.style=[link,->,>=latex,thick]]
+  \begin{tikzpicture}[node distance=6mm,xscale=1.2,yscale=0.7,.every edge/.style=[link,->,>=latex,thick]]
     \node (U) {$\kun$} ;
@@
   \end{tikzpicture}
+  \vspace{-5pt}
     \caption{Graph representing the example constraints}
@@
 
-We consider the expression $\lam{f}{\lam{x}{(\app{f}{x},x)}}$.
-The application of our inference algorithm yields the following constraints:
+Consider the expression $\lam{f}{\lam{x}{(\app{f}{x},x)}}$.
+The inference algorithm yields the following constraints:
 %
@@
 
-The first step of the algorithm is to use Herbrand unification to obtain
-a type skeleton. We thus obtain
+The first step of the algorithm uses Herbrand unification to obtain
+a type skeleton. 
 $$\alpha_r =
@@
 
-and the following kind constraints: 
+In addition, we obtain the following kind constraints: 
 \[\begin{aligned}
@@
     (\kvar_x \leq \kvar_r)
-    \Cand\\
-    &(\kvar_\beta \leq \kvar_r)
+    \Cand
+    (\kvar_\beta \leq \kvar_r)
     \Cand
@@
 
-The algorithm we presented does not use variance, however we can use it to
-simplify this type further by noticing that $\kvar_1$ and $\kvar_2$ are only
-used in covariant position, and can thus be replaced by their lower bound.
-By removing the unneeded quantifier, we obtain the equivalent type:
+We can further simplify this type by exploiting variance. As $\kvar_1$
+and $\kvar_2$ are only used in covariant position, they can be
+replaced by their lower bounds, $\kvar_3$ and $\kun$. 
+By removing the unused quantifiers, we obtain a much simplified equivalent type:
 $$

### 2020-02-28 kindly/infer/infer.tex
Adapt the figures to fit better.
soundness

@@
 An important contribution of \affe is its principal type inference.
-We now formulate our inference algorithm
+Our type inference algorithm is
 based on the \hmx framework~\citep{DBLP:journals/tapos/OderskySW99},
 a Hindley-Milner type system for a language
-with qualified types where constraints are expressed in an arbitrary
+with constrained types where constraints are expressed in an arbitrary
 theory $X$.
-Most importantly,
-\hmx ensures principal type inference
-when $X$ respects the specified properties.
+%Most importantly,
+If $X$ has certain properties, then
+\hmx guarantees principal type inference.
 We apply \hmx to a concrete constraint language which we name $\CL$.
-We adapt and extend \hmx's existing rules to support kind inference,
-track linearity and handle borrows and regions. Finally, we
-formulate an appropriate constraint solving algorithm for $\CL$
-which solves and simplifies constraints in a principled way.
+We adapt and extend \hmx's rules to support kind inference,
+track linearity, and handle borrows and regions. Finally, we
+formulate constraint solving and simplification algorithms for $\CL$.
 
@@
 
-In the context of inference, it's critical to know which elements
-are input and output of inference judgements.
-In the following, when presenting a new judgement,
-we write input parameters in \textbf{\textcolor{DarkSlateBlue}{bold blue}}. The rest will be
-output parameters.
+In the context of inference, it is critical to know which elements
+are input and output of inference judgments.
+In the following, when presenting a new judgment,
+we write input parameters in \textbf{\textcolor{DarkSlateBlue}{bold
+    blue}}. The remaining parameters are output parameters.
 
-\paragraph{Usage environments}
+\paragraph{Usage Environments}
 
-% One novelty of our inference  judgement now
+% One novelty of our inference  judgment now
 % returns a type environment, which we call ``usage environment'' and commonly
@@
 
-To determine if a variable is used in an affine manner, we must track
+To determine if a variable is used in an affine manner, we track
 its uses and the associated kinds. In the expression
@@
 %
-To infer such constraints, our inference judgement not only
-take an environment as parameter but also return an environment called
-``usage environment'', denoted $\Sv$,
+To infer such constraints, our inference judgment not only
+takes an environment as parameter but also returns a \emph{usage
+  environment}, denoted $\Sv$, 
 which summarizes usages of variables and borrows.
-Usage environments follow the grammar
-of normal environments.
+Usage environments are defined like normal environments.
 %
-In \cref{sdtyping}, we used relations to split environments in two and to
+In \cref{sdtyping}, we use relations to split environments and to
 transform suspended bindings into borrows inside a region.
-These relations take as argument a constraint which validates
+These relations take a constraint parameter which validates
 the transformations.
-In the context of inference, we define new judgements which \emph{infer}
+In the context of inference, we define new judgments which \emph{infer}
 the constraints.
@@
 
-\paragraph{Constraint normalization}
+\paragraph{Constraint Normalization}
 
-The \hmx framework assumes the existence of a ``$\operatorname{normalize}$''
-function which takes a constraint $\inP{C}$, a substitution $\inP\psi$ and returns a
+The \hmx framework assumes the existence of a function
+``$\operatorname{normalize}$'' which takes a constraint $\inP{C}$ and a
+substitution $\inP\psi$ and returns a 
 simplified constraint $C'$
@@
 assume the existence of such a function for our constraint system
-and delay details to \cref{infer:solving}.
+and defer details to \cref{infer:solving}.
 
-\subsection{Type inference}
+\subsection{Type Inference}
 
@@
 
-\paragraph{Environments and bindings}
+\paragraph{Environments and Bindings}
 \label{infer:envs}
@@
 direction: we collect data from the leaves and enforce linearity
-at \emph{binders}. This is reflected in the {\sc Var$_I$} and
+at \emph{binders}. This policy is reflected in the {\sc Var$_I$} and
 {\sc Abs$_I$} rules.
-Typing for variables is very similar to traditional (non-affine) HM.
-The difference is that we record
+Typing for variables is very similar to traditional Hindley-Milner
+type inference. To keep track of linearity, we record
 that $x$ was used with the scheme $\schm$ by returning
@@
 This usage environment is in turn used at each binder to enforce proper
-usage of linear variable via the $\Weaken$ property.
-We demonstrate this for lambda expressions
-in the {\sc Abs$_I$} rule.
+usage of linear variable via the $\Weaken$ property as shown for
+lambda expressions in the {\sc Abs$_I$} rule.
 First, we typecheck the body of the lambda and obtain a usage
-environment $\Sv_x$. Like in the syntax-directed type system,
+environment $\Sv_x$. As in the syntax-directed type system,
 we introduce the constraint
@@
 the body of the lambda expression. We then introduce the constraint
-$\Weaken_{\bvar{x}{\sigma}}(\Sv)$, which is false only if weakening
-is necessary-yet-impossible due to linearity.
-The $\Weaken$ constraint is introduced at each binding point in the program.
+$\Weaken_{\bvar{x}{\sigma}}(\Sv)$, which fails if we try
+to abandon a linear variable. 
+The $\Weaken$ constraint is introduced at each binding construct.
 %
@@
 
-\subsection{Constraint solving}
+\subsection{Constraint Solving}
 \label{infer:solving}
@@
 
-For concision, we demonstrate our constraint solving
-algorithm on an example. The complete constraint system is
+For concision, we demonstrate the constraint solving
+algorithm with an example. The complete constraint system is
 defined in \cref{appendix:constraints}.
@@
 
-We now state that our extended inference algorithm still satisfies the soundness
+The extended inference algorithm still satisfies the soundness
 and completeness properties of \hmx.
@@
     \vee \svar{x}{\tau}^n\in\E
-    \right\}\\
-  &\cup \left\{ \bvar{\tvar}{k} \mid \bvar{\tvar}{k}\in\E \right\}
+    \right\}
+     \cup \left\{ \bvar{\tvar}{k} \mid \bvar{\tvar}{k}\in\E \right\}
 \end{align*}
@@
 \begin{theorem}[Principality]
-  Let $\inferS{\Ctrue}{\E}{e}{\schm}$ a closed typing judgement.
+  Let $\inferS{\Ctrue}{\E}{e}{\schm}$ a closed typing judgment.
   Then $\inferW{\Sv}{(C,\unif)}{\Eflat\E}{e}{\tau}$

### 2020-02-28 kindly/soa.tex
rectifications

--- 2019-11-23 kindly/soa.tex
@@
 
-Many systems have attempted to combine
+Many systems propose combinations of
 functional programming and linear types in a practical setting.
-One contribution of \lang is to combine several ingredients
-from these different languages while still preserving
+\lang is to combine key ingredients
+from these proposals while still preserving
 complete type inference.
-Many of the following languages supports linear or affine types, but rarely
+Many of the following languages support linear or affine types, but rarely
 both. In many cases, it is easy to adapt a system to support both, as
@@
 between linear and unrestricted types.
-\citet{DBLP:conf/tldi/MazurakZZ10} provide
+The authors provide
 a linearity-aware semantics with a soundness proof.
@@
 
-Quill~\citep{DBLP:conf/icfp/Morris16} is a Haskell-like language with affine
-types through the use of qualified types.
+Quill~\citep{DBLP:conf/icfp/Morris16} is a Haskell-like language with linear
+types.
 Quill does not expose a kind language, but
-uses typeclass-like mechanism and annotation on arrows.
-It supports both a most general type for function composition and
-principal type inference.
-We were able to express the type signatures of all Quill examples
-in \lang by leveraging kind inequalities.
-Unlike Quill,
-we provide a linearity-aware semantics and prove its soundness.
+uses the framework of qualified types to govern linearity annotations on arrows.
+Its type inference algorithm is proven sound and complete.
+\lang infers type signatures for all Quill examples, but often with
+simpler types because Quill does not support subkinding.
+Quill comes with a linearity-aware semantics and soundness proof.
 Quill does not support borrows.
@@
 affine types and ML modules, similar to \lang.
-Their system often relies on existential types to track the identity
+Alms examples often rely on existential types to track the identity
 of objects. For instance
 \lstinline/Array.create : int -> 'a -> \E 'b. ('a, 'b) array/ where
-\lstinline/'b/ uniquely identify the array.
+\lstinline/'b/ uniquely identifies the array.
 Due to the reliance on existentials, Alms does not support complete type inference.
-Furthermore, Alms does not support borrow and often rely
-on manual passing of capabilities.
+Furthermore, Alms does not support borrows and often relies
+on explicit capability passing.
 In our experience, \affe's limited support for existential types through
-regions is sufficient to express Alms' examples and leads to
+regions is sufficient to express many of Alms' examples and leads to
 a more convenient programming style for imperative code.
 %
-Kind-wise, Alms relies
-on unions, intersections and dependent kinds while
-\lang uses qualified types.
+Alms kind structure features unions, intersections and dependent kinds while
+\lang uses constrained types.
 We believe most of Alms' kind signatures can be expressed equivalently in
@@
 (where $\alpha$ and $\beta$ are types and $\Pi$ is the dependent function)
-in Alms and $\kvar\to\kvar\to\kvar$ in \lang thanks
+in Alms compared to $\kvar\to\kvar\to\kvar$ in \lang thanks
 to subkinding.
@@
 (LH) is to retrofit linear types to Haskell.
-LH introduces a new notion
-of ``linear arrows'', written $\multimap$ similar to linear logic,
+LH relies on ``linear arrows'', written $\multimap$ as in linear logic,
 which are functions that \emph{use} their argument exactly once.
@@
 \item
-  The GHC implementation of LH performs type inference, but
-  it is neither formalized nor complete.
+  While the LH paper specifies a full type system along with a
+  linearity-aware soundness proof, there is neither formal description of
+  the type inference algorithm nor a proof of the properties of inference.
+  However, there is an implementation of the inference as part of GHC.
 \item
@@
 for safe low-level programming.
-\lang is inspired from Rust's borrowing system to apply
-it in a functional setting with type inference, a GC, and
-an ML module system.
+\lang is inspired by Rust's borrowing system and transfers some of its
+ideas  to a functional setting with type inference, garbage collection, and
+an ML-like module system.
 Rust's lifetime system is more explicit and more expressive than \lang,
@@
 We believe this work
-could help us improve our handling of regions and of successive exclusive borrows.
+could help improve our handling of regions and of successive exclusive borrows.
 
@@
 These results make us confident that the system developed in \lang
-could be applied to larger and more complex languages such as OCaml.
+could be applied to larger and more complex languages such as OCaml
+and the full range of features based on ad-hoc polymorphism.
 % Alternatively, we could have based \lang on qualified
@@
 is at the leaves.
-Such discipline is known to be friendly to inference and has been used in many
+Such a discipline is known to be friendly to inference and has been used in many
 contexts, including OCaml, and has been combined

### 2020-02-28 kindly/motivation-pool.tex
Improve layout of figures in the appendix.

--- 2019-11-19 kindly/motivation-pool.tex
@@
   \centering
-  \begin{subfigure}[t]{1\linewidth}
+  \begin{subfigure}[t]{0.5\linewidth}
 \begin{lstlisting}
@@
 \end{lstlisting}
-    \vspace{-10pt}
+    \vspace{-8pt}
     \caption{Signature}
@@
 \end{lstlisting}
-    \vspace{-10pt}
+    \vspace{-8pt}
     \caption{Example of use}
@@
   \end{subfigure}
-
-  \begin{subfigure}[t]{1\linewidth}
+  \hfill
+  \begin{subfigure}[t]{0.45\linewidth}
 \begin{lstlisting}
-type ('a:'k) pool : 'k = {spawn: unit -> 'a; queue: 'a CQueue.t}
-let create spawn = { spawn ; queue = CQueue.create () }
+type ('a:'k) pool : 'k =
+  { spawn: unit -> 'a; queue: 'a CQueue.t }
+let create spawn = 
+  { spawn ; queue = CQueue.create () }
 let consume f c = CQueue.iter f c.queue
@@
   let r = {| f &!o |} in(*@\label{line:pool:region}*)
-  Queue.push o &queue;
+  CQueue.push o &queue;
   r
 \end{lstlisting}
-    \vspace{-15pt}
+    \vspace{-8pt}
     \caption{Implementation}

### 2020-02-28 kindly/extensions.tex

\lstMakeShortInline[keepspaces,basicstyle=\small\ttfamily]@
\section{Extensions}

\subsection{Algebraic Data Types}

Algebraic Data Types are a staple of functional programming and fit nicely
in our paradigm. Indeed, it is sufficient to check that the kinds of constructors
are strictly less than the kind of the datatype.
For instance, it is allowed to define a linear datatype containing unrestricted
parts.
On the contrary, it is forbidden to include affine elements in an unrestricted
datatype.
Here is the definition of a linear lists. The objects inside the list do not have to be linear.

\begin{lstlisting}
type ('a : 'k) llist : lin = Nil | Cons of ('a * 'a llist)
\end{lstlisting}

This extension is implemented in our prototype.

\subsection{Ad-hoc polymorphism and borrows}

In our formalization, we use two operators, $\borrow{x}$ and $\reborrow{x}$ to
distinguish between borrows and borrows of borrows.
Such a distinction is inconvenient for programming.
Using typeclass-like mechanisms, we can replace these operators
by a single one, $\borrow{x}$, which expects $x$ to be @Borrowable@ and
would then desugar to the more precise operators.
A similar solution is used in Rust through the @Borrow@ and @Defer@
traits.
It also allows method calls on objects without
explicit borrows, such as @foo.len()@ where len expects a shared borrow.

Ad-hoc polymorphism fits nicely in the framework of constrained
types and would preserve all the good properties of our language such
as principal type inference. Its soundness is orthogonal to linear types
and has been explored in the literature~\citep{DBLP:journals/scp/Jones94}.

\subsection{Non lexical lifetimes}

A recent important addition to Rust in the notion of ``non-lexical lifetimes''.
It allows code where borrowing doesn't follow
lexical scoping:

\begin{lstlisting}
let a = &x in
f a;
g (&!x)
\end{lstlisting}

Such patterns are dynamically safe, since $a$ is not used after
the second line.
Non-lexical lifetimes handles such patterns by ``removing'' the instructions
that do not mention $a$ from its region, here the last one.
%
In \lang, regions are purely lexical and marked by the
expressions $\region{b}{e}$.
During inference, region typing is only done when *existing* the region:
we modify the usage environment and introduce a kind constraint to prevent
escaping.

We can thus replace the lexical region by
an annotation on each subexpression indicating which borrowing it is associated
to. When exploring a subexpression, we compare the annotations, and automatically
apply the $\textsc{Region}$ rule when they differ.
%
This is equivalent to inlining the $\textsc{Region}$ rule in all the other
rules.

On the program above, the first two line will be annotated
to be ``in the region associated with @&x@'', but the last line will not.
When typing the sequence, since the annotations are different, we will apply
the region rule to the right hand side.

\lstDeleteShortInline@

%%% Local Variables:
%%% mode: latex
%%% TeX-master: "main"
%%% End:

### 2020-02-28 kindly/soa.tex

@@
 (LH) is to retrofit linear types to Haskell.
-LH relies on ``linear arrows'', written $\multimap$ as in linear logic,
+Unlike the previously discussed approaches, LH relies on ``linear
+arrows'', written $\multimap$ as in linear logic, 
 which are functions that \emph{use} their argument exactly once.
-% By contrast, in \lang linearity is only decided
-% by the kinds, and $\tarr{\klin}$ simply denotes \emph{single-use functions}.
 This design is easy to retrofit on top of an existing compiler
@@
   the type inference algorithm nor a proof of the properties of inference.
+  Subsequent work~\cite{DBLP:journals/corr/abs-1911-00268}
+  formalizes the inference for rank 1 qualified-types.
   However, there is an implementation of the inference as part of GHC.
@@
 formalized Rust's ownership discipline, including non-lexical lifetimes.
-We believe this work
-could help improve our handling of regions and of successive exclusive borrows.
+Another important difference is that Rust allows
+to precisely specify the memory layout of objects
+and to pass arguments by value.
+These features are crucial for the efficiency goals of Rust.
+In \lang, we assume that all arguments are passed by reference and
+that a garbage collector is available. This forgoes numerous
+issues regarding interior mutability and algebraic data types.
+In particular, it
+allows us to easily nest mutable references inside objects, regardless
+whether they are linear or unrestricted.
 

### 2020-02-28 kindly/extensions.tex
cosmetics

@@
 
-Algebraic Data Types are a staple of functional programming and fit nicely
+Algebraic data types are a staple of functional programming and fit nicely
 in our paradigm. Indeed, it is sufficient to check that the kinds of constructors
 are strictly less than the kind of the datatype.
-For instance, it is allowed to define a linear datatype containing unrestricted
+For instance, we may define a linear datatype containing unrestricted
 parts.
@@
 
-\subsection{Ad-hoc polymorphism and borrows}
+\subsection{Ad-hoc Polymorphism and Borrows}
 
@@
 Such a distinction is inconvenient for programming.
-Using typeclass-like mechanisms, we can replace these operators
-by a single one, $\borrow{x}$, which expects $x$ to be @Borrowable@ and
+Using a typeclass-like mechanism, we can replace these operators
+by a single overloaded operator, $\borrow{x}$, which expects $x$ to be @Borrowable@ and
 would then desugar to the more precise operators.
@@
 traits.
-It also allows method calls on objects without
+This approach also enables method calls on objects without
 explicit borrows, such as @foo.len()@ where len expects a shared borrow.
@@
 Ad-hoc polymorphism fits nicely in the framework of constrained
-types and would preserve all the good properties of our language such
+types and preserves all properties of our language such
 as principal type inference. Its soundness is orthogonal to linear types
@@
 
-\subsection{Non lexical lifetimes}
+\subsection{Non-Lexical Lifetimes}
 
-A recent important addition to Rust in the notion of ``non-lexical lifetimes''.
-It allows code where borrowing doesn't follow
-lexical scoping:
+A recent important addition to Rust is the notion of non-lexical lifetimes.
+With this feature code is acceptable even if borrowing does not respect
+lexical scoping as in this example:
 
@@
 
-Such patterns are dynamically safe, since $a$ is not used after
+Patterns like this one are dynamically safe because $a$ is not used after
 the second line.
-Non-lexical lifetimes handles such patterns by ``removing'' the instructions
-that do not mention $a$ from its region, here the last one.
+Non-lexical lifetimes handle such patterns by removing expressions
+that do not mention $a$ from its region; in this example,  the
+expression on the last line.
 %
 In \lang, regions are purely lexical and marked by the
-expressions $\region{b}{e}$.
-During inference, region typing is only done when *existing* the region:
+expression $\region{b}{e}$.
+During inference, region typing is only done when exiting the region:
 we modify the usage environment and introduce a kind constraint to prevent
@@
 %
-This is equivalent to inlining the $\textsc{Region}$ rule in all the other
+This approach is equivalent to inlining the $\textsc{Region}$ rule in all the other
 rules.
 
-On the program above, the first two line will be annotated
-to be ``in the region associated with @&x@'', but the last line will not.
+Applied to the program above, only the first two lines would be annotated
+to be ``in the region associated with @&x@'', but not the last line.
 When typing the sequence, since the annotations are different, we will apply

### 2020-02-28 kindly/extensions.tex
Fix some bits about non-lexical regions.

@@
 
-\subsection{Algebraic Data Types}
+\subsection{Algebraic Datatypes}
 
 Algebraic data types are a staple of functional programming and fit nicely
-in our paradigm. Indeed, it is sufficient to check that the kinds of constructors
-are strictly less than the kind of the datatype.
-For instance, we may define a linear datatype containing unrestricted
-parts.
-On the contrary, it is forbidden to include affine elements in an unrestricted
-datatype.
-Here is the definition of a linear lists. The objects inside the list do not have to be linear.
+in our paradigm. Indeed, it is sufficient to ensure  that the kinds of
+the constructor arguments are less than or equal to the kind of the datatype.
+Hence, it is forbidden to include affine elements in an unrestricted
+datatype, whereas the elements in a linear list may be linear or
+unrestricted. 
+Here is the definition of a datatype for linear lists.
 
@@
 
-Ad-hoc polymorphism fits nicely in the framework of constrained
+Ad-hoc polymorphism fits demonstrably in the \hmx framework of constrained
 types and preserves all properties of our language such
 as principal type inference. Its soundness is orthogonal to linear types
-and has been explored in the literature~\citep{DBLP:journals/scp/Jones94}.
+and has been explored in the literature~\citep{DBLP:conf/fpca/OderskyWW95}.
 
-\subsection{Non-Lexical Lifetimes}
+\subsection{Non-Lexical Regions}
 
@@
 to be ``in the region associated with @&x@'', but not the last line.
-When typing the sequence, since the annotations are different, we will apply
-the region rule to the right hand side.
+Thanks to these annotations, when type checking the sequence we check
+that the borrow does not escape the left-hand side (i.e., the second line).
 

### 2020-03-04 kindly/extensions.tex
linear lists
Small additions.

@@
 
+\subsection{Branching constructs}
+
+Our formalization of \lang does not contain branching constructs. As a consequence,
+subexpressions are always executed sequentially. This is of great importance
+for the splitting relation introduced in \cref{sdtyping}, as it must
+coincide with the order of evaluation. To account for borrows and linear
+values in different branches, we need a new (symmetric) join relation.
+This relation immediately follows from the semantics of borrows and linearity
+shown so far, and allow to easily define branching constructs such as if-then-else
+and pattern matching.
+This extension is implemented in our prototype.
+
 \subsection{Non-Lexical Regions}
@@
 
-Patterns like this one are dynamically safe because $a$ is not used after
+Such code patterns are dynamically safe since $a$ is not used after
 the second line.
-Non-lexical lifetimes handle such patterns by removing expressions
+Non-lexical lifetimes handle this by removing expressions
 that do not mention $a$ from its region; in this example,  the
@@
 expression $\region{b}{e}$.
-During inference, region typing is only done when exiting the region:
-we modify the usage environment and introduce a kind constraint to prevent
-escaping.
+During inference, we introduce kind constraint to prevent escaping from
+a region.
 
-We can thus replace the lexical region by
-an annotation on each subexpression indicating which borrowing it is associated
-to. When exploring a subexpression, we compare the annotations, and automatically
+We can replace the lexical region by
+an annotation on each expression indicating which borrows are live
+in this expression.
+When exploring a subexpression, we compare the annotations, and automatically
 apply the $\textsc{Region}$ rule when they differ.

### 2020-03-04 kindly/split-jam.sh
Pre-submit fixes.

@@ -1,2 +1,2 @@
-pdfjam -o paper.pdf main.pdf '1-14'
-pdfjam -o appendix.pdf main.pdf '15-'
+pdfjam -o paper.pdf main.pdf '1-25'
+pdfjam -o appendix.pdf main.pdf '26-'

### 2020-03-04 kindly/split.sh

@@ -1,2 +1,2 @@
-qpdf --pages main.pdf 1-14 -- main.pdf paper.pdf
-qpdf --pages main.pdf 15-z -- main.pdf appendix.pdf
+qpdf --pages main.pdf 1-25 -- main.pdf paper.pdf
+qpdf --pages main.pdf 26-z -- main.pdf appendix.pdf

### 2020-03-04 kindly/extensions.tex
NLL Extension: made it more clear that this is not currently implemented in Affe.

@@
 expression $\region{b}{e}$.
-During inference, we introduce kind constraint to prevent escaping from
+During inference, we introduce kind constraints to prevent escaping from
 a region.
 
-We can replace the lexical region by
-an annotation on each expression indicating which borrows are live
-in this expression.
-When exploring a subexpression, we compare the annotations, and automatically
+To add support for non-lexical lifetimes, we could replace the
+lexical region by an annotation on each expression indicating which borrows are
+live in this expression.
+When exploring a subexpression, we would compare the annotations, and automatically
 apply the $\textsc{Region}$ rule when they differ.
@@
 to be ``in the region associated with @&x@'', but not the last line.
-Thanks to these annotations, when type checking the sequence we check
+Thanks to these annotations, when type checking the sequence we would check
 that the borrow does not escape the left-hand side (i.e., the second line).

### 2020-03-04 kindly/motivation-pool.tex
fixed syntax rewriting

@@
 \begin{lstlisting}
-(* Using the pool in queries. *)
+(*Using the pool in queries.*)
 let create_user pool name =
@@
 let main users =
-  (* Create a database connection pool. *)
+  (*Create a database connection pool.*)
   let pool = Pool.create (fun _ -> Db.connect uri) in

### 2020-03-04 kindly/intro.tex
extensions

@@
 \item A prototype implementation of the type inference algorithm, including all
-  constraint simplification and extended with pattern matching,
-  available at \url{https://affe.netlify.com/}.
+  constraint simplification and extended with algebraic datatypes and
+  pattern matching,
+  available at \url{https://affe.netlify.com/}\footnote{This link is safe for double-blind.}.
 \end{itemize}
@@
 traits, etc is possible and orthogonal to the presentation in this paper. 
+While the system is geared towards type inference, it is nonetheless
+compatible with type annotations and thereby amenable to extensions
+where type inference may no longer be possible.
 

### 2020-04-23 kindly/icfp20-rebuttal.md
First version of the ICFP rebuttal.

We thank the reviewers for their thoughtful comments.
We will take their comments into account and revise our article as follows.
- Expand and improve our examples
- Precise further the limitations of our system and the comparison to other systems

# Reviewer A

> The use of "absolute" natural numbers to identify regions seems not modular.

Affe requires that each region is identified by an index along with a partial
order on indices that is compatible with the nesting of regions.
This can be implemented in many ways, including region variables
often used in algebraic effects systems, existentials, etc.

Natural numbers are the absolute simplest way to implement such indices
which is why
we use them in the formalisation. Our formalisation only relies on
the partial order and could easily use any of these
more sophisticated systems. This would certainly be the case in a 
realistic implementation for OCaml, for instance, to leverage
the on-going work on algebraic effects.
The properties of the system would stay identical.

We plan to add this point to section 6.

> It is very hard to get a grasp what kind of programming idioms can be type checked using Affe.

Affe can easily express rich data-structures (notably 
persistent ones) or manipulate resources and protocols (for instance with session types).

The main limitations of Affe is that it is not flow sensitive.
Code idioms that relies of subtle flow-sensitive usage of 
permissions and linearity will not work in Affe. Such patterns often
do not work in Rust either, and require richer logic more akin to theorem
proving, such as Mezzo.

Rust slightly relaxes these constraints through the use of non-lexical lifetimes
and double-borrows. 
Following the work by Weiss et al. [38] on the Rust semantics, we believe
these techniques can also be applied to Affe and 
sketch how in section 6.

> What is the reason the borrow is fixed to `U` in the splitting rules `SuspB` and `SuspS` (page 10).

Peter/Hannes ? Is this due to a proof thingy ? In principle, doing an exclusive reborrow is ok.

> Is Affe's kind mechanism also powerful enough to keep track of things like `Send` and `Sync`?

This is probably better implemented via ad-hoc polymorphism. It is very
desirable to extend Affe with some form of typeclasses, as we point out in
Section 6.2, which would allow to implement such mechanisms.

# Reviewer B

> What could go wrong, for example, if the set function of the array module would use a non-exclusive borrow type in the argument?

This question should be reframed in a larger context: "What does my API guarantee ?".
For instance, ML references and arrays allow concurrent writes in a perfectly
safe manner. In a more general concurrent context, 
many data-structures are designed to allow concurrent mutations, via mutexes for
instance.

However, if the API prevents concurrent mutations, this ensures that no data-races
can happen on this data-structure. Data-races are often the source of subtle
and hard-to-fix bugs.
Exclusive borrow allows the author to choose if they want or not to provide
such guarantee. This has been used with huge success in the Rust community to 
write safe and fast concurrent programs.

# Reviewer C

> in other systems, one might use existential quantification tricks or magic wands to express temporarily borrowing an *element* of an array before returning the element to the array. What would be the equivalent here?

Indeed, this is something that we would like to add to the system. 
Such functions can already be implemented via closures, 
as demonstrated in the iteration examples in section 2.3:

    val get_borrow :
      ('b <= affₖ) < &('a t) * int * (&(linₖ₊₁, 'a) -{linₖ₊₁}> 'b) -> 'b

This API is however a bit inconvenient. We can improve it using
so called "existential tricks" to allow matching on borrows:

    let &x = get_borrow (a,3) in (* &x must not escape here *)

In general, existentials tricks are not incompatible with Affe and are in fact
one of the foremost way to extend its expressivity.

> Do you envision a similar discipline for your system where one might write, say, a doubly-linked list in regular GC'd ML, then assume an interface for it that exposes it as a resource?

This pattern is indeed essential and was well demonstrated by Alms.
Affe already allow such patterns via subkinding and module type abstraction:
a type can be defined as unrestricted and exposed as linear.

Integrating "forced freeing" of such objects with the GC is more delicate
and is explored by Munch-Maccagnoni [21].

### 2020-04-25 kindly/icfp20-rebuttal.md
Some small fixes to the rebuttal.

@@
 
-Affe requires that each region is identified by an index along with a partial
+Affe requires that each region is identified by an index, along with a partial
 order on indices that is compatible with the nesting of regions.
 This can be implemented in many ways, including region variables
-often used in algebraic effects systems, existentials, etc.
+often used in algebraic effect systems, existentials, etc.
 
@@
 which is why
-we use them in the formalisation. Our formalisation only relies on
+we use them in the formalization. Our formalization only relies on
 the partial order and could easily use any of these
@@
 realistic implementation for OCaml, for instance, to leverage
-the on-going work on algebraic effects.
+the ongoing work on algebraic effects.
 The properties of the system would stay identical.
@@
 The main limitations of Affe is that it is not flow sensitive.
-Code idioms that relies of subtle flow-sensitive usage of 
+Code idioms that relies on subtle flow-sensitive usage of 
 permissions and linearity will not work in Affe. Such patterns often
@@
 
-This question should be reframed in a larger context: "What does my API guarantee ?".
-For instance, ML references and arrays allow concurrent writes in a perfectly
-safe manner. In a more general concurrent context, 
-many data-structures are designed to allow concurrent mutations, via mutexes for
-instance.
+This refers to a larger question: "What does my API guarantee ?".
+ML references and arrays allow concurrent writes but are subject to data-races.
+Data-races are the source of subtle and hard-to-fix bugs.
 
-However, if the API prevents concurrent mutations, this ensures that no data-races
-can happen on this data-structure. Data-races are often the source of subtle
-and hard-to-fix bugs.
-Exclusive borrow allows the author to choose if they want or not to provide
-such guarantee. This has been used with huge success in the Rust community to 
-write safe and fast concurrent programs.
+One can prevent such issues dynamically with mutexes. Exclusive borrows
+aims to forbid concurrent mutations *statically*. The choice then falls on the 
+API designer: a file handle is probably better checked statically.
+A distributed queue on the other hand would 
+feature a `push` function taking a shared borrow precisely
+to allow concurrent writes.
+
+While a full description of a concurrent story for Affe is out of scope 
+of this paper, such concepts have been used with huge success in the Rust
+community to write safe and fast concurrent programs.
 
@@
     val get_borrow :
-      ('b <= affₖ) < &('a t) * int * (&(linₖ₊₁, 'a) -{linₖ₊₁}> 'b) -> 'b
+      ('b <= linₖ) => &('a t) * int * (&(linₖ₊₁, 'a) -{linₖ₊₁}> 'b) -> 'b
 
@@
 
-In general, existentials tricks are not incompatible with Affe and are in fact
-one of the foremost way to extend its expressivity.
+In general, existential types are not incompatible with Affe and are in fact
+one of the first way to extend its expressivity. They also happen to be already implemented in several ML-family languages (OCaml, notably).
 
@@
 This pattern is indeed essential and was well demonstrated by Alms.
-Affe already allow such patterns via subkinding and module type abstraction:
+Affe already allows such patterns via subkinding and module type abstraction:
 a type can be defined as unrestricted and exposed as linear.
 
-Integrating "forced freeing" of such objects with the GC is more delicate
-and is explored by Munch-Maccagnoni [21].
+Integrating "forced freeing" of such objects with the GC is more delicate,
+and we hope to reuse the work by Munch-Maccagnoni [21] for this purpose.

### 2020-04-25 kindly/icfp20-rebuttal.md
updated

--- 2020-04-23 kindly/icfp20-rebuttal.md
@@
 - Expand and improve our examples
-- Precise further the limitations of our system and the comparison to other systems
+- Provide a comparison table with other systems 
+- Further discuss limitations of our system
 
@@
 
-Affe requires that each region is identified by an index along with a partial
-order on indices that is compatible with the nesting of regions.
-This can be implemented in many ways, including region variables
-often used in algebraic effects systems, existentials, etc.
+Affe requires that each region is identified by an index drawn from a partial
+order that is compatible with the nesting of regions.
+This order can be implemented in many ways, including region variables
+as often used in algebraic effects systems, existentials, etc.
 
-Natural numbers are the absolute simplest way to implement such indices
-which is why
-we use them in the formalisation. Our formalisation only relies on
-the partial order and could easily use any of these
-more sophisticated systems. This would certainly be the case in a 
-realistic implementation for OCaml, for instance, to leverage
-the on-going work on algebraic effects.
-The properties of the system would stay identical.
+For simplicity, the formalization uses the concrete implementation with 
+natural numbers for indices. Our proofs only rely on
+the partial order and could be adapted to one of the
+more abstract approaches.
 
@@
 
-The main limitations of Affe is that it is not flow sensitive.
-Code idioms that relies of subtle flow-sensitive usage of 
-permissions and linearity will not work in Affe. Such patterns often
-do not work in Rust either, and require richer logic more akin to theorem
-proving, such as Mezzo.
+The main limitation of Affe is that it is not flow sensitive.
+Code idioms that rely on subtle flow-sensitive usage of 
+permissions and linearity do not typecheck in Affe. Such patterns often
+do not work in Rust either and require a richer logic as in Mezzo.
 
@@
 
-Peter/Hannes ? Is this due to a proof thingy ? In principle, doing an exclusive reborrow is ok.
+For SuspB it should be ok to relax the `U` to `b`. For SuspS, the idea
+is that an exclusive borrow is affine. It would most likely be fine in
+the present system, but it wouldn't scale to a concurrent setting.
+We will check the proofs, implement the most general setting in the
+rule, and comment on possible restrictions in a concurrent setting.
 
@@
 
-This question should be reframed in a larger context: "What does my API guarantee ?".
+This question should be reframed in a larger context: "What guarantee
+is offered by an API?".
 For instance, ML references and arrays allow concurrent writes in a perfectly
@@
 This API is however a bit inconvenient. We can improve it using
-so called "existential tricks" to allow matching on borrows:
+so called "existential tricks" to match on borrows:
 
@@
 
-This pattern is indeed essential and was well demonstrated by Alms.
-Affe already allow such patterns via subkinding and module type abstraction:
+This pattern is indeed essential and is well demonstrated by Alms.
+Affe allows such patterns via subkinding and module type abstraction:
 a type can be defined as unrestricted and exposed as linear.

### 2020-04-25 kindly/icfp20-rebuttal.md
merged
formatting

@@
 
-This question should be reframed in a larger context: "What guarantee
-is offered by an API?".
-For instance, ML references and arrays allow concurrent writes in a perfectly
-safe manner. In a more general concurrent context, 
-many data-structures are designed to allow concurrent mutations, via mutexes for
-instance.
+The real question here is: "What guarantee is offered by an API?".
+ML references and arrays permit concurrent writes but are prone to
+data races, which give rise to subtle bugs.
 
-However, if the API prevents concurrent mutations, this ensures that no data-races
-can happen on this data-structure. Data-races are often the source of subtle
-and hard-to-fix bugs.
-Exclusive borrow allows the author to choose if they want or not to provide
-such guarantee. This has been used with huge success in the Rust community to 
-write safe and fast concurrent programs.
+One can prevent such issues dynamically with mutexes, whereas exclusive borrows
+forbid concurrent mutations *statically*. The choice then falls on the 
+API designer: a file handle is probably better checked statically.
+A distributed queue on the other hand would 
+feature a `push` function taking a shared borrow precisely
+to permit concurrent writes.
+
+While a full description of a concurrent story for Affe is out of scope 
+of this paper, such concepts have been used with huge success in the Rust
+community to write safe and fast concurrent programs.
 
@@
 This pattern is indeed essential and is well demonstrated by Alms.
-Affe allows such patterns via subkinding and module type abstraction:
+Affe supports such patterns via subkinding and module type abstraction:
 a type can be defined as unrestricted and exposed as linear.
 
-Integrating "forced freeing" of such objects with the GC is more delicate
-and is explored by Munch-Maccagnoni [21].
+Integrating "forced freeing" of such objects with the GC is more delicate,
+and we hope to reuse the work by Munch-Maccagnoni [21] for this purpose.

### 2020-04-25 kindly/icfp20-rebuttal.md
Add the merge list example.

@@
 do not work in Rust either and require a richer logic as in Mezzo.
+As a simple example, Affe can not directly type functions such as the 
+merge on linear lists:
+```
+let rec merge l1 l2 = match l1, l2 with
+  | h1::t1, h2::t2 ->
+    if &h1 < &h2 
+    then h1::(merge t1 l2) (* Must expand l2 to h2::t2 here *)
+    else h2::(merge l1 t2)
+  | ....
+```
 
@@
 these techniques can also be applied to Affe and 
-sketch how in section 6.
+sketch how in section 6, which would help for the example above.
 

### 2020-05-15 kindly/intro.tex
first round of revision

@@
 \begin{itemize}
-\item @withFile : path -> (handle o-> r) -> r@, which creates a new
-  file handle and takes a continuation that uses the @handle@ linearly,
+\item @withFile : path -> (handle o-> Unrestricted r) -> r@, which creates a new
+  file handle and takes a continuation that uses the @handle@
+  linearly, but returns an unrestricted value,
 \item @writeFile : string -> handle o-> handle@, which
@@
 \begin{itemize}
-\item A type system encoding linearity and affinity with
-  borrowing. The type system is polymorphic for types and
-  for kinds that express constraints on the number of uses of a value.
+\item A polymorphic type system that encodes linearity and affinity with
+  borrowing in lexical regions. Polymorphism covers types and kinds that
+  express multiplicity constraints on the number of uses of a
+  value. This type system is a conservative extension of systems for
+  existing ML-like languages.
+\item Expressive type soundness theorem with respect to a big-step linearity-aware semantics. 
 \item An extension of the \hmx framework
-  \cite{DBLP:journals/tapos/OderskySW99} with kinds, regions, and
-  multiplicity constraints that supports full, principal type inference.
+  \cite{DBLP:journals/tapos/OderskySW99} for constrained type
+  inference to equip the type system with full, principal type inference.
 \item Soundness proof of the inference algorithm.
-\item Big-step linearity aware semantics with an expressive soundness theorem.
-\item Constraint simplification techniques.
+% \item Constraint simplification techniques.
 \item Automatic inference of regions for borrows.
@@
   pattern matching,
-  available at \url{https://affe.netlify.com/}\footnote{This link is safe for double-blind.}.
+  available at \url{https://affe.netlify.com/}.
 \end{itemize}

### 2020-05-15 kindly/lang.tex
first round of revision

@@
 A borrow $\borrow{x}$ is always taken from a variable $x$. The
-\emph{borrow annotation} $\BORROW$ indicates whether the borrow is exclusive
-($\MBORROW$) or shared ($\IBORROW$). If $x$ is already borrowed,
+\emph{borrow annotation} $\BORROW$ indicates whether the borrow is exclusive/affine
+($\MBORROW$) or shared/unrestricted ($\IBORROW$). If $x$ is already borrowed,
 then we need to use the reborrow expression $\reborrow{x}$.
@@
 $\create$, $\observe$, $\update$ and $\destroy$.
+They serve a prototypes to demonstrate the typing and handling of
+resources. For a concrete type of resource, there are further
+arguments and perhaps different versions of the operations. But the
+typing and behavior of the operations is analogous to the prototype
+operations.
 

### 2020-05-15 kindly/motivation-persistent.tex
first round of revision

@@
 The signature differs slightly from the @Array@
-signature. For simplicity, the array elements are unrestricted, but
+signature. As our application requires the @get@ function, the array
+elements must be unrestricted, but
 the structure itself remains linear so as to be implemented in terms of @Array@.
@@
 modifications. The function
-@mapi: (int -> &'a -> 'b) -> &('a t) -> 'b t@
+@mapi: (int * &'a -> 'b) * &('a t) -> 'b t@
 is a simple variation on @Array.map@ where the mapping function also
@@
 \begin{lstlisting}
-module Cow : HYBARRAY = struct
+module CowArray : HYBARRAY = struct
   include Array
   let set (a, i0, x0) =
-    Array.mapi (fun i x -> if i = i0 then x0 else x) a
+    Array.mapi ((fun (i, x) -> if i = i0 then x0 else x), a)
   let set_mut = Array.set
@@
   \begin{subfigure}[t]{0.55\linewidth}
-  \lstinputlisting[linerange=14-31]{code/sukodu.affe}
+  \lstinputlisting[linerange=14-31,firstnumber=last]{code/sukodu.affe}
   \end{subfigure}
@@
 The board is represented as a 2D-matrix (\cref{line:boardtype}), where
-the @Matrix@ type uses the same API as @PersistArray@
+the @Matrix@ type uses the same API as @CowArray@
 but with two indices.

### 2020-05-15 kindly/motivation.tex
first round of revision

@@
   type ('a : 'k) t : lin
-  val create : ('a : un) => int -> 'a -> 'a t
+  val create : ('a : un) => int * 'a -> 'a t
   val free : ('a : aff) => 'a t -> unit
@@
   val set : ('a : aff) => &!('a t) * int * 'a -> unit
-  val map : (&'a -> 'b) -> &('a t) -> 'b t
+  val map : (&'a -> 'b) * &('a t) -> 'b t
 end
@@
 deallocation of arrays that may contain affine elements.
-A program would first use @create n v@ to create
+A program would first use @create (n, v)@ to create
 an array of size @n@ initialized with value
@@
 let mk_fib_array n =
-  let a = create n 1 in
+  let a = create (n, 1) in
   for i = 2 to n - 1 do
@@
 prevents access to the ``raw'' resource inside the loop's
-body. \cref{ex:array:get} contains two shared borrows
+body. In particular, the resource cannot be freed through a
+borrow. \cref{ex:array:get} contains two shared borrows 
 in the same expression which forms a region by itself. These borrows
@@
   for i = 2 to n - 1 do {|
-    let x = {| get &a (i-1) + get &a (i-2) |} in(*@\label{ex:array:region:get}*)
-    set &!a i x
+    let x = {| get (&a, i-1) + get (&a, i-2) |} in(*@\label{ex:array:region:get}*)
+    set (&!a, i, x)
   |} done;
@@
 \begin{lstlisting}
-val fold : ('a : un) => ('a -> 'b -> 'b) -> 'a array -> 'b -> 'b
+val fold : ('a : un) => ('a -> 'b -> 'b) -> 'a Array.t -> 'b -> 'b
 \end{lstlisting}
@@
   ('b:'k),('k <= aff_r) =>
-  (&(aff_r1,'a) -> 'b -{aff_r1}> 'b) -> &('k_1,'a array) -> 'b -{'k_1}> 'b
+  (&(aff_r1,'a) -> 'b -{aff_r1}> 'b) -> &('k_1,'a Array.t) -> 'b -{'k_1}> 'b
 \end{lstlisting}
@@
 \begin{lstlisting}
-val iter_mut : &!('a array) -> (&!(aff_r1,'a) -{un}> unit) -{aff}> unit
+val iter_mut : &!('a Array.t) -> (&!(aff_r1,'a) -{un}> unit) -{aff}> unit
 \end{lstlisting}
@@
 may be composed and optimized.
+It also enables different implementations such as
+using parallel iterations in the immutable case.
 \lang's region variables ensure that the content iterated on can never

### 2020-05-16 kindly/lang.tex
macro for borrow binding

@@
     |&\ \bvar{x}{\schm} \tag{Variables} \\
-    |&\ \bvar{\borrow{x}}{\borrowty{k}{\tau}} \tag{Borrows}\\
-    |&\ \svar{x}{\schm}^n \tag{Suspended}
+    |&\ \svar{x}{\schm}^n \tag{Suspended} \\
+    |&\ \bbvar x k \schm \tag{Borrows}
+    % \bvar{\borrow{x}}{\borrowty{k}{\tau}} \tag{Borrows}\\
     % |&\ \tydecl{t}{\kschm}{K}{\tau}\tag{Type Declaration}\\
   \end{align*}
-  \vspace{-10pt}
+  \vspace{-15pt}
   \caption{Type environments}
@@
         \lregion{C}{}
-        {\svar[\BORROW]{x}{\tau}^n}
-        {\bvar{\borrow[\BORROW]{x}}{\borrowty[\BORROW] k{\tau}}}
+        {\svar[\BORROW]{x}{\schm}^n}
+        {\bbvar x k \schm}
+        % {\bvar{\borrow[\BORROW]{x}}{\borrowty[\BORROW] k{\schm}}}
       }
@@
 cannot be used directly.
-A borrow binding, $\bvar{\borrow{x}}{\borrowty{k}{\schm}}$, replaces
+A borrow binding,
+$\bbvar x k \schm$,
+%$\bvar{\borrow{x}}{\borrowty{k}{\schm}}$,
+replaces
 such a suspended binding on entry to the $x$-region. It indicates
 that the borrow $\borrow{x}$ can be used directly. Kind $k$
-restricts the lifetime of the borrow to the region (see \cref{selectrules:borrow}).
+restricts the lifetime of the borrow to the region (see rules
+\TirName{Region} and \TirName{BorrowBinding} in
+\cref{selectrules:borrow} and the upcoming discussion of these rules).
 
@@
 corresponding to $\E$ where the suspended binding is replaced by the
-borrow binding  $\bvar{\borrow[\IBORROW]{c}}{\borrowty[\IBORROW]{k}{\text{channel}}}$.
+borrow binding  $\bbvar[\IBORROW]{c}{k}{\text{channel}}$.
 To constrain the borrow to this region we impose the constraint

### 2020-05-17 kindly/lang.tex
more comments by Rev#A processed

@@
 typing and behavior of the operations is analogous to the prototype
-operations.
+operations. Moreover, $\observe$ and $\update$ serve as eliminators
+for borrow types.
 
@@
       \and
+      \ruleSDReBorrow
+      \and
       \ruleSDLam
@@
 unrestricted ($\kun_0$) and which is equipped with the four
-operations described in \cref{motivation}: 
+operations introduced in \cref{syntax}: 
 \begin{itemize}[topsep=0pt]
@@
 \lstMakeShortInline[keepspaces,basicstyle=\normalsize\normalfont]@
+
+It is straightforward to give a standard semantics for \lang, but such
+a semantics would not be very informative. In this section, we give a
+big-step semantics that performs explicit bookkeeping of the number of
+times a value is used and of the mode in which a reference to a
+resource is used (e.g., borrowed or not). This bookkeeping is based on a set of permissions
+that regulate the currently allowed mode of access to resources and
+closures. It enables us to state and prove a highly informative type
+soundness result (see \cref{sec:metatheory}) with expressive invariants that ensure proper
+resource usage.
 

### 2020-05-17 kindly/soa.tex
macro for borrow binding
more comments by Rev#A processed

@@
 
-\paragraph{Substructural type-systems in functional languages}
+\subsection{Substructural type-systems in functional languages}
 
@@
 functional programming and linear types in a practical setting.
-\lang is to combine key ingredients
+The goal of \lang is to combine key ingredients
 from these proposals while still preserving
@@
 such as GHC, but has proven quite controversial\footnote{
-  See the in-depth discussion attached to the GHC proposal starting  \href{https://github.com/ghc-proposals/ghc-proposals/pull/111\#issuecomment-403349707}{here}.}.
+  See the in-depth discussion attached to the GHC proposal for LH on GitHub: \url{https://github.com/ghc-proposals/ghc-proposals/pull/111\#issuecomment-403349707}.}.
 Most relevant to \lang:
-\begin{itemize}[leftmargin=*,topsep=0pt]
+\begin{itemize}[leftmargin=*]
 \item LH does not admit subtyping for arrows and requires
@@
   LH promotes a continuation-passing style with functions such as
-  \lstinline/withFile : path -> (file ->. file) ->. unit/
+  \lstinline/withFile : path -> (file ->. Unrestricted r) ->. r/
   to ensure linear use of resources. This style leads to problems with
@@
 
-\paragraph{Other substructural type-systems}
+\subsection{Other substructural type-systems}
 
@@
 
-\paragraph{Type-system features}
+\subsection{Type-system features}
 %

### 2020-05-20 kindly/motivation.tex
linear array api extended

@@
   val set : ('a : aff) => &!('a t) * int * 'a -> unit
-  val map : (&'a -> 'b) * &('a t) -> 'b t
+  val map : ('a -> 'b) * 'a t -> 'b t (* destructive *)
+  val map_copy : (&'a -> 'b) * &('a t) -> 'b t (* copying *)
 end

### 2020-05-20 kindly/intro.tex
cosmetic changes

@@
 \begin{figure}[tp]
-  \begin{subfigure}[t]{0.53\linewidth}
+  \begin{subfigure}[t]{0.35\linewidth}
     \lstinputlisting{code/writefiles.affe}
@@
   \end{subfigure}~
-  \begin{subfigure}[t]{0.43\linewidth}
+  \begin{subfigure}[t]{0.3\linewidth}
 \begin{lstlisting}
@@
     \label{fig:writing-files-example}
+  \end{subfigure}~
+  \begin{subfigure}[t]{0.3\linewidth}
+\begin{lstlisting}
+let main () =
+  let h = File.fopen "foo" in
+  {| File.write &!h "Hello " |};(*@\label{line:hello}*)
+  {| File.write &!h "world!" |};(*@\label{line:world}*)
+  File.close h
+\end{lstlisting}
+    \vspace{-10pt}
+    \caption{File example with regions}
+    \label{fig:writing-files-example-region}
   \end{subfigure}
@@
 scoped program fragment from which the borrow must not escape. In
-\cref{fig:writing-files-example}, there are two regions, one 
+\cref{fig:writing-files-example}, there are two regions visualized in \cref{fig:writing-files-example-region}, one 
 consisting of \cref{line:hello} and another consisting of
@@
   file handle and takes a continuation that uses the @handle@
-  linearly, but returns an unrestricted value,
+  linearly, but returns an unrestricted value\footnote{For technical
+    reasons, LH  requires the programmer to use a type like
+    \lstinline/Unrestricted/ at this  point.},
 \item @writeFile : string -> handle o-> handle@, which
@@
 \end{lstlisting}
-The reading of the type inferred by our system is straightforward. If
+The reading of the inferred type is straightforward. If
 @f@ is a @'k@-restricted function, then so is
@@
 simplification is part of our type inference algorithm. As before,
-printing the type scheme only mentions the required constraint 
+printing the type scheme only mentions the non-trivial constraint 
 $\kappa\le\kappa_1$ and omits the prenex quantification over $\kappa,

### 2020-05-20 kindly/motivation.tex
explanation added

@@
 @v@. The value @v@ must be unrestricted as it is duplicated to
-initialize all array elements. Using the @map@ function we
+initialize all array elements. Using the @map@ and @map_copy@ functions we
 can transform the unrestricted elements into linear (affine)
-ones. The resulting array can be processed further using @map@ (or @set@ if
-affine). The @get@ function is only applicable if the element type
-is unrestricted as one element is duplicated. The @length@
+ones. The @map@ function consumes its input array, whereas @map_copy@
+preserves the input. The implementation of @map@ \emph{might} perform
+an in-place update, but this is neither required nor guaranteed. The
+resulting array can be processed further using @map@ or 
+@map_copy@ (or @set@ if affine). The @get@ function is only applicable
+if the element type is unrestricted as one element is duplicated. The @length@
 function is always applicable. To @free@ an array the elements must be
-affine.
+affine. An array with linear elements can only be freed by @map@ping
+an explicit destructor over its elements.
 

### 2020-05-20 kindly/infer/infer.tex
first pass reviewer #A

@@
 We adapt and extend \hmx's rules to support kind inference,
-track linearity, and handle borrows and regions. Finally, we
-formulate constraint solving and simplification algorithms for $\CL$.
+track linearity, and handle borrows and regions. We
+formulate constraint solving and simplification algorithms for
+$\CL$. Finally, we prove that the inference algorithm computes
+principal types.
 
@@
 In the following, when presenting a new judgment,
-we write input parameters in \textbf{\textcolor{DarkSlateBlue}{bold
-    blue}}. The remaining parameters are output parameters.
+we write input parameters in \textbf{\textcolor{ForestGreen}{bold
+    green}}. The remaining parameters are output parameters.
 
@@
   of constraints that must be respected.
-\item $\bregion[\inP{n}]{C}{\inP{x}}{\Sv}{\inP\Sv'}$.
+\item $\bregion[\inP{n}]{C}{\inP{x}}{\Sv}{\inP{\Sv'}}$.
   Given a usage environment $\inP{\Sv'}$, a nesting level $\inP{n}$,
@@
 $\bsplit{C}{\Sv}{\inP{\Sv_1}}{\inP{\Sv_2}}$
-and $\bregion{C}{\inP{x}}{\Sv}{\inP\Sv'}$,
+and $\bregion{C}{\inP{x}}{\Sv}{\inP{\Sv'}}$,
 which returns the necessary constraints.

### 2020-05-24 kindly/intro.tex
Worked on 'Typos and Focused Remarks' from Reviewer C.

@@
 the existing foreign function interface, but managing the underlying
-resources like malloc'd storage is not: It cannot be left to the
+resources like malloc'd storage is not: it cannot be left to the
 garbage collector, but proper release of the storage via calls to \texttt{free()} must be

### 2020-05-24 kindly/lang.tex
instance
Worked on 'Typos and Focused Remarks' from Reviewer C.

@@
   \lam{a}{}\region[1]{\Sone a\MBORROW}{\ &%
-  \letin{x}{\app{f}{\borrow[\kaff]{a}}}{}\\
+  \letin{x}{\app{f}{\borrow[\kun]{a}}}{}\\
   &\region[2]{\Sone x\MBORROW}{g\ (\borrow[\kaff]{x})};\\
@@
     \begin{mathpar}
+      \ruleInstance
+      \and
       \ruleSDVar
+      \and
+      \ruleSDLam
       \and
@@
       \ruleSDReBorrow
-      \and
-      \ruleSDLam
       \and
@@
 a pointer into a store, and a stack of modifiers that indicates the
-borrows that have been taken from the original object. From the raw
-location, we may take affine borrows and reborrows. Once we have
-taken an unrestricted borrow (from a raw location or a borrowed one),
-then we can take further unrestricted borrows from it, but no more
-affine ones.
+borrows and reborrows that have been taken from $\Loc$.
+Once we have taken an unrestricted borrow (from a raw location or a borrowed
+one), then we can take further unrestricted borrows from it, but no more affine
+ones.
 

### 2020-05-26 kindly/motivation.tex
fixing linear map

--- 2020-05-15 kindly/motivation.tex
@@
   val map : (&'a -> 'b) * &('a t) -> 'b t
+  val iter : ('a -> unit) * 'a t -> unit
 end
 \end{lstlisting}
+%  val map : ('a -> 'b) * 'a t -> 'b t (* destructive *)
+%  val map_copy : (&'a -> 'b) * &('a t) -> 'b t (* copying *)
   \vspace{-15pt}
@@
 aims to safely handle manual allocation and
-deallocation of arrays that may contain affine elements.
+deallocation of arrays that may contain affine or linear elements.
 A program would first use @create (n, v)@ to create
@@
 @v@. The value @v@ must be unrestricted as it is duplicated to
-initialize all array elements. Using the @map@ function we
-can transform the unrestricted elements into linear (affine)
-ones. The resulting array can be processed further using @map@ (or @set@ if
-affine). The @get@ function is only applicable if the element type
-is unrestricted as one element is duplicated. The @length@
-function is always applicable. To @free@ an array the elements must be
-affine.
+initialize all array elements.
+To @free@ an array the elements must be affine. 
+The @length@ function is always applicable.
+The @get@ function is only applicable
+if the element type is unrestricted as one element is duplicated.
+To @set@ an array element displaces the previous content, which must
+be at least affine.
+
+The @map@ function can transform arrays with arbitrary elements. In 
+particular, it can turn unrestricted elements into linear (affine)
+ones. It takes a borrow of the input array and returns a newly created
+output array. As @free@ing requires affine elements, we provide the
+@iter@ function which takes a
+suitable finalizer and an array with arbitrary elements, which is
+consumed. Indeed such an iteration is the only way to free an array
+with linear elements. A real-life
+API would provide a combination of @map@ and @iter@ as a
+``destructive'' map that consumes its input array. Assuming a uniform
+representation, such a destructive map might be implemented by
+in-place update.
 
@@
 Observer functions such as @length@ and @get@ expect a shared borrow
-argument, mutator functions such a @set@ expect
+argument, mutator functions such as @set@ expect
 an exclusive borrow.

### 2020-05-26 kindly/motivation.tex
folding

@@
 For instance, how should we fold on an array of linear objects?
-A naive fold function would only work on unrestricted content:
+Here is a first attempt at the type of a fold function:
 \begin{lstlisting}
-val fold : ('a : un) => ('a -> 'b -> 'b) -> 'a Array.t -> 'b -> 'b
+val fold : ('a : 'k) => ('a -> 'b -{'k}> 'b) -> 'a Array.t -> 'b -{lin}> 'b
 \end{lstlisting}
+This type puts no restrictions on the element type of the array, but
+it requires the @fold@ function to consume the array and all its
+elements. The last function arrow is linear because the array type (from
+\cref{sec:imper-progr}) is linear.
 
-If we want to work on linear and affine resources,
-we must make sure those resources are not leaked in the result.
-content. We obtain the following signature:
+If we want to work on borrows of linear and affine resources, then the
+typing gets more involved because
+we must make sure those borrows are not leaked in the result.
+We obtain the following signature for @bfold@, the borrowing fold operation:
 \begin{lstlisting}
-val fold :
-  ('b:'k),('k <= aff_r) =>
-  (&(aff_r1,'a) -> 'b -{aff_r1}> 'b) -> &('k_1,'a Array.t) -> 'b -{'k_1}> 'b
+val bfold : ('b:'k),('k <= aff_r) => (&(aff_r1,'a) -> 'b -{aff_r1}> 'b) -> &('k_1,'a Array.t) -> 'b -{'k_1}> 'b
 \end{lstlisting}
@@
 
-As an easy example, we fold over an array of files to compute the sum of their sizes:
+As an easy example, we fold over an array of files
+@all_files : File.t Array.t@ to compute the sum of their sizes:
 \begin{lstlisting}
-let total_size = fold (fun f s -> File.size f + s) &all_files 0
+let total_size_of files = bfold (fun f s -> File.size f + s) files 0
+let total_size = total_size_of &all_files
 \end{lstlisting}
@@
 \begin{lstlisting}
-val iter_mut : &!('a Array.t) -> (&!(aff_r1,'a) -{un}> unit) -{aff}> unit
+val iter_mut : (&!(aff_r1,'a) -{un}> unit) -> &!('a Array.t) -> unit
 \end{lstlisting}

### 2020-05-26 kindly/soa.tex
related work rustbelt

@@
 Rust~\citep{rust} is the first
-mainstream language to popularize the idea of borrowing and ownership
-for safe low-level programming.
+mainstream language that builds on the concepts of borrowing and ownership
+to enable safe low-level programming.
 \lang is inspired by Rust's borrowing system and transfers some of its
@@
 an ML-like module system.
-Rust's lifetime system is more explicit and more expressive than \lang,
-but Rust does not provide type inference
-and only provides \emph{partial} lifetime inference.
+Rust's lifetime system is more explicit and more expressive than \lang.
+While Rust provides partial lifetime inference, it
+does not support full type inference. 
 Recently, \citet{DBLP:journals/corr/abs-1903-00982}
 formalized Rust's ownership discipline, including non-lexical lifetimes.
-Another important difference is that Rust allows
-to precisely specify the memory layout of objects
-and to pass arguments by value.
+Moreover, Rust programmers have full control over memory allocation
+and memory layout of objects; they can pass arguments by value or by reference.
 These features are crucial for the efficiency goals of Rust.
-In \lang, we assume that all arguments are passed by reference and
-that a garbage collector is available. This forgoes numerous
+In contrast, \lang is garbage collected, assumes a uniform object
+representation, and all arguments are passed by reference. This choice
+forgoes numerous  
 issues regarding interior mutability and algebraic data types.
@@
 whether they are linear or unrestricted.
+
+In Rust, programmers can implement their low-level abstractions by
+using unsafe code fragments. Unsafe code is not typechecked with the
+full force of the Rust type system, but with a watered down version
+that ignores ownership and lifetimes. This loophole is needed to implement
+datastructures like doubly-linked lists or advanced concurrency
+abstractions. When unsafe code occurs as part of a 
+function body, the Rust typechecker leaves the adherence of the unsafe
+code to the function's type signature as a proof obligation to
+the programmer. The RustBelt project
+\cite{DBLP:journals/pacmpl/0002JKD18} provides a formal 
+foundation for creating such proofs by exhibiting a framework for
+semantic soundness of the Rust type system that incorporates aspects
+of concurrecy (i.e., data-race freedom). Similar proof obligations
+would be needed in \lang to check that an implementation of the module
+types or the type of fold shown in \cref{motivation} matches the
+semantics of the typings. We aim to develop a suitable framework for
+this task for \lang. At present, the metatheory of \lang does not
+cover concurrency.
 

### 2020-05-27 kindly/motivation-persistent.tex
improved 2.2 as requested by rev#A

@@
 \subsection{Solving sudokus with hybrid data-structures}
+
+This section presents an implementation of a
+backtracking Sudoku solver using a safe API for persistent arrays that
+supports both mutable updates and immutable versioning.
+The implementation showcases safe mixing of resource allocation and
+deallocation in the presence of exclusive (mutable) and immutable
+borrows. It also demonstrates two new aspects: the interaction between
+closures and borrows and the notion of reborrowing.
+%% A function that closes over a borrow must not escape the borrow's region.
 
@@
 
-In this section, we present
-a safe API for persistent arrays that support both immutable and mutable features,
-and use it to implement a backtracking Sudoku solver.
+Our implementation of a backtracking Sudoku solver abstracts this scenario.
 % show how \lang allows us to define
-The solver maintains an array to represent the state of
+The solver maintains a two-dimensional array to represent the state of
 the game and uses backtracking when there are several choices to proceed.
-The use of backtracking suggests a persistent data structure for the array.
-However, only changes that correspond to a choice point need to use
-the persistence mechanism, others may be implemented as
+As choice points may be revisited several times, it seems advantageous
+to select a persistent data structure for the array.
+However, local changes between choice points may be implemented as
 cheap in-place mutations.
 
-\cref{sig:hybarray} contains an API @HYBARRAY@
+\cref{sig:hybarray} contains an API @HYBARRAY@ along with an
+implementation @CowArray@
 that enables using mutable and immutable modifications to
@@
 The in-place mutation function @set_mut@ with type
-@&!('a t) * int * 'a -> unit@ works on a exclusive borrow  whereas the persistent
+@&!('a t) * int * 'a -> unit@ works on an exclusive borrow  whereas the persistent
 @set@ operation has type @&('a t) * int * 'a -> 'a t@. It
@@
 argument array, but returns a fresh,  modified structure.
-The code in \cref{ex:cow} contains a very simple implementation of
+The module @CowArray@, also in \cref{ex:cow}, contains a very simple
+implementation of 
 @HYBARRAY@ that represents hybrid arrays
@@
 is a simple variation on @Array.map@ where the mapping function also
-takes the position of the element.
+takes the position of the element. Recall that @Array.map@ always
+creates a new array for the result.
 % A more elaborate implementation could use persistent arrays
 % \cite{DBLP:conf/ml/ConchonF07}.
+
+In the example code, we make use of a two-dimensional version
+@Matrix@ of the @CowArray@ data\-structure. The only difference is
+that the API functions @get@, @set@, and @set_mut@ now take two index
+parameters of type @int@ instead of one. The internal working is exactly the same.
 
@@
   \begin{subfigure}[t]{0.45\linewidth}
-\begin{lstlisting}
+\begin{lstlisting}[numbers=none]
 module type HYBARRAY = sig
@@
   \begin{subfigure}[t]{0.55\linewidth}
-\begin{lstlisting}
+\begin{lstlisting}[numbers=none]
 module CowArray : HYBARRAY = struct
@@
 
+\begin{lstlisting}[numbers=none]
+val propagate : int -> int -> &!Matrix.t -> int -> unit
+val solve : int -> int -> Matrix.t -> unit
+\end{lstlisting}
+
   \begin{subfigure}[t]{0.45\linewidth}
@@
 
-Our implementation of a Sudoku solver (\cref{ex:sudoku}) performs
-modifications that correspond to choice points using @set@,
-which makes it trivial to come back to the previous version
-of the array, while other modifications use @set_mut@, which
-cannot be reverted.
+% Our implementation of a Sudoku solver (\cref{ex:sudoku}) performs
+% modifications that correspond to choice points using @set@,
+% which makes it trivial to come back to the previous version
+% of the array, while other modifications use @set_mut@, which
+% cannot be reverted.
 
-The board is represented as a 2D-matrix (\cref{line:boardtype}), where
-the @Matrix@ type uses the same API as @CowArray@
-but with two indices.
-Each cell contains an integer set that represents admissible solutions so far.
-The Sudoku solver iterates over the cells and tries each possible solution (\cref{line:try_solution}).
-When a value is picked for the current cell, we create a choice point,
-change the cell with an immutable modification (\cref{line:immutset}), and propagate
+Our implementation of a Sudoku solver (~\cref{ex:sudoku}) represents
+the board as a 2D-matrix (\cref{line:boardtype}).
+Each cell contains an integer set of type @IntSet.t@ that represents
+admissible solutions so far. This type is immutable, i.e.,
+@IntSet.remove@ produces a new value.
+
+The main functions are @solve@ and @propagate@ with the typings shown
+on top of \cref{ex:sudoku}. The types of @propagate_line@ etc are the
+same as for @propagate@. From the types, we can see that @solve@ takes
+ownership of the board, whereas @propagate@ only takes a mutable
+borrow. Hence, the @propagate@ functions can only modify or read the board,
+whereas @solve@ has full control.
+
+The Sudoku solver @solve@ iterates over the cells and tries each possible
+solution (\cref{line:try_solution}). 
+When a value is picked for the current cell, it creates a choice point
+in @new_board@, where the current cell is updated with an immutable modification (\cref{line:immutset}), and propagate
 the changes with the @propagate@ function.
@@
 
+As the @board@ parameter to the @propagate@ function is an exclusive
+borrow, it should be handled 
+in an affine manner. To pass it safely to the three helper functions,
+the body of @propagate@ \emph{reborrows} (i.e., it takes a borrow of a
+borrow) the board three times in 
+\cref{line:reborrow1}-\cref{line:reborrow3}.
+The function @propagate_line@ also contains two reborrows of the
+exclusive borrow argument @board@, an
+immutable one (\cref{line:reborrow_i}) and an exclusive one
+(\cref{line:reborrow_m}). It demonstrates the facility 
+to take immutable borrows from exclusive ones.
+
 The typing ensures that the mutations do not compromise the state at
-the choice point, because they operate on a new state @new_g@ created for one
+the choice point, because they operate on a new state @new_board@ created for one
 particular branch of the choice.
@@
 the closure @try_solution@ remains unrestricted even though
-it captures the borrow @&g@.
+it captures the borrow @&board@.
 The price is that @try_solution@ cannot escape from
-@&g@'s region. In this example, the inferred region corresponds to the
+@&board@'s region. In this example, the inferred region corresponds to the
 @begin@/@end@ scope. 
@@
 % instead of a sequential one.
-As @g@ is linear we must free it outside of the region
+As @board@ is linear we must free it outside of the region
 before returning (\cref{line:free:g}).

### 2020-05-28 kindly/intro.tex
removed unnecessary line numbers

@@
 would be an error to use the affine closure twice as in
-\begin{lstlisting}
+\begin{lstlisting}[numbers=none]
 let w = File.write &!h in w "Hello "; w "world!" (*type error*)
@@
 For application, \lang{} infers the following type.
-\begin{lstlisting}
+\begin{lstlisting}[numbers=none]
 let app f x = f x
@@
 For @compose@, \lang{} infers this type.
-\begin{lstlisting}
+\begin{lstlisting}[numbers=none]
 let compose f g x = f (g x)

### 2020-05-28 kindly/motivation.tex
removed unnecessary line numbers

@@
   \centering
-\begin{lstlisting}
+\begin{lstlisting}[numbers=none]
 module Array : sig
@@
 Here is a first attempt at the type of a fold function:
-\begin{lstlisting}
+\begin{lstlisting}[numbers=none]
 val fold : ('a : 'k) => ('a -> 'b -{'k}> 'b) -> 'a Array.t -> 'b -{lin}> 'b
@@
 We obtain the following signature for @bfold@, the borrowing fold operation:
-\begin{lstlisting}
+\begin{lstlisting}[numbers=none]
 val bfold : ('b:'k),('k <= aff_r) => (&(aff_r1,'a) -> 'b -{aff_r1}> 'b) -> &('k_1,'a Array.t) -> 'b -{'k_1}> 'b
@@
 @all_files : File.t Array.t@ to compute the sum of their sizes:
-\begin{lstlisting}
+\begin{lstlisting}[numbers=none]
 let total_size_of files = bfold (fun f s -> File.size f + s) files 0
@@
 on:
-\begin{lstlisting}
+\begin{lstlisting}[numbers=none]
 val iter_mut : (&!(aff_r1,'a) -{un}> unit) -> &!('a Array.t) -> unit

### 2020-05-28 kindly/motivation.tex
low level stuff added

@@
 \begin{lstlisting}[numbers=none]
-val bfold : ('b:'k),('k <= aff_r) => (&(aff_r1,'a) -> 'b -{aff_r1}> 'b) -> &('k_1,'a Array.t) -> 'b -{'k_1}> 'b
+val bfold : ('b:'k),('k <= lin_r) => (&(aff_r1,'a) -> 'b -{aff_r1}> 'b) -> &('k_1,'a Array.t) -> 'b -{'k_1}> 'b
 \end{lstlisting}
@@
 shared borrow of the resource with an affine kind at region nesting $r+1$. The return
-type of the callback is constrained to kind @'k <= aff_r@
-so that the callback certainly cannot return the borrowed argument.
+type of the callback is constrained to kind @'k <= lin_r@. The
+important part of this constraint is the $_r$ index, which ensures
+that the callback cannot return the borrowed argument from the more
+deeply nested scope. 
 %
@@
 
+To close this discussion, let's see which primitives are needed to implement
+functions like @bfold@ and @iter_mut@. The obvious sequential implementation of
+both functions boils down to a loop over the index range of the array:
+\begin{lstlisting}
+let rec bfold_helper f a z i =
+  if i >= 0
+  then bfold_helper f &a (f (get_sb (&a, i))) (i - 1)
+  else z
+let bfold f a z =
+  bfold_helper f a z (length &a - 1)
+\end{lstlisting}
+\begin{lstlisting}
+let rec iter_helper f a i =
+  if i >= 0
+  then (f (get_eb (&!a, i)); iter_helper f &!a (i - 1))
+  else ()
+let iter_mut f a =
+  iter_mut f a (length &a - 1)
+\end{lstlisting}
+Observe that we are forced to use different primitives
+\begin{itemize}
+\item @get_sb : &('r, 'a Array.t) -> int -> &('r, 'a)@ to get a
+  shared borrow from a shared borrow of an array and
+\item @get_eb : &!('r, 'a Array.t) -> int -> &('r, 'a)@ to get an
+  exclusive borrow from an exclusive borrow of  an array.
+\end{itemize}
+These primitives have the same underlying implementing (the same as
+@get@ from \cref{sec:imper-progr}). Their types arise from the
+intuition that the borrow of a structure should entitle to borrows of
+its substructures, roughly, the borrow of an array could be considered as
+array of borrows.
+
+%%% PJT: TBC
+
 \lstDeleteShortInline@

### 2020-05-29 kindly/motivation.tex
more discussion of getters in 2.3

@@
 \begin{lstlisting}[numbers=none]
-val fold : ('a : 'k) => ('a -> 'b -{'k}> 'b) -> 'a Array.t -> 'b -{lin}> 'b
+val fold : ('a:'k) => ('a -> 'b -{'k}> 'b) -> 'a Array.t -> 'b -{lin}> 'b
 \end{lstlisting}
@@
 \begin{lstlisting}[numbers=none]
-val iter_mut : (&!(aff_r1,'a) -{un}> unit) -> &!('a Array.t) -> unit
+val iter_mut : (&!(aff_r1,'a) -{un}> unit) -> &!('k_1,'a Array.t) -> unit
 \end{lstlisting}
@@
 To close this discussion, let's see which primitives are needed to implement
-functions like @bfold@ and @iter_mut@. The obvious sequential implementation of
-both functions boils down to a loop over the index range of the array:
+functions like @bfold@ and @iter_mut@. The naive sequential implementations of
+both functions boil down to a loop over the index range of the array:
+\begin{flushleft}\vspace{-\baselineskip}
+  \begin{minipage}[t]{0.49\linewidth}
 \begin{lstlisting}
 let rec bfold_helper f a z i =
-  if i >= 0
-  then bfold_helper f &a (f (get_sb (&a, i))) (i - 1)
-  else z
+  if i < 0 then z
+  else bfold_helper f &a (f (get_sb (&a, i))) (i-1)
 let bfold f a z =
@@
 \end{lstlisting}
+  \end{minipage}
+  \begin{minipage}[t]{0.49\linewidth}
 \begin{lstlisting}
 let rec iter_helper f a i =
-  if i >= 0
-  then (f (get_eb (&!a, i)); iter_helper f &!a (i - 1))
-  else ()
+  if i < 0 then ()
+  else (f (get_eb (&!a, i)); iter_helper f &!a (i-1))
 let iter_mut f a =
@@
 \end{lstlisting}
-Observe that we are forced to use different primitives
+  \end{minipage}
+\end{flushleft}
+Observe that we are proposing two different primitives
 \begin{itemize}
-\item @get_sb : &('r, 'a Array.t) -> int -> &('r, 'a)@ to get a
+\item @get_sb : &('k, 'a Array.t) -> int -> &('k, 'a)@ \\to get a
   shared borrow from a shared borrow of an array and
-\item @get_eb : &!('r, 'a Array.t) -> int -> &('r, 'a)@ to get an
+\item @get_eb :  &!('k, 'a Array.t) -> int -> &!('k, 'a)@ \hfill (unsafe!) \\to get an
   exclusive borrow from an exclusive borrow of  an array.
 \end{itemize}
-These primitives have the same underlying implementing (the same as
+These primitives have the same underlying implemention (the same as
 @get@ from \cref{sec:imper-progr}). Their types arise from the
@@
 array of borrows.
+However, only @get_sb@ is safe: the shared borrow of the array entitles us to
+obtain shared borrows for the elements \emph{in the same region} as
+the shared borrow for the array freezes the array inside the
+region. This freeze extends to the elements because 
+lifetime of the array fully overlaps with the lifetime of its
+elements. 
+Considering @get_eb@, we see that we may obtain \emph{two different
+exclusive borrows} of the same array element inside a region. Clearly,
+the exclusive borrow for the element should live in a nested region
+where the array is not accessible. Hence, the safe alternative is
+to use a different function for obtaining borrows of elements
+\begin{lstlisting}[numbers=none]
+with_eborrow : ('b:'k),('k <= lin_r) => &!('k_1,'a Array.t) -> int -> (&!(aff_r1,'a) -> 'b) -{'k_1}> 'b
+\end{lstlisting}
+Hence, the helper function for @iter_mut@ should read like this
+\begin{lstlisting}
+let rec iter_helper f a i =
+  if i < 0 then ()
+  else (with_eborrow &!a i f; iter_helper f &!a (i-1))
+\end{lstlisting}
+which also explains the occurrence of @aff_r1@ in the type of
+@iter_mut@.
 
-%%% PJT: TBC
+We conclude that borrows of datastructures create the need for
+differently typed access functions that are tailored for different use
+cases. The argumentation whether such an access function is safe is
+sometimes subtle and gives rise to nonobvious types.
 

### 2020-05-29 kindly/soa.tex
preliminary comparison matrix (w/o callout)

@@
 % \cite{DBLP:conf/popl/BoylandR05}
+
+\begin{figure}[tp]
+  \begin{center}
+    \begin{tabular}{l|c|c|c}
+      Language& URAL & (Re)borrowing & Impure
+      \\\hline
+      System F\degree~\citep{DBLP:conf/tldi/MazurakZZ10} & UL & --- & N
+      \\
+      Alms~\citep{DBLP:conf/popl/TovP11} & UA & --- & Y \\
+      Quill~\citep{DBLP:conf/icfp/Morris16} & UL & --- & N \\
+      Linear Haskell~\citep{DBLP:journals/pacmpl/BernardyBNJS18} & UL
+                     & --- & N \\
+      Rust~\citep{rust} & UAL & (Y)Y & Y
+      \\\hline
+      \lang & UAL & (Y)Y & Y
+      \\
+    \end{tabular}
+  \end{center}
+  \caption{Comparison matrix}
+  \label{fig:comparison-matrix}
+\end{figure}
 

### 2020-05-29 kindly/motivation.tex
remarks by rev#C

@@
 initialize all array elements.
-To @free@ an array the elements must be affine. 
+To @free@ an array the elements must be affine. Thanks to subkinding,
+the type of @free@ is pleasingly simple: any type @'a@ whose kind is
+less than or equal to @aff@ is acceptable. 
 The @length@ function is always applicable.
@@
 borrow. \cref{ex:array:get} contains two shared borrows 
-in the same expression which forms a region by itself. These borrows
+in the same expression which forms a region by itself (recall that
+shared borrows are unrestricted and may thus be duplicated). These borrows
 are split off the exclusive borrow used in \cref{ex:array:set} which

### 2020-05-30 kindly/soa.tex
Expand the comparison table.

--- 2020-05-26 kindly/soa.tex
@@
 % \cite{DBLP:conf/popl/BoylandR05}
+
+\begin{figure}[tp]
+  \begin{center}
+    \newcommand\YC{\color{Green}}
+    \newcommand\NC{\color{Red}}
+    \newcommand\MC{\color{Orange}}
+    \newcommand\Y{{\YC{\ding{52}}}\xspace}
+    \newcommand\N{{\NC{\ding{56}}}\xspace}
+    \newcommand\M{\MC{\textasciitilde}} % "Meh"
+
+    % Little bit of magic to make angled headers
+    \newcolumntype{R}[2]{%
+      >{\adjustbox{angle=#1,lap=\width-(#2)}\bgroup}%
+      c%
+      <{\egroup}%
+    }
+    \newcommand*\rot{\multicolumn{1}{R{45}{1em}}}
+
+    \begin{tabular}{l|cccccccc}
+      Language
+      & URAL & \rot{Borrows} & \rot{State}
+      & \rot{R. Subsumption}
+      & \rot{R. Polymorphism}
+      & \rot{Inference} & \rot{Formalisation}
+      & \rot{Basis}
+      \\\hline
+      System F\degree~\citep{DBLP:conf/tldi/MazurakZZ10}
+      & UL & --- & \N & \N & \N & \N & \YC Coq & System F
+      \\
+      Alms~\citep{DBLP:conf/popl/TovP11}
+      & UA & --- & \Y & \M & \Y & \NC Local & \N & ML \\
+      Quill~\citep{DBLP:conf/icfp/Morris16}
+      & UL & --- & \N & \N & \Y & \YC Principal & \MC Manual & Qualified types\\
+      Linear Haskell~\citep{DBLP:journals/pacmpl/BernardyBNJS18}
+      & UL & --- & \M & \N & \Y & \MC Non-principal & \MC Manual & Haskell \\
+      Rust~\citep{rust}
+      & UA & \Y & \Y & \Y & \M & \NC Local & \YC Coq & ---
+      \\\hline
+      \lang
+      & UAL & \Y & \Y & \Y & \Y & \YC Principal & \MC Manual & ML/HM(X)
+      \\
+    \end{tabular}
+  \end{center}
+  \caption{Comparison matrix}
+  \label{fig:comparison-matrix}
+\end{figure}
 

### 2020-05-30 kindly/soa.tex
Extend the comparison table a bit more.

@@
     \begin{tabular}{l|cccccccc}
-      Language
+      \textbf{Language}
       & URAL & \rot{Borrows} & \rot{State}
@@
       Alms~\citep{DBLP:conf/popl/TovP11}
-      & UA & --- & \Y & \M & \Y & \NC Local & \N & ML \\
+      & UA & --- & \Y & \M & \Y & \MC Local & \N & ML \\
       Quill~\citep{DBLP:conf/icfp/Morris16}
@@
       & UL & --- & \M & \N & \Y & \MC Non-principal & \MC Manual & Haskell \\
+      Mezzo~\citep{DBLP:phd/hal/Protzenko14}
+      & $\star$ & --- & \Y & \M & \Y & \N & \MC Manual & --- \\
+      \hline
       Rust~\citep{rust}
-      & UA & \Y & \Y & \Y & \M & \NC Local & \YC Coq & ---
-      \\\hline
+      & UA & \Y & \Y & \Y & \M & \MC Local & \YC Coq & --- \\
+      Vault~\citep{DBLP:conf/pldi/DeLineF01}/
+      Plaid~\citep{DBLP:conf/oopsla/AldrichSSS09}
+      & UA & --- & \Y & \Y & \Y & \N & \MC Manual & --- \\
+      \hline
       \lang

### 2020-05-30 kindly/extensions.tex
NLL Extension: fixed typo.
Start presenting limitations (and potential extensions)

@@
 \lstMakeShortInline[keepspaces,basicstyle=\small\ttfamily]@
-\section{Extensions}
+\section{Limitations and Extensions}
 
-\subsection{Algebraic Datatypes}
+\subsection{Flow sensitivity}
 
-Algebraic data types are a staple of functional programming and fit nicely
-in our paradigm. Indeed, it is sufficient to ensure  that the kinds of
-the constructor arguments are less than or equal to the kind of the datatype.
-Hence, it is forbidden to include affine elements in an unrestricted
-datatype, whereas the elements in a linear list may be linear or
-unrestricted. 
-Here is the definition of a datatype for linear lists.
+The type system defined so far does not support any
+form of flow sensitivity. This means that code patterns that relies on
+subtle flow-sensitive usage of permissions and linearity likely will not
+typecheck in \lang. As an example the following piece of code,
+which defines the merge function on linear lists,
+cannot be expressed directly.
 
 \begin{lstlisting}
-type ('a : 'k) llist : lin = Nil | Cons of ('a * 'a llist)
+let rec merge l1 l2 = match l1, l2 with
+  | h1::t1, h2::t2 ->
+    if &h1 < &h2 
+    then h1::(merge t1 l2) (* Must expand l2 to h2::t2 here *)
+    else h2::(merge l1 t2)
+  | ....
 \end{lstlisting}
 
-This extension is implemented in our prototype.
+Such pattern usually requires richer logic, such as
+Mezzo~\citep{DBLP:phd/hal/Protzenko14}.
+However
+\citet{DBLP:journals/corr/abs-1903-00982} formalize Rust's
+notion of non-lexical lifetimes which allow to partially support such
+code patterns. We believe this notion can be adapted to \lang's notion of
+regions
 
-\subsection{Ad-hoc Polymorphism and Borrows}
-
-In our formalization, we use two operators, $\borrow{x}$ and $\reborrow{x}$ to
-distinguish between borrows and borrows of borrows.
-Such a distinction is inconvenient for programming.
-Using a typeclass-like mechanism, we can replace these operators
-by a single overloaded operator, $\borrow{x}$, which expects $x$ to be @Borrowable@ and
-would then desugar to the more precise operators.
-A similar solution is used in Rust through the @Borrow@ and @Defer@
-traits.
-This approach also enables method calls on objects without
-explicit borrows, such as @foo.len()@ where len expects a shared borrow.
-
-Ad-hoc polymorphism fits demonstrably in the \hmx framework of constrained
-types and preserves all properties of our language such
-as principal type inference. Its soundness is orthogonal to linear types
-and has been explored in the literature~\citep{DBLP:conf/fpca/OderskyWW95}.
-
-\subsection{Branching constructs}
-
-Our formalization of \lang does not contain branching constructs. As a consequence,
-subexpressions are always executed sequentially. This is of great importance
-for the splitting relation introduced in \cref{sdtyping}, as it must
-coincide with the order of evaluation. To account for borrows and linear
-values in different branches, we need a new (symmetric) join relation.
-This relation immediately follows from the semantics of borrows and linearity
-shown so far, and allow to easily define branching constructs such as if-then-else
-and pattern matching.
-This extension is implemented in our prototype.
-
-\subsection{Non-Lexical Regions}
+\subsubsection{Non-Lexical Regions}
 
@@
 
+\subsection{Capabilities and Identity}
+
+One primary feature of \lang is that its tracking of linearity does
+not use any notion of ``identity'': the type system can't specify that two
+objects are the same, simply that they share the same usage pattern with
+regards to linearity.
+Languages such as Alms~\citep{DBLP:conf/popl/TovP11}, on the other hand,
+often relies on a notion of identity to express capabilities.
+For instance, the Alms function
+@Array.create@ of type @int -> 'a -> \E 'b. ('a, 'b) array@ uses
+@'b@ to uniquely identifies the array.
+Functions such as @Array.acquire : ('a, 'b) array -> 'b cap@
+are then used to manipulate capabilities pertaining to the array.
+
+While such uses are partially covered by borrows and regions,
+a notion of identity associated to regions
+would allow expressing regions directly in type signatures.
+For instance, the @get_eb@ function shown in
+\cref{tuto:vector} could be made safe
+by creating a restricted inner region on function application,
+with the signature:
+@&!('k, 'a Array.t) -> int -> \E ('k' < 'k) &!('k', 'a)@
+
+This notion of identity often relies on existential types.
+\lang does not support existentials so far as it would
+forgo principal type inference.
+However existentials
+are compatible with the HM(X) framework~\citep{DBLP:conf/icfp/Simonet03}
+and would make a very desirable addition to \lang. OCaml also show
+that existential types can be made usable without compromising
+inference in the rest of the language, by integrating unpacking
+and pattern matching.
+
+\subsection{Ad-hoc Polymorphism and Borrows}
+
+In our formalization, we use two operators, $\borrow{x}$ and $\reborrow{x}$ to
+distinguish between borrows and borrows of borrows.
+Such a distinction is inconvenient for programming.
+Using a typeclass-like mechanism, we can replace these operators
+by a single overloaded operator, $\borrow{x}$, which expects $x$ to be @Borrowable@ and
+would then desugar to the more precise operators.
+A similar solution is used in Rust through the @Borrow@ and @Defer@
+traits.
+This approach also enables method calls on objects without
+explicit borrows, such as @foo.len()@ where len expects a shared borrow.
+
+Ad-hoc polymorphism fits demonstrably in the \hmx framework of constrained
+types and preserves all properties of our language such
+as principal type inference. Its soundness is orthogonal to linear types
+and has been explored in the literature~\citep{DBLP:conf/fpca/OderskyWW95}.
+
+\subsection{Traditional functional features}
+
+\subsubsection{Algebraic Datatypes}
+
+Algebraic data types are a staple of functional programming and fit nicely
+in our paradigm. Indeed, it is sufficient to ensure  that the kinds of
+the constructor arguments are less than or equal to the kind of the datatype.
+Hence, it is forbidden to include affine elements in an unrestricted
+datatype, whereas the elements in a linear list may be linear or
+unrestricted. 
+Here is the definition of a datatype for linear lists.
+
+\begin{lstlisting}
+type ('a : 'k) llist : lin = Nil | Cons of ('a * 'a llist)
+\end{lstlisting}
+
+This extension is implemented in our prototype.
+
+\subsubsection{Branching constructs}
+
+Our formalization of \lang does not contain branching constructs. As a consequence,
+subexpressions are always executed sequentially. This is of great importance
+for the splitting relation introduced in \cref{sdtyping}, as it must
+coincide with the order of evaluation. To account for borrows and linear
+values in different branches, we need a new (symmetric) join relation.
+This relation immediately follows from the semantics of borrows and linearity
+shown so far, and allow to easily define branching constructs such as if-then-else
+and pattern matching.
+This extension is implemented in our prototype.
+
 \lstDeleteShortInline@

### 2020-05-30 kindly/motivation.tex
Tiny presentation fix.
Small code fixes for the new iterators.

@@
 
-This approach is not sufficient if we want to mutate the array during iteration.
+This approach is not sufficient if we want to mutate the elements of
+the array during iteration.
 To this end, we need to take an exclusive borrow of the structure to iterate
@@
 \begin{lstlisting}[numbers=none]
-val iter_mut : (&!(aff_r1,'a) -{un}> unit) -> &!('k_1,'a Array.t) -> unit
+val iter_mut : (&!(aff_r1,'a) -> unit) -> &!('k_1,'a Array.t) -> unit
 \end{lstlisting}
-
-The callback is explicitly annotated with @un@ because  it will be called multiple times.
 
@@
   if i < 0 then z
-  else bfold_helper f &a (f (get_sb (&a, i))) (i-1)
+  else bfold_helper f &&a (f (get_sb &&a i) z) (i-1)
 let bfold f a z =
-  bfold_helper f a z (length &a - 1)
+  let l = length &&a - 1 in
+  bfold_helper f &&a z l
 \end{lstlisting}
@@
   if i < 0 then ()
-  else (f (get_eb (&!a, i)); iter_helper f &!a (i-1))
+  else (f (get_eb &&!a i); iter_helper f &&!a (i-1))
 let iter_mut f a =
-  iter_mut f a (length &a - 1)
+  let l = length &&a - 1 in
+  iter_helper f &&!a l
 \end{lstlisting}
@@
 \begin{itemize}
-\item @get_sb : &('k, 'a Array.t) -> int -> &('k, 'a)@ \\to get a
+\item
+  @get_sb : &('k, 'a Array.t) -> int -{'k}> &('k, 'a)@ \\to get a
   shared borrow from a shared borrow of an array and
-\item @get_eb :  &!('k, 'a Array.t) -> int -> &!('k, 'a)@ \hfill (unsafe!) \\to get an
+\item
+  @get_eb : &!('k, 'a Array.t) -> int -{'k}> &!('k, 'a) (* Unsafe! *)@
+\\to get an
   exclusive borrow from an exclusive borrow of  an array.

### 2020-05-30 kindly/soa.tex
callout of the comparison matrix

--- 2020-05-30 kindly/soa.tex
@@
       Alms~\citep{DBLP:conf/popl/TovP11}
-      & UA & --- & \Y & \M & \Y & \NC Local & \N & ML \\
+      & UA & \N & \Y & \M & \Y & \NC Local & \N & ML \\
       Quill~\citep{DBLP:conf/icfp/Morris16}
@@
 None of the following languages support borrows.
+
+The comparison matrix in \cref{fig:comparison-matrix} gives an
+overview over the systems discussed in this section. The column URAL
+specifies the substructural features. Borrows and state are
+self-explanatory. Subsumption indicates that kind U is promoted
+to A and further to L. Polymorphism refers to polymorphism over
+substructural kinds. Inference is self-explanatory, formalisation
+refers to the formal semantics and type soundness proof. Basis is self-explanatory.
 

### 2020-06-02 kindly/extensions.tex
Add a small note about a richer region system.

@@
 
+\subsection{A richer region system}
+
+\lang requires that each region is identified by an index drawn from a partial
+order that is compatible with the nesting of regions.
+This order can be implemented in many ways, including region variables
+as often used in algebraic effects systems, existentials, etc.
+
+For simplicity, the formalization uses the concrete implementation with 
+natural numbers for indices. The proofs only rely on
+the existence of a partial order and could be adapted to one of the
+more abstract approaches.
+In particular, \lang could reuse regions variables provided by
+the ongoing work on effect systems
+for OCaml~\cite{DBLP:conf/sfp/DolanEHMSW17}.
+
 \subsection{Traditional functional features}

### 2020-06-02 kindly/infer/infer.tex
Present the properties required on the constraint system.

@@
 
-\subsection{Constraint Solving}
+\subsection{Constraints}
 \label{infer:solving}
@@
 
-For concision, we demonstrate the constraint solving
-algorithm with an example. The complete constraint system is
+To properly define our type system, we need to define $\CL$, a constraint
+system equipped with an entailment relation noted $\operatorname{\vdash_e}$
+and a normalizing function.
+For concision, we first demonstrate the constraint solving
+algorithm with an example. We then state the various properties
+that make it suitable for use in the \hmx framework.
+The complete constraint system is
 defined in \cref{appendix:constraints}.
-%
+
+\subsubsection{Constraints normalization by example}
+
 \input{infer/exampleconstr}
 
-Our algorithm respect the properties
-of \hmx, computes principal normal forms
-and simplifies constraints significantly.
+\subsubsection{Properties of the constraint system}
+
+To apply \hmx to $\CL$, normalize must compute principal normal forms
+and $\CL$ must be regular.
+
+\begin{property}[Principal normal form]
+  Normalization computes principal normal forms for $\CL$, i.e.
+  given a constraint $D\in\CL$, a substitution $\phi$ and
+  $(C,\unif) = \normalize{D}{\phi}$,
+  then $\phi\leq\unif$,
+  $C \equivC \unif D$ and
+  $\unif C = C$.
+\end{property}
+
+\begin{property}[Regular constraint system]
+  $\CL$ is regular, ie, for $x, x'$ two types or kinds,
+  $\entail{}{\Ceq{x}{x'}}$ implies
+  $\fv{x} = \fv{x'}$
+\end{property}
+
+These properties
+are sufficient to state that $HM(\CL)$ provides principal type inference.
+The next section states that this properties carries over to our extension
+of \hmx
+with kind inference, affine types and borrows.
+%
+This algorithm also simplifies constraints significantly.
 All the simplification mechanisms presented
@@
 \citep{DBLP:conf/aplas/Simonet03}.
-
-%
-% We now show that this algorithm is well-behaved with respect to entailment
-% and computes unique normal forms.
-
-% \begin{property}[Principal normal form]
-%   Normalization computes principal normal forms for $\CL$, i.e.
-%   given a constraint $D\in\CL$, a substitution $\phi$ and
-%   $(C,\unif) = \normalize{D}{\phi}$,
-%   then $\phi\leq\unif$,
-%   $C \equivC \unif D$ and
-%   $\unif C = C$.
-% \end{property}
-
-% \begin{property}[Principal constraints]
-%   $\CL$ has the principal constraint property, i.e.
-%   for every $D\in\CL$, and a substitution $\unif$,
-%   either $D$ does not have a normal form, or it has
-%   a principal normal form.
-% \end{property}
-
-% \begin{property}[Regular constraint system]
-%   $\CL$ is regular, ie, for $x, x'$ two types or kinds,
-%   $\entail{}{\Ceq{x}{x'}}$ implies
-%   $\fv{x} = \fv{x'}$
-% \end{property}
-
-% The proofs are given in \cref{appendix:constraints}. These three properties
-% are sufficient to state that $HM(\CL)$ provides principal type inference.
-% However, we use an extension of \hmx with kind inference
-% and affine types and borrows.
-% We now show that \hmx's completeness and soundness theorems are still valid on
-% this extended system.
 

### 2020-06-02 kindly/extensions.tex
cosmetics

--- 2020-03-04 kindly/extensions.tex
@@
 \lstMakeShortInline[keepspaces,basicstyle=\small\ttfamily]@
-\section{Extensions}
+\section{Limitations and Extensions}
 
-\subsection{Algebraic Datatypes}
+\subsection{Flow sensitivity}
 
-Algebraic data types are a staple of functional programming and fit nicely
-in our paradigm. Indeed, it is sufficient to ensure  that the kinds of
-the constructor arguments are less than or equal to the kind of the datatype.
-Hence, it is forbidden to include affine elements in an unrestricted
-datatype, whereas the elements in a linear list may be linear or
-unrestricted. 
-Here is the definition of a datatype for linear lists.
+The type system defined so far does not support any
+form of flow sensitivity. Therefore, code patterns that rely on
+subtle flow-sensitive usage of permissions and linearity will most likely not
+typecheck in \lang. For example, the following merge function on linear lists
+cannot be expressed directly, because matching against
+\lstinline/l1/ and \lstinline/l2/ consumes both lists.
 
 \begin{lstlisting}
-type ('a : 'k) llist : lin = Nil | Cons of ('a * 'a llist)
+let rec merge l1 l2 = match l1, l2 with
+  | h1::t1, h2::t2 ->
+    if &h1 < &h2 
+    then h1::(merge t1 l2) (* Must expand l2 to h2::t2 here *)
+    else h2::(merge l1 t2)
+  | ....
 \end{lstlisting}
 
-This extension is implemented in our prototype.
+Patterns like this require a richer logic, such as provided by
+Mezzo~\citep{DBLP:phd/hal/Protzenko14}.
+However,
+\citet{DBLP:journals/corr/abs-1903-00982} formalize Rust's
+notion of non-lexical lifetimes which partially supports such
+code patterns. We believe this notion can be adapted to \lang's notion of
+regions.
+
+\subsubsection*{Non-Lexical Regions}
+
+The notion of non-lexical lifetimes is 
+a recent addition to Rust.
+With this feature code is acceptable even if borrowing does not respect
+lexical scoping as in this example:
+
+\begin{lstlisting}[numbers=none]
+let a = &x in (f a; g (&!x))
+\end{lstlisting}
+
+This code pattern is dynamically safe because $a$ is not used after
+the function call @f a@.
+Non-lexical lifetimes (NLL) handle such a pattern by removing expressions
+that do not mention $a$ from its region; in this example,  NLL removes
+the last expression.
+%
+In \lang, regions are lexical and marked by the
+expression $\region{b}{e}$.
+During inference, kind constraints prevent escaping from
+a region.
+
+To add support for non-lexical lifetimes, we could replace the
+lexical region by an annotation on each expression indicating which borrows are
+live in this expression.
+When exploring a subexpression, we would compare the annotations, and automatically
+apply the $\textsc{Region}$ rule when they differ.
+%
+This approach is equivalent to inlining the $\textsc{Region}$ rule in all the other
+rules.
+
+Applied to the program above, only the first two expressions would be annotated
+to be ``in the region associated with @&x@'', but not the last expression.
+Thanks to these annotations, type checking the sequence would check
+that the borrow does not escape the left-hand side (i.e., the second
+expression @f a@).
+
+\subsection{Capabilities and Identity}
+
+In \lang the tracking of linearity does
+not rely on any notion of ``identity'': the type system cannot specify that two
+objects are the same, simply that they share the same usage pattern with
+regards to linearity.
+A language like Alms~\citep{DBLP:conf/popl/TovP11}, on the other hand,
+often relies on a notion of identity to express capabilities.
+For instance, the Alms typing
+@Array.create : int -> 'a -> \E 'b. ('a, 'b) array@ uses
+@'b@ as a  unique identification of the array.
+Functions such as @Array.acquire : ('a, 'b) array -> 'b cap@
+are used to obtain capabilities to operate on the array.
+
+While such uses are partially covered by borrows and regions,
+a notion of identity associated to regions
+would enable us to express regions directly in type signatures.
+For instance, the @get_eb@ function shown in
+\cref{tuto:vector} could be made safe
+by creating a restricted inner region on function application,
+with the signature:
+@&!('k, 'a Array.t) -> int -> \E ('k' < 'k) &!('k', 'a)@
+
+This approach relies on existential types to model identities.
+At present, \lang does not support existentials as it would
+forgo principal type inference.
+However, existentials
+are compatible with the HM(X) framework~\citep{DBLP:conf/icfp/Simonet03}
+and would make a very desirable addition to \lang. Work on GADTs in
+OCaml and Haskell demonstrates
+%% PJT: citation?
+that existential types can be put to use without compromising
+inference in the rest of the language, by integrating unpacking
+and pattern matching.
 
@@
 
-\subsection{Branching constructs}
+\subsection{A Richer Region System}
+
+\lang requires that each region is identified by an index drawn from a partial
+order that is compatible with the nesting of regions.
+This order can be implemented in many ways, including region variables
+as often used in algebraic effects systems, existentials, etc.
+
+For simplicity, the formalization uses the concrete implementation with 
+natural numbers for indices. The proofs only rely on
+the existence of a partial order and could be adapted to one of the
+more abstract approaches.
+In particular, \lang could reuse regions variables provided by
+the ongoing work on effect systems
+for OCaml~\cite{DBLP:conf/sfp/DolanEHMSW17}.
+
+\subsection{Standard Features}
+
+\subsubsection{Algebraic Datatypes}
+
+Algebraic data types are a staple of functional programming and fit nicely
+in our paradigm. Indeed, it is sufficient to ensure  that the kinds of
+the constructor arguments are less than or equal to the kind of the datatype.
+Hence, it is forbidden to include affine elements in an unrestricted
+datatype, whereas the elements in a linear list may be linear or
+unrestricted. 
+Here is the definition of a datatype for linear lists.
+
+\begin{lstlisting}
+type ('a : 'k) llist : lin = Nil | Cons of ('a * 'a llist)
+\end{lstlisting}
+
+This extension is implemented in our prototype.
+
+\subsubsection{Branching constructs}
 
 Our formalization of \lang does not contain branching constructs. As a consequence,
-subexpressions are always executed sequentially. This is of great importance
+subexpressions are always executed sequentially. Ordering is import
 for the splitting relation introduced in \cref{sdtyping}, as it must
 coincide with the order of evaluation. To account for borrows and linear
-values in different branches, we need a new (symmetric) join relation.
-This relation immediately follows from the semantics of borrows and linearity
-shown so far, and allow to easily define branching constructs such as if-then-else
-and pattern matching.
+values in different branches, we need a different, straightforward (symmetric) join relation.
+Its definition follows immediately from the semantics of borrows and linearity
+shown so far and it must be used in the typing rules for branching
+constructs such as if-then-else and pattern matching.
 This extension is implemented in our prototype.
-
-\subsection{Non-Lexical Regions}
-
-A recent important addition to Rust is the notion of non-lexical lifetimes.
-With this feature code is acceptable even if borrowing does not respect
-lexical scoping as in this example:
-
-\begin{lstlisting}
-let a = &x in
-f a;
-g (&!x)
-\end{lstlisting}
-
-Such code patterns are dynamically safe since $a$ is not used after
-the second line.
-Non-lexical lifetimes handle this by removing expressions
-that do not mention $a$ from its region; in this example,  the
-expression on the last line.
-%
-In \lang, regions are purely lexical and marked by the
-expression $\region{b}{e}$.
-During inference, we introduce kind constraints to prevent escaping from
-a region.
-
-To add support for non-lexical lifetimes, we could replace the
-lexical region by an annotation on each expression indicating which borrows are
-live in this expression.
-When exploring a subexpression, we would compare the annotations, and automatically
-apply the $\textsc{Region}$ rule when they differ.
-%
-This approach is equivalent to inlining the $\textsc{Region}$ rule in all the other
-rules.
-
-Applied to the program above, only the first two lines would be annotated
-to be ``in the region associated with @&x@'', but not the last line.
-Thanks to these annotations, when type checking the sequence we would check
-that the borrow does not escape the left-hand side (i.e., the second line).
 

### 2020-06-02 kindly/sdtyping/extra.tex

@@
 %
-Following \hmx, we define a cylindric constraint system as
-a system which respects the following properties for any variable $x$:
+
+We directly reuse the following definitions from \hmx.
+
+\begin{property}[Cylindric constraint system]
+  A cylindric constraint system is a constraint system
+  such that, for any constraint $C$:
 \begin{align*}
@@
 \end{align*}
+\end{property}
 
-Furthermore, we define a term rewriting system as a system where, for every
+\begin{property}[Term rewriting system]
+A term rewriting system is a system where, for every
 types $\tau$,$\tau'$, there exists an equality predicates $\Ceq{\tau}{\tau'}$
@@
 \end{align*}
+\end{property}
+
+\begin{definition}[Constraint system with lattice]
 $\CL$ is defined as the smallest cylindric term constraint system that
 satisfies the axiom shown in \cref{rules:entail}.
+\end{definition}
 

### 2020-06-02 kindly/lang.tex
Small fix.
Add some explanation.
Address a few more comments.

@@
   \lam{a}{}\ &
-  \letin{x}{\app{f}{\borrow[\kun]{a}}}{}\\
+  \letin{x}{\app{f}{\borrow[\kaff]{a}}}{}\\
   &g\ (\borrow[\kaff]{x});\\
@@
   \lam{a}{}\region[1]{\Sone a\MBORROW}{\ &%
-  \letin{x}{\app{f}{\borrow[\kun]{a}}}{}\\
+  \letin{x}{\app{f}{\borrow[\kaff]{a}}}{}\\
   &\region[2]{\Sone x\MBORROW}{g\ (\borrow[\kaff]{x})};\\
@@
 error.
+In particular, it is essential to place the second borrow around \emph{both}
+occurrence of $(\borrow[\kun]{x}$: we want the function to receive both
+borrows, but its result must not contain any borrow (otherwise it would escape).
 Region indices are assigned after placement, so it is trivial to
@@
                                 $C$. \\
-    $\entail{D}{C}$ & Constraint $D$ entails constraint $C$.
+    $\entail{D}{C}$ & Constraint $D$ entails constraint $C$.\\
+    $D \equivC C$ & Constraints $C$ and $D$ are equivalent.
   \end{tabular}
@@
       }{
-        \lregion{C}{}
+        \lregion{C}{x}
         {\svar[\BORROW]{x}{\schm}^n}

### 2020-06-02 kindly/soa.tex
Improve the comparison matrix.

--- 2020-05-30 kindly/soa.tex
@@
 
-    \begin{tabular}{l|cccccccc}
-      Language
+    \begin{tabular}{l|ccccccccc}
+      \textbf{Language}
       & URAL & \rot{Borrows} & \rot{State}
@@
       & \rot{R. Polymorphism}
+      & \rot{Identity}
       & \rot{Inference} & \rot{Formalisation}
@@
       System F\degree~\citep{DBLP:conf/tldi/MazurakZZ10}
-      & UL & --- & \N & \N & \N & \N & \YC Coq & System F
+      & UL & --- & \N & \N & \N & \N & \N & \YC Coq & System F
       \\
       Alms~\citep{DBLP:conf/popl/TovP11}
-      & UA & \N & \Y & \M & \Y & \NC Local & \N & ML \\
+      & UA & --- & \Y & \M & \Y & \Y & \MC Local & \N & ML \\
       Quill~\citep{DBLP:conf/icfp/Morris16}
-      & UL & --- & \N & \N & \Y & \YC Principal & \MC Manual & Qualified types\\
-      Linear Haskell~\citep{DBLP:journals/pacmpl/BernardyBNJS18}
-      & UL & --- & \M & \N & \Y & \MC Non-principal & \MC Manual & Haskell \\
+      & UL & --- & \N & \N & \Y & \N & \YC Principal & \MC Manual & Qualified types\\
+      Lin. Haskell~\citep{DBLP:journals/pacmpl/BernardyBNJS18}
+      & UL & --- & \M & \N & \Y & \N & \MC Non-principal & \MC Manual & Haskell \\
+      Mezzo~\citep{DBLP:phd/hal/Protzenko14}
+      & UAL & --- & \Y & \M & \Y & \Y & \N & \MC Manual & --- \\
+      \hline
       Rust~\citep{rust}
-      & UA & \Y & \Y & \Y & \M & \NC Local & \YC Coq & ---
-      \\\hline
+      & UA & \Y & \Y & \Y & \M & \N & \MC Local & \YC Coq & --- \\
+      Plaid~\citep{DBLP:conf/oopsla/AldrichSSS09}
+      & UA & --- & \Y & \Y & \Y & \Y & \N & \MC Manual & --- \\
+      \hline
       \lang
-      & UAL & \Y & \Y & \Y & \Y & \YC Principal & \MC Manual & ML/HM(X)
+      & UAL & \Y & \Y & \Y & \Y & \N & \YC Principal & \MC Manual & ML/HM(X)
       \\
@@
 The comparison matrix in \cref{fig:comparison-matrix} gives an
-overview over the systems discussed in this section. The column URAL
+overview over the systems discussed in this section. The column UAL
 specifies the substructural features. Borrows and state are
-self-explanatory. Subsumption indicates that kind U is promoted
-to A and further to L. Polymorphism refers to polymorphism over
-substructural kinds. Inference is self-explanatory, formalisation
+self-explanatory. ``Subsumption'' indicates that unrestricted elements
+can be promoted to affine and then linear. This applies to resources,
+borrows, or closures. ``Polymorphism'' refers to polymorphism over
+usage (unrestricted, affine, linear), notably for closures.
+``Identity'' indicates that the language supports a notion
+of identity, usually through existential types, as described in
+\cref{identity}.
+Inference is self-explanatory, formalisation
 refers to the formal semantics and type soundness proof. Basis is self-explanatory.

### 2020-06-03 kindly/motivation.tex
preview in first paragraph

@@
 
-\lang{} is amenable to a functional resource-passing style
+\lang{} supports the resource-passing style
 common in functional encodings of session types (e.g.,
@@
 example of linear arrays demonstrates simple borrowing and
-imperative programming; the second example is a Sudoko solver using a
-hybrid copy-on-write data structure; the third example demonstrates advanced uses
-of regions with iterators on linear values.
+imperative programming; the second example demonstrates reborrowing
+and the interaction between closures and borrowing by implementing a
+Sudoko solver based on a hybrid copy-on-write data structure; the
+third example demonstrates advanced uses 
+of regions with iterators on linear values and the low-level
+primitives needed to implement them.
 Further examples are available in \cref{sec:extra-example}.

### 2020-06-03 kindly/infer/infer.tex
cosmetics

@@
 are sufficient to state that $HM(\CL)$ provides principal type inference.
-The next section states that this properties carries over to our extension
+The next section shows that these properties carry over to the
+inference algorithm for our extension
 of \hmx
-with kind inference, affine types and borrows.
+with kind inference, affine types, and borrows.
 %
-This algorithm also simplifies constraints significantly.
-All the simplification mechanisms presented
-here, including the variance-based one, are complete.
-It is also possible to add ``best-effort'' simplification
-rules which help reduce the size of inferred signatures even further
+This algorithm includes sound and complete constraint simplification.
+In addition, we may add ``best-effort'' simplification
+rules which help reduce the size of inferred signatures 
 \citep{DBLP:conf/aplas/Simonet03}.
@@
 
-The extended inference algorithm still satisfies the soundness
-and completeness properties of \hmx.
+The extended inference algorithm is sound
+and complete with respect to our extension of \hmx.
 % This is achieved by extending
@@
 %
-The first theorem states that our inference algorithm is sound
+The first theorem states that inference is sound
 with respect to the syntax-directed type system.
@@
 The syntax-directed derivation holds with the usage environment $\Sv$ instead of the originally provided environment $\E$. Indeed,
-$\E$ does not contain suspended and borrow bindings, since those
-are discovered on the fly and recorded in $\Sv$. The type binding, however,
-are directly taken from the syntax-directed derivation.
+$\E$ does not contain suspended and borrow bindings. Those
+are discovered on the fly and recorded in $\Sv$. Type bindings
+are taken directly from the syntax-directed derivation.
 
-The second theorem states that our algorithm is complete: for any given
+The second theorem states that inference is complete: for any given
 syntax-directed typing derivation, our inference algorithm can find

### 2020-06-03 kindly/extensions.tex
Address a few more comments.
Improve the comparison matrix.
algebraic datatypes

@@
 \begin{lstlisting}[numbers=none]
-let a = &x in (f a; g (&!x))
+let a = &x in (f a; g &!x)
 \end{lstlisting}
@@
 the function call @f a@.
+Here, this can be made explicit by transforming
+the code to @(let a = &x in f a); g &!x@. However, this is not possible
+in programs with branches who uses different dynamic patterns.
 Non-lexical lifetimes (NLL) handle such a pattern by removing expressions
@@
 \subsection{Capabilities and Identity}
+\label{identity}
 
@@
 
-\begin{lstlisting}
+\begin{lstlisting}[numbers=none]
 type ('a : 'k) llist : lin = Nil | Cons of ('a * 'a llist)
@@
 
-This extension is implemented in our prototype.
+Our prototype implements non-recursive algebraic datatypes with
+pattern matching.
 

### 2020-06-03 kindly/lang.tex
cosmetics
Make a pass on \Instance and \bbvar.

@@
       \inferrule[BorrowBinding]{
-        \entail{C}{(\BORROW_n\lk k) \wedge (k \lk \BORROW_\infty)}\\
+        \entail{C}{(\BORROW_n\lk k) \wedge (k \lk \BORROW_\infty)}\\\\
         \BORROW\in\left\{\kun,\kaff\right\}
@@
 the environment $\E$
-and instantiates it using the substitution $\unif$. The rule also
+and instantiates it with $\Instance(\E,\schm)$.
+Instantiation follows the \hmx formulation and
+takes as input a scheme $\schm$ and an environment $\E$
+and returns a constraint $C$ and a type $\tau$.
+The rule also
 checks that the other bindings in $\E$ can be safely discarded by
@@
 and then later as a full resource on the right. The {\sc Susp} rule
-works with a full resource, the {\sc SuspB}
-with a borrow and {\sc SuspS} with a suspended binding.
+works with a full resource, rule {\sc SuspB}
+with a borrow, and rule {\sc SuspS} with a suspended binding.
 

### 2020-06-04 kindly/soa.tex
paragraph on Rust traits Send, Sync, and Copy

@@
 overview over the systems discussed in this section. The column UAL
-specifies the substructural features. Borrows and state are
-self-explanatory. ``Subsumption'' indicates that unrestricted elements
-can be promoted to affine and then linear. This applies to resources,
+specifies the substructural features supported (unrestricted, affine,
+linear). Borrows and state indicate support for the respective
+feature. ``Subsumption'' indicates that unrestricted elements 
+can be promoted to affine and then linear. This promotion applies to
+all objects, resources,
 borrows, or closures. ``Polymorphism'' refers to polymorphism over
-usage (unrestricted, affine, linear), notably for closures.
+substructural features, notably for closures.
 ``Identity'' indicates that the language supports a notion
@@
 \cref{identity}.
-Inference is self-explanatory, formalisation
-refers to the formal semantics and type soundness proof. Basis is self-explanatory.
+Inference is self-explanatory, formalization
+refers to the existence of a formal semantics and type soundness proof. Basis is self-explanatory.
 
@@
 Alms examples often rely on existential types to track the identity
-of objects. For instance
+of objects. For instance, consider the signature
 \lstinline/Array.create : int -> 'a -> \E 'b. ('a, 'b) array/ where
@@
 an ML-like module system.
-Rust's lifetime system is more explicit and more expressive than \lang.
+Instead of kinds, Rust uses marker traits like \lstinline/Copy/,
+\lstinline/Send/, and \lstinline/Sync/ to distinguish characteristics 
+of types, which would just be unrestricted in \lang.
+Also, Rust's lifetime system is more explicit and more expressive than \lang.
 While Rust provides partial lifetime inference, it

### 2020-06-04 kindly/extensions.tex
Remove the misleading definition of lists.
subsection concurrency

@@
 
-\subsubsection{Algebraic Datatypes}
+\subsubsection*{Algebraic Datatypes}
 
@@
 unrestricted.
-Here is the definition of a datatype for linear lists.
-
-\begin{lstlisting}[numbers=none]
-type ('a : 'k) llist : lin = Nil | Cons of ('a * 'a llist)
-\end{lstlisting}
-
 Our prototype implements non-recursive algebraic datatypes with
@@
 
-\subsubsection{Branching constructs}
+\subsubsection*{Branching constructs}
 
@@
 
+\subsection{Concurrency}
+\label{sec:concurrency}
+
+While our present semantic model does not consider concurrency, some
+design decisions were taken with a possible extension to concurrency
+in mind. The main impact of this foresight is the distinction between
+exclusive borrows and shared borrows, which materializes in the
+metatheory. The intended contract of the shared 
+borrow is that it can be duplicated and that the program consistently
+observes the same state of the underlying resource inside its region
+even in the presence of concurrency.
+
+The exclusive borrow, on the other hand, is propagated according to
+the evaluation order with the intention that any suspended binding
+split of from an exclusive borrow has finished its action on the
+resource before the borrow gets exercised. In the presence of
+concurrency, this intended semantics of the exclusive borrow should
+guarantee freedom of data races.
+
+That is, if a thread closes over a borrow, that thread should have
+terminated before the parent thread leaves the borrow's region. Rust
+addresses this lifetime issue with the \lstinline/move/ qualification
+for a thread which transfers ownership of the free variables to the
+thread. However, moving (in Rust) only applies to the resource itself, but not
+to borrows. A more discerning kind system would be needed for \lang to
+enable safe sharing of synchronizable resources or borrows analogously
+to Rust's \texttt{Sync} trait.
+
 \lstDeleteShortInline@

### 2020-06-04 kindly/split-jam.sh
update for final paper

@@ -1,2 +1,2 @@
-pdfjam -o paper.pdf main.pdf '1-25'
-pdfjam -o appendix.pdf main.pdf '26-'
+pdfjam -o paper.pdf main.pdf '1-29'
+pdfjam -o appendix.pdf main.pdf '30-'

### 2020-06-15 kindly/soa.tex
First set of fixes for the comparison matrix

@@
       \textbf{Language}
-      & URAL & \rot{Borrows} & \rot{State}
-      & \rot{R. Subsumption}
-      & \rot{R. Polymorphism}
+      & UAL & \rot{Borrows} & \rot{State}
+      & \rot{Resource Subsumption}
+      & \rot{Resource Polymorphism}
       & \rot{Identity}
@@
       Mezzo~\citep{DBLP:phd/hal/Protzenko14}
-      & UAL & --- & \Y & \M & \Y & \Y & \N & \MC Manual & --- \\
+      & UA & --- & \Y & \M & \Y & \Y & \MC Local & \YC Coq & --- \\
       \hline
-      Rust~\citep{rust}
+      Rust~\citep{rust,DBLP:journals/pacmpl/0002JKD18}
       & UA & \Y & \Y & \Y & \M & \N & \MC Local & \YC Coq & --- \\
       Plaid~\citep{DBLP:conf/oopsla/AldrichSSS09}
-      & UA & --- & \Y & \Y & \Y & \Y & \N & \MC Manual & --- \\
+      & UA & --- & \Y & \Y & \Y & \Y & \N & \MC Manual & Java \\
       \hline
@@
 linear). Borrows and state indicate support for the respective
-feature. ``Subsumption'' indicates that unrestricted elements 
+feature. ``Resource Subsumption'' indicates that unrestricted elements 
 can be promoted to affine and then linear. This promotion applies to
 all objects, resources,
-borrows, or closures. ``Polymorphism'' refers to polymorphism over
+borrows, or closures. ``Resource Polymorphism'' refers to polymorphism over
 substructural features, notably for closures.

### 2020-06-20 kindly/soa.tex

@@
 
-\begin{figure}[tp]
-  \begin{center}
 \newcommand\YC{\color{Green}}
@@
 \newcommand\N{{\NC{\ding{56}}}\xspace}
-    \newcommand\M{\MC{\textasciitilde}} % "Meh"
+\newcommand\M{{\MC{\textasciitilde}}\xspace} % "Meh"
+\begin{figure}[tp]
+  \begin{center}
 
@@
       System F\degree~\citep{DBLP:conf/tldi/MazurakZZ10}
-      & UL & --- & \N & \N & \N & \N & \N & \YC Coq & System F
+      & UL & \N & \N & \N & \N & \N & \N & \YC Coq & System F
       \\
       Alms~\citep{DBLP:conf/popl/TovP11}
-      & UA & --- & \Y & \M & \Y & \Y & \MC Local & \N & ML \\
+      & UA & \N & \Y & \M & \Y & \Y & \MC Local & \N & ML \\
       Quill~\citep{DBLP:conf/icfp/Morris16}
-      & UL & --- & \N & \N & \Y & \N & \YC Principal & \MC Manual & Qualified types\\
+      & UL & \N & \N & \N & \Y & \N & \YC Principal & \MC Manual & Qualified types\\
       Lin. Haskell~\citep{DBLP:journals/pacmpl/BernardyBNJS18}
-      & UL & --- & \M & \N & \Y & \N & \MC Non-principal & \MC Manual & Haskell \\
+      & UL & \N & \M & \N & \Y & \N & \MC Non-principal & \MC Manual & Haskell \\
       Mezzo~\citep{DBLP:phd/hal/Protzenko14}
-      & UA & --- & \Y & \M & \Y & \Y & \MC Local & \YC Coq & --- \\
+      & UA & \M & \Y & \M & \Y & \Y & \MC Local & \YC Coq & --- \\
       \hline
@@
       Plaid~\citep{DBLP:conf/oopsla/AldrichSSS09}
-      & UA & --- & \Y & \Y & \Y & \Y & \N & \MC Manual & Java \\
+      & UA & \N & \Y & \Y & \Y & \Y & \N & \MC Manual & Java \\
       \hline
@@
 The comparison matrix in \cref{fig:comparison-matrix} gives an
-overview over the systems discussed in this section. The column UAL
+overview over the systems discussed in this section.
+Each column indicates if a feature is present (\Y), absent (\N) or
+partially supported (\M) such as when the feature is limited or can
+only be obtained through a non-trivial encoding.
+The column UAL
 specifies the substructural features supported (unrestricted, affine,

### 2020-06-20 kindly/soa.tex
matrix and explanation

--- 2020-06-15 kindly/soa.tex
@@
 
-    \begin{tabular}{l|ccccccccc}
+    \begin{tabular}{l|c@{~}ccccccccc}
       \textbf{Language}
@@
       & \rot{Identity}
+      & \rot{Concurrency}
       & \rot{Inference} & \rot{Formalisation}
@@
       System F\degree~\citep{DBLP:conf/tldi/MazurakZZ10}
-      & UL & --- & \N & \N & \N & \N & \N & \YC Coq & System F
+      & UL & \N & \N & \N & \N & \N & \N & \N & \YC Coq & System F
       \\
       Alms~\citep{DBLP:conf/popl/TovP11}
-      & UA & --- & \Y & \M & \Y & \Y & \MC Local & \N & ML \\
+      & UA & \N & \Y & \M & \Y & \Y & \Y & \MC Local & \N & ML \\
       Quill~\citep{DBLP:conf/icfp/Morris16}
-      & UL & --- & \N & \N & \Y & \N & \YC Principal & \MC Manual & Qualified types\\
+      & UL & \N & \N & \N & \Y & \N & \N & \YC Principal & \MC Manual & Qualified types\\
       Lin. Haskell~\citep{DBLP:journals/pacmpl/BernardyBNJS18}
-      & UL & --- & \M & \N & \Y & \N & \MC Non-principal & \MC Manual & Haskell \\
-      Mezzo~\citep{DBLP:phd/hal/Protzenko14}
-      & UA & --- & \Y & \M & \Y & \Y & \MC Local & \YC Coq & --- \\
+      & UL & \N & \M & \N & \Y & \N & \M & \MC Non-pr. & \MC Manual & Haskell \\
+      Mezzo~\citep{DBLP:phd/hal/Protzenko14,DBLP:journals/toplas/BalabonskiPP16}
+      & UA & \M & \Y & \M & \Y & \Y & \Y & \MC Local & \YC Coq & ML \\
       \hline
       Rust~\citep{rust,DBLP:journals/pacmpl/0002JKD18}
-      & UA & \Y & \Y & \Y & \M & \N & \MC Local & \YC Coq & --- \\
-      Plaid~\citep{DBLP:conf/oopsla/AldrichSSS09}
-      & UA & --- & \Y & \Y & \Y & \Y & \N & \MC Manual & Java \\
+      & UA & \Y & \Y & \Y & \M & \N & \Y & \MC Local & \YC Coq & --- \\
+      Plaid~\citep{DBLP:conf/oopsla/AldrichSSS09,DBLP:journals/toplas/GarciaTWA14}
+      & UA & \N & \Y & \Y & \Y & \Y & \N   & \N & \MC Manual & Java \\
       \hline
       \lang
-      & UAL & \Y & \Y & \Y & \Y & \N & \YC Principal & \MC Manual & ML/HM(X)
+      & UAL & \Y & \Y & \Y & \Y & \N & \N & \YC Principal & \MC Manual & ML/HM(X)
       \\
@@
 The comparison matrix in \cref{fig:comparison-matrix} gives an
-overview over the systems discussed in this section. The column UAL
+overview over the systems discussed in this section.
+Each column indicates whether a feature is present (\Y), absent (\N), or
+partially supported (\M), i.e., if the feature is limited or can
+only be obtained through a non-trivial encoding.
+Features are selected according to their contribution to expressivity
+and programmer convenience.
+The column UAL
 specifies the substructural features supported (unrestricted, affine,
@@
 \cref{identity}.
-Inference is self-explanatory, formalization
-refers to the existence of a formal semantics and type soundness proof. Basis is self-explanatory.
+``Concurrency'' indicates whether the language supports
+concurrency. For example, the implementation of Linear Haskell
+supports state and concurrency, but its theory covers neither.
+Type ``inference''  can be local, principal, or non-principal (if the
+inferred type is not necessary the most general). ``Formalization''
+refers to the existence of a formal semantics and type soundness
+proof. ``Basis'' indicates the heritage or inspiration of the language.
 
@@
 
-Mezzo~\citep{DBLP:phd/hal/Protzenko14} is an ML-like language
+Mezzo~\citep{DBLP:phd/hal/Protzenko14,DBLP:journals/toplas/BalabonskiPP16} is an ML-like language
 with a rich capability system which is able to encode numerous
@@
 Vault~\citep{DBLP:conf/pldi/DeLineF01}
-and Plaid~\citep{DBLP:conf/oopsla/AldrichSSS09}
+and Plaid~\citep{DBLP:conf/oopsla/AldrichSSS09,DBLP:journals/toplas/GarciaTWA14}
 leverage typestate and capabilities

### 2020-06-22 kindly/soa.tex
language
Add some notes on escape hatch.

@@
 
-    \begin{tabular}{l|c@{~}ccccccccc}
+    \begin{tabular}{l|c@{~}cccccccccc}
       \textbf{Language}
@@
       & \rot{Concurrency}
+      & \rot{Escape hatch}
       & \rot{Inference} & \rot{Formalisation}
@@
       System F\degree~\citep{DBLP:conf/tldi/MazurakZZ10}
-      & UL & \N & \N & \N & \N & \N & \N & \N & \YC Coq & System F
+      & UL & \N & \N & \N & \N & \N & \N & \N & \N & \YC Coq & System F
       \\
       Alms~\citep{DBLP:conf/popl/TovP11}
-      & UA & \N & \Y & \M & \Y & \Y & \Y & \MC Local & \N & ML \\
+      & UA & \N & \Y & \M & \Y & \Y & \Y & \Y & \MC Local & \N & ML \\
       Quill~\citep{DBLP:conf/icfp/Morris16}
-      & UL & \N & \N & \N & \Y & \N & \N & \YC Principal & \MC Manual & Qualified types\\
+      & UL & \N & \N & \N & \Y & \N & \N & \N & \YC Principal & \MC Manual & Qual. types\\
       Lin. Haskell~\citep{DBLP:journals/pacmpl/BernardyBNJS18}
-      & UL & \N & \M & \N & \Y & \N & \M & \MC Non-pr. & \MC Manual & Haskell \\
+      & UL & \N & \M & \N & \Y & \N & \M & \M & \MC Non-pr. & \MC Manual & Haskell \\
       Mezzo~\citep{DBLP:phd/hal/Protzenko14,DBLP:journals/toplas/BalabonskiPP16}
-      & UA & \M & \Y & \M & \Y & \Y & \Y & \MC Local & \YC Coq & ML \\
+      & UA & \M & \Y & \M & \Y & \Y & \Y & \Y & \MC Local & \YC Coq & ML \\
       \hline
       Rust~\citep{rust,DBLP:journals/pacmpl/0002JKD18}
-      & UA & \Y & \Y & \Y & \M & \N & \Y & \MC Local & \YC Coq & --- \\
+      & UA & \Y & \Y & \Y & \M & \N & \Y & \Y & \MC Local & \YC Coq & --- \\
       Plaid~\citep{DBLP:conf/oopsla/AldrichSSS09,DBLP:journals/toplas/GarciaTWA14}
-      & UA & \N & \Y & \Y & \Y & \Y & \N   & \N & \MC Manual & Java \\
+      & UA & \N & \Y & \Y & \Y & \Y & \N & \Y & \N & \MC Manual & Java \\
       \hline
       \lang
-      & UAL & \Y & \Y & \Y & \Y & \N & \N & \YC Principal & \MC Manual & ML/HM(X)
+      & UAL & \Y & \Y & \Y & \Y & \N & \N & \M & \YC Principal & \MC Manual & ML/HM(X)
       \\
@@
 only be obtained through a non-trivial encoding.
-Features are selected according to their contribution to expressivity
-and programmer convenience.
+Features are selected according to their relevance for resource
+management and programmer convenience.
 The column UAL
@@
 supports state and concurrency, but its theory covers neither.
+``Escape hatch'' indicates whether users can (locally) opt out of
+the resource management through language-integrated means such
+as Rust's \texttt{unsafe}. Partial support, in the case of \lang
+for instance, indicates this feature is available, but not formalized.
 Type ``inference''  can be local, principal, or non-principal (if the

### 2020-06-23 kindly/extensions.tex
fixed unfortunate language about branching

@@
 
-Our formalization of \lang does not contain branching constructs. As a consequence,
-subexpressions are always executed sequentially. Ordering is import
-for the splitting relation introduced in \cref{sdtyping}, as it must
-coincide with the order of evaluation. To account for borrows and linear
-values in different branches, we need a different, straightforward (symmetric) join relation.
-Its definition follows immediately from the semantics of borrows and linearity
-shown so far and it must be used in the typing rules for branching
-constructs such as if-then-else and pattern matching.
+Our formalization of \lang does not cover conditionals. In the
+typing rules, a conditional is supported as usual by checking all
+branches with the same constraint and typing environment and requiring
+the return types to match.
+It materializes in the inference algorithm as a straightforward
+(symmetric) join relation which is used for all elimination rules on
+sum types.
 This extension is implemented in our prototype.

### 2020-06-23 kindly/soa.tex
fixed final requirements on related work

@@
       \textbf{Language}
-      & UAL & \rot{Borrows} & \rot{State}
-      & \rot{Resource Subsumption}
-      & \rot{Resource Polymorphism}
+      & UAL
+      & \rot{State}
+      & \rot{Borrows}
+      & \rot{Multiplicity Subsumption}
+      & \rot{Multiplicity Polymorphism}
       & \rot{Identity}
@@
       Alms~\citep{DBLP:conf/popl/TovP11}
-      & UA & \N & \Y & \M & \Y & \Y & \Y & \Y & \MC Local & \N & ML \\
+      & UA & \Y & \N & \M & \Y & \Y & \Y & \Y & \MC Local & \N & ML \\
       Quill~\citep{DBLP:conf/icfp/Morris16}
@@
       Lin. Haskell~\citep{DBLP:journals/pacmpl/BernardyBNJS18}
-      & UL & \N & \M & \N & \Y & \N & \M & \M & \MC Non-pr. & \MC Manual & Haskell \\
+      & UL & \M & \N & \N & \Y & \N & \M & \M & \MC Non-pr. & \MC Manual & Haskell \\
       Mezzo~\citep{DBLP:phd/hal/Protzenko14,DBLP:journals/toplas/BalabonskiPP16}
-      & UA & \M & \Y & \M & \Y & \Y & \Y & \Y & \MC Local & \YC Coq & ML \\
+      & UA & \Y & \M & \M & \Y & \Y & \Y & \Y & \MC Local & \YC Coq & ML \\
       \hline
@@
       Plaid~\citep{DBLP:conf/oopsla/AldrichSSS09,DBLP:journals/toplas/GarciaTWA14}
-      & UA & \N & \Y & \Y & \Y & \Y & \N & \Y & \N & \MC Manual & Java \\
+      & UA & \Y & \N & \Y & \Y & \Y & \N & \Y & \N & \MC Manual & Java \\
       \hline
@@
 \end{figure}
+
+The comparison matrix in \cref{fig:comparison-matrix} gives an
+overview over the systems discussed in this section.
+Each column indicates whether a feature is present (\Y), absent (\N), or
+partially supported (\M), i.e., if the feature is limited or can
+only be obtained through a non-trivial encoding.
+Features are selected according to their relevance for type-based resource
+management and programmer convenience.
+
+The column UAL
+specifies the substructural features supported (Unrestricted, Affine,
+Linear). The columns ``State'' and ``Borrows'' indicate support for the respective
+feature. In an ideal world, the presence of linearity and state
+indicates that the system is able to support safe manual memory
+management as linearity enforces manual deallocation. True affinity
+and state only works with garbage collection, which eventually
+automatically finalizes an object that is no longer referenced. In practice, this
+distinction is often watered down. For example, Rust automatically
+destructs objects at the end of their lifetime, which creates the
+illusion of affinity while the low-level code is strictly linear. On
+the other hand, there are ways to consume an object at the source
+level without invoking its destructor (using
+\lstinline/mem::forget/)\footnote{See
+  \url{https://doc.rust-lang.org/nomicon/leaking.html} which contains
+  further examples and discussion. Thanks to
+  Derek Dreyer and Ralf Jung for pointing this out.}
+where the high-level code exhibits linearity, but the low-level code
+is affine.
+
+``Multiplicity Subsumption'' indicates that unrestricted elements 
+can be promoted to affine and then linear. This promotion applies to
+objects, resources, borrows, and closures. ``Multiplicity Polymorphism'' refers to polymorphism over
+substructural features: a function can be parameterized over the multiplicity
+restriction of an object. For instance, the type of function
+composition should express that applies to functions with linear,
+affine, and unrestricted multiplicity and returns a function with the
+same multiplicity. 
+``Identity'' indicates that the language supports a notion
+of identity, usually through existential types, as described in
+\cref{identity}.
+``Concurrency'' indicates whether the language supports
+concurrency. For example, the implementation of Linear Haskell
+supports state and concurrency, but its theory covers neither.
+``Escape hatch'' indicates whether a programmer can (locally) opt out of
+resource management through language-integrated means such
+as Rust's \texttt{unsafe}. Partial support ``\M'', in the case of \lang
+for instance, indicates that this feature is available, but not formalized.
+Type ``inference''  can be local, principal, or non-principal (if the
+inferred type is not necessarily the most general one). ``Formalization''
+refers to the existence of a formal semantics and type soundness
+proof. ``Basis'' indicates the heritage or inspiration of the language.
 
@@
 None of the following languages support borrows.
-
-The comparison matrix in \cref{fig:comparison-matrix} gives an
-overview over the systems discussed in this section.
-Each column indicates whether a feature is present (\Y), absent (\N), or
-partially supported (\M), i.e., if the feature is limited or can
-only be obtained through a non-trivial encoding.
-Features are selected according to their relevance for resource
-management and programmer convenience.
-The column UAL
-specifies the substructural features supported (unrestricted, affine,
-linear). Borrows and state indicate support for the respective
-feature. ``Resource Subsumption'' indicates that unrestricted elements 
-can be promoted to affine and then linear. This promotion applies to
-all objects, resources,
-borrows, or closures. ``Resource Polymorphism'' refers to polymorphism over
-substructural features, notably for closures.
-``Identity'' indicates that the language supports a notion
-of identity, usually through existential types, as described in
-\cref{identity}.
-``Concurrency'' indicates whether the language supports
-concurrency. For example, the implementation of Linear Haskell
-supports state and concurrency, but its theory covers neither.
-``Escape hatch'' indicates whether users can (locally) opt out of
-the resource management through language-integrated means such
-as Rust's \texttt{unsafe}. Partial support, in the case of \lang
-for instance, indicates this feature is available, but not formalized.
-Type ``inference''  can be local, principal, or non-principal (if the
-inferred type is not necessary the most general). ``Formalization''
-refers to the existence of a formal semantics and type soundness
-proof. ``Basis'' indicates the heritage or inspiration of the language.
 
@@
 an ML-like module system.
-Instead of kinds, Rust uses marker traits like \lstinline/Copy/,
-\lstinline/Send/, and \lstinline/Sync/ to distinguish characteristics 
-of types, which would just be unrestricted in \lang.
-Also, Rust's lifetime system is more explicit and more expressive than \lang.
+Everything is affine in Rust and marker traits like \lstinline/Copy/,
+\lstinline/Send/, and \lstinline/Sync/ are used to modulate the characteristics 
+of types. \lang relies on kinds to express substructural properties of
+types and marker traits may be considered as implementing a
+fine-grained kind structure on Rust types.
+Rust's lifetime system is more explicit and more expressive than
+\lang's regions.
 While Rust provides partial lifetime inference, it
 does not support full type inference.
-Recently, \citet{DBLP:journals/corr/abs-1903-00982}
-formalized Rust's ownership discipline, including non-lexical lifetimes.
 Moreover, Rust programmers have full control over memory allocation
@@
 foundation for creating such proofs by exhibiting a framework for
-semantic soundness of the Rust type system that incorporates aspects
-of concurrecy (i.e., data-race freedom). Similar proof obligations
+semantic soundness of the Rust type system in terms of a low-level
+core language that incorporates aspects
+of concurrency (i.e., data-race freedom). Similar proof obligations
 would be needed in \lang to check that an implementation of the module
@@
 cover concurrency.
+
+\citet{DBLP:journals/corr/abs-1903-00982}
+formalize Rust's ownership discipline from a source-level
+perspective. Their approach is purely syntactic and is therefore not
+able to reason about unsafe fragments of Rust code. However, their
+flow-sensitive type discipline enables soundness proofs for
+non-lexical lifetimes, which have been adopted in Rust, but cannot be
+expressed in \lang at present.
 

### 2020-06-25 kindly/soa.tex
The spacey magic as worked, the paper fits!

@@
 and state only works with garbage collection, which eventually
-automatically finalizes an object that is no longer referenced. In practice, this
+automatically finalizes an object no longer referenced. In practice, this
 distinction is often watered down. For example, Rust automatically
-destructs objects at the end of their lifetime, which creates the
-illusion of affinity while the low-level code is strictly linear. On
-the other hand, there are ways to consume an object at the source
+destructs objects at the end of their lifetime, creating the
+illusion of affinity while the low-level code is strictly linear.
+However, there are ways to consume an object at the source
 level without invoking its destructor (using

### 2020-06-30 kindly/main.tex
Add a section with some potential extensions.
Pre-submit fixes.
first round of revision
Fixed compilation error '\Bbbk already defined'.
Expand the comparison table.
acknowledgments
Prepare for camery ready.

@@
 \PassOptionsToPackage{svgnames}{xcolor}
-\documentclass[acmsmall,review,anonymous]{acmart}
-\settopmatter{printfolios=true,printccs=false,printacmref=false}
+\documentclass[acmsmall,screen]{acmart}
+
+% The page limit for final papers is 27 pages, not including references (but 
+% including appendices). Additional materials may be uploaded to ACM's 
+% Digital Library, in a manner that the publisher will clarify in 
+% instructions sent to you separately. In extraordinary circumstances, I'm 
+% able to grant further pages still, and please write to me if you think your 
+% paper might qualify. There's no fee associated with the number of pages 
+% you use, but we do want to be fair to all authors in allocating space.
 
@@
 \usepackage{amsfonts}
-\usepackage{amssymb}
+% \usepackage{amssymb}
 \usepackage{amsthm}
@@
 \usepackage{array}
+\usepackage{adjustbox}
 \usepackage{bm}%Decent bolding for math symbols
+\usepackage{pifont}%% check mark
 
@@
 \bibliographystyle{ACM-Reference-Format}
+\citestyle{acmauthoryear}
+
+%%% The following is specific to ICFP '20 and the paper
+%%% 'Kindly Bent to Free Us'
+%%% by Gabriel Radanne, Hannes Saffrich, and Peter Thiemann.
+%%%
+\copyrightyear{2020}
+\acmSubmissionID{icfp20main-p74-p}
+\setcopyright{rightsretained}
+\acmJournal{PACMPL}
+\acmYear{2020} \acmVolume{4} \acmNumber{ICFP} \acmArticle{103} \acmMonth{8} \acmPrice{}\acmDOI{10.1145/3408985}
+
+\begin{CCSXML}
+<ccs2012>
+   <concept>
+       <concept_id>10011007.10011006.10011039.10011311</concept_id>
+       <concept_desc>Software and its engineering~Semantics</concept_desc>
+       <concept_significance>500</concept_significance>
+       </concept>
+   <concept>
+       <concept_id>10011007.10011006.10011008.10011009.10011012</concept_id>
+       <concept_desc>Software and its engineering~Functional languages</concept_desc>
+       <concept_significance>500</concept_significance>
+       </concept>
+   <concept>
+       <concept_id>10011007.10011006.10011008.10011009.10011010</concept_id>
+       <concept_desc>Software and its engineering~Imperative languages</concept_desc>
+       <concept_significance>300</concept_significance>
+       </concept>
+ </ccs2012>
+\end{CCSXML}
+
+\ccsdesc[500]{Software and its engineering~Semantics}
+\ccsdesc[500]{Software and its engineering~Functional languages}
+\ccsdesc[300]{Software and its engineering~Imperative languages}
 
@@
 \affiliation{
-  \institution{University of Freiburg}
-  \country{Germany}
+  \institution{Inria}
+  \country{Paris}
 }
-\email{radanne@informatik.uni-freiburg.de}
+\email{gabriel.radanne@inria.fr}
 
@@
 \input{metatheory}
+\input{extensions}
 \input{soa}
 \input{conclusion}
+
+\begin{acks}
+  This material is based upon work supported by the German Research
+  Council, \grantsponsor{sponsor01}{DFG}{https://www.dfg.de/}, project reference number
+  \grantnum{sponsor01}{TH 665/11-1}. We are indebted to the anonymous reviewers
+  for their thoughtful and constructive comments.
+\end{acks}
 
@@
 
-\section{Further Examples}
+\section{Further examples}
 \label{sec:extra-example}

### 2020-06-30 kindly/infer/infer.tex
Fix and shrinking.

@@
 
-\subsection{Type Inference}
+\subsection{Type inference}
 
@@
 
-\paragraph{Environments and Bindings}
+\paragraph{Environments and bindings}
 \label{infer:envs}
@@
 
-\paragraph{Splitting and Regions}
+\paragraph{Splitting and regions}
 \label{infer:split}
@@
 
-\subsection{Soundness and Principality}
+\subsection{Soundness and principality}
 
@@
 
-\begin{definition}[Flattened Environment]
+\begin{definition}[Flattened environment]
 A flattened environment,

### 2020-06-30 kindly/intro.tex
remarks by rev#C
Fix and shrinking.

@@
 the resource, but it cannot release the resource. Only the original
-owner of the resource has all rights to it and can release it.
+owner of the resource has the rights to release it.
 
@@
 \end{lstlisting}
-The remaining arrows in
-the API are unrestricted and we write @->@ instead of the
+The remaining arrows are unrestricted and we write @->@ instead of the
 explicitly annotated @-{un}>@.
@@
 region. In both regions the borrow is consumed immediately by passing
-it to @File.write@. Regions are generally inferred by \lang.
+it to @File.write@. \lang elaborates regions automatically before type
+inference. Alternatively, programmers may mark regions explicitly (See
+\cref{sec:imper-progr}).
 
@@
 application and composition are the archetypical
-examples for functions exploiting that feature.\footnote{Compared to
-  Quill \cite{DBLP:conf/icfp/Morris16} the signatures of application and
-  composition are simpler because \lang{} supports kind subsumption.}
+examples for functions exploiting that feature.\footnote{\lang{}'s signature
+  for application and composition are simpler than
+  Quill \cite{DBLP:conf/icfp/Morris16} thanks to kind subsumption.}
 For application, \lang{} infers the following type.
@@
   express multiplicity constraints on the number of uses of a
-  value. This type system is a conservative extension of systems for
+  value. This type system is a conservative extension of
   existing ML-like languages.

### 2020-06-30 kindly/motivation-persistent.tex
Fix and shrinking.

@@
 % \cite{DBLP:conf/ml/ConchonF07}.
-
+%
 In the example code, we make use of a two-dimensional version
@@
   \begin{subfigure}[t]{0.45\linewidth}
-\begin{lstlisting}[numbers=none]
+\begin{lstlisting}
 module type HYBARRAY = sig
@@
   \begin{subfigure}[t]{0.55\linewidth}
-\begin{lstlisting}[numbers=none]
+\begin{lstlisting}
 module CowArray : HYBARRAY = struct
@@
 
-\begin{lstlisting}[numbers=none]
-val propagate : int -> int -> &!Matrix.t -> int -> unit
-val solve : int -> int -> Matrix.t -> unit
-\end{lstlisting}
-
   \begin{subfigure}[t]{0.45\linewidth}
@@
   % \lstinputlisting[linerange=1-31,multicols=2]{code/sukodu.affe}
-  \vspace{-15pt}
+  \vspace{-10pt}
+\begin{lstlisting}
+val propagate : int -> int -> &!Matrix.t -> int -> unit
+val solve : int -> int -> Matrix.t -> unit
+\end{lstlisting}
+  \vspace{-10pt}
   \caption{Excerpt of the Sudoku solver}
@@
 
-Our implementation of a Sudoku solver (~\cref{ex:sudoku}) represents
+Our implementation of a Sudoku solver (\cref{ex:sudoku}) represents
 the board as a 2D-matrix (\cref{line:boardtype}).
@@
 @IntSet.remove@ produces a new value.
-
+%
 The main functions are @solve@ and @propagate@ with the typings shown
@@
 
-As the @board@ parameter to the @propagate@ function is an exclusive
+The @board@ parameter of the @propagate@ function is an exclusive
 borrow, it should be handled 
@@
 borrow) the board three times in 
-\cref{line:reborrow1}-\cref{line:reborrow3}.
+\crefrange{line:reborrow1}{line:reborrow3}.
 The function @propagate_line@ also contains two reborrows of the
-exclusive borrow argument @board@, an
+exclusive borrow @board@, an
 immutable one (\cref{line:reborrow_i}) and an exclusive one

### 2020-06-30 kindly/motivation.tex
Fix and shrinking.

@@
   \centering
-\begin{lstlisting}[numbers=none]
+  \begin{minipage}{0.5\linewidth}
+\begin{lstlisting}
 module Array : sig
@@
   \vspace{-10pt}
+\end{minipage}~
+\begin{minipage}{0.5\linewidth}
+\begin{lstlisting}
+let mk_fib_array n =
+  let a = create (n, 1) in
+  for i = 2 to n - 1 do
+    let x = get (&a, i-1) + get (&a, i-2) in(*@\label{ex:array:get}*)
+    set (&!a, i, x)(*@\label{ex:array:set}*)
+  done;
+  a(*@\label{ex:array:return}*)
+# mk_fib_array : int -> int Array.t
+\end{lstlisting}
+  \vspace{-5pt}
+  \caption{An array of Fibonacci numbers}
+  \label{ex:fibo}
+  \vspace{-10pt}
+\end{minipage}
+
 \end{figure}
@@
 we can create an
-array of Fibonacci numbers in an imperative coding style:
-\begin{lstlisting}
-let mk_fib_array n =
-  let a = create (n, 1) in
-  for i = 2 to n - 1 do
-    let x = get (&a, i-1) + get (&a, i-2) in(*@\label{ex:array:get}*)
-    set (&!a, i, x)(*@\label{ex:array:set}*)
-  done;
-  a(*@\label{ex:array:return}*)
-# mk_fib_array : int -> int Array.t
-\end{lstlisting}
-
+array of Fibonacci numbers in an imperative coding style, as shown
+in \cref{ex:fibo}.
+%
 After

### 2020-06-30 kindly/soa.tex
Fix and shrinking.

@@
       \\\hline
-      System F\degree~\citep{DBLP:conf/tldi/MazurakZZ10}
+      System F\degree%~\citep{DBLP:conf/tldi/MazurakZZ10}
       & UL & \N & \N & \N & \N & \N & \N & \N & \N & \YC Coq & System F
       \\
-      Alms~\citep{DBLP:conf/popl/TovP11}
+      Alms%~\citep{DBLP:conf/popl/TovP11}
       & UA & \Y & \N & \M & \Y & \Y & \Y & \Y & \MC Local & \N & ML \\
-      Quill~\citep{DBLP:conf/icfp/Morris16}
+      Quill%~\citep{DBLP:conf/icfp/Morris16}
       & UL & \N & \N & \N & \Y & \N & \N & \N & \YC Principal & \MC Manual & Qual. types\\
-      Lin. Haskell~\citep{DBLP:journals/pacmpl/BernardyBNJS18}
+      Lin. Haskell%~\citep{DBLP:journals/pacmpl/BernardyBNJS18}
       & UL & \M & \N & \N & \Y & \N & \M & \M & \MC Non-pr. & \MC Manual & Haskell \\
-      Mezzo~\citep{DBLP:phd/hal/Protzenko14,DBLP:journals/toplas/BalabonskiPP16}
+      Mezzo%~\citep{DBLP:phd/hal/Protzenko14,DBLP:journals/toplas/BalabonskiPP16}
       & UA & \Y & \M & \M & \Y & \Y & \Y & \Y & \MC Local & \YC Coq & ML \\
       \hline
-      Rust~\citep{rust,DBLP:journals/pacmpl/0002JKD18}
+      Rust%~\citep{rust,DBLP:journals/pacmpl/0002JKD18}
       & UA & \Y & \Y & \Y & \M & \N & \Y & \Y & \MC Local & \YC Coq & --- \\
-      Plaid~\citep{DBLP:conf/oopsla/AldrichSSS09,DBLP:journals/toplas/GarciaTWA14}
+      Plaid%~\citep{DBLP:conf/oopsla/AldrichSSS09,DBLP:journals/toplas/GarciaTWA14}
       & UA & \Y & \N & \Y & \Y & \Y & \N & \Y & \N & \MC Manual & Java \\

### 2020-06-30 kindly/main.tex
Final preparation for preprint.

@@
 \usepackage{graphicx}
-
+\usepackage{microtype}
 \usepackage{xspace}
@@
        </concept>
-   <concept>
-       <concept_id>10011007.10011006.10011008.10011009.10011010</concept_id>
-       <concept_desc>Software and its engineering~Imperative languages</concept_desc>
-       <concept_significance>300</concept_significance>
-       </concept>
  </ccs2012>
@@
 \ccsdesc[500]{Software and its engineering~Functional languages}
-\ccsdesc[300]{Software and its engineering~Imperative languages}
 
@@
 
+\keywords{Functional programming, Linear types, Ownership, Type inference}
+
 \maketitle

### 2020-07-16 kindly/infer/infer.tex
More final version tweaks
Urk.

@@
 in terms of these new relations by using constraint entailment.
-All relations are fully described in \cref{typ:extra:envs}.
+All relations are fully described in Appendix D.
 
@@
 
-\subsection{Type inference}
+\subsection{Type Inference}
 
@@
 system in \cref{rule:infer:envs}.
-The complete type inference rules are shown in \cref{appendix:infer}.
+The complete type inference rules are shown in Appendix E.
 
-\paragraph{Environments and bindings}
+\paragraph{Environments and Bindings}
 \label{infer:envs}
@@
 
-\paragraph{Splitting and regions}
+\paragraph{Splitting and Regions}
 \label{infer:split}
@@
 The complete constraint system is
-defined in \cref{appendix:constraints}.
+defined in Appendix C.
 
-\subsubsection{Constraints normalization by example}
+\subsubsection{Constraints Normalization by Example}
 
@@
 
-\subsubsection{Properties of the constraint system}
+\subsubsection{Properties of the Constraint System}
 
@@
 
-\subsection{Soundness and principality}
+\subsection{Soundness and Principality}
 

### 2020-07-16 kindly/lang.tex
More final version tweaks
Urk.

@@
-\section{The \lang language}
+\section{The \lang Language}
 
@@
 
-\subsection{Automatic region annotation}
+\subsection{Automatic Region Annotation}
 \label{regionannot}
@@
 We give an informal presentation of our code transformation and defer
-the complete definition to \cref{appendix:regionannot}. This code transformation
+the complete definition to Appendix B. This code transformation
 aims to find, for each borrow $\borrow{x}$, the biggest region satisfying
@@
 parts of the type system. A complete
-description is available in \cref{appendix:sdtyping}.
+description is available in Appendix D.
 % In this section, we focus on chosen parts of our system
@@
 \cref{fig:big-step-interpretation} and leave the full
-definition for \cref{sec:semant-defin}.
+definition for Appendix F.
 % The appendix also

### 2020-07-16 kindly/motivation.tex
More final version tweaks
Urk.

@@
 \cite{DBLP:journals/jfp/Padovani17}; see also
-\cref{sec:session-linearity} in the supplement) as well as other
+Appendix A.1 in the supplement) as well as other
 functional resource handling. But it really shines
@@
 primitives needed to implement them.
-Further examples are available in \cref{sec:extra-example}.
+Further examples are available in Appendix A.
 
-\subsection{Imperative programming with linear arrays}
+\subsection{Imperative Programming with Linear Arrays}
 \label{sec:imper-progr}
@@
 
-\subsection{Iterators and regions}
+\subsection{Iterators and Regions}
 \label{tuto:vector}
@@
 This pattern is essential in many use cases of linearity such
-as pools of linear objects (see \cref{tuto:pool}).
+as pools of linear objects (see Appendix A.2).
 

### 2020-07-28 kindly/slides-icfp.tex
First draft of the ICFP slides.

\documentclass[aspectratio=169,dvipsnames,svgnames,10pt]{beamer}

\usepackage[T1]{fontenc}
\usepackage[utf8]{inputenc}
\usepackage[french, english]{babel}
\selectlanguage{english}

\usepackage{graphicx}

% Math
\usepackage{amsmath}
% \usepackage{amsfonts}
\usepackage{amssymb}
\usepackage{amsthm}
% \usepackage{mathrsfs}
\usepackage{mathtools}
\usepackage{textcomp}
% \usepackage{textgreek}

% Specialized packages
% \usepackage{syntax} % Grammar definitions
\usepackage{verbatim}
\usepackage{listings} % Code
\usepackage{xspace} % Useful for macros
\usepackage{natbib}% Good citations and bibliography
\usepackage{mathpartir} % Syntax trees
\usepackage{colortbl}
\usepackage{hhline}
\usepackage{multicol}%multicolumn lists
\usepackage{pifont}%% check mark
\usepackage{minted}

% \usepackage{mathptmx}
% \usepackage[scaled=0.9]{helvet}
\usepackage{beramono}

\bibliographystyle{ACM-Reference-Format}
\citestyle{acmauthoryear}
\usepackage{appendixnumberbeamer}

\usetheme{metropolis}
\beamertemplatenavigationsymbolsempty
\setbeamercovered{transparent=20}

\usepackage{wasysym}

\usepackage{tikz}
\usetikzlibrary{decorations.text,backgrounds,positioning,shapes,
  shadings,shadows,arrows,decorations.markings,calc,fit,fadings,
  tikzmark
}

\def\HUGE{\fontsize{35pt}{15pt}\selectfont}

\input{prelude}
\input{caml}
\input{notations}
\input{rulemacros}

\newcommand\Y{{\color{Green}{\ding{52}}}}
\newcommand\N{{\color{Red}{\ding{56}}}}
\newcommand\M{{\color{Orange}{\textasciitilde}}}

\lstset{
  tabsize=4,
  aboveskip={0.5\baselineskip},
  belowcaptionskip=0.5\baselineskip,
  basicstyle=\scriptsize\ttfamily,
}

\title{Kindly Bent To Free Us}
\author{\textbf{Gabriel \textsc{Radanne}}
  \and Hannes \textsc{Saffrich}
  \and Peter \textsc{Thiemann}}

\begin{document}

\lstMakeShortInline[keepspaces,basicstyle=\small\ttfamily]@

\frame[plain]{\titlepage}

\begin{frame}
  \frametitle{Motivation}

  Memory safety is the cause of numerous bugs and security issues.

  Recent classification (2015-2020) of \og high severity security bugs\fg in Chromium:

  \begin{figure}[h]
    \centering
    \includegraphics[width=0.55\textwidth]
    {chromium-use-after-free}
  \end{figure}

  \pause
  Chromium is written in C/C++!\\
  Surely these bugs don't happen in high-level typed languages \dots

\end{frame}

\begin{frame}[fragile]
  \frametitle{Let's write some OCaml code}

\begin{minted}{OCaml}
let db = Dbm.opendbm "mydb" [Dbm_rdwr] 0o666 in
Dbm.add db "hello" "world";
(* ... *)
Dbm.close db;
(* ... *)
print_string (Dbm.find db "hello") (* bug! *)
\end{minted}

\end{frame}

\begin{frame}
  \frametitle{Solutions?}

  Existing ways to solve this

\end{frame}

\begin{frame}[fragile]
  \frametitle{Affe!}

\begin{minted}[texcomments=true]{OCaml}
let db = Dbm.opendbm "firstnames" [Dbm_rdwr] 0o666 in
Dbm.add &!db "Radanne" "Gabriel";
(* ... *)
Dbm.close db;
(* ... *)
print_string (Dbm.find &db "Saffrich") (* \N Compile time error! *)
\end{minted}

  \TODO{Add lot's of annotation w. arrows/pointers to explain the code piece by piece}

\end{frame}

\begin{frame}[fragile]{A recap on borrows}

  \TODO{C/C, to refine}

  Borrows allow to lend usage of something to someone else.

  There are different types of borrows:
  \begin{itemize}
  \item Shared borrows @&a@ which are \emph{Unrestricted} (@un@)
  \item Exclusive borrows @&!a@ which are \emph{Affine} (@aff@)
  \end{itemize}

  We cannot use a borrow of @a@ and @a@ itself at the same time.\\
  A borrow must not escape.
\end{frame}

\begin{frame}[fragile]
  \frametitle{Borrows}
  \begin{onlyenv}<1>
\begin{minted}[texcomments=true]{OCaml}
let gradeDB = ... in
let avg =
  (Dbm.find &gradeDB "math" + Dbm.find &gradeDB "compsci") / 2
  (* \Y Multiple shared borrows *)
in
Dbm.add &!gradeDB "average" avg (* \Y One exclusive borrow *)
\end{minted}
\end{onlyenv}

% \begin{onlyenv}<2>
% \begin{minted}[texcomments=true]{OCaml}
% let gradeDB = ... in
% let avg =
%  let grade topic = Dbm.find &gradeDB topic in (* \Y Capture *)
%  (grade "math" + grade "compsci") / 2
% in
% Dbm.add &!gradeDB "average" avg (* \Y Exclusive borrow *)
% \end{minted}
% \end{onlyenv}

\begin{onlyenv}<2>
\begin{minted}[texcomments=true]{OCaml}
let gradeDB = ... in
f (&!gradeDB, &gradeDB) (* \N Conflicting borrows *)
\end{minted}
\end{onlyenv}
\begin{onlyenv}<3>
\begin{minted}[texcomments=true]{OCaml}
let gradeDB = ... in
f (gradeDB, &gradeDB) (* \N Use and borrow *)
\end{minted}
\end{onlyenv}
\end{frame}

\begin{frame}[fragile]{API}

\begin{minted}[texcomments=true]{OCaml}
type t : lin
val find : &(Dbm.t,'k) -> string -{'k}> int
val add : &!(Dbm.t,'k) -> string -{'k}> int -{'k}> unit
\end{minted}

\end{frame}

\begin{frame}[fragile]
  \frametitle{Captures}
  \begin{onlyenv}<1>
\begin{minted}[texcomments=true]{OCaml}
let gradeDB = ... in
(Dbm.find &gradeDB "math" + Dbm.find &gradeDB "compsci") / 2
\end{minted}
  \end{onlyenv}
  \begin{onlyenv}<2>
\begin{minted}[texcomments=true]{OCaml}
let gradeDB = ... in
let grade subject = Dbm.find &gradeDB subject in (* \Y Capture *)
(grade "math" + grade "compsci") / 2
\end{minted}
  \end{onlyenv}
  \begin{onlyenv}<3>
\begin{minted}[texcomments=true]{OCaml}
let gradeDB = ... in
let grade = Dbm.find &gradeDB in (* \Y Partial application *)
(grade "math" + grade "compsci") / 2
\end{minted}
  \end{onlyenv}
  \begin{onlyenv}<4->
\begin{minted}[texcomments=true]{OCaml}
(* \Y Easy functional abstraction *)
let average db subjects = 
  List.map (Dbm.find db) subjects / List.length subjects

let main () =
  let gradeDB = ... in
  average &gradeDB ["math";"compsci";...]
  ...
\end{minted}
  \end{onlyenv}
\end{frame}

\begin{frame}[fragile]{Region}

With explicit regions:

\begin{minted}[texcomments=true]{OCaml}
let average db subjects = 
  List.map (Dbm.find db) subjects / List.length subjects

let main () =
  let gradeDB = ... in
  let avg = {| average &gradeDB ["math";"compsci";...] |} in
  {| Dbm.add &!gradeDB "average" avg |}
\end{minted}

\end{frame}

\begin{frame}
  \frametitle{Region -- details}
  Explain the indexed types in slightly more detailed
\end{frame}

\begin{frame}
  \frametitle{High order functions}

  Talk about fancy iterators ?

\end{frame}

\begin{frame}
  \frametitle{Affe -- Recap}
  Recap all the nice properties we showed
\end{frame}

\begin{frame}
  \frametitle{Theory}
  Give the theoretical contributions of the paper
\end{frame}

\begin{frame}
  \frametitle{Conclusion}

\end{frame}

\bibliography{biblio}

\end{document}

%%% Local Variables:
%%% mode: latex
%%% TeX-command-extra-options: "-shell-escape"
%%% TeX-master: "slides-icfp"
%%% End:

### 2020-07-29 kindly/slides-icfp.tex
Start adding arrows.

@@
 
+\tikzset{
+  annot/.style={align=center,draw=black,scale=0.9,text depth=0pt,rounded corners=6pt, thick},
+  onslide/.code args={<#1>#2}{\only<#1>{\pgfkeysalso{#2}}},
+}
+\tikzstyle{link}=[->,>=latex, rounded corners=6pt, thick]
+\newcommand\tikzcoord[1]{%
+  \tikz[baseline,remember picture]{\node[coordinate] (#1) {};}}
+
 \def\HUGE{\fontsize{35pt}{15pt}\selectfont}
@@
   Chromium is written in C/C++!\\
-  Surely these bugs don't happen in high-level typed languages \dots
+  Surely these bugs don't happen in high-level typed languages\alt<3>{ \dots right?}{.}
 
@@
 \begin{minted}{OCaml}
-let db = Dbm.opendbm "mydb" [Dbm_rdwr] 0o666 in
+let db : database = Dbm.opendbm "mydb" [Dbm_rdwr] 0o666 in
 Dbm.add db "hello" "world";
@@
 
-\begin{minted}[texcomments=true]{OCaml}
-let db = Dbm.opendbm "firstnames" [Dbm_rdwr] 0o666 in
-Dbm.add &!db "Radanne" "Gabriel";
+\begin{minted}[texcomments=true,escapeinside=\#\#]{OCaml}
+let db : data#\tikzcoord{db}#base = Dbm.opendbm "mydb" [Dbm_rdwr] 0o666 in
+Dbm.add &!#\tikzcoord{borrow1}#db "hello" "world"#\tikzcoord{string}#;
 (* ... *)
@@
 (* ... *)
-print_string (Dbm.find &db "Saffrich") (* \N Compile time error! *)
+print_string (Dbm.find &#\tikzcoord{borrow2}#db "hello") (* \N Compile time error! *)
 \end{minted}
 
-  \TODO{Add lot's of annotation w. arrows/pointers to explain the code piece by piece}
-
+  \begin{tikzpicture}[remember picture,overlay]
+    \begin{onlyenv}<2->
+      \draw[overlay,link] ($(db)+(0,1em)$) to[out=90,in=180]
+      +(2,1) node[annot,anchor=west]
+      {@type database : lin (* Linear kind *)@};
+      \draw[overlay,link] ($(string)+(-1em,-0.2em)$) to[out=-30,in=180]
+      +(1.5,-0.1) node[annot,anchor=west]
+      {@type string : un (* Unrestricted kind *)@};
+    \end{onlyenv}
+    \begin{onlyenv}<3->
+      \node[annot] at ($(borrow1)+(1.8,-0.8)$) (borrows) {Borrows!};
+      \draw[overlay,link] ($(borrow1)-(0,0.2em)$) to[out=-60,in=120] (borrows) ;
+      \draw[overlay,link] ($(borrow2)+(0,1em)$) to[out=120,in=-60] (borrows) ;
+    \end{onlyenv}
+  \end{tikzpicture}
 \end{frame}

### 2020-07-30 kindly/slides-icfp.tex
Lot's of improvements (and arrows).

--- 2020-07-28 kindly/slides-icfp.tex
@@
 \usepackage{minted}
+\usepackage{wasysym}
 
@@
 \setbeamercovered{transparent=20}
-
-\usepackage{wasysym}
+\metroset{
+  sectionpage=none,
+  numbering=fraction,
+  progressbar=foot,
+}
 
@@
   shadings,shadows,arrows,decorations.markings,calc,fit,fadings,
-  tikzmark
+  tikzmark,scopes
 }
+\input{tikz}
 
@@
 \begin{minted}{OCaml}
-let db = Dbm.opendbm "mydb" [Dbm_rdwr] 0o666 in
-Dbm.add db "hello" "world";
+let db : database = Dbm.opendbm "mydb" [Dbm_rdwr] 0o666 in
+Dbm.add db "hello" 42;
 (* ... *)
@@
 (* ... *)
-print_string (Dbm.find db "hello") (* bug! *)
+print_int (Dbm.find db "hello") (* bug! *)
 \end{minted}
@@
 
-\begin{minted}[texcomments=true]{OCaml}
-let db = Dbm.opendbm "firstnames" [Dbm_rdwr] 0o666 in
-Dbm.add &!db "Radanne" "Gabriel";
+  \begin{onlyenv}<1-3>
+\begin{minted}[texcomments=true,escapeinside=\#\#]{OCaml}
+let db : #\tikzcoord{db1}#database#\tikzcoord{db2}# = Dbm.opendbm "mydb" [Dbm_rdwr] 0o666 in
+Dbm.add #\tikzcoord{A1}#&!db#\tikzcoord{A2}# #\tikzcoord{string1}#"hello"#\tikzcoord{string2}# 42;
 (* ... *)
@@
 (* ... *)
-print_string (Dbm.find &db "Saffrich") (* \N Compile time error! *)
+print_int (Dbm.find #\tikzcoord{U1}#&db#\tikzcoord{U2}# "world") (* \N Compile time error! *)
+\end{minted}
+  \end{onlyenv}
+  \begin{onlyenv}<4>
+\begin{minted}[texcomments=true,escapeinside=\#\#]{OCaml}
+let db #\tikzcoord{eq1}#=#\tikzcoord{eq2}# Dbm.opendbm "mydb" [Dbm_rdwr] 0o666 in
+Dbm.add &!db "hello" "world";
+(* ... *)
+Dbm.close db;
+(* ... *)
+print_string (Dbm.find &db "hello") (* \N Compile time error! *)
+\end{minted}
+  \end{onlyenv}
+
+  \begin{tikzpicture}[remember picture,overlay]
+    \begin{onlyenv}<2>
+      \node[code,fit=(db1) (db2)] (db) {};
+      \node[code,fit=(string1) (string2)] (string) {};
+      \node[overlay,text width=8.7cm,annot,scale=1] at (5,-0.7) {
+        \textbf{Kinds determine usage:}
+        \begin{itemize}
+        \item Linear (@lin@): Used exactly once $[1]$
+        \item Affine (@aff@): Used at most once $[0-1]$
+        \item Unrestricted (@un@): Used arbitrarily many time $[0-\infty]$
+        \end{itemize}
+      };
+      \draw[overlay,link] (db) to[out=60,in=180]
+      +(2,1) node[annot,anchor=west]
+      {@type database : lin (* Linear kind *)@};
+      \draw[overlay,link] (string) to[out=-30,in=180]
+      +(1.5,-0.5) node[annot,anchor=west]
+      {@type string : un (* Unrestricted kind *)@};
+    \end{onlyenv}
+    \begin{onlyenv}<3>
+      \node[code,fit=(A1) (A2)] (A) {};
+      \node[code,fit=(U1) (U2)] (U) {};
+      \node[annot,scale=1.2] at ($(A)+(1.8,-0.8)$) (borrows) {Borrows!};
+      \draw[overlay,link] (A) to[out=-30,in=150] (borrows) ;
+      \draw[overlay,link] (U) to[out=120,in=-60] (borrows) ;
+    \end{onlyenv}
+    \begin{onlyenv}<4>
+      \node[code,fit=(eq1) (eq2)] (eq) {};
+      \draw[overlay,link] (eq) to[out=90,in=-90]
+      +(0.5,1) node[annot,scale=1.2,anchor=south]
+      {Complete Type Inference};
+    \end{onlyenv}
+  \end{tikzpicture}
+\end{frame}
+
+\begin{frame}{Table of contents}
+  \setbeamertemplate{section in toc}[sections numbered]
+  \tableofcontents[hideallsubsections]
+\end{frame}
+
+\section{Linearity through kinds}
+
+\begin{frame}[fragile]
+  \frametitle{Linearity through kinds}
+
+  \textbf{Kinds determine usage:}
+  \begin{itemize}
+  \item Linear (@lin@): Used exactly once $[1]$
+  \item Affine (@aff@): Used at most once $[0-1]$
+  \item Unrestricted (@un@): Used arbitrarily many time $[0-\infty]$
+  \end{itemize}
+
+  \textbf{Examples:}
+  \begin{onlyenv}<1>
+\begin{minted}[texcomments=true]{OCaml}
+type database : lin
+type string : un
+\end{minted}
+  \end{onlyenv}
+  \begin{onlyenv}<2>
+\begin{minted}[texcomments=true]{OCaml}
+type ('a : 'k) list : 'k
+\end{minted}
+  \end{onlyenv}
+  \begin{onlyenv}<3>
+\begin{minted}[texcomments=true]{OCaml}
+type ('a : 'k) list : 'k
+val create_list : ('a : un) => int -> 'a -> 'a list
 \end{minted}
 
-  \TODO{Add lot's of annotation w. arrows/pointers to explain the code piece by piece}
-
+    We also use \textbf{kind constraints} and \textbf{subkinding}\\
+    $\Rightarrow$ Type signatures are still reasonably simple!
+    \TODO{More on subkinding ?}
+  \end{onlyenv}
 \end{frame}
 
-\begin{frame}[fragile]{A recap on borrows}
+\section{Functions and captures}
 
-  \TODO{C/C, to refine}
+\begin{frame}[fragile]
+  \frametitle{Functions and captures}
+
+  What about functions ? We can capture: 
+\begin{minted}[texcomments=true,escapeinside=\#\#]{OCaml}
+let db = Dbm.open ...
+
+let #\tikzcoord{fun1}#conclude#\tikzcoord{fun2}# x = 
+  Dbm.add &db "conclude" x;
+  Dbm.close db
+\end{minted}
+  \begin{onlyenv}<3->
+    We infer the type:
+\begin{minted}[texcomments=true,escapeinside=\#\#]{OCaml}
+val oups : int #\tikzcoord{arr1}#-{lin}>#\tikzcoord{arr2}# unit
+\end{minted}
+  \end{onlyenv}
+
+  \begin{tikzpicture}[remember picture,overlay]
+    \begin{onlyenv}<2>
+      \node[code,fit=(fun1) (fun2)] (fun) {};
+      \draw[overlay,link] (fun) to[out=30,in=180]
+      +(3,0.5) node[annot,scale=1,anchor=west] {
+        This function must behave \textbf{linearly}
+      };
+    \end{onlyenv}
+    \begin{onlyenv}<3->
+      \node[code,fit=(arr1) (arr2)] (arr) {};
+      \draw[overlay,link] (arr) to[out=-30,in=180]
+      +(2,-0.5) node[annot,scale=1,anchor=west] {
+        Arrows are annotated with their usage mode
+      };
+      \node<4>[annot,scale=1.2,red] at ($(arr)+(5,1)$) {
+        \textbf{Warning}: This does not say anything about the arguments!!
+      };
+    \end{onlyenv}
+  \end{tikzpicture}
+\end{frame}
+
+\section{Borrows and regions}
+
+\begin{frame}[fragile]{Borrows}
+
+\begin{minted}[texcomments=true,escapeinside=\#\#]{OCaml}
+let gradeDB = ... in
+let grade = Dbm.find #\tikzcoord{U1}#&gradeDB#\tikzcoord{U2}# "compsci" in
+Dbm.add #\tikzcoord{A1}#&!gradeDB#\tikzcoord{A2}# "best" grade;
+Dbm.close db
+\end{minted}
 
   Borrows allow to lend usage of something to someone else.
-
-  There are different types of borrows:
   \begin{itemize}
-  \item Shared borrows @&a@ which are \emph{Unrestricted} (@un@)
-  \item Exclusive borrows @&!a@ which are \emph{Affine} (@aff@)
+  \item \textbf{Shared} borrows \tikzcoord{U3}@&a@\tikzcoord{U4} which are Unrestricted (@un@)
+  \item \textbf{Exclusive} borrows \tikzcoord{A3}@&!a@\tikzcoord{A4} which are Affine (@aff@)
   \end{itemize}
 
-  We cannot use a borrow of @a@ and @a@ itself at the same time.\\
-  A borrow must not escape.
+  \visible<2>{
+    \textbf{There are rules to respect!}
+  }
+
+  % \textbf{Rules:}
+  % \begin{itemize}
+  % \item Cannot use a borrow of @a@ and @a@ itself at the same time.
+  % \item Only one exclusive borrows at a time
+  % \item Borrows must not escape
+  % \end{itemize}
+
+  \begin{tikzpicture}[remember picture,overlay]
+    \node[code,color=yellow,fit=(A1) (A2)] {};
+    \node[code,color=red,fit=(U1) (U2)] {};
+    \node[code,color=yellow,fit=(A3) (A4)] {};
+    \node[code,color=red,fit=(U3) (U4)] {};
+  \end{tikzpicture}
 \end{frame}
@@
 \begin{frame}[fragile]
-  \frametitle{Borrows}
+  \frametitle{Borrows -- Example of uses}
   \begin{onlyenv}<1>
+    Good use of borrows:
 \begin{minted}[texcomments=true]{OCaml}
@@
 \begin{onlyenv}<2>
+  \textbf{Rule 1}: Cannot use a borrow and the thing itself simultaneously
 \begin{minted}[texcomments=true]{OCaml}
 let gradeDB = ... in
-f (&!gradeDB, &gradeDB) (* \N Conflicting borrows *)
+f (gradeDB, &gradeDB) (* \N Conflicting use and borrow! *)
 \end{minted}
@@
 \begin{onlyenv}<3>
+  \textbf{Rule 2}: Cannot use an exclusive borrow and any other borrow simultaneously
 \begin{minted}[texcomments=true]{OCaml}
 let gradeDB = ... in
-f (gradeDB, &gradeDB) (* \N Use and borrow *)
+f (&!gradeDB, &gradeDB) (* \N Conflicting borrows! *)
+\end{minted}
+\end{onlyenv}
+\begin{onlyenv}<4>
+  \textbf{Rule 3}: Borrows must not escape
+\begin{minted}[texcomments=true]{OCaml}
+let f () = 
+  let gradeDB = ... in
+  let x = (&gradeDb, "mygrades") in
+  x (* \N Borrow escaping its scope! *)
 \end{minted}
@@
 
-\begin{frame}[fragile]{API}
-
-\begin{minted}[texcomments=true]{OCaml}
-type t : lin
-val find : &(Dbm.t,'k) -> string -{'k}> int
-val add : &!(Dbm.t,'k) -> string -{'k}> int -{'k}> unit
-\end{minted}
-
+\begin{frame}
+  \frametitle{Regions/Scoping}
+  \TODO{Very quick example of region/scoping}
 \end{frame}
 
+% \begin{frame}[fragile]{API}
+
+%   {\Large The \texttt{Dbm} API:}
+% \begin{minted}[texcomments=true]{OCaml}
+% type database : lin
+% val find : &(database,'k) -> string -{'k}> int
+% val add : &!(database,'k) -> string -{'k}> int -{'k}> unit
+% \end{minted}
+
+% \end{frame}
+
 \begin{frame}[fragile]
-  \frametitle{Captures}
+  \frametitle{Everything together}
+  \TODO{Make that straighter/punchier}
+
   \begin{onlyenv}<1>
@@
   let gradeDB = ... in
-  average &gradeDB ["math";"compsci";...]
-  ...
+  let avg = average &gradeDB ["math";"compsci";...] in
+  Dbm.add &!gradeDB "average" avg
 \end{minted}
@@
 
-\begin{frame}[fragile]{Region}
+\section{Inference and constraints}
 
-With explicit regions:
+\begin{frame}[fragile]{Inference and constraints}
+
+\TODO{Show the global step-by-step process}
 
@@
 
-\begin{frame}
-  \frametitle{Region -- details}
-  Explain the indexed types in slightly more detailed
-\end{frame}
-
-\begin{frame}
-  \frametitle{High order functions}
-
-  Talk about fancy iterators ?
-
-\end{frame}
+\section*{Conclusion}
 

### 2020-07-31 kindly/slides-icfp.tex
smooth out the first half of the talk.

@@
 \begin{minted}{OCaml}
-let db : database = Dbm.opendbm "mydb" [Dbm_rdwr] 0o666 in
-Dbm.add db "hello" 42;
+let gradeDB : database = Dbm.opendbm "gradeDB" ... in
+Dbm.add gradeDB "math" 42;
 (* ... *)
-Dbm.close db;
+Dbm.close gradeDB;
 (* ... *)
-print_int (Dbm.find db "hello") (* bug! *)
+print_int (Dbm.find gradeDB "literature") (* bug! *)
 \end{minted}
@@
 \begin{minted}[texcomments=true,escapeinside=\#\#]{OCaml}
-let db : #\tikzcoord{db1}#database#\tikzcoord{db2}# = Dbm.opendbm "mydb" [Dbm_rdwr] 0o666 in
-Dbm.add #\tikzcoord{A1}#&!db#\tikzcoord{A2}# #\tikzcoord{string1}#"hello"#\tikzcoord{string2}# 42;
+let gradeDB : #\tikzcoord{db1}#database#\tikzcoord{db2}# = Dbm.opendbm "gradeDB" ... in
+Dbm.add #\tikzcoord{A1}#&!gradeDB#\tikzcoord{A2}# #\tikzcoord{string1}#"math"#\tikzcoord{string2}# 42;
 (* ... *)
-Dbm.close db;
+Dbm.close gradeDB;
 (* ... *)
-print_int (Dbm.find #\tikzcoord{U1}#&db#\tikzcoord{U2}# "world") (* \N Compile time error! *)
+print_int (Dbm.find #\tikzcoord{U1}#&gradeDB#\tikzcoord{U2}# "literature") (* \N Compile time error! *)
 \end{minted}
@@
 \begin{minted}[texcomments=true,escapeinside=\#\#]{OCaml}
-let db #\tikzcoord{eq1}#=#\tikzcoord{eq2}# Dbm.opendbm "mydb" [Dbm_rdwr] 0o666 in
-Dbm.add &!db "hello" "world";
+let gradeDB #\tikzcoord{eq1}#=#\tikzcoord{eq2}# Dbm.opendbm "gradeDB" ... in
+Dbm.add &!gradeDB "math" 42;
 (* ... *)
-Dbm.close db;
+Dbm.close gradeDB;
 (* ... *)
-print_string (Dbm.find &db "hello") (* \N Compile time error! *)
+print_string (Dbm.find &gradeDB "literature") (* \N Compile time error! *)
 \end{minted}
@@
       \node[code,fit=(string1) (string2)] (string) {};
-      \node[overlay,text width=8.7cm,annot,scale=1] at (5,-0.7) {
-        \textbf{Kinds determine usage:}
-        \begin{itemize}
-        \item Linear (@lin@): Used exactly once $[1]$
-        \item Affine (@aff@): Used at most once $[0-1]$
-        \item Unrestricted (@un@): Used arbitrarily many time $[0-\infty]$
-        \end{itemize}
+      \node[overlay% ,text width=8.7cm
+      ,annot,scale=1.2] at (5,-0.7) {
+        Kinds determine usage
+        % \begin{itemize}
+        % \item Linear (@lin@): Used exactly once $[1]$
+        % \item Affine (@aff@): Used at most once $[0-1]$
+        % \item Unrestricted (@un@): Used arbitrarily many time $[0-\infty]$
+        % \end{itemize}
       };
@@
       \node[code,fit=(U1) (U2)] (U) {};
-      \node[annot,scale=1.2] at ($(A)+(1.8,-0.8)$) (borrows) {Borrows!};
-      \draw[overlay,link] (A) to[out=-30,in=150] (borrows) ;
-      \draw[overlay,link] (U) to[out=120,in=-60] (borrows) ;
+      \node[annot,scale=1.2] at ($(A)+(2,-0.8)$) (borrows) {Borrows!};
+      \draw[overlay,link,<-] (A) to[out=-30,in=150] (borrows) ;
+      \draw[overlay,link,<-] (U) to[out=120,in=-70] (borrows) ;
     \end{onlyenv}
@@
       \node[code,fit=(eq1) (eq2)] (eq) {};
-      \draw[overlay,link] (eq) to[out=90,in=-90]
+      \draw[overlay,link,<-] (eq) to[out=90,in=-90]
       +(0.5,1) node[annot,scale=1.2,anchor=south]
@@
 
-\begin{frame}[fragile]
+\begin{frame}[t,fragile]
   \frametitle{Functions and captures}
@@
 \begin{minted}[texcomments=true,escapeinside=\#\#]{OCaml}
-let db = Dbm.open ...
+let gradeDB = Dbm.open ...
 
-let #\tikzcoord{fun1}#conclude#\tikzcoord{fun2}# x = 
-  Dbm.add &db "conclude" x;
-  Dbm.close db
+let #\tikzcoord{fun1}#log_n_close#\tikzcoord{fun2}# msg = 
+  printf "Closing: %s" msg;
+  Dbm.close gradeDB
 \end{minted}
@@
 \begin{minted}[texcomments=true,escapeinside=\#\#]{OCaml}
-val oups : int #\tikzcoord{arr1}#-{lin}>#\tikzcoord{arr2}# unit
+val conclude : string #\tikzcoord{arr1}#-{lin}>#\tikzcoord{arr2}# unit
 \end{minted}
@@
       \node[code,fit=(fun1) (fun2)] (fun) {};
-      \draw[overlay,link] (fun) to[out=30,in=180]
+      \draw[overlay,link,<-] (fun) to[out=30,in=180]
       +(3,0.5) node[annot,scale=1,anchor=west] {
-        This function must behave \textbf{linearly}
+        This function \textbf{must} be used exactly once
       };
@@
       \node[code,fit=(arr1) (arr2)] (arr) {};
-      \draw[overlay,link] (arr) to[out=-30,in=180]
-      +(2,-0.5) node[annot,scale=1,anchor=west] {
+      \draw[overlay,link,<-] (arr) to[out=-110,in=70]
+      +(-1,-1) node[annot,scale=1,anchor=north] {
         Arrows are annotated with their usage mode
       };
-      \node<4>[annot,scale=1.2,red] at ($(arr)+(5,1)$) {
+      \node<4>[annot,scale=1.2,red] at ($(arr)+(4,1)$) {
         \textbf{Warning}: This does not say anything about the arguments!!
@@
 
-\begin{frame}[fragile]{Borrows}
-
-\begin{minted}[texcomments=true,escapeinside=\#\#]{OCaml}
-let gradeDB = ... in
-let grade = Dbm.find #\tikzcoord{U1}#&gradeDB#\tikzcoord{U2}# "compsci" in
-Dbm.add #\tikzcoord{A1}#&!gradeDB#\tikzcoord{A2}# "best" grade;
-Dbm.close db
-\end{minted}
+\begin{frame}[t,fragile]{Borrows}
 
@@
   \begin{itemize}
-  \item \textbf{Shared} borrows \tikzcoord{U3}@&a@\tikzcoord{U4} which are Unrestricted (@un@)
-  \item \textbf{Exclusive} borrows \tikzcoord{A3}@&!a@\tikzcoord{A4} which are Affine (@aff@)
+  \item \textbf{Shared} borrows \tikzcoord{UX1}@&a@\tikzcoord{UX2} which are Unrestricted (@un@)
+  \item \textbf{Exclusive} borrows \tikzcoord{AX1}@&!a@\tikzcoord{AX2} which are Affine (@aff@)
   \end{itemize}
 
-  \visible<2>{
-    \textbf{There are rules to respect!}
-  }
-
-  % \textbf{Rules:}
-  % \begin{itemize}
-  % \item Cannot use a borrow of @a@ and @a@ itself at the same time.
-  % \item Only one exclusive borrows at a time
-  % \item Borrows must not escape
-  % \end{itemize}
+  \begin{onlyenv}<2->
+  A correct usage of borrows:
+\begin{minted}[texcomments=true,escapeinside=\#\#]{OCaml}
+let avg =
+  (Dbm.find #\tikzcoord{U1}#&gradeDB#\tikzcoord{U2}# "math" + Dbm.find #\tikzcoord{U3}#&gradeDB#\tikzcoord{U4}# "compsci") / 2
+  (* \Y Multiple shared borrows *)
+in
+Dbm.add #\tikzcoord{A1}#&!gradeDB#\tikzcoord{A2}# "average" avg (* \Y One exclusive borrow *)
+\end{minted}
+  \end{onlyenv}
+  \visible<3>{There are rules to respect!}
 
   \begin{tikzpicture}[remember picture,overlay]
+    \begin{onlyenv}<2->
       \node[code,color=yellow,fit=(A1) (A2)] {};
       \node[code,color=red,fit=(U1) (U2)] {};
-    \node[code,color=yellow,fit=(A3) (A4)] {};
       \node[code,color=red,fit=(U3) (U4)] {};
+    \end{onlyenv}
+    \node[code,color=yellow,fit=(AX1) (AX2)] {};
+    \node[code,color=red,fit=(UX1) (UX2)] {};
   \end{tikzpicture}
@@
   \frametitle{Borrows -- Example of uses}
-  \begin{onlyenv}<1>
-    Good use of borrows:
-\begin{minted}[texcomments=true]{OCaml}
-let gradeDB = ... in
-let avg =
-  (Dbm.find &gradeDB "math" + Dbm.find &gradeDB "compsci") / 2
-  (* \Y Multiple shared borrows *)
-in
-Dbm.add &!gradeDB "average" avg (* \Y One exclusive borrow *)
-\end{minted}
-\end{onlyenv}
 
@@
 % \end{onlyenv}
-
-\begin{onlyenv}<2>
+\begin{onlyenv}<1>
   \textbf{Rule 1}: Cannot use a borrow and the thing itself simultaneously
@@
 \end{minted}
-\end{onlyenv}
-\begin{onlyenv}<3>
+\end{onlyenv}%
+\begin{onlyenv}<2>
   \textbf{Rule 2}: Cannot use an exclusive borrow and any other borrow simultaneously
@@
 \end{minted}
+\end{onlyenv}%
+\begin{onlyenv}<3->
+  \textbf{Rule 3}: Borrows must not escape
+  \begin{onlyenv}<3>
+\begin{minted}[texcomments=true,escapeinside=\#\#]{OCaml}
+let f () = 
+  let gradeDB = ... in
+  #\tikzcoord{R1}#let x = (&gradeDb, "mygrades") in#\tikzcoord{R2}#
+  #\tikzcoord{R3}#x#\tikzcoord{R4}#
+  (* \N Borrow escaping its scope! *)
+\end{minted}
   \end{onlyenv}
   \begin{onlyenv}<4>
-  \textbf{Rule 3}: Borrows must not escape
-\begin{minted}[texcomments=true]{OCaml}
+\begin{minted}[texcomments=true,escapeinside=\#\#]{OCaml}
 let f () = 
   let gradeDB = ... in
-  let x = (&gradeDb, "mygrades") in
-  x (* \N Borrow escaping its scope! *)
+  #\tikzcoord{R1}#{| let x = (&gradeDb, "mygrades") in#\tikzcoord{R2}#
+  #\tikzcoord{R3}#x |}#\tikzcoord{R4}#
+  (* \N Borrow escaping its scope! *)
 \end{minted}
   \end{onlyenv}
-\end{frame}
-
-\begin{frame}
-  \frametitle{Regions/Scoping}
-  \TODO{Very quick example of region/scoping}
+  \begin{tikzpicture}[remember picture,overlay]%
+    \begin{onlyenv}<4>
+      \node[code,rounded corners=0,color=green,fit=(R1) (R2)] {};
+      \node[code,rounded corners=0,minimum height=15pt,yshift=1pt,color=green,fit=(R3) (R4)] {};
+    \end{onlyenv}
+  \end{tikzpicture}%
+\end{onlyenv}%
 \end{frame}

### 2020-08-03 kindly/slides-icfp.tex
Cleanup the "everything" slide.

@@
     \begin{onlyenv}<4>
-      \node[code,rounded corners=0,color=green,fit=(R1) (R2)] {};
-      \node[code,rounded corners=0,minimum height=15pt,yshift=1pt,color=green,fit=(R3) (R4)] {};
+      \node[code,rounded corners=0,color=red,fit=(R1) (R2)] {};
+      \node[code,rounded corners=0,minimum height=15pt,yshift=1pt,color=red,fit=(R3) (R4)] {};
     \end{onlyenv}
@@
 
+  % \begin{tikzpicture}[remember picture,overlay]
+  %   \begin{onlyenv}<2->
+  %     \node[code,color=yellow,fit=(A1) (A2)] {};
+  %     \node[code,color=red,fit=(U1) (U2)] {};
+  %     \node[code,color=red,fit=(U3) (U4)] {};
+  %   \end{onlyenv}
+  %   \node[code,color=yellow,fit=(AX1) (AX2)] {};
+  %   \node[code,color=red,fit=(UX1) (UX2)] {};
+  % \end{tikzpicture}
+
 \begin{frame}[fragile]
   \frametitle{Everything together}
-  \TODO{Make that straighter/punchier}
 
@@
 let gradeDB = ... in
+let avg =
   (Dbm.find &gradeDB "math" + Dbm.find &gradeDB "compsci") / 2
+
+in
+Dbm.add &!gradeDB "average" avg
 \end{minted}
@@
 let gradeDB = ... in
+let avg =
   let grade subject = Dbm.find &gradeDB subject in (* \Y Capture *)
   (grade "math" + grade "compsci") / 2
+in
+Dbm.add &!gradeDB "average" avg
 \end{minted}
@@
 let gradeDB = ... in
+let avg =
   let grade = Dbm.find &gradeDB in (* \Y Partial application *)
   (grade "math" + grade "compsci") / 2
+in
+Dbm.add &!gradeDB "average" avg
 \end{minted}
@@
   \begin{onlyenv}<4->
+    \begin{onlyenv}<4>%
 \begin{minted}[texcomments=true]{OCaml}
-(* \Y Easy functional abstraction *)
 let average db subjects = 
@@
 \end{minted}
+    \end{onlyenv}%
+    \begin{onlyenv}<5>%
+\begin{minted}[texcomments=true,escapeinside=\#\#]{OCaml}
+let average db subjects #\tikzcoord{eq1}#=#\tikzcoord{eq2}#
+  List.map (Dbm.find db) subjects / List.length subjects
+
+let main () =
+  let gradeDB = ... in
+  let avg = #\tikzcoord{R1}#{| average &gradeDB ["math";"compsci";...] |}#\tikzcoord{R2}# in
+  #\tikzcoord{R3}#{| Dbm.add &!gradeDB "average" avg |}#\tikzcoord{R4}#
+\end{minted}
+    \end{onlyenv}%
+  \end{onlyenv}%
+  \begin{tikzpicture}[remember picture,overlay]%
+    \begin{onlyenv}<5>
+      \node[code,fit=(eq1) (eq2)] (eq) {};
+      \draw[overlay,link,<-] (eq) to[out=90,in=-90]
+      +(0.5,1) node[annot,scale=1.2,anchor=south]
+      {No type annotation};
+      \node[code,rounded corners=0,color=red,fit=(R1) (R2)] (RA) {};
+      \node[code,rounded corners=0,color=yellow,fit=(R3) (R4)] (RB) {};
+      \node[annot,scale=1.2] at ($(R4)+(1,-1)$) (AR) {Disjoint regions};
+      \draw[link] (AR) to[out=90,in=-40] (RA.-8);
+      \draw[link] (AR) to[out=180,in=-40] (RB.-8);
     \end{onlyenv}
+  \end{tikzpicture}%
 \end{frame}

### 2020-08-03 kindly/slides-icfp.tex
Add notes on the region slide.

@@
   \end{onlyenv}
-  \begin{onlyenv}<4>
+  \begin{onlyenv}<4->
 \begin{minted}[texcomments=true,escapeinside=\#\#]{OCaml}
@@
   let gradeDB = ... in
-  #\tikzcoord{R1}#{| let x = (&gradeDb, "mygrades") in#\tikzcoord{R2}#
+  #\tikzcoord{R1}#{| let x = (#\tikzcoord{B1}#&gradeDb#\tikzcoord{B2}#, "mygrades") in#\tikzcoord{R2}#
   #\tikzcoord{R3}#x |}#\tikzcoord{R4}#
@@
 \end{minted}
-  \end{onlyenv}
+  \end{onlyenv}%
   \begin{tikzpicture}[remember picture,overlay]%
-    \begin{onlyenv}<4>
+    \begin{onlyenv}<4->
       \node[code,rounded corners=0,color=red,fit=(R1) (R2)] {};
       \node[code,rounded corners=0,minimum height=15pt,yshift=1pt,color=red,fit=(R3) (R4)] {};
+      \node[code,opacity=0,fit=(B1) (B2)] (B) {};
+    \end{onlyenv}
+    \begin{onlyenv}<5->
+      \draw[link,<-] (B.10) to[in=180,out=40] ++(2,1) node[annot, anchor=west] {
+        @&gradeDb : &(database, un_2)@
+      };
+      \draw[link,<-] (R2) to[in=90,out=-10] ++(1.5,-0.5) node[annot, anchor=north] {
+        Region nesting level: 1
+      };
     \end{onlyenv}
   \end{tikzpicture}%
+  \visible<5>{%
+    We use \textbf{indexed} kinds to check that borrows do not escape!
+
+    Borrows of index $2$ cannot escape a region of index $1$.
+  }
 \end{onlyenv}%

### 2020-08-04 kindly/slides-icfp.tex
First version of the schema

@@
 
-\begin{frame}[fragile]{Inference and constraints}
+\begin{frame}{Inference and constraints}
 
-\TODO{Show the global step-by-step process}
+  Inference proceeds in the following steps:
 
+  \begin{enumerate}
+  \item Infer where regions goes
+  \item Turn the type, linearity and escape checks into constraints
+  \item Solve the constraints
+  \item Simplify the resulting types
+  \end{enumerate}
+\end{frame}
+
+% \begin{frame}[fragile]
+%   \frametitle{Inference in action}
+%   \begin{onlyenv}<1>%
+% \begin{minted}[texcomments=true]{OCaml}
+% let average db subjects = 
+%   List.map (Dbm.find db) subjects / List.length subjects
+
+% let main () =
+%   let gradeDB = ... in
+%   let avg = average &gradeDB ["math";"compsci";...] in
+%   Dbm.add &!gradeDB "average" avg
+% \end{minted}
+%   \end{onlyenv}%
+%   \begin{onlyenv}<2>%
+% \begin{minted}[texcomments=true,escapeinside=\#\#]{OCaml}
+% let average db subjects =
+%   List.map (Dbm.find db) subjects / List.length subjects
+
+% let main () =
+%   let gradeDB = ... in
+%   let avg = #\tikzcoord{R1}#{| average &gradeDB ["math";"compsci";...] |}#\tikzcoord{R2}# in
+%   #\tikzcoord{R3}#{| Dbm.add &!gradeDB "average" avg |}#\tikzcoord{R4}#
+% \end{minted}
+%   \end{onlyenv}%
+%   \begin{tikzpicture}[remember picture,overlay]%
+%     \begin{onlyenv}<2>
+%       \node[code,rounded corners=0,color=red,fit=(R1) (R2)] (RA) {};
+%       \node[code,rounded corners=0,color=yellow,fit=(R3) (R4)] (RB) {};
+%     \end{onlyenv}
+%   \end{tikzpicture}%
+% \end{frame}
+
+\begin{frame}[fragile,shrink]
+  \frametitle{Inference in action}
+  \begin{center}
+    \begin{tikzpicture}
+      \node[annot,text width=11.5cm,scale=0.5] at (180:5) (S1) {%
 \begin{minted}[texcomments=true]{OCaml}
@@
   let gradeDB = ... in
-  let avg = {| average &gradeDB ["math";"compsci";...] |} in
-  {| Dbm.add &!gradeDB "average" avg |}
+  let avg = average &gradeDB ["math";"compsci";...] in
+  Dbm.add &!gradeDB "average" avg
 \end{minted}
+      };
+
+      \node[annot,text width=12cm,scale=0.5] at (108:3) (S2) {%
+\begin{minted}[texcomments=true,escapeinside=\#\#]{OCaml}
+let average db subjects =
+  List.map (Dbm.find db) subjects / List.length subjects
+
+let main () =
+  let gradeDB = ... in
+  let avg = #\tikzcoord{R1}#{| average &gradeDB ["math";"compsci";...] |}#\tikzcoord{R2}# in
+  #\tikzcoord{R3}#{| Dbm.add &!gradeDB "average" avg |}#\tikzcoord{R4}#
+\end{minted}
+        \tikz[remember picture,overlay]{
+          \node[code,rounded corners=0,color=red,fit=(R1) (R2)] (RA) {};
+          \node[code,rounded corners=0,color=yellow,fit=(R3) (R4)] (RB) {};
+        }%
+      };
+
+      \node[annot] at (36:4) (S3) {
+        CONTRAINTES
+      };
+
+      \node[annot] at (-36:4) (S4) {
+        SCHEMES
+      };
+
+      \node[annot] at (-108:3) (S5) {
+        SIMPLIFIED SCHEMES
+      };
+
+      \draw[link,->] (S1) to[bend left] node[midway] {Infer regions} (S2) ;
+      \draw[link,->] (S2) to[bend left] node[midway] {Convert into constraints} (S3);
+      \draw[link,->] (S3) to[bend left] node[midway] {Solve constraints} (S4);
+      \draw[link,->] (S4) to[bend left] node[midway] {Simplify types} (S5);
+    \end{tikzpicture}
+  \end{center}
 
 \end{frame}
-
 \section*{Conclusion}

### 2020-08-05 kindly/slides-icfp.tex
Last chunk of the talk.

@@
 
-  Recent classification (2015-2020) of \og high severity security bugs\fg in Chromium:
+  Recent classification (2015-2020) of\\ \og high severity security bugs\fg in Chromium:
 
@@
     \centering
-    \includegraphics[width=0.55\textwidth]
+    \includegraphics[width=0.5\textwidth]
     {chromium-use-after-free}
@@
 
-  Existing ways to solve this
+  \TODO{Existing ways to solve this}
 
@@
       };
-      \draw[overlay,link] (db) to[out=60,in=180]
-      +(2,1) node[annot,anchor=west]
+      \draw[overlay,link] (db) to[out=60,in=-90]
+      +(0,1) node[annot,anchor=south]
       {@type database : lin (* Linear kind *)@};
@@
 \begin{minted}[texcomments=true,escapeinside=\#\#]{OCaml}
-val conclude : string #\tikzcoord{arr1}#-{lin}>#\tikzcoord{arr2}# unit
+val conclude : string #\tikzcoord{arr1}##$\tarr{\texttt{lin}}$##\tikzcoord{arr2}# unit
 \end{minted}
@@
       \draw[overlay,link,<-] (fun) to[out=30,in=180]
-      +(3,0.5) node[annot,scale=1,anchor=west] {
-        This function \textbf{must} be used exactly once
+      +(3,0.2) node[annot,scale=1,anchor=west,text width=3.2cm] {
+        This function \textbf{must}\\ be used exactly once
       };
@@
     \begin{onlyenv}<3->
-      \node[code,fit=(arr1) (arr2)] (arr) {};
+      \node[code,minimum height=18pt,fit=(arr1) (arr2)] (arr) {};
       \draw[overlay,link,<-] (arr) to[out=-110,in=70]
@@
     \begin{onlyenv}<5->
-      \draw[link,<-] (B.10) to[in=180,out=40] ++(2,1) node[annot, anchor=west] {
+      \draw[link,<-] (B.10) to[in=180,out=40] ++(0.5,0.5) node[annot, anchor=west] {
         @&gradeDb : &(database, un_2)@
@@
 
-  \begin{onlyenv}<1>
+  The @Database@ API:
+\begin{minted}[texcomments=true,escapeinside=\#\#]{OCaml}
+type database : lin
+val find : &(database,'k) -> string #$\tarr{\texttt{'k}}$# int
+val add : &!(database,'k) -> string #$\tarr{\texttt{'k}}$# int #$\tarr{\texttt{'k}}$# unit
+\end{minted}
+
+  \begin{onlyenv}<2->
+    A simple use:
+  \end{onlyenv}
+  \begin{onlyenv}<2>
 \begin{minted}[texcomments=true]{OCaml}
@@
   \end{onlyenv}
-  \begin{onlyenv}<2>
+  \begin{onlyenv}<3>
 \begin{minted}[texcomments=true]{OCaml}
@@
   \end{onlyenv}
-  \begin{onlyenv}<3>
+  \begin{onlyenv}<4>
 \begin{minted}[texcomments=true]{OCaml}
@@
   \end{onlyenv}
-  \begin{onlyenv}<4->
-    \begin{onlyenv}<4>%
+  \begin{onlyenv}<5->
+    \begin{onlyenv}<5>%
 \begin{minted}[texcomments=true]{OCaml}
@@
   List.map (Dbm.find db) subjects / List.length subjects
-
 let main () =
@@
     \end{onlyenv}%
-    \begin{onlyenv}<5>%
+    \begin{onlyenv}<6>%
 \begin{minted}[texcomments=true,escapeinside=\#\#]{OCaml}
@@
   \begin{tikzpicture}[remember picture,overlay]%
-    \begin{onlyenv}<5>
+    \begin{onlyenv}<6>
       \node[code,fit=(eq1) (eq2)] (eq) {};
-      \draw[overlay,link,<-] (eq) to[out=90,in=-90]
-      +(0.5,1) node[annot,scale=1.2,anchor=south]
+      \draw[overlay,link,<-] (eq) to[out=60,in=180]
+      +(0.5,0.5) node[annot,scale=1.2,anchor=west]
       {No type annotation};
@@
 
-\begin{frame}[fragile,shrink]
+\begin{frame}[fragile,t,shrink]
   \frametitle{Inference in action}
   \begin{center}
-    \begin{tikzpicture}
-      \node[annot,text width=11.5cm,scale=0.5] at (180:5) (S1) {%
+    \begin{tikzpicture}[xscale=1,yscale=0.8]
+      \node[annot,text width=11.5cm,scale=0.5] at (0,4) (S1) {%
 \begin{minted}[texcomments=true]{OCaml}
@@
 
-      \node[annot,text width=12cm,scale=0.5] at (108:3) (S2) {%
+      \node[annot,text width=12cm,scale=0.5] at (8,4) (S2) {%
 \begin{minted}[texcomments=true,escapeinside=\#\#]{OCaml}
@@
 
-      \node[annot] at (36:4) (S3) {
-        CONTRAINTES
+      \draw[link,->] (S1) to node[midway,above,sloped] {1. Infer regions} (S2) ;
+
+      \visible<2->{
+        \node[annot,text width=4.8cm] at (8,0) (S3) {%
+          \vspace{-1em}
+          \begin{align*}
+            \E = &(\tvar_f : \kvar_f)
+                 (\tvar_x : \kvar_x)\dots\\
+            C = &(\tvar_f \leq \gamma \tarr{\kvar_1} \beta )
+                \Cand
+                (\gamma \leq \tvar_x)\\
+                &\Cand
+                (\beta \times  \tvar_x \leq \alpha_r)
+                \Cand
+                  (\kvar_x \leq \kun)\\
+                 &\Cand \dots
+          \end{align*}
         };
+        \draw[link,->] (S2) to
+        node[midway,right,text width=3cm] {2. Convert\\ into constraints}
+        (S3);
+      }
 
-      \node[annot] at (-36:4) (S4) {
-        SCHEMES
+      \begin{onlyenv}<3->
+        \node[annot,text width=10.5cm,scale=0.7] at (0.7,1) (S4) {
+          @val average : 'k_2 . &(database,'k) -{'k_2}> string list -{'k}> int list@
         };
+        \draw[link,->] ($(S3.west)+(0,1)$) to
+        node[midway, above, sloped] {3. Solve}
+        node[midway, below, sloped] {constraints}
+        (S4.east);
+      \end{onlyenv}
 
-      \node[annot] at (-108:3) (S5) {
-        SIMPLIFIED SCHEMES
+      \begin{onlyenv}<4->
+        \node[annot,text width=9.7cm,scale=0.7] at (0.7,-1) (S5) {
+          @val average : &(database,'k) -> string list -{'k}> int list@
         };
+        \draw[link,->] (S4) to
+        node[midway, left] {4. Simplify}
+        node[midway, right] {types}
+        (S5);
+      \end{onlyenv}
 
-      \draw[link,->] (S1) to[bend left] node[midway] {Infer regions} (S2) ;
-      \draw[link,->] (S2) to[bend left] node[midway] {Convert into constraints} (S3);
-      \draw[link,->] (S3) to[bend left] node[midway] {Solve constraints} (S4);
-      \draw[link,->] (S4) to[bend left] node[midway] {Simplify types} (S5);
     \end{tikzpicture}
   \end{center}
+
+  \only<1>{%
+    1. Infer the placement of region based on the position of borrows and the borrowing rules.
+  }%
+  \only<2>{%
+    2. Convert into a custom constraint system based on HM(X).
+  }%
+  \only<3>{%
+    3. Solve the constraints using a custom algorithm and obtain principal type schemes.
+  }%
+  \only<4>{%
+    4. Simplify the obtained type scheme by leveraging subkinding
+  }%
 
@@
 \begin{frame}
-  \frametitle{Affe -- Recap}
-  Recap all the nice properties we showed
+  \frametitle{The \lang language -- Summary}
+
+  Prototype: \url{https://affe.netlify.com/}
+
+  \begin{itemize}
+  \item[\Y] Linearity, Closures, Borrows and Regions\\
+    $\Rightarrow$ Good support for both imperative and functional programming
+  \item[\Y] Support managed and unmanaged objects
+  \item[\Y] Principal type inference
+  \item[\N] No flow sensitivity
+  \item[\N] No concurrency story (yet)
+  \end{itemize}
+
+  Many examples in the paper: files, session types, semi-persistent arrays, iterators, connection pools, \dots
 \end{frame}
@@
 \begin{frame}
-  \frametitle{Theory}
-  Give the theoretical contributions of the paper
+  \frametitle{The theory -- Summary}
+  In the paper, the theoretical contributions:
+
+  \begin{itemize}
+  \item A Syntax-directed type system for \lang\\
+    $\Rightarrow$ How to encode borrows into an ML-style type-system
+  \item A formal semantics that takes allocations into account (+ proof of soundness)
+  \item An inference algorithm for \lang:
+    \begin{itemize}
+    \item An extension of HM(X) with kinds
+    \item A novel constraint systems to encode linearity and borrows
+    \item A constraint solving algorithm, and its proof of completeness
+    \end{itemize}
+  \end{itemize}
 \end{frame}
 
-\begin{frame}
-  \frametitle{Conclusion}
+\begin{frame}[standout]
+
+  \centering\Huge\texttt{Close(Talk)}
 

### 2020-08-06 kindly/slides-icfp.tex
Apply peter's remarks

@@
 \begin{frame}
-  \frametitle{Motivation}
-
-  Memory safety is the cause of numerous bugs and security issues.
-
-  Recent classification (2015-2020) of\\ \og high severity security bugs\fg in Chromium:
+  \frametitle{\og high severity security bugs\fg in Chromium}
 
@@
     \centering
-    \includegraphics[width=0.5\textwidth]
+    \includegraphics[width=0.7\textwidth]
     {chromium-use-after-free}
@@
 
-  \pause
+  \visible<2->{
     Chromium is written in C/C++!\\
-  Surely these bugs don't happen in high-level typed languages\alt<3>{ \dots right?}{.}
+    Surely these bugs don't happen in high-level typed languages%
+    \alt<3>{ \dots right?}{.}
+  }
 
@@
 (* ... *)
-print_int (Dbm.find gradeDB "literature") (* bug! *)
+print_int (Dbm.find gradeDB "literature") (* run-time error! *)
 \end{minted}
-
-\end{frame}
-
-\begin{frame}
-  \frametitle{Solutions?}
-
-  \TODO{Existing ways to solve this}
 
@@
 (* ... *)
-print_int (Dbm.find #\tikzcoord{U1}#&gradeDB#\tikzcoord{U2}# "literature") (* \N Compile time error! *)
+print_int (Dbm.find #\tikzcoord{U1}#&gradeDB#\tikzcoord{U2}# "literature") (* \N compile-time error! *)
 \end{minted}
@@
 (* ... *)
-print_string (Dbm.find &gradeDB "literature") (* \N Compile time error! *)
+print_string (Dbm.find &gradeDB "literature") (* \N compile-time error! *)
 \end{minted}
@@
   \end{onlyenv}
-  \begin{onlyenv}<3>
+  \begin{onlyenv}<3->
 \begin{minted}[texcomments=true]{OCaml}
@@
 
-    We also use \textbf{kind constraints} and \textbf{subkinding}\\
-    $\Rightarrow$ Type signatures are still reasonably simple!
-    \TODO{More on subkinding ?}
+    \begin{onlyenv}<4>
+      We also use \textbf{subkinding}: @un <= aff <= lin@
+    \end{onlyenv}
   \end{onlyenv}
@@
 
-  What about functions ? We can capture: 
 \begin{minted}[texcomments=true,escapeinside=\#\#]{OCaml}
@@
   printf "Closing: %s" msg;
-  Dbm.close gradeDB
+  Dbm.close #\tikzcoord{db1}#gradeDB#\tikzcoord{db2}#
 \end{minted}
@@
 \begin{minted}[texcomments=true,escapeinside=\#\#]{OCaml}
-val conclude : string #\tikzcoord{arr1}##$\tarr{\texttt{lin}}$##\tikzcoord{arr2}# unit
+val log_n_close : string #\tikzcoord{arr1}##$\tarr{\texttt{lin}}$##\tikzcoord{arr2}# unit
 \end{minted}
@@
   \begin{tikzpicture}[remember picture,overlay]
-    \begin{onlyenv}<2>
+    \begin{onlyenv}<2->
       \node[code,fit=(fun1) (fun2)] (fun) {};
-      \draw[overlay,link,<-] (fun) to[out=30,in=180]
-      +(3,0.2) node[annot,scale=1,anchor=west,text width=3.2cm] {
-        This function \textbf{must}\\ be used exactly once
+      \node[code,fit=(db1) (db2)] (db) {};
+      \node[annot,scale=1.2]
+      at ($(fun)+(5,0.2)$) (capture) {
+        Capture!
       };
+      \draw[overlay,link] (fun) to[out=20,in=170] (capture) ;
+      \draw[overlay,link] (capture) to[out=-90,in=0] (db) ;
     \end{onlyenv}
@@
       \draw[overlay,link,<-] (arr) to[out=-110,in=70]
-      +(-1,-1) node[annot,scale=1,anchor=north] {
-        Arrows are annotated with their usage mode
+      +(-1,-1) node[annot,scale=1.2,anchor=north] {
+        Usage mode
       };
-      \node<4>[annot,scale=1.2,red] at ($(arr)+(4,1)$) {
-        \textbf{Warning}: This does not say anything about the arguments!!
+    \end{onlyenv}
+    \begin{onlyenv}<4>
+      \node[annot,scale=1.2,red] at ($(arr)+(4,1)$) (warn) {
+        \textbf{Warning}: Does not say anything about the arguments!!
       };
+      \draw[overlay,link,<-,red] (arr.north) to[out=70,in=-150] (warn.-175) ;
     \end{onlyenv}
@@
 
-  Borrows allow to lend usage of something to someone else.
+  A borrow is a temporary loan of a resource @a@
   \begin{itemize}
-  \item \textbf{Shared} borrows \tikzcoord{UX1}@&a@\tikzcoord{UX2} which are Unrestricted (@un@)
-  \item \textbf{Exclusive} borrows \tikzcoord{AX1}@&!a@\tikzcoord{AX2} which are Affine (@aff@)
+  \item
+    \textbf{Shared} borrows \tikzcoord{UX1}@&a@\tikzcoord{UX2} are for observing the resource (Unrestricted -- @un@)
+    % \textbf{Shared} borrows \tikzcoord{UX1}@&a@\tikzcoord{UX2} which are Unrestricted (@un@)
+  \item
+    \textbf{Exclusive} borrows \tikzcoord{AX1}@&!a@\tikzcoord{AX2} are for modifying the resource (Affine -- @aff@)
+    % \textbf{Exclusive} borrows \tikzcoord{AX1}@&!a@\tikzcoord{AX2} which are Affine (@aff@)
   \end{itemize}
@@
   \end{onlyenv}  
-  \visible<3>{There are rules to respect!}
-
   \begin{tikzpicture}[remember picture,overlay]
@@
 \begin{onlyenv}<1>
-  \textbf{Rule 1}: Cannot use a borrow and the thing itself simultaneously
+  \textbf{Rule 1}: Cannot use a borrow and the resource itself simultaneously
 \begin{minted}[texcomments=true]{OCaml}
@@
   \visible<5>{%
-    We use \textbf{indexed} kinds to check that borrows do not escape!
+    \textbf{Indexed} kinds ensure that borrows do not escape!
 
@@
 
-  Inference proceeds in the following steps:
-
   \begin{enumerate}
-  \item Infer where regions goes
-  \item Turn the type, linearity and escape checks into constraints
+  \item Elaborate regions
+  \item Generate constraints for type, linearity and escape checks
   \item Solve the constraints
@@
 
-      \draw[link,->] (S1) to node[midway,above,sloped] {1. Infer regions} (S2) ;
+      \draw[link,->] (S1) to node[midway,above,sloped] {1. Elaborate regions} (S2) ;
 
@@
         \draw[link,->] (S2) to
-        node[midway,right,text width=3cm] {2. Convert\\ into constraints}
+        node[midway,right,text width=3cm] {2. Generate\\ constraints}
         (S3);
@@
   \only<2>{%
-    2. Convert into a custom constraint system based on HM(X).
+    2. Generate custom constraints based on HM(X).
   }%
@@
   \frametitle{The theory -- Summary}
-  In the paper, the theoretical contributions:
 

### 2020-08-06 kindly/slides-icfp.tex
Polish some bits.

@@
 (* ... *)
-print_string (Dbm.find &gradeDB "literature") (* \N compile-time error! *)
+print_int (Dbm.find &gradeDB "literature") (* \N compile-time error! *)
 \end{minted}
@@
 
+\begin{frame}{Table of contents}
+  \setbeamertemplate{section in toc}[sections numbered]
+  \tableofcontents[hideallsubsections,currentsection]
+\end{frame}
+
 \begin{frame}[t,fragile]
@@
 
+\begin{frame}{Table of contents}
+  \setbeamertemplate{section in toc}[sections numbered]
+  \tableofcontents[hideallsubsections,currentsection]
+\end{frame}
+
 \begin{frame}[t,fragile]{Borrows}
@@
   \item
-    \textbf{Shared} borrows \tikzcoord{UX1}@&a@\tikzcoord{UX2} are for observing the resource (Unrestricted -- @un@)
+    \textbf{Shared} borrows \tikzcoord{UX1}@&a@\tikzcoord{UX2} are for observing the resource
     % \textbf{Shared} borrows \tikzcoord{UX1}@&a@\tikzcoord{UX2} which are Unrestricted (@un@)
   \item
-    \textbf{Exclusive} borrows \tikzcoord{AX1}@&!a@\tikzcoord{AX2} are for modifying the resource (Affine -- @aff@)
+    \textbf{Exclusive} borrows \tikzcoord{AX1}@&!a@\tikzcoord{AX2} are for modifying the resource
     % \textbf{Exclusive} borrows \tikzcoord{AX1}@&!a@\tikzcoord{AX2} which are Affine (@aff@)
@@
     \begin{onlyenv}<2->
-      \node[code,color=yellow,fit=(A1) (A2)] {};
-      \node[code,color=red,fit=(U1) (U2)] {};
-      \node[code,color=red,fit=(U3) (U4)] {};
+      \node[code,color=yellow,fit=(A1) (A2)] (A) {};
+      \node[code,color=red,fit=(U1) (U2)] (UA) {};
+      \node[code,color=red,fit=(U3) (U4)] (UB) {};
+      \node[annot] at ($(UA)+(3,1)$) (un) { Unrestricted -- @un@ } ;
+      \draw[link] (un) to[out=180,in=80] (UA) ;
+      \draw[link] (un) to[out=0,in=100] (UB) ;
+      \node[annot] at ($(A)+(2,-1)$) (aff) { Affine -- @aff@ } ;
+      \draw[link] (aff) to[out=180,in=-80] (A) ;
     \end{onlyenv}
@@
 
-\begin{frame}{Inference and constraints}
+\begin{frame}{Table of contents}
+  \setbeamertemplate{section in toc}[sections numbered]
+  \tableofcontents[hideallsubsections,currentsection]
+\end{frame}
 
-  \begin{enumerate}
-  \item Elaborate regions
-  \item Generate constraints for type, linearity and escape checks
-  \item Solve the constraints
-  \item Simplify the resulting types
-  \end{enumerate}
-\end{frame}
+% \begin{frame}{Inference and constraints}
+
+%   \begin{enumerate}
+%   \item Elaborate regions
+%   \item Generate constraints for type, linearity and escape checks
+%   \item Solve the constraints
+%   \item Simplify the resulting types
+%   \end{enumerate}
+% \end{frame}
 
@@
 
-      \draw[link,->] (S1) to node[midway,above,sloped] {1. Elaborate regions} (S2) ;
+      \draw[link,->] (S1) to node[midway,above,sloped] {1. Elaborate} (S2) ;
+      \draw[link,->] (S1) to node[midway,below,sloped] {regions} (S2) ;
 

### 2022-12-07 kindly/split-jam.sh
sync

@@ -1,2 +1,2 @@
-pdfjam -o paper.pdf main.pdf '1-29'
-pdfjam -o appendix.pdf main.pdf '30-'
+pdfjam -o paper.pdf main.pdf '1-30'
+pdfjam -o appendix.pdf main.pdf '31-'

### 2022-12-07 kindly/slides-icfp.tex
Remove the date.
slides v2022

@@
 \usepackage[utf8]{inputenc}
-\usepackage[french, english]{babel}
-\selectlanguage{english}
+%\usepackage[french, english]{babel}
+%\selectlanguage{english}
 
@@
 \usepackage{xspace} % Useful for macros
-\usepackage{natbib}% Good citations and bibliography
-\usepackage{mathpartir} % Syntax trees
+%\usepackage{natbib}% Good citations and bibliography
+%\usepackage{mathpartir} % Syntax trees
 \usepackage{colortbl}
@@
 
-\bibliographystyle{ACM-Reference-Format}
-\citestyle{acmauthoryear}
+% \bibliographystyle{ACM-Reference-Format}
+% \citestyle{acmauthoryear}
 \usepackage{appendixnumberbeamer}
@@
 \title{Kindly Bent To Free Us}
-\author{\textbf{Gabriel \textsc{Radanne}}
-  \and Hannes \textsc{Saffrich}
-  \and Peter \textsc{Thiemann}}
+\author{{Gabriel {Radanne}}
+  \and Hannes {Saffrich}
+  \and \textbf{Peter {Thiemann}}}
+\date{ICFP 2020}
 
@@
 \begin{minted}{OCaml}
-let gradeDB : database = Dbm.opendbm "gradeDB" ... in
-Dbm.add gradeDB "math" 42;
+let gradeKV : kv_store = KVs.opendbm "grades" ... in
+KVs.add gradeKV "math" 42;
 (* ... *)
-Dbm.close gradeDB;
+KVs.close gradeKV;
 (* ... *)
-print_int (Dbm.find gradeDB "literature") (* run-time error! *)
+print_int (KVs.find gradeKV "literature") (* run-time error! *)
 \end{minted}
@@
 \begin{minted}[texcomments=true,escapeinside=\#\#]{OCaml}
-let gradeDB : #\tikzcoord{db1}#database#\tikzcoord{db2}# = Dbm.opendbm "gradeDB" ... in
-Dbm.add #\tikzcoord{A1}#&!gradeDB#\tikzcoord{A2}# #\tikzcoord{string1}#"math"#\tikzcoord{string2}# 42;
+let gradeKV : #\tikzcoord{db1}#kv_store#\tikzcoord{db2}# = KVs.opendbm "grades" ... in
+KVs.add #\tikzcoord{A1}#&!gradeKV#\tikzcoord{A2}# #\tikzcoord{string1}#"math"#\tikzcoord{string2}# 42;
 (* ... *)
-Dbm.close gradeDB;
+KVs.close gradeKV;
 (* ... *)
-print_int (Dbm.find #\tikzcoord{U1}#&gradeDB#\tikzcoord{U2}# "literature") (* \N compile-time error! *)
+print_int (KVs.find #\tikzcoord{U1}#&gradeKV#\tikzcoord{U2}# "literature") (* \N compile-time error! *)
 \end{minted}
@@
 \begin{minted}[texcomments=true,escapeinside=\#\#]{OCaml}
-let gradeDB #\tikzcoord{eq1}#=#\tikzcoord{eq2}# Dbm.opendbm "gradeDB" ... in
-Dbm.add &!gradeDB "math" 42;
+let gradeKV #\tikzcoord{eq1}#=#\tikzcoord{eq2}# KVs.opendbm "grades" ... in
+KVs.add &!gradeKV "math" 42;
 (* ... *)
-Dbm.close gradeDB;
+KVs.close gradeKV;
 (* ... *)
-print_int (Dbm.find &gradeDB "literature") (* \N compile-time error! *)
+print_int (KVs.find &gradeKV "literature") (* \N compile-time error! *)
 \end{minted}
@@
       ,annot,scale=1.2] at (5,-0.7) {
-        Kinds determine usage
+        Kinds restrict usage
         % \begin{itemize}
@@
       +(0,1) node[annot,anchor=south]
-      {@type database : lin (* Linear kind *)@};
+      {@type kv_store : lin (* Linear kind *)@};
       \draw[overlay,link] (string) to[out=-30,in=180]
@@
 
-  \textbf{Kinds determine usage:}
+  \textbf{Kinds restrict usage:}
   \begin{itemize}
@@
   \item Unrestricted (@un@): Used arbitrarily many time $[0-\infty]$
+  \item<2-> Kinds can be polymorphic!
   \end{itemize}
@@
 \begin{minted}[texcomments=true]{OCaml}
-type database : lin
+type kv_store : lin
 type string : un
@@
 \begin{minted}[texcomments=true,escapeinside=\#\#]{OCaml}
-let gradeDB = Dbm.open ...
+let gradeKV = KVs.open ...
 
@@
   printf "Closing: %s" msg;
-  Dbm.close #\tikzcoord{db1}#gradeDB#\tikzcoord{db2}#
+  KVs.close #\tikzcoord{db1}#gradeKV#\tikzcoord{db2}#
 \end{minted}
@@
       \node[annot,scale=1.2,red] at ($(arr)+(4,1)$) (warn) {
-        \textbf{Warning}: Does not say anything about the arguments!!
+        \textbf{Warning}: Does not say anything about the arguments
       };
@@
 
-  A borrow is a temporary loan of a resource @a@
+  Let @a@ be a (linear or affine) resource. 
+  A \emph{borrow} is a temporary loan of @a@.
   \begin{itemize}
@@
 let avg =
-  (Dbm.find #\tikzcoord{U1}#&gradeDB#\tikzcoord{U2}# "math" + Dbm.find #\tikzcoord{U3}#&gradeDB#\tikzcoord{U4}# "compsci") / 2
+  (KVs.find #\tikzcoord{U1}#&gradeKV#\tikzcoord{U2}# "math" + KVs.find #\tikzcoord{U3}#&gradeKV#\tikzcoord{U4}# "compsci") / 2
   (* \Y Multiple shared borrows *)
 in
-Dbm.add #\tikzcoord{A1}#&!gradeDB#\tikzcoord{A2}# "average" avg (* \Y One exclusive borrow *)
+KVs.add #\tikzcoord{A1}#&!gradeKV#\tikzcoord{A2}# "average" avg (* \Y One exclusive borrow *)
 \end{minted}
@@
 % \begin{minted}[texcomments=true]{OCaml}
-% let gradeDB = ... in
+% let gradeKV = ... in
 % let avg =
-%  let grade topic = Dbm.find &gradeDB topic in (* \Y Capture *)
+%  let grade topic = KVs.find &gradeKV topic in (* \Y Capture *)
 %  (grade "math" + grade "compsci") / 2
 % in
-% Dbm.add &!gradeDB "average" avg (* \Y Exclusive borrow *)
+% KVs.add &!gradeKV "average" avg (* \Y Exclusive borrow *)
 % \end{minted}
@@
 \begin{minted}[texcomments=true]{OCaml}
-let gradeDB = ... in
-f (gradeDB, &gradeDB) (* \N Conflicting use and borrow! *)
+let gradeKV = ... in
+f (gradeKV, &gradeKV) (* \N Conflict: direct use and borrow! *)
 \end{minted}
@@
 \begin{minted}[texcomments=true]{OCaml}
-let gradeDB = ... in
-f (&!gradeDB, &gradeDB) (* \N Conflicting borrows! *)
+let gradeKV = ... in
+f (&!gradeKV, &gradeKV) (* \N Conflicting borrows! *)
+\end{minted}
+  Two shared borrows are ok
+\begin{minted}[texcomments=true]{OCaml}
+let gradeKV = ... in
+f (&gradeKV, &gradeKV) (* \Y two shared borrows *)
 \end{minted}
@@
 let f () = 
-  let gradeDB = ... in
-  #\tikzcoord{R1}#let x = (&gradeDb, "mygrades") in#\tikzcoord{R2}#
+  let gradeKV = ... in
+  #\tikzcoord{R1}#let x = (&gradeKV, "mygrades") in#\tikzcoord{R2}#
+  let clean_up = KVs.close gradeKV in
   #\tikzcoord{R3}#x#\tikzcoord{R4}#
@@
 let f () = 
-  let gradeDB = ... in
-  #\tikzcoord{R1}#{| let x = (#\tikzcoord{B1}#&gradeDb#\tikzcoord{B2}#, "mygrades") in#\tikzcoord{R2}#
+  let gradeKV = ... in
+  #\tikzcoord{R1}#{| let x = (#\tikzcoord{B1}#&gradeKV#\tikzcoord{B2}#, "mygrades") in#\tikzcoord{R2}#
+  let clean_up = KVs.close gradeKV in
   #\tikzcoord{R3}#x |}#\tikzcoord{R4}#
@@
     \begin{onlyenv}<4->
-      \node[code,rounded corners=0,color=red,fit=(R1) (R2)] {};
+      \node[code,rounded corners=0,color=red,fit=(R1) (R2) (R4)] {};
       \node[code,rounded corners=0,minimum height=15pt,yshift=1pt,color=red,fit=(R3) (R4)] {};
@@
       \draw[link,<-] (B.10) to[in=180,out=40] ++(0.5,0.5) node[annot, anchor=west] {
-        @&gradeDb : &(database, un_2)@
+        @&gradeDb : &(kv_store, un_2)@
       };
@@
 
-%   {\Large The \texttt{Dbm} API:}
+%   {\Large The \texttt{KVs} API:}
 % \begin{minted}[texcomments=true]{OCaml}
-% type database : lin
-% val find : &(database,'k) -> string -{'k}> int
-% val add : &!(database,'k) -> string -{'k}> int -{'k}> unit
+% type kv_store : lin
+% val find : &(kv_store,'k) -> string -{'k}> int
+% val add : &!(kv_store,'k) -> string -{'k}> int -{'k}> unit
 % \end{minted}
@@
 
-  The @Database@ API:
+  The @Kv_Store@ API:
 \begin{minted}[texcomments=true,escapeinside=\#\#]{OCaml}
-type database : lin
-val find : &(database,'k) -> string #$\tarr{\texttt{'k}}$# int
-val add : &!(database,'k) -> string #$\tarr{\texttt{'k}}$# int #$\tarr{\texttt{'k}}$# unit
+type kv_store : lin
+val find : &(kv_store,'k) -> string #$\tarr{\texttt{'k}}$# int
+val add : &!(kv_store,'k) -> string #$\tarr{\texttt{'k}}$# int #$\tarr{\texttt{'k}}$# unit
 \end{minted}
@@
 \begin{minted}[texcomments=true]{OCaml}
-let gradeDB = ... in
+let gradeKV = ... in
 let avg =
-  (Dbm.find &gradeDB "math" + Dbm.find &gradeDB "compsci") / 2
+  (KVs.find &gradeKV "math" + KVs.find &gradeKV "compsci") / 2
 
 in
-Dbm.add &!gradeDB "average" avg
+KVs.add &!gradeKV "average" avg
 \end{minted}
@@
 \begin{minted}[texcomments=true]{OCaml}
-let gradeDB = ... in
+let gradeKV = ... in
 let avg =
-  let grade subject = Dbm.find &gradeDB subject in (* \Y Capture *)
+  let grade subject = KVs.find &gradeKV subject in (* \Y Capture *)
   (grade "math" + grade "compsci") / 2
 in
-Dbm.add &!gradeDB "average" avg
+KVs.add &!gradeKV "average" avg
 \end{minted}
@@
 \begin{minted}[texcomments=true]{OCaml}
-let gradeDB = ... in
+let gradeKV = ... in
 let avg =
-  let grade = Dbm.find &gradeDB in (* \Y Partial application *)
+  let grade = KVs.find &gradeKV in (* \Y Partial application *)
   (grade "math" + grade "compsci") / 2
 in
-Dbm.add &!gradeDB "average" avg
+KVs.add &!gradeKV "average" avg
 \end{minted}
@@
 let average db subjects = 
-  List.map (Dbm.find db) subjects / List.length subjects
+  List.map (KVs.find db) subjects / List.length subjects
 let main () =
-  let gradeDB = ... in
-  let avg = average &gradeDB ["math";"compsci";...] in
-  Dbm.add &!gradeDB "average" avg
+  let gradeKV = ... in
+  let avg = average &gradeKV ["math";"compsci";...] in
+  KVs.add &!gradeKV "average" avg
 \end{minted}
@@
 let average db subjects #\tikzcoord{eq1}#=#\tikzcoord{eq2}#
-  List.map (Dbm.find db) subjects / List.length subjects
+  List.map (KVs.find db) subjects / List.length subjects
 let main () =
-  let gradeDB = ... in
-  let avg = #\tikzcoord{R1}#{| average &gradeDB ["math";"compsci";...] |}#\tikzcoord{R2}# in
-  #\tikzcoord{R3}#{| Dbm.add &!gradeDB "average" avg |}#\tikzcoord{R4}#
+  let gradeKV = ... in
+  let avg = #\tikzcoord{R1}#{| average &gradeKV ["math";"compsci";...] |}#\tikzcoord{R2}# in
+  #\tikzcoord{R3}#{| KVs.add &!gradeKV "average" avg |}#\tikzcoord{R4}#
 \end{minted}
@@
 % let average db subjects = 
-%   List.map (Dbm.find db) subjects / List.length subjects
+%   List.map (KVs.find db) subjects / List.length subjects
 
 % let main () =
-%   let gradeDB = ... in
-%   let avg = average &gradeDB ["math";"compsci";...] in
-%   Dbm.add &!gradeDB "average" avg
+%   let gradeKV = ... in
+%   let avg = average &gradeKV ["math";"compsci";...] in
+%   KVs.add &!gradeKV "average" avg
 % \end{minted}
@@
 % let average db subjects =
-%   List.map (Dbm.find db) subjects / List.length subjects
+%   List.map (KVs.find db) subjects / List.length subjects
 
 % let main () =
-%   let gradeDB = ... in
-%   let avg = #\tikzcoord{R1}#{| average &gradeDB ["math";"compsci";...] |}#\tikzcoord{R2}# in
-%   #\tikzcoord{R3}#{| Dbm.add &!gradeDB "average" avg |}#\tikzcoord{R4}#
+%   let gradeKV = ... in
+%   let avg = #\tikzcoord{R1}#{| average &gradeKV ["math";"compsci";...] |}#\tikzcoord{R2}# in
+%   #\tikzcoord{R3}#{| KVs.add &!gradeKV "average" avg |}#\tikzcoord{R4}#
 % \end{minted}
@@
 let average db subjects = 
-  List.map (Dbm.find db) subjects / List.length subjects
+  List.map (KVs.find db) subjects / List.length subjects
 
 let main () =
-  let gradeDB = ... in
-  let avg = average &gradeDB ["math";"compsci";...] in
-  Dbm.add &!gradeDB "average" avg
+  let gradeKV = ... in
+  let avg = average &gradeKV ["math";"compsci";...] in
+  KVs.add &!gradeKV "average" avg
 \end{minted}
@@
 let average db subjects =
-  List.map (Dbm.find db) subjects / List.length subjects
+  List.map (KVs.find db) subjects / List.length subjects
 
 let main () =
-  let gradeDB = ... in
-  let avg = #\tikzcoord{R1}#{| average &gradeDB ["math";"compsci";...] |}#\tikzcoord{R2}# in
-  #\tikzcoord{R3}#{| Dbm.add &!gradeDB "average" avg |}#\tikzcoord{R4}#
+  let gradeKV = ... in
+  let avg = #\tikzcoord{R1}#{| average &gradeKV ["math";"compsci";...] |}#\tikzcoord{R2}# in
+  #\tikzcoord{R3}#{| KVs.add &!gradeKV "average" avg |}#\tikzcoord{R4}#
 \end{minted}
@@
         \node[annot,text width=10.5cm,scale=0.7] at (0.7,1) (S4) {
-          @val average : 'k_2 . &(database,'k) -{'k_2}> string list -{'k}> int list@
+          @val average : 'k_2 . &(kv_store,'k) -{'k_2}> string list -{'k}> int list@
         };
@@
         \node[annot,text width=9.7cm,scale=0.7] at (0.7,-1) (S5) {
-          @val average : &(database,'k) -> string list -{'k}> int list@
+          @val average : &(kv_store,'k) -> string list -{'k}> int list@
         };
@@
 
-\bibliography{biblio}
+% \bibliography{biblio}
 
